### 2021-12-04 proposal/research_proposal.Rmd
updated before_body.tex and set up template for research proposal

---
title: "Overlooked dynamics of algorithmic recourse"
subtitle: "Research Proposal"
author:
- "Patrick Altmeyer (p.altmeyer@tudelft.nl)"
date: "`r format(Sys.Date(), '%B, %Y')`"
output: 
  bookdown::pdf_document2: 
    toc: false
    includes:
      in_header: ../tex/preamble.tex
  bookdown::html_document2: default
fontsize: 12pt
bibliography: ../bib.bib
link-citations: true
abstract: "..."
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  out.width = 450,
  fig.align = "center"
)
```

### 2021-12-07 work/preprocessing.Rmd
work on endogenous shifts notebook

---
title: "Data preprocessing"
author:
- "Patrick Altmeyer (p.altmeyer@tudelft.nl)"
date: "`r format(Sys.Date(), '%B, %Y')`"
output: 
  bookdown::pdf_document2: 
    toc: false
    includes:
      in_header: ../tex/preamble.tex
  bookdown::html_document2: default
fontsize: 12pt
bibliography: ../bib.bib
link-citations: true
abstract: "..."
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  out.width = 450,
  fig.align = "center"
)
```

# German credit data

```{r}
library(data.table)
dat <- read.table("../data/raw/SouthGermanCredit/SouthGermanCredit.asc", header=TRUE) 
source("../data/raw/SouthGermanCredit/read_SouthGermanCredit.R") # apply preprocessing 
dt <- data.table(dat)
```

```{r}
library(mltools)
X_scm <- c("personal_status_sex", "age", "amount", "duration")
y <- "credit_risk"
vars <- c(y,X_scm)
dt_scm <- copy(dt[,.SD,.SDcols=vars])
dt_scm[,credit_risk:=as.numeric(credit_risk)-1] # turn into binary numerical
dt_scm <- one_hot(dt_scm) # one-hot encode personal_status_sex
setcolorder(dt_scm, c("credit_risk", "age", "amount", "duration"))
setnames(dt_scm, colnames(dt_scm), c("y", "x1", "x2", "x3", sprintf("i%i", 1:4)))
fwrite(dt_scm,"../data/credit.csv")
```

### 2021-12-07 julia/generate_recourse.jl

@@ -1,6 +1,7 @@
 struct Recourse
     x_cf::Vector{Float64}
     path::Matrix{Float64}
+    valid::Bool
 end;
 
 # ------------------------- Wachter et al (2018) -------------------------
@@ -8,7 +9,8 @@ function gradient_cost(x_f, x_cf)
     (x_f - x_cf) ./ norm(x_f - x_cf)
 end;
 
-function generate_recourse_wachter(x, gradient, w, target; α=1, τ=1e-5, λ=0.25, gradient_cost=gradient_cost, T=1000)
+function generate_recourse_wachter(x, gradient, classifier, target; α=1, τ=1e-5, λ=0.1, gradient_cost=gradient_cost, T=1000)
+    w = coef(classifier)
     D = length(x) # input dimension
     path = reshape(x, 1, D) # storing the path
     # Initialize:
@@ -32,23 +34,58 @@ function generate_recourse_wachter(x, gradient, w, target; α=1, τ=1e-5, λ=0.2
         converged = convergence_condition(vcat(1,x_cf), gradient, w, target, τ) # check if converged
         path = vcat(path, reshape(x_cf, 1, D))
     end
+    # Validity:
+    valid = predict(classifier, vcat(1,x_cf), false)[1] == target * 1.0
     # Output:
-    recourse = Recourse(x_cf, path) 
+    recourse = Recourse(x_cf, path, valid) 
+end;
+
+# ------------------------- Upadhyay et al (2021) -------------------------
+function gradient_cost(x_f, x_cf)
+    (x_f - x_cf) ./ norm(x_f - x_cf)
+end;
+
+function generate_recourse_roar(x, gradient, classifier, target; α=1, τ=1e-5, λ=0.25, gradient_cost=gradient_cost, T=1000)
+    w = coef(classifier)
+    D = length(x) # input dimension
+    path = reshape(x, 1, D) # storing the path
+    # Initialize:
+    x_cf = copy(x) # start from factual
+    t = 1 # counter
+    function convergence_condition(x_cf, gradient, w, target, tol)
+        all(gradient(x_cf,w,target) .<= τ)
+    end
+    converged = convergence_condition(vcat(1,x_cf), gradient, w, target, τ)
+    # Recursion:
+    while !converged && t < T 
+        𝐠_t = gradient(vcat(1,x_cf),w,target)[2:length(w)] # compute gradient
+        𝐠_cost_t = gradient_cost(vcat(1,x),vcat(1,x_cf))[2:length(w)] # compute gradient of cost function
[…]

### 2022-01-11 proposal/research_proposal.Rmd
work on endogenous shifts notebook
after work on delft device

@@
 ```
+
+# Introduction
+
+In their NeurIPS 2021 paper @upadhyay2021towards demonstrate that existing approaches to algorithmic recourse are not robust to exogenous domain and model shifts.
+A related and so far unexplored idea is that recourse itself may lead to endogenous domain and model shifts, if a substantial share of individuals acts on it.
+We believe that this can have implications on the validity and cost of algorithmic recourse that have so far been overlooked.
+
+# Related work
+
+## @wachter2017counterfactual
+
+## @schut2021generating
+
+## @upadhyay2021towards
+
+## Other
+
+- @joshi2019towards
+- @ustun2019actionable
+- @karimi2021algorithmic
+
+# Preliminary findings
+
+## Synthetic data
+
+## German credit data

### 2022-01-12 paper/paper.Rmd
Almost finished tutorial on loss functions

---
title: "Overlooked dynamics of algorithmic recourse"
author:
- "Patrick Altmeyer (p.altmeyer@tudelft.nl)"
date: "`r format(Sys.Date(), '%B, %Y')`"
output: 
  bookdown::pdf_document2: 
    toc: false
    includes:
      in_header: ../tex/preamble.tex
      before_body: ../tex/before_body.tex
  bookdown::html_document2: default
fontsize: 12pt
bibliography: ../bib.bib
link-citations: true
abstract: "..."
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  out.width = 450,
  fig.align = "center"
)
```

```{r introduction, child="introduction.Rmd"}
```

\FloatBarrier
\pagebreak

# References {-}

<div id="refs"></div>

\pagebreak

```{r annex, child="appendix.Rmd"}
```

### 2022-02-03 students/project_counterfactual_generators.qmd
synthetic and boston housing experiments done. Finished student project proposal.

---
title: "Endogenous model shifts in algorithmic recourse"
author: "Patrick Altmeyer"
date: "February 03, 2022"
format: gfm
bibliography: ../bib.bib
---

Responsible Professor: Dr. Cynthia Liem
Supervisor: Patrick Altmeyer

## Project overview

Algorithmic recourse is an approach towards not only explaining black-box machine learning algorithms, but provide subjects 
who have been adversely affected by such with a realistic set of actions they can take in order to revise their outcome. 
The approach has been quickly growing in popularity during recent years, perhaps because the underlying tend to be methodologies tend to be intuitive and easy to implement. This project aims to shed some light on what happens *after* recourse is actually implemented. In particular, we want to investigate if and how the resulting domain shifts lead to model shifts and compare outcomes for different recourse generators.

## Motivation and background

The topic is also of great importance to society since automated decision-making through black-box algorithm affects all of us - inlcuding you! Suppose, for example, that upon graduation you apply for a job at a large coorperation. Unfortunately, you soon receive an email stating that you have not "met the short-listing criteria and no further feedback can be provided". Unbeknown to you, the company has used an algorithm to decide who makes it onto the short list. The HR teams does not fully understand the inner workings of the algorithm, so they genuinely cannot provide you with feedback.

Algorithmic recourse aims to resolve these types of issues. But existing work has paid little attention to the dynamics of 
recourse. The first paper to consider the problem that algorithms are updated in response to (exogenous) domain shifts was presented at NeurIPS in December 2021 [@upadhyay2021towards]. The authors propose a recourse generator that is robust to "small (and not drastic)" shifts. But rather than small shifts, we find that the actual implementation of recourse can lead to potentially quite large domain shifts and consequently model shifts. The animation below illustrates this: after a small share of individuals moves across the decision boundary, the classifier is updated in response. Evidently, the model shifts are quite significant in this toy example. Below we refer to these types of shifts as *endogenous* shifts.

We think that endogenous shifts deserve attention for various reasons:

1. Our initial experimental results suggests that even generators that aim to produce realisitic counterfactuals can induce significant domain and model shifts.
2. Model shifts in response to some individuals' recourse may lead to undesired consequences (label switch) for other individuals.
3. Individuals that have implemented recourse may end up forming their own distinct cluster in the feature domain. The original classifier could therefore be adapted to distinguish "recourse" indviduals in the target class from those that were always in the target class.

![](../work/www/synthetic/bayesian.gif)

## Research questions

Collectively, students will discuss and agree on an answer to the following question:

- How can we quantify endogenous domain and model shifts?

Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts and quantify their magnitude.

Collectively, the students will finally tackle the following questions:

- Does the magnitude of induced model shifts vary by recourse generator? 
- If so, what factors might be playing a role here?
- Based on your findings, what appear to be good ways to deal mitigate endogenous shifts?

While the individual part should really be done individually, do keep in mind that you will end up comparing the results of your experiments and you should therefore all generate the same, comparable output.

## Methodology

Overall, this will be a project focused on benchmarking and experimental design: you will learn about how to use existing tools for benchmarking to asses outcomes from your own experiments.

For a gentle introduction to the topic of algorithmic recourse you can check out this [primer](https://towardsdatascience.com/individual-recourse-for-black-box-models-5e9ed1e4b4cc) written by Patrick as part of his PhD application. To familiarize yourself further with the topic and [CARLA](https://github.com/carla-recourse/CARLA) you should all read this [paper](https://arxiv.org/pdf/2108.00783.pdf).

You are free to choose any of the generators already implemented [CARLA](https://github.com/carla-recourse/CARLA), but below are five suggestions:

1. [Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)
2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf)
3. [CEM](https://arxiv.org/pdf/1802.07623.pdf)
4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf)
5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf)

Since the generators are all conveniently implemented in the same library, running the experiment should require roughly the 
same effort for all of them. While you are not expected to fully understand in detail the methodology underlying your
chosen generator, you should study the relevant paper and ask yourself: what makes your generator particular? What assumptions do the authors make? How do you expect these particularities to affect recourse outcome with respect to endogenous shifts?

**Experiment**

The idea is that you replicate a variation of the following experiement already implemented by Patrick:

1. Train an algorithm $\mathcal{M}$ for a binary classification task. 
2. Determine a target class. Generate recourse for a share $\mu$ of randomly selected individuals in the other class to revise their label (i.e. move to the target class). Store some of the conventional benchmark measures readily implemented in CARLA (cost, success rate, ...)
3. Implement recourse for those indviduals and quantify the domain shift.
4. Retrain $\mathcal{M}$ and quantify the model shift.
5. Repeat steps 1-4 for $K$ rounds.

You are free to either follow Patrick's recipe exactly or tweak the various parameters that determine the experiemntal design as you see fit. Make sure that you end up running the exact same experiment though, in order for the results to be comparable.

**Hypothesis**

- I expect to find that generators that focus on creating realisitic counterfactuals in high-density regions of the feature domain (REVISE, CEM-VAE) tend to be associated with endogenous shifts of a smaller magnitude compared to generators that do not specifically address this (CEM, Wachter, AR-LIME).

## Relation between sub-projects

The students will collectively design and establish an evaluation benchmark, while each focusing on a particular framework implementation to ultimately evaluate in the benchmark. Researching the intricacies of each framework will lead to clearly individual contributions, while part of the relevant literature is shared, and the final comparison connects the individual insights.

## Responsible Engineering / Computer Science / Research

As noted above, algorithmic recourse was designed in the context of responsible machine learning. Work on this topic should foster an awareness that automated decision-making can have detrimental consequences for individuals and that many researchers are working on approaches to deal with some of those consequences. Comparing different approaches to the issue will also make it clear that in the context of research there is rarely (if ever) one right solution. More to the point of this particular research project, solutions to one particular issue usually open up new questions.

## Possible publication

In 2021 there was an [ICML workshop](https://icml.cc/Conferences/2021/ScheduleMultitrack?event=8363) on algorithmic recourse. Should this workshop take place again, the results of this project could lead to a poster presentation. Later in the year NeurIPS is coming up, which in 2021 also attracted conference and workshop papers on algorithmic recourse and benchmarking.

## References

### 2022-02-10 students/project_counterfactual_generators.qmd
updated the student project proposal

@@
 Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts and quantify their magnitude.
+You will also each produce results for the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) and benchmark the results from both generators.
 
-Collectively, the students will finally tackle the following questions:
+Use the results to answer the following questions with respect to your chosen generator:
 
-- Does the magnitude of induced model shifts vary by recourse generator? 
+- Does the magnitude of induced model shifts differ compared to the baseline generator? 
 - If so, what factors might be playing a role here?
-- Based on your findings, what appear to be good ways to deal mitigate endogenous shifts?
+- Based on your findings, what appear to be good ways to mitigate endogenous shifts?
 
@@
 
-1. [Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)
+1. [FACE](https://arxiv.org/pdf/1909.09369.pdf)
 2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf)
@@
 
-Since the generators are all conveniently implemented in the same library, running the experiment should require roughly the 
-same effort for all of them. While you are not expected to fully understand in detail the methodology underlying your
+Since the generators are all conveniently implemented in the same library, running the experiment should require roughly the same effort for all of them.
+While you are not expected to fully understand in detail the methodology underlying your
 chosen generator, you should study the relevant paper and ask yourself: what makes your generator particular?
@@
 Determine a target class.
-Generate recourse for a share $\mu$ of randomly selected individuals in the other class to revise their label (i.e.
+Using your chosen generator and the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) generate recourse for a share $\mu$ of randomly selected individuals in the other class to revise their label (i.e.
 move to the target class).

### 2022-02-17 students/research-plan-counterfactual-explanations.qmd
updated student projects

--- 2022-02-10 students/project_counterfactual_generators.qmd
@@
 title: "Endogenous model shifts in algorithmic recourse"
-author: "Patrick Altmeyer"
-date: "February 03, 2022"
-format: gfm
+format: 
+    gfm:
+        wrap: none
 bibliography: ../bib.bib
+csl: https://raw.githubusercontent.com/citation-style-language/styles/master/elsevier-vancouver.csl
 ---
@@
 Supervisor: Patrick Altmeyer
-
-## Project overview
-
-Algorithmic recourse is an approach towards not only explaining black-box machine learning algorithms, but provide subjects 
-who have been adversely affected by such with a realistic set of actions they can take in order to revise their outcome. 
-The approach has been quickly growing in popularity during recent years, perhaps because the underlying tend to be methodologies tend to be intuitive and easy to implement.
-This project aims to shed some light on what happens *after* recourse is actually implemented.
-In particular, we want to investigate if and how the resulting domain shifts lead to model shifts and compare outcomes for different recourse generators.
-
-## Motivation and background
-
-The topic is also of great importance to society since automated decision-making through black-box algorithm affects all of us - inlcuding you!
-Suppose, for example, that upon graduation you apply for a job at a large coorperation.
-Unfortunately, you soon receive an email stating that you have not "met the short-listing criteria and no further feedback can be provided".
-Unbeknown to you, the company has used an algorithm to decide who makes it onto the short list.
-The HR teams does not fully understand the inner workings of the algorithm, so they genuinely cannot provide you with feedback.
-
-Algorithmic recourse aims to resolve these types of issues. But existing work has paid little attention to the dynamics of 
-recourse.
-The first paper to consider the problem that algorithms are updated in response to (exogenous) domain shifts was presented at NeurIPS in December 2021 [@upadhyay2021towards].
-The authors propose a recourse generator that is robust to "small (and not drastic)" shifts.
-But rather than small shifts, we find that the actual implementation of recourse can lead to potentially quite large domain shifts and consequently model shifts.
-The animation below illustrates this: after a small share of individuals moves across the decision boundary, the classifier is updated in response.
-Evidently, the model shifts are quite significant in this toy example.
-Below we refer to these types of shifts as *endogenous* shifts.
-
-We think that endogenous shifts deserve attention for various reasons:
-
-1. Our initial experimental results suggests that even generators that aim to produce realisitic counterfactuals can induce significant domain and model shifts.
-2. Model shifts in response to some individuals' recourse may lead to undesired consequences (label switch) for other individuals.
-3.
-Individuals that have implemented recourse may end up forming their own distinct cluster in the feature domain.
-The original classifier could therefore be adapted to distinguish "recourse" indviduals in the target class from those that were always in the target class.
-
-![](../work/www/synthetic/bayesian.gif)
 
@@
 
-## Methodology
+## Research Methods
 
@@
 
-1. [FACE](https://arxiv.org/pdf/1909.09369.pdf)
-2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf)
-3. [CEM](https://arxiv.org/pdf/1802.07623.pdf)
-4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf)
-5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf)
+1. [FACE](https://arxiv.org/pdf/1909.09369.pdf), [@poyiadzi2020face]
+2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf), [@joshi2019towards]
+3. [CEM](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
+4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
+5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf), [@ustun2019actionable]
 
@@
 
-- I expect to find that generators that focus on creating realisitic counterfactuals in high-density regions of the feature domain (REVISE, CEM-VAE) tend to be associated with endogenous shifts of a smaller magnitude compared to generators that do not specifically address this (CEM, Wachter, AR-LIME).
+I expect to find that generators that focus on creating realistic counterfactuals in high-density regions of the feature domain (REVISE, CEM-VAE) tend to be associated with endogenous shifts of a smaller magnitude compared to generators that do not specifically address this (CEM, Wachter, AR-LIME).
 

### 2022-04-12 dev/students/research-proposal-counterfactual-explanations.qmd
setting up for next steps, including AIES submission

--- 2022-02-10 students/project_counterfactual_generators.qmd
@@
 
-## Project overview
+### Background and motivation
 
@@
 In particular, we want to investigate if and how the resulting domain shifts lead to model shifts and compare outcomes for different recourse generators.
-
-## Motivation and background
 
@@
 
-## Research questions
+For a gentle introduction to the topic of algorithmic recourse you can check out this [primer](https://towardsdatascience.com/individual-recourse-for-black-box-models-5e9ed1e4b4cc) written by Patrick as part of his PhD application.
+To familiarize yourself further with the topic and [CARLA](https://github.com/carla-recourse/CARLA) you should all read this [paper](https://arxiv.org/pdf/2108.00783.pdf).
+
+### Research Questions for the Sub-Projects
 
@@
 
-Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts and quantify their magnitude.
+Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts - that is, provide and implement recourse - and quantify their magnitude.
 You will also each produce results for the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) and benchmark the results from both generators.
+
+You are free to choose any of the generators already implemented [CARLA](https://github.com/carla-recourse/CARLA), but below are five suggestions:
+
+1. [FACE](https://arxiv.org/pdf/1909.09369.pdf), [@poyiadzi2020face]
+2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf), [@joshi2019towards]
+3. [CEM](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
+4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
+5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf), [@ustun2019actionable]
 
@@
 
-## Methodology
+### Prerequisites
 
-Overall, this will be a project focused on benchmarking and experimental design: you will learn about how to use existing tools for benchmarking to asses outcomes from your own experiments.
+For this assignment, prior experience is Python is a requirement. A basic understanding of gradient-based optimization is helpful, but not a strict requirement.
 
-For a gentle introduction to the topic of algorithmic recourse you can check out this [primer](https://towardsdatascience.com/individual-recourse-for-black-box-models-5e9ed1e4b4cc) written by Patrick as part of his PhD application.
-To familiarize yourself further with the topic and [CARLA](https://github.com/carla-recourse/CARLA) you should all read this [paper](https://arxiv.org/pdf/2108.00783.pdf).
+### Q&A Session
 
-You are free to choose any of the generators already implemented [CARLA](https://github.com/carla-recourse/CARLA), but below are five suggestions:
+If you're interested in this project and/or have follow-up questions, please feel free to join us through [Zoom]() on one of these two sessions:
 
-1. [FACE](https://arxiv.org/pdf/1909.09369.pdf)
-2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf)
-3. [CEM](https://arxiv.org/pdf/1802.07623.pdf)
-4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf)
-5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf)
+-   Tuesday, March 8, 12:45-13:25 (Zoom [link](https://us04web.zoom.us/j/77787187346?pwd=QFBh8Lc9iliVRzSSa9Yz9XFwZzUrXM.1))
+-   Wednesday, March 9, 12:45-13:25 (Zoom [link](https://us04web.zoom.us/j/77845264541?pwd=QoHXseljKn_PMTDawWFrhHPX7AUaua.1))
 
-Since the generators are all conveniently implemented in the same library, running the experiment should require roughly the same effort for all of them.
-While you are not expected to fully understand in detail the methodology underlying your
-chosen generator, you should study the relevant paper and ask yourself: what makes your generator particular?
-What assumptions do the authors make?
-How do you expect these particularities to affect recourse outcome with respect to endogenous shifts?
-
-**Experiment**
-
-The idea is that you replicate a variation of the following experiement already implemented by Patrick:
-
-1. Train an algorithm $\mathcal{M}$ for a binary classification task. 
-2.
-Determine a target class.
-Using your chosen generator and the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) generate recourse for a share $\mu$ of randomly selected individuals in the other class to revise their label (i.e.
-move to the target class).
-Store some of the conventional benchmark measures readily implemented in CARLA (cost, success rate, ...)
-3. Implement recourse for those indviduals and quantify the domain shift.
-4. Retrain $\mathcal{M}$ and quantify the model shift.
-5. Repeat steps 1-4 for $K$ rounds.
-
-You are free to either follow Patrick's recipe exactly or tweak the various parameters that determine the experiemntal design as you see fit.
-Make sure that you end up running the exact same experiment though, in order for the results to be comparable.
-
-**Hypothesis**
-
-- I expect to find that generators that focus on creating realisitic counterfactuals in high-density regions of the feature domain (REVISE, CEM-VAE) tend to be associated with endogenous shifts of a smaller magnitude compared to generators that do not specifically address this (CEM, Wachter, AR-LIME).
-
-## Relation between sub-projects
-
-The students will collectively design and establish an evaluation benchmark, while each focusing on a particular framework implementation to ultimately evaluate in the benchmark.
-Researching the intricacies of each framework will lead to clearly individual contributions, while part of the relevant literature is shared, and the final comparison connects the individual insights.
-
-## Responsible Engineering / Computer Science / Research
-
-As noted above, algorithmic recourse was designed in the context of responsible machine learning.
-Work on this topic should foster an awareness that automated decision-making can have detrimental consequences for individuals and that many researchers are working on approaches to deal with some of those consequences.
-Comparing different approaches to the issue will also make it clear that in the context of research there is rarely (if ever) one right solution.
-More to the point of this particular research project, solutions to one particular issue usually open up new questions.
-
-## Possible publication
-
-In 2021 there was an [ICML workshop](https://icml.cc/Conferences/2021/ScheduleMultitrack?event=8363) on algorithmic recourse.
-Should this workshop take place again, the results of this project could lead to a poster presentation.
-Later in the year NeurIPS is coming up, which in 2021 also attracted conference and workshop papers on algorithmic recourse and benchmarking.
-
-## References
+### References

### 2022-04-13 paper/paper.tex
cant get documentclass to work for quarto

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Overlooked Dynamics in Algorithmic Recourse},
  pdfauthor={Patrick Altmeyer (p.altmeyer@tudelft.nl)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Packages:
\usepackage{placeins}
\usepackage{hyperref}
\usepackage{acmart}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\renewcommand*\contentsname{Table of contents}
\renewcommand*\listfigurename{List of Figures}
\renewcommand*\listtablename{List of Tables}
\renewcommand*\figurename{Figure}
\renewcommand*\tablename{Table}
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Overlooked Dynamics in Algorithmic Recourse}
\author{Patrick Altmeyer (p.altmeyer@tudelft.nl)}
\date{April, 2022}

\begin{document}
\maketitle
\begin{abstract}
\ldots{}
\end{abstract}

% Include before body: ----
\pagenumbering{gobble}

\newenvironment{acknowledgements}%
{\begin{center}%
\bfseries Acknowledgements
\end{center}}%
{\vfill\null}
\begin{acknowledgements}
...
\end{acknowledgements}
{\vfill\null}
\textbf{DISCLAIMER:}
This is a work-in-progress. Please do not distribute or share.
\pagebreak

\pagenumbering{arabic}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\ldots{}

\FloatBarrier
\pagebreak

References

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\pagebreak

\hypertarget{appendix-appendix}{%
\section{(APPENDIX) Appendix}\label{appendix-appendix}}

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\ldots{}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\ldots{}

\hypertarget{code}{%
\section{Code}\label{code}}

\ldots{}

\end{document}

### 2022-04-13 dev/work/boston_housing.qmd
moved to qmd docs

---
jupyter: julia-1.6
---

# Boston Housing

```{julia}
include("../src/load.jl")
using AlgorithmicRecourse, MLDatasets, Flux
using Plots, PlotThemes
theme(:juno)
using Logging
disable_logging(Logging.Info)
output_folder = "output/boston_housing"
www_folder = "www/boston_housing"
using DataFrames, CSV
using BSON: @save, @load
using Base.Filesystem: joinpath
```

## Training the classifier

```{julia}
using MLDatasets, Statistics
X = BostonHousing.features()
y = BostonHousing.targets()
y = Float64.(y .>= median(y)); # binary target
```

To start off we will just train a single neural network for the binary classification task.

```{julia}
# Prepare data and model:
using Random
Random.seed!(1234)
using StatsBase
dt = fit(ZScoreTransform, X, dims=2)
StatsBase.transform!(dt, X)
xs = Flux.unstack(X,2)
data = zip(xs,y)
nn = Models.build_model(input_dim=size(X)[1], n_hidden=100)
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
```

The model achieves decent training accuracy

```{julia}
run = false
if run
  # Train model:
  using Flux.Optimise: update!, ADAM
  using Statistics, StatsBase
  opt = ADAM()
  epochs = 100
  avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
  accuracy(data) = sum(map(d ->round.(Flux.σ.(nn(d[1]))) .== d[2], data))[1]/length(data)

  using Plots
  anim = Animation()
  avg_l = [avg_loss(data)]
  p1 = scatter( ylim=(0,avg_l[1]), xlim=(0,epochs), legend=false, xlab="Epoch", title="Average loss")
  acc = [accuracy(data)]
  p2 = scatter( ylim=(0.5,1), xlim=(0,epochs), legend=false, xlab="Epoch", title="Accuracy")

  τ = 1.0
  stopping_criterium_reached = accuracy(data) >= τ
  epoch = 1

  while epoch <= epochs && !stopping_criterium_reached
    for d in data
      gs = gradient(params(nn)) do
        l = loss(d...)
      end
      update!(opt, params(nn), gs)
    end
    avg_l = vcat(avg_l,avg_loss(data))
    plot!(p1, [0:epoch], avg_l, color=1)
    scatter!(p1, [0:epoch], avg_l, color=1)
    acc = vcat(acc,accuracy(data))
    plot!(p2, [0:epoch], acc, color=1)
    scatter!(p2, [0:epoch], acc, color=1)
    plt=plot(p1,p2, size=(600,300))
    frame(anim, plt)

    # Check if desired accuracy reached:
    stopping_criterium_reached = accuracy(data) >= τ

    epoch += 1

  end

  gif(anim, joinpath(www_folder, "single_nn.gif"), fps=10)

  using BSON: @save
  @save joinpath(output_folder, "nn.bson") nn

end
```

![](www/boston_housing/single_nn.gif)

Next we will build and train a deep ensemble.

```{julia}
opt = ADAM()
loss_type = :logitbinarycrossentropy
run = false
if run
    K = 50
    𝓜 = Models.build_ensemble(K,kw=(input_dim=size(X)[1], n_hidden=100));
    𝓜, anim = Models.forward(𝓜, data, opt, n_epochs=30, plot_every=10, loss_type=loss_type); # fit the ensemble
    Models.save_ensemble(𝓜, root=joinpath(output_folder, "ensemble")) # save to disk
    gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
end
```

![](www/boston_housing/ensemble_loss.gif)

## Experiment

```{julia}
using BSON: @load
@load joinpath(output_folder, "nn.bson") nn
𝑴ₙₙ = Models.FittedNeuralNet(nn, opt, loss);
```

Prepare ensemble for use with AlgorithmicRecourse.jl:

```{julia}
𝓜 = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
𝑴 = Models.FittedEnsemble(𝓜, opt, loss_type);
```

Prepare grid of variables for experiment:

```{julia}
# Variables:
μ = [0.01,0.1,0.25]
γ = [0.50,0.75,0.9]
grid_ = Experiments.GridVariables(μ, γ)
n_rounds = 10
target=1.0
T = 1000;
```

### Generic generator for MLP

```{julia}
# Experiment:
experiment = Experiments.Experiment(X,y,𝑴ₙₙ,target,grid_,n_rounds);
```

```{julia}
run = false
if run  
    Λ = [0.0, 0.1, 0.25, 0.5] 
    outcome = DataFrame()
    for λ in Λ
        generator = GenericGenerator(λ,0.1,1e-5,:logitbinarycrossentropy,nothing)
        outcome_λ, path = Experiments.run_experiment(experiment, generator, 5, T=T)
        outcome_λ = DataFrame(outcome_λ)
        insertcols!(outcome_λ, :λ => λ)
        outcome = vcat(outcome, outcome_λ)
    end
    CSV.write(joinpath(output_folder, "mlp_generic.csv"), outcome)
end
```

```{julia}
outcome = CSV.read(joinpath(output_folder, "mlp_generic.csv"), DataFrame)

function prepare_results(outcome; id_vars=[:t, :μ, :γ])
    validity = groupby(outcome, id_vars) |>
        gdf -> combine(gdf, :pct_valid .=> [mean, std] .=> [:mean, :std])
    validity[!,:ymin] = validity[!,:mean] - validity[!,:std]
    validity[!,:ymax] = validity[!,:mean] + validity[!,:std]

    cost = groupby(outcome, id_vars) |>
        gdf -> combine(gdf, :avg_cost .=> [mean, std] .=> [:mean, :std])
    cost[!,:ymin] = cost[!,:mean] - cost[!,:std]
    cost[!,:ymax] = cost[!,:mean] + cost[!,:std];

    return validity, cost
end

validity₀, cost₀ = prepare_results(outcome[outcome.λ .== 0.0,:]);
insertcols!(validity₀, :generator => "Generic")
insertcols!(cost₀, :generator => "Generic");
```

### Greedy generator for ensemble

```{julia}
# Experiment:
experiment = Experiments.Experiment(X,y,𝑴,target,grid_,n_rounds);
```

```{julia}
run = true
if run
    n = round(T/size(X)[1])
    δ = 0.1
    generator = GreedyGenerator(δ,n,:logitbinarycrossentropy,nothing)
    outcome, path = Experiments.run_experiment(experiment, generator, 5, T=T)
    CSV.write(joinpath(output_folder, "ensemble_greedy.csv"), DataFrame(outcome))
end
```

```{julia}
outcome = CSV.read(joinpath(output_folder, "ensemble_greedy.csv"), DataFrame)

validity₁, cost₁ = prepare_results(outcome,id_vars=[:t, :μ, :γ])
insertcols!(validity₁, :generator => "Ensemble (greedy)")
insertcols!(cost₁, :generator => "Ensemble (greedy)");
```

### Generic generator for ensemble

```{julia}
run = true
if run  
    generator = GenericGenerator(0.0,0.1,1e-5,:logitbinarycrossentropy,nothing)
    outcome, path = Experiments.run_experiment(experiment, generator, 5, T=T)
    CSV.write(joinpath(output_folder, "ensemble_generic.csv"), DataFrame(outcome))
end
```

```{julia}
outcome = CSV.read(joinpath(output_folder, "ensemble_generic.csv"), DataFrame)

validity₂, cost₂ = prepare_results(outcome,id_vars=[:t, :μ, :γ])
insertcols!(validity₂, :generator => "Ensemble (generic)")
insertcols!(cost₂, :generator => "Ensemble (generic)");
```

### Results

```{julia}
validity = vcat(validity₀,validity₁,validity₂)
cost = vcat(cost₀,cost₁,cost₂);
```

```{julia}
using Gadfly
function plot_results(results; title="")
    set_default_plot_size(700px, 600px)
    Gadfly.plot(
        results, 
        xgroup="μ", ygroup="γ", x="t", y="mean", ymin=:ymin, ymax=:ymax, color=:generator,
        Geom.subplot_grid(Geom.point, Geom.errorbar),
        Guide.title(title)
    )
end
```

```{julia}
plot_results(cost, title="Cost")
```

```{julia}
plot_results(validity, title="Validity")
```

### 2022-04-13 dev/work/german_credit.qmd
moved to qmd docs

---
jupyter: julia-1.6
---

## German credit data

```{julia}
include("../src/load.jl")
using AlgorithmicRecourse
using CSV;
using DataFrames;
using LinearAlgebra;
```

```{julia}
df = CSV.read("../data/credit.csv", DataFrame);
df[!,:] = convert.(Float64, df[!,:]); # convert to floats
y = df[:,:y]
N = length(y)
X = Matrix(df[:,Not(:y)])';
X_train = copy(X)
y_train = copy(y);
```

```{julia}
using MLDataUtils, Random, Flux
Random.seed!(1234)

# # Preproc:
# X_train, y_train = oversample((X_train,y_train))
# # Standardize continuous features:
# using StatsBase
# dt = fit(ZScoreTransform, X_train[1:3,:], dims=1)
# X_train[1:3,:] = StatsBase.transform(dt, X_train[1:3,:])

# Stack:
xs = Flux.unstack(X_train,2)
ys = Flux.flatten(y_train)
data = zip(xs, ys)
nn = Models.build_model(input_dim=size(X_train)[1], n_hidden=32)
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
ps = params(nn);
```

```{julia}
using Flux.Optimise: update!, ADAM
using Statistics
opt = ADAM()
epochs = 50
avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))

using Plots
anim = Animation()
plt = plot(ylim=(0,avg_loss(data)), xlim=(0,epochs), legend=false, xlab="Epoch")
avg_l = []

for epoch = 1:epochs
  for d in data
    gs = gradient(params(nn)) do
      l = loss(d...)
    end
    update!(opt, params(nn), gs)
  end
  avg_l = vcat(avg_l,avg_loss(data))
  plot!(plt, avg_l, color=1, title="Average (training) loss")
  frame(anim, plt)
end

gif(anim)
```

**EVEN THE NEURAL NET JUST CLASSIFIES ALL AS 1, WHAT IS GOING ON HERE?**

```{julia}
sum(round.(Flux.σ.(nn(X_train)))' .== 1)
```

```{julia}
using EvalMetrics
using Plots
y_hat = predict(model,X)
rocplot(y,y_hat)
```

```{julia}
generators = (wachter = generate_recourse_wachter, schut = generate_recourse_schut);
generator_args = (wachter=(λ=0.01,), schut=(T=1000,δ=0.05));
target = 1;
```

```{julia}
props = [0.01,0.05,0.1,0.25]
results = DataFrame()
for prop in props
    results_prop = run_experiment(X,y,bayes_logreg,target,generators,generator_args,experiment_dynamic)
    insertcols!(results_prop, :prop => prop)
    results = vcat(results, results_prop)
end
```

Possible that these results are driven by the following fact:

- classifier is biased towards target class (try random over/undersampling)

```{julia}
dt_plot = groupby(results, [:period, :generator, :prop]) |>
    gdf -> combine(gdf, :validity .=> [mean, std] .=> [:mean, :std])
dt_plot[!,:ymin] = dt_plot[!,:mean] - dt_plot[!,:std]
dt_plot[!,:ymax] = dt_plot[!,:mean] + dt_plot[!,:std]
ggplot(data=dt_plot, aes(x=:period, y=:mean, colour=:generator)) +
    geom_line() + 
    geom_errorbar(aes(ymin=:ymin, ymax=:ymax), width=.2) + 
    geom_point() +
    facet_wrap(R".~prop") +
    theme_bw() |>
    p -> ggsave("www/german_dynamic_validity.png", plot = p);
load("www/german_dynamic_validity.png")
```

Interesting to see that there are some clear trends in the cost of recourse over time:

- what could be the reason for these dynamics?

```{julia}
dt_plot = groupby(results, [:period, :generator, :prop]) |>
    gdf -> combine(gdf, :cost .=> [mean, std] .=> [:mean, :std])
dt_plot[!,:ymin] = dt_plot[!,:mean] - dt_plot[!,:std]
dt_plot[!,:ymax] = dt_plot[!,:mean] + dt_plot[!,:std]
ggplot(data=dt_plot, aes(x=:period, y=:mean, colour=:generator)) +
    geom_line() + 
    geom_errorbar(aes(ymin=:ymin, ymax=:ymax), width=.2) + 
    geom_point() +
    facet_wrap(R".~prop") +
    theme_bw() |>
    p -> ggsave("www/german_dynamic_cost.png", plot = p);
load("www/german_dynamic_cost.png")
```

### Adjusting for imbalance

```{julia}
using MLDataUtils
X_train, y_train = oversample((transpose(X),y))
X_train = transpose(X_train);
model = bayes_logreg(X_train,y_train);
```

```{julia}
y_hat = predict(model,X)
rocplot(y,y_hat)
```

### 2022-04-13 dev/work/preprocessing.qmd
moved to qmd docs

--- 2021-12-07 work/preprocessing.Rmd
@@
 date: "`r format(Sys.Date(), '%B, %Y')`"
-output: 
-  bookdown::pdf_document2: 
-    toc: false
-    includes:
-      in_header: ../tex/preamble.tex
-  bookdown::html_document2: default
-fontsize: 12pt
-bibliography: ../bib.bib
-link-citations: true
-abstract: "..."
 ---

### 2022-04-13 dev/work/synthetic_bayesian_ar.qmd
moved to qmd docs

---
jupyter: julia-1.6
---

# Algorithmic recourse for Bayesian classifiers (Schut et al. 2021)

In this notebook we see how algorithmic recourse can be implemented for Bayesian classifiers. In particular, we follow the implementation in Schut et al (2021).

```{julia}
include("../src/load.jl")
using AlgorithmicRecourse
using CSV;
using DataFrames;
using LinearAlgebra;
using Distributions;
output_folder = "output/synthetic"
www_folder = "www/synthetic"
using Base.Filesystem: joinpath
using Images, FileIO
```

## Fitting the Bayesian classifier

```{julia}
df = CSV.read("../data/cats_dogs_large.csv", DataFrame);
```

For plotting we will use the R package `ggplot2` which can be readily integrated into Julia (just like many other R packages, see for example [here](https://avt.im/blog/2018/03/23/R-packages-ggplot-in-julia#fn:rcall)).

```{julia}
using RCall
@rlibrary ggplot2
@rlibrary ggimage
@rlibrary emojifont
@rlibrary viridis
@rlibrary magick
```

```{julia}
y = df[:,:y];
N = length(y);
X = Matrix(df[:,Not(:y)]);
X = [ones(N) X]; # add for constant
d = size(X)[2]; # number of features
w_0 = zeros(d); # zero-mean prior
λ = 0.005
H_0 = UniformScaling(λ);
```

```{julia}
model = Models.bayes_logreg(X,y);
w = model.μ;
```

```{julia}
step = 1;
expand = 20;
W = expandgrid((minimum(df.x1)-expand):step:(maximum(df.x1)+expand),(minimum(df.x2)-expand):step:(maximum(df.x2)+expand));
N_grid = size(W)[1];
```

```{julia}
y_probit = Models.predict(model, hcat(ones(N_grid), W));
estimates = DataFrame(hcat(y_probit,W), :auto);
features = copy(df);
transform!(features, :y => (x -> ifelse.(x.==0,"../www/cat.png","../www/dog.png"))  => :emoji);
```

```{julia}
ggplot() + 
  geom_contour_filled(data=estimates, aes(x = :x2, y = :x3, z = :x1), bins=25, alpha=0.8) +
  geom_image(data=features, aes(x=:x1, y=:x2, image=:emoji), size=0.1) +
  guides(fill="none") +
  scale_fill_viridis(option="C",discrete=true) +
  scale_x_continuous(expand = R"c(0, 0)") + scale_y_continuous(expand = R"c(0, 0)") +
  labs(x="Height", y="Tail") +
  theme_bw() |>
  p -> ggsave(joinpath(www_folder, "predictive.png"), plot = p, width=5, height=5);
```

![](www/synthetic/predictive.png)

## Recourse algorithm

We have our fitted logistic regression classifier $f(\mathbf{x}^\text{CF})$. The loss for our recourse objective is $\ell(f(\mathbf{x}^\text{CF}),1))$. Using Hinge loss we have

$$\ell(f(\mathbf{x}^\text{CF}),1))=(1-z\cdot logits(f(\mathbf{x}^\text{CF}),1)))_{+}$$

where $z=1$ if $y=1$ and $z=-1$ if $y=0$. Furthermore, $logits(\cdot)$ in this particular case just refers to the linear predictions that enter the sigmoid function to produce predicted probabilities.

This yields the following gradient for our recourse problem:

$$
\begin{aligned}
&& \nabla_{\mathbf{x}^\text{CF}} \left( \ell(f(\mathbf{x}^\text{CF}),1)) \right) &= \begin{cases} -z \cdot \hat{\mathbf{w}}  & \text{if} \ \ \ \hat{\mathbf{w}}^T\mathbf{x}^\text{CF} z \le 1\\ \mathbf{0} & \text{otherwise} \end{cases} \\
\end{aligned}
$$

### Schut et al.

The algorithm used to generate recourse in Schut et al. (2021) is based on the Jacobian-based Saliency Map Attack (JSMA). Let's refer to the recourse variant as Jacobian-based Saliency Map CE (JSMCE). It can be coded up as follows:

## Generating recourse

```{julia}
using AlgorithmicRecourse.Models: BayesianLogisticModel
𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ); # repare for usage with AlgorithmicRecourse.jl
x̅ = [50.0, 60.0]; # factual
using Random
Random.seed!(1234)
y̅ = round(AlgorithmicRecourse.Models.probs(𝑴, x̅)[1])
target = ifelse(y̅==1.0,0.0,1.0) # opposite label as target
γ = ifelse(target==1.0,0.9,0.1); # desired threshold based on target
```

```{julia}
generic = GenericGenerator(0.0,0.1,1e-5,:logitbinarycrossentropy,nothing)
greedy = GreedyGenerator(1,30,:logitbinarycrossentropy,nothing)
```

```{julia}
recourse_generic = generate_recourse(generic, x̅, 𝑴, target, γ) # generate recourse
recourse_greedy = generate_recourse(greedy, x̅, 𝑴, target, γ);
```

```{julia}
path_df = DataFrame();
recourses = (generic = recourse_generic, greedy = recourse_greedy)
for i in 1:length(recourses)
    recourse_name = keys(recourses)[i]
    recourse = recourses[i]
    path = DataFrame(recourse.path, :auto)
    insertcols!(path, :type => titlecase(String(recourse_name)))
    insertcols!(path, :emoji=> "../www/cat.png")
    path_df = vcat(path_df, path)
end;
```

## Results

```{julia}
ggplot() + 
    geom_contour_filled(data=estimates, aes(x = :x2, y = :x3, z = :x1), bins=25, alpha=0.8) +
    geom_abline(intercept=1-w[1], slope=-w[2]/w[3]) +
    geom_image(data=features, aes(x=:x1, y=:x2, image=:emoji), size=0.1, image_fun=R"function(img) magick::image_fx(img, expression = '0.5*a', channel = 'alpha')") +
    geom_image(data=path_df, aes(x=:x1, y=:x2, image=:emoji), size=0.1) +
    guides(fill="none") +
    scale_fill_viridis(option="C",discrete=true) +
    scale_x_continuous(expand = R"c(0, 0)") + scale_y_continuous(expand = R"c(0, 0)") +
    labs(x="Height", y="Tail") +
    facet_wrap(R".~type") +
    theme_bw() |>
    p -> ggsave(joinpath(www_folder,"path.png"), plot = p, width=8, height=4);
```

As we can see in the figure below, the approach by Schut et al. results in a counter factual explanation with high posterior predictive likelihood. Conversely, the conventional approach (Wachter) generates recourse that lies right on the decision boundary where predictive uncertainty is very high.

```{julia}
load(joinpath(www_folder,"path.png"))
```

### 2022-04-13 dev/work/synthetic_endogenous_shifts.qmd
moved to qmd docs

---
jupyter: julia-1.6
---

# Endogenous model shifts

In this notebook we will use toy data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.

```{julia}
using Pkg; Pkg.activate("..")
```

```{julia}
include("../src/load.jl")
using AlgorithmicRecourse
using CSV, DataFrames
using LinearAlgebra;
using Distributions;
using Images, FileIO;
using Plots, PlotThemes
theme(:wong)
using Logging
disable_logging(Logging.Info)
output_folder = "output/synthetic"
www_folder = "www/synthetic"
using Base.Filesystem: joinpath
```

## Single-period

### Training the classifier

We start by training a simple linear classifier: Bayesian Logisitic Regression.

```{julia}
df = CSV.read("../data/cats_dogs_large.csv", DataFrame);
y = df[:,:y];
N = length(y);
X = Matrix(df[:,Not(:y)]);
model = Models.bayes_logreg(X,y); # get MAP estimates
w = model.μ;
```

```{julia}
# Plot data points.
plt = plot()
using Main.Experiments: plot_data!
plot_data!(plt,X,y)
```

```{julia}
using AlgorithmicRecourse.Models: BayesianLogisticModel, LogisticModel
𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ) # fully Bayesian model
𝑴_plugin = LogisticModel(reshape(model.μ[2:3],1,2), [model.μ[1]]); # Plugin estimate (non-Bayesian)
models = (Bayesian=𝑴, Plugin=𝑴_plugin);
```

```{julia}
using Main.Experiments: plot_contour
p1 = plot_contour(X,y,𝑴,title="Bayesian")
p2 = plot_contour(X,y,𝑴_plugin,title="Plugin")
plot(p1,p2,size=(1000,400))
```

### Generating recourse

Now we will generate and implement algorithmic recourse for a subset of cats.

```{julia}
generators = (Bayesian=GreedyGenerator(1,500,:logitbinarycrossentropy, nothing), Plugin=GenericGenerator(0.1,0.1,1e-5,:logitbinarycrossentropy,nothing));
proportion = 0.2
using StatsBase
cats = findall(df.y.==0);
chosen_individuals = StatsBase.sample(cats,Int(round(proportion * length(cats))),replace=false);
target = 1.0
γ = 0.90;
```

```{julia}
function generate_counterfactuals(generator, 𝑴, target, γ, chosen_individuals; X=X, y=y)
    X̲ = copy(X)
    y̲ = copy(y)
    for i in chosen_individuals
        x̅ = X[i,:]
        recourse = generate_recourse(generator, x̅, 𝑴, target, γ);
        X̲[i,:] = recourse.x̲
        y̲[i] = recourse.y̲
    end
    return X̲, y̲
end
```

```{julia}
plt = []
for i in 1:length(generators)
    X̲, y̲ = generate_counterfactuals(generators[i], models[i], target, γ, chosen_individuals)
    pltᵢ = plot_contour(X̲,y̲,models[i],title=keys(generators)[i])
    plt = vcat(plt, pltᵢ)
end
plot(plt..., size = (1000,400))
```

### Retraining classifier

Finally, we will retrain the classifier (model shift) and plot the resulting new decision boundary and posterior predictive distribution.

```{julia}
plt = []
for i in 1:length(generators)
    X̲, y̲ = generate_counterfactuals(generators[i], models[i], target, γ, chosen_individuals)
    data = Models.prepare_data(X̲', y̲')
    𝑴 = Models.retrain(models[i], data)
    pltᵢ = plot_contour(X̲,y̲,𝑴,title=keys(generators)[i])
    plt = vcat(plt, pltᵢ)
end
plot(plt..., size = (1000,400))
```

## Dynamic

```{julia}
# Variables:
μ = [0.01,0.1,0.25]
γ = [0.50,0.75,0.9]
grid_ = Experiments.GridVariables(μ, γ)
n_rounds = 10
n_folds = 5
target = 1.0
T = 1000;
# Experiments:
experiment = Experiments.Experiment(X',y',models[:Bayesian],target,grid_,n_rounds)
experiment_plugin = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
experiments = (experiment, experiment_plugin);
```

```{julia}
run = false
if run
    using Random
    Random.seed!(1234)
    for j in 1:length(generators)
        outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
        Experiments.save_path(joinpath(output_folder, string(keys(generators)[j])),path)
        CSV.write(joinpath(output_folder, string(keys(generators)[j]) * ".csv"), DataFrame(outcome))
    end
end
```

```{julia}
run = false
if run
    using Main.Experiments: plot_path
    paths = [Experiments.load_path(joinpath(output_folder, string(keys(generators)[j]))) for j ∈ 1:length(generators)];
    anim = Experiments.plot_path(paths[1],Γ=γ,𝙈=μ)
    gif(anim,joinpath(www_folder,"bayesian.gif"), fps=2)
    anim = Experiments.plot_path(paths[2],Γ=γ,𝙈=μ)
    gif(anim,joinpath(www_folder,"plugin.gif"), fps=2)
end
```

For the Bayesian classifier with greedy recourse...

![](www/synthetic/bayesian.gif)

For the non-Bayesian classifier with generic recourse...

![](www/synthetic/plugin.gif)

```{julia}
using Main.Experiments: prepare_results, plot_results
validity = DataFrame()
cost = DataFrame()
files = (Bayesian=joinpath(output_folder, "Bayesian.csv"), Plugin=joinpath(output_folder, "Plugin.csv"))
results = map(file -> CSV.read(file, DataFrame), files)
p_val, p_cost = plot_results(results);
```

```{julia}
p_val
```

```{julia}
p_cost
```

## Neural network

```{julia}
using Flux
data = Models.prepare_data(X', y')
nn = Models.build_model(input_dim=size(X')[1], n_hidden=100)
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y);
```

Training a single neural network...

```{julia}
using BSON
run = false
opt = ADAM()
if run
  # Train model:
  using Flux.Optimise: update!, ADAM
  using Statistics, StatsBase
  epochs = 10
  avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
  accuracy(data) = sum(map(d ->round.(Flux.σ.(nn(d[1]))) .== d[2], data))[1]/length(data)

  using Plots
  anim = Animation()
  avg_l = [avg_loss(data)]
  p1 = scatter(ylim=(0,avg_l[1]), xlim=(0,epochs), legend=false, xlab="Epoch", title="Average loss")
  acc = [accuracy(data)]
  p2 = scatter(ylim=(0.5,1), xlim=(0,epochs), legend=false, xlab="Epoch", title="Accuracy")

  τ = 1.1
  stopping_criterium_reached = accuracy(data) >= τ
  epoch = 1

  while epoch <= epochs && !stopping_criterium_reached
    for d in data
      gs = gradient(Flux.params(nn)) do
        l = loss(d...)
      end
      update!(opt, Flux.params(nn), gs)
    end
    avg_l = vcat(avg_l,avg_loss(data))
    plot!(p1, [0:epoch], avg_l, color=1)
    scatter!(p1, [0:epoch], avg_l, color=1)
    acc = vcat(acc,accuracy(data))
    plot!(p2, [0:epoch], acc, color=1)
    scatter!(p2, [0:epoch], acc, color=1)
    plt=plot(p1,p2, size=(600,300))
    frame(anim, plt)

    # Check if desired accuracy reached:
    stopping_criterium_reached = accuracy(data) >= τ

    epoch += 1

  end

  gif(anim, joinpath(www_folder, "single_nn.gif"), fps=10)

  BSON.@save joinpath(output_folder, "nn.bson") nn

end
```

```{julia}
using BSON: @load
@load joinpath(output_folder, "nn.bson") nn
𝑴ₙₙ = Models.FittedNeuralNet(nn, opt, loss);
```

Training a deep ensemble...

```{julia}
opt = ADAM()
loss_type = :logitbinarycrossentropy
run = false
if run
    K = 50
    𝓜 = Models.build_ensemble(K,kw=(input_dim=size(X')[1], n_hidden=100));
    𝓜, anim = Models.forward(𝓜, data, opt, n_epochs=10, plot_every=10, loss_type=loss_type, τ=1.1); # fit the ensemble
    Models.save_ensemble(𝓜, root=joinpath(output_folder, "ensemble")) # save to disk
    gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
end
```

```{julia}
𝓜 = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
𝑴 = Models.FittedEnsemble(𝓜, opt, loss_type)
models = (Bayesian=𝑴, Plugin=𝑴ₙₙ);
```

```{julia}
using Main.Experiments: plot_contour
p1 = plot_contour(X,y,models[1],title="Bayesian")
p2 = plot_contour(X,y,models[2],title="Plugin")
plot(p1,p2,size=(1000,400))
```

```{julia}
# Experiments:
experiment = Experiments.Experiment(X',y',models[:Bayesian],target,grid_,n_rounds)
experiment_plugin = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
experiments = (experiment, experiment_plugin);
```

```{julia}
run = true
if run
    using Random
    Random.seed!(1234)
    for j in 1:length(generators)
        outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
        Experiments.save_path(joinpath(output_folder, string(keys(generators)[j])),path)
        CSV.write(joinpath(output_folder, string(keys(generators)[j]) * "_deep.csv"), DataFrame(outcome))
    end
end
```

### 2022-04-13 paper/sections/introduction.qmd
more work on setting up submission

---
format: gfm
---

## Introduction {#intro}

...

### 2022-04-13 dev/submissions/aies/extended_abstract/submission.qmd
more on aies

---
title: "Overlooked Dynamics in Algorithmic Recourse"
format: 
    html: default
    pdf:
        documentclass: acmart
        # include-in-header: preamble.tex
author:
  - name: Patrick Altmeyer
---

### Author

- Full Name: {{< var author.name >}}
- Home page: {{< var author.url >}}
- Affiliation: {{< var author.affiliation >}}
- Affiliation URL: {{< var author.affiliation-url >}}
- Country of citizenship: {{< var author.nationality >}}

### PhD Progress

- Expected graduation date: {{< var phd.graduation-date >}}
- Years in program: {{< var phd.years >}}

### Research Advisor

- Full Name: {{< var advisor.name >}}

*Key words*: {{< var keywords >}}

### Letter of Support

\pagebreak

## Extended abstract

### 2022-04-13 dev/submissions/aies/sections/support.qmd
aies

### Letter of Support

Patrick has been my supervisee for ...

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.qmd
Fun with latex templates

---
title: Dynamics in Algorithmic Recourse
format:
    pdf:
        documentclass: acmconf
        number-sections: true
        include-in-header: preamble.tex
        keep-tex: true
---

Recent advances in artificial intelligence (AI) have propelled its adoption in domains outside of computer science including health care, bioinformatics and genetics. In finance, economics and other social sciences, applications of AI are still relatively limited. Decision-making in these fields has traditionally been guided by Generalized Linear Models (GLM), which are theoretically founded, interpretable and often sufficient to model relationships between variables. Model interpretability is crucial in the social sciences context, because inference is typically at least as important as predictive performance. Decision-makers in the social sciences are also typically required to explain their decisions to human stakeholders: central bankers, for example, are held accountable by the public for the policies they decide on. It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that ultimately cannot be trusted. Deep learning models, for example, are generally considered as black boxes and therefore difficult to apply in a context that demands explanations.

In my research I explore and develop methodologies that improve the trustworthiness of AI. I would like to understand how we can unlock the enormous potential of AI without sacrificing the human aspect of decision-making in finance and economics. My work so far has focused primarily on counterfactual explanations, algorithmic recourse and probabilistic machine learning. Counterfactual explanations are intuitive, largely model-agnostic and straight-forward to implement. They are also intrinsically linked to the potential outcome framework for causal inference and therefore should be somewhat familiar to social scientists. Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of algorithmic recourse to help individuals facing adverse decisions. Probabilistic machine learning can be leveraged in this context and more generally facilitates inference and interpretability. It is also closely related to Bayesian statistics, which has played an important role in both finance and economics for many years.

In the following (@sec-main), I will first briefly present one particular research question I have explored during the first months of my PhD: how do counterfactual explanations handle dynamics? I will also briefly present related projects I have worked on (@sec-related) and ideas for future projects (@sec-future).

## Dynamics in Algorithmic Recourse {#sec-main}

Existing work on counterfactual explanations and algorithmic recourse has largely been limited to the following static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], plausible [@joshi2019towards, @antoran2020getting, @schut2021generating], sparse [@schut2021generating], diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$. The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual]. Only one recent work considers the implications of **exogenous** domain shifts on the validity of recourse [@upadhyay2021towards]. The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal domain and model shift. They show that their approach yields more robust counterfactuals than existing approaches. In my project I investigate **endogenous** domain and model shifts, i.e. shifts that occur as algorithmic recourse is actually impemented by a proportion of individuals. Preliminary findings indicate that individuals who receive and implement algorithmic recourse end up forming a distinct subgroup inside the target class, which may leave them vulnerable to discrimination (@fig-dynamics). This is a work-in-progress that I would like to present and discuss at AIES.

![PLACEHOLDER: The dynamics of algorithmic recourse.](www/dynamics.png){#fig-dynamics fig.pos="h" width=250px}

## Related Projects {#sec-related}

Alongside my research I have developed open-source implementations related to explainable AI. [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/) is a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R. I have recently submitted the package along with a companion paper as a proposal for a main talk at [JuliaCon](https://juliacon.org/2022/). [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/) is a small Julia package that can be used to recover Bayesian representations of deep neural networks through Laplace approximation in a post-hoc manner. It is inspired by a recent paper [@daxberger2021laplace] and has also been submitted to JuliaCon. Finally, [deepvars](https://github.com/pat-alt/deepvars) is an R package that implements an approach towards vector autoregression that leverages deep learning. This was originally my master's thesis and later presented at the NeurIPS 2021 MLECON workshop. I have also published several blog posts on explainable AI and probabilisitic ML in an effort to make my research accessible to a broad audience.

## Future Projects {#sec-future}

Data sets in finance and economics typically involve time series data. Therefore, I am naturally interested in the application of explainable AI to sequential data, an area which has so far not been explored extensively. In the future, I want to work on counterfactual explanations for time series models. I am also interested in seeing if and how Laplace approximation can be used for Bayesian deep learning with time series data. I hope that the findings from both of these projects can ultimately be used to build complex but interpretable time series models for classification and forecasting in finance and economics. For example, I would like to leverage effortless Bayesian deep learning to make our proposed Deep Vector Autoregression model explainable.

## References

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.tex
Fun with latex templates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{acmconf}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Dynamics in Algorithmic Recourse},
  pdfauthor={Patrick Altmeyer},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\ConferenceShortName{AIES '22}
\ConferenceName{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Dynamics in Algorithmic Recourse}
\author{Patrick Altmeyer}
\date{April 25, 2022}

\begin{document}
\maketitle

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

Recent advances in artificial intelligence (AI) have propelled its
adoption in domains outside of computer science including health care,
bioinformatics and genetics. In finance, economics and other social
sciences, applications of AI are still relatively limited.
Decision-making in these fields has traditionally been guided by
Generalized Linear Models (GLM), which are theoretically founded,
interpretable and often sufficient to model relationships between
variables. Model interpretability is crucial in the social sciences
context, because inference is typically at least as important as
predictive performance. Decision-makers in the social sciences are also
typically required to explain their decisions to human stakeholders:
central bankers, for example, are held accountable by the public for the
policies they decide on. It is therefore not surprising that
practitioners and academics in these fields are reluctant to adopt AI
technologies that ultimately cannot be trusted. Deep learning models,
for example, are generally considered as black boxes and therefore
difficult to apply in a context that demands explanations.

In my research I explore and develop methodologies that improve the
trustworthiness of AI. I would like to understand how we can unlock the
enormous potential of AI without sacrificing the human aspect of
decision-making in finance and economics. My work so far has focused
primarily on counterfactual explanations, algorithmic recourse and
probabilistic machine learning. Counterfactual explanations are
intuitive, largely model-agnostic and straight-forward to implement.
They are also intrinsically linked to the potential outcome framework
for causal inference and therefore should be somewhat familiar to social
scientists. Counterfactual explanations that involve realistic and
actionable changes can be used for the purpose of algorithmic recourse
to help individuals facing adverse decisions. Probabilistic machine
learning can be leveraged in this context and more generally facilitates
inference and interpretability. It is also closely related to Bayesian
statistics, which has played an important role in both finance and
economics for many years.

In the following (Section~\ref{sec-main}), I will first briefly present
one particular research question I have explored during the first months
of my PhD: how do counterfactual explanations handle dynamics? I will
also briefly present related projects I have worked on
(Section~\ref{sec-related}) and ideas for future projects
(Section~\ref{sec-future}).

\hypertarget{sec-main}{%
\section{Dynamics in Algorithmic Recourse}\label{sec-main}}

Existing work on counterfactual explanations and algorithmic recourse
has largely been limited to the following static setting: given some
classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are interested in
finding close (Wachter, Mittelstadt, and Russell 2017), actionable
(Ustun, Spangher, and Liu 2019), plausible Schut et al. (2021), sparse
(Schut et al. 2021), diverse (Mothilal, Sharma, and Tan 2020) and
ideally causally founded counterfactual explanations (Karimi, Schölkopf,
and Valera 2021) for some individual \(x\). The ability of
counterfactual explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
(Verma, Dickerson, and Hines 2020). Only one recent work considers the
implications of \textbf{exogenous} domain shifts on the validity of
recourse (Upadhyay, Joshi, and Lakkaraju 2021). The authors propose a
simple minimax objective, that minimizes the counterfactual loss
function for a maximal domain and model shift. They show that their
approach yields more robust counterfactuals than existing approaches. In
my project I investigate \textbf{endogenous} domain and model shifts,
i.e.~shifts that occur as algorithmic recourse is actually impemented by
a proportion of individuals. Preliminary findings indicate that
individuals who receive and implement algorithmic recourse end up
forming a distinct subgroup inside the target class, which may leave
them vulnerable to discrimination (Figure~\ref{fig-dynamics}). This is a
work-in-progress that I would like to present and discuss at AIES.

\begin{figure}

{\centering \includegraphics[width=2.60417in,height=\textheight]{www/dynamics.png}

}

\caption{\label{fig-dynamics}PLACEHOLDER: The dynamics of algorithmic
recourse.}

\end{figure}

\hypertarget{sec-related}{%
\section{Related Projects}\label{sec-related}}

Alongside my research I have developed open-source implementations
related to explainable AI.
\href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}
is a Julia package that can be used to generate counterfactual
explanations for models developed and trained not only in Julia, but
also in other popular programming languages like Python and R. I have
recently submitted the package along with a companion paper as a
proposal for a main talk at \href{https://juliacon.org/2022/}{JuliaCon}.
\href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}
is a small Julia package that can be used to recover Bayesian
representations of deep neural networks through Laplace approximation in
a post-hoc manner. It is inspired by a recent paper (Daxberger et al.
2021) and has also been submitted to JuliaCon. Finally,
\href{https://github.com/pat-alt/deepvars}{deepvars} is an R package
that implements an approach towards vector autoregression that leverages
deep learning. This was originally my master's thesis and later
presented at the NeurIPS 2021 MLECON workshop. I have also published
several blog posts on explainable AI and probabilisitic ML in an effort
to make my research accessible to a broad audience.

\hypertarget{sec-future}{%
\section{Future Projects}\label{sec-future}}

Data sets in finance and economics typically involve time series data.
Therefore, I am naturally interested in the application of explainable
AI to sequential data, an area which has so far not been explored
extensively. In the future, I want to work on counterfactual
explanations for time series models. I am also interested in seeing if
and how Laplace approximation can be used for Bayesian deep learning
with time series data. I hope that the findings from both of these
projects can ultimately be used to build complex but interpretable time
series models for classification and forecasting in finance and
economics. For example, I would like to leverage effortless Bayesian
deep learning to make our proposed Deep Vector Autoregression model
explainable.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
Miguel Hernández-Lobato. 2020. {``Getting a Clue: A Method for
Explaining Uncertainty Estimates.''} \emph{arXiv Preprint
arXiv:2006.06848}.

\leavevmode\vadjust pre{\hypertarget{ref-daxberger2021laplace}{}}%
Daxberger, Erik, Agustinus Kristiadi, Alexander Immer, Runa Eschenhagen,
Matthias Bauer, and Philipp Hennig. 2021. {``Laplace Redux-Effortless
Bayesian Deep Learning.''} \emph{Advances in Neural Information
Processing Systems} 34.

\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems.''}
\emph{arXiv Preprint arXiv:1907.09615}.

\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera. 2021.
{``Algorithmic Recourse: From Counterfactual Explanations to
Interventions.''} In \emph{Proceedings of the 2021 ACM Conference on
Fairness, Accountability, and Transparency}, 353--62.

\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan. 2020.
{``Explaining Machine Learning Classifiers Through Diverse
Counterfactual Explanations.''} In \emph{Proceedings of the 2020
Conference on Fairness, Accountability, and Transparency}, 607--17.

\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
Sacaleanu, Yarin Gal, et al. 2021. {``Generating Interpretable
Counterfactual Explanations by Implicit Minimisation of Epistemic and
Aleatoric Uncertainties.''} In \emph{International Conference on
Artificial Intelligence and Statistics}, 1756--64. PMLR.

\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021.
{``Towards Robust and Reliable Algorithmic Recourse.''} \emph{arXiv
Preprint arXiv:2102.13620}.

\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable
Recourse in Linear Classification.''} In \emph{Proceedings of the
Conference on Fairness, Accountability, and Transparency}, 10--19.

\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual
Explanations for Machine Learning: A Review.''} \emph{arXiv Preprint
arXiv:2010.10596}.

\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
{``Counterfactual Explanations Without Opening the Black Box: Automated
Decisions and the GDPR.''} \emph{Harv. JL \& Tech.} 31: 841.

\end{CSLReferences}

\end{document}

### 2022-04-25 dev/submissions/aies/sections/personal_statement.qmd
Fun with latex templates

---
format:
    pdf: default
---

## Personal Statement

As a first-year PhD student in computer science with a background in finance and economics, I was very happy to see that AIES offers a student track. I have worked on several projects during the first months of my PhD and while the one I hope to present at AIES is the most mature research project at this point, it is still very much a work in progress. I think that the question I pose about the dynamics of algorithmic recourse is an interesting, albeit still fairly broad and open one. Through my participation in AIES, I would hope to get feedback that helps me refine the research question. As experiments related to the question are now being carried out by my students, the work can be expected to progress further until the actual AIES event and hence I could share more experimental results that may be of interest to the broader community. AIES would also be an opportunity to discuss ideas for potential remedies to the issues this work highlights.

I also think that my broader PhD research agenda on the intersection of trustworthy AI and the social sciences should be of interest to the AIES community. Thanks to my academic background in economics, finance and data science as well my previous professional experience in monetary policy, I believe that I can contribute insights to a wide range of discussions revolving around the diverse set of topics relevant to AIES. It would be immensely helpful to learn from more experienced colleagues as well as fellow early-stage researchers, who are working on related research questions. At the main conference and the student event I expect to be exposed to new research ideas that I have never thought off, but may well be highly relevant to my own research. I would even hope to potentially find opportunities for future collaborations.

This would be the first student program I have attended, although it is worth mentioning that in the past I have presented research at the European Central Bank (2019), IFABS (2021), NeurIPS (2021) and the Irving Fisher Committee on Central Bank Statistics (2022). Since the main PhD research project that I would like to dicuss is still in its early stages, I have not submitted anything to the main track of AIES. I would hope that following my participation in the student track I can use the feedback to carry the work forward and submit it to the main track next year.

### 2022-04-25 dev/work/boston_housing.qmd
Fun with latex templates

@@
     K = 50
-    𝓜 = Models.build_ensemble(K,kw=(input_dim=size(X)[1], n_hidden=100));
-    𝓜, anim = Models.forward(𝓜, data, opt, n_epochs=30, plot_every=10, loss_type=loss_type); # fit the ensemble
-    Models.save_ensemble(𝓜, root=joinpath(output_folder, "ensemble")) # save to disk
+    ensemble = Models.build_ensemble(K,kw=(input_dim=size(X)[1], n_hidden=100));
+    ensemble, anim = Models.forward(ensemble, data, opt, n_epochs=30, plot_every=10, loss_type=loss_type); # fit the ensemble
+    Models.save_ensemble(ensemble, root=joinpath(output_folder, "ensemble")) # save to disk
     gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
@@
 ```{julia}
-𝓜 = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
-𝑴 = Models.FittedEnsemble(𝓜, opt, loss_type);
+ensemble = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
+𝑴 = Models.FittedEnsemble(ensemble, opt, loss_type);
 ```

### 2022-04-25 dev/work/synthetic_endogenous_shifts.qmd
Fun with latex templates

@@
 ```{julia}
-using Pkg; Pkg.activate("..")
-```
-
-```{julia}
-include("../src/load.jl")
-using AlgorithmicRecourse
-using CSV, DataFrames
-using LinearAlgebra;
-using Distributions;
-using Images, FileIO;
-using Plots, PlotThemes
+include("dev/utils.jl")
+using CounterfactualExplanations, CSV, DataFrames, LinearAlgebra, Distributions, Images, FileIO, Plots, PlotThemes
+using Base.Filesystem: joinpath
 theme(:wong)
@@
 www_folder = "www/synthetic"
-using Base.Filesystem: joinpath
 ```
@@
     K = 50
-    𝓜 = Models.build_ensemble(K,kw=(input_dim=size(X')[1], n_hidden=100));
-    𝓜, anim = Models.forward(𝓜, data, opt, n_epochs=10, plot_every=10, loss_type=loss_type, τ=1.1); # fit the ensemble
-    Models.save_ensemble(𝓜, root=joinpath(output_folder, "ensemble")) # save to disk
+    ensemble = Models.build_ensemble(K,kw=(input_dim=size(X')[1], n_hidden=100));
+    ensemble, anim = Models.forward(ensemble, data, opt, n_epochs=10, plot_every=10, loss_type=loss_type, τ=1.1); # fit the ensemble
+    Models.save_ensemble(ensemble, root=joinpath(output_folder, "ensemble")) # save to disk
     gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
@@
 ```{julia}
-𝓜 = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
-𝑴 = Models.FittedEnsemble(𝓜, opt, loss_type)
+ensemble = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
+𝑴 = Models.FittedEnsemble(ensemble, opt, loss_type)
 models = (Bayesian=𝑴, Plugin=𝑴ₙₙ);

### 2022-04-25 src/models/deep_ensemble.jl

@@ -77,7 +77,7 @@ using AlgorithmicRecourse
 import AlgorithmicRecourse.Models: logits, probs # import functions in order to extend
 
 """
-    FittedNeuralNet(𝓜::AbstractArray,opt::Any,loss::Function)
+    FittedNeuralNet(ensemble::AbstractArray,opt::Any,loss::Function)
 
 A simple subtype that is compatible with the AlgorithmicRecourse.jl package.
 """
@@ -115,45 +115,45 @@ end
 
 using Statistics
 """
-    forward(𝓜, data, opt; loss_type=:logitbinarycrossentropy, plot_loss=true, n_epochs=200, plot_every=20) 
+    forward(ensemble, data, opt; loss_type=:logitbinarycrossentropy, plot_loss=true, n_epochs=200, plot_every=20) 
 
 Trains a deep ensemble by separately training each neural network.
 """
-function forward(𝓜, data, opt; loss_type=:logitbinarycrossentropy, plot_loss=true, n_epochs=200, plot_every=20, τ=1.0) 
+function forward(ensemble, data, opt; loss_type=:logitbinarycrossentropy, plot_loss=true, n_epochs=200, plot_every=20, τ=1.0) 
 
     anim = nothing
     if plot_loss
         anim = Animation()
         plt = plot(ylim=(0,1), xlim=(0,n_epochs), legend=false, xlab="Epoch", title="Average (training) loss")
-        for i in 1:length(𝓜)
-            nn = 𝓜[i]
+        for i in 1:length(ensemble)
+            nn = ensemble[i]
             loss(x, y) = getfield(Flux.Losses,loss_type)(nn(x), y)
             nn = forward_nn(nn, loss, data, opt, n_epochs=n_epochs, plotting=(plt, anim, i, plot_every), τ=τ)
-            𝓜[i] = nn
+            ensemble[i] = nn
         end
     else
         plt = nothing
-        for i in 1:length(𝓜)
-            nn = 𝓜[i]
+        for i in 1:length(ensemble)
+            nn = ensemble[i]
             loss(x, y) = getfield(Flux.Losses,loss_type)(nn(x), y)
             nn = forward_nn(nn, loss, data, opt, n_epochs=n_epochs, plotting=plt, τ=τ)
-            𝓜[i] = nn
[…]

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.qmd
further progress on extended abstract

@@
 title: Dynamics in Algorithmic Recourse
+subtitle: Trustworthy Artificial Intelligence for Finance and Economics
 format:
@@
 
+## Introduction
+
 Recent advances in artificial intelligence (AI) have propelled its adoption in domains outside of computer science including health care, bioinformatics and genetics.
@@
 Deep learning models, for example, are generally considered as black boxes and therefore difficult to apply in a context that demands explanations.
+This PhD project is focused on exploring and developing methodologies that improve the trustworthiness of AI and thereby enable its application in Finance and Economics.
 
-In my research I explore and develop methodologies that improve the trustworthiness of AI.
-I would like to understand how we can unlock the enormous potential of AI without sacrificing the human aspect of decision-making in finance and economics.
-My work so far has focused primarily on counterfactual explanations, algorithmic recourse and probabilistic machine learning.
-Counterfactual explanations are intuitive, largely model-agnostic and straight-forward to implement.
-They are also intrinsically linked to the potential outcome framework for causal inference and therefore should be somewhat familiar to social scientists.
-Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of algorithmic recourse to help individuals facing adverse decisions.
-Probabilistic machine learning can be leveraged in this context and more generally facilitates inference and interpretability.
-It is also closely related to Bayesian statistics, which has played an important role in both finance and economics for many years.
-
-In the following (@sec-main), I will first briefly present one particular research question I have explored during the first months of my PhD: how do counterfactual explanations handle dynamics?
-I will also briefly present related projects I have worked on (@sec-related) and ideas for future projects (@sec-future).
+The remainder of this extended abstract is structured as follows: @sec-main presents one of the research questions I have investigated during the first months of my PhD: how do counterfactual explanations handle dynamics?
+@sec-related places this work in the broader context of Trustworthy AI for Finance and Economics.
 
@@
 
-Existing work on counterfactual explanations and algorithmic recourse has largely been limited to the following static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], plausible [@joshi2019towards, @antoran2020getting, @schut2021generating], sparse [@schut2021generating], diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
+**Counterfactual explanations** (CE) explain how inputs into a model need to change for it to produce different outputs.
+They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with.
+Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals facing adverse decisions.
+An example relevant to the Finance and Econonomics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .
+
+Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], plausible [@joshi2019towards, @antoran2020getting, @schut2021generating], sparse [@schut2021generating], diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
 The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
@@
 They show that their approach yields more robust counterfactuals than existing approaches.
-In my project I investigate **endogenous** domain and model shifts, i.e.
-shifts that occur as algorithmic recourse is actually impemented by a proportion of individuals.
-Preliminary findings indicate that individuals who receive and implement algorithmic recourse end up forming a distinct subgroup inside the target class, which may leave them vulnerable to discrimination (@fig-dynamics).
-This is a work-in-progress that I would like to present and discuss at AIES.
+
+This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually impemented by a proportion of individuals and the classifier is updated in response.
+@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift.
+As this game of implementing AR and updating the classifier is repeated, individuals who receive and implement algorithmic recourse end up forming a distinct subgroup inside the target class, which may leave them vulnerable to discrimination.
+Through future experiments we want to investigate if this phenomenon is robust across different benchmark datasets and counterfactual generators.
 
@@
 
-## Related Projects {#sec-related}
+## Related and Future Work {#sec-related}
+
+### Benchmarking CE in Julia
 
@@
 
-## Future Projects {#sec-future}
+### Probabilistic methods for realistic counterfactual explanations
+
+Probabilistic machine learning can be leveraged in this context and more generally facilitates inference and interpretability.
+It is also closely related to Bayesian statistics, which has played an important role in both finance and economics for many years.
+
+To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not.
+To quantify their joint likelihood, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions.
+While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient.
+In my work I want to explore how recent advances in post-hoc uncertainty quantification can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.
+
+### CE for time series
 
@@
 I hope that the findings from both of these projects can ultimately be used to build complex but interpretable time series models for classification and forecasting in finance and economics.
+
+### Explainable black-box models for time series
+
 For example, I would like to leverage effortless Bayesian deep learning to make our proposed Deep Vector Autoregression model explainable.

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.tex
further progress on extended abstract

@@
 \title{Dynamics in Algorithmic Recourse}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Trustworthy Artificial Intelligence for Finance and Economics}
 \author{Patrick Altmeyer}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 for example, are generally considered as black boxes and therefore
-difficult to apply in a context that demands explanations.
+difficult to apply in a context that demands explanations. This PhD
+project is focused on exploring and developing methodologies that
+improve the trustworthiness of AI and thereby enable its application in
+Finance and Economics.
 
-In my research I explore and develop methodologies that improve the
-trustworthiness of AI. I would like to understand how we can unlock the
-enormous potential of AI without sacrificing the human aspect of
-decision-making in finance and economics. My work so far has focused
-primarily on counterfactual explanations, algorithmic recourse and
-probabilistic machine learning. Counterfactual explanations are
-intuitive, largely model-agnostic and straight-forward to implement.
-They are also intrinsically linked to the potential outcome framework
-for causal inference and therefore should be somewhat familiar to social
-scientists. Counterfactual explanations that involve realistic and
-actionable changes can be used for the purpose of algorithmic recourse
-to help individuals facing adverse decisions. Probabilistic machine
-learning can be leveraged in this context and more generally facilitates
-inference and interpretability. It is also closely related to Bayesian
-statistics, which has played an important role in both finance and
-economics for many years.
-
-In the following (Section~\ref{sec-main}), I will first briefly present
-one particular research question I have explored during the first months
-of my PhD: how do counterfactual explanations handle dynamics? I will
-also briefly present related projects I have worked on
-(Section~\ref{sec-related}) and ideas for future projects
-(Section~\ref{sec-future}).
+The remainder of this extended abstract is structured as follows:
+Section~\ref{sec-main} presents one of the research questions I have
+investigated during the first months of my PhD: how do counterfactual
+explanations handle dynamics? Section~\ref{sec-related} places this work
+in the broader context of Trustworthy AI for Finance and Economics.
 
@@
 
-Existing work on counterfactual explanations and algorithmic recourse
-has largely been limited to the following static setting: given some
-classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are interested in
-finding close (Wachter, Mittelstadt, and Russell 2017), actionable
-(Ustun, Spangher, and Liu 2019), plausible Schut et al. (2021), sparse
-(Schut et al. 2021), diverse (Mothilal, Sharma, and Tan 2020) and
-ideally causally founded counterfactual explanations (Karimi, Schölkopf,
-and Valera 2021) for some individual \(x\). The ability of
-counterfactual explanations to handle dynamics like data and model
-shifts remains a largely unexplored research challenge at this point
-(Verma, Dickerson, and Hines 2020). Only one recent work considers the
-implications of \textbf{exogenous} domain shifts on the validity of
+\textbf{Counterfactual explanations} (CE) explain how inputs into a
+model need to change for it to produce different outputs. They are
+intuitive, simple and intrinsically linked to the potential outcome
+framework for causal inference, which social scientists are familiar
+with. Counterfactual explanations that involve realistic and actionable
+changes can be used for the purpose of \textbf{Algorithmic Recourse}
+(AR) to help individuals facing adverse decisions. An example relevant
+to the Finance and Econonomics domain is consumer credit: in this
+context AR can be used to guide individuals to improve their credit
+worthiness, should they have previously been denied access to credit
+based on a black-box decision-making system .
+
+Existing work on CE and AR has largely been limited to the static
+setting: given some classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we
+are interested in finding close (Wachter, Mittelstadt, and Russell
+2017), actionable (Ustun, Spangher, and Liu 2019), plausible Schut et
+al. (2021), sparse (Schut et al. 2021), diverse (Mothilal, Sharma, and
+Tan 2020) and ideally causally founded counterfactual explanations
+(Karimi, Schölkopf, and Valera 2021) for some individual \(x\). The
+ability of counterfactual explanations to handle dynamics like data and
+model shifts remains a largely unexplored research challenge at this
+point (Verma, Dickerson, and Hines 2020). Only one recent work considers
+the implications of \textbf{exogenous} domain shifts on the validity of
 recourse (Upadhyay, Joshi, and Lakkaraju 2021). The authors propose a
@@
 function for a maximal domain and model shift. They show that their
-approach yields more robust counterfactuals than existing approaches. In
-my project I investigate \textbf{endogenous} domain and model shifts,
-i.e.~shifts that occur as algorithmic recourse is actually impemented by
-a proportion of individuals. Preliminary findings indicate that
-individuals who receive and implement algorithmic recourse end up
-forming a distinct subgroup inside the target class, which may leave
-them vulnerable to discrimination (Figure~\ref{fig-dynamics}). This is a
-work-in-progress that I would like to present and discuss at AIES.
+approach yields more robust counterfactuals than existing approaches.
+
+This project investigates \textbf{endogenous} domain and model shifts,
+that is shifts that occur when AR is actually impemented by a proportion
+of individuals and the classifier is updated in response.
+Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
+involving a probabilistic classifier and a greedy counterfactual
+generator proposed by Schut et al. (2021): AR leads to a domain shift,
+which in turn causes a drastic model shift. As this game of implementing
+AR and updating the classifier is repeated, individuals who receive and
+implement algorithmic recourse end up forming a distinct subgroup inside
+the target class, which may leave them vulnerable to discrimination.
+Through future experiments we want to investigate if this phenomenon is
+robust across different benchmark datasets and counterfactual
+generators.
 
@@
 \hypertarget{sec-related}{%
-\section{Related Projects}\label{sec-related}}
+\section{Related and Future Work}\label{sec-related}}
+
+\hypertarget{benchmarking-ce-in-julia}{%
+\subsection{Benchmarking CE in Julia}\label{benchmarking-ce-in-julia}}
 
@@
 
-\hypertarget{sec-future}{%
-\section{Future Projects}\label{sec-future}}
+\hypertarget{probabilistic-methods-for-realistic-counterfactual-explanations}{%
+\subsection{Probabilistic methods for realistic counterfactual
+explanations}\label{probabilistic-methods-for-realistic-counterfactual-explanations}}
+
+Probabilistic machine learning can be leveraged in this context and more
+generally facilitates inference and interpretability. It is also closely
+related to Bayesian statistics, which has played an important role in
+both finance and economics for many years.
+
+To ensure that the generated explanations are realistic it is important
+to understand which input-output pairs are likely and which are not. To
+quantify their joint likelihood, previous work has either relied on
+generative models or restricted the analysis to probabilistic models
+that incorporate uncertainty in their predictions. While the former
+approach is more versatile since it is applicable to both deterministic
+and probabilistic models, the latter is computationally much more
+efficient. In my work I want to explore how recent advances in post-hoc
+uncertainty quantification can be leveraged to generate realistic and
+unambiguous counterfactual explanations for any model.
+
+\hypertarget{ce-for-time-series}{%
+\subsection{CE for time series}\label{ce-for-time-series}}
 
@@
 series models for classification and forecasting in finance and
-economics. For example, I would like to leverage effortless Bayesian
-deep learning to make our proposed Deep Vector Autoregression model
-explainable.
+economics.
+
+\hypertarget{explainable-black-box-models-for-time-series}{%
+\subsection{Explainable black-box models for time
+series}\label{explainable-black-box-models-for-time-series}}
+
+For example, I would like to leverage effortless Bayesian deep learning
+to make our proposed Deep Vector Autoregression model explainable.
 

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.qmd
more work on abstract

@@
 @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift.
-As this game of implementing AR and updating the classifier is repeated, individuals who receive and implement algorithmic recourse end up forming a distinct subgroup inside the target class, which may leave them vulnerable to discrimination.
-Through future experiments we want to investigate if this phenomenon is robust across different benchmark datasets and counterfactual generators.
+As this game of implementing AR and updating the classifier is repeated, the decision boundary moves in the opposite direction of the original training samples in the target class.
+We consider several aspects of these dynamics as problematic.
+Firstly, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as credit worthy.
+Average default risk across all borrowers can therefore be expected to increase.
+Conversely, lenders that anticipate or detect these sorts of dynamics may choose to still deny credit to individuals that have implemented AR, thereby comprimising the validity of AR.
+
+To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse.
+In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets (including German credit [@germancredit1994], Boston Housing [@boston1993] and COMPAS [@compas2016]).
+Furthermore, I want to asses to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
+To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.
 
@@
 
-Alongside my research I have developed open-source implementations related to explainable AI.
-[CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/) is a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R.
-I have recently submitted the package along with a companion paper as a proposal for a main talk at [JuliaCon](https://juliacon.org/2022/).
-[BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/) is a small Julia package that can be used to recover Bayesian representations of deep neural networks through Laplace approximation in a post-hoc manner.
-It is inspired by a recent paper [@daxberger2021laplace] and has also been submitted to JuliaCon.
-Finally, [deepvars](https://github.com/pat-alt/deepvars) is an R package that implements an approach towards vector autoregression that leverages deep learning.
-This was originally my master's thesis and later presented at the NeurIPS 2021 MLECON workshop.
-I have also published several blog posts on explainable AI and probabilisitic ML in an effort to make my research accessible to a broad audience.
+Until recently there existed only one open-source library that provides a unifying approach to generate and benchmark counterfactual explanations for models built and trained in Python [@pawelczyk2021carla].
+To address this limitation I have developed [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/): a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R.
+The package and companion paper are currently pending acceptance for a main talk at [JuliaCon](https://juliacon.org/2022/).
 
-### Probabilistic methods for realistic counterfactual explanations
-
-Probabilistic machine learning can be leveraged in this context and more generally facilitates inference and interpretability.
-It is also closely related to Bayesian statistics, which has played an important role in both finance and economics for many years.
+### Probabilistic Methods for Realistic CE
 
@@
 While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient.
-In my work I want to explore how recent advances in post-hoc uncertainty quantification can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.
+The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category.
+In future work I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]
 
-### CE for time series
+### XAI for Time Series Models
 
-Data sets in finance and economics typically involve time series data.
-Therefore, I am naturally interested in the application of explainable AI to sequential data, an area which has so far not been explored extensively.
-In the future, I want to work on counterfactual explanations for time series models.
-I am also interested in seeing if and how Laplace approximation can be used for Bayesian deep learning with time series data.
-I hope that the findings from both of these projects can ultimately be used to build complex but interpretable time series models for classification and forecasting in finance and economics.
+Datasets in Finance and Economics often involve autoregressive time series, which is modelled differently than static, cross-sectional data.
+One popular approach to macroeconomic forecasting, for example, is Vector Autoregression (VAR), which involves an interpretable system of linear equations.
+Recent work has shown that machine learning can be leveraged in this context to improve forecasting performance, albeit at the sacrifice of explainability [@altmeyer2021deep].^[For an open-source implementation of our approach see [deepvars](https://github.com/pat-alt/deepvars): an R package Deep Vector Autoregression].
+Through my PhD research I hope to ultimately adress that shortfall through explainable AI and probabilistic methods.
 
-### Explainable black-box models for time series
+#### Counterfactual Explanations
 
-For example, I would like to leverage effortless Bayesian deep learning to make our proposed Deep Vector Autoregression model explainable.
+I would like to investigate if and how CE can be used to leverage machine learning for time series forecasting without sacrificing explainability.
+In the context of conventional VAR counterfactual outcomes are actually well-studied [@kilian2017structural], which makes macroeconomic forecasting and interesting case study.
+More generally though, work on applying CE to black-box models for time series data has so far been limited and therefore warrants attention.
+
+#### Probabilistic Methods
+
+I am also interested in seeing if and how Laplace Redux can be applied to reccurent neural networks.
 

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.tex
more work on abstract

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 which in turn causes a drastic model shift. As this game of implementing
-AR and updating the classifier is repeated, individuals who receive and
-implement algorithmic recourse end up forming a distinct subgroup inside
-the target class, which may leave them vulnerable to discrimination.
-Through future experiments we want to investigate if this phenomenon is
-robust across different benchmark datasets and counterfactual
-generators.
+AR and updating the classifier is repeated, the decision boundary moves
+in the opposite direction of the original training samples in the target
+class. We consider several aspects of these dynamics as problematic.
+Firstly, as the decision boundary moves in the direction of the
+non-target class, counterfactual paths become shorter: in the loan
+example, individuals that previously would have been denied credit based
+on their input features are suddenly considered as credit worthy.
+Average default risk across all borrowers can therefore be expected to
+increase. Conversely, lenders that anticipate or detect these sorts of
+dynamics may choose to still deny credit to individuals that have
+implemented AR, thereby comprimising the validity of AR.
+
+To the best of my knowledge this is the first work investigating
+endogenous dynamic in algorithmic recourse. In future experiments I want
+to investigate how this phenomenon plays out across different benchmark
+datasets (including German credit ({``German Credit Data''} 1994),
+Boston Housing ({``Boston Housing Data''} 1993) and COMPAS (Larson et
+al. 2016)). Furthermore, I want to asses to what extent the magnitude
+and direction of domain and model shifts depends on the choice of the
+counterfactual generator. To this end, I am currently supervising a
+group of undergraduate students, who are tackling some of these tasks in
+their final-year research project.
 
@@
 
-Alongside my research I have developed open-source implementations
-related to explainable AI.
-\href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}
-is a Julia package that can be used to generate counterfactual
-explanations for models developed and trained not only in Julia, but
-also in other popular programming languages like Python and R. I have
-recently submitted the package along with a companion paper as a
-proposal for a main talk at \href{https://juliacon.org/2022/}{JuliaCon}.
-\href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}
-is a small Julia package that can be used to recover Bayesian
-representations of deep neural networks through Laplace approximation in
-a post-hoc manner. It is inspired by a recent paper (Daxberger et al.
-2021) and has also been submitted to JuliaCon. Finally,
-\href{https://github.com/pat-alt/deepvars}{deepvars} is an R package
-that implements an approach towards vector autoregression that leverages
-deep learning. This was originally my master's thesis and later
-presented at the NeurIPS 2021 MLECON workshop. I have also published
-several blog posts on explainable AI and probabilisitic ML in an effort
-to make my research accessible to a broad audience.
+Until recently there existed only one open-source library that provides
+a unifying approach to generate and benchmark counterfactual
+explanations for models built and trained in Python (Pawelczyk et al.
+2021). To address this limitation I have developed
+\href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}:
+a Julia package that can be used to generate counterfactual explanations
+for models developed and trained not only in Julia, but also in other
+popular programming languages like Python and R. The package and
+companion paper are currently pending acceptance for a main talk at
+\href{https://juliacon.org/2022/}{JuliaCon}.
 
-\hypertarget{probabilistic-methods-for-realistic-counterfactual-explanations}{%
-\subsection{Probabilistic methods for realistic counterfactual
-explanations}\label{probabilistic-methods-for-realistic-counterfactual-explanations}}
-
-Probabilistic machine learning can be leveraged in this context and more
-generally facilitates inference and interpretability. It is also closely
-related to Bayesian statistics, which has played an important role in
-both finance and economics for many years.
+\hypertarget{probabilistic-methods-for-realistic-ce}{%
+\subsection{Probabilistic Methods for Realistic
+CE}\label{probabilistic-methods-for-realistic-ce}}
 
@@
 and probabilistic models, the latter is computationally much more
-efficient. In my work I want to explore how recent advances in post-hoc
-uncertainty quantification can be leveraged to generate realistic and
-unambiguous counterfactual explanations for any model.
+efficient. The approach proposed by Schut et al. (2021) and used to
+generate the examples in Figure~\ref{fig-dynamics} falls into the latter
+category. In future work I want to explore how recent advances in
+post-hoc uncertainty quantification, most notably Laplace Redux
+(Daxberger et al. 2021), can be leveraged to generate realistic and
+unambiguous counterfactual explanations for any model.\footnote{For some
+  initial work on this see my Julia implementation of Laplace Redux:
+  \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}
 
-\hypertarget{ce-for-time-series}{%
-\subsection{CE for time series}\label{ce-for-time-series}}
+\hypertarget{xai-for-time-series-models}{%
+\subsection{XAI for time series
+models}\label{xai-for-time-series-models}}
 
-Data sets in finance and economics typically involve time series data.
-Therefore, I am naturally interested in the application of explainable
-AI to sequential data, an area which has so far not been explored
-extensively. In the future, I want to work on counterfactual
-explanations for time series models. I am also interested in seeing if
-and how Laplace approximation can be used for Bayesian deep learning
-with time series data. I hope that the findings from both of these
-projects can ultimately be used to build complex but interpretable time
-series models for classification and forecasting in finance and
-economics.
+Datasets in Finance and Economics often involve autoregressive time
+series, which is modelled differently than static, cross-sectional data.
+One popular approach to macroeconomic forecasting, for example, is
+Vector Autoregression (VAR), which involves an interpretable system of
+linear equations. Recent work has shown that machine learning can be
+leveraged in this context to improve forecasting performance, albeit at
+the sacrifice of explainability (Altmeyer, Agusti, and Vidal-Quadras
+Costa 2021).\footnote{For an open-source implementation of our approach
+  see \href{https://github.com/pat-alt/deepvars}{deepvars}: an R package
+  Deep Vector Autoregression}. Through my PhD research I hope to
+ultimately adress that shortfall through explainable AI and
+probabilistic methods.
 
-\hypertarget{explainable-black-box-models-for-time-series}{%
-\subsection{Explainable black-box models for time
-series}\label{explainable-black-box-models-for-time-series}}
+\hypertarget{counterfactual-explanations}{%
+\subsubsection{Counterfactual
+Explanations}\label{counterfactual-explanations}}
 
-For example, I would like to leverage effortless Bayesian deep learning
-to make our proposed Deep Vector Autoregression model explainable.
+I would like to investigate if and how CE can be used to leverage
+machine learning for time series forecasting without sacrificing
+explainability. In the context of conventional VAR counterfactual
+outcomes are actually well-studied (Kilian and Lütkepohl 2017), which
+makes macroeconomic forecasting and interesting case study. More
+generally though, work on applying CE to black-box models for time
+series data has so far been limited and therefore warrants attention.
+
+\hypertarget{probabilistic-methods}{%
+\subsubsection{Probabilistic Methods}\label{probabilistic-methods}}
+
+I am also interested in seeing if and how Laplace approximation can be
+used for Bayesian deep learning with time series data. I hope that the
+findings from both of these projects can ultimately be used to build
+complex but interpretable time series models for classification and
+forecasting in finance and economics.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2021deep}{}}%
+Altmeyer, Patrick, Marc Agusti, and Ignacio Vidal-Quadras Costa. 2021.
+{``Deep Vector Autoregression for Macroeconomic Data.''}
+\url{https://thevoice.bse.eu/wp-content/uploads/2021/07/ds21-project-agusti-et-al.pdf}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-boston1993}{}}%
+{``Boston Housing Data.''} 1993.
+\url{https://archive.ics.uci.edu/ml/machine-learning-databases/housing/}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-daxberger2021laplace}{}}%
@@
 Processing Systems} 34.
+
+\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
+{``German Credit Data.''} 1994.
+\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kilian2017structural}{}}%
+Kilian, Lutz, and Helmut Lütkepohl. 2017. \emph{Structural Vector
+Autoregressive Analysis}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-compas2016}{}}%
+Larson, Jeff, Surya Mattu, Lauren Kirchner, and Julia Angwin. 2016.
+{``COMPAS.''} \url{https://github.com/propublica/compas-analysis}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
@@
 Conference on Fairness, Accountability, and Transparency}, 607--17.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
+Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias
+Richter, and Gjergji Kasneci. 2021. {``Carla: A Python Library to
+Benchmark Algorithmic Recourse and Counterfactual Explanation
+Algorithms.''} \emph{arXiv Preprint arXiv:2108.00783}.
 

### 2022-04-26 dev/submissions/aies/extended_abstract/extended_abstract.qmd
more work on AIES

@@
     pdf:
-        documentclass: acmconf
+        documentclass: acmart
+        classoption: sigconf
         number-sections: true
@@
 In finance, economics and other social sciences, applications of AI are still relatively limited.
-Decision-making in these fields has traditionally been guided by Generalized Linear Models (GLM), which are theoretically founded, interpretable and often sufficient to model relationships between variables.
-Model interpretability is crucial in the social sciences context, because inference is typically at least as important as predictive performance.
+Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations.
+Explainability is crucial in the social sciences context, because inference is typically at least as important as predictive performance.
 Decision-makers in the social sciences are also typically required to explain their decisions to human stakeholders: central bankers, for example, are held accountable by the public for the policies they decide on.
-It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that ultimately cannot be trusted.
-Deep learning models, for example, are generally considered as black boxes and therefore difficult to apply in a context that demands explanations.
+It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that cannot be explained.
+Deep neural networks, for example, are generally considered as black boxes and therefore not trustworthy in a context that demands explanations.
 This PhD project is focused on exploring and developing methodologies that improve the trustworthiness of AI and thereby enable its application in Finance and Economics.
@@
 Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals facing adverse decisions.
-An example relevant to the Finance and Econonomics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .
+An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .
 
-Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], plausible [@joshi2019towards, @antoran2020getting, @schut2021generating], sparse [@schut2021generating], diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
+Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
 The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
@@
 
-This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually impemented by a proportion of individuals and the classifier is updated in response.
+This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
 @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift.
@@
 To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse.
-In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets (including German credit [@germancredit1994], Boston Housing [@boston1993] and COMPAS [@compas2016]).
-Furthermore, I want to asses to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
+In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.
+Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
 To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.
 
-![PLACEHOLDER: The dynamics of algorithmic recourse.](www/dynamics.png){#fig-dynamics fig.pos="h" width=250px}
+![PLACEHOLDER: The dynamics of algorithmic recourse.](){#fig-dynamics fig.pos="h" height=100px}
 
@@
 To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not.
-To quantify their joint likelihood, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions.
-While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient.
+With respect to the work-in-progress presented here, I expect that this may help in mitigating endogenous domain and model shifts.
+To quantify the joint likelihood of input-output pairs, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions.
+While the former approach is more versatile, since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient.
 The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category.
-In future work I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]
-
-### XAI for Time Series Models
-
-Datasets in Finance and Economics often involve autoregressive time series, which is modelled differently than static, cross-sectional data.
-One popular approach to macroeconomic forecasting, for example, is Vector Autoregression (VAR), which involves an interpretable system of linear equations.
-Recent work has shown that machine learning can be leveraged in this context to improve forecasting performance, albeit at the sacrifice of explainability [@altmeyer2021deep].^[For an open-source implementation of our approach see [deepvars](https://github.com/pat-alt/deepvars): an R package Deep Vector Autoregression].
-Through my PhD research I hope to ultimately adress that shortfall through explainable AI and probabilistic methods.
-
-#### Counterfactual Explanations
-
-I would like to investigate if and how CE can be used to leverage machine learning for time series forecasting without sacrificing explainability.
-In the context of conventional VAR counterfactual outcomes are actually well-studied [@kilian2017structural], which makes macroeconomic forecasting and interesting case study.
-More generally though, work on applying CE to black-box models for time series data has so far been limited and therefore warrants attention.
-
-#### Probabilistic Methods
-
-I am also interested in seeing if and how Laplace Redux can be applied to reccurent neural networks.
+In future work, I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]
 

### 2022-04-26 dev/submissions/aies/extended_abstract/extended_abstract.tex
more work on AIES

@@
 \documentclass[
-]{acmconf}
-\usepackage{amsmath,amssymb}
+  sigconf]{acmart}
+\usepackage{amsmath}
 \usepackage{lmodern}
@@
   pdftitle={Dynamics in Algorithmic Recourse},
-  pdfauthor={Patrick Altmeyer},
   colorlinks=true,
@@
 
-\ConferenceShortName{AIES '22}
-\ConferenceName{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}
+% \ConferenceShortName{AIES '22}
+% \ConferenceName{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}
+
+\setcopyright{acmcopyright}
+\copyrightyear{2022}
+\acmYear{2022}
+\acmDOI{XXXXXXX.XXXXXXX}
+
+\acmConference[AIES '22]{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}{Oxford, UK}
+
+\author{Patrick Altmeyer}
+\affiliation{%
+  \institution{Delft University of Technology}
+  \city{Delft}
+  \country{The Netherlands}
+}
+\email{p.altmeyer@tudelft.nl}
+\orcid{0000-0003-4726-8613}
 \makeatletter
@@
 \subtitle{Trustworthy Artificial Intelligence for Finance and Economics}
-\author{Patrick Altmeyer}
-\date{April 25, 2022}
+\author{}
+\date{April 26, 2022}
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 Decision-making in these fields has traditionally been guided by
-Generalized Linear Models (GLM), which are theoretically founded,
-interpretable and often sufficient to model relationships between
-variables. Model interpretability is crucial in the social sciences
-context, because inference is typically at least as important as
-predictive performance. Decision-makers in the social sciences are also
-typically required to explain their decisions to human stakeholders:
-central bankers, for example, are held accountable by the public for the
-policies they decide on. It is therefore not surprising that
-practitioners and academics in these fields are reluctant to adopt AI
-technologies that ultimately cannot be trusted. Deep learning models,
-for example, are generally considered as black boxes and therefore
-difficult to apply in a context that demands explanations. This PhD
-project is focused on exploring and developing methodologies that
-improve the trustworthiness of AI and thereby enable its application in
-Finance and Economics.
+interpretable models that facilitate explanations. Explainability is
+crucial in the social sciences context, because inference is typically
+at least as important as predictive performance. Decision-makers in the
+social sciences are also typically required to explain their decisions
+to human stakeholders: central bankers, for example, are held
+accountable by the public for the policies they decide on. It is
+therefore not surprising that practitioners and academics in these
+fields are reluctant to adopt AI technologies that cannot be explained.
+Deep neural networks, for example, are generally considered as black
+boxes and therefore not trustworthy in a context that demands
+explanations. This PhD project is focused on exploring and developing
+methodologies that improve the trustworthiness of AI and thereby enable
+its application in Finance and Economics.
 
@@
 (AR) to help individuals facing adverse decisions. An example relevant
-to the Finance and Econonomics domain is consumer credit: in this
-context AR can be used to guide individuals to improve their credit
-worthiness, should they have previously been denied access to credit
-based on a black-box decision-making system .
+to the Finance and Economics domain is consumer credit: in this context
+AR can be used to guide individuals to improve their credit worthiness,
+should they have previously been denied access to credit based on a
+black-box decision-making system .
 
@@
 are interested in finding close (Wachter, Mittelstadt, and Russell
-2017), actionable (Ustun, Spangher, and Liu 2019), plausible Schut et
-al. (2021), sparse (Schut et al. 2021), diverse (Mothilal, Sharma, and
-Tan 2020) and ideally causally founded counterfactual explanations
-(Karimi, Schölkopf, and Valera 2021) for some individual \(x\). The
-ability of counterfactual explanations to handle dynamics like data and
-model shifts remains a largely unexplored research challenge at this
-point (Verma, Dickerson, and Hines 2020). Only one recent work considers
-the implications of \textbf{exogenous} domain shifts on the validity of
-recourse (Upadhyay, Joshi, and Lakkaraju 2021). The authors propose a
-simple minimax objective, that minimizes the counterfactual loss
-function for a maximal domain and model shift. They show that their
-approach yields more robust counterfactuals than existing approaches.
+2017), actionable (Ustun, Spangher, and Liu 2019), realistic Schut et
+al. (2021), sparse, diverse (Mothilal, Sharma, and Tan 2020) and ideally
+causally founded counterfactual explanations (Karimi, Schölkopf, and
+Valera 2021) for some individual \(x\). The ability of counterfactual
+explanations to handle dynamics like data and model shifts remains a
+largely unexplored research challenge at this point (Verma, Dickerson,
+and Hines 2020). Only one recent work considers the implications of
+\textbf{exogenous} domain shifts on the validity of recourse (Upadhyay,
+Joshi, and Lakkaraju 2021). The authors propose a simple minimax
+objective, that minimizes the counterfactual loss function for a maximal
+domain and model shift. They show that their approach yields more robust
+counterfactuals than existing approaches.
 
 This project investigates \textbf{endogenous} domain and model shifts,
-that is shifts that occur when AR is actually impemented by a proportion
-of individuals and the classifier is updated in response.
+that is shifts that occur when AR is actually implemented by a
+proportion of individuals and the classifier is updated in response.
 Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
@@
 to investigate how this phenomenon plays out across different benchmark
-datasets (including German credit ({``German Credit Data''} 1994),
-Boston Housing ({``Boston Housing Data''} 1993) and COMPAS (Larson et
-al. 2016)). Furthermore, I want to asses to what extent the magnitude
-and direction of domain and model shifts depends on the choice of the
-counterfactual generator. To this end, I am currently supervising a
-group of undergraduate students, who are tackling some of these tasks in
-their final-year research project.
+datasets including German credit, Boston Housing and COMPAS.
+Furthermore, I want to assess to what extent the magnitude and direction
+of domain and model shifts depends on the choice of the counterfactual
+generator. To this end, I am currently supervising a group of
+undergraduate students, who are tackling some of these tasks in their
+final-year research project.
 
@@
 
-{\centering \includegraphics[width=2.60417in,height=\textheight]{www/dynamics.png}
+{\centering \includegraphics[width=\textwidth,height=1.04167in]{.pdf}
 
@@
 To ensure that the generated explanations are realistic it is important
-to understand which input-output pairs are likely and which are not. To
-quantify their joint likelihood, previous work has either relied on
-generative models or restricted the analysis to probabilistic models
-that incorporate uncertainty in their predictions. While the former
-approach is more versatile since it is applicable to both deterministic
-and probabilistic models, the latter is computationally much more
-efficient. The approach proposed by Schut et al. (2021) and used to
-generate the examples in Figure~\ref{fig-dynamics} falls into the latter
-category. In future work I want to explore how recent advances in
-post-hoc uncertainty quantification, most notably Laplace Redux
-(Daxberger et al. 2021), can be leveraged to generate realistic and
-unambiguous counterfactual explanations for any model.\footnote{For some
-  initial work on this see my Julia implementation of Laplace Redux:
+to understand which input-output pairs are likely and which are not.
+With respect to the work-in-progress presented here, I expect that this
+may help in mitigating endogenous domain and model shifts. To quantify
+the joint likelihood of input-output pairs, previous work has either
+relied on generative models or restricted the analysis to probabilistic
+models that incorporate uncertainty in their predictions. While the
+former approach is more versatile, since it is applicable to both
+deterministic and probabilistic models, the latter is computationally
+much more efficient. The approach proposed by Schut et al. (2021) and
+used to generate the examples in Figure~\ref{fig-dynamics} falls into
+the latter category. In future work, I want to explore how recent
+advances in post-hoc uncertainty quantification, most notably Laplace
+Redux (Daxberger et al. 2021), can be leveraged to generate realistic
+and unambiguous counterfactual explanations for any model.\footnote{For
+  some initial work on this see my Julia implementation of Laplace
+  Redux:
   \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}
-
-\hypertarget{xai-for-time-series-models}{%
-\subsection{XAI for time series
-models}\label{xai-for-time-series-models}}
-
-Datasets in Finance and Economics often involve autoregressive time
-series, which is modelled differently than static, cross-sectional data.
-One popular approach to macroeconomic forecasting, for example, is
-Vector Autoregression (VAR), which involves an interpretable system of
-linear equations. Recent work has shown that machine learning can be
-leveraged in this context to improve forecasting performance, albeit at
-the sacrifice of explainability (Altmeyer, Agusti, and Vidal-Quadras
-Costa 2021).\footnote{For an open-source implementation of our approach
-  see \href{https://github.com/pat-alt/deepvars}{deepvars}: an R package
-  Deep Vector Autoregression}. Through my PhD research I hope to
-ultimately adress that shortfall through explainable AI and
-probabilistic methods.
-
-\hypertarget{counterfactual-explanations}{%
-\subsubsection{Counterfactual
-Explanations}\label{counterfactual-explanations}}
-
-I would like to investigate if and how CE can be used to leverage
-machine learning for time series forecasting without sacrificing
-explainability. In the context of conventional VAR counterfactual
-outcomes are actually well-studied (Kilian and Lütkepohl 2017), which
-makes macroeconomic forecasting and interesting case study. More
-generally though, work on applying CE to black-box models for time
-series data has so far been limited and therefore warrants attention.
-
-\hypertarget{probabilistic-methods}{%
-\subsubsection{Probabilistic Methods}\label{probabilistic-methods}}
-
-I am also interested in seeing if and how Laplace approximation can be
-used for Bayesian deep learning with time series data. I hope that the
-findings from both of these projects can ultimately be used to build
-complex but interpretable time series models for classification and
-forecasting in finance and economics.
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2021deep}{}}%
-Altmeyer, Patrick, Marc Agusti, and Ignacio Vidal-Quadras Costa. 2021.
-{``Deep Vector Autoregression for Macroeconomic Data.''}
-\url{https://thevoice.bse.eu/wp-content/uploads/2021/07/ds21-project-agusti-et-al.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
-Miguel Hernández-Lobato. 2020. {``Getting a Clue: A Method for
-Explaining Uncertainty Estimates.''} \emph{arXiv Preprint
-arXiv:2006.06848}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-boston1993}{}}%
-{``Boston Housing Data.''} 1993.
-\url{https://archive.ics.uci.edu/ml/machine-learning-databases/housing/}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-daxberger2021laplace}{}}%
@@
 Processing Systems} 34.
-
-\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
-{``German Credit Data.''} 1994.
-\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.
 
@@
 Fairness, Accountability, and Transparency}, 353--62.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kilian2017structural}{}}%
-Kilian, Lutz, and Helmut Lütkepohl. 2017. \emph{Structural Vector
-Autoregressive Analysis}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-compas2016}{}}%
-Larson, Jeff, Surya Mattu, Lauren Kirchner, and Julia Angwin. 2016.
-{``COMPAS.''} \url{https://github.com/propublica/compas-analysis}.
 

### 2022-04-29 dev/work/synthetic_bayesian_ar.qmd
aies

@@
 include("../src/load.jl")
-using AlgorithmicRecourse
+using CounterfactualExplanations
 using CSV;
@@
 ```{julia}
-using AlgorithmicRecourse.Models: BayesianLogisticModel
-𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ); # repare for usage with AlgorithmicRecourse.jl
+using CounterfactualExplanations.Models: BayesianLogisticModel
+𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ); # repare for usage with CounterfactualExplanations.jl
 x̅ = [50.0, 60.0]; # factual
@@
 Random.seed!(1234)
-y̅ = round(AlgorithmicRecourse.Models.probs(𝑴, x̅)[1])
+y̅ = round(CounterfactualExplanations.Models.probs(𝑴, x̅)[1])
 target = ifelse(y̅==1.0,0.0,1.0) # opposite label as target

### 2022-04-29 dev/work/synthetic_endogenous_shifts.qmd
aies

@@
 ---
+title: Synthetic data
 jupyter: julia-1.6
 ---
-
-# Endogenous model shifts
 
@@
 ```{julia}
-include("dev/utils.jl")
-using CounterfactualExplanations, CSV, DataFrames, LinearAlgebra, Distributions, Images, FileIO, Plots, PlotThemes
+include("src/load.jl")
+using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, BayesLaplace, LinearAlgebra
 using Base.Filesystem: joinpath
@@
 disable_logging(Logging.Info)
-output_folder = "output/synthetic"
-www_folder = "www/synthetic"
+output_path = "dev/work/output/synthetic"
+www_path = "dev/work/www/synthetic"
 ```
 
-## Single-period
-
-### Training the classifier
-
-We start by training a simple linear classifier: Bayesian Logisitic Regression.
+## Classifiers
 
 ```{julia}
-df = CSV.read("../data/cats_dogs_large.csv", DataFrame);
-y = df[:,:y];
-N = length(y);
-X = Matrix(df[:,Not(:y)]);
-model = Models.bayes_logreg(X,y); # get MAP estimates
-w = model.μ;
+# Number of points to generate.
+N = 100
+using CounterfactualExplanations.Data
+xs, ys = Data.toy_data_linear(N)
+X = hcat(xs...) # bring into tabular format
+data = zip(xs,ys)
+```
+
+### Logistic Regression
+
+```{julia}
+nn = Chain(Dense(2,1))
+λ = 0.5
+sqnorm(x) = sum(abs2, x)
+weight_regularization(λ=λ) = 1/2 * λ^2 * sum(sqnorm, Flux.params(nn))
+loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y) + weight_regularization();
 ```
@@
 ```{julia}
-# Plot data points.
-plt = plot()
-using Main.Experiments: plot_data!
-plot_data!(plt,X,y)
+using Flux.Optimise: update!, ADAM
+opt = ADAM()
+epochs = 50
+function _forward_nn(nn, data, epochs)
+    for epoch = 1:epochs
+        for d in data
+            gs = gradient(params(nn)) do
+            l = loss(d...)
+            end
+            update!(opt, params(nn), gs)
+        end
+    end
+    return nn
+end
+_forward_nn(nn, data, epochs)
 ```
@@
 ```{julia}
-using AlgorithmicRecourse.Models: BayesianLogisticModel, LogisticModel
-𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ) # fully Bayesian model
-𝑴_plugin = LogisticModel(reshape(model.μ[2:3],1,2), [model.μ[1]]); # Plugin estimate (non-Bayesian)
-models = (Bayesian=𝑴, Plugin=𝑴_plugin);
+using CounterfactualExplanations.Models
+import CounterfactualExplanations.Models: logits, probs # import functions in order to extend
+import Main.Models: retrain
+
+# Step 1)
+struct NeuralNetwork <: Models.AbstractFittedModel
+    model::Any
+end
+
+# Step 2)
+logits(M::NeuralNetwork, X::AbstractArray) = M.model(X)
+probs(M::NeuralNetwork, X::AbstractArray)= σ.(logits(M, X))
+
+# Step 3)
+function retrain(M::NeuralNetwork, data; n_epochs=10, τ=1.0) 
+    nn = M.model
+    nn = _forward_nn(nn, data, n_epochs)
+    M = NeuralNetwork(nn)
+    return M
+end
+
+M = NeuralNetwork(nn)
 ```
@@
 ```{julia}
-using Main.Experiments: plot_contour
-p1 = plot_contour(X,y,𝑴,title="Bayesian")
-p2 = plot_contour(X,y,𝑴_plugin,title="Plugin")
-plot(p1,p2,size=(1000,400))
+#| echo: false
+# Plot the posterior distribution with a contour plot.
+plt = plot_contour(X',ys,M);
+savefig(plt, joinpath(www_path, "nn_contour.png"))
 ```
 
-### Generating recourse
+![](www/nn_contour.png)
 
-Now we will generate and implement algorithmic recourse for a subset of cats.
+### Laplace Redux
 
 ```{julia}
-generators = (Bayesian=GreedyGenerator(1,500,:logitbinarycrossentropy, nothing), Plugin=GenericGenerator(0.1,0.1,1e-5,:logitbinarycrossentropy,nothing));
-proportion = 0.2
-using StatsBase
-cats = findall(df.y.==0);
-chosen_individuals = StatsBase.sample(cats,Int(round(proportion * length(cats))),replace=false);
-target = 1.0
-γ = 0.90;
+la = Laplace(nn, λ=λ)
+BayesLaplace.fit!(la, data)
 ```
@@
 ```{julia}
-function generate_counterfactuals(generator, 𝑴, target, γ, chosen_individuals; X=X, y=y)
-    X̲ = copy(X)
-    y̲ = copy(y)
-    for i in chosen_individuals
-        x̅ = X[i,:]
-        recourse = generate_recourse(generator, x̅, 𝑴, target, γ);
-        X̲[i,:] = recourse.x̲
-        y̲[i] = recourse.y̲
+# Step 1)
+struct LaplaceNeuralNetwork <: Models.AbstractFittedModel
+    la::BayesLaplace.Laplace
 end
-    return X̲, y̲
+
+# Step 2)
+logits(M::LaplaceNeuralNetwork, X::AbstractArray) = M.la.model(X)
+probs(M::LaplaceNeuralNetwork, X::AbstractArray)= BayesLaplace.predict(M.la, X)
+
+# Step 3)
+function retrain(M::LaplaceNeuralNetwork, data; n_epochs=10, τ=1.0) 
+    nn = M.la.model
+    nn = _forward_nn(nn, data, n_epochs)
+    la = Laplace(nn, λ=λ)
+    BayesLaplace.fit!(la, data)
+    M = LaplaceNeuralNetwork(la)
+    return M
 end
+
+Mᴸ = LaplaceNeuralNetwork(la)
 ```
@@
 ```{julia}
-plt = []
-for i in 1:length(generators)
-    X̲, y̲ = generate_counterfactuals(generators[i], models[i], target, γ, chosen_individuals)
-    pltᵢ = plot_contour(X̲,y̲,models[i],title=keys(generators)[i])
-    plt = vcat(plt, pltᵢ)
-end
-plot(plt..., size = (1000,400))
+#| echo: false
+# Plot the posterior distribution with a contour plot.
+plt = plot_contour(X',ys,Mᴸ);
+savefig(plt, joinpath(www_path, "la_contour.png"))
 ```
 
-### Retraining classifier
+![](www/la_contour.png)
 
-Finally, we will retrain the classifier (model shift) and plot the resulting new decision boundary and posterior predictive distribution.
+## Experiment
 
 ```{julia}
-plt = []
-for i in 1:length(generators)
-    X̲, y̲ = generate_counterfactuals(generators[i], models[i], target, γ, chosen_individuals)
-    data = Models.prepare_data(X̲', y̲')
-    𝑴 = Models.retrain(models[i], data)
-    pltᵢ = plot_contour(X̲,y̲,𝑴,title=keys(generators)[i])
-    plt = vcat(plt, pltᵢ)
-end
-plot(plt..., size = (1000,400))
-```
+# Models:
+models = (Bayesian=Mᴸ, Plugin=M)
 
-## Dynamic
+# Generators:
+generators = (Bayesian=GreedyGenerator(), Plugin=GenericGenerator())
 
-```{julia}
 # Variables:
@@
 target = 1.0
-T = 1000;
-# Experiments:
-experiment = Experiments.Experiment(X',y',models[:Bayesian],target,grid_,n_rounds)
-experiment_plugin = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
-experiments = (experiment, experiment_plugin);
+T = 1000
 ```
@@
 ```{julia}
-run = false
-if run
-    using Random
+# Experiments:
+experiment = Experiments.Experiment(X,ys,models[:Bayesian],target,grid_,n_rounds)
+experiment_plugin = Experiments.Experiment(X,ys,models[:Plugin],target,grid_,n_rounds)
+experiments = (experiment, experiment_plugin)
+```
+
+```{julia}
+#| eval: false
+using Random, CSV, DataFrames
 Random.seed!(1234)
@@
     outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
-        Experiments.save_path(joinpath(output_folder, string(keys(generators)[j])),path)
-        CSV.write(joinpath(output_folder, string(keys(generators)[j]) * ".csv"), DataFrame(outcome))
-    end
+    Experiments.save_path(joinpath(output_path, string(keys(generators)[j])),path)
+    CSV.write(joinpath(output_path, string(keys(generators)[j]) * ".csv"), DataFrame(outcome))
 end
@@
 ```{julia}
-run = false
-if run
+#| eval: false
 using Main.Experiments: plot_path
-    paths = [Experiments.load_path(joinpath(output_folder, string(keys(generators)[j]))) for j ∈ 1:length(generators)];
+paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)];
 anim = Experiments.plot_path(paths[1],Γ=γ,𝙈=μ)
-    gif(anim,joinpath(www_folder,"bayesian.gif"), fps=2)
+gif(anim,joinpath(www_path,"bayesian.gif"), fps=2)
 anim = Experiments.plot_path(paths[2],Γ=γ,𝙈=μ)
-    gif(anim,joinpath(www_folder,"plugin.gif"), fps=2)
-end
+gif(anim,joinpath(www_path,"plugin.gif"), fps=2)
 ```
@@
 cost = DataFrame()
-files = (Bayesian=joinpath(output_folder, "Bayesian.csv"), Plugin=joinpath(output_folder, "Plugin.csv"))
+files = (Bayesian=joinpath(output_path, "Bayesian.csv"), Plugin=joinpath(output_path, "Plugin.csv"))
 results = map(file -> CSV.read(file, DataFrame), files)
@@
 
-  gif(anim, joinpath(www_folder, "single_nn.gif"), fps=10)
+  gif(anim, joinpath(www_path
+, "single_nn.gif"), fps=10)
 
-  BSON.@save joinpath(output_folder, "nn.bson") nn
+  BSON.@save joinpath(output_path
+, "nn.bson") nn
 
@@
 using BSON: @load
-@load joinpath(output_folder, "nn.bson") nn
-𝑴ₙₙ = Models.FittedNeuralNet(nn, opt, loss);
+@load joinpath(output_path, "nn.bson") nn
+Mₙₙ = Models.FittedNeuralNet(nn, opt, loss);
 ```
@@
     ensemble, anim = Models.forward(ensemble, data, opt, n_epochs=10, plot_every=10, loss_type=loss_type, τ=1.1); # fit the ensemble
-    Models.save_ensemble(ensemble, root=joinpath(output_folder, "ensemble")) # save to disk
-    gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
+    Models.save_ensemble(ensemble, root=joinpath(output_path
+, "ensemble")) # save to disk
+    gif(anim, joinpath(www_path
+, "ensemble_loss.gif"), fps=25);
 end
@@
 ```{julia}
-ensemble = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
-𝑴 = Models.FittedEnsemble(ensemble, opt, loss_type)
-models = (Bayesian=𝑴, Plugin=𝑴ₙₙ);
+ensemble = Models.load_ensemble(root=joinpath(output_path, "ensemble"))
+M = Models.FittedEnsemble(ensemble, opt, loss_type)
+models = (Bayesian=M, Plugin=Mₙₙ);
 ```
@@
         outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
-        Experiments.save_path(joinpath(output_folder, string(keys(generators)[j])),path)
-        CSV.write(joinpath(output_folder, string(keys(generators)[j]) * "_deep.csv"), DataFrame(outcome))
+        Experiments.save_path(joinpath(output_path
+    , string(keys(generators)[j])),path)
+        CSV.write(joinpath(output_path
+    , string(keys(generators)[j]) * "_deep.csv"), DataFrame(outcome))
     end

### 2022-04-29 dev/utils.jl

@@ -1,51 +1,0 @@
-# Compute cartesian product over two vectors:
-function expandgrid(x,y)
-    N = length(x) * length(y)
-    grid = Iterators.product(x,y) |>
-        Iterators.flatten |>
-        collect |>
-        z -> reshape(z, (2,N)) |>
-        transpose |>
-        Matrix
-    return grid
-end
-
-# Train-test split
-using StatsBase
-function train_test_split(X,y;test_size=0.2)
-    N = size(X)[1]
-    rows = 1:N
-    test_rows = StatsBase.sample(rows,Int(floor(test_size * N)),replace=false)
-    train_rows = setdiff(rows, test_rows)
-    X_train = X[train_rows,:]
-    y_train = y[train_rows]
-    X_test = X[test_rows,:]
-    y_test = y[test_rows]
-    return X_train, y_train, X_test, y_test
-end
-
-# K-fold
-using MLDataUtils
-function kfolds_(N;k=5)
-    rows = 1:N
-    rows_shuffled = StatsBase.sample(rows,N,replace=false)
-    folds = kfolds(rows_shuffled, k)
-    train_indices = [folds.data[idx] for idx in folds.train_indices]
-    test_indices = [folds.data[idx] for idx in folds.val_indices]
-    return train_indices, test_indices
-end
-
-using StatsBase
-function scale(X, dim)
-    dt = fit(ZScoreTransform, X, dim=dim)
-    X_scaled = StatsBase.transform(dt, X)
-    return X_scaled, dt
-end
-
-function rescale(X, dt)
-    X_rescaled = StatsBase.reconstruct(dt, X)
-    return X_rescaled
-end
-    
-
-

### 2022-04-29 src/experiments/utils.jl

@@ -93,7 +93,7 @@ function plot_data!(plt,X,y)
     Plots.scatter!(plt, X[y.==0.0,1],X[y.==0.0,2], color=0, clim = (0,1), label="y=0")
 end
 
-using Plots, AlgorithmicRecourse
+using Plots, CounterfactualExplanations
 """
     plot_contour(X,y,𝑴;clegend=true,title="",length_out=100)
 
@@ -102,7 +102,7 @@ Helper function to plot contour of predictive probabilities.
 function plot_contour(X,y,𝑴;clegend=true,title="",length_out=100)
     x_range = collect(range(minimum(X[:,1]),stop=maximum(X[:,1]),length=length_out))
     y_range = collect(range(minimum(X[:,2]),stop=maximum(X[:,2]),length=length_out))
-    Z = [AlgorithmicRecourse.Models.probs(𝑴,[x, y])[1] for x=x_range, y=y_range]
+    Z = [CounterfactualExplanations.Models.probs(𝑴,[x, y])[1] for x=x_range, y=y_range]
     plt = contourf(x_range, y_range, Z', color=:viridis, legend=clegend, title=title, linewidth=0)
     plot_data!(plt,X,y)
 end

### 2022-04-29 src/generate_recourse.jl

@@ -1,189 +1,0 @@
-# ------------------------- RECOURSE constructor -------------------------
-mutable struct Recourse
-    x_cf::Vector{Float64}
-    y_cf::Float64
-    path::Matrix{Float64}
-    target::Float64
-    valid::Bool
-    cost::Float64
-    factual::Vector{Float64}
-end;
-
-# --------------- Outer constructor methods: 
-function valid(self::Recourse; classifier=nothing)
-    if isnothing(classifier)
-        valid = self.valid
-    else 
-        valid = predict(classifier, vcat(1, self.x_cf); proba=false)[1] == self.target
-    end
-    return valid
-end
-
-function cost(self::Recourse; cost_fun=nothing, cost_fun_kargs)
-    if isnothing(cost_fun)
-        cost = self.cost
-    else 
-        cost = cost_fun(self.factual .- self.x_cf; cost_fun_kargs...)
-        self.cost = cost
-    end
-    return cost
-end
-
-# ------------------------- RECOUSE generators -------------------------
-
-# --------------- Wachter et al (2018): 
-function gradient_cost(x_f, x_cf)
-    (x_cf .- x_f) ./ norm(x_cf .- x_f)
-end;
-
-function generate_recourse_wachter(x, gradient, classifier, target; T=1000, immutable_=[], a=1, τ=1e-5, λ=0.1, gradient_cost=gradient_cost)
-    # Setup:
-    w = coef(classifier)
-    constant_needed = length(w) > length(x) # is adjustment for constant needed?
-    if (constant_needed)
-        x = vcat(1, x)
-        immutable_ = vcat(1, immutable_ .+ 1) # adjust mask for immutable features
-    end
-    x_cf = copy(x) # start from factual
-    D = length(x_cf)
-    path = reshape(x, 1, length(x)) # storing the path
-    function convergence_condition(x_cf, gradient, w, target, tol)
-        all(gradient(x_cf,w,target) .<= τ)
-    end
-    
-    # Initialize:
-    t = 1 # counter
-    converged = convergence_condition(x_cf, gradient, w, target, τ)
-    
-    # Recursion:
-    while !converged && t < T 
-        g_t = gradient(x_cf,w,target) # compute gradient
-        g_t[immutable_] .= 0 # set gradient of immutable features to zero
-        g_cost_t = gradient_cost(x,x_cf) # compute gradient of cost function
-        g_cost_t[immutable_] .= 0 # set gradient of immutable features to zero
[…]

### 2022-04-29 src/gradients.jl

@@ -1,24 +1,0 @@
-# Hinge:
-function gradient(x,w,y)
-    z = ifelse(y==1,1,-1)
-    if z.*w'x <= 1
-        𝐠 = -z .* w
-    else
-        𝐠 = zeros(length(w))
-    end
-    return 𝐠
-end;
-
-# MSE:
-# function gradient(x,w,y)
-#     𝐠 = 2 * 𝛔(w'x) * (1-𝛔(w'x)) * (𝛔(w'x) - y) .* w
-#     return 𝐠
-# end;
-
-# Cross-entropy:
-function gradient(x,w,y)
-    𝐠 = (𝛔(w'x) - y) .* w
-    return 𝐠
-end;
-
-

### 2022-04-29 src/models/utils.jl

@@ -5,7 +5,6 @@ Helper function that prepares the data for training and moves arrays to GPU if a
 """
 function prepare_data(X::AbstractArray, y::AbstractArray)
     xs = Flux.unstack(X,2) |> gpu
-    y = Flux.unstack(y,2) |> gpu
     data = zip(xs,y)
     return data
 end

### 2022-04-29 src/utils.jl

@@ -1,0 +1,345 @@
+# Compute cartesian product over two vectors:
+function expandgrid(x,y)
+    N = length(x) * length(y)
+    grid = Iterators.product(x,y) |>
+        Iterators.flatten |>
+        collect |>
+        z -> reshape(z, (2,N)) |>
+        transpose |>
+        Matrix
+    return grid
+end
+
+# Train-test split
+using StatsBase
+function train_test_split(X,y;test_size=0.2)
+    N = size(X)[1]
+    rows = 1:N
+    test_rows = StatsBase.sample(rows,Int(floor(test_size * N)),replace=false)
+    train_rows = setdiff(rows, test_rows)
+    X_train = X[train_rows,:]
+    y_train = y[train_rows]
+    X_test = X[test_rows,:]
+    y_test = y[test_rows]
+    return X_train, y_train, X_test, y_test
+end
+
+# K-fold
+using MLDataUtils
+function kfolds_(N;k=5)
+    rows = 1:N
+    rows_shuffled = StatsBase.sample(rows,N,replace=false)
+    folds = kfolds(rows_shuffled, k)
+    train_indices = [folds.data[idx] for idx in folds.train_indices]
+    test_indices = [folds.data[idx] for idx in folds.val_indices]
+    return train_indices, test_indices
+end
+
+using StatsBase
+function scale(X, dim)
+    dt = fit(ZScoreTransform, X, dim=dim)
+    X_scaled = StatsBase.transform(dt, X)
+    return X_scaled, dt
+end
+
+function rescale(X, dt)
+    X_rescaled = StatsBase.reconstruct(dt, X)
+    return X_rescaled
+end
+    
+
+"""
+    build_model()
+
+Helper function to build simple MLP.
+
+# Examples
+
+```julia-repl
+using BayesLaplace
+nn = build_model()
+```
+
+"""
+function build_model(;input_dim=2,n_hidden=32,output_dim=1,batch_norm=false,dropout=false,activation=Flux.relu)
+    
+    if batch_norm
+        nn = Chain(
+            Dense(input_dim, n_hidden),
+            BatchNorm(n_hidden, activation),
+            Dense(n_hidden, output_dim),
+            BatchNorm(output_dim)
+        )  
+    elseif dropout
+        nn = Chain(
+            Dense(input_dim, n_hidden, activation),
+            Dropout(0.1),
+            Dense(n_hidden, output_dim)
[…]

### 2022-04-29 dev/submissions/aies/sections/support.qmd
Fun with latex templates

@@
+---
+format:
+    pdf: default
+---
+
 ### Letter of Support
 
-Patrick has been my supervisee for ...
+As substantiated in more detail in my full letter of support (attached with this application), as Patrick's advisor, I am very happy to support his application to the AIES student track.
+Patrick is an eager student, with a pro-active enthusiasm to learn from others, and contribute himself to the learning progress of relevant communities.
+Getting him to present at AIES will be very valuable for receiving feedback on his current work, while I would more broadly be happy to receive the community's guidance on how to best navigate questions of ethics and responsibility in a project like his.
+In case of a positive outcome of this application, we are in a position to fully cover Patrick's costs, and thus will not lay any claim on the financial waivers.

### 2022-04-29 dev/submissions/aies/extended_abstract/extended_abstract.tex
adapted code to changes to CounterfactualExplanations package. Still need to rerun remaining experiments, for now just ran POC for AIES.

@@
 \urlstyle{same} % disable monospaced font for URLs
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
 \newlength{\cslhangindent}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
 % \ConferenceShortName{AIES '22}
@@
 \AtBeginDocument{%
-\ifdefined\contentsname
 \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
 \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
 \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
 \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
 \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
 }
@@
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
 \makeatother
@@
 \author{}
-\date{April 26, 2022}
+\date{today}
 
@@
 \maketitle
-
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-{\centering \includegraphics[width=\textwidth,height=1.04167in]{.pdf}
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/poc.png}
 
@@
 
-\caption{\label{fig-dynamics}PLACEHOLDER: The dynamics of algorithmic
-recourse.}
+\caption{\label{fig-dynamics}Dynamics in Algorithmic Recourse.}
 

### 2022-04-29 dev/work/synthetic_endogenous_shifts.qmd
adapted code to changes to CounterfactualExplanations package. Still need to rerun remaining experiments, for now just ran POC for AIES.

@@
 ```{julia}
-# Number of points to generate.
-N = 100
-using CounterfactualExplanations.Data
-xs, ys = Data.toy_data_linear(N)
-X = hcat(xs...) # bring into tabular format
+Random.seed!(123)
+using MLJ
+N = 1000
+X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-2 => 2), cluster_std=0.1)
+ys .= ys.==2
+X = X'
+xs = Flux.unstack(X,2)
 data = zip(xs,ys)
+plt = plot()
+plot_data!(plt,X',ys)
 ```
@@
 nn = Chain(Dense(2,1))
-λ = 0.5
-sqnorm(x) = sum(abs2, x)
-weight_regularization(λ=λ) = 1/2 * λ^2 * sum(sqnorm, Flux.params(nn))
-loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y) + weight_regularization();
+loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y) 
 ```
@@
 opt = ADAM()
-epochs = 50
+epochs = 10
 function _forward_nn(nn, data, epochs)
@@
         for d in data
-            gs = gradient(params(nn)) do
+            gs = gradient(Flux.params(nn)) do
             l = loss(d...)
             end
-            update!(opt, params(nn), gs)
+            update!(opt, Flux.params(nn), gs)
         end
@@
 # Plot the posterior distribution with a contour plot.
-plt = plot_contour(X',ys,M);
-savefig(plt, joinpath(www_path, "nn_contour.png"))
+plt_nn = plot_contour(X',ys,M;zoom=0);
+savefig(plt_nn, joinpath(www_path, "nn_contour.png"))
 ```
@@
 ```{julia}
+λ = 0.1
 la = Laplace(nn, λ=λ)
@@
 # Plot the posterior distribution with a contour plot.
-plt = plot_contour(X',ys,Mᴸ);
+plt = plot_contour(X',ys,Mᴸ;zoom=0)
 savefig(plt, joinpath(www_path, "la_contour.png"))
@@
 
-## Experiment
+## Single Round
 
@@
 models = (Bayesian=Mᴸ, Plugin=M)
+# Generators:
+generators = (Bayesian=GreedyGenerator(), Plugin=GenericGenerator(ϵ=0.01))
+```
 
-# Generators:
-generators = (Bayesian=GreedyGenerator(), Plugin=GenericGenerator())
+### Generate counterfactual
 
+```{julia}
+cb = false
+γ = 0.75
+mod = Mᴸ
+gen = generators[:Plugin]
+plt_original = plt_nn = plot_contour(X',ys,mod;zoom=0,colorbar=cb,title="(a)")
+```
+
+```{julia}
+counterfactual_data = CounterfactualData(X,ys')
+μ = 0.10
+Random.seed!(123)
+candidates = findall(ys.==0)
+chosen = rand(candidates, Int(round(μ*length(candidates))))
+X′ = copy(X)
+y′ = copy(ys)
+using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
+for i in chosen
+    x = X[:,i]
+    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen,γ=γ)
+    X′[:,i] = counterfactual(outcome)
+    y′[i] = counterfactual_label(outcome)
+end
+plt_single = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(b)")
+```
+
+### Retrain
+
+```{julia}
+xs′ = Flux.unstack(X′,2)
+data = zip(xs′,y′)
+mod = retrain(mod, data)
+plt_single_retrained = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(c)")
+```
+
+### Repeat
+
+```{julia}
+i = 2
+while i <= 10
+    counterfactual_data = CounterfactualData(X′,y′')
+    Random.seed!(123)
+    candidates = findall(y′.==0)
+    chosen = rand(candidates, Int(round(μ*length(candidates))))
+    xs′ = Flux.unstack(X′,2)
+    data = zip(xs′,y′)
+    mod = retrain(mod, data)
+    for i in chosen
+        x = X′[:,i]
+        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen; γ=γ)
+        X′[:,i] = counterfactual(outcome)
+        y′[i] = counterfactual_label(outcome)
+    end
+    i += 1
+end
+plt_single_repeat = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(d)")
+```
+
+```{julia}
+plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
+savefig(plt, joinpath(www_path, "poc.png"))
+```
+
+## Experiments
+
+```{julia}
 # Variables:
-μ = [0.01,0.1,0.25]
+μ = [0.05,0.1,0.25]
 γ = [0.50,0.75,0.9]
@@
 # Experiments:
-experiment = Experiments.Experiment(X,ys,models[:Bayesian],target,grid_,n_rounds)
+experiment = Experiments.Experiment(X,ys,models[:Plugin],target,grid_,n_rounds)
 experiment_plugin = Experiments.Experiment(X,ys,models[:Plugin],target,grid_,n_rounds)
@@
 ```{julia}
+paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)]
+```
+
+```{julia}
+function plot_state(path, t; γ=0.75, μ = 0.05, k = 1, length_out=500, size=(150,150), title="")
+    chosen = map(p -> p.γ == γ && p.μ == μ && p.t==t && p.k==k, path)
+    path = path[chosen][1]
+    X = path.X̲'
+    y = vec(path.y̲')
+    plt = plot_contour(X,y,path.𝑴,length_out=length_out,colorbar=cb,title=title)
+    plt = plot(plt, axis=nothing, size=size, legend=false)
+end
+plot_state(paths[2], 1)
+```
+
+```{julia}
 #| eval: false
 using Main.Experiments: plot_path
-paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)];
+paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)]
 anim = Experiments.plot_path(paths[1],Γ=γ,𝙈=μ)
@@
     for d in data
-      gs = gradient(Flux.params(nn)) do
+      gs = gradient(Flux.Flux.params(nn)) do
         l = loss(d...)
       end
-      update!(opt, Flux.params(nn), gs)
+      update!(opt, Flux.Flux.params(nn), gs)
     end
@@
 # Experiments:
-experiment = Experiments.Experiment(X',y',models[:Bayesian],target,grid_,n_rounds)
+experiment = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
 experiment_plugin = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)

### 2022-04-29 dev/submissions/aies/extended_abstract/extended_abstract.qmd
first complete draft

@@
 Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations.
-Explainability is crucial in the social sciences context, because inference is typically at least as important as predictive performance.
-Decision-makers in the social sciences are also typically required to explain their decisions to human stakeholders: central bankers, for example, are held accountable by the public for the policies they decide on.
-It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that cannot be explained.
+Explainability is crucial in this context, since decision-makers are typically held accountable by the public: central banks, for example, are heavily scrutinized for the policies they impose.
+It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies they cannot trust.
 Deep neural networks, for example, are generally considered as black boxes and therefore not trustworthy in a context that demands explanations.
@@
 The remainder of this extended abstract is structured as follows: @sec-main presents one of the research questions I have investigated during the first months of my PhD: how do counterfactual explanations handle dynamics?
-@sec-related places this work in the broader context of Trustworthy AI for Finance and Economics.
+@sec-related places this work in the broader context of my research.
 
@@
 They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with.
-Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals facing adverse decisions.
-An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .
+Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes.
+An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
 
@@
 The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
-Only one recent work considers the implications of **exogenous** domain shifts on the validity of recourse [@upadhyay2021towards].
-The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal domain and model shift.
-They show that their approach yields more robust counterfactuals than existing approaches.
+Only one recent work considers the implications of **exogenous** domain and model shifts [@upadhyay2021towards].
+The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal model shift.
+They show that their approach yields more robust counterfactuals in this context than existing approaches.
 
 This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
-@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift.
-As this game of implementing AR and updating the classifier is repeated, the decision boundary moves in the opposite direction of the original training samples in the target class.
-We consider several aspects of these dynamics as problematic.
-Firstly, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as credit worthy.
+@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
+
+These dynamics may be problematic.
+As the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
 Average default risk across all borrowers can therefore be expected to increase.
-Conversely, lenders that anticipate or detect these sorts of dynamics may choose to still deny credit to individuals that have implemented AR, thereby comprimising the validity of AR.
+Conversely, lenders that anticipate such dynamics may choose to deny credit to individuals that have implemented AR, thereby compromising the validity of AR.
 
-To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse.
-In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.
-Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
+To the best of my knowledge this is the first work investigating endogenous dynamic in AR.
+Through future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.^[These benchmark datasets have their issues and controversies, which is one of the challenges I would like to discuss at AIES.] Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
 To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.
 
-![PLACEHOLDER: The dynamics of algorithmic recourse.](){#fig-dynamics fig.pos="h" height=100px}
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}
 
@@
 
-Until recently there existed only one open-source library that provides a unifying approach to generate and benchmark counterfactual explanations for models built and trained in Python [@pawelczyk2021carla].
-To address this limitation I have developed [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/): a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R.
-The package and companion paper are currently pending acceptance for a main talk at [JuliaCon](https://juliacon.org/2022/).
+Until recently there existed only one open-source library that provides a unifying approach to generate and benchmark counterfactual explanations for Python models [@pawelczyk2021carla].
+To address this limitation I have developed [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/): a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages.
+The package and companion paper are pending acceptance for a main talk at [JuliaCon '22](https://juliacon.org/2022/).
 
@@
 
-To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not.
-With respect to the work-in-progress presented here, I expect that this may help in mitigating endogenous domain and model shifts.
-To quantify the joint likelihood of input-output pairs, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions.
-While the former approach is more versatile, since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient.
-The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category.
-In future work, I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]
+To ensure that the generated counterfactuals are realistic it helps to understand which input-output pairs are likely to occur under the data generating process.
+To this end, previous work has either relied on generative models or restricted the analysis to probabilistic classifiers that incorporate uncertainty in their predictions.
+While the former approach is more generally applicable, the latter is computationally more efficient.
+In future work, I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).] With respect to the work-in-progress presented here, I expect that these efforts may help in mitigating endogenous domain and model shifts.
 

### 2022-04-29 dev/submissions/aies/extended_abstract/extended_abstract.tex
first complete draft

--- 2022-04-26 dev/submissions/aies/extended_abstract/extended_abstract.tex
@@
 \author{}
-\date{April 26, 2022}
+\date{April 29, 2022}
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 interpretable models that facilitate explanations. Explainability is
-crucial in the social sciences context, because inference is typically
-at least as important as predictive performance. Decision-makers in the
-social sciences are also typically required to explain their decisions
-to human stakeholders: central bankers, for example, are held
-accountable by the public for the policies they decide on. It is
-therefore not surprising that practitioners and academics in these
-fields are reluctant to adopt AI technologies that cannot be explained.
-Deep neural networks, for example, are generally considered as black
-boxes and therefore not trustworthy in a context that demands
-explanations. This PhD project is focused on exploring and developing
-methodologies that improve the trustworthiness of AI and thereby enable
-its application in Finance and Economics.
+crucial in this context, since decision-makers are typically held
+accountable by the public: central banks, for example, are heavily
+scrutinized for the policies they impose. It is therefore not surprising
+that practitioners and academics in these fields are reluctant to adopt
+AI technologies they cannot trust. Deep neural networks, for example,
+are generally considered as black boxes and therefore not trustworthy in
+a context that demands explanations. This PhD project is focused on
+exploring and developing methodologies that improve the trustworthiness
+of AI and thereby enable its application in Finance and Economics.
 
@@
 explanations handle dynamics? Section~\ref{sec-related} places this work
-in the broader context of Trustworthy AI for Finance and Economics.
+in the broader context of my research.
 
@@
 changes can be used for the purpose of \textbf{Algorithmic Recourse}
-(AR) to help individuals facing adverse decisions. An example relevant
+(AR) to help individuals who face adverse outcomes. An example relevant
 to the Finance and Economics domain is consumer credit: in this context
-AR can be used to guide individuals to improve their credit worthiness,
-should they have previously been denied access to credit based on a
-black-box decision-making system .
+AR can be used to guide individuals in improving their creditworthiness,
+should they have previously been denied access to credit based on an
+automated decision-making system.
 
@@
 and Hines 2020). Only one recent work considers the implications of
-\textbf{exogenous} domain shifts on the validity of recourse (Upadhyay,
-Joshi, and Lakkaraju 2021). The authors propose a simple minimax
-objective, that minimizes the counterfactual loss function for a maximal
-domain and model shift. They show that their approach yields more robust
-counterfactuals than existing approaches.
+\textbf{exogenous} domain and model shifts (Upadhyay, Joshi, and
+Lakkaraju 2021). The authors propose a simple minimax objective, that
+minimizes the counterfactual loss function for a maximal model shift.
+They show that their approach yields more robust counterfactuals in this
+context than existing approaches.
 
@@
 Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
-involving a probabilistic classifier and a greedy counterfactual
-generator proposed by Schut et al. (2021): AR leads to a domain shift,
-which in turn causes a drastic model shift. As this game of implementing
-AR and updating the classifier is repeated, the decision boundary moves
-in the opposite direction of the original training samples in the target
-class. We consider several aspects of these dynamics as problematic.
-Firstly, as the decision boundary moves in the direction of the
-non-target class, counterfactual paths become shorter: in the loan
-example, individuals that previously would have been denied credit based
-on their input features are suddenly considered as credit worthy.
-Average default risk across all borrowers can therefore be expected to
-increase. Conversely, lenders that anticipate or detect these sorts of
-dynamics may choose to still deny credit to individuals that have
-implemented AR, thereby comprimising the validity of AR.
+involving a probabilistic classifier and the counterfactual generator
+proposed by Wachter, Mittelstadt, and Russell (2017): the implementation
+of AR for a subset of individuals leads to a domain shift (b), which in
+turn triggers a model shift (c). As this game of implementing AR and
+updating the classifier is repeated, the decision boundary moves away
+from training samples that were originally in the target class (d).
+
+These dynamics may be problematic. As the decision boundary moves in the
+direction of the non-target class, counterfactual paths become shorter:
+in the loan example, individuals that previously would have been denied
+credit based on their input features are suddenly considered as
+creditworthy. Average default risk across all borrowers can therefore be
+expected to increase. Conversely, lenders that anticipate such dynamics
+may choose to deny credit to individuals that have implemented AR,
+thereby compromising the validity of AR.
 
 To the best of my knowledge this is the first work investigating
-endogenous dynamic in algorithmic recourse. In future experiments I want
-to investigate how this phenomenon plays out across different benchmark
-datasets including German credit, Boston Housing and COMPAS.
-Furthermore, I want to assess to what extent the magnitude and direction
-of domain and model shifts depends on the choice of the counterfactual
-generator. To this end, I am currently supervising a group of
-undergraduate students, who are tackling some of these tasks in their
-final-year research project.
+endogenous dynamic in AR. Through future experiments I want to
+investigate how this phenomenon plays out across different benchmark
+datasets including German credit, Boston Housing and COMPAS.\footnote{These
+  benchmark datasets have their issues and controversies, which is one
+  of the challenges I would like to discuss at AIES.} Furthermore, I
+want to assess to what extent the magnitude and direction of domain and
+model shifts depends on the choice of the counterfactual generator. To
+this end, I am currently supervising a group of undergraduate students,
+who are tackling some of these tasks in their final-year research
+project.
 
@@
 
-\caption{\label{fig-dynamics}PLACEHOLDER: The dynamics of algorithmic
-recourse.}
+\caption{\label{fig-dynamics}Dynamics in Algorithmic Recourse: we have a
+simple Bayesian model trained for binary classification (a); the
+implementation of AR for a random subset of individuals leads to a
+domain shift (b); as the classifier is retrained we observe a model
+shift (c); as this process is repeated, the decision boundary moves away
+from the target class (d).}
 
@@
 a unifying approach to generate and benchmark counterfactual
-explanations for models built and trained in Python (Pawelczyk et al.
-2021). To address this limitation I have developed
+explanations for Python models (Pawelczyk et al. 2021). To address this
+limitation I have developed
 \href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}:
@@
 for models developed and trained not only in Julia, but also in other
-popular programming languages like Python and R. The package and
-companion paper are currently pending acceptance for a main talk at
-\href{https://juliacon.org/2022/}{JuliaCon}.
+popular programming languages. The package and companion paper are
+pending acceptance for a main talk at
+\href{https://juliacon.org/2022/}{JuliaCon '22}.
 
@@
 
-To ensure that the generated explanations are realistic it is important
-to understand which input-output pairs are likely and which are not.
-With respect to the work-in-progress presented here, I expect that this
-may help in mitigating endogenous domain and model shifts. To quantify
-the joint likelihood of input-output pairs, previous work has either
-relied on generative models or restricted the analysis to probabilistic
-models that incorporate uncertainty in their predictions. While the
-former approach is more versatile, since it is applicable to both
-deterministic and probabilistic models, the latter is computationally
-much more efficient. The approach proposed by Schut et al. (2021) and
-used to generate the examples in Figure~\ref{fig-dynamics} falls into
-the latter category. In future work, I want to explore how recent
-advances in post-hoc uncertainty quantification, most notably Laplace
-Redux (Daxberger et al. 2021), can be leveraged to generate realistic
-and unambiguous counterfactual explanations for any model.\footnote{For
-  some initial work on this see my Julia implementation of Laplace
-  Redux:
+To ensure that the generated counterfactuals are realistic it helps to
+understand which input-output pairs are likely to occur under the data
+generating process. To this end, previous work has either relied on
+generative models or restricted the analysis to probabilistic
+classifiers that incorporate uncertainty in their predictions. While the
+former approach is more generally applicable, the latter is
+computationally more efficient. In future work, I want to explore how
+recent advances in post-hoc uncertainty quantification, most notably
+Laplace Redux (Daxberger et al. 2021), can be leveraged to generate
+realistic and unambiguous counterfactual explanations for any
+model.\footnote{For some initial work on this see my Julia
+  implementation of Laplace Redux:
   \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}
+With respect to the work-in-progress presented here, I expect that these
+efforts may help in mitigating endogenous domain and model shifts.
 

### 2022-04-29 dev/submissions/aies/sections/cover.qmd
aies
Fun with latex templates
first complete draft

--- 2022-04-13 dev/submissions/aies/extended_abstract/submission.qmd
@@
 ---
-title: "Overlooked Dynamics in Algorithmic Recourse"
 format:
-    html: default
-    pdf:
-        documentclass: acmart
-        # include-in-header: preamble.tex
-author:
-  - name: Patrick Altmeyer
+    pdf: default
 ---
 
-### Author
+## Personal details
 
-- Full Name: {{< var author.name >}}
-- Home page: {{< var author.url >}}
-- Affiliation: {{< var author.affiliation >}}
-- Affiliation URL: {{< var author.affiliation-url >}}
-- Country of citizenship: {{< var author.nationality >}}
-
-### PhD Progress
-
-- Expected graduation date: {{< var phd.graduation-date >}}
-- Years in program: {{< var phd.years >}}
-
-### Research Advisor
-
-- Full Name: {{< var advisor.name >}}
+|  | Author | Research Advisor |
+|---------|:--------|:--------|
+| Name     | {{< var author.name >}} | {{< var advisor.name >}} |  
+| Email | {{< var author.email >}} | {{< var advisor.email >}} | 
+| Phone | {{< var author.tel >}} | {{< var advisor.tel >}} | 
+| Home page | {{< var author.url >}} | {{< var advisor.url >}} | 
+| Affiliation | {{< var author.affiliation >}} | {{< var advisor.affiliation >}} | 
+| Affiliation URL | {{< var author.affiliation-url >}} | {{< var advisor.affiliation-url >}} | 
+| Citizenship | {{< var author.nationality >}} | {{< var advisor.nationality >}} | 
+| Expected graduation date | {{< var phd.graduation-date >}} | --- | 
+| Years in program | {{< var phd.years >}} | --- |
 
 *Key words*: {{< var keywords >}}
-
-### Letter of Support
-
-\pagebreak
-
-## Extended abstract

### 2022-04-29 dev/submissions/aies/sections/support.qmd
first complete draft

@@
 
-### Letter of Support
+## Letter of Support
 
-As substantiated in more detail in my full letter of support (attached with this application), as Patrick's advisor, I am very happy to support his application to the AIES student track.
+> As substantiated in more detail in my full letter of support (attached with this application), as Patrick's advisor, I am very happy to support his application to the AIES student track.
 Patrick is an eager student, with a pro-active enthusiasm to learn from others, and contribute himself to the learning progress of relevant communities.
@@
 In case of a positive outcome of this application, we are in a position to fully cover Patrick's costs, and thus will not lay any claim on the financial waivers.
+> --- {{< var advisor.name >}}

### 2022-04-30 dev/submissions/aies/extended_abstract/extended_abstract.qmd
AIES submission done

@@
 
-Recent advances in artificial intelligence (AI) have propelled its adoption in domains outside of computer science including health care, bioinformatics and genetics.
-In finance, economics and other social sciences, applications of AI are still relatively limited.
+Recent advances in Artificial Intelligence (AI) have propelled its adoption in domains outside of Computer Science including Healthcare, Bioinformatics and Genetics.
+In Finance, Economics and other social sciences, applications of AI are still relatively limited.
 Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations.
@@
 
-To the best of my knowledge this is the first work investigating endogenous dynamic in AR.
+To the best of my knowledge this is the first work investigating endogenous dynamics in AR.
 Through future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.^[These benchmark datasets have their issues and controversies, which is one of the challenges I would like to discuss at AIES.] Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.

### 2022-04-30 dev/submissions/aies/extended_abstract/extended_abstract.tex
AIES submission done

@@
 \author{}
-\date{April 29, 2022}
+\date{}
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-Recent advances in artificial intelligence (AI) have propelled its
-adoption in domains outside of computer science including health care,
-bioinformatics and genetics. In finance, economics and other social
+Recent advances in Artificial Intelligence (AI) have propelled its
+adoption in domains outside of Computer Science including Healthcare,
+Bioinformatics and Genetics. In Finance, Economics and other social
 sciences, applications of AI are still relatively limited.
@@
 To the best of my knowledge this is the first work investigating
-endogenous dynamic in AR. Through future experiments I want to
+endogenous dynamics in AR. Through future experiments I want to
 investigate how this phenomenon plays out across different benchmark

### 2022-04-30 dev/submissions/aies/sections/personal_statement.qmd
AIES submission done

@@
 ---
+title: Personal Statement
 format:
@@
 
-As a first-year PhD student in computer science with a background in finance and economics, I was very happy to see that AIES offers a student track.
-I have worked on several projects during the first months of my PhD and while the one I hope to present at AIES is the most mature research project at this point, it is still very much a work in progress.
-I think that the question I pose about the dynamics of algorithmic recourse is an interesting, albeit still fairly broad and open one.
+As a first-year PhD student in Computer Science with a background in Finance and Economics, I was very happy to see that AIES offers a student track.
+I have worked on several projects during the first months of my PhD and while the one I hope to present at AIES is the most mature research project at this point, it is still very much a work-in-progress.
+I think that the question I pose about the dynamics in Algorithmic Recourse is an interesting, albeit still fairly broad and open one.
 Through my participation in AIES, I would hope to get feedback that helps me refine the research question.
-As experiments related to the question are now being carried out by my students, the work can be expected to progress further until the actual AIES event and hence I could share more experimental results that may be of interest to the broader community.
+As experiments related to the question are now being carried out by a group of students under my supervision, the work can be expected to progress further until the actual AIES event.
+I therefore expect that I can share more experimental results that may be of interest to the broader community.
 AIES would also be an opportunity to discuss ideas for potential remedies to the issues this work highlights.
@@
 I also think that my broader PhD research agenda on the intersection of trustworthy AI and the social sciences should be of interest to the AIES community.
-Thanks to my academic background in economics, finance and data science as well my previous professional experience in monetary policy, I believe that I can contribute insights to a wide range of discussions revolving around the diverse set of topics relevant to AIES.
+Thanks to my academic background in Economics, Finance and Data Science as well my previous professional experience in Monetary Policy, I believe that I can contribute insights to a wide range of discussions revolving around the diverse set of topics relevant to AIES.
 It would be immensely helpful to learn from more experienced colleagues as well as fellow early-stage researchers, who are working on related research questions.
-At the main conference and the student event I expect to be exposed to new research ideas that I have never thought off, but may well be highly relevant to my own research.
+At the main conference and the student event I expect to be exposed to new research ideas that I have not thought of myself, but that may well be highly relevant to my own research.
 I would even hope to potentially find opportunities for future collaborations.
 
-This would be the first student program I have attended, although it is worth mentioning that in the past I have presented research at the European Central Bank (2019), IFABS (2021), NeurIPS (2021) and the Irving Fisher Committee on Central Bank Statistics (2022).
+This would be the first student program I have attended, although it is worth mentioning that in the past I have presented research at the European Central Bank (2019), Bank of England, IFABS (2021), NeurIPS (2021) and the Irving Fisher Committee on Central Bank Statistics (2022).
+More informations on this can be found in my resume.
 Since the main PhD research project that I would like to dicuss is still in its early stages, I have not submitted anything to the main track of AIES.

### 2022-07-05 README.qmd
integrating Aleks' code

---
format: gfm
---

# Algorithmic Recourse

## Abstract

{{< include paper/sections/abstract.qmd >}}

![](www/poc.png)

## Code

...

### 2022-07-05 paper/sections/abstract.qmd
integrating Aleks' code

Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting: given some classifier we are interested in finding close, actionable, realistic, sparse, diverse and ideally causally founded counterfactuals. The ability of CE to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. Only one recent work considers the implications of exogenous domain and model shifts. This project instead focuses on endogenous dynamics, that is shifts that occur when AR is actually implemented by a proportion of individuals. Early findings suggest that the involved shifts may be large with important implications on the validity of AR and the overall characteristics of the sample population.

### 2022-07-05 paper/sections/introduction.qmd
integrating Aleks' code

@@
----
-format: gfm
----
+Recent advances in Artificial Intelligence (AI) have propelled its adoption in domains outside of Computer Science including Healthcare, Bioinformatics and Genetics.
+In Finance, Economics and other social sciences, applications of AI are still relatively limited.
+Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations.
+Explainability is crucial in this context, since decision-makers are typically held accountable by the public: central banks, for example, are heavily scrutinized for the policies they impose.
+It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies they cannot trust.
+Deep neural networks, for example, are generally considered as black boxes and therefore not trustworthy in a context that demands explanations.
 
-## Introduction {#intro}
+**Counterfactual explanations** (CE) explain how inputs into a model need to change for it to produce different outputs.
+They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with.
+Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes.
+An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
 
-...
+Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
+The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
+Only one recent work considers the implications of **exogenous** domain and model shifts [@upadhyay2021towards].
+The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal model shift.
+They show that their approach yields more robust counterfactuals in this context than existing approaches.
+
+This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
+@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
+
+These dynamics may be problematic.
+As the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+Average default risk across all borrowers can therefore be expected to increase.
+Conversely, lenders that anticipate such dynamics may choose to deny credit to individuals that have implemented AR, thereby compromising the validity of AR.
+
+To the best of my knowledge this is the first work investigating endogenous dynamics in AR.
+Through future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.^[These benchmark datasets have their issues and controversies, which is one of the challenges I would like to discuss at AIES.] Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
+To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.
+
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](poc.png){#fig-dynamics fig.pos="h" width=45%}

### 2022-07-05 src/python/__init__.py

@@ -1,0 +1,14 @@
+# flake8: noqa
+
+from .datasets.generate_dataset import generate_continuous_dataset, generate_categorical_dataset
+from .dynamic_recourse.dynamic_benchmark import DynamicBenchmark
+from .dynamic_recourse.dynamic_csv_catalog import DynamicCsvCatalog
+from .dynamic_recourse.dynamic_online_catalog import DynamicOnlineCatalog
+from .dynamic_recourse.dynamic_mlmodel_catalog import DynamicMLModelCatalog
+from .dynamic_recourse.recourse_experiment import RecourseExperiment
+from .dynamic_recourse.recourse_generator import RecourseGenerator
+from .dynamic_recourse.recourse_generator import train_model
+from .plotting.plot_experiment import plot_experiment
+from .plotting.plot_dataset import calculate_boundary
+from .plotting.plot_tsne_boundary import plot_tsne_boundary
+from .plotting.visualize_recourse import generate_gif

### 2022-07-05 src/python/dynamic_recourse/recourse_generator.py

@@ -1,0 +1,261 @@
+import os
+import pandas as pd
+import sys
+
+
+from carla import log
+from .dynamic_mlmodel_catalog import DynamicMLModelCatalog
+from func_timeout import func_timeout, FunctionTimedOut
+from inspect import signature
+
+
+class RecourseGenerator():
+    """
+    Wrapper class for the CARLA generators that contains utilities required to conduct experiments.
+
+    Attributes:
+        name (str):
+            Name of the generator.
+        dataset (DataCatalog):
+            Catalog containing a dataframe, set of train and test records, and the target.
+        model (MLModelCatalog):
+            Classifier with additional utilities required by CARLA.
+        recourse_method (RecourseMethod):
+            A generator of algorithmic recourse which follows the API of CARLA.
+        generator_params (dict):
+            Dictionary of parameters that affect the actions of the generator.
+        model_params (dict):
+            Hyper-parameters for the underlying black-box model.
+        timeout (int):
+            Number of seconds after which the generation of a counterfactual should be considered a failure.
+    """
+    def __init__(self, name, dataset, model, recourse_method,
+                 generator_params, model_params, timeout=None):
+        self.name = name
+        self.dataset = dataset
+        self.model = model
+        self.recourse_method = recourse_method
+        self.generator_params = generator_params
+        self.model_params = model_params
+        self.timeout = timeout
+        self.positive_class_proba = 0
+
+        self.update_generator()
+
+    def update_generator(self):
+        """
+        Re-creates the generator on an updated model.
+        """
+        sig = signature(self.recourse_method.__init__)
+        params = [param for param in sig.parameters]
+
+        # RecourseMethods of CARLA do not follow a common interface
+        if 'data' in params:
+            self.generator = self.recourse_method(mlmodel=self.model,
+                                                  data=self.dataset,
+                                                  hyperparams=self.generator_params)
+        else:
+            self.generator = self.recourse_method(mlmodel=self.model,
+                                                  hyperparams=self.generator_params)
+
+    def apply(self, factuals):
+        """
+        Generate (a set of) counterfactuals with a method relevant for the generator.
+
+        Args:
+            factuals (pandas.DataFrame):
+                One or more records from a dataset used to train the black-box model.
+
+        Returns:
+            int: Number of successfully generated counterfactuals.
+        """
+        if self.timeout is not None:
[…]

### 2022-07-05 src/python/metrics/model.py

@@ -1,0 +1,126 @@
+import numpy as np
+
+
+from .distribution import MMD, MMD_null_hypothesis
+
+
+def disagreement_distance(dataset, target, initial_model, updated_model):
+    """
+    Calculates the Disagreement pseudo-distance defined in https://doi.org/10.1145/1273496.1273541
+    as Pr(h(x) != h'(x)), that is the probability that labels assigned by one classifier do not agree
+    with the labels assigned by another classifier. Simply put, it measures the overlap between models.
+    As this is an empirical measure, we can vary the number of records in `data`.
+
+    Args:
+        dataset (DataCatalog):
+            Catalog containing a dataframe, set of train and test records, and the target.
+        target (str):
+            The target column in the dataset.
+        initial_model (MLModelCatalog):
+            A model which was trained before recourse has been applied.
+        updated_model (MLModelCatalog):
+            A model retrained on a dataset with induced recourse.
+
+    Returns:
+        float: Probability that the two classifiers disagree on the label of a sample.
+    """
+    data = dataset._df
+    # Check how the initial model would assign labels to the dataset
+    initial_pred = np.argmax(initial_model.predict_proba(data.loc[:, data.columns != target]), axis=1)
+
+    # Check how the updated model would assign labels to the dataset
+    updated_pred = np.argmax(updated_model.predict_proba(data.loc[:, data.columns != target]), axis=1)
+
+    count_mismatch = 0
+    for index, prediction in enumerate(initial_pred):
+        if updated_pred[index] != prediction:
+            count_mismatch += 1
+
+    # Find the disagreement pseudo-distance
+    return count_mismatch / len(initial_pred)
+
+
+def class_decisiveness(data, target, model, cls):
+    """Calculates the pseudo-distance of points to the decision boundary
+    measured as the average probability of classification centered around 0.5.
+    High value corresponds to a large margin of classification.
+
+    Args:
+        data (pandas.DataFrame):
+            Records along with their labels.
+        target (str):
+            Name of the column that contains the target variable.
+        model (MLModelCatalog):
+            Classifier with additional utilities required by CARLA.
+        cls (str):
+            Encoding of the class to measure (positive or negative).
+
+    Returns:
+        float: Pseudo-distance from decision boundary
+    """
[…]

### 2022-07-05 src/python/plotting/visualize_recourse.py

@@ -1,0 +1,22 @@
+import imageio
+import os
+
+
+def generate_gif(experiment_path, generator_name):
+    """
+    Collect the images of data distribution over time and process them into a .gif file.
+
+    Args:
+        experiment_path (str): path to the experiment directory where images are stored
+        generator_name (str): name of the generator which should be processed
+    """
+    # Find corresponding filenames
+    filenames = [name for name in os.listdir(experiment_path) if os.path.isfile(os.path.join(experiment_path, name))]
+    filenames = list(filter(lambda name: name.startswith(generator_name), filenames))
+
+    # Generate the GIF and save to the original directory
+    images = []
+    for filename in sorted(filenames):
+        for _ in range(3):
+            images.append(imageio.imread(f'{experiment_path}/{filename}'))
+    imageio.mimsave(f'{experiment_path}/{generator_name}.gif', images)

### 2022-07-05 paper/sections/introduction.qmd
set up template for paper

@@
+# Introduction {#intro}
+
 Recent advances in Artificial Intelligence (AI) have propelled its adoption in domains outside of Computer Science including Healthcare, Bioinformatics and Genetics.
@@
 
-![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](poc.png){#fig-dynamics fig.pos="h" width=45%}
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}

### 2022-07-05 paper/tex/before_body.tex
set up template for paper

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

### 2022-07-05 README.qmd
some work on intro

@@
 
+### Python
+
+The bulk of the work for this project was done in Python. The relevant source code is under `/src/python`. Use the command below to install the necessary dependencies.
+
+```{zsh}
+pip install -r src/python/requirements.txt
+```
+
+### Julia
+
+Some exploratory work was initially done in Julia. The relevant source code is under `/src/julia`.
+
 ...

### 2022-07-05 paper/sections/introduction.qmd
some work on intro

--- 2022-04-13 paper/sections/introduction.qmd
@@
----
-format: gfm
----
+# Introduction {#intro}
 
-## Introduction {#intro}
+Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences.
+While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of principal-agent problem in the context of data-driven decision-making.
+It involves a group of **principals** - i.e.
+human stakeholders - that fail to understand the behaviour of their **agent** - i.e.
+the model used for automated decision-making [@borch2022machine].
 
-...
+Models or algorithms that fall into this category are typically referred to **black-box** models.
+Despite their shortcomings, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons].
+The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently iterpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models.
+These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.
+
+Among the approaches to XAI that have recently grown in popularity are **Counterfactual Explanations** (CE).
+They explain how inputs into a model need to change for it to produce different outputs.
+Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes.
+An example relevant to the Social Sciences is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
+
+Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
+The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
+Only one recent work considers the implications of **exogenous** domain and model shifts [@upadhyay2021towards].
+The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal model shift.
+They show that their approach yields more robust counterfactuals in this context than existing approaches.
+
+This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
+@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
+
+These dynamics may be problematic.
+As the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+Average default risk across all borrowers can therefore be expected to increase.
+Conversely, lenders that anticipate such dynamics may choose to deny credit to individuals that have implemented AR, thereby compromising the validity of AR.
+
+To the best of our knowledge this is the first work investigating endogenous dynamics in AR.
+Our contributions to the state of knowledge are two-fold.
+Firstly, we introduce an experimental framework built on top of CARLA, which allows for benchmarking different counterfactual generators in terms of the endogenous domain and model shifts they introduce.
+To this end we propose a number of novel evaluation metrix.
+Secondly, we use this framework to provide the first in-depth analysis of the dynamics of recourse induced by various diverse approaches to CE.
+In particular we focus on the following methodologies:  introduced in @wachter2017counterfactual, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
+
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}

### 2022-07-06 paper/sections/introduction.qmd
another attempt at py 3.8, but not working. Experiment code throwing and error that looks CARLA related.

@@
-# Introduction {#intro}
+# Introduction {#sec-intro}
 
@@
 
-Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
-The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
-Only one recent work considers the implications of **exogenous** domain and model shifts [@upadhyay2021towards].
-The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal model shift.
-They show that their approach yields more robust counterfactuals in this context than existing approaches.
+Existing work in this field has largely worked in a static setting: various approaches have been proposed to generate counterfactuals for a given individual that is subject to some pre-trained model.
+More recent work has compared different approaches within this static setting [@pawelczyk2021carla].
+In this work we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly?
+What types of dynamics are introduced and how do different counterfactual generators compare in this context?
 
-This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
 @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
+We refer to these types of dynamics as **endogenous** because they are induced by the implementation of recourse itself.
 
-These dynamics may be problematic.
-As the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}
+
+We find reasons to believe that these types of endogenous dynamics may be problematic.
+In @fig-dynamics, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
 Average default risk across all borrowers can therefore be expected to increase.
@@
 Our contributions to the state of knowledge are two-fold.
-Firstly, we introduce an experimental framework built on top of CARLA, which allows for benchmarking different counterfactual generators in terms of the endogenous domain and model shifts they introduce.
-To this end we propose a number of novel evaluation metrix.
-Secondly, we use this framework to provide the first in-depth analysis of the dynamics of recourse induced by various diverse approaches to CE.
-In particular we focus on the following methodologies:  introduced in @wachter2017counterfactual, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
+Firstly, we introduce an experimental framework extending previous work by @pawelczyk2021carla, which allows for benchmarking different counterfactual generators in terms of the endogenous domain and model shifts they introduce.
+To this end we propose a number of novel evaluation metrics.
+Secondly, we use this framework to provide the first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
 
-![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}
+The paper is structured as follows: @sec-related ... .

### 2022-07-06 paper/sections/related.qmd
another attempt at py 3.8, but not working. Experiment code throwing and error that looks CARLA related.

--- 2022-04-30 dev/submissions/aies/extended_abstract/extended_abstract.qmd
@@
----
-title: Dynamics in Algorithmic Recourse
-subtitle: Trustworthy Artificial Intelligence for Finance and Economics
-format:
-    pdf:
-        documentclass: acmart
-        classoption: sigconf
-        number-sections: true
-        include-in-header: preamble.tex
-        keep-tex: true
----
-
-## Introduction
-
-Recent advances in Artificial Intelligence (AI) have propelled its adoption in domains outside of Computer Science including Healthcare, Bioinformatics and Genetics.
-In Finance, Economics and other social sciences, applications of AI are still relatively limited.
-Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations.
-Explainability is crucial in this context, since decision-makers are typically held accountable by the public: central banks, for example, are heavily scrutinized for the policies they impose.
-It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies they cannot trust.
-Deep neural networks, for example, are generally considered as black boxes and therefore not trustworthy in a context that demands explanations.
-This PhD project is focused on exploring and developing methodologies that improve the trustworthiness of AI and thereby enable its application in Finance and Economics.
-
-The remainder of this extended abstract is structured as follows: @sec-main presents one of the research questions I have investigated during the first months of my PhD: how do counterfactual explanations handle dynamics?
-@sec-related places this work in the broader context of my research.
-
-## Dynamics in Algorithmic Recourse {#sec-main}
-
-**Counterfactual explanations** (CE) explain how inputs into a model need to change for it to produce different outputs.
-They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with.
-Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes.
-An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
+# Related Work {#sec-related}
 
@@
 They show that their approach yields more robust counterfactuals in this context than existing approaches.
-
-This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response.
-@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
-As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
-
-These dynamics may be problematic.
-As the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
-Average default risk across all borrowers can therefore be expected to increase.
-Conversely, lenders that anticipate such dynamics may choose to deny credit to individuals that have implemented AR, thereby compromising the validity of AR.
-
-To the best of my knowledge this is the first work investigating endogenous dynamics in AR.
-Through future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS.^[These benchmark datasets have their issues and controversies, which is one of the challenges I would like to discuss at AIES.] Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator.
-To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.
-
-![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}
-
-## Related and Future Work {#sec-related}
-
-### Benchmarking CE in Julia
-
-Until recently there existed only one open-source library that provides a unifying approach to generate and benchmark counterfactual explanations for Python models [@pawelczyk2021carla].
-To address this limitation I have developed [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/): a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages.
-The package and companion paper are pending acceptance for a main talk at [JuliaCon '22](https://juliacon.org/2022/).
-
-### Probabilistic Methods for Realistic CE
-
-To ensure that the generated counterfactuals are realistic it helps to understand which input-output pairs are likely to occur under the data generating process.
-To this end, previous work has either relied on generative models or restricted the analysis to probabilistic classifiers that incorporate uncertainty in their predictions.
-While the former approach is more generally applicable, the latter is computationally more efficient.
-In future work, I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).] With respect to the work-in-progress presented here, I expect that these efforts may help in mitigating endogenous domain and model shifts.
-
-## References

### 2022-07-06 paper/sections/introduction.qmd
work on paper

@@
 An example relevant to the Social Sciences is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
+In the remainder of this paper we will use both terminologies interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
 
@@
 
-@fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+@fig-poc illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 
-![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-dynamics fig.pos="h" width=45%}
+![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}
 
-We find reasons to believe that these types of endogenous dynamics may be problematic.
-In @fig-dynamics, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+We think that these types of endogenous dynamics may be problematic.
+Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+Secondly and relatedly, we observe in @fig-poc that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
 Average default risk across all borrowers can therefore be expected to increase.

### 2022-07-06 paper/sections/limitations.qmd
work on paper

## Limitations and Future Work {#sec-limit}

The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion. In practice, models are in fact updated on a regular basis [@upadhyay2021towards]. We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$. That being said, our experimental design is a vast over-simplification of potential real-world scenarios. In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021towards. We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed counterfactual explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class. Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary. In practice it is doubtful that agents behave according to such simple rules.

Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).

### 2022-07-06 paper/sections/methodology.qmd
work on paper

# Methodology {#sec-method}

In the following we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general to introduce the various counterfactual generators we have chosen to use in our experiments. We then describe the experimental setup in @sec-method-experiment and introduce several evaluation metrics used to benchmark the different generators.

## A Generalized Framework for Gradient-Based Counterfactual Search {#sec-method-general}

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
$$
\mathbf{s}\prime = \arg \min_{\mathbf{s}\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}\prime)) \right\}
$$ {#eq-general}

Here $\mathbf{s}\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards]. CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. To this end the authors use a complexity penalty $h(\mathbf{s}\prime)$ that favours diverse outcomes, in the sense that $s_1, ... , s_K$ look as different from each other as possible.

Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

## Experimental Setup {#sec-method-experiment}

The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice. We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class. Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse. In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in ...

[PSEUDO CODE]

A noteworthy practical consideration is the choice of $T$ and $B$. The higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time $T$. At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions. In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $25-50\%$ of the negative instances from the initial dataset. As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals. Using our framework extenson of @pawelczyk2021carla the experiment can be conducted on an arbitrary number of algorithmic recourse generators in parallel. As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.

## Data {#sec-method-data}

We have chosen to work with both synthetic and real-world datasets. Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics. Following @upadhyay2021towards we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results. Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.

### Synthetic data

We use 6 synthetic binary classification datasets consisting of 200-400 samples grouped in normally-distributed clusters.^[To see how the data is generated see here: [https://github.com/pat-alt/algorithmic_recourse_dynamics/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/algorithmic_recourse_dynamics/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

![A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}

Ex-ante we expect to see that Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary. Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (@joshi2019towards, @antoran2020getting) or rely on (and have access to) probabilistic models (@antoran2020getting, @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class.

### Real-world data

Additionally, we use two real-world datasets from the Finance domain. Firstly, we use the Give Me Some Credit dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes, the dataset was randomly undersampled to result in a balanced subsample made up of 3000 individuals. Secondly, we the German Credit dataset which involves the task of predicting if bank customers are credit-worthy or not [@germancredit1994]. It consists of 700 positive and 300 negative instances charaterized by 7 numerical and 13 categorical attributes. We process the dataset in two ways: (1) the values of the "Personal status and sex" feature are aggregated by the two represented genders; (2) the most common values are calculated for all categorical features such that a feature $x_d$ with the mode $\bar{x}_d$ is transformed into a new binary feature $\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}$. Binarization ensures that we can use all counterfactual generators in the benchmark.

## Evaluation Metrics {#sec-method-metrics}

We formulate two desiderata for the set of metrics used to measure domain and model shifts induced by recourse. First, the metrics should be applicable regardless of the dataset or classification technique so that they allow for the meaningful comparison of the generators in various scenarios. As the knowledge of the underlying probability distribution is rarely available, the metrics should be empirical and non-parametric. This further ensures that we can also measure large datasets by sampling from the available data. Moreover, while our study was conducted in a two-class classification setting, our choice of metrics should remain applicable in the future research on multi- class recourse problems. Second, the set of metrics should allow to capture various aspects of the previously mentioned magnitude, path, and tempo of changes while remaining as small as possible.

### Domain Shifts

To quantify the magnitude of domain shifts we rely on an unbiased estimate of the squared population **Maximum Mean Discrepancy (MMD)** given as:

$$
MMD^2_u
$$ {#eq-mmd}

where $\mathcal{F}$ is a unit ball in a Reproducing Kernel Hilbert Space H [27], and $X$, $\tilde{X}$ represent independent and identically distributed samples drawn from probability distributions $p$ and $q$ respectively [28]. MMD is a measure of the distance between the kernel mean embeddings of $p$ and $q$ in RKHS $\mathcal{H}$. An important consideration is the choice of the kernel function $k(\cdot,\cdot)$. In our implementation we make use of the radial basis function (RBF) kernel with a constant length-scale parameter of $0.5$. As RBF captures all moments of distributions $p$ and $q$, we have that $MMD_u^2[F,X,\tilde{X}]=0$ if and only if $X=\tilde{X}$.

The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X\prime$ and $\tilde{X}\prime$ having the same size as the original samples. Then under the null hypothesis we should have that $MMD_u^2[F,X\prime,\tilde{X}\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

We calculate the MMD for both classes individually based on the ground truth labels, i.e. the labels that samples were assigned in time $t=0$. Throughout our experiments we generally do not expect the distribution of the negative class to change over time – application of recourse reduces the size of this class, but since individuals are sampled uniformly the distribution should remain unaffected. Conversely, unless a recourse generator can perfectly replicate the original probability distribution, we expect the MMD of the positive class to increase. Thus, when discussing MMD, we generally mean the shift in the distribution of the positive class.

The **Operating Point** or **elbow** of the $k$-means algorithm is used to measure whether the counterfactual instances tend to lie close to the positively labelled training samples or make up a new, separate cluster of data points within the positive class. In the latter case the operating point of the $k$-means algorithm can be expected to change. There exist several algorithms for the automated discovery of elbow points and we resort to the Kneedle method [@satopaa2011finding]. Kneedle selects the optimal value of $k$ by rotating the inertia curve onto the horizontal axis and selecting a local minimum. While working with the synthetic datasets we also use domain knowledge to confirm the results of Kneedle through the visual inspection of data.

Finally, **feature mean and feature standard deviation** are also calculated to verify how the implementation of recourse impacts every attribute in the dataset. Although MMD already captures information about the expected value and variance, we may also be interested in a more granular look at individual features.

### Model Shifts

We extend the metric in @eq-mmd for the purpose of quantifying model shifts. Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying @eq-mmd to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$. If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. It is worth noting that while we apply the technique to samples drawn uniformly from the dataset, it can also be employed on arbitrary points in the entire feature space (or a subspace). The latter approach is theoretically more robust. Unfortunately, in practice this approachs suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.

As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.

Finally, we introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample. We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function. This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes. Although the exact values for this metric are not important for our study, they can be used to detect model shifts. If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.

### 2022-07-06 paper/sections/related.qmd
work on paper

@@
 
-Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$.
+In this Section we provide a review of the relevant literature.
+First, @sec-related-recourse discusses the existing research within the domain of counterfactual explanations and algorithmic recourse.
+Then, @sec-related-shifts presents some of the previous work on the measurement of dataset and model shifts.
+
+## Algorithmic Recourse {#sec-related-recourse}
+
+A framework for Counterfactual Explanations was first proposed in 2017 by @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then.
+Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.
+Then we are interested in minimizing the complexity or effort $H=h(x\prime)$ associated with moving an individual $x$ to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $t$:
+
+$$
+\min_{x\prime \in \mathcal{x}} c(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
+$$ {#eq-obj}
+
+For implementation purposes, @eq-obj is typically approximated through regularization:
+
+$$
+x\prime = \arg \min_{x\prime}  \ell(M(x\prime),t) + \lambda h(x\prime)
+$$ {#eq-solution}
+
+In the baseline work and many subsequent approaches the complexity function $h: \mathcal{X} \mapsto \mathbb{R}$ is proxied by some distance metric based on the simple intuition that large perturbations of $x$ are costly.
+
+Many approaches for the generation of algorithmic recourse have been described in the literature.
+An October 2020 survey by Karimi et al.
+layed out 60 algorithms that have been proposed since 2014 [@karimi2020survey].
+Another survey published around the same time by Verma et al.
+described 29 algorithms [@verma2020counterfactual].
+Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met.
+Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs.
+To name but a few, we are interested in generating counterfactuals that close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and if possible causally founded [@karimi2021algorithmic].
+
+Efforts so far have largely been directed at improving the quality of counterfactual explanations within a static context: given some pre-trained classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in generating one or multiple meaningful counterfactual explanations for some individual characterized by $x_i$.
 The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
-Only one recent work considers the implications of **exogenous** domain and model shifts [@upadhyay2021towards].
-The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal model shift.
-They show that their approach yields more robust counterfactuals in this context than existing approaches.
+We have been able to identify only one recent work that considers the implications of **exogenous** domain and model shifts in the context of AR [@upadhyay2021towards].
+Exogenous shifts are strictly of external origin.
+For example, they might stem from data correction, temporal shifts or geospatial changes [@upadhyay2021towards].
+The authors of [@upadhyay2021towards] propose framework for algorithmic recourse (ROAR) that evidently improves robustness to such exogenous shifts.
+
+## Domain and Model Shifts {#sec-related-shifts}
+
+Much attention has been paid to the detection of dataset shifts – situations where the distribution of data changes over time.
+Rabanser et al.
+suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests [@rabanser2019failing].
+This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution.
+Numerous approaches to anomaly detection have been summarized [@chandola2009anomaly].
+Another well-established research topic is that of concept drift – situations where external variables influence the patterns between the input and the output of a model [@widmer1996learning].
+For instance, Gama et al.
+offer a review of the adaptive learning techniques which can handle concept drift [@gama2014survey].
+Less previous work is available on the related topic of model drift - changes in model performance over time.
+Nelson et al.
+review how resistant different machine learning models are to the model drift [@nelson2015evaluating].
+Ackerman et al.
+offer a method to detect changes in model performance when ground truth is not available [@ackerman2021machine].
+
+In the context of algorithmic recourse, domain and model shifts were first brought up by the authors behind ROAR [@upadhyay2021towards].
+In their work they refer to model shifts as simply any perturbation $\Delta$ to the parameters of the model in question: $M$.
+While this also sets the baseline for our analysis here, it is worth noting that in @upadhyay2021towards these perturbations are mechanically introduced.
+In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
+In addition to quantifying the magnitude of shifts $\Delta$, we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model.
+We have not been able to identify previous work on this topic.
+
+## Benchmarking Counterfactual Generators {#sec-related-benchmark}
+
+...

### 2022-07-06 paper/sections/introduction.qmd
added pseudo-code placeholder

@@
 
-The paper is structured as follows: @sec-related ... .
+We find that ...
+
+The paper is structured as follows: @sec-related places our work in the broader context of related literature.
+@sec-method presents our methodology and data.
+@sec-empirical presents our empirical findings which are then discussed in @sec-discussion.
+We also point to some of the limitations or our work as well as avenues for future research in @sec-limit.
+Finally, @sec-conclusion concludes.

### 2022-07-06 paper/sections/methodology.qmd
added pseudo-code placeholder

@@
 $$
-\mathbf{s}\prime = \arg \min_{\mathbf{s}\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}\prime)) \right\}
+\mathbf{s}^\prime = \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}^\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}^\prime)) \right\}
 $$ {#eq-general}
 
-Here $\mathbf{s}\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
+Here $\mathbf{s}^\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
 In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one.
@@
 Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals.
-To this end the authors use a complexity penalty $h(\mathbf{s}\prime)$ that favours diverse outcomes, in the sense that $s_1, ...
+To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ...
 , s_K$ look as different from each other as possible.
@@
 To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population.
-The experimental setup is summarized in ...
+The experimental setup is summarized in Algorithm \ref{euclid}
 
-[PSEUDO CODE]
+\begin{algorithm}
+\caption{PLACEHOLDER}\label{euclid}
+\begin{algorithmic}[1]
+\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
+\State $r\gets a\bmod b$
+\While{$r\not=0$}\Comment{We have the answer if r is 0}
+\State $a\gets b$
+\State $b\gets r$
+\State $r\gets a\bmod b$
+\EndWhile\label{euclidendwhile}
+\State \textbf{return} $b$\Comment{The gcd is b}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
 
@@
 $$
-MMD^2_u
+\begin{aligned}
+MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
 $$ {#eq-mmd}
@@
 To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$.
-Then, we split the permuted data into two new samples $X\prime$ and $\tilde{X}\prime$ having the same size as the original samples.
-Then under the null hypothesis we should have that $MMD_u^2[F,X\prime,\tilde{X}\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$.
+Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples.
+Then under the null hypothesis we should have that $MMD_u^2[F,X^\prime,\tilde{X}^\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$.
 The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.
@@
 As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement).
-This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.

### 2022-07-07 paper/sections/methodology.qmd
minor changes to paper draft

@@
 To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population.
-The experimental setup is summarized in Algorithm \ref{euclid}
+The experimental setup is summarized in Algorithm \ref{algo-experiment}
 
 \begin{algorithm}
-\caption{PLACEHOLDER}\label{euclid}
+\caption{Experiment}\label{algo-experiment}
 \begin{algorithmic}[1]
-\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-\State $r\gets a\bmod b$
-\While{$r\not=0$}\Comment{We have the answer if r is 0}
-\State $a\gets b$
-\State $b\gets r$
-\State $r\gets a\bmod b$
-\EndWhile\label{euclidendwhile}
-\State \textbf{return} $b$\Comment{The gcd is b}
+\Procedure{Experiment}{$M,D,G$}
+\State $t\gets 0$
+\While{$t<T$}
+\State $D_B \subset D$
+\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
+\State $M\gets M(D)$ \Comment{Retrain model.}
+\EndWhile
+\State \textbf{return} $M,D$
 \EndProcedure
@@
 
-![A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}
+![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}
 
@@
 
+As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021towards.
+We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$.
+We shall refer to this baseline metric simply as **Perturbations**.
+
 We extend the metric in @eq-mmd for the purpose of quantifying model shifts.

### 2022-07-08 paper/sections/methodology.qmd
moving to julia

@@
 We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
+
+[AGREE ON NOTATION]
+
 $$
-\mathbf{s}^\prime = \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}^\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}^\prime)) \right\}
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left(  {\left\{\ell(M(f(s_k^\prime)),t)\right\}_K}^\mathsf{T} \mathbf{1}_K + \lambda {\left\{h(f(s_k^\prime)) \right\}_K}^\mathsf{T} \mathbf{1}_K \right) \\
+&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \left(\ell . ((M\circ f).(\mathbf{s}^\prime),t) + \lambda (h \circ f).(\mathbf{s}^\prime)\right)^\mathsf{T} \mathbf{1}_K \right\} \\
+&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
+\end{aligned}
 $$ {#eq-general}
 
-Here $\mathbf{s}^\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
+Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
 In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one.
@@
 As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
-Using our framework extenson of @pawelczyk2021carla the experiment can be conducted on an arbitrary number of algorithmic recourse generators in parallel.
+Using our framework the experiment can be conducted on an arbitrary number of algorithmic recourse generators.
 As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.

### 2022-07-08 paper/sections/related.qmd
moving to julia

@@
 
-...
+Despite the large and growing number of approaches to counterfactual search there have been surprisingly few benchmark studies that compare different methodologies.
+This may be partially due to limited software availability in this space.
+Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework that can be used to apply and benchmark different methodology; finally, @altmeyer2022CounterfactualExplanations provides an extensible and language-agnostic implementation in Julia.
+For the experiments conducted in this work we relied on the latter.

### 2022-07-12 dev/python/notebooks/experiments.qmd
setting up Julia framework

---
jupyter: python3
---

```{python}
import warnings
warnings.filterwarnings("ignore", message=r"Passing", category=FutureWarning)
warnings.filterwarnings("ignore", message=r"Implicit", category=UserWarning)
warnings.filterwarnings("ignore", message=r"cannot", category=UserWarning)
```

```{python}
import os
import sys

from carla.recourse_methods import Dice, Wachter
from copy import deepcopy
from src.python import (
    DynamicCsvCatalog,
    DynamicOnlineCatalog,
    RecourseExperiment,
    RecourseGenerator,
    train_model,
    plot_experiment,
    generate_gif
)

os.environ["CUDA_VISIBLE_DEVICES"] = ""
data_path = "data"
```

We load the dataset as a `DynamicCsvCatalog` and provide it with the lists of categorical and continuous features, a list of immutable features that should not be taken into consideration by the recourse generator, the target, encoding of positive and negative classes, size of the test set which is used to assess the error of the classifier, and a name.

```{python}
dataset = DynamicCsvCatalog(file_path=os.path.join(data_path,"give_me_some_credit_balanced.csv"), 
                            categorical=[],
                            continuous=['RevolvingUtilizationOfUnsecuredLines',
                                        'age', 'NumberOfTime30-59DaysPastDueNotWorse',
                                        'DebtRatio', 'MonthlyIncome',
                                        'NumberOfOpenCreditLinesAndLoans',
                                        'NumberOfTimes90DaysLate',
                                        'NumberRealEstateLoansOrLines',
                                        'NumberOfTime60-89DaysPastDueNotWorse',
                                        'NumberOfDependents'],
                            immutables=['age'], target='SeriousDlqin2yrs',
                            positive=1, negative=0, test_size=0.3, name="GMSC")
```

Our frameworks supports simultaneous assessment of multiple generators. To that end, we pass a list of `RecourseGenerators` which should be put to test in the same experimental conditions.

Data about the dynamics of recourse is collected for all of the declared generators. It can be saved to a `.csv` file.

```{python}
hyper_params = {'learning_rate': 0.01, 'epochs': 10, 'batch_size': 1, 'hidden_size': [5]}
name='GMSC_experiment'

# SET NAME BEFORE THE EXPERIMENT
for index in range(5):

  # Recourse generated by DICE is compared with the Wachter generator, as they may modify data differently
  # we need to keep track of two models and two datasets and update them independently
  model = train_model(deepcopy(dataset), hyper_params, model_type='ann')

  generators = [
              RecourseGenerator('Wachter', deepcopy(dataset), deepcopy(model),
                                Wachter, {"loss_type": "BCE", "t_max_min": 2}, hyper_params, 120),
              RecourseGenerator('DICE_3', deepcopy(dataset), deepcopy(model),
                                Dice, {"num": 3}, hyper_params, 120),
              ]

  experiment = RecourseExperiment(dataset, model, generators, name, hyper_params)
  experiment.run(epochs=15, recourse_per_epoch=10, calculate_p=1500)
  experiment.save_data()
```

We also provide multiple functions that simplify the visualization of collected data.

```{python}
experiment_path = f'../experiment_data/{experiment.experiment_name}'
for g in generators:
    generate_gif(experiment_path, g.name)
```

```{python}
config = [
    {'type': 'pos_MMD', 'dict_path': ['MMD', 'positive', 'value']},
    {'type': 'neg_MMD', 'dict_path': ['MMD', 'negative', 'value']},
    {'type': 'pos_distance', 'dict_path': ['decisiveness', 'positive']},
    {'type': 'neg_distance', 'dict_path': ['decisiveness', 'negative']},
    {'type': 'disagreement', 'dict_path': ['disagreement']},
    {'type': 'model_MMD', 'dict_path': ['model_MMD', 'value']}
]

for c in config:
    plot_experiment(experiment_path,
                    [g.name for g in generators],
                    c['type'],
                    c['dict_path'],
                    show_plot=False)
```

### 2022-07-12 dev/resources/misc/boston_housing.qmd
setting up Julia framework

--- 2022-04-25 dev/work/boston_housing.qmd
@@
 include("../src/load.jl")
-using AlgorithmicRecourse, MLDatasets, Flux
+using CounterfactualExplanations, MLDatasets, Flux
 using Plots, PlotThemes
@@
 
-Prepare ensemble for use with AlgorithmicRecourse.jl:
+Prepare ensemble for use with CounterfactualExplanations.jl:
 

### 2022-07-12 notebooks/synthetic.qmd
setting up Julia framework

--- 2022-04-29 dev/work/synthetic_endogenous_shifts.qmd
@@
 ---
-title: Synthetic data
-jupyter: julia-1.6
+title: Experiments with Synthetic Data
+jupyter: julia-1.7
 ---
@@
 ```{julia}
-include("src/load.jl")
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, BayesLaplace, LinearAlgebra
-using Base.Filesystem: joinpath
+include("src/AlgorithmicRecourseDynamics.jl")
+using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
 theme(:wong)
-using Logging
-disable_logging(Logging.Info)
-output_path = "dev/work/output/synthetic"
-www_path = "dev/work/www/synthetic"
+output_path = output_dir("synthetic")
+www_path = www_dir("synthetic")
 ```

### 2022-07-12 paper/sections/methodology.qmd
bleh
setting up Julia framework

@@
 
-[AGREE ON NOTATION]
-
 $$
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left(  {\left\{\ell(M(f(s_k^\prime)),t)\right\}_K}^\mathsf{T} \mathbf{1}_K + \lambda {\left\{h(f(s_k^\prime)) \right\}_K}^\mathsf{T} \mathbf{1}_K \right) \\
-&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \left(\ell . ((M\circ f).(\mathbf{s}^\prime),t) + \lambda (h \circ f).(\mathbf{s}^\prime)\right)^\mathsf{T} \mathbf{1}_K \right\} \\
-&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
 \end{aligned}
@@
 
-We use 6 synthetic binary classification datasets consisting of 200-400 samples grouped in normally-distributed clusters.^[To see how the data is generated see here: [https://github.com/pat-alt/algorithmic_recourse_dynamics/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/algorithmic_recourse_dynamics/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
+We use 6 synthetic binary classification datasets consisting of 200-400 samples grouped in normally-distributed clusters.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
 Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

### 2022-07-12 src/AlgorithmicRecourseDynamics.jl

@@ -1,7 +1,15 @@
-# Load modules:
+module AlgorithmicRecourseDynamics
 
+# Load modules:
 include("utils.jl")
+
 include("models/Models.jl")
 using .Models
+
+include("metrics/Metrics.jl")
+using .Metrics
+
 include("experiments/Experiments.jl")
 using .Experiments
+
+end

### 2022-07-12 src/metrics/Metrics.jl

@@ -1,0 +1,4 @@
+module Metrics
+
+
+end

### 2022-07-12 test/runtests.jl

@@ -1,0 +1,6 @@
+using AlgorithmicRecourseDynamics
+using Test
+
+@testset "AlgorithmicRecourseDynamics.jl" begin
+    # Write your tests here.
+end

### 2022-07-12 dev/paper/sections/introduction.qmd
prepared artifacts

--- 2022-07-06 paper/sections/introduction.qmd
@@
 An example relevant to the Social Sciences is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
-In the remainder of this paper we will use both terminologies interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
+A meaningful recourse recommendation for a denied applicant could be: *"If your net savings rate had been 10% of your monthly income instead of the actual 8%, your application would have been successful.
+See if you can temporarily cut down on consumption."* In the remainder of this paper we will use both terminologies - recourse and counterfactual - interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
 

### 2022-07-12 dev/artifacts/generate_artifacts.jl

@@ -1,0 +1,76 @@
+using Pkg; Pkg.activate("dev")
+using Pkg.Artifacts, LibGit2, ghr_jll
+
+function generate_artifacts(
+    datafiles=readdir("dev/artifacts/upload/"); 
+    data_dir="dev/artifacts/upload", 
+    root=".",
+    artifact_toml=joinpath(root, "Artifacts.toml"), 
+    deploy=true,
+    tag="artifacts"
+)
+
+    if deploy && !haskey(ENV, "GITHUB_TOKEN")
+        @warn "For automatic github deployment, need GITHUB_TOKEN. Not found in ENV, attemptimg global git config."
+    end
+
+    if deploy
+        # Where we will put our tarballs
+        tempdir = mktempdir()
+    
+        function get_git_remote_url(repo_path::String)
+            repo = LibGit2.GitRepo(repo_path)
+            origin = LibGit2.get(LibGit2.GitRemote, repo, "origin")
+            return LibGit2.url(origin)
+        end
+    
+        # Try to detect where we should upload these weights to (or just override
+        # as shown in the commented-out line)
+        origin_url = get_git_remote_url(root)
+        deploy_repo = "$(basename(dirname(origin_url)))/$(splitext(basename(origin_url))[1])"
+    
+    end
+
+    # For each BSON file, generate its own artifact:
+    for datafile in datafiles
+
+        # Name for hash/artifact:
+        name = splitext(datafile)[1]
+
+        # create_artifact() returns the content-hash of the artifact directory once we're finished creating it
+        hash = create_artifact() do artifact_dir
+            cp(joinpath(data_dir, datafile), joinpath(artifact_dir, datafile))
+        end
+
+        # Spit tarballs to be hosted out to local temporary directory:
+        if deploy
+            
+            tarball_hash = archive_artifact(hash, joinpath(tempdir, "$(name).tar.gz"))
+
+            # Calculate tarball url
+            tarball_url = "https://github.com/$(deploy_repo)/releases/download/$(tag)/$(name).tar.gz"
+
+            # Bind this to an Artifacts.toml file
+            @info("Binding $(name) in Artifacts.toml...")
+            bind_artifact!(
+                artifact_toml, name, hash; 
+                download_info=[(tarball_url, tarball_hash)], lazy=true, force=true
+            )
+        end
+
+    end
+
+    if deploy
+        # Upload tarballs to a special github release
[…]

### 2022-07-12 src/utils.jl

@@ -4,7 +4,7 @@
 Sets up the directory to save computational outputs and returns the path.
 """
 function output_dir(dir="")
-    root_ = "output"
+    root_ = "dev/artifacts/upload/output"
     if !isdir(root_)
         mkdir(root_)
     end
@@ -18,7 +18,7 @@ end
 Sets up the directory to save images and returns the path.
 """
 function www_dir(dir="")
-    root_ = "www"
+    root_ = "dev/artifacts/upload/www"
     if !isdir(root_)
         mkdir(root_)
     end

### 2022-07-12 dev/notebooks/synthetic.qmd
bleh

--- 2022-07-12 notebooks/synthetic.qmd
@@
 ```{julia}
-include("src/AlgorithmicRecourseDynamics.jl")
+include("dev/utils.jl")
 using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra

### 2022-07-12 src/data/functions.jl

@@ -1,3 +1,5 @@
+using Pkg.Artifacts
+
 function download_artifacts()
     data_dir = artifact"data"
 end

### 2022-07-13 dev/notebooks/synthetic.qmd
work on streamlining code for experiments

@@
 
+```{julia}
+using Pkg; Pkg.activate("dev")
+```
+
 In this notebook we will use toy data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.
@@
 include("dev/utils.jl")
+using AlgorithmicRecourseDynamics
 using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
@@
 ```{julia}
-Random.seed!(123)
+Random.seed!(1234)
 using MLJ
@@
 data = zip(xs,ys)
-plt = plot()
-plot_data!(plt,X',ys)
+counterfactual_data = CounterfactualData(X,ys')
 ```
@@
 ```{julia}
-nn = Chain(Dense(2,1))
-loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y) 
+using Flux
+model = Chain(Dense(2,1))
+M = FluxModel(model)
 ```
@@
 ```{julia}
-using Flux.Optimise: update!, ADAM
-opt = ADAM()
-epochs = 10
-function _forward_nn(nn, data, epochs)
-    for epoch = 1:epochs
-        for d in data
-            gs = gradient(Flux.params(nn)) do
-            l = loss(d...)
-            end
-            update!(opt, Flux.params(nn), gs)
-        end
-    end
-    return nn
-end
-_forward_nn(nn, data, epochs)
+n_epochs = 10
+using AlgorithmicRecourseDynamics.Models
+M = Models.train(M,counterfactual_data; n_epochs=n_epochs)
 ```
@@
 ```{julia}
-using CounterfactualExplanations.Models
-import CounterfactualExplanations.Models: logits, probs # import functions in order to extend
-import Main.Models: retrain
-
-# Step 1)
-struct NeuralNetwork <: Models.AbstractFittedModel
-    model::Any
-end
-
-# Step 2)
-logits(M::NeuralNetwork, X::AbstractArray) = M.model(X)
-probs(M::NeuralNetwork, X::AbstractArray)= σ.(logits(M, X))
-
-# Step 3)
-function retrain(M::NeuralNetwork, data; n_epochs=10, τ=1.0) 
-    nn = M.model
-    nn = _forward_nn(nn, data, n_epochs)
-    M = NeuralNetwork(nn)
-    return M
-end
-
-M = NeuralNetwork(nn)
-```
-
-```{julia}
-#| echo: false
-# Plot the posterior distribution with a contour plot.
-plt_nn = plot_contour(X',ys,M;zoom=0);
+plt_nn = plot(M,counterfactual_data;zoom=0)
 savefig(plt_nn, joinpath(www_path, "nn_contour.png"))
@@
 
-![](www/nn_contour.png)
+![](../artifacts/upload/www/synthetic/nn_contour.png)
 
@@
 λ = 0.1
-la = Laplace(nn, λ=λ)
-BayesLaplace.fit!(la, data)
+model_laplace = Laplace(M.model, λ=λ)
+LaplaceRedux.fit!(model_laplace, data)
 ```
@@
 ```{julia}
-# Step 1)
-struct LaplaceNeuralNetwork <: Models.AbstractFittedModel
-    la::BayesLaplace.Laplace
-end
-
-# Step 2)
-logits(M::LaplaceNeuralNetwork, X::AbstractArray) = M.la.model(X)
-probs(M::LaplaceNeuralNetwork, X::AbstractArray)= BayesLaplace.predict(M.la, X)
-
-# Step 3)
-function retrain(M::LaplaceNeuralNetwork, data; n_epochs=10, τ=1.0) 
-    nn = M.la.model
-    nn = _forward_nn(nn, data, n_epochs)
-    la = Laplace(nn, λ=λ)
-    BayesLaplace.fit!(la, data)
-    M = LaplaceNeuralNetwork(la)
-    return M
-end
-
-Mᴸ = LaplaceNeuralNetwork(la)
+M_laplace = Models.LaplaceModel(model_laplace)
 ```
@@
 ```{julia}
-#| echo: false
-# Plot the posterior distribution with a contour plot.
-plt = plot_contour(X',ys,Mᴸ;zoom=0)
-savefig(plt, joinpath(www_path, "la_contour.png"))
+plt_laplace = plot(M_laplace, counterfactual_data;zoom=0)
+savefig(plt_laplace, joinpath(www_path, "la_contour.png"))
 ```
 
-![](www/la_contour.png)
+![](../artifacts/upload/www/synthetic/la_contour.png)
 
@@
 # Models:
-models = (Bayesian=Mᴸ, Plugin=M)
+models = (Bayesian=M_laplace, Plugin=M)
 # Generators:
-generators = (Bayesian=GreedyGenerator(), Plugin=GenericGenerator(ϵ=0.01))
+generators = (
+    Bayesian=GreedyGenerator(loss=:logitbinarycrossentropy), 
+    Plugin=GenericGenerator(loss=:logitbinarycrossentropy,ϵ=0.01)
+)
 ```
@@
 γ = 0.75
-mod = Mᴸ
+mod = M_laplace
 gen = generators[:Plugin]
-plt_original = plt_nn = plot_contour(X',ys,mod;zoom=0,colorbar=cb,title="(a)")
+plt_original = plot(mod,counterfactual_data;zoom=0,colorbar=false,title="(a)")
 ```
@@
 ```{julia}
-counterfactual_data = CounterfactualData(X,ys')
 μ = 0.10
@@
 end
-plt_single = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(b)")
+counterfactual_data′ = CounterfactualData(X′,y′')
+plt_single = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(b)")
 ```
@@
 ```{julia}
-xs′ = Flux.unstack(X′,2)
-data = zip(xs′,y′)
-mod = retrain(mod, data)
-plt_single_retrained = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(c)")
+mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
+plt_single_retrained = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(c)")
 ```
@@
 while i <= 10
-    counterfactual_data = CounterfactualData(X′,y′')
+    counterfactual_data′ = CounterfactualData(X′,y′')
     Random.seed!(123)
@@
     chosen = rand(candidates, Int(round(μ*length(candidates))))
-    xs′ = Flux.unstack(X′,2)
-    data = zip(xs′,y′)
-    mod = retrain(mod, data)
+    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
     for i in chosen
@@
 end
-plt_single_repeat = plot_contour(X′',y′,mod;zoom=0,colorbar=cb,title="(d)")
+plt_single_repeat = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(d)")
 ```

### 2022-07-13 src/models/bayes_logreg.jl

@@ -1,130 +1,0 @@
-# Packages
-using LinearAlgebra
-include("optimization.jl")
-
-∑(vector)=sum(vector)
-
-# Sigmoid function:
-function 𝛔(a)
-    trunc = 8.0 # truncation to avoid numerical over/underflow
-    a = clamp.(a,-trunc,trunc)
-    p = exp.(a)
-    p = p ./ (1 .+ p)
-    return p
-end
-
-# Negative log likelihood
-function 𝓁(w,w_0,H_0,X,y)
-    N = length(y)
-    D = size(X)[2]
-    μ = 𝛔(X*w)
-    Δw = w-w_0
-    return - ∑( y[n] * log(μ[n]) + (1-y[n]) * log(1-μ[n]) for n=1:N) + 1/2 * Δw'H_0*Δw
-end
-
-# Gradient:
-function ∇𝓁(w,w_0,H_0,X,y)
-    N = length(y)
-    μ = 𝛔(X*w)
-    Δw = w-w_0
-    g = ∑((μ[n]-y[n]) * X[n,:] for n=1:N)
-    return g + H_0*Δw
-end
-
-# Hessian:
-function ∇∇𝓁(w,w_0,H_0,X,y)
-    N = length(y)
-    μ = 𝛔(X*w)
-    H = ∑(μ[n] * (1-μ[n]) * X[n,:] * X[n,:]' for n=1:N)
-    return H + H_0
-end
-
-# Main function:
-struct BayesLogreg
-    μ::Vector{Float64}
-    Σ::Matrix{Float64}
-end
-function bayes_logreg(X,y;w_0=nothing,H_0=nothing,𝓁=𝓁,∇𝓁=∇𝓁,∇∇𝓁=∇∇𝓁,constant=true,λ=0.005,optim_options...)
-    # Setup:
-    if constant
-        if !all(X[:,1] .== 1)
-            X = hcat(ones(size(X)[1]), X)
-        else
-        end
-    end
-    N, D = size(X)
-    if isnothing(w_0)
-        w_0 = zeros(D)
-    end
-    if isnothing(H_0)
-        H_0 = UniformScaling(λ)
-    end
-    
-    # Model:
-    w_map, H_map = newton(𝓁, w_0, ∇𝓁, ∇∇𝓁, (w_0=w_0, H_0=H_0, X=X, y=y), optim_options...) # fit the model (find mode of posterior distribution)
[…]

### 2022-07-13 dev/notebooks/synthetic.qmd
finally sorted out stuff on CounterfactualExplanations side and basic experiment code now running.

@@
     chosen = rand(candidates, Int(round(μ*length(candidates))))
-    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
+    H₀ = mod.model.H # use posterior as new prior
+    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs, H₀=H₀)
     for i in chosen
@@
 
+![](../artifacts/upload/www/synthetic/poc.png)
+
 ## Experiments
@@
 ```{julia}
+using AlgorithmicRecourseDynamics.Experiments
 # Variables:
@@
 grid_ = Experiments.GridVariables(μ, γ)
-n_rounds = 10
-n_folds = 5
+n_rounds = 5
+n_folds = 1
 target = 1.0

### 2022-07-13 src/models/laplace.jl

@@ -14,7 +14,7 @@ using Parameters
 @with_kw struct LaplaceModelParams 
     loss::Symbol = :logitbinarycrossentropy
     opt::Symbol = :Adam
-    n_epochs::Int = 200
+    n_epochs::Int = 10
     λ::Real = 1
     H₀::Union{Nothing, AbstractMatrix} = nothing
 end
@@ -27,7 +27,7 @@ using CounterfactualExplanations
 
 Wrapper function to retrain `LaplaceModel`.
 """
-function train(M::LaplaceModel, data::CounterfactualData; kwargs...)
+function train(M::LaplaceModel, data::CounterfactualData; τ=nothing, kwargs...)
 
     args = LaplaceModelParams(; kwargs...)
 

### 2022-07-13 src/models/mlp.jl

@@ -2,7 +2,7 @@ using Parameters
 @with_kw struct FluxModelParams 
     loss::Symbol = :logitbinarycrossentropy
     opt::Symbol = :Adam
-    n_epochs::Int = 200 
+    n_epochs::Int = 10
 end
 
 using Flux
@@ -13,7 +13,7 @@ using CounterfactualExplanations
 
 Wrapper function to retrain `FluxModel`.
 """
-function train(M::FluxModel, data::CounterfactualData; kwargs...)
+function train(M::FluxModel, data::CounterfactualData; τ=nothing, kwargs...)
 
     args = FluxModelParams(; kwargs...)
 
@@ -35,7 +35,7 @@ function train(M::FluxModel, data::CounterfactualData; kwargs...)
 end
 
 using Statistics
-function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=200)
+function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
 
     # Loss:
     loss_(x, y) = getfield(Flux.Losses, loss)(model(x), y) 

### 2022-07-13 src/post_processing/functions.jl

@@ -60,27 +60,7 @@ function plot_results(results::NamedTuple; title="", id_vars=[:t, :μ, :γ])
 
 end
 
-using BSON
-"""
-    save_path(root,path)
 
-Helper function to save `path` output from `run_experiment` to BSON.
-"""
-function save_path(root,path)
-    bson(root * "_path.bson",Dict(i => path[i] for i ∈ 1:length(path)))
-end
-
-using BSON
-"""
-    load_path(root,path)
-
-Helper function to load `path` output.
-"""
-function load_path(root)
-    dict = BSON.load(root * "_path.bson")
-    path = [dict[i] for i ∈ 1:length(dict)]
-    return path
-end
 
 using Plots
 """

### 2022-07-18 dev/notebooks/synthetic.qmd
some progress on experiment set up

--- 2022-07-13 dev/notebooks/synthetic.qmd
@@
 ```{julia}
-Random.seed!(1234)
 using MLJ
@@
 μ = 0.10
-Random.seed!(123)
 candidates = findall(ys.==0)
@@
     X′[:,i] = counterfactual(outcome)
-    y′[i] = counterfactual_label(outcome)
+    y′[i] = first(counterfactual_label(outcome))
 end
@@
     counterfactual_data′ = CounterfactualData(X′,y′')
-    Random.seed!(123)
     candidates = findall(y′.==0)
     chosen = rand(candidates, Int(round(μ*length(candidates))))
-    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
+    H₀ = mod.model.H # use posterior as new prior
+    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs, H₀=H₀)
     for i in chosen
@@
         X′[:,i] = counterfactual(outcome)
-        y′[i] = counterfactual_label(outcome)
+        y′[i] = first(counterfactual_label(outcome))
     end
@@
 γ = [0.50,0.75,0.9]
-grid_ = Experiments.GridVariables(μ, γ)
-n_rounds = 10
-n_folds = 5
+grid_ = Experiments.GridVariables(models, generators, μ, γ)
+n_rounds = 5
+n_folds = 1
 target = 1.0
@@
     y = vec(path.y̲')
-    plt = plot_contour(X,y,path.𝑴,length_out=length_out,colorbar=cb,title=title)
+    plt = plot_contour(X,y,path.M,length_out=length_out,colorbar=cb,title=title)
     plt = plot(plt, axis=nothing, size=size, legend=false)

### 2022-07-18 dev/resources/misc/boston_housing.qmd
some progress on experiment set up

@@
 @load joinpath(output_folder, "nn.bson") nn
-𝑴ₙₙ = Models.FittedNeuralNet(nn, opt, loss);
+Mₙₙ = Models.FittedNeuralNet(nn, opt, loss);
 ```
@@
 ensemble = Models.load_ensemble(root=joinpath(output_folder, "ensemble"))
-𝑴 = Models.FittedEnsemble(ensemble, opt, loss_type);
+M = Models.FittedEnsemble(ensemble, opt, loss_type);
 ```
@@
 # Experiment:
-experiment = Experiments.Experiment(X,y,𝑴ₙₙ,target,grid_,n_rounds);
+experiment = Experiments.Experiment(X,y,Mₙₙ,target,grid_,n_rounds);
 ```
@@
 # Experiment:
-experiment = Experiments.Experiment(X,y,𝑴,target,grid_,n_rounds);
+experiment = Experiments.Experiment(X,y,M,target,grid_,n_rounds);
 ```

### 2022-07-18 dev/resources/misc/synthetic_bayesian_ar.qmd
some progress on experiment set up

--- 2022-04-29 dev/work/synthetic_bayesian_ar.qmd
@@
 using CounterfactualExplanations.Models: BayesianLogisticModel
-𝑴 = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ); # repare for usage with CounterfactualExplanations.jl
+M = BayesianLogisticModel(reshape(model.μ,1,length(model.μ)), model.Σ); # repare for usage with CounterfactualExplanations.jl
 x̅ = [50.0, 60.0]; # factual
@@
 Random.seed!(1234)
-y̅ = round(CounterfactualExplanations.Models.probs(𝑴, x̅)[1])
+y̅ = round(CounterfactualExplanations.Models.probs(M, x̅)[1])
 target = ifelse(y̅==1.0,0.0,1.0) # opposite label as target
@@
 ```{julia}
-recourse_generic = generate_recourse(generic, x̅, 𝑴, target, γ) # generate recourse
-recourse_greedy = generate_recourse(greedy, x̅, 𝑴, target, γ);
+recourse_generic = generate_recourse(generic, x̅, M, target, γ) # generate recourse
+recourse_greedy = generate_recourse(greedy, x̅, M, target, γ);
 ```

### 2022-07-18 src/evaluation/Evaluation.jl

@@ -1,0 +1,16 @@
+module Evaluation
+
+using Distances
+using StatsBase
+using LinearAlgebra
+using ..Experiments
+using ..Experiments: Experiment, RecourseSystem
+const MetricOrFun = Union{PreMetric,Function}
+
+include("kernels.jl")
+include("mmd.jl")
+include("domain_shifts.jl")
+include("model_shifts.jl")
+
+
+end

### 2022-07-18 src/evaluation/domain_shifts.jl

@@ -1,0 +1,8 @@
+using CounterfactualExplanations
+using CounterfactualExplanations.DataPreprocessing: unpack
+function mmd_domain(experiment::Experiment; target_only::Bool=true)
+    X, y = unpack(experiment.data)
+    new_X, new_y = unpack(experiment.newdata)
+    _classes = target_only ? experiment.target : sort(unique(y))
+    return map(cls -> (mmd(X[:,y.==cls], new_X[:,new_y.==cls])..., cls), _classes)
+end

### 2022-07-18 src/models/deep_ensemble.jl

@@ -88,29 +88,29 @@ struct FittedNeuralNet <: CounterfactualExplanations.AbstractFittedModel
 end
 
 """
-    logits(𝑴::FittedNeuralNet, X::AbstractArray)
+    logits(M::FittedNeuralNet, X::AbstractArray)
 
 A method (extension) that computes predicted logits for a single deep neural network.
 """
-logits(𝑴::FittedNeuralNet, X::AbstractArray) = 𝑴.nn(X)
+logits(M::FittedNeuralNet, X::AbstractArray) = M.nn(X)
 
 """
-    probs(𝑴::FittedNeuralNet, X::AbstractArray)
+    probs(M::FittedNeuralNet, X::AbstractArray)
 
 A method (extension) that computes predicted probabilities for a single deep neural network.
 """
-probs(𝑴::FittedNeuralNet, X::AbstractArray) = Flux.σ.(logits(𝑴, X))
+probs(M::FittedNeuralNet, X::AbstractArray) = Flux.σ.(logits(M, X))
 
 """
-    retrain(𝑴::FittedNeuralNet, data; n_epochs=10)
+    retrain(M::FittedNeuralNet, data; n_epochs=10)
 
 Retrains a fitted a neural network for (new) data.
 """
-function retrain(𝑴::FittedNeuralNet, data; n_epochs=10, τ=1.0) 
-    nn = 𝑴.nn
-    nn = forward_nn(nn, 𝑴.loss, data, 𝑴.opt, n_epochs=n_epochs, τ=τ)
-    𝑴 = FittedNeuralNet(nn, 𝑴.opt, 𝑴.loss)
-    return 𝑴
+function retrain(M::FittedNeuralNet, data; n_epochs=10, τ=1.0) 
+    nn = M.nn
+    nn = forward_nn(nn, M.loss, data, M.opt, n_epochs=n_epochs, τ=τ)
+    M = FittedNeuralNet(nn, M.opt, M.loss)
+    return M
 end
 
 using Statistics
@@ -192,28 +192,28 @@ struct FittedEnsemble <: CounterfactualExplanations.AbstractFittedModel
 end
 
 """
-    logits(𝑴::FittedEnsemble, X::AbstractArray)
+    logits(M::FittedEnsemble, X::AbstractArray)
 
 A method (extension) that computes predicted logits for a deep ensemble.
 """
[…]

### 2022-07-19 dev/notebooks/synthetic.qmd
code for experiment running, including with mutli-threading

@@
 generators = (
-    Bayesian=GreedyGenerator(loss=:logitbinarycrossentropy), 
-    Plugin=GenericGenerator(loss=:logitbinarycrossentropy,ϵ=0.01)
+    Greedy=GreedyGenerator(loss=:logitbinarycrossentropy), 
+    Generic=GenericGenerator(loss=:logitbinarycrossentropy),
+    REVISE=REVISEGenerator(loss=:logitbinarycrossentropy)
 )
@@
 mod = M_laplace
-gen = generators[:Plugin]
+gen = generators[:Greedy]
 plt_original = plot(mod,counterfactual_data;zoom=0,colorbar=false,title="(a)")
@@
 using AlgorithmicRecourseDynamics.Experiments
+using AlgorithmicRecourseDynamics.Experiments: Experiment
+target = 1.0
+experiment = Experiment(counterfactual_data, target, models, generators)
+```
+
+```{julia}
 # Variables:
-μ = [0.05,0.1,0.25]
-γ = [0.50,0.75,0.9]
-grid_ = Experiments.GridVariables(models, generators, μ, γ)
-n_rounds = 5
-n_folds = 1
-target = 1.0
+μ = 0.1
+γ = 0.75
+n_rounds = 10
+n_folds = 5
 T = 1000
+```
+
+```{julia}
+run_experiment(experiment; n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
 ```

### 2022-07-19 src/evaluation/domain_shifts.jl

@@ -1,8 +1,18 @@
 using CounterfactualExplanations
 using CounterfactualExplanations.DataPreprocessing: unpack
-function mmd_domain(experiment::Experiment; target_only::Bool=true)
+using DataFrames
+function mmd_domain(experiment::Experiment, recourse_system::RecourseSystem, n=1000; target_only::Bool=true, as_dataframe=true, kwargs...)
     X, y = unpack(experiment.data)
-    new_X, new_y = unpack(experiment.newdata)
+    new_X, new_y = unpack(recourse_system.data)
     _classes = target_only ? experiment.target : sort(unique(y))
-    return map(cls -> (mmd(X[:,y.==cls], new_X[:,new_y.==cls])..., cls), _classes)
+    metric = map(cls -> (mmd(X[:,vec(y.==cls)], new_X[:,vec(new_y.==cls)], n; kwargs...)..., cls), _classes)
+    if as_dataframe
+        if !isa(metric, AbstractVector)
+            metric = [metric]
+        end
+        metric = DataFrame(NamedTuple{(:value, :p_value, :class)}.(metric))
+        metric.metric .= :mmd
+        metric.scope .= :domain
+    end
+    return metric
 end

### 2022-07-19 src/experiments/functions.jl

@@ -1,7 +1,7 @@
 using CounterfactualExplanations
 
 using Parameters
-@with_kw mutable struct FixedParamters
+@with_kw mutable struct FixedParameters
     n_rounds::Int = 5
     n_folds::Int = 5
     seed::Union{Nothing, Int} = nothing
@@ -9,34 +9,43 @@ using Parameters
     μ::AbstractFloat = 0.05
     γ::AbstractFloat = 0.75
     intersect_::Bool = true
+    τ::AbstractFloat = 1.0
 end
 
 mutable struct Experiment
     data::CounterfactualExplanations.CounterfactualData
     target::Number
-    grid::AbstractArray
-    fixed_parameters::Union{Nothing,FixedParamters}
+    recourse_systems::AbstractArray
+    system_identifiers::Base.Iterators.ProductIterator
+    fixed_parameters::Union{Nothing,FixedParameters}
 end
 
+"""
+    Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple)
+
+
+"""
 function Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple)
     
     # Set up grid
     grid = Base.Iterators.product(models, generators)
-    grid = map(grid) do vars
-        model = vars[1]
+    recourse_systems = map(grid) do vars
+        newdata = deepcopy(data)
+        model = vars[1] # initial model is owned by the recourse systems
+        newmodel = deepcopy(model)
         generator = vars[2]
-        recourse_system = RecourseSystem(data, model, generator)
-        grid_element = Dict(
-            :model => model,
-            :recourse_system => recourse_system
-        )
-        return grid_element
+        recourse_system = RecourseSystem(newdata, newmodel, generator, model)
+        return recourse_system
     end
 
+    # Add system identifiers:
+    system_identifiers = Base.Iterators.product(keys(models), keys(generators))
+
     experiment = Experiment(
-        data,
+        data, # initial data is owned by the experiment, shared across recourse systems
         target,
-        grid,
+        recourse_systems,
+        system_identifiers,
         nothing
     )
 
@@ -55,6 +64,7 @@ mutable struct RecourseSystem
     data::CounterfactualExplanations.CounterfactualData
     model::CounterfactualExplanations.AbstractFittedModel
     generator::CounterfactualExplanations.Generators.AbstractGenerator
+    initial_model::CounterfactualExplanations.AbstractFittedModel
 end
 
 using StatsBase
[…]

### 2022-07-19 src/models/mlp.jl

@@ -52,7 +52,6 @@ function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
             end
             update!(opt_, Flux.params(model), gs)
         end
-        @info "Epoch $epoch. Average loss: $(avg_loss(data))" 
     end
 
 end

### 2022-07-20 dev/notebooks/synthetic.qmd
more work on streamlingin experiment

@@
 γ = 0.75
-n_rounds = 10
+n_rounds = 2
 n_folds = 5
@@
 ```{julia}
-run_experiment(experiment; n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
+output = run_experiment(experiment; n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
 ```
@@
 ```{julia}
-# Experiments:
-experiment = Experiments.Experiment(X,ys,models[:Plugin],target,grid_,n_rounds)
-experiment_plugin = Experiments.Experiment(X,ys,models[:Plugin],target,grid_,n_rounds)
-experiments = (experiment, experiment_plugin)
+using CSV, BSON
+CSV.write(joinpath(output_path,"synthetic_data.csv"), output)
+BSON.@save joinpath(output_path,"synthetic_data.bson") output
 ```
-
-```{julia}
-#| eval: false
-using Random, CSV, DataFrames
-Random.seed!(1234)
-for j in 1:length(generators)
-    outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
-    Experiments.save_path(joinpath(output_path, string(keys(generators)[j])),path)
-    CSV.write(joinpath(output_path, string(keys(generators)[j]) * ".csv"), DataFrame(outcome))
-end
-```
-
-```{julia}
-paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)]
-```
-
-```{julia}
-function plot_state(path, t; γ=0.75, μ = 0.05, k = 1, length_out=500, size=(150,150), title="")
-    chosen = map(p -> p.γ == γ && p.μ == μ && p.t==t && p.k==k, path)
-    path = path[chosen][1]
-    X = path.X̲'
-    y = vec(path.y̲')
-    plt = plot_contour(X,y,path.M,length_out=length_out,colorbar=cb,title=title)
-    plt = plot(plt, axis=nothing, size=size, legend=false)
-end
-plot_state(paths[2], 1)
-```
-
-```{julia}
-#| eval: false
-using Main.Experiments: plot_path
-paths = [Experiments.load_path(joinpath(output_path, string(keys(generators)[j]))) for j ∈ 1:length(generators)]
-anim = Experiments.plot_path(paths[1],Γ=γ,𝙈=μ)
-gif(anim,joinpath(www_path,"bayesian.gif"), fps=2)
-anim = Experiments.plot_path(paths[2],Γ=γ,𝙈=μ)
-gif(anim,joinpath(www_path,"plugin.gif"), fps=2)
-```
-
-For the Bayesian classifier with greedy recourse...
-
-![](www/synthetic/bayesian.gif)
-
-For the non-Bayesian classifier with generic recourse...
-
-![](www/synthetic/plugin.gif)
-
-```{julia}
-using Main.Experiments: prepare_results, plot_results
-validity = DataFrame()
-cost = DataFrame()
-files = (Bayesian=joinpath(output_path, "Bayesian.csv"), Plugin=joinpath(output_path, "Plugin.csv"))
-results = map(file -> CSV.read(file, DataFrame), files)
-p_val, p_cost = plot_results(results);
-```
-
-```{julia}
-p_val
-```
-
-```{julia}
-p_cost
-```
-
-## Neural network
-
-```{julia}
-using Flux
-data = Models.prepare_data(X', y')
-nn = Models.build_model(input_dim=size(X')[1], n_hidden=100)
-loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y);
-```
-
-Training a single neural network...
-
-```{julia}
-using BSON
-run = false
-opt = ADAM()
-if run
-  # Train model:
-  using Flux.Optimise: update!, ADAM
-  using Statistics, StatsBase
-  epochs = 10
-  avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
-  accuracy(data) = sum(map(d ->round.(Flux.σ.(nn(d[1]))) .== d[2], data))[1]/length(data)
-
-  using Plots
-  anim = Animation()
-  avg_l = [avg_loss(data)]
-  p1 = scatter(ylim=(0,avg_l[1]), xlim=(0,epochs), legend=false, xlab="Epoch", title="Average loss")
-  acc = [accuracy(data)]
-  p2 = scatter(ylim=(0.5,1), xlim=(0,epochs), legend=false, xlab="Epoch", title="Accuracy")
-
-  τ = 1.1
-  stopping_criterium_reached = accuracy(data) >= τ
-  epoch = 1
-
-  while epoch <= epochs && !stopping_criterium_reached
-    for d in data
-      gs = gradient(Flux.Flux.params(nn)) do
-        l = loss(d...)
-      end
-      update!(opt, Flux.Flux.params(nn), gs)
-    end
-    avg_l = vcat(avg_l,avg_loss(data))
-    plot!(p1, [0:epoch], avg_l, color=1)
-    scatter!(p1, [0:epoch], avg_l, color=1)
-    acc = vcat(acc,accuracy(data))
-    plot!(p2, [0:epoch], acc, color=1)
-    scatter!(p2, [0:epoch], acc, color=1)
-    plt=plot(p1,p2, size=(600,300))
-    frame(anim, plt)
-
-    # Check if desired accuracy reached:
-    stopping_criterium_reached = accuracy(data) >= τ
-
-    epoch += 1
-
-  end
-
-  gif(anim, joinpath(www_path
-, "single_nn.gif"), fps=10)
-
-  BSON.@save joinpath(output_path
-, "nn.bson") nn
-
-end
-```
-
-```{julia}
-using BSON: @load
-@load joinpath(output_path, "nn.bson") nn
-Mₙₙ = Models.FittedNeuralNet(nn, opt, loss);
-```
-
-Training a deep ensemble...
-
-```{julia}
-opt = ADAM()
-loss_type = :logitbinarycrossentropy
-run = false
-if run
-    K = 50
-    ensemble = Models.build_ensemble(K,kw=(input_dim=size(X')[1], n_hidden=100));
-    ensemble, anim = Models.forward(ensemble, data, opt, n_epochs=10, plot_every=10, loss_type=loss_type, τ=1.1); # fit the ensemble
-    Models.save_ensemble(ensemble, root=joinpath(output_path
-, "ensemble")) # save to disk
-    gif(anim, joinpath(www_path
-, "ensemble_loss.gif"), fps=25);
-end
-```
-
-```{julia}
-ensemble = Models.load_ensemble(root=joinpath(output_path, "ensemble"))
-M = Models.FittedEnsemble(ensemble, opt, loss_type)
-models = (Bayesian=M, Plugin=Mₙₙ);
-```
-
-```{julia}
-using Main.Experiments: plot_contour
-p1 = plot_contour(X,y,models[1],title="Bayesian")
-p2 = plot_contour(X,y,models[2],title="Plugin")
-plot(p1,p2,size=(1000,400))
-```
-
-```{julia}
-# Experiments:
-experiment = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
-experiment_plugin = Experiments.Experiment(X',y',models[:Plugin],target,grid_,n_rounds)
-experiments = (experiment, experiment_plugin);
-```
-
-```{julia}
-run = true
-if run
-    using Random
-    Random.seed!(1234)
-    for j in 1:length(generators)
-        outcome, path = Experiments.run_experiment(experiments[j], generators[j], n_folds, T=T, store_path=true)
-        Experiments.save_path(joinpath(output_path
-    , string(keys(generators)[j])),path)
-        CSV.write(joinpath(output_path
-    , string(keys(generators)[j]) * "_deep.csv"), DataFrame(outcome))
-    end
-end
-```

### 2022-07-20 src/base.jl

@@ -1,53 +1,60 @@
 using Random, StatsBase, LinearAlgebra, Flux
-using .Experiments: Experiment, FixedParameters, choose_individuals, update!
+using .Experiments: Experiment, FixedParameters, choose_individuals, update!, set_up_system_grid!
 using .Evaluation: evaluate_system
 using DataFrames
-using ProgressBars
+using ProgressMeter
+using Logging
+
+is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true")
+
 """
     run_experiment(experiment::Experiment, generator::CounterfactualExplanations.AbstractGenerator, n_folds=5; seed=nothing, T=1000)
 
 A wrapper function that runs the experiment for endogenous models shifts.
 """
-function run_experiment(experiment::Experiment; store_path=true, fixed_parameters...)
+function run_experiment(experiment::Experiment; store_path=true, forward=false, show_progress=!is_logging(stderr), fixed_parameters...)
 
     # Load fixed hyperparameters:
     args = FixedParameters(;fixed_parameters...)
     experiment.fixed_parameters = args
     K, N, intersect_ = args.n_folds, args.n_rounds, args.intersect_
-    M = length(experiment.recourse_systems)
+    M = length(experiment.system_identifiers)
 
     # Setup:
     if !isnothing(args.seed)
         Random.seed!(args.seed)
     end
+    if !forward
+        set_up_system_grid!(experiment, K)
+    else
+        @assert !isnothing(experiment.recourse_systems) "Cannot forward an experiment that has never been run."
+    end
 
     # Pre-allocate memory:
     output = DataFrame()
 
-    fold_iter = ProgressBar(1:K)
-    round_iter = ProgressBar(1:N)
-    system_iter = ProgressBar(1:M)
-    for k in fold_iter
-        set_description(fold_iter, string("Fold $k out of $K"))
-        chosen_individuals = zeros(size(experiment.recourse_systems))
-        for n in round_iter
-            set_description(round_iter, string("Round $n out of $N"))
-
+    p_fold = Progress(K; desc="Progress on folds:", showspeed=true, enabled=show_progress, output = stderr)
+    p_round = Progress(N; desc="Progress on rounds:", showspeed=true, enabled=show_progress, output = stderr)
+    # p_system = Progress(M; desc="Progress on systems:", showspeed=true, enabled=show_progress, output = stderr)
+    for k in 1:K
+        @info "Fold $k"
[…]

### 2022-07-20 src/experiments/functions.jl

@@ -15,9 +15,11 @@ end
 mutable struct Experiment
     data::CounterfactualExplanations.CounterfactualData
     target::Number
-    recourse_systems::AbstractArray
+    recourse_systems::Union{Nothing,AbstractArray}
     system_identifiers::Base.Iterators.ProductIterator
     fixed_parameters::Union{Nothing,FixedParameters}
+    models::NamedTuple
+    generators::NamedTuple
 end
 
 """
@@ -27,31 +29,40 @@ end
 """
 function Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple)
     
-    # Set up grid
-    grid = Base.Iterators.product(models, generators)
-    recourse_systems = map(grid) do vars
-        newdata = deepcopy(data)
-        model = vars[1] # initial model is owned by the recourse systems
-        newmodel = deepcopy(model)
-        generator = vars[2]
-        recourse_system = RecourseSystem(newdata, newmodel, generator, model)
-        return recourse_system
-    end
-
     # Add system identifiers:
     system_identifiers = Base.Iterators.product(keys(models), keys(generators))
 
     experiment = Experiment(
         data, # initial data is owned by the experiment, shared across recourse systems
         target,
-        recourse_systems,
+        nothing,
         system_identifiers,
-        nothing
+        nothing,
+        models,
+        generators
     )
 
     return experiment
 end
 
+function set_up_system_grid!(experiment::Experiment, K::Int=1)
+
+    data = experiment.data
+    grid = Base.Iterators.product(experiment.models, experiment.generators)
+    
+    # Set up systems grid
+    recourse_systems = map(1:K) do k
+        recourse_systems = map(grid) do vars
+            newdata = deepcopy(data)
+            model = vars[1] # initial model is owned by the recourse systems
+            newmodel = deepcopy(model)
+            generator = vars[2]
+            recourse_system = RecourseSystem(newdata, newmodel, generator, model, nothing)
+            return recourse_system
+        end
+    end
+    experiment.recourse_systems = recourse_systems
+end
 
 using CounterfactualExplanations
 
@@ -65,6 +76,7 @@ mutable struct RecourseSystem
     model::CounterfactualExplanations.AbstractFittedModel
     generator::CounterfactualExplanations.Generators.AbstractGenerator
     initial_model::CounterfactualExplanations.AbstractFittedModel
[…]

### 2022-07-20 dev/notebooks/synthetic.qmd
some more progress on experiment setup

@@
 ```{julia}
+# Models:
+models = (Bayesian=M_laplace, Plugin=M)
+# Generators:
+loss = :logitbinarycrossentropy
+generators = (
+    Greedy=GreedyGenerator(loss=loss), 
+    Generic=GenericGenerator(loss=loss),
+    REVISE=REVISEGenerator(loss=loss)
+)
+```
+
+```{julia}
 using AlgorithmicRecourseDynamics.Experiments
@@
 γ = 0.75
-n_rounds = 2
+n_rounds = 10
 n_folds = 5
@@
 output = run_experiment(experiment; n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
-```
-
-```{julia}
 using CSV, BSON

### 2022-07-20 src/base.jl

@@ -12,7 +12,7 @@ is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true
 
 A wrapper function that runs the experiment for endogenous models shifts.
 """
-function run_experiment(experiment::Experiment; store_path=true, forward=false, show_progress=!is_logging(stderr), fixed_parameters...)
+function run_experiment(experiment::Experiment; evaluate_every=10, store_path=true, forward=false, show_progress=!is_logging(stderr), fixed_parameters...)
 
     # Load fixed hyperparameters:
     args = FixedParameters(;fixed_parameters...)
@@ -34,14 +34,12 @@ function run_experiment(experiment::Experiment; store_path=true, forward=false, 
     output = DataFrame()
 
     p_fold = Progress(K; desc="Progress on folds:", showspeed=true, enabled=show_progress, output = stderr)
-    p_round = Progress(N; desc="Progress on rounds:", showspeed=true, enabled=show_progress, output = stderr)
-    # p_system = Progress(M; desc="Progress on systems:", showspeed=true, enabled=show_progress, output = stderr)
+    @info "Running experiment ..."
     for k in 1:K
-        @info "Fold $k"
         recourse_systems = experiment.recourse_systems[k]
         chosen_individuals = zeros(size(recourse_systems))
+        p_round = Progress(N; desc="Progress on rounds:", showspeed=true, enabled=show_progress, output = stderr)
         for n in 1:N
-            @info "Round $n"
             # Choose individuals that shall receive recourse:
             chosen_individuals_n = choose_individuals(experiment, recourse_systems; intersect_=intersect_)
             chosen_individuals = map((x,y) -> union(x,y),chosen_individuals,chosen_individuals_n)
@@ -54,18 +52,23 @@ function run_experiment(experiment::Experiment; store_path=true, forward=false, 
                     update!(experiment, recourse_system, chosen_individuals_m)
                 end
                 # Evaluate:
-                evaluation = evaluate_system(recourse_system, experiment)
-                # Store results:
-                evaluation.k .= k
-                evaluation.n .= n
-                evaluation.model .= collect(experiment.system_identifiers)[m][1]
-                evaluation.generator .= collect(experiment.system_identifiers)[m][2]
-                evaluation.n_individuals .= length(chosen_individuals[m])
-                evaluation.pct_total .= length(chosen_individuals[m])/size(experiment.data.y,2)
-                output = vcat(output, evaluation)
-                # next!(p_system)
[…]

### 2022-07-20 src/evaluation/Evaluation.jl

@@ -7,16 +7,20 @@ using ..Experiments
 using ..Experiments: Experiment, RecourseSystem
 const MetricOrFun = Union{PreMetric,Function}
 
+include("utils.jl")
 include("kernels.jl")
 include("mmd.jl")
 include("domain_shifts.jl")
 include("model_shifts.jl")
 
 using DataFrames
-function evaluate_system(recourse_system::RecourseSystem, experiment::Experiment)
+function evaluate_system(recourse_system::RecourseSystem, experiment::Experiment; n=1000)
     vcat(
-        mmd_model(experiment, recourse_system, 1000),
-        mmd_domain(experiment, recourse_system, 1000)
+        mmd_domain(experiment, recourse_system; n=n),
+        mmd_model(experiment, recourse_system; n=n),
+        mmd_model(experiment, recourse_system; n=n, grid_search=true),
+        disagreement(experiment, recourse_system),
+        decisiveness(experiment, recourse_system)
     )
 end
 

### 2022-07-20 src/evaluation/utils.jl

@@ -1,0 +1,9 @@
+function metric_to_dataframe(metric, name, scope)
+    if !isa(metric, AbstractVector)
+        metric = [metric]
+    end
+    metric = DataFrame(NamedTuple{(:value, :p_value, :class)}.(metric))
+    metric.metric .= name
+    metric.scope .= scope
+    return metric
+end

### 2022-07-21 dev/notebooks/synthetic.qmd
some more work on streamlining experiments.

--- 2022-07-20 dev/notebooks/synthetic.qmd
@@
 ```{julia}
+# Models:
+models = (Bayesian=M_laplace, Plugin=M)
+# Generators:
+loss = :logitbinarycrossentropy
+generators = (
+    Greedy=GreedyGenerator(loss=loss), 
+    Generic=GenericGenerator(loss=loss),
+    REVISE=REVISEGenerator(loss=loss),
+    DICE=DiCEGenerator(loss=loss)
+)
+```
+
+```{julia}
 using AlgorithmicRecourseDynamics.Experiments
@@
 target = 1.0
-experiment = Experiment(counterfactual_data, target, models, generators)
+experiment = Experiment(counterfactual_data, target, models, generators, 5)
 ```
@@
 γ = 0.75
-n_rounds = 2
-n_folds = 5
+n_rounds = 10
+n_folds = 1
 T = 1000

### 2022-07-21 src/experiments/functions.jl

@@ -20,6 +20,7 @@ mutable struct Experiment
     fixed_parameters::Union{Nothing,FixedParameters}
     models::NamedTuple
     generators::NamedTuple
+    num_counterfactuals::Int
 end
 
 """
@@ -27,7 +28,7 @@ end
 
 
 """
-function Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple)
+function Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple, num_counterfactuals::Int=1)
     
     # Add system identifiers:
     system_identifiers = Base.Iterators.product(keys(models), keys(generators))
@@ -39,7 +40,8 @@ function Experiment(data::CounterfactualExplanations.CounterfactualData, target:
         system_identifiers,
         nothing,
         models,
-        generators
+        generators,
+        num_counterfactuals
     )
 
     return experiment
@@ -132,9 +134,12 @@ function update!(experiment::Experiment, recourse_system::RecourseSystem, chosen
     # Generate recourse:
     for i in chosen_individuals
         x = X[:,i]
-        outcome = generate_counterfactual(x, target, counterfactual_data, M, generator; T=T, γ=γ)
-        X[:,i] = counterfactual(outcome) # update individuals' features
-        y[:,i] .= first(counterfactual_label(outcome)) # update individuals' predicted label
+        outcome = generate_counterfactual(x, target, counterfactual_data, M, generator; T=T, γ=γ, num_counterfactuals=experiment.num_counterfactuals)
+        X′ = counterfactual(outcome)
+        y′ = counterfactual_label(outcome)
+        idx = rand(1:experiment.num_counterfactuals) # randomly draw from generated counterfactuals
+        X[:,i] = selectdim(X′,3,idx) # update individuals' features
+        y[:,i] .= selectdim(y′,3,idx) # update individuals' predicted label
     end
 
     # Update data and classifier:

### 2022-07-23 dev/notebooks/real_world.qmd
final streamlining

---
title: Experiments with Real-World Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("real_world")
www_path = www_dir("real_world")
```

## Models

```{julia}
# Models:
model = Models.build_mlp()
λ = 0.1
model_laplace = Laplace(model, λ=λ)
M = FluxModel(model)
M_laplace = LaplaceReduxModel(model_laplace)
models = Dict(:laplace=>M_laplace, :mlp=>M)
```

## Generators

### 2022-07-23 dev/notebooks/synthetic.qmd
final streamlining

--- 2022-07-20 dev/notebooks/synthetic.qmd
@@
 
-## Classifiers
+## Experiments
+
+### Data
 
 ```{julia}
-using MLJ
-N = 1000
-X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-2 => 2), cluster_std=0.1)
-ys .= ys.==2
-X = X'
-xs = Flux.unstack(X,2)
-data = zip(xs,ys)
-counterfactual_data = CounterfactualData(X,ys')
-```
-
-### Logistic Regression
-
-```{julia}
-using Flux
-model = Chain(Dense(2,1))
-M = FluxModel(model)
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
 ```
@@
 ```{julia}
-n_epochs = 10
 using AlgorithmicRecourseDynamics.Models
-M = Models.train(M,counterfactual_data; n_epochs=n_epochs)
+using Flux
+using AlgorithmicRecourseDynamics.Experiments
+using AlgorithmicRecourseDynamics.Experiments: Experiment
+using CSV, BSON
+
+function _run_one()
+    @info "Starting $(key)"
+
+    # Models:
+    model = Models.build_mlp()
+    λ = 0.1
+    model_laplace = Laplace(model, λ=λ)
+    M = FluxModel(model)
+    M_laplace = LaplaceReduxModel(model_laplace)
+    models = (Bayesian=M_laplace, Plugin=M)
+
+    # Pretrain:
+    data_train, data_test = Models.train_test_split(counterfactual_data)
+    models = map(model -> Models.train(model, data_train; n_epochs=100), models)
+
+    # Generators:
+    generators = (
+        Greedy=GreedyGenerator(), 
+        Generic=GenericGenerator(),
+        REVISE=REVISEGenerator(),
+        DICE=DiCEGenerator()
+    )
+
+    # Variables:
+    μ = 0.1
+    γ = 0.75
+    n_rounds = 10
+    n_folds = 5
+    T = 1000
+    target = 1.0
+    experiment = Experiment(data_train, data_test, target, models, generators, 5)
+
+    # Run:
+    output = run_experiment(experiment; evaluate_every=2, n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
+
+    CSV.write(joinpath(output_path,"output_$(key).csv"), output)
+    BSON.@save joinpath(output_path,"output_$(key).bson") output
+    BSON.@save joinpath(output_path,"experiment_$(key).bson") experiment
+
+    @info "Successfully saved $(key)"
+
+    return output, experiment
+end
+
+import AlgorithmicRecourseDynamics: run_experiment
+function run_experiment(catalogue::Dict)
+    output = [_run_one() for (key,counterfactual_data) in catalogue]
+    return output
+end
 ```
@@
 ```{julia}
-plt_nn = plot(M,counterfactual_data;zoom=0)
-savefig(plt_nn, joinpath(www_path, "nn_contour.png"))
-```
-
-![](../artifacts/upload/www/synthetic/nn_contour.png)
-
-### Laplace Redux
-
-```{julia}
-λ = 0.1
-model_laplace = Laplace(M.model, λ=λ)
-LaplaceRedux.fit!(model_laplace, data)
+choices = [:linearly_separable, :overlapping, :plus_shaped, :balanced_positive_clusters]
+data_sets = filter(p -> p[1] in choices, catalogue)
+run_experiment(data_sets)
 ```
@@
 ```{julia}
-M_laplace = Models.LaplaceModel(model_laplace)
+BSON.@load joinpath(output_path,"output_linearly_separable.bson") output
 ```
@@
 ```{julia}
-plt_laplace = plot(M_laplace, counterfactual_data;zoom=0)
-savefig(plt_laplace, joinpath(www_path, "la_contour.png"))
+output[output.name.==:mmd_grid,:]
 ```
-
-![](../artifacts/upload/www/synthetic/la_contour.png)
-
-## Single Round
-
-```{julia}
-# Models:
-models = (Bayesian=M_laplace, Plugin=M)
-# Generators:
-generators = (
-    Greedy=GreedyGenerator(loss=:logitbinarycrossentropy), 
-    Generic=GenericGenerator(loss=:logitbinarycrossentropy),
-    REVISE=REVISEGenerator(loss=:logitbinarycrossentropy)
-)
-```
-
-### Generate counterfactual
-
-```{julia}
-cb = false
-γ = 0.75
-mod = M_laplace
-gen = generators[:Greedy]
-plt_original = plot(mod,counterfactual_data;zoom=0,colorbar=false,title="(a)")
-```
-
-```{julia}
-μ = 0.10
-candidates = findall(ys.==0)
-chosen = rand(candidates, Int(round(μ*length(candidates))))
-X′ = copy(X)
-y′ = copy(ys)
-using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
-for i in chosen
-    x = X[:,i]
-    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen,γ=γ)
-    X′[:,i] = counterfactual(outcome)
-    y′[i] = first(counterfactual_label(outcome))
-end
-counterfactual_data′ = CounterfactualData(X′,y′')
-plt_single = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(b)")
-```
-
-### Retrain
-
-```{julia}
-mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
-plt_single_retrained = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(c)")
-```
-
-### Repeat
-
-```{julia}
-i = 2
-while i <= 10
-    counterfactual_data′ = CounterfactualData(X′,y′')
-    candidates = findall(y′.==0)
-    chosen = rand(candidates, Int(round(μ*length(candidates))))
-    H₀ = mod.model.H # use posterior as new prior
-    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs, H₀=H₀)
-    for i in chosen
-        x = X′[:,i]
-        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen; γ=γ)
-        X′[:,i] = counterfactual(outcome)
-        y′[i] = first(counterfactual_label(outcome))
-    end
-    i += 1
-end
-plt_single_repeat = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(d)")
-```
-
-```{julia}
-plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
-savefig(plt, joinpath(www_path, "poc.png"))
-```
-
-![](../artifacts/upload/www/synthetic/poc.png)
-
-## Experiments
-
-```{julia}
-using AlgorithmicRecourseDynamics.Experiments
-using AlgorithmicRecourseDynamics.Experiments: Experiment
-target = 1.0
-experiment = Experiment(counterfactual_data, target, models, generators)
-```
-
-```{julia}
-# Variables:
-μ = 0.1
-γ = 0.75
-n_rounds = 2
-n_folds = 5
-T = 1000
-```
-
-```{julia}
-output = run_experiment(experiment; n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
-```
-
-```{julia}
-using CSV, BSON
-CSV.write(joinpath(output_path,"synthetic_data.csv"), output)
-BSON.@save joinpath(output_path,"synthetic_data.bson") output
-```

### 2022-07-23 src/AlgorithmicRecourseDynamics.jl

@@ -3,6 +3,7 @@ module AlgorithmicRecourseDynamics
 # Load modules:
 include("data/Data.jl")
 using .Data
+export load_synthetic, load_real_world
 
 include("models/Models.jl")
 using .Models
@@ -12,6 +13,7 @@ using .Experiments
 
 include("evaluation/Evaluation.jl")
 using .Evaluation
+export evaluate_system
 
 include("base.jl")
 export run_experiment

### 2022-07-23 src/base.jl

@@ -1,18 +1,22 @@
-using Random, StatsBase, LinearAlgebra, Flux
-using .Experiments: Experiment, FixedParameters, choose_individuals, update!, set_up_system_grid!
+using .Experiments: Experiment, FixedParameters, set_up_system_grid!, update!
 using .Evaluation: evaluate_system
+using Random, StatsBase, LinearAlgebra, Flux
 using DataFrames
 using ProgressMeter
 using Logging
+using Statistics
+using CounterfactualExplanations
 
 is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true")
 
 """
-    run_experiment(experiment::Experiment, generator::CounterfactualExplanations.AbstractGenerator, n_folds=5; seed=nothing, T=1000)
+    run!(experiment::Experiment, generator::CounterfactualExplanations.AbstractGenerator, n_folds=5; seed=nothing, T=1000)
 
 A wrapper function that runs the experiment for endogenous models shifts.
 """
-function run_experiment(experiment::Experiment; evaluate_every=10, store_path=true, forward=false, show_progress=!is_logging(stderr), fixed_parameters...)
+function run!(
+    experiment::Experiment; evaluate_every=10, n_boostrap=1000, forward=false, show_progress=!is_logging(stderr), fixed_parameters...
+)
 
     # Load fixed hyperparameters:
     args = FixedParameters(;fixed_parameters...)
@@ -53,14 +57,20 @@ function run_experiment(experiment::Experiment; evaluate_every=10, store_path=tr
                 end
                 # Evaluate:
                 if n % evaluate_every == 0 
-                    evaluation = evaluate_system(recourse_system, experiment)
+                    evaluation = evaluate_system(recourse_system, experiment, n=n_boostrap)
                     # Store results:
                     evaluation.k .= k
                     evaluation.n .= n
                     evaluation.model .= collect(experiment.system_identifiers)[m][1]
                     evaluation.generator .= collect(experiment.system_identifiers)[m][2]
                     evaluation.n_individuals .= length(chosen_individuals[m])
-                    evaluation.pct_total .= length(chosen_individuals[m])/size(experiment.data.y,2)
+                    evaluation.pct_total .= length(chosen_individuals[m])/size(experiment.train_data.y,2)
+                    # Add recourse measures:
[…]

### 2022-07-23 src/evaluation/utils.jl

@@ -1,9 +1,0 @@
-function metric_to_dataframe(metric, name, scope)
-    if !isa(metric, AbstractVector)
-        metric = [metric]
-    end
-    metric = DataFrame(NamedTuple{(:value, :p_value, :class)}.(metric))
-    metric.metric .= name
-    metric.scope .= scope
-    return metric
-end

### 2022-07-23 src/experiments/functions.jl

@@ -1,4 +1,4 @@
-using CounterfactualExplanations
+using CounterfactualExplanations, DataFrames
 
 using Parameters
 @with_kw mutable struct FixedParameters
@@ -14,6 +14,8 @@ end
 
 mutable struct Experiment
     data::CounterfactualExplanations.CounterfactualData
+    train_data::CounterfactualExplanations.CounterfactualData
+    test_data::CounterfactualExplanations.CounterfactualData
     target::Number
     recourse_systems::Union{Nothing,AbstractArray}
     system_identifiers::Base.Iterators.ProductIterator
@@ -23,18 +25,28 @@ mutable struct Experiment
     num_counterfactuals::Int
 end
 
+using CounterfactualExplanations.DataPreprocessing
 """
     Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple)
 
 
 """
-function Experiment(data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple, num_counterfactuals::Int=1)
+function Experiment(
+    train_data::CounterfactualExplanations.CounterfactualData, test_data::CounterfactualExplanations.CounterfactualData, target::Number, models::NamedTuple, generators::NamedTuple, num_counterfactuals::Int=1
+)
     
     # Add system identifiers:
     system_identifiers = Base.Iterators.product(keys(models), keys(generators))
 
+    # Full data:
+    X_train, y_train = DataPreprocessing.unpack(train_data)
+    X_test, y_test = DataPreprocessing.unpack(train_data)
+    data = CounterfactualData(hcat(X_train, X_test), hcat(y_train, y_test))
+
     experiment = Experiment(
-        data, # initial data is owned by the experiment, shared across recourse systems
+        data, # initial data is owned by the experiment, shared across recourse systems,
+        train_data,
+        test_data,
         target,
         nothing,
         system_identifiers,
@@ -49,17 +61,17 @@ end
 
 function set_up_system_grid!(experiment::Experiment, K::Int=1)
 
-    data = experiment.data
+    data = experiment.train_data
     grid = Base.Iterators.product(experiment.models, experiment.generators)
     
     # Set up systems grid
     recourse_systems = map(1:K) do k
-        recourse_systems = map(grid) do vars
+        map(grid) do vars
             newdata = deepcopy(data)
[…]

### 2022-07-23 src/experiments/utils.jl

@@ -9,28 +9,3 @@ function expandgrid(x,y)
         Matrix
     return grid
 end
-
-# Train-test split
-using StatsBase
-function train_test_split(X,y;test_size=0.2)
-    N = size(X)[1]
-    rows = 1:N
-    test_rows = StatsBase.sample(rows,Int(floor(test_size * N)),replace=false)
-    train_rows = setdiff(rows, test_rows)
-    X_train = X[train_rows,:]
-    y_train = y[train_rows]
-    X_test = X[test_rows,:]
-    y_test = y[test_rows]
-    return X_train, y_train, X_test, y_test
-end
-
-# K-fold
-using MLDataUtils
-function kfolds_(N;k=5)
-    rows = 1:N
-    rows_shuffled = StatsBase.sample(rows,N,replace=false)
-    folds = kfolds(rows_shuffled, k)
-    train_indices = [folds.data[idx] for idx in folds.train_indices]
-    test_indices = [folds.data[idx] for idx in folds.val_indices]
-    return train_indices, test_indices
-end

### 2022-07-23 src/models/mlp.jl

@@ -3,6 +3,7 @@ using Parameters
     loss::Symbol = :logitbinarycrossentropy
     opt::Symbol = :Adam
     n_epochs::Int = 10
+    data_loader::Function = data_loader
 end
 
 using Flux
@@ -18,7 +19,8 @@ function train(M::FluxModel, data::CounterfactualData; τ=nothing, kwargs...)
     args = FluxModelParams(; kwargs...)
 
     # Prepare data:
-    data = data_loader(data)
+    data = args.data_loader(data)
+
     # Training:
     model = M.model
     forward!(
@@ -27,8 +29,6 @@ function train(M::FluxModel, data::CounterfactualData; τ=nothing, kwargs...)
         opt = args.opt,
         n_epochs = args.n_epochs
     )
-    # Declare type:
-    M = FluxModel(model)
 
     return M
     
@@ -56,8 +56,7 @@ function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
 
 end
 
-using MLDataUtils
-
+using MLUtils
 """
     data_loader(data::CounterfactualData)
 
@@ -65,7 +64,7 @@ Prepares data for training.
 """
 function data_loader(data::CounterfactualData)
     X, y = CounterfactualExplanations.DataPreprocessing.unpack(data)
-    xs = Flux.unstack(X,2) 
+    xs = MLUtils.unstack(X,dims=2) 
     data = zip(xs,y)
     return data
 end

### 2022-07-24 dev/notebooks/real_world.qmd
final streamlining and now running for credit data

@@
 
-## Models
+```{julia}
+catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
+choices = [:german_credit_numerical, :give_me_some_credit_balanced]
+data_sets = filter(p -> p[1] in choices, catalogue)
+```
 
 ```{julia}
-# Models:
-model = Models.build_mlp()
-λ = 0.1
-model_laplace = Laplace(model, λ=λ)
-M = FluxModel(model)
-M_laplace = LaplaceReduxModel(model_laplace)
-models = Dict(:laplace=>M_laplace, :mlp=>M)
+models = [:FluxModel, :LaplaceReduxModel]
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator()
+)
 ```
 
-## Generators
+```{julia}
+run_experiments(data_sets,models,generators)
+```

### 2022-07-24 src/base.jl

@@ -1,4 +1,4 @@
-using .Experiments: Experiment, FixedParameters, set_up_system_grid!, update!
+using .Experiments: Experiment, FixedParameters, set_up_system_grid!, update!, choose_individuals
 using .Evaluation: evaluate_system
 using Random, StatsBase, LinearAlgebra, Flux
 using DataFrames
@@ -90,11 +90,12 @@ end
 using DataFrames, CSV, BSON
 function run_experiment(
     data::CounterfactualData,
-    models::Dict{Symbol, CounterfactualExplanations.Models.AbstractFittedModel},
-    generators::Dict{Symbol, CounterfactualExplanations.Generators.AbstractGenerator};
+    models::Dict{Symbol, <: CounterfactualExplanations.Models.AbstractFittedModel},
+    generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
     target::Int=1,
     num_counterfactuals::Int=5,
-    pre_train_models::Union{Nothing, Int}=100,
+    evaluate_every::Int=2,
+    pre_train_models::Union{Nothing,Int}=100,
     save_path::Union{Nothing,String}=nothing,
     save_name::Union{Nothing,String}=nothing,
     kwargs...
@@ -107,11 +108,11 @@ function run_experiment(
 
     # Pretrain:
     if !isnothing(pre_train_models)
-        models = map(model -> Models.train(model, data_train; n_epochs=pre_train_models), models)
+        map!(model -> Models.train(model, data_train; n_epochs=pre_train_models), values(models))
     end
 
     # Run:
-    experiment = Experiment(data_train, data_test, target, models, generators, num_counterfactuals)
+    experiment = Experiment(data_train, data_test, target, models, deepcopy(generators), num_counterfactuals)
     output = run!(experiment; evaluate_every=evaluate_every, kwargs...)
 
     # Save to disk:
@@ -127,12 +128,41 @@ function run_experiment(
     
 end
 
-function run_experiments(
-    catalogue::Dict{Symbol, CounterfactualData},
-    models::Dict{Symbol, CounterfactualExplanations.Models.AbstractFittedModel},
-    generators::Dict{Symbol, CounterfactualExplanations.Generators.AbstractGenerator};
+function run_experiment(
+    data::CounterfactualData,
+    models::Vector{Symbol},
+    generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
     target::Int=1,
     num_counterfactuals::Int=5,
[…]

### 2022-07-25 dev/notebooks/real_world.qmd
added plotting function for experiment results

@@
 catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
-choices = [:german_credit_numerical, :give_me_some_credit_balanced]
+choices = [:give_me_some_credit_balanced]
 data_sets = filter(p -> p[1] in choices, catalogue)
@@
 ```{julia}
-run_experiments(data_sets,models,generators)
+results = run_experiments(data_sets,models,generators;save_path=output_path,evaluate_every=10,n_rounds=50)
 ```
+
+```{julia}
+[plot(res) for res in results]
+```

### 2022-07-25 dev/notebooks/synthetic.qmd
added plotting function for experiment results

@@
 ```
-
-In this notebook we will use toy data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.
 
@@
 
-## Experiments
-
-### Data
-
 ```{julia}
 catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
+choices = [:linearly_separable, :overlapping, :plus_shaped, :balanced_positive_clusters]
+data_sets = filter(p -> p[1] in choices, catalogue)
 ```
@@
 ```{julia}
-using AlgorithmicRecourseDynamics.Models
-using Flux
-using AlgorithmicRecourseDynamics.Experiments
-using AlgorithmicRecourseDynamics.Experiments: Experiment
-using CSV, BSON
-
-function _run_one()
-    @info "Starting $(key)"
-
-    # Models:
-    model = Models.build_mlp()
-    λ = 0.1
-    model_laplace = Laplace(model, λ=λ)
-    M = FluxModel(model)
-    M_laplace = LaplaceReduxModel(model_laplace)
-    models = (Bayesian=M_laplace, Plugin=M)
-
-    # Pretrain:
-    data_train, data_test = Models.train_test_split(counterfactual_data)
-    models = map(model -> Models.train(model, data_train; n_epochs=100), models)
-
-    # Generators:
-    generators = (
-        Greedy=GreedyGenerator(), 
-        Generic=GenericGenerator(),
-        REVISE=REVISEGenerator(),
-        DICE=DiCEGenerator()
+models = [:FluxModel, :LaplaceReduxModel]
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator()
 )
-
-    # Variables:
-    μ = 0.1
-    γ = 0.75
-    n_rounds = 10
-    n_folds = 5
-    T = 1000
-    target = 1.0
-    experiment = Experiment(data_train, data_test, target, models, generators, 5)
-
-    # Run:
-    output = run_experiment(experiment; evaluate_every=2, n_rounds=n_rounds, μ=μ, n_folds=n_folds, γ=γ, T=T)
-
-    CSV.write(joinpath(output_path,"output_$(key).csv"), output)
-    BSON.@save joinpath(output_path,"output_$(key).bson") output
-    BSON.@save joinpath(output_path,"experiment_$(key).bson") experiment
-
-    @info "Successfully saved $(key)"
-
-    return output, experiment
-end
-
-import AlgorithmicRecourseDynamics: run_experiment
-function run_experiment(catalogue::Dict)
-    output = [_run_one() for (key,counterfactual_data) in catalogue]
-    return output
-end
 ```
@@
 ```{julia}
-choices = [:linearly_separable, :overlapping, :plus_shaped, :balanced_positive_clusters]
-data_sets = filter(p -> p[1] in choices, catalogue)
-run_experiment(data_sets)
+results = run_experiments(data_sets,models,generators;save_path=output_path,evaluate_every=10,n_rounds=50)
 ```
@@
 ```{julia}
-BSON.@load joinpath(output_path,"output_linearly_separable.bson") output
+[plot(res) for res in results]
 ```
-
-```{julia}
-output[output.name.==:mmd_grid,:]
-```

### 2022-07-25 src/base.jl

@@ -86,6 +86,10 @@ function run!(
 
 end
 
+struct ExperimentResults 
+    output::DataFrame
+    experiment::Experiment
+end
 
 using DataFrames, CSV, BSON
 function run_experiment(
@@ -115,16 +119,22 @@ function run_experiment(
     experiment = Experiment(data_train, data_test, target, models, deepcopy(generators), num_counterfactuals)
     output = run!(experiment; evaluate_every=evaluate_every, kwargs...)
 
+    @info "Completed experiment."
+
+    results = Experiment(output,experiment)
+
     # Save to disk:
     if !isnothing(save_path)
         save_name = isnothing(save_name) ? "experiment" : "experiment_$(save_name)"
-        mkpath(joinpath(save_path,save_name))
+        save_path = joinpath(save_path,save_name)
+        mkpath(save_path)
         CSV.write(joinpath(save_path,"output.csv"), output)
         BSON.@save joinpath(save_path,"output.bson") output
         BSON.@save joinpath(save_path,"experiment.bson") experiment
+        BSON.@save joinpath(save_path,"results.bson") results
     end
 
-    @info "Completed experiment."
+    return results
     
 end
 
@@ -146,12 +156,14 @@ function run_experiment(
 
     models = Dict([(model,getfield(AlgorithmicRecourseDynamics.Models, model)(data)) for model in models])
 
-    run_experiment(
+    output = run_experiment(
         data, models, generators;
         target=target,num_counterfactuals=num_counterfactuals,pre_train_models=100,
         save_path=save_path,save_name=save_name,evaluate_every=evaluate_every,
         kwargs...
     )
+
+    return output
 
 end
 

### 2022-07-25 src/data/functions.jl

@@ -23,7 +23,7 @@ function load_real_world()
     data = map(files) do file
         df = CSV.read(joinpath(data_dir, file), DataFrame)
         X = Matrix(df[:,Not(:target)])
-        X = reshape(X,size(X,2),size(X,1))
+        X = permutedims(X)
         y = convert(Matrix, df.target')
         data = CounterfactualData(X,y)
         (Symbol(replace(file, ".csv" => "")) => data)

### 2022-07-25 src/plotting.jl

@@ -1,0 +1,23 @@
+using Gadfly
+import Plots: plot
+using DataFrames
+
+function plot(results::ExperimentResults, variable::Symbol=:mmd)
+    
+    df = results.output
+    @assert variable in unique(df.name) "Not a valid variable."
+
+    gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+    df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+
+
+    # Plot validity:
+    plt = Gadfly.plot(
+        df_plot[df_plot.name.==variable,:], 
+        ygroup=:scope, xgroup=:model, x=:n, y=:mean, ymin=:ymin, ymax=:ymax, color=:generator,
+        Geom.subplot_grid(Geom.line, Geom.ribbon, free_y_axis=true),
+    )
+
+    return plt
+
+end

### 2022-07-26 dev/notebooks/credit.qmd
bleh

---
title: Experiments with Credit Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("real_world")
www_path = www_dir("real_world")
```

```{julia}
catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
choices = [:give_me_some_credit_balanced]
data_sets = filter(p -> p[1] in choices, catalogue)
```

```{julia}
using CounterfactualExplanations.DataPreprocessing: unpack
bs = 50
function data_loader(data::CounterfactualData)
    X, y = unpack(data)
    data = Flux.DataLoader((X,y),batchsize=bs)
    return data
end
model_params = (batch_norm=false,n_hidden=128,dropout=true,p_dropout=0.3)
```

```{julia}
models = [:FluxModel, :FluxEnsemble]
generators = Dict(
    :Greedy=>GreedyGenerator(), 
    :Generic=>GenericGenerator(),
    :REVISE=>REVISEGenerator(),
    :DICE=>DiCEGenerator()
)
```

```{julia}
experiments = set_up_experiments(data_sets,models,generators; pre_train_models=100, model_params=model_params, data_loader=data_loader)
```

```{julia}
plts = []
for (exp_name, exp) in experiments
    for (M_name, M) in exp.models
        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
    end
end
plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
savefig(plt, joinpath(www_path,"models_test_before.png"))
```

```{julia}
n_rounds = 10
evaluate_every = n_rounds/5
using Serialization
results = run_experiments(experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds)
Serialization.serialize(joinpath(output_path,"results.bson"),results)
```

```{julia}
plts = []
for (exp_name, exp) in experiments
    for (M_name, M) in exp.models
        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
    end
end
plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
savefig(plt, joinpath(www_path,"models_test_after.png"))
```

```{julia}
using Serialization
Serialization.deserialize(joinpath(output_path,"results.bson"),results)
using Gadfly
plts = [plot(res) for res in values(results)]
gridstack(reshape(plts,2,2))
```

### 2022-07-26 dev/notebooks/mnist.qmd
bleh

---
title: Experiments with Credit Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LinearAlgebra
theme(:wong)
output_path = output_dir("mnist")
www_path = www_dir("mnist")
```

```{julia}
using Flux
using CounterfactualExplanations.Data: mnist_data, mnist_model, mnist_ensemble, mnist_vae
X, ys = mnist_data()
model = mnist_model()
ensemble = mnist_ensemble()
vae = mnist_vae()
```

```{julia}
n_per_class = 100
test_ratio = 0.2
train_data = CounterfactualData(X,ys)
using AlgorithmicRecourseDynamics.Models: undersample
train_data = undersample(train_data, n_per_class)
train_data.generative_model = vae
using MLDatasets
test_data = MNIST(:test)
X_test, y_test = test_data[:]
X_test = Flux.flatten(X_test)
y_test = Flux.onehotbatch(y_test,0:9)
test_data = undersample(CounterfactualData(X_test, y_test),Int(round(n_per_class*test_ratio)))
```

```{julia}
M = FluxModel(model, :classification_multi)
M_ens = FluxEnsemble(ensemble, :classification_multi)
models = Dict(:FluxModel => M, :FluxEnsemble => M_ens)
generators = Dict(
    :Greedy=>GreedyGenerator(), 
    :Generic=>GenericGenerator(),
    :REVISE=>REVISEGenerator(),
    # :DICE=>DiCEGenerator(),
)
```

```{julia}
experiment = set_up_experiment(train_data,test_data,models,generators;target=5)
```

```{julia}
n_evaluations = 1
n_rounds = 5
evaluate_every = Int(round(n_rounds/n_evaluations))
using Serialization
results = run_experiment(experiment;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds)
Serialization.serialize(joinpath(output_path,"results.bson"),results)
```

### 2022-07-26 src/AlgorithmicRecourseDynamics.jl

@@ -16,7 +16,7 @@ using .Evaluation
 export evaluate_system
 
 include("base.jl")
-export run_experiment, run_experiments, ExperimentResults 
+export run_experiment, set_up_experiment, run_experiments, set_up_experiments, ExperimentResults 
 
 include("plotting.jl")
 

### 2022-07-26 src/base.jl

@@ -6,6 +6,7 @@ using ProgressMeter
 using Logging
 using Statistics
 using CounterfactualExplanations
+using Serialization
 
 is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true")
 
@@ -86,15 +87,186 @@ function run!(
 
 end
 
+"""
+    set_up_experiment(
+        data::CounterfactualData,
+        models::Dict{Symbol, <: CounterfactualExplanations.Models.AbstractFittedModel},
+        generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
+        target::Int=1,
+        num_counterfactuals::Int=5,
+        pre_train_models::Union{Nothing,Int}=100,
+        kwargs...
+    )
+    
+Sets up one experiment for the provided data, models and generators.
+"""
+function set_up_experiment(
+    data_train::CounterfactualData,
+    data_test::CounterfactualData,
+    models::Dict{Symbol, <: CounterfactualExplanations.Models.AbstractFittedModel},
+    generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
+    target::Int=1,
+    num_counterfactuals::Int=5,
+    kwargs...
+)
+
+    experiment = Experiment(data_train, data_test, target, models, deepcopy(generators), num_counterfactuals)
+
+    # Sanity check:
+    @info "Initial model scores:"
+    println(experiment.initial_model_scores)
+
+    return experiment
+    
+end
+
+"""
+    set_up_experiment(
+        data::CounterfactualData,
+        models::Dict{Symbol, <: CounterfactualExplanations.Models.AbstractFittedModel},
+        generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
+        target::Int=1,
+        num_counterfactuals::Int=5,
+        pre_train_models::Union{Nothing,Int}=100,
+        kwargs...
+    )
+    
+Sets up one experiment for the provided data, models and generators.
+"""
+function set_up_experiment(
+    data::CounterfactualData,
+    models::Vector{Symbol},
+    generators::Dict{Symbol, <: CounterfactualExplanations.Generators.AbstractGenerator};
+    model_params::NamedTuple=(batch_norm=false,dropout=false,activation=Flux.relu),
+    target::Int=1,
+    num_counterfactuals::Int=5,
+    pre_train_models::Union{Nothing,Int}=100,
+    kwargs...
+)
+
[…]

### 2022-07-26 src/evaluation/Evaluation.jl

@@ -3,6 +3,8 @@ module Evaluation
 using Distances
 using StatsBase
 using LinearAlgebra
+using ..Models
+import ..Models: perturbation
 using ..Experiments
 using ..Experiments: Experiment, RecourseSystem
 const MetricOrFun = Union{PreMetric,Function}
@@ -19,6 +21,7 @@ function evaluate_system(recourse_system::RecourseSystem, experiment::Experiment
     
     metrics = [
         mmd_domain(experiment, recourse_system; n=n),
+        perturbation(experiment, recourse_system),
         mmd_model(experiment, recourse_system; n=n),
         mmd_model(experiment, recourse_system; n=n, grid_search=true),
         disagreement(experiment, recourse_system),

### 2022-07-26 src/evaluation/mmd.jl

@@ -30,7 +30,10 @@ function mmd(x::AbstractArray, y::AbstractArray, k::AbstractKernel=GaussianKerne
     end
     return mmd_, p_val
 end
-mmd(x::AbstractArray, y::AbstractArray, n::Int, k::AbstractKernel=GaussianKernel(), dist=pairwisel2; compute_p::Union{Nothing,Int}=1000) = mmd(samplecolumns(x,n), samplecolumns(y,n), k, dist; compute_p=compute_p)
+function mmd(x::AbstractArray, y::AbstractArray, n::Int, k::AbstractKernel=GaussianKernel(), dist=pairwisel2; compute_p::Union{Nothing,Int}=1000)
+    n = minimum([size(x,2),n])
+    mmd(samplecolumns(x,n), samplecolumns(y,n), k, dist; compute_p=compute_p)
+end
 
 using Random: shuffle
 """

### 2022-07-26 src/evaluation/model_shifts.jl

@@ -17,6 +17,22 @@ function DataFrame(metric::ModelMetric)
     return df  
 end
 
+using LinearAlgebra
+function perturbation(experiment::Experiment, recourse_system::RecourseSystem)
+
+    # Initial:
+    M = recourse_system.initial_model
+    # New:
+    new_M = recourse_system.model
+
+    value = perturbation(M, new_M)
+
+    metric = ModelMetric(value,missing,:perturbation)
+
+    return metric
+
+end
+
 using CounterfactualExplanations
 using CounterfactualExplanations.Models: probs
 using CounterfactualExplanations.DataPreprocessing: unpack

### 2022-07-26 src/models/Models.jl

@@ -1,10 +1,11 @@
 module Models
 
-# Models:
-include("mlp.jl")
-include("laplace.jl")
-
 include("utils.jl")
+
+# Models:
+include("mlp.jl") # including logistic regression
+include("laplace.jl")
+include("deep_ensemble.jl")
 
 abstract type AbstractTrainableModel end
 

### 2022-07-26 src/models/mlp.jl

@@ -14,7 +14,7 @@ using CounterfactualExplanations
 
 Wrapper function to retrain `FluxModel`.
 """
-function train(M::FluxModel, data::CounterfactualData; τ=nothing, kwargs...)
+function train(M::FluxModel, data::CounterfactualData; kwargs...)
 
     args = FluxModelParams(; kwargs...)
 
@@ -35,7 +35,7 @@ function train(M::FluxModel, data::CounterfactualData; τ=nothing, kwargs...)
 end
 
 using Statistics
-function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
+function forward!(model::Flux.Chain, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
 
     # Loss:
     loss_(x, y) = getfield(Flux.Losses, loss)(model(x), y) 
@@ -56,19 +56,6 @@ function forward!(model, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
 
 end
 
-using MLUtils
-"""
-    data_loader(data::CounterfactualData)
-
-Prepares data for training.
-"""
-function data_loader(data::CounterfactualData)
-    X, y = CounterfactualExplanations.DataPreprocessing.unpack(data)
-    xs = MLUtils.unstack(X,dims=2) 
-    data = zip(xs,y)
-    return data
-end
-
 """
     build_mlp()
 
@@ -83,12 +70,20 @@ nn = build_mlp()
 """
 function build_mlp(;
     input_dim::Int=2,n_hidden::Int=32,n_layers::Int=2,output_dim::Int=1,
-    batch_norm::Bool=false,dropout::Bool=false,activation=Flux.relu
+    batch_norm::Bool=false,dropout::Bool=false,activation=Flux.relu,
+    p_dropout=0.25
 )
 
-    @assert n_layers >= 2 "Need at least two layers."
-    
-    if batch_norm
+    @assert n_layers >= 1 "Need at least one layer."
+
+    if n_layers == 1
+
+        @assert output_dim==1 "Expected output dimension of 1 for logisitic regression, got $output_dim."
+
+        # Logistic regression:
+        model = Chain(Dense(input_dim, output_dim))
+
+    elseif batch_norm
 
         hidden_ = repeat([Dense(n_hidden,n_hidden),BatchNorm(n_hidden,activation)],n_layers-2)
 
@@ -102,11 +97,11 @@ function build_mlp(;
 
     elseif dropout
 
[…]

### 2022-07-27 dev/notebooks/housing.qmd
uh

--- 2022-07-26 dev/notebooks/credit.qmd
@@
 ---
-title: Experiments with Credit Data
+title: Experiments with Housing Data
 jupyter: julia-1.7
@@
 ```
-
-```{julia}
-catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
-choices = [:give_me_some_credit_balanced]
-data_sets = filter(p -> p[1] in choices, catalogue)
-```
-
-```{julia}
-using CounterfactualExplanations.DataPreprocessing: unpack
-bs = 50
-function data_loader(data::CounterfactualData)
-    X, y = unpack(data)
-    data = Flux.DataLoader((X,y),batchsize=bs)
-    return data
-end
-model_params = (batch_norm=false,n_hidden=128,dropout=true,p_dropout=0.3)
-```
-
-```{julia}
-models = [:FluxModel, :FluxEnsemble]
-generators = Dict(
-    :Greedy=>GreedyGenerator(), 
-    :Generic=>GenericGenerator(),
-    :REVISE=>REVISEGenerator(),
-    :DICE=>DiCEGenerator()
-)
-```
-
-```{julia}
-experiments = set_up_experiments(data_sets,models,generators; pre_train_models=100, model_params=model_params, data_loader=data_loader)
-```
-
-```{julia}
-plts = []
-for (exp_name, exp) in experiments
-    for (M_name, M) in exp.models
-        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
-    end
-end
-plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-savefig(plt, joinpath(www_path,"models_test_before.png"))
-```
-
-```{julia}
-n_rounds = 10
-evaluate_every = n_rounds/5
-using Serialization
-results = run_experiments(experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds)
-Serialization.serialize(joinpath(output_path,"results.bson"),results)
-```
-
-```{julia}
-plts = []
-for (exp_name, exp) in experiments
-    for (M_name, M) in exp.models
-        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
-    end
-end
-plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-savefig(plt, joinpath(www_path,"models_test_after.png"))
-```
-
-```{julia}
-using Serialization
-Serialization.deserialize(joinpath(output_path,"results.bson"),results)
-using Gadfly
-plts = [plot(res) for res in values(results)]
-gridstack(reshape(plts,2,2))
-```

### 2022-07-27 dev/notebooks/synthetic.qmd
uh

@@
 catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
-choices = [:linearly_separable, :overlapping, :plus_shaped, :balanced_positive_clusters]
+choices = [
+    :linearly_separable, 
+    :overlapping, 
+    :plus_shaped, 
+    :balanced_positive_clusters,
+]
 data_sets = filter(p -> p[1] in choices, catalogue)
@@
 ```{julia}
-models = [:FluxModel, :LaplaceReduxModel]
+models = [
+    :LogisticRegression, 
+    :FluxModel, 
+    :FluxEnsemble,
+]
 generators = Dict(
@@
 ```{julia}
-results = run_experiments(data_sets,models,generators;save_path=output_path,evaluate_every=10,n_rounds=50)
+experiments = set_up_experiments(data_sets,models,generators)
 ```
@@
 ```{julia}
-[plot(res) for res in results]
+plts = []
+for (exp_name, exp) in experiments
+    for (M_name, M) in exp.models
+        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
+    end
+end
+plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+savefig(plt, joinpath(www_path,"models_test_before.png"))
 ```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+using Serialization
+results = run_experiments(experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+```{julia}
+plts = []
+for (exp_name, exp) in experiments
+    for (M_name, M) in exp.models
+        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
+    end
+end
+plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+savefig(plt, joinpath(www_path,"models_test_after.png"))
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+```
+
+```{julia}
+using Serialization
+results = [Serialization.deserialize(joinpath(output_path,subfolder,"results.jls")) for subfolder in readdir(output_path)];
+```
+
+```{julia}
+using Gadfly
+plts = [plot(res) for res in values(results)]
+```

### 2022-07-27 src/base.jl

@@ -1,4 +1,4 @@
-using .Experiments: Experiment, FixedParameters, set_up_system_grid!, update!, choose_individuals
+using .Experiments: Experiment, FixedParameters, RecourseSystem, set_up_system_grid!, update!, choose_individuals
 using .Evaluation: evaluate_system
 using Random, StatsBase, LinearAlgebra, Flux
 using DataFrames
@@ -10,13 +10,40 @@ using Serialization
 
 is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true")
 
+function collect_output(
+    experiment::Experiment, recourse_system::RecourseSystem, chosen_individuals::Union{Nothing,AbstractArray}, k::Int, n::Int, m::Int;
+    n_bootstrap=1000
+)
+
+    # Evaluate:
+    output = evaluate_system(recourse_system, experiment, n=n_bootstrap)
+    
+    # Add additional information:
+    output.k .= k
+    output.n .= n
+    output.model .= collect(experiment.system_identifiers)[m][1]
+    output.generator .= collect(experiment.system_identifiers)[m][2]
+    output.n_individuals .= isnothing(chosen_individuals) ? 0 : length(chosen_individuals)
+    output.pct_total .= isnothing(chosen_individuals) ? 0 : length(chosen_individuals)/size(experiment.train_data.y,2)
+
+    # Add recourse measures:
+    if n > 0 
+        bmk = mapcols(mean, recourse_system.benchmark)
+        output.success_rate .=  bmk.success_rate
+        output.distance .= bmk.distance
+        output.redundancy .= bmk.redundancy
+    end
+
+    return output
+end
+
 """
     run!(experiment::Experiment, generator::CounterfactualExplanations.AbstractGenerator, n_folds=5; seed=nothing, T=1000)
 
 A wrapper function that runs the experiment for endogenous models shifts.
 """
 function run!(
-    experiment::Experiment; evaluate_every=10, n_boostrap=1000, forward=false, show_progress=!is_logging(stderr), fixed_parameters...
+    experiment::Experiment; evaluate_every=10, n_bootstrap=1000, forward=false, show_progress=!is_logging(stderr), fixed_parameters...
 )
 
     # Load fixed hyperparameters:
@@ -42,6 +69,12 @@ function run!(
     @info "Running experiment ..."
     for k in 1:K
         recourse_systems = experiment.recourse_systems[k]
+        # Initial evaluation:
+        for m in 1:M
[…]

### 2022-07-27 src/evaluation/kernels.jl

@@ -1,28 +1,16 @@
-export GaussianKernel
+using KernelFunctions
 
-abstract type AbstractKernel end
-
-function kernelsum(k::AbstractKernel, x::AbstractMatrix, y::AbstractMatrix, dist::MetricOrFun)
-    sum(k(dist(x,y))) / (size(x,2) * size(y,2))
+function kernelsum(k::KernelFunctions.Kernel, x::AbstractMatrix, y::AbstractMatrix)
+    m = size(x,2)
+    n = size(y,2)
+    sum(kernelmatrix(k,x,y)) / (m * n)
 end
 
-function kernelsum(k::AbstractKernel, x::AbstractMatrix{T}, dist::MetricOrFun) where T
+LinearAlgebra
+function kernelsum(k::KernelFunctions.Kernel, x::AbstractMatrix)
     l = size(x,2)
-    (sum(k(dist(x,x))) - l*k(T(0)))/(l^2 - l)
+    (sum(kernelmatrix(k,x,x)) - tr(kernelmatrix(k,x,x)))/(l^2 - l)
 end
 
-kernelsum(k::AbstractKernel, x::AbstractVector, dist::MetricOrFun) = zero(eltype(x))
+kernelsum(k::KernelFunctions.Kernel, x::AbstractVector) = zero(eltype(x))
 
-"""
-    GaussianKernel(γ)
-implements the standard Gaussian kernel ``exp(-γ * x)
-"""
-struct GaussianKernel <: AbstractKernel
-    γ::AbstractFloat
-    GaussianKernel(γ) = γ < 0 ? throw(DomainError(γ, "Scale should be positive.")) : new(γ)
-end
-
-GaussianKernel() = GaussianKernel(0.5)
-
-(m::GaussianKernel)(x::Number) = exp(-m.γ * x)
-(m::GaussianKernel)(x::AbstractArray) = exp.(-m.γ .* x)

### 2022-07-27 src/evaluation/model_shifts.jl

@@ -43,12 +43,12 @@ using StatsBase
 
 Calculates the MMD on the probabilities of classification assigned by the model to the set of (all) instances. Allows to quantify the model shift.
 """
-function mmd_model(experiment::Experiment, recourse_system::RecourseSystem; n=1000, grid_search=false, kwargs...)
+function mmd_model(experiment::Experiment, recourse_system::RecourseSystem; n=1000, grid_search=false, n_samples=1000, kwargs...)
     
     X, _ = unpack(experiment.data)
 
     if grid_search
-        X = reduce(hcat,[map(x -> rand(range(x..., length=100)), extrema(X, dims=2)) for i in 1:n])
+        X = reduce(hcat,[map(x -> rand(range(x..., length=100)), extrema(X, dims=2)) for i in 1:n_samples])
     end
     
     # Initial:
@@ -59,7 +59,7 @@ function mmd_model(experiment::Experiment, recourse_system::RecourseSystem; n=10
     new_M = recourse_system.model
     new_proba = probs(new_M, X)
 
-    value, p_value = mmd(proba, new_proba, 1000; compute_p=n, kwargs...)
+    value, p_value = mmd(proba, new_proba, n_samples; compute_p=n, kwargs...)
     metric_name = grid_search ? :mmd_grid : :mmd
 
     metric = ModelMetric(value,p_value,metric_name)
@@ -120,25 +120,28 @@ function fscore(experiment::Experiment, recourse_system::RecourseSystem)
 
     X, y = unpack(experiment.test_data)
     m = MulticlassFScore()
-    binary = length(unique(y)) == 2
+    binary = recourse_system.initial_model.likelihood == :classification_binary
+
+    function compute_fscore(model) 
+        if binary
+            proba = reduce(hcat, map(x -> binary ? [1-x,x] : x, probs(model,X)))
+            ŷ = Flux.onecold(proba, 0:1)
+        else
+            y = Flux.onecold(y,1:size(y,1))
+            ŷ = Flux.onecold(probs(M,X), sort(unique(y)))
+        end
+        fscore = m(ŷ, vec(y))
+        return fscore
+    end
 
     # Initial:
     M = recourse_system.initial_model
-    proba = reduce(hcat, map(x -> length(x) == 1 ? [x,1-x] : x, probs(M,X)))
-    ŷ = Flux.onecold(proba, 1:size(proba,2))
[…]

### 2022-07-27 src/plotting.jl

@@ -2,7 +2,7 @@ using Gadfly
 import Plots: plot
 using DataFrames
 
-function plot(results::ExperimentResults, variable::Symbol=:mmd)
+function plot(results::ExperimentResults, variable::Symbol=:mmd; kwargs...)
     
     df = results.output
     @assert variable in unique(df.name) "Not a valid variable."
@@ -16,6 +16,7 @@ function plot(results::ExperimentResults, variable::Symbol=:mmd)
         df_plot[df_plot.name.==variable,:], 
         ygroup=:scope, xgroup=:model, x=:n, y=:mean, ymin=:ymin, ymax=:ymax, color=:generator,
         Geom.subplot_grid(Geom.line, Geom.ribbon, free_y_axis=true),
+        kwargs...
     )
 
     return plt

### 2022-07-28 dev/notebooks/credit.qmd
implemented idea about gravitational generator

@@
 data_sets = filter(p -> p[1] in choices, catalogue)
+map!(data_sets) data do 
+    using StatsBase
+    dt = fit(ZScoreTransform, X, dims=2)
+    StatsBase.transform!(dt, X)
+end
 ```

### 2022-07-28 dev/notebooks/housing.qmd
implemented idea about gravitational generator

@@
 ```
+
+## Training the classifiers
+
+```{julia}
+using MLDatasets, Statistics, StatsBase
+X = BostonHousing.features()
+y = BostonHousing.targets()
+y = Float64.(y .>= median(y)); # binary target
+dt = fit(ZScoreTransform, X, dims=2)
+StatsBase.transform!(dt, X)
+data = CounterfactualData(X,y)
+```
+
+```{julia}
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 50
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=true,n_hidden=128,dropout=true,p_dropout=0.1)
+models = [:FluxModel, :FluxEnsemble]
+models = Dict([(model,getfield(AlgorithmicRecourseDynamics.Models, model)(data; model_params...)) for model in models])
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models
+# Data:
+data_train, data_test = Models.train_test_split(data)
+n_epochs=100
+map!(model -> Models.train(model, data_train; n_epochs=n_epochs, data_loader=data_loader), values(models))
+
+```
+
+```{julia}
+
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator()
+)
+```
+
+```{julia}
+# Prepare data and model:
+using Random
+Random.seed!(1234)
+using StatsBase
+dt = fit(ZScoreTransform, X, dims=2)
+StatsBase.transform!(dt, X)
+xs = Flux.unstack(X,2)
+data = zip(xs,y)
+nn = Models.build_model(input_dim=size(X)[1], n_hidden=100)
+loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
+```
+
+The model achieves decent training accuracy
+
+```{julia}
+run = false
+if run
+  # Train model:
+  using Flux.Optimise: update!, ADAM
+  using Statistics, StatsBase
+  opt = ADAM()
+  epochs = 100
+  avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
+  accuracy(data) = sum(map(d ->round.(Flux.σ.(nn(d[1]))) .== d[2], data))[1]/length(data)
+
+  using Plots
+  anim = Animation()
+  avg_l = [avg_loss(data)]
+  p1 = scatter( ylim=(0,avg_l[1]), xlim=(0,epochs), legend=false, xlab="Epoch", title="Average loss")
+  acc = [accuracy(data)]
+  p2 = scatter( ylim=(0.5,1), xlim=(0,epochs), legend=false, xlab="Epoch", title="Accuracy")
+
+  τ = 1.0
+  stopping_criterium_reached = accuracy(data) >= τ
+  epoch = 1
+
+  while epoch <= epochs && !stopping_criterium_reached
+    for d in data
+      gs = gradient(params(nn)) do
+        l = loss(d...)
+      end
+      update!(opt, params(nn), gs)
+    end
+    avg_l = vcat(avg_l,avg_loss(data))
+    plot!(p1, [0:epoch], avg_l, color=1)
+    scatter!(p1, [0:epoch], avg_l, color=1)
+    acc = vcat(acc,accuracy(data))
+    plot!(p2, [0:epoch], acc, color=1)
+    scatter!(p2, [0:epoch], acc, color=1)
+    plt=plot(p1,p2, size=(600,300))
+    frame(anim, plt)
+
+    # Check if desired accuracy reached:
+    stopping_criterium_reached = accuracy(data) >= τ
+
+    epoch += 1
+
+  end
+
+  gif(anim, joinpath(www_folder, "single_nn.gif"), fps=10)
+
+  using BSON: @save
+  @save joinpath(output_folder, "nn.bson") nn
+
+end
+```
+
+![](www/boston_housing/single_nn.gif)
+
+Next we will build and train a deep ensemble.
+
+```{julia}
+opt = ADAM()
+loss_type = :logitbinarycrossentropy
+run = false
+if run
+    K = 50
+    ensemble = Models.build_ensemble(K,kw=(input_dim=size(X)[1], n_hidden=100));
+    ensemble, anim = Models.forward(ensemble, data, opt, n_epochs=30, plot_every=10, loss_type=loss_type); # fit the ensemble
+    Models.save_ensemble(ensemble, root=joinpath(output_folder, "ensemble")) # save to disk
+    gif(anim, joinpath(www_folder, "ensemble_loss.gif"), fps=25);
+end
+```
+
+![](www/boston_housing/ensemble_loss.gif)

### 2022-07-28 dev/notebooks/synthetic.qmd
implemented idea about gravitational generator

@@
     :REVISE=>REVISEGenerator(),
-    :DICE=>DiCEGenerator()
+    :DICE=>DiCEGenerator(),
+    :Gravitational=>GravitationalGenerator(),
 )
@@
 ```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
 plts = []
-for (exp_name, exp) in experiments
-    for (M_name, M) in exp.models
-        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
+for (exp_name, exp_) in experiments
+    for (M_name, M) in exp_.models
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$exp_name;\n $M_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plts = vcat(plts..., plt)
     end
@@
 savefig(plt, joinpath(www_path,"models_test_before.png"))
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plts = []
+for (exp_name, exp_) in experiments
+    for (M_name, M) in exp_.models
+        score = round(model_evaluation(M, exp_.train_data),digits=2)
+        plt = plot(M, exp_.train_data, title="$exp_name;\n $M_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.train_data.X)) .!= exp_.train_data.y))
+        x_wrongly_labelled = exp_.train_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plts = vcat(plts..., plt)
+    end
+end
+plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+savefig(plt, joinpath(www_path,"models_train_before.png"))
 ```
@@
 ```{julia}
-plts = []
-for (exp_name, exp) in experiments
-    for (M_name, M) in exp.models
-        plts = vcat(plts..., plot(M, exp.test_data, title="$exp_name;\n $M_name"))
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
     end
 end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
     plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-savefig(plt, joinpath(www_path,"models_test_after.png"))
+    savefig(plt, joinpath(www_path,"models_test_after_$name.png"))
+end
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        data = rec_sys[m].data
+        score = round(model_evaluation(M, data),digits=2)
+        plt = plot(M, data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+        x_wrongly_labelled = data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_train_after_$name.png"))
+end
 ```

### 2022-07-28 src/models/utils.jl

@@ -53,9 +53,10 @@ function data_loader(data::CounterfactualData)
 end
 
 using CounterfactualExplanations
+using CounterfactualExplanations.DataPreprocessing
 using MLJ, Flux
 function model_evaluation(M::CounterfactualExplanations.AbstractFittedModel, test_data::CounterfactualData)
-    X, y = unpack(test_data)
+    X, y = DataPreprocessing.unpack(test_data)
     m = MulticlassFScore()
     binary = M.likelihood == :classification_binary
     if binary

### 2022-07-29 dev/notebooks/gravitational_generator.qmd
more work on gravitational generator. It works as expected if we choose strict convergence: we consider concergence as when the probability threshold is reached AND the reduction in our loss function from one iteration to another is small.

---
title: Experiments with Housing Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("generator")
www_path = www_dir("generator")
```

# `GravitationalGenerator`

```{julia}
using MLJ
N = 1000
X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-5 => 5), cluster_std=0.5)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')
```

```{julia}
using Flux
nn = Chain(Dense(2,1))
using Flux.Optimise: update!, ADAM
opt = ADAM()
epochs = 100
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
show_every = epochs/10
for epoch = 1:epochs
  for d in data
    gs = gradient(Flux.params(nn)) do
      l = loss(d...)
    end
    update!(opt, Flux.params(nn), gs)
  end
  if epoch % show_every == 0
    println("Epoch " * string(epoch))
    @show avg_loss(data)
  end
end
```

```{julia}
M = FluxModel(nn)
```

```{julia}
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
```

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals = vcat(
      counterfactuals...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict)
    )
end
```

```{julia}
theme(:wong)
T = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=T, title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator_strict.png"))
```

![Generating diverse counterfactuals through `GravitationalGenerator`. The magnitude of attraction ($\lambda_2) increases from left to right.](../artifacts/upload/www/generator/gravitational_generator_strict.png){#fig-strict}

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals = vcat(
      counterfactuals...,
      generate_counterfactual(x, target, counterfactual_data, M, generator)
    )
end
```

```{julia}
theme(:wong)
T = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=T, title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1400,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator.png"))
```

![Generating diverse counterfactuals through `GravitationalGenerator`, but converging when threshold is reached. The magnitude of attraction ($\lambda_2) increases from left to right.](../artifacts/upload/www/generator/gravitational_generator.png){#fig-conventional}

# References

### 2022-07-29 dev/notebooks/housing.qmd
more work on gravitational generator. It works as expected if we choose strict convergence: we consider concergence as when the probability threshold is reached AND the reduction in our loss function from one iteration to another is small.

@@
 map!(model -> Models.train(model, data_train; n_epochs=n_epochs, data_loader=data_loader), values(models))
-
+initial_model_scores = [(name,Models.model_evaluation(model,data_train),Models.model_evaluation(model,data_test)) for (name,model) in pairs(models)]
 ```

### 2022-07-31 dev/notebooks/housing.qmd
wuoohoo

--- 2022-07-27 dev/notebooks/housing.qmd
@@
 ---
-title: Experiments with Housing Data
+title: Preprocessing Housing Data
 jupyter: julia-1.7
@@
 www_path = www_dir("real_world")
+data_path = data_dir("real_world")
 ```
+
+```{julia}
+using MLDatasets, Statistics, StatsBase
+X = Matrix(BostonHousing().features)
+dt = fit(ZScoreTransform, X, dims=1)
+StatsBase.transform!(dt, X)
+using DataFrames
+df = DataFrame(X,:auto)
+# Target:
+y = BostonHousing().targets.MEDV
+y = Float64.(y .>= median(y)); # binary target
+df.target .= y
+using CSV
+CSV.write(joinpath(data_path, "housing.csv"), df)
+```

### 2022-07-31 dev/notebooks/real_world.qmd
wuoohoo

--- 2022-07-23 dev/notebooks/real_world.qmd
@@
 ---
-title: Experiments with Real-World Data
+title: Experiments with Credit Data
 jupyter: julia-1.7
@@
 
-## Models
+```{julia}
+catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
+choices = [:give_me_some_credit_balanced, :housing]
+data_sets = filter(p -> p[1] in choices, catalogue)
+```
 
 ```{julia}
-# Models:
-model = Models.build_mlp()
-λ = 0.1
-model_laplace = Laplace(model, λ=λ)
-M = FluxModel(model)
-M_laplace = LaplaceReduxModel(model_laplace)
-models = Dict(:laplace=>M_laplace, :mlp=>M)
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 50
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=false,n_hidden=128,dropout=true,p_dropout=0.1)
 ```
 
-## Generators
+```{julia}
+models = [:FluxModel, :FluxEnsemble]
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator(),
+    :Gravitational=>GravitationalGenerator(),
+)
+```
+
+```{julia}
+experiments = set_up_experiments(data_sets,models,generators; pre_train_models=100, model_params=model_params, data_loader=data_loader)
+```
+
+```{julia}
+# #| eval: false
+# using AlgorithmicRecourseDynamics.Models: model_evaluation
+# plts = []
+# for (exp_name, exp_) in experiments
+#     for (M_name, M) in exp_.models
+#         score = round(model_evaluation(M, exp_.test_data),digits=2)
+#         plt = plot(M, exp_.test_data; title="$exp_name;\n $M_name ($score)", dim_red=:umap)
+#         # # Errors:
+#         # ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+#         # x_wrongly_labelled = exp_.test_data.X[:,ids]
+#         # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+#         plts = vcat(plts..., plt)
+#     end
+# end
+# plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+# savefig(plt, joinpath(www_path,"models_test_before.png"))
+```
+
+```{julia}
+# #| eval: false
+# using AlgorithmicRecourseDynamics.Models: model_evaluation
+# plts = []
+# for (exp_name, exp_) in experiments
+#     for (M_name, M) in exp_.models
+#         score = round(model_evaluation(M, exp_.train_data),digits=2)
+#         plt = plot(M, exp_.train_data; title="$exp_name;\n $M_name ($score)", dim_red=:umap)
+#         # # Errors:
+#         # ids = findall(vec(round.(probs(M, exp_.train_data.X)) .!= exp_.train_data.y))
+#         # x_wrongly_labelled = exp_.train_data.X[:,ids]
+#         # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+#         plts = vcat(plts..., plt)
+#     end
+# end
+# plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+# savefig(plt, joinpath(www_path,"models_train_before.png"))
+```
+
+## Simple convergence
+
+```{julia}
+n_evals = 1
+n_rounds = 10
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 2
+n_bootstrap = 1
+using Serialization
+results = run_experiments(
+    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap
+)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+```{julia}
+# using AlgorithmicRecourseDynamics.Models: model_evaluation
+# plot_dict = Dict(key => Dict() for (key,val) in results)
+# fold = 1
+# for (name, res) in results
+#     exp_ = res.experiment
+#     plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+#     rec_sys = exp_.recourse_systems[fold]
+#     sys_ids = collect(exp_.system_identifiers)
+#     M = length(rec_sys)
+#     for m in 1:M
+#         model_name, generator_name = sys_ids[m]
+#         M = rec_sys[m].model
+#         score = round(model_evaluation(M, exp_.test_data),digits=2)
+#         plt = plot(M, exp_.test_data; title="$name;\n $model_name ($score)", dim_red=:umap)
+#         # # Errors:
+#         # ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+#         # x_wrongly_labelled = exp_.test_data.X[:,ids]
+#         # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+#         plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+#     end
+# end
+# plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+# for (name, plts) in plot_dict
+#     plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+#     savefig(plt, joinpath(www_path,"models_test_after_$name.png"))
+# end
+```
+
+```{julia}
+# using AlgorithmicRecourseDynamics.Models: model_evaluation
+# plot_dict = Dict(key => Dict() for (key,val) in results)
+# fold = 1
+# for (name, res) in results
+#     exp_ = res.experiment
+#     plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+#     rec_sys = exp_.recourse_systems[fold]
+#     sys_ids = collect(exp_.system_identifiers)
+#     M = length(rec_sys)
+#     for m in 1:M
+#         model_name, generator_name = sys_ids[m]
+#         M = rec_sys[m].model
+#         data = rec_sys[m].data
+#         score = round(model_evaluation(M, data),digits=2)
+#         plt = plot(M, data; title="$name;\n $model_name ($score)", dim_red=:umap)
+#         # # Errors:
+#         # ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+#         # x_wrongly_labelled = data.X[:,ids]
+#         # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+#         plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+#     end
+# end
+# plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+# for (name, plts) in plot_dict
+#     plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+#     savefig(plt, joinpath(www_path,"models_train_after_$name.png"))
+# end
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+```
+
+## Strict convergence
+
+```{julia}
+n_evals = 1
+n_rounds = 10
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 2
+n_bootstrap = 1
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
+    convergence=:strict, save_name_suffix="strict"
+)
+Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data; title="$name;\n $model_name ($score)", dim_red=:umap)
+        # # Errors:
+        # ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        # x_wrongly_labelled = exp_.test_data.X[:,ids]
+        # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_test_after_$(name)_strict.png"))
+end
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        data = rec_sys[m].data
+        score = round(model_evaluation(M, data),digits=2)
+        plt = plot(M, data; title="$name;\n $model_name ($score)", dim_red=:umap)
+        # # Errors:
+        # ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+        # x_wrongly_labelled = data.X[:,ids]
+        # scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_train_after_$(name)_strict.png"))
+end
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
+```

### 2022-07-31 dev/notebooks/synthetic.qmd
more work on gravitational generator. It works as expected if we choose strict convergence: we consider concergence as when the probability threshold is reached AND the reduction in our loss function from one iteration to another is small.
wuoohoo

@@
 
+## Simple convergence
+
 ```{julia}
-n_evals = 5
-n_rounds = 50
+n_evals = 1
+n_rounds = 1
 evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
+n_folds = 1
+n_bootstrap = 1
 using Serialization
-results = run_experiments(experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds)
+results = run_experiments(
+    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds,n_folds=n_folds,n_bootstrap=n_bootstrap
+)
 Serialization.serialize(joinpath(output_path,"results.jls"),results)
@@
 
+## Strict convergence
+
 ```{julia}
+n_evals = 1
+n_rounds = 1
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 1
+n_bootstrap = 1
 using Serialization
-results = [Serialization.deserialize(joinpath(output_path,subfolder,"results.jls")) for subfolder in readdir(output_path)];
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
+    convergence=:strict, save_name_suffix="strict"
+)
+Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
 ```
@@
 ```{julia}
-using Gadfly
-plts = [plot(res) for res in values(results)]
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_test_after_$(name)_strict.png"))
+end
 ```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        data = rec_sys[m].data
+        score = round(model_evaluation(M, data),digits=2)
+        plt = plot(M, data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+        x_wrongly_labelled = data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_train_after_$(name)_strict.png"))
+end
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
+```

### 2022-07-31 dev/paper/paper.tex
wuoohoo

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  conference]{IEEEtran}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Endogenous Dynamics in Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Endogenous Dynamics in Algorithmic Recourse}
\author{\author{
\IEEEauthorblockN{Patrick Altmeyer}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
p.altmeyer[at]tudelft.nl}
\and
\IEEEauthorblockN{Giovan Angela}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
g.j.a.angela[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Aleksander Buszydlik}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
a.j.buszydlik[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Karol Dobiczek}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
k.t.dobiczek[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Cynthia C. S. Liem}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
c.c.s.liem[at]tudelft.nl}
}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

Recent advances in Artificial Intelligence (AI) have propelled its
adoption in scientific domains outside of Computer Science including
Healthcare, Bioinformatics, Genetics and the Social Sciences. While this
has in many cases brought benefits in terms of efficiency,
state-of-the-art models like Deep Neural Networks (DNN) have also given
rise a new type of principal-agent problem in the context of data-driven
decision-making. It involves a group of \textbf{principals} - i.e.~human
stakeholders - that fail to understand the behaviour of their
\textbf{agent} - i.e.~the model used for automated decision-making
\protect\hyperlink{ref-borch2022machine}{{[}1{]}}.

Models or algorithms that fall into this category are typically referred
to \textbf{black-box} models. Despite their shortcomings, black-box
models have grown in popularity in recent years and have at times
created undesirable societal outcomes
\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
has tackled this issue from two different angles: while some have
appealed for a strict focus on inherently iterpretable models
\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
different ways to explain the behaviour of black-box models. These two
sub-domains can be broadly referred to as \textbf{interpretable AI} and
\textbf{explainable AI} (XAI), respectively.

Among the approaches to XAI that have recently grown in popularity are
\textbf{Counterfactual Explanations} (CE). They explain how inputs into
a model need to change for it to produce different outputs.
Counterfactual Explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals who face adverse outcomes. An example relevant
to the Social Sciences is consumer credit: in this context AR can be
used to guide individuals in improving their creditworthiness, should
they have previously been denied access to credit based on an automated
decision-making system. A meaningful recourse recommendation for a
denied applicant could be: \emph{``If your net savings rate had been
10\% of your monthly income instead of the actual 8\%, your application
would have been successful. See if you can temporarily cut down on
consumption.''} In the remainder of this paper we will use both
terminologies - recourse and counterfactual - interchangeably to refer
to situations where counterfactuals are generated with the intent to
provide individual recourse.

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself.

\begin{figure}

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/poc.png}

}

\caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
simple Bayesian model trained for binary classification (a); the
implementation of AR for a random subset of individuals leads to a
domain shift (b); as the classifier is retrained we observe a model
shift (c); as this process is repeated, the decision boundary moves away
from the target class (d).}

\end{figure}

We think that these types of endogenous dynamics may be problematic.
Firstly, model shifts may inadvertently change classification outcomes
for individuals who never received and implemented recourse. Secondly
and relatedly, we observe in Figure~\ref{fig-poc} that as the decision
boundary moves in the direction of the non-target class, counterfactual
paths become shorter: in the consumer credit example, individuals that
previously would have been denied credit based on their input features
are suddenly considered as creditworthy. Average default risk across all
borrowers can therefore be expected to increase. Conversely, lenders
that anticipate such dynamics may choose to deny credit to individuals
that have implemented AR, thereby compromising the validity of AR.

To the best of our knowledge this is the first work investigating
endogenous dynamics in AR. Our contributions to the state of knowledge
are two-fold. Firstly, we introduce an experimental framework extending
previous work by \protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
which allows for benchmarking different counterfactual generators in
terms of the endogenous domain and model shifts they introduce. To this
end we propose a number of novel evaluation metrics. Secondly, we use
this framework to provide the first in-depth analysis of endogenous
recourse dynamics induced by various popular counterfactual generators
including \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}.

We find that \ldots{}

The paper is structured as follows: Section~\ref{sec-related} places our
work in the broader context of related literature.
Section~\ref{sec-method} presents our methodology and data.
Section~\ref{sec-empirical} presents our empirical findings which are
then discussed in Section~\ref{sec-discussion}. We also point to some of
the limitations or our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

\hypertarget{sec-related}{%
\section{Related Work}\label{sec-related}}

In this Section we provide a review of the relevant literature. First,
Section~\ref{sec-related-recourse} discusses the existing research
within the domain of counterfactual explanations and algorithmic
recourse. Then, Section~\ref{sec-related-shifts} presents some of the
previous work on the measurement of dataset and model shifts.

\hypertarget{sec-related-recourse}{%
\subsection{Algorithmic Recourse}\label{sec-related-recourse}}

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
complexity or effort \(H=h(x\prime)\) associated with moving an
individual \(x\) to a counterfactual state \(x\prime\) such that the
predicted outcome \(M(x\prime)\) corresponds to some target outcome
\(t\):

\begin{equation}\protect\hypertarget{eq-obj}{}{
\min_{x\prime \in \mathcal{x}} c(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
}\label{eq-obj}\end{equation}

For implementation purposes, Equation~\ref{eq-obj} is typically
approximated through regularization:

\begin{equation}\protect\hypertarget{eq-solution}{}{
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),t) + \lambda h(x\prime)
}\label{eq-solution}\end{equation}

In the baseline work and many subsequent approaches the complexity
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that large perturbations
of \(x\) are costly.

Many approaches for the generation of algorithmic recourse have been
described in the literature. An October 2020 survey by Karimi et
al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}9{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}11{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

Efforts so far have largely been directed at improving the quality of
counterfactual explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are
interested in generating one or multiple meaningful counterfactual
explanations for some individual characterized by \(x_i\). The ability
of counterfactual explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose framework
for algorithmic recourse (ROAR) that evidently improves robustness to
such exogenous shifts.

\hypertarget{sec-related-shifts}{%
\subsection{Domain and Model Shifts}\label{sec-related-shifts}}

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}15{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}16{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}17{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}18{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}19{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}20{]}}.

In the context of algorithmic recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

\hypertarget{sec-related-benchmark}{%
\subsection{Benchmarking Counterfactual
Generators}\label{sec-related-benchmark}}

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}21{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different methodology;
finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}22{]}}
provides an extensible and language-agnostic implementation in Julia.
For the experiments conducted in this work we relied on the latter.

\hypertarget{sec-method}{%
\section{Methodology}\label{sec-method}}

In the following we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
to introduce the various counterfactual generators we have chosen to use
in our experiments. We then describe the experimental setup in
Section~\ref{sec-method-experiment} and introduce several evaluation
metrics used to benchmark the different generators.

\hypertarget{sec-method-general}{%
\subsection{A Generalized Framework for Gradient-Based Counterfactual
Search}\label{sec-method-general}}

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks. We can restate Equation~\ref{eq-solution} in a more
general form that encompasses most gradient-based approaches to
counterfactual search:

\begin{equation}\protect\hypertarget{eq-general}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
}\label{eq-general}\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In the case of the baseline
counterfactual generator
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
the idendity function and the number of counterfactuals \(K\) is equal
to one. This generator, which we shall refer to as \textbf{Wachter} in
the following, shall serve as the baseline against which all other
gradient-based methodologies will be compared. In particular, we include
include the following generator in our benchmarking exercises: REVISE
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}, DICE
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and a greedy
approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}.

Both \textbf{REVISE} and \textbf{CLUE} search counterfactuals in some
latent embedding \(S \subset \mathcal{S}\) instead of the feature space
directly. The latent embedding is learned by a separate generative model
that is tasked with learning the data generating process (DGP) of \(X\).
In this case \(f\) in Equation~\ref{eq-general} corresponds to the
decoder part of the generative model, in other words the deterministic
function that maps back from the latent embedding to the feature space.
Provided the generative model is well-specified, traversing the latent
embedding typically results in realistic and plausible counterfactuals,
because they are implicitly generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}. CLUE distinguishes
itself from REVISE and other counterfactual generators in that it aims
to minimize the predictive uncertainty of the model in question \(M\).
To quantify predictive uncertainty the authors rely on entropy estimates
for probabilistic models. The \textbf{Greedy} approach proposed by
\protect\hyperlink{ref-schut2021generating}{{[}12{]}} also works with
the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that
can produce predictive uncertainty estimates. The authors show that in
this setting the complexity penalty \(h(\cdot)\) in
Equation~\ref{eq-general} is redundant and meaningful counterfactuals
can be generated in a fast and efficient manner through a modified
Jacobian-based Saliency Map Attack (JSMA). Finally, \textbf{DICE}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. To this
end the authors use a complexity penalty \(h(\mathbf{s}^\prime)\) that
favours diverse outcomes, in the sense that \(s_1, ... , s_K\) look as
different from each other as possible.

Our motivation for including these different generators in our analysis,
is that they all offer slightly different approaches to generate
meaningful counterfactuals for differentiable black-box models. We
hypothesize that generating more \textbf{meaningful} counterfactuals
should mitigate the endogenous dynamics illustrated in
Figure~\ref{fig-poc} in Section~\ref{sec-intro}. This intuition stems
from the underlying idea that more meaningful counterfactuals are
generated by the same or at least a very similar data generating process
as the training data. All else equal, counterfactuals that fulfill this
basic requirement should be less prone to trigger domain and model
shifts.

\hypertarget{sec-method-experiment}{%
\subsection{Experimental Setup}\label{sec-method-experiment}}

The dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro} were generated through a simple experiment that
aims to simulate the process of algorithmic recourse in practice. We
begin in the static setting at time time \(t=0\): given some pre-trained
classifier \(M\) we generate recourse for a random batch of \(B\)
individuals in the non-target class. Note that we focus our attention on
classification problems, since classification poses the most common
practical use-case for algorithmic recourse. In order to simulate the
dynamical process we suppose that the model \(M\) is retrained following
the actual implementation of recourse in time \(t=0\). Following the
update to the model, we assume that at time \(t=1\) recourse is
generated for yet another random subset of individuals in the non-target
class. This process is repeated for a number of time periods \(T\). To
get a clean read on endogenous dynamics we keep the total population of
samples closed: we allow existing samples to move from factual to
counterfactual states, but do not allow any entirely new samples to
enter the population. The experimental setup is summarized in Algorithm
\ref{algo-experiment}

\begin{algorithm}
\caption{Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,D,G$}
\State $t\gets 0$
\While{$t<T$}
\State $D_B \subset D$
\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(D)$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,D$
\EndProcedure
\end{algorithmic}
\end{algorithm}

A noteworthy practical consideration is the choice of \(T\) and \(B\).
The higher these values, the more factual instances undergo recourse
throughout the entire experiment. Of course, this is likely to lead to
more pronounced domain and model shifts by time \(T\). At the same time,
it is generally improbable that a very large part of the population
would request an explanation of the algorithm's decisions. In our
experiments, we choose the values such that \(T \cdot B\) corresponds to
the application of recourse on \(25-50\%\) of the negative instances
from the initial dataset. As we collect data at each time \(t\), we can
also verify the impact of recourse when it is implemented for a smaller
number of individuals. Using our framework the experiment can be
conducted on an arbitrary number of algorithmic recourse generators. As
all generators make use of the same initial model and initial dataset,
the differences in domain and model shifts observed throughout the
rounds depend solely on the employed generator.

\hypertarget{sec-method-data}{%
\subsection{Data}\label{sec-method-data}}

We have chosen to work with both synthetic and real-world datasets.
Using synthetic data allows us to impose distributional properties that
may affect the resulting recourse dynamics. Following
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} we generate
synthetic data in \(\mathbb{R}^2\) to also allow for a visual
interpretation of the results. Real-world data is used in order to
assess if endogenous dynamics also occur in higher-dimensional settings.

\hypertarget{synthetic-data}{%
\subsubsection{Synthetic data}\label{synthetic-data}}

We use 6 synthetic binary classification datasets consisting of 200-400
samples grouped in normally-distributed clusters.\footnote{To see how
  the data is generated see here:
  \url{https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb}}
The datasets are presented in Figure~\ref{fig-synthetic-data} (see also
Appendix A for a formal description). Samples from the negative class
are marked in blue while samples of the positive class are marked in
orange.

\begin{figure}

{\centering \includegraphics{www/synthetic_data.png}

}

\caption{\label{fig-synthetic-data}PLACEHOLDER: A visualization of the
synthetic classification datasets used in our experiments.}

\end{figure}

Ex-ante we expect to see that Wachter will create a new cluster of
counterfactual instances in the proximity of the initial decision
boundary. Thus, the choice of a black-box model may have an impact on
the paths of the recourse. For generators that use latent space search
(\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}) or rely on (and
have access to) probabilistic models
(\protect\hyperlink{ref-antoran2020getting}{{[}8{]}},
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}) we expect that
counterfactuals will end up in regions of the target domain that are
densely populated by training samples. Finally, we expect to see the
counterfactuals generated by DiCE to be uniformly spread around the
feature space inside the target class.

\hypertarget{real-world-data}{%
\subsubsection{Real-world data}\label{real-world-data}}

Additionally, we use two real-world datasets from the Finance domain.
Firstly, we use the Give Me Some Credit dataset which was open-sourced
on Kaggle for the task to predict whether a borrower is likely to
experience financial difficulties in the next two years
\protect\hyperlink{ref-gmsc_data}{{[}23{]}}. Originally consisting of
250,000 instances with 11 numerical attributes, the dataset was randomly
undersampled to result in a balanced subsample made up of 3000
individuals. Secondly, we the German Credit dataset which involves the
task of predicting if bank customers are credit-worthy or not
\protect\hyperlink{ref-germancredit1994}{{[}24{]}}. It consists of 700
positive and 300 negative instances charaterized by 7 numerical and 13
categorical attributes. We process the dataset in two ways: (1) the
values of the ``Personal status and sex'' feature are aggregated by the
two represented genders; (2) the most common values are calculated for
all categorical features such that a feature \(x_d\) with the mode
\(\bar{x}_d\) is transformed into a new binary feature
\(\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}\). Binarization ensures
that we can use all counterfactual generators in the benchmark.

\hypertarget{sec-method-metrics}{%
\subsection{Evaluation Metrics}\label{sec-method-metrics}}

We formulate two desiderata for the set of metrics used to measure
domain and model shifts induced by recourse. First, the metrics should
be applicable regardless of the dataset or classification technique so
that they allow for the meaningful comparison of the generators in
various scenarios. As the knowledge of the underlying probability
distribution is rarely available, the metrics should be empirical and
non-parametric. This further ensures that we can also measure large
datasets by sampling from the available data. Moreover, while our study
was conducted in a two-class classification setting, our choice of
metrics should remain applicable in the future research on multi- class
recourse problems. Second, the set of metrics should allow to capture
various aspects of the previously mentioned magnitude, path, and tempo
of changes while remaining as small as possible.

\hypertarget{domain-shifts}{%
\subsubsection{Domain Shifts}\label{domain-shifts}}

To quantify the magnitude of domain shifts we rely on an unbiased
estimate of the squared population \textbf{Maximum Mean Discrepancy
(MMD)} given as:

\begin{equation}\protect\hypertarget{eq-mmd}{}{
\begin{aligned}
MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
\end{aligned}
}\label{eq-mmd}\end{equation}

where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
identically distributed samples drawn from probability distributions
\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
between the kernel mean embeddings of \(p\) and \(q\) in RKHS
\(\mathcal{H}\). An important consideration is the choice of the kernel
function \(k(\cdot,\cdot)\). In our implementation we make use of the
radial basis function (RBF) kernel with a constant length-scale
parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
\(X=\tilde{X}\).

The evaluation metric in Equation~\ref{eq-mmd} is computed after every
round \(t=1,...,T\) of the experiment. To assess the statistical
significance of the observed shifts under the null hypothesis that
samples \(X\) and \(\tilde{X}\) were drawn from the same probability
distribution we follow
\protect\hyperlink{ref-arcones1992bootstrap}{{[}25{]}}. To that end, we
combine the two samples and generate a large number of permutations of
\(X + \tilde{X}\). Then, we split the permuted data into two new samples
\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
original samples. Then under the null hypothesis we should have that
\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
calculated by counting how these two quantities are not equal.

We calculate the MMD for both classes individually based on the ground
truth labels, i.e.~the labels that samples were assigned in time
\(t=0\). Throughout our experiments we generally do not expect the
distribution of the negative class to change over time -- application of
recourse reduces the size of this class, but since individuals are
sampled uniformly the distribution should remain unaffected. Conversely,
unless a recourse generator can perfectly replicate the original
probability distribution, we expect the MMD of the positive class to
increase. Thus, when discussing MMD, we generally mean the shift in the
distribution of the positive class.

The \textbf{Operating Point} or \textbf{elbow} of the \(k\)-means
algorithm is used to measure whether the counterfactual instances tend
to lie close to the positively labelled training samples or make up a
new, separate cluster of data points within the positive class. In the
latter case the operating point of the \(k\)-means algorithm can be
expected to change. There exist several algorithms for the automated
discovery of elbow points and we resort to the Kneedle method
\protect\hyperlink{ref-satopaa2011finding}{{[}26{]}}. Kneedle selects
the optimal value of \(k\) by rotating the inertia curve onto the
horizontal axis and selecting a local minimum. While working with the
synthetic datasets we also use domain knowledge to confirm the results
of Kneedle through the visual inspection of data.

Finally, \textbf{feature mean and feature standard deviation} are also
calculated to verify how the implementation of recourse impacts every
attribute in the dataset. Although MMD already captures information
about the expected value and variance, we may also be interested in a
more granular look at individual features.

\hypertarget{model-shifts}{%
\subsubsection{Model Shifts}\label{model-shifts}}

As our baseline for quantifying model shifts we measure perturbations to
the model parameters at each point in time \(t\) following
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We define
\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
between the vectors of parameters before and after retraining the model
\(M\). We shall refer to this baseline metric simply as
\textbf{Perturbations}.

We extend the metric in Equation~\ref{eq-mmd} for the purpose of
quantifying model shifts. Specifically, we introduce \textbf{Predicted
Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
features directly, we apply it to the predicted probabilities assigned
to a set of samples by the model \(M\). If the model shifts, the
probabilities assigned to each sample will change; again, this metric
will equal 0 only if the two classifiers are the same. It is worth
noting that while we apply the technique to samples drawn uniformly from
the dataset, it can also be employed on arbitrary points in the entire
feature space (or a subspace). The latter approach is theoretically more
robust. Unfortunately, in practice this approachs suffers from the curse
of dimensionality, since it becomes increasingly difficult to select
enough points to overcome noise as the dimension \(D\) grows.

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
the likelihood that a model assigns a high probability to its
classification of any given sample. We define the metric simply as
\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
predicted logits from a binary classifier and \(\sigma\) denotes the
sigmoid function. This metric provides an unbiased estimate of the
binary classifier's tendency to produce high-confidence predictions in
either one of the two classes. Although the exact values for this metric
are not important for our study, they can be used to detect model
shifts. If decisiveness changes over time, then this is indicative of
the decision boundary moves towards either one of the two classes.

\hypertarget{sec-empirical}{%
\section{Experiments}\label{sec-empirical}}

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

\hypertarget{sec-limit}{%
\subsection{Limitations and Future Work}\label{sec-limit}}

The experimental setup proposed here is designed to mimic a real-world
recourse process in a simple fashion. In practice, models are in fact
updated on a regular basis
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also find it
plausible to assume that the implementation of recourse happens
periodically for different individuals, rather that all at once at time
\(t=0\). That being said, our experimental design is a vast
over-simplification of potential real-world scenarios. In practice, any
endogenous shifts that may occur can be expected to be entangled with
exogenous shifts of the nature investigated in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also make
implicit assumptions about the utility functions of the involved agents
that may well be too simple: individuals seeking recourse are assumed to
always implement the proposed counterfactual explanations; conversely,
the agent in charge of the model \(M\) is assumed to always treat
individuals that have implemented valid recourse as if they were truly
now in the target class. Relating this back to the consumer credit
example, we assume that the would-be borrowers are always willing and
able to implement recourse and the bank is always willing to provide
credit as would-be borrowers move across the decision boundary. In
practice it is doubtful that agents behave according to such simple
rules. Nonetheless, we think that our simple framework offers a starting
point for future work on recourse dynamics (both endogenous and
exogenous dynamics).

\hypertarget{sec-conclusion}{%
\section{Concluding Remarks}\label{sec-conclusion}}

\hypertarget{acknowledgment}{%
\section*{Acknowledgment}\label{acknowledgment}}
\addcontentsline{toc}{section}{Acknowledgment}

P. A. thanks \ldots{}

\pagebreak
\FloatBarrier

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
\CSLLeftMargin{{[}1{]} }%
\CSLRightInline{C. Borch, {``Machine learning, knowledge risk, and
principal-agent problems in automated trading,''} \emph{Technology in
Society}, p. 101852, 2022.}

\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
\CSLLeftMargin{{[}2{]} }%
\CSLRightInline{C. O'neil, \emph{Weapons of math destruction: How big
data increases inequality and threatens democracy}. Crown, 2016.}

\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
\CSLLeftMargin{{[}3{]} }%
\CSLRightInline{C. Rudin, {``Stop explaining black box machine learning
models for high stakes decisions and use interpretable models
instead,''} \emph{Nature Machine Intelligence}, vol. 1, no. 5, pp.
206--215, 2019.}

\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
\CSLLeftMargin{{[}4{]} }%
\CSLRightInline{M. Pawelczyk, S. Bielawski, J. van den Heuvel, T.
Richter, and G. Kasneci, {``Carla: A python library to benchmark
algorithmic recourse and counterfactual explanation algorithms,''}
\emph{arXiv preprint arXiv:2108.00783}, 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
\CSLLeftMargin{{[}5{]} }%
\CSLRightInline{S. Wachter, B. Mittelstadt, and C. Russell,
{``Counterfactual explanations without opening the black box: Automated
decisions and the GDPR,''} \emph{Harv. JL \& Tech.}, vol. 31, p. 841,
2017.}

\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
\CSLLeftMargin{{[}6{]} }%
\CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
Ghosh, {``Towards realistic individual recourse and actionable
explanations in black-box decision making systems,''} \emph{arXiv
preprint arXiv:1907.09615}, 2019.}

\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
\CSLLeftMargin{{[}7{]} }%
\CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
machine learning classifiers through diverse counterfactual
explanations,''} in \emph{Proceedings of the 2020 conference on
fairness, accountability, and transparency}, 2020, pp. 607--617.}

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
\CSLLeftMargin{{[}8{]} }%
\CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty
estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
\CSLLeftMargin{{[}9{]} }%
\CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
{``A survey of algorithmic recourse: Definitions, formulations,
solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050},
2020.}

\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
\CSLLeftMargin{{[}10{]} }%
\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
explanations for machine learning: A review,''} \emph{arXiv preprint
arXiv:2010.10596}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
\CSLLeftMargin{{[}11{]} }%
\CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
recourse in linear classification,''} in \emph{Proceedings of the
conference on fairness, accountability, and transparency}, 2019, pp.
10--19.}

\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
\CSLLeftMargin{{[}12{]} }%
\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
counterfactual explanations by implicit minimisation of epistemic and
aleatoric uncertainties,''} in \emph{International conference on
artificial intelligence and statistics}, 2021, pp. 1756--1764.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
\CSLLeftMargin{{[}13{]} }%
\CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
{``Algorithmic recourse: From counterfactual explanations to
interventions,''} in \emph{Proceedings of the 2021 ACM conference on
fairness, accountability, and transparency}, 2021, pp. 353--362.}

\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
\CSLLeftMargin{{[}14{]} }%
\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
robust and reliable algorithmic recourse,''} \emph{arXiv preprint
arXiv:2102.13620}, 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
\CSLLeftMargin{{[}15{]} }%
\CSLRightInline{S. Rabanser, S. Günnemann, and Z. Lipton, {``Failing
loudly: An empirical study of methods for detecting dataset shift,''}
\emph{Advances in Neural Information Processing Systems}, vol. 32,
2019.}

\leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
\CSLLeftMargin{{[}16{]} }%
\CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly
detection: A survey,''} \emph{ACM computing surveys (CSUR)}, vol. 41,
no. 3, pp. 1--58, 2009.}

\leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
\CSLLeftMargin{{[}17{]} }%
\CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of
concept drift and hidden contexts,''} \emph{Machine learning}, vol. 23,
no. 1, pp. 69--101, 1996.}

\leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
\CSLLeftMargin{{[}18{]} }%
\CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A.
Bouchachia, {``A survey on concept drift adaptation,''} \emph{ACM
computing surveys (CSUR)}, vol. 46, no. 4, pp. 1--37, 2014.}

\leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
\CSLLeftMargin{{[}19{]} }%
\CSLRightInline{K. Nelson, G. Corbin, M. Anania, M. Kovacs, J. Tobias,
and M. Blowers, {``Evaluating model drift in machine learning
algorithms,''} in \emph{2015 IEEE symposium on computational
intelligence for security and defense applications (CISDA)}, 2015, pp.
1--8.}

\leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
\CSLLeftMargin{{[}20{]} }%
\CSLRightInline{S. Ackerman, P. Dube, E. Farchi, O. Raz, and M.
Zalmanovici, {``Machine learning model drift detection via weak data
slices,''} in \emph{2021 IEEE/ACM third international workshop on deep
learning for testing and testing for deep learning (DeepTest)}, 2021,
pp. 1--8.}

\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
\CSLLeftMargin{{[}21{]} }%
\CSLRightInline{R. M. B. de Oliveira and D. Martens, {``A framework and
benchmarking study for counterfactual generating methods on tabular
data,''} \emph{Applied Sciences}, vol. 11, no. 16, p. 7274, 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
\CSLLeftMargin{{[}22{]} }%
\CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
Julia package for Counterfactual Explanations and Algorithmic
Recourse}}. 2022. Available:
\url{https://github.com/pat-alt/CounterfactualExplanations.jl}}

\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
\CSLLeftMargin{{[}23{]} }%
\CSLRightInline{K. Competition, {``Give me some credit, improve on the
state of the art in credit scoring by predicting the probability that
somebody will experience financial distress in the next two years.''}
\url{https://www.kaggle.com/c/GiveMeSomeCredit}}

\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
\CSLLeftMargin{{[}24{]} }%
\CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}}

\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
\CSLLeftMargin{{[}25{]} }%
\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}

\leavevmode\vadjust pre{\hypertarget{ref-satopaa2011finding}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{V. Satopaa, J. Albrecht, D. Irwin, and B. Raghavan,
{``Finding a" kneedle" in a haystack: Detecting knee points in system
behavior,''} in \emph{2011 31st international conference on distributed
computing systems workshops}, 2011, pp. 166--171.}

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}

\end{CSLReferences}

\pagebreak

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\ldots{}

\pagebreak

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\ldots{}

\pagebreak

\hypertarget{code}{%
\section{Code}\label{code}}

\ldots{}

\end{document}

### 2022-07-31 dev/utils.jl

@@ -26,3 +26,16 @@ function www_dir(dir="")
     return www_dir
 end
 
+"""
+    data_dir(dir="")
+
+Sets up the directory to save images and returns the path.
+"""
+function data_dir(dir="")
+    root_ = "dev/artifacts/upload/data"
+    data_dir = joinpath(root_, dir)
+    if !isdir(data_dir)
+        mkpath(data_dir)
+    end
+    return data_dir
+end

### 2022-07-31 src/base.jl

@@ -91,7 +91,7 @@ function run!(
                 end
                 # Evaluate:
                 if n % evaluate_every == 0 
-                    output_checkpoint = collect_output(experiment, recourse_system, chosen_individuals_m, k, n, m, n_bootstrap=1000)
+                    output_checkpoint = collect_output(experiment, recourse_system, chosen_individuals[m], k, n, m, n_bootstrap=n_bootstrap)
                     output[m] = vcat(output[m], output_checkpoint, cols=:union)
                 end
             end
@@ -245,12 +245,15 @@ function run_experiment(
     save_name::Union{Nothing,String}=nothing,
     kwargs...
 )
-    @info "Starting experiment"
+
+    exp_name = isnothing(save_name) ? "unnamed" : save_name
+
+    @info "Starting experiment: $exp_name"
 
     # Run:
     output = run!(experiment; evaluate_every=evaluate_every, kwargs...)
 
-    @info "Completed experiment."
+    @info "Completed experiment: $exp_name"
 
     results = ExperimentResults(output,experiment)
 
@@ -263,6 +266,9 @@ function run_experiment(
         Serialization.serialize(joinpath(save_path,"output.jls"), output)
         Serialization.serialize(joinpath(save_path,"experiment.jls"), experiment)
         Serialization.serialize(joinpath(save_path,"results.jls"), results)
+
+        @info "Saved experiment: $exp_name"
+
     end
 
     return results
@@ -302,12 +308,14 @@ function run_experiment(
         target=target,num_counterfactuals=num_counterfactuals,pre_train_models=pre_train_models
     )
 
-    @info "Starting experiment"
+    exp_name = isnothing(save_name) ? "unnamed" : save_name
+
+    @info "Starting experiment: $exp_name"
 
     # Run:
     output = run!(experiment; evaluate_every=evaluate_every, kwargs...)
 
-    @info "Completed experiment."
+    @info "Completed experiment: $exp_name"
 
     results = ExperimentResults(output,experiment)
 
@@ -320,6 +328,9 @@ function run_experiment(
         Serialization.serialize(joinpath(save_path,"output.jls"), output)
         Serialization.serialize(joinpath(save_path,"experiment.jls"), experiment)
         Serialization.serialize(joinpath(save_path,"results.jls"), results)
+
+        @info "Saved experiment: $exp_name"
+
     end
 
     return results
[…]

### 2022-07-31 src/evaluation/Evaluation.jl

@@ -26,7 +26,7 @@ function evaluate_system(recourse_system::RecourseSystem, experiment::Experiment
         mmd_model(experiment, recourse_system; n=n, grid_search=true),
         disagreement(experiment, recourse_system),
         decisiveness(experiment, recourse_system),
-        fscore(experiment, recourse_system)
+        model_performance(experiment, recourse_system)
     ]
 
     if to_dataframe

### 2022-07-31 src/evaluation/model_shifts.jl

@@ -1,4 +1,4 @@
-struct ModelMetric <: AbstractMetric 
+struct ModelMetric <: AbstractMetric
     metric::Number
     p_value::Union{Missing,Number}
     name::Symbol
@@ -14,7 +14,7 @@ Turns an instance of class `ModelMetric` into `DataFrame`.
 function DataFrame(metric::ModelMetric)
     vals = (metric.metric, metric.p_value, metric.name, :model)
     df = DataFrame(NamedTuple{(:value, :p_value, :name, :scope)}.([vals]))
-    return df  
+    return df
 end
 
 using LinearAlgebra
@@ -27,7 +27,7 @@ function perturbation(experiment::Experiment, recourse_system::RecourseSystem)
 
     value = perturbation(M, new_M)
 
-    metric = ModelMetric(value,missing,:perturbation)
+    metric = ModelMetric(value, missing, :perturbation)
 
     return metric
 
@@ -44,13 +44,13 @@ using StatsBase
 Calculates the MMD on the probabilities of classification assigned by the model to the set of (all) instances. Allows to quantify the model shift.
 """
 function mmd_model(experiment::Experiment, recourse_system::RecourseSystem; n=1000, grid_search=false, n_samples=1000, kwargs...)
-    
+
     X, _ = unpack(experiment.data)
 
     if grid_search
-        X = reduce(hcat,[map(x -> rand(range(x..., length=100)), extrema(X, dims=2)) for i in 1:n_samples])
+        X = reduce(hcat, [map(x -> rand(range(x..., length=100)), extrema(X, dims=2)) for i in 1:n_samples])
     end
-    
+
     # Initial:
     M = recourse_system.initial_model
     proba = probs(M, X)
@@ -62,7 +62,7 @@ function mmd_model(experiment::Experiment, recourse_system::RecourseSystem; n=10
     value, p_value = mmd(proba, new_proba, n_samples; compute_p=n, kwargs...)
     metric_name = grid_search ? :mmd_grid : :mmd
 
-    metric = ModelMetric(value,p_value,metric_name)
+    metric = ModelMetric(value, p_value, metric_name)
 
     return metric
 
@@ -75,20 +75,20 @@ using LinearAlgebra
 Calculates the pseudo-distance of points to the decision boundary measured as the average probability of classification centered around 0.5. High value corresponds to a large margin of classification.
 """
[…]

### 2022-07-31 src/experiments/functions.jl

@@ -9,6 +9,7 @@ using Parameters
     μ::AbstractFloat = 0.05
     γ::AbstractFloat = 0.75
     intersect_::Bool = true
+    convergence::Symbol = :threshold_only
 end
 
 mutable struct Experiment

### 2022-07-31 src/models/laplace.jl

@@ -67,6 +67,6 @@ using LinearAlgebra, Flux, Statistics
 function perturbation(model::LaplaceReduxModel, new_model::LaplaceReduxModel; agg=mean)
     mlp = model.model
     new_mlp = new_model.model
-    Δ = agg(norm.(collect(Flux.params(new_mlp)) .- collect(Flux.params(mlp))))
+    Δ = mean(map(x -> norm(x)/length(x),Flux.params(new_mlp).-Flux.params(mlp)))
     return Δ
 end

### 2022-08-08 dev/notebooks/data_preprocessing.qmd/real_world_data.qmd
2/3 real world data done

---
title: Preprocessing Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("real_world")
www_path = www_dir("real_world")
data_path = data_dir("real_world")
```

## California Housing Data

Fetching the data using Python's `sklearn`:

```{python}
from sklearn.datasets import fetch_california_housing
df, y = fetch_california_housing(return_X_y=True, as_frame=True)
df["target"] = y.values
data_path = "../../artifacts/upload/data/real_world"
import os
df.to_csv(os.path.join(data_path,"raw/cal_housing.csv"), index=False)
```

Loading the data into Julia session:

```{julia}
using CSV, DataFrames, Statistics, StatsBase
df = CSV.read(joinpath(data_path, "raw/cal_housing.csv"), DataFrame)
# Features:
X = Matrix(df[:,Not(:target)])
dt = fit(ZScoreTransform, X, dims=1)
StatsBase.transform!(dt, X)
# Target:
y = df.target
y = Float64.(y .>= median(y)); # binary target
# Data:
df = DataFrame(X,:auto)
df.target = y
```

```{julia}
using MLUtils: undersample
# Make DataFrames.jl work
MLUtils.getobs(data::DataFrame, i) = data[i,:]
MLUtils.numobs(data::DataFrame) = nrow(data)
df_balanced = getobs(undersample(df, df.target;shuffle=true))
```

```{julia}
CSV.write(joinpath(data_path, "cal_housing.csv"), df)
```

## Give Me Some Credit

```{julia}
using CSV, DataFrames, Statistics, StatsBase
df = CSV.read(joinpath(data_path, "raw/cs-training.csv"), DataFrame)
select!(df, Not([:Column1]))
rename!(df, :SeriousDlqin2yrs => :target)
mapcols!(x -> [ifelse(x_=="NA", missing, x_) for x_ in x], df)
dropmissing!(df)
mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
# Features:
X = Matrix(df[:,Not(:target)])
dt = fit(ZScoreTransform, X, dims=1)
StatsBase.transform!(dt, X)
# Target:
y = df.target
# Data:
df = DataFrame(X,:auto)
df.target = y
```

```{julia}
using MLUtils
using MLUtils: undersample
# Make DataFrames.jl work
MLUtils.getobs(data::DataFrame, i) = data[i,:]
MLUtils.numobs(data::DataFrame) = nrow(data)
df_balanced = getobs(undersample(df, df.target;shuffle=true))
```

```{julia}
CSV.write(joinpath(data_path, "gmsc.csv"), df_balanced)
```

## UCI Credit Card Default

```{julia}
using CSV, DataFrames, Statistics, StatsBase
df = CSV.read(joinpath(data_path, "raw/UCI_Credit_Card.csv"), DataFrame)
select!(df, Not([:ID, :SEX, :EDUCATION, :MARRIAGE]))
rename!(df, "default.payment.next.month" => :target)
dropmissing!(df)
mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
# Features:
X = Matrix(df[:,Not(:target)])
dt = fit(ZScoreTransform, X, dims=1)
StatsBase.transform!(dt, X)
# Target:
y = df.target
# Data:
df = DataFrame(X,:auto)
df.target = y
```

```{julia}
using MLUtils
using MLUtils: undersample
# Make DataFrames.jl work
MLUtils.getobs(data::DataFrame, i) = data[i,:]
MLUtils.numobs(data::DataFrame) = nrow(data)
df_balanced = getobs(undersample(df, df.target;shuffle=true))
```

```{julia}
CSV.write(joinpath(data_path, "credit_default.csv"), df_balanced)
```

### 2022-08-08 dev/notebooks/gravitational_generator.qmd
2/3 real world data done

@@
 target = ifelse(y==1.0,0.0,1.0) # opposite label as target
+T = 100
 ```
@@
       counterfactuals...,
-      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict)
+      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
     )
@@
 theme(:wong)
-T = 500
+T_ = 500
 plts = []
@@
     counterfactual = counterfactuals[i]  
-    plt = plot(counterfactual, plot_up_to=T, title="λ₂=$(λ₂)")
+    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
     x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
@@
 
+```{julia}
+
+```
+
 # References

### 2022-08-08 dev/notebooks/real_world.qmd
2/3 real world data done

@@
 ```{julia}
-catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
-choices = [:give_me_some_credit_balanced, :housing]
-data_sets = filter(p -> p[1] in choices, catalogue)
+max_obs = 10000
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
 ```
@@
 end
-model_params = (batch_norm=false,n_hidden=128,dropout=true,p_dropout=0.1)
+model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
 ```
@@
 ```{julia}
-models = [:FluxModel, :FluxEnsemble]
+models = [
+    :LogisticRegression, 
+    :FluxModel, 
+    :FluxEnsemble
+]
 generators = Dict(
@@
     :DICE=>DiCEGenerator(),
-    :Gravitational=>GravitationalGenerator(),
+    :Gravitational=>GravitationalGenerator(λ=[0.1,2]),
 )
@@
 ```{julia}
-experiments = set_up_experiments(data_sets,models,generators; pre_train_models=100, model_params=model_params, data_loader=data_loader)
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
 ```
@@
 
+## Strict convergence
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+T = 500
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
+    convergence=:strict, save_name_suffix="strict"
+)
+Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+```
+
 ## Simple convergence
@@
 ```{julia}
-n_evals = 1
-n_rounds = 10
+n_evals = 5
+n_rounds = 50
 evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 2
+n_folds = 5
 n_bootstrap = 1
@@
 
-## Strict convergence
+### Plots
 
 ```{julia}
-n_evals = 1
-n_rounds = 10
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 2
-n_bootstrap = 1
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
-    convergence=:strict, save_name_suffix="strict"
-)
-Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+# plot_dict = Dict()
+# for (data_name, res) in results
+#     output_data = res.output
+#     metrics = unique(output_data.name)
+#     plots = Dict(metric => plot(res, metric) for metric in metrics)
+#     plot_dict[data_name] = plots
+# end
 ```

### 2022-08-08 dev/notebooks/synthetic.qmd
2/3 real world data done

@@
 ```{julia}
-n_evals = 1
-n_rounds = 1
+n_evals = 10
+n_rounds = 50
 evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 1
+n_folds = 5
 n_bootstrap = 1
@@
 
+```{julia}
+plot_dict = Dict()
+```
+
 ## Strict convergence
@@
 ```
+
+```{julia}
+plot_dict = Dict()
+for (data_name, res) in results
+    output_data = res.output
+    metrics = unique(output_data.name)
+    plots = Dict(metric => plot(res, metric) for metric in metrics)
+    plot_dict[data_name] = plots
+end
+```

### 2022-08-08 dev/paper/paper.tex
2/3 real world data done

@@
 
+\usepackage{cite}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{graphicx}
+\usepackage{textcomp}
+\usepackage{xcolor}
+\usepackage{placeins}
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
+    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Endogenous Dynamics in Algorithmic Recourse},
+  pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
   colorlinks=true,
@@
 
-\title{Endogenous Dynamics in Algorithmic Recourse}
-\author{\author{
-\IEEEauthorblockN{Patrick Altmeyer}
-\IEEEauthorblockA{\textit{EEMCS} \\
-\textit{Delft University of Technology}\\
-Delft, Netherlands \\
-p.altmeyer[at]tudelft.nl}
-\and
-\IEEEauthorblockN{Giovan Angela}
-\IEEEauthorblockA{\textit{EEMCS} \\
-\textit{Delft University of Technology}\\
-Delft, Netherlands \\
-g.j.a.angela[at]student.tudelft.nl}
-\and
-\IEEEauthorblockN{Aleksander Buszydlik}
-\IEEEauthorblockA{\textit{EEMCS} \\
-\textit{Delft University of Technology}\\
-Delft, Netherlands \\
-a.j.buszydlik[at]student.tudelft.nl}
-\and
-\IEEEauthorblockN{Karol Dobiczek}
-\IEEEauthorblockA{\textit{EEMCS} \\
-\textit{Delft University of Technology}\\
-Delft, Netherlands \\
-k.t.dobiczek[at]student.tudelft.nl}
-\and
-\IEEEauthorblockN{Cynthia C. S. Liem}
-\IEEEauthorblockA{\textit{EEMCS} \\
-\textit{Delft University of Technology}\\
-Delft, Netherlands \\
-c.c.s.liem[at]tudelft.nl}
-}}
+\title{Endogenous Macrodynamics in Algorithmic Recourse}
+\author{}
 \date{}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-The \textbf{Operating Point} or \textbf{elbow} of the \(k\)-means
-algorithm is used to measure whether the counterfactual instances tend
-to lie close to the positively labelled training samples or make up a
-new, separate cluster of data points within the positive class. In the
-latter case the operating point of the \(k\)-means algorithm can be
-expected to change. There exist several algorithms for the automated
-discovery of elbow points and we resort to the Kneedle method
-\protect\hyperlink{ref-satopaa2011finding}{{[}26{]}}. Kneedle selects
-the optimal value of \(k\) by rotating the inertia curve onto the
-horizontal axis and selecting a local minimum. While working with the
-synthetic datasets we also use domain knowledge to confirm the results
-of Kneedle through the visual inspection of data.
-
 Finally, \textbf{feature mean and feature standard deviation} are also
@@
 \textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}26{]}} and
 estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
@@
 \section{Experiments}\label{sec-empirical}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Shift of focus from individual to group of individuals (related:
+  https://www.researchgate.net/publication/353073138\_Generating\_Collective\_Counterfactual\_Explanations\_in\_Score-Based\_Classification\_via\_Mathematical\_Optimization)
+\item
+  Convergence criterium matters: terminating once threshold probability
+  is reached may not be optimal (see e.g.~REVISE)
+\item
+  Optimizer choice matters: dimensionality is typically low, so no
+  obvious benefit to using ADAM.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    This might be better placed in JuliaCon proceedings, perhaps backed
+    by small blog post on the matter.
+  \end{itemize}
+\item
+  Mitigating strategy: penaliye distance from centroid.
+\end{enumerate}
 
@@
 
+\hypertarget{data}{%
+\subsubsection{Data}\label{data}}
+
+The number of real-world datasets considered in this work is limited,
+constrained by the fact that at the time of writing
+\texttt{CounterfactualExplanations.jl} only supported continuous
+features.
+
 \hypertarget{sec-conclusion}{%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-satopaa2011finding}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
 \CSLLeftMargin{{[}26{]} }%
-\CSLRightInline{V. Satopaa, J. Albrecht, D. Irwin, and B. Raghavan,
-{``Finding a" kneedle" in a haystack: Detecting knee points in system
-behavior,''} in \emph{2011 31st international conference on distributed
-computing systems workshops}, 2011, pp. 166--171.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}27{]} }%
 \CSLRightInline{S. Hanneke, {``A bound on the label complexity of

### 2022-08-08 dev/paper/sections/empirical.qmd
2/3 real world data done

# Experiments {#sec-empirical}

1. Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
4. Mitigating strategy: penaliye distance from centroid.

### 2022-08-08 dev/paper/sections/methodology.qmd
2/3 real world data done

--- 2022-07-12 paper/sections/methodology.qmd
@@
 
-The **Operating Point** or **elbow** of the $k$-means algorithm is used to measure whether the counterfactual instances tend to lie close to the positively labelled training samples or make up a new, separate cluster of data points within the positive class.
-In the latter case the operating point of the $k$-means algorithm can be expected to change.
-There exist several algorithms for the automated discovery of elbow points and we resort to the Kneedle method [@satopaa2011finding].
-Kneedle selects the optimal value of $k$ by rotating the inertia curve onto the horizontal axis and selecting a local minimum.
-While working with the synthetic datasets we also use domain knowledge to confirm the results of Kneedle through the visual inspection of data.
-
 Finally, **feature mean and feature standard deviation** are also calculated to verify how the implementation of recourse impacts every attribute in the dataset.

### 2022-08-08 src/data/utils.jl

@@ -1,0 +1,40 @@
+# Train-test split
+using StatsBase
+using CounterfactualExplanations
+using CounterfactualExplanations.DataPreprocessing: unpack
+using Flux
+"""
+    train_test_split(data::CounterfactualData;test_size=0.2)
+
+Splits data into train and test split.
+"""
+function train_test_split(data::CounterfactualData;test_size=0.2)
+    X,y = unpack(data)
+    N = size(y,2)
+    classes_ = sort(unique(y))
+    n_per_class = round(N/length(classes_))
+    test_idx = sort(reduce(vcat,[sample(findall(vec(y.==cls)), Int(floor(test_size * n_per_class)),replace=false) for cls in classes_]))
+    train_idx = setdiff(1:N, test_idx)
+    train_data = CounterfactualData(X[:,train_idx], y[:,train_idx])
+    test_data = CounterfactualData(X[:,test_idx], y[:,test_idx])
+    return train_data, test_data
+end
+
+
+function undersample(data::CounterfactualData, n_per_class::Int)
+    
+    X,y = unpack(data)
+    n_classes, N = size(y)
+    if n_classes > 2
+        y_cls = Flux.onecold(y,1:n_classes)
+    else
+        y_cls = y
+    end
+    classes_ = sort(unique(y_cls))
+
+    idx = sort(reduce(vcat,[sample(findall(vec(y_cls.==cls)), n_per_class,replace=false) for cls in classes_]))
+    data = CounterfactualData(X[:,idx], y[:,idx])
+
+    return data
+
+end

### 2022-08-08 src/experiments/functions.jl

@@ -162,7 +162,8 @@ function update!(experiment::Experiment, recourse_system::RecourseSystem, chosen
     factuals = select_factual(counterfactual_data,chosen_individuals)
     results = generate_counterfactual(
         factuals, target, counterfactual_data, M, generator; 
-        T=T, γ=γ, num_counterfactuals=experiment.num_counterfactuals
+        T=T, γ=γ, num_counterfactuals=experiment.num_counterfactuals,
+        convergence=args.convergence
     );
     
     indices_ = rand(1:experiment.num_counterfactuals,length(results)) # randomly draw from generated counterfactuals

### 2022-08-08 src/generators/GravitationalGenerator.jl

@@ -4,18 +4,18 @@ mutable struct GravitationalGenerator <: AbstractGradientBasedGenerator
     loss::Union{Nothing,Symbol} # loss function
     complexity::Function # complexity function
     λ::Union{AbstractFloat,AbstractVector} # strength of penalty
-    ϵ::AbstractFloat # learning rate
+    opt::Any # optimizer
     τ::AbstractFloat # tolerance for convergence
     K::Int # number of K randomly chosen neighbours
     centroid::Union{Nothing,AbstractArray}
 end
 
 # API streamlining:
-using Parameters
+using Parameters, Flux
 @with_kw struct GravitationalGeneratorParams
-    ϵ::AbstractFloat=0.1
+    opt::Any=Flux.Optimise.Descent()
     τ::AbstractFloat=1e-5
-    K::Int=5
+    K::Int=50
     centroid::Union{Nothing,AbstractArray}=nothing
 end
 
@@ -25,7 +25,7 @@ end
         loss::Symbol=:logitbinarycrossentropy,
         complexity::Function=norm,
         λ::AbstractFloat=0.1,
-        ϵ::AbstractFloat=0.1,
+        opt::Any=Flux.Optimise.Descent(),
         τ::AbstractFloat=1e-5
     )
 
@@ -36,9 +36,9 @@ An outer constructor method that instantiates a generic generator.
 generator = GravitationalGenerator()
 ```
 """
-function GravitationalGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=[0.1, 5.0],kwargs...)
+function GravitationalGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=[0.1,1.0],kwargs...)
     params = GravitationalGeneratorParams(;kwargs...)
-    GravitationalGenerator(loss, complexity, λ, params.ϵ, params.τ, params.K, params.centroid)
+    GravitationalGenerator(loss, complexity, λ, params.opt, params.τ, params.K, params.centroid)
 end
 
 # Complexity:

### 2022-08-08 src/plotting.jl

@@ -1,25 +1,55 @@
-using Gadfly
-import Plots: plot
-using DataFrames
+# using RCall
+# import Plots: plot
+# using DataFrames
+# using Statistics
+# using Images
 
-function plot(results::ExperimentResults, variable::Symbol=:mmd; kwargs...)
+# function plot(results::ExperimentResults, variable::Symbol=:mmd, scope::Symbol=:model; kwargs...)
     
-    df = results.output
-    @assert variable in unique(df.name) "Not a valid variable."
+#     df = results.output
+#     @assert variable in unique(df.name) "Not a valid variable."
 
-    gdf = groupby(df, [:generator, :model, :n, :name, :scope])
-    df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+#     gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+#     df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+#     df_plot = df_plot[(df_plot.scope .== scope),:]
+#     df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
 
+#     ncol = length(unique(df_plot.model))
+#     @rimport ggplot2 as gg
+#     gg.ggplot(df_plot,aes(x=:n, y=:mean, color=:generator)) +
+#         gg.geom_line() +
+#         gg.facet_wrap(R"name ~ model", scale="free_y", ncol=ncol) 
+#     temp_path = joinpath(tempdir(),"plot.png")
+#     gg.ggsave(temp_path)
+#     img = Images.load(temp_path)
 
-    # Plot validity:
-    plt = Gadfly.plot(
-        df_plot[df_plot.name.==variable,:], 
-        ygroup=:scope, xgroup=:model, x=:n, y=:mean, ymin=:ymin, ymax=:ymax, color=:generator,
-        Geom.subplot_grid(Geom.line, Geom.ribbon, free_y_axis=true),
-        kwargs...
-    )
+#     return img
 
-    return plt
+# end
 
-end
+# function plot(results::ExperimentResults, n::Int, variable::Symbol=:mmd, scope::Symbol=:model; kwargs...)
+    
+#     df = results.output
[…]

### 2022-08-08 dev/notebooks/real_world.qmd
added/improved functionality to plot experiment results. Using ggplot backend because it is still just so much better than anything else sorry

--- 2022-07-25 dev/notebooks/real_world.qmd
@@
 ```{julia}
-catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
-choices = [:give_me_some_credit_balanced]
-data_sets = filter(p -> p[1] in choices, catalogue)
+max_obs = 10000
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
 ```
@@
 ```{julia}
-models = [:FluxModel, :LaplaceReduxModel]
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 50
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
+```
+
+```{julia}
+models = [
+    :LogisticRegression, 
+    :FluxModel, 
+    :FluxEnsemble
+]
 generators = Dict(
@@
     :REVISE=>REVISEGenerator(),
-    :DICE=>DiCEGenerator()
+    :DICE=>DiCEGenerator(),
+    :Gravitational=>GravitationalGenerator(λ=[0.1,2]),
 )
@@
 ```{julia}
-results = run_experiments(data_sets,models,generators;save_path=output_path,evaluate_every=10,n_rounds=50)
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
+```
+
+## Strict convergence
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+T = 500
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
+    convergence=:strict, save_name_suffix="strict"
+)
+Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
 ```
@@
 ```{julia}
-[plot(res) for res in results]
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
 ```
+
+```{julia}
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
+end
+```
+
+## Simple convergence
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+using Serialization
+results = run_experiments(
+    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap
+)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+```
+
+### Plots
+
+```{julia}
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
+end
+```

### 2022-08-08 dev/notebooks/synthetic.qmd
added/improved functionality to plot experiment results. Using ggplot backend because it is still just so much better than anything else sorry

@@
 ```{julia}
-plot_dict = Dict()
+line_charts = Dict()
+errorbar_charts = Dict()
 for (data_name, res) in results
-    output_data = res.output
-    metrics = unique(output_data.name)
-    plots = Dict(metric => plot(res, metric) for metric in metrics)
-    plot_dict[data_name] = plots
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
 end

### 2022-08-09 dev/paper/paper.tex
some work on rephrasing narrative in intro

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+Research on algorithmic recourse has also so far typically addressed the
+issue from the perspective of one single individual and has indeed been
+referred to as \textbf{individual recourse} in some places. Arguably
+though, most real-world applications that warrant algorithmic recourse
+involve potentially large groups of individuals typically competing for
+scarce resources. Our work demonstrates that in such scenarios, choices
+made by or for one single individual are likely to affect the broader
+collective of individuals in ways that current approaches to AR fail to
+account for. More specifically, we argue that a strict focus on
+minimizing the private costs faced by individuals may be too narrow an
+objective.
+
 Figure~\ref{fig-poc} illustrates this idea for a binary problem
@@
 \textbf{endogenous} because they are induced by the implementation of
-recourse itself.
+recourse itself. The term \textbf{macrodynamics} is borrowed from the
+economics literature and used to describe processes involving whole
+groups or societies.
 
@@
 
-We think that these types of endogenous dynamics may be problematic.
-Firstly, model shifts may inadvertently change classification outcomes
-for individuals who never received and implemented recourse. Secondly
-and relatedly, we observe in Figure~\ref{fig-poc} that as the decision
-boundary moves in the direction of the non-target class, counterfactual
-paths become shorter: in the consumer credit example, individuals that
-previously would have been denied credit based on their input features
-are suddenly considered as creditworthy. Average default risk across all
-borrowers can therefore be expected to increase. Conversely, lenders
-that anticipate such dynamics may choose to deny credit to individuals
-that have implemented AR, thereby compromising the validity of AR.
+We think that these types of endogenous dynamics may be problematic and
+warrant our attention. Firstly, model shifts may inadvertently change
+classification outcomes for individuals who never received and
+implemented recourse. Secondly and relatedly, we observe in
+Figure~\ref{fig-poc} that as the decision boundary moves in the
+direction of the non-target class, counterfactual paths become shorter:
+in the consumer credit example, individuals that previously would have
+been denied credit based on their input features are suddenly considered
+as creditworthy. Average default risk across all borrowers can therefore
+be expected to increase. Conversely, lenders that anticipate such
+dynamics may choose to refrain from offering recourse (and hence credit)
+to more than just a tiny share of individuals. In that latter and
+perhaps more likely scenario, the probability of being offered recourse
+decreases with every individual that implements recourse: in other
+words, the actions of first-movers exert a negative externality on
+future would-be borrowers.
 
 To the best of our knowledge this is the first work investigating
-endogenous dynamics in AR. Our contributions to the state of knowledge
-are two-fold. Firstly, we introduce an experimental framework extending
-previous work by \protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
-which allows for benchmarking different counterfactual generators in
-terms of the endogenous domain and model shifts they introduce. To this
-end we propose a number of novel evaluation metrics. Secondly, we use
-this framework to provide the first in-depth analysis of endogenous
-recourse dynamics induced by various popular counterfactual generators
-including \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
-\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
-\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and
-\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}.
-
-We find that \ldots{}
-
-The paper is structured as follows: Section~\ref{sec-related} places our
-work in the broader context of related literature.
-Section~\ref{sec-method} presents our methodology and data.
-Section~\ref{sec-empirical} presents our empirical findings which are
-then discussed in Section~\ref{sec-discussion}. We also point to some of
-the limitations or our work as well as avenues for future research in
+endogenous macrodynamics in AR. Our contributions to the state of
+knowledge are as follows: firstly, we posit a compelling argument that
+calls for a novel perspective on algorithmic recourse extending our
+focus from single individuals to groups. Secondly, we introduce an
+experimental framework extending previous work by
+\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
+which enables us to study macrodynamics of algorithmic recourse through
+simulations. To this end we propose a number of novel evaluation metrics
+that can be used to qunatify and benchmark the macrodynmics introduced
+by different counterfactual generators. Thirdly, we use this framework
+to provide a first in-depth analysis of endogenous recourse dynamics
+induced by various popular counterfactual generators including
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
+\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
+\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. Finally, we
+discuss what drives endogenous dynamics and strategies to mitigate them.
+
+The remainder of the paper is structured as follows:
+Section~\ref{sec-related} places our work in the broader context of
+related literature. Section~\ref{sec-method} presents our methodology
+and data. Section~\ref{sec-empirical} presents our empirical findings
+which are then discussed in the broader context of the literature in
+Section~\ref{sec-discussion}. We also point to some of the limitations
+or our work as well as avenues for future research in
 Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
@@
 al.~layed out 60 algorithms that have been proposed since 2014
-\protect\hyperlink{ref-karimi2020survey}{{[}9{]}}. Another survey
+\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
 published around the same time by Verma et al.~described 29 algorithms
-\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. Different
+\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
 approaches vary primarily in terms of the objective functions they
@@
 \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
-\protect\hyperlink{ref-ustun2019actionable}{{[}11{]}}, realistic
-\protect\hyperlink{ref-schut2021generating}{{[}12{]}}, sparse, diverse
-\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and if possible
+\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
+\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
 causally founded
-\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
+\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.
 
@@
 shifts remains a largely unexplored research challenge at this point
-\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. We have been
+\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
 able to identify only one recent work that considers the implications of
 \textbf{exogenous} domain and model shifts in the context of AR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
 are strictly of external origin. For example, they might stem from data
 correction, temporal shifts or geospatial changes
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose framework
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose framework
 for algorithmic recourse (ROAR) that evidently improves robustness to
@@
 samples through the application of two-sample tests
-\protect\hyperlink{ref-rabanser2019failing}{{[}15{]}}. This task is a
+\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
 generalization of the anomaly detection problem for large datasets,
@@
 anomaly detection have been summarized
-\protect\hyperlink{ref-chandola2009anomaly}{{[}16{]}}. Another
+\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
 well-established research topic is that of concept drift -- situations
@@
 the output of a model
-\protect\hyperlink{ref-widmer1996learning}{{[}17{]}}. For instance, Gama
+\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
 et al.~offer a review of the adaptive learning techniques which can
-handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}18{]}}.
+handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
 Less previous work is available on the related topic of model drift -
@@
 resistant different machine learning models are to the model drift
-\protect\hyperlink{ref-nelson2015evaluating}{{[}19{]}}. Ackerman et
+\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
 al.~offer a method to detect changes in model performance when ground
 truth is not available
-\protect\hyperlink{ref-ackerman2021machine}{{[}20{]}}.
+\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.
 
@@
 first brought up by the authors behind ROAR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
 they refer to model shifts as simply any perturbation \(\Delta\) to the
@@
 baseline for our analysis here, it is worth noting that in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
 perturbations are mechanically introduced. In contrast we are interested
@@
 availability in this space. Recent work has started to address this gap:
-firstly, \protect\hyperlink{ref-de2021framework}{{[}21{]}} run a large
+firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
 benchmarking study using different algorithmic aproaches and numerous
@@
 finally,
-\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}22{]}}
+\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
 provides an extensible and language-agnostic implementation in Julia.
@@
 include the following generator in our benchmarking exercises: REVISE
-\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}, CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}, DICE
-\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and a greedy
+\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DICE
+\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and a greedy
 approach that relies on probabilistic models
-\protect\hyperlink{ref-schut2021generating}{{[}12{]}}.
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}.
 
@@
 because they are implicitly generated by the (learned) DGP
-\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}. CLUE distinguishes
+\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}. CLUE distinguishes
 itself from REVISE and other counterfactual generators in that it aims
@@
 for probabilistic models. The \textbf{Greedy} approach proposed by
-\protect\hyperlink{ref-schut2021generating}{{[}12{]}} also works with
-the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that
-can produce predictive uncertainty estimates. The authors show that in
-this setting the complexity penalty \(h(\cdot)\) in
-Equation~\ref{eq-general} is redundant and meaningful counterfactuals
-can be generated in a fast and efficient manner through a modified
-Jacobian-based Saliency Map Attack (JSMA). Finally, \textbf{DICE}
-distinguishes itself from all other generators considered here in that
-it aims to generate a diverse set of \(K>1\) counterfactuals. To this
-end the authors use a complexity penalty \(h(\mathbf{s}^\prime)\) that
-favours diverse outcomes, in the sense that \(s_1, ... , s_K\) look as
-different from each other as possible.
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}} also works with the
+subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can
+produce predictive uncertainty estimates. The authors show that in this
+setting the complexity penalty \(h(\cdot)\) in Equation~\ref{eq-general}
+is redundant and meaningful counterfactuals can be generated in a fast
+and efficient manner through a modified Jacobian-based Saliency Map
+Attack (JSMA). Finally, \textbf{DICE} distinguishes itself from all
+other generators considered here in that it aims to generate a diverse
+set of \(K>1\) counterfactuals. To this end the authors use a complexity
+penalty \(h(\mathbf{s}^\prime)\) that favours diverse outcomes, in the
+sense that \(s_1, ... , s_K\) look as different from each other as
+possible.
 
@@
 may affect the resulting recourse dynamics. Following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} we generate
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} we generate
 synthetic data in \(\mathbb{R}^2\) to also allow for a visual
@@
 the paths of the recourse. For generators that use latent space search
-(\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
-\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}) or rely on (and
+(\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on (and
 have access to) probabilistic models
-(\protect\hyperlink{ref-antoran2020getting}{{[}8{]}},
-\protect\hyperlink{ref-schut2021generating}{{[}12{]}}) we expect that
+(\protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
 counterfactuals will end up in regions of the target domain that are
@@
 the model parameters at each point in time \(t\) following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We define
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
 \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
@@
 
+\hypertarget{sec-discussion}{%
+\section{Discussion}\label{sec-discussion}}
+
 \begin{enumerate}
@@
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
-
 \hypertarget{sec-limit}{%
@@
 updated on a regular basis
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also find it
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also find it
 plausible to assume that the implementation of recourse happens
@@
 exogenous shifts of the nature investigated in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also make
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also make
 implicit assumptions about the utility functions of the involved agents
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
+\CSLLeftMargin{{[}6{]} }%
+\CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
+Julia package for Counterfactual Explanations and Algorithmic
+Recourse}}. 2022. Available:
+\url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
+\CSLLeftMargin{{[}7{]} }%
+\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
+counterfactual explanations by implicit minimisation of epistemic and
+aleatoric uncertainties,''} in \emph{International conference on
+artificial intelligence and statistics}, 2021, pp. 1756--1764.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
-\CSLLeftMargin{{[}6{]} }%
+\CSLLeftMargin{{[}8{]} }%
 \CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
-\CSLLeftMargin{{[}7{]} }%
+\CSLLeftMargin{{[}9{]} }%
 \CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
@@
 \leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-\CSLLeftMargin{{[}8{]} }%
+\CSLLeftMargin{{[}10{]} }%
 \CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
-\CSLLeftMargin{{[}9{]} }%
+\CSLLeftMargin{{[}11{]} }%
 \CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
-\CSLLeftMargin{{[}10{]} }%
+\CSLLeftMargin{{[}12{]} }%
 \CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
-\CSLLeftMargin{{[}11{]} }%
+\CSLLeftMargin{{[}13{]} }%
 \CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
-\CSLLeftMargin{{[}12{]} }%
-\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
-counterfactual explanations by implicit minimisation of epistemic and
-aleatoric uncertainties,''} in \emph{International conference on
-artificial intelligence and statistics}, 2021, pp. 1756--1764.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
-\CSLLeftMargin{{[}13{]} }%
+\CSLLeftMargin{{[}14{]} }%
 \CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
-\CSLLeftMargin{{[}14{]} }%
+\CSLLeftMargin{{[}15{]} }%
 \CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
-\CSLLeftMargin{{[}15{]} }%
+\CSLLeftMargin{{[}16{]} }%
 \CSLRightInline{S. Rabanser, S. Günnemann, and Z. Lipton, {``Failing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
-\CSLLeftMargin{{[}16{]} }%
+\CSLLeftMargin{{[}17{]} }%
 \CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly
@@
 \leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
-\CSLLeftMargin{{[}17{]} }%
+\CSLLeftMargin{{[}18{]} }%
 \CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
-\CSLLeftMargin{{[}18{]} }%
+\CSLLeftMargin{{[}19{]} }%
 \CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
-\CSLLeftMargin{{[}19{]} }%
+\CSLLeftMargin{{[}20{]} }%
 \CSLRightInline{K. Nelson, G. Corbin, M. Anania, M. Kovacs, J. Tobias,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
-\CSLLeftMargin{{[}20{]} }%
+\CSLLeftMargin{{[}21{]} }%
 \CSLRightInline{S. Ackerman, P. Dube, E. Farchi, O. Raz, and M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
-\CSLLeftMargin{{[}21{]} }%
+\CSLLeftMargin{{[}22{]} }%
 \CSLRightInline{R. M. B. de Oliveira and D. Martens, {``A framework and
@@
 data,''} \emph{Applied Sciences}, vol. 11, no. 16, p. 7274, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
-\CSLLeftMargin{{[}22{]} }%
-\CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
-Julia package for Counterfactual Explanations and Algorithmic
-Recourse}}. 2022. Available:
-\url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
 

### 2022-08-09 dev/paper/sections/introduction.qmd
some work on rephrasing narrative in intro

@@
 
+Research on algorithmic recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as **individual recourse** in some places.
+Arguably though, most real-world applications that warrant algorithmic recourse involve potentially large groups of individuals typically competing for scarce resources.
+Our work demonstrates that in such scenarios, choices made by or for one single individual are likely to affect the broader collective of individuals in ways that current approaches to AR fail to account for.
+More specifically, we argue that a strict focus on minimizing the private costs faced by individuals may be too narrow an objective.
+
 @fig-poc illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
@@
 We refer to these types of dynamics as **endogenous** because they are induced by the implementation of recourse itself.
+The term **macrodynamics** is borrowed from the economics literature and used to describe processes involving whole groups or societies.
 
@@
 
-We think that these types of endogenous dynamics may be problematic.
+We think that these types of endogenous dynamics may be problematic and warrant our attention.
 Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
@@
 Average default risk across all borrowers can therefore be expected to increase.
-Conversely, lenders that anticipate such dynamics may choose to deny credit to individuals that have implemented AR, thereby compromising the validity of AR.
+Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals.
+In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.
 
-To the best of our knowledge this is the first work investigating endogenous dynamics in AR.
-Our contributions to the state of knowledge are two-fold.
-Firstly, we introduce an experimental framework extending previous work by @pawelczyk2021carla, which allows for benchmarking different counterfactual generators in terms of the endogenous domain and model shifts they introduce.
-To this end we propose a number of novel evaluation metrics.
-Secondly, we use this framework to provide the first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
+To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on algorithmic recourse extending our focus from single individuals to groups.
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations.
+To this end we propose a number of novel evaluation metrics that can be used to qunatify and benchmark the macrodynmics introduced by different counterfactual generators.
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
+Finally, we discuss what drives endogenous dynamics and strategies to mitigate them.
 
-We find that ...
-
-The paper is structured as follows: @sec-related places our work in the broader context of related literature.
+The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature.
 @sec-method presents our methodology and data.
-@sec-empirical presents our empirical findings which are then discussed in @sec-discussion.
+@sec-empirical presents our empirical findings which are then discussed in the broader context of the literature in @sec-discussion.
 We also point to some of the limitations or our work as well as avenues for future research in @sec-limit.

### 2022-08-09 dev/notebooks/gravitational_generator.qmd
extended methodology to include Gravitational generator

@@
 Λ₂ = [0.1, 1, 5]
-counterfactuals = []
+counterfactuals_strict = []
 generators = []
@@
     generators = vcat(generators..., generator)
-    counterfactuals = vcat(
-      counterfactuals...,
+    counterfactuals_strict = vcat(
+      counterfactuals_strict...,
       generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
@@
     λ₂ = Λ₂[i]
-    counterfactual = counterfactuals[i]  
+    counterfactual = counterfactuals_strict[i]  
     plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
@@
 
+### Comparison - simple vs. strict convergence
+
 ```{julia}
-
+idx = 2
+x1,x2 = generators[idx].centroid[1], generators[idx].centroid[2]
+plt1 = plot(counterfactuals[idx], plot_up_to=minimum([T,T_]), title="Simple")
+scatter!(plt1, [x1], [x2], colour=:purple, label="attractor")
+plt2 = plot(counterfactuals_strict[idx], plot_up_to=minimum([T,T_]), title="Strict")
+scatter!(plt2, [x1], [x2], colour=:purple, label="attractor")
+plt = plot(plt1, plt2, size=(850,350), layout=(1,2))
+savefig(plt, joinpath(www_path,"gravitational_generator_comparison.png"))
 ```

### 2022-08-09 dev/paper/paper.tex
extended methodology to include Gravitational generator

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 search is well-suited for differentiable black-box models like deep
-neural networks. We can restate Equation~\ref{eq-solution} in a more
-general form that encompasses most gradient-based approaches to
-counterfactual search:
+neural networks.
+
+\hypertarget{from-individual-recourse}{%
+\subsubsection{From individual recourse
+\ldots{}}\label{from-individual-recourse}}
+
+We can restate Equation~\ref{eq-solution} in a more general form that
+encompasses most gradient-based approaches to counterfactual search:
 
@@
 
+\hypertarget{towards-collective-recourse}{%
+\subsubsection{\ldots{} towards collective
+recourse}\label{towards-collective-recourse}}
+
+To explicitly address the notion that individual recourse may affect the
+outcome and prospect of other individuals, we propose to extend
+Equation~\ref{eq-general} as follows:
+
+\begin{equation}\protect\hypertarget{eq-collective}{}{
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
+\end{aligned}
+}\label{eq-collective}\end{equation}
+
+Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
+the individual as before and \(\lambda_1\) governs to what extent that
+private cost ought to be penalised. The newly introduced term
+\(g(f(s_k^\prime))\) is meant to capture and address social costs
+incurred by the collective of individuals in response to changes in
+\(\mathbf{s}^\prime\). The trade-off between private and social costs is
+determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
+individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
+nor do we intend to provide a definite answer in this work, if such even
+exists. A straight-forward choice simply extends the baseline approach
+by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
+of only penalizing the distance of the individuals' counterfactual to
+its factual, we propose penalizing its distance to some sensible point
+in the target domain, for example the sample average:
+\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
+\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
+gravitate towards the specified point in the target domain. In the
+remainder of this paper we will therefore refer to this approach as
+\textbf{Gravitational} generator, when we investigate its potential
+usefulness for mitigating endongenous macrodynamics\footnote{Note that
+  despite the naming convention our goal here is not to provide yet
+  another counterfactual generator, but merely investigate the most
+  simple penalty we can think of with respect to its effectiveness.}.
+
+\hypertarget{a-note-on-convergence}{%
+\subsubsection{A note on convergence}\label{a-note-on-convergence}}
+
+For this simple mitigating strategy underlying the Gravitational
+generator to work as expected, one needs to ensure that counterfactual
+search continues, even after a predetermined threshold probability
+\(\gamma\) has potentially already been reached.
+Figure~\ref{fig-convergence} illustrates this distinction: if one
+chooses to terminate search once the desired threshold is reached (left
+panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
+actually satisfied (compare to right panel). More generally, if
+convergence is defined simply in terms of flipping the predicted label
+with some desired degree of confidence, this corresponds to essentially
+ignoring any parts of the counterfactual search objective that do not
+involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
+be appropriate for some applications, in general this seems like an odd
+convention. Since we nonetheless seen convergence specified simply in
+terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
+  define convergence of Algorithm 1 in this way. The implementation of
+  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
+  also defined in this way.}, we thought it worth making this
+distinction explicit.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}
+
+}
+
+\caption{\label{fig-convergence}Comparison of counterfactual search
+outcome with simple (left) and strict convergence (right).}
+
+\end{figure}
+
 \hypertarget{sec-method-experiment}{%
@@
 \hypertarget{sec-limit}{%
-\subsection{Limitations and Future Work}\label{sec-limit}}
+\section{Limitations and Future Work}\label{sec-limit}}
+
+While we believe that this work constitutes a valuable starting point
+for addressing existing issues in algorithmic recourse from a fresh
+perspective, we are aware of several of its limitations. In the
+following we highlight some of these limitations and point to avenues
+for future research.
+
+\hypertarget{experimental-setup}{%
+\subsection{Experimental Setup}\label{experimental-setup}}
 
@@
 \hypertarget{data}{%
-\subsubsection{Data}\label{data}}
-
-The number of real-world datasets considered in this work is limited,
-constrained by the fact that at the time of writing
-\texttt{CounterfactualExplanations.jl} only supported continuous
-features.
+\subsection{Data}\label{data}}
+
+Largely in line with the existing literature on algorithmic recourse, we
+have limited our analysis of real-world data to three commonly used
+benchmark datasets that involve binary prediction tasks. Future work may
+benefit from including novel datasets or extending the analysis to
+multi-class or regression problems, the latter arguably representing the
+most common objective in Finance and Economics. It is also worth
+mentioning that the use of real-world datasets considered in this work
+is constrained by the fact that at the time of writing
+\texttt{CounterfactualExplanations.jl} only supports continuous
+features, at least of some of the counterfactual generators considered
+here. The fact that we therefore had to discard discrete features led to
+relatively poor initial performance of our classifiers in some cases.
+While this is indeed a limitation we intend to address in future and
+derivative work, our findings with respect to endogenous macrodynamics
+do not hinge on strong classifier performance.
+
+\hypertarget{classifiers}{%
+\subsection{Classifiers}\label{classifiers}}
+
+For reasons stated earlier we have limited our analysis to
+differentiable linear and non-linear classifiers, in particular logistic
+regression and deep neural networks. While these sorts of classifiers
+have also typically been analyzed in the existing literature on
+counterfactual explanations and algorithmic recourse, they represent
+only a subset of popular machine learning models employed in practice -
+both black-box and glass-box. Despite the success and popularity of deep
+learning in the context of high-dimensional data such as image, audio
+and video, empirical evidence suggests that other models such as boosted
+decision trees may have an edge when it comes to lower-dimensional
+tabular datasets, such as the ones considered here
+\protect\hyperlink{ref-grinsztajn2022tree}{{[}28{]}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
+\CSLLeftMargin{{[}27{]} }%
+\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
+Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: A
+survey,''} \emph{arXiv preprint arXiv:2110.01889}, 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
+\CSLLeftMargin{{[}28{]} }%
+\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do
+tree-based models still outperform deep learning on tabular data?''}
+\emph{arXiv preprint arXiv:2207.08815}, 2022.}
+
 \end{CSLReferences}

### 2022-08-09 dev/paper/sections/methodology.qmd
extended methodology to include Gravitational generator

@@
 Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks.
+
+### From individual recourse ...
+
 We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
@@
 All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
+
+### ... towards collective recourse
+
+To explicitly address the notion that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:
+
+$$
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
+\end{aligned}
+$$ {#eq-collective}
+
+Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalised.
+The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address social costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
+The trade-off between private and social costs is determined by the ratio between $\lambda_1$ and $\lambda_2$.
+As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
+A straight-forward choice simply extends the baseline approach by @wachter2017counterfactual: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{\mathbf{x}}$.
+For such a recourse objective, higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point in the target domain.
+In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+
+### A note on convergence
+
+For this simple mitigating strategy underlying the Gravitational generator to work as expected, one needs to ensure that counterfactual search continues, even after a predetermined threshold probability $\gamma$ has potentially already been reached.
+@fig-convergence illustrates this distinction: if one chooses to terminate search once the desired threshold is reached (left panel) the gravitational pull towards $\bar{\mathbf{x}}$ is never actually satisfied (compare to right panel).
+More generally, if convergence is defined simply in terms of flipping the predicted label with some desired degree of confidence, this corresponds to essentially ignoring any parts of the counterfactual search objective that do not involve $\ell(M(f(s_k^\prime)),t)$ beyond that point.
+While this may be appropriate for some applications, in general this seems like an odd convention.
+Since we nonetheless seen convergence specified simply in terms of reaching the threshold probability in some places^[@joshi2019towards define convergence of Algorithm 1 in this way.
+The implementation of @wachter2017counterfactual in CARLA is also defined in this way.], we thought it worth making this distinction explicit.
+
+![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}
 

### 2022-08-09 dev/paper/sections/limitations.qmd
2/3 real world data done

--- 2022-07-06 paper/sections/limitations.qmd
@@
-## Limitations and Future Work {#sec-limit}
+# Limitations and Future Work {#sec-limit}
+
+While we believe that this work constitutes a valuable starting point for addressing existing issues in algorithmic recourse from a fresh perspective, we are aware of several of its limitations.
+In the following we highlight some of these limitations and point to avenues for future research.
+
+## Experimental Setup
 
@@
 In practice it is doubtful that agents behave according to such simple rules.
+Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).
 
-Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).
+## Data
+
+Largely in line with the existing literature on algorithmic recourse, we have limited our analysis of real-world data to three commonly used benchmark datasets that involve binary prediction tasks.
+Future work may benefit from including novel datasets or extending the analysis to multi-class or regression problems, the latter arguably representing the most common objective in Finance and Economics.
+It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing `CounterfactualExplanations.jl` only supports continuous features, at least of some of the counterfactual generators considered here.
+The fact that we therefore had to discard discrete features led to relatively poor initial performance of our classifiers in some cases.
+While this is indeed a limitation we intend to address in future and derivative work, our findings with respect to endogenous macrodynamics do not hinge on strong classifier performance.
+
+## Classifiers
+
+For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks.
+While these sorts of classifiers have also typically been analyzed in the existing literature on counterfactual explanations and algorithmic recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here [@borisov2021deep, @grinsztajn2022tree].

### 2022-08-09 dev/notebooks/data_preprocessing.qmd/synthetic_data.qmd
new and larger synthetic datasets

---
title: Generating Synthetic Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
theme(:wong)
output_path = output_dir("synthetic")
www_path = www_dir("synthetic")
data_path = data_dir("synthetic")
```

```{julia}
using MLJ, DataFrames, CSV, VegaLite
n = 1000
p = 2

using Random
Random.seed!(42)

# Linearly separable:
X, y = make_blobs(n, p; centers=2, center_box=(-2 => 2), cluster_std=0.1)
df = DataFrame(X)
df.target = y
df |> @vlplot(:point, x=:x1, y=:x2, color = :"target:n")
CSV.write(joinpath(data_path, "linearly_separable.csv"),df)

# Overlapping:
X, y = make_blobs(n, p; centers=2, center_box=(-2 => 2), cluster_std=0.5)
df = DataFrame(X)
df.target = y
df |> @vlplot(:point, x=:x1, y=:x2, color = :"target:n")
CSV.write(joinpath(data_path, "overlapping.csv"),df)

# Circles:
X, y = make_circles(n; noise=0.15, factor=0.01)
df = DataFrame(X)
df.target = y
df |> @vlplot(:point, x=:x1, y=:x2, color = :"target:n")
CSV.write(joinpath(data_path, "circles.csv"),df)

# Moon:
X, y = make_moons(n)
df = DataFrame(X)
df.target = y
df |> @vlplot(:point, x=:x1, y=:x2, color = :"target:n")
CSV.write(joinpath(data_path, "moons.csv"),df)
```

### 2022-08-09 dev/notebooks/real_world.qmd
new and larger synthetic datasets

@@
 ```{julia}
-max_obs = 10000
+max_obs = 1000
 data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
@@
 using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
 ```

### 2022-08-09 dev/paper/sections/introduction.qmd
minor things

@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on algorithmic recourse extending our focus from single individuals to groups.
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations.
-To this end we propose a number of novel evaluation metrics that can be used to qunatify and benchmark the macrodynmics introduced by different counterfactual generators.
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations that can be fully parallelized.
 Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
-Finally, we discuss what drives endogenous dynamics and strategies to mitigate them.
+To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators.
+Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.
 

### 2022-08-09 dev/paper/sections/related.qmd
minor things

--- 2022-07-06 paper/sections/related.qmd
@@
 
+As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances.
+We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals [@carrizosa2021generating].
+The authors propose an optimization framework that generates collective counterfactuals.
+We provide a motivation for doing so from the perspective of endogenous macrodynamics of algorithmic recourse.
+
 ## Domain and Model Shifts {#sec-related-shifts}
@@
 
-...
+Despite the large and growing number of approaches to counterfactual search there have been surprisingly few benchmark studies that compare different methodologies.
+This may be partially due to limited software availability in this space.
+Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework that can be used to apply and benchmark different methodologies; finally, @altmeyer2022CounterfactualExplanations provides an extensible, fast and language-agnostic implementation in Julia.
+Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.

### 2022-08-09 dev/notebooks/data_preprocessing.qmd/synthetic_data.qmd

@@
 df = DataFrame(X)
-df.target = y
+df.target .= ifelse.(y.==1,0,1)
 df |> @vlplot(:point, x=:x1, y=:x2, color = :"target:n")

### 2022-08-09 dev/notebooks/real_world.qmd

@@
 n_bootstrap = 1
-T = 500
+T = 250
+generative_model_params = (epochs=250, latent_dim=8)
 using Serialization
@@
     save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
-    convergence=:strict, save_name_suffix="strict"
+    convergence=:strict, generative_model_params=generative_model_params, save_name_suffix="strict"
 )

### 2022-08-09 dev/notebooks/synthetic.qmd
also added basic functionality to generate summary tables using Rs kableExtra package
some work on rephrasing narrative in intro

@@
     :overlapping, 
-    :plus_shaped, 
-    :balanced_positive_clusters,
+    :circles, 
+    :moons,
 ]
@@
 ```
+
+## Strict convergence
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
+    convergence=:strict, save_name_suffix="strict"
+)
+Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_test_after_$(name)_strict.png"))
+end
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        data = rec_sys[m].data
+        score = round(model_evaluation(M, data),digits=2)
+        plt = plot(M, data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+        x_wrongly_labelled = data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_train_after_$(name)_strict.png"))
+end
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
+```
+
+### Plots
+
+```{julia}
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
+end
+```
+
+### Tables
 
@@
 ```
-
-## Strict convergence
-
-```{julia}
-n_evals = 10
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-n_bootstrap = 1
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
-    convergence=:strict, save_name_suffix="strict"
-)
-Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        score = round(model_evaluation(M, exp_.test_data),digits=2)
-        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
-        x_wrongly_labelled = exp_.test_data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_test_after_$(name)_strict.png"))
-end
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        data = rec_sys[m].data
-        score = round(model_evaluation(M, data),digits=2)
-        plt = plot(M, data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
-        x_wrongly_labelled = data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_train_after_$(name)_strict.png"))
-end
-```
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
-```
-
-```{julia}
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
-end
-```

### 2022-08-09 dev/notebooks/gravitational_generator.qmd
small change to default params for gravitational generator

@@
 ```{julia}
-idx = 2
+idx = 1
 x1,x2 = generators[idx].centroid[1], generators[idx].centroid[2]

### 2022-08-09 src/generators/GravitationalGenerator.jl

@@ -36,7 +36,7 @@ An outer constructor method that instantiates a generic generator.
 generator = GravitationalGenerator()
 ```
 """
-function GravitationalGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=[0.1,1.0],kwargs...)
+function GravitationalGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=0.1,kwargs...)
     params = GravitationalGeneratorParams(;kwargs...)
     GravitationalGenerator(loss, complexity, λ, params.opt, params.τ, params.K, params.centroid)
 end

### 2022-08-09 dev/paper/paper.tex
bleh

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 which enables us to study macrodynamics of algorithmic recourse through
-simulations. To this end we propose a number of novel evaluation metrics
-that can be used to qunatify and benchmark the macrodynmics introduced
-by different counterfactual generators. Thirdly, we use this framework
-to provide a first in-depth analysis of endogenous recourse dynamics
-induced by various popular counterfactual generators including
+simulations that can be fully parallelized. Thirdly, we use this
+framework to provide a first in-depth analysis of endogenous recourse
+dynamics induced by various popular counterfactual generators including
 \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
@@
 \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. Finally, we
-discuss what drives endogenous dynamics and strategies to mitigate them.
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
+propose a number of novel evaluation metrics that can be used to
+quantify and benchmark the macrodynamics introduced by the different
+generators. Finally, we also discuss what drives endogenous dynamics and
+propose strategies to mitigate them.
 
@@
 such exogenous shifts.
+
+As mentioned earlier, research has so far also generally focused on
+generating counterfactuals for single individuals or instances. We have
+been able to identify only one existing work that investigates black-box
+model behavior towards a group of individuals
+\protect\hyperlink{ref-carrizosa2021generating}{\textbf{carrizosa2021generating?}}.
+The authors propose an optimization framework that generates collective
+counterfactuals. We provide a motivation for doing so from the
+perspective of endogenous macrodynamics of algorithmic recourse.
 
@@
 \protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
-framework that can be used to apply and benchmark different methodology;
-finally,
+framework that can be used to apply and benchmark different
+methodologies; finally,
 \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
-provides an extensible and language-agnostic implementation in Julia.
-For the experiments conducted in this work we relied on the latter.
+provides an extensible, fast and language-agnostic implementation in
+Julia. Since the experiments presented here involve extensive
+simulations, we have relied on and extended the Julia implementation due
+to the associated performance benefits.
 

### 2022-08-09 dev/notebooks/synthetic.qmd

@@
 ```
-
-### Tables
-
-## Simple convergence
-
-```{julia}
-n_evals = 10
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-n_bootstrap = 1
-using Serialization
-results = run_experiments(
-    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds,n_folds=n_folds,n_bootstrap=n_bootstrap
-)
-Serialization.serialize(joinpath(output_path,"results.jls"),results)
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        score = round(model_evaluation(M, exp_.test_data),digits=2)
-        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
-        x_wrongly_labelled = exp_.test_data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_test_after_$name.png"))
-end
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        data = rec_sys[m].data
-        score = round(model_evaluation(M, data),digits=2)
-        plt = plot(M, data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
-        x_wrongly_labelled = data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_train_after_$name.png"))
-end
-```
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results.jls"))
-```
-
-```{julia}
-plot_dict = Dict()
-```

### 2022-08-09 dev/notebooks/synthetic.qmd
bluh

@@
 ```{julia}
-catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
+max_obs = 1000
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
 choices = [
@@
 n_bootstrap = 1
+T = 100
 using Serialization
@@
     experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
     convergence=:strict, save_name_suffix="strict"

### 2022-08-09 dev/paper/paper.tex

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 model behavior towards a group of individuals
-\protect\hyperlink{ref-carrizosa2021generating}{\textbf{carrizosa2021generating?}}.
-The authors propose an optimization framework that generates collective
+\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
+propose an optimization framework that generates collective
 counterfactuals. We provide a motivation for doing so from the
@@
 samples through the application of two-sample tests
-\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
+\protect\hyperlink{ref-rabanser2019failing}{{[}17{]}}. This task is a
 generalization of the anomaly detection problem for large datasets,
@@
 anomaly detection have been summarized
-\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
+\protect\hyperlink{ref-chandola2009anomaly}{{[}18{]}}. Another
 well-established research topic is that of concept drift -- situations
@@
 the output of a model
-\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
+\protect\hyperlink{ref-widmer1996learning}{{[}19{]}}. For instance, Gama
 et al.~offer a review of the adaptive learning techniques which can
-handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
+handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}20{]}}.
 Less previous work is available on the related topic of model drift -
@@
 resistant different machine learning models are to the model drift
-\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
+\protect\hyperlink{ref-nelson2015evaluating}{{[}21{]}}. Ackerman et
 al.~offer a method to detect changes in model performance when ground
 truth is not available
-\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.
+\protect\hyperlink{ref-ackerman2021machine}{{[}22{]}}.
 
@@
 availability in this space. Recent work has started to address this gap:
-firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
+firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
 benchmarking study using different algorithmic aproaches and numerous
@@
 
-We use 6 synthetic binary classification datasets consisting of 200-400
-samples grouped in normally-distributed clusters.\footnote{To see how
-  the data is generated see here:
+We use four synthetic binary classification datasets consisting of 1000
+samples each.\footnote{To see how the data is generated see here:
   \url{https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb}}
@@
 
-{\centering \includegraphics{www/synthetic_data.png}
+{\centering \includegraphics[width=8cm,height=8cm]{www/placeholder.png}
 
@@
 experience financial difficulties in the next two years
-\protect\hyperlink{ref-gmsc_data}{{[}23{]}}. Originally consisting of
+\protect\hyperlink{ref-gmsc_data}{{[}24{]}}. Originally consisting of
 250,000 instances with 11 numerical attributes, the dataset was randomly
@@
 task of predicting if bank customers are credit-worthy or not
-\protect\hyperlink{ref-germancredit1994}{{[}24{]}}. It consists of 700
+\protect\hyperlink{ref-germancredit1994}{{[}25{]}}. It consists of 700
 positive and 300 negative instances charaterized by 7 numerical and 13
@@
 distribution we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}25{]}}. To that end, we
+\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
 combine the two samples and generate a large number of permutations of
@@
 \textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}26{]}} and
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
 estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
@@
 tabular datasets, such as the ones considered here
-\protect\hyperlink{ref-grinsztajn2022tree}{{[}28{]}}.
+\protect\hyperlink{ref-grinsztajn2022tree}{{[}29{]}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
+\CSLLeftMargin{{[}16{]} }%
+\CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero,
+{``Generating collective counterfactual explanations in score-based
+classification via mathematical optimization,''} 2021.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
-\CSLLeftMargin{{[}16{]} }%
+\CSLLeftMargin{{[}17{]} }%
 \CSLRightInline{S. Rabanser, S. Günnemann, and Z. Lipton, {``Failing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
-\CSLLeftMargin{{[}17{]} }%
+\CSLLeftMargin{{[}18{]} }%
 \CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly
@@
 \leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
-\CSLLeftMargin{{[}18{]} }%
+\CSLLeftMargin{{[}19{]} }%
 \CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
-\CSLLeftMargin{{[}19{]} }%
+\CSLLeftMargin{{[}20{]} }%
 \CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
-\CSLLeftMargin{{[}20{]} }%
+\CSLLeftMargin{{[}21{]} }%
 \CSLRightInline{K. Nelson, G. Corbin, M. Anania, M. Kovacs, J. Tobias,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
-\CSLLeftMargin{{[}21{]} }%
+\CSLLeftMargin{{[}22{]} }%
 \CSLRightInline{S. Ackerman, P. Dube, E. Farchi, O. Raz, and M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
-\CSLLeftMargin{{[}22{]} }%
+\CSLLeftMargin{{[}23{]} }%
 \CSLRightInline{R. M. B. de Oliveira and D. Martens, {``A framework and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}23{]} }%
+\CSLLeftMargin{{[}24{]} }%
 \CSLRightInline{K. Competition, {``Give me some credit, improve on the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
-\CSLLeftMargin{{[}24{]} }%
+\CSLLeftMargin{{[}25{]} }%
 \CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}25{]} }%
+\CSLLeftMargin{{[}26{]} }%
 \CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}26{]} }%
+\CSLLeftMargin{{[}27{]} }%
 \CSLRightInline{S. Hanneke, {``A bound on the label complexity of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}27{]} }%
+\CSLLeftMargin{{[}28{]} }%
 \CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-\CSLLeftMargin{{[}28{]} }%
+\CSLLeftMargin{{[}29{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do

### 2022-08-09 dev/paper/sections/methodology.qmd

@@
 
-We use 6 synthetic binary classification datasets consisting of 200-400 samples grouped in normally-distributed clusters.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
+We use four synthetic binary classification datasets consisting of 1000 samples each.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
 Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
 
-![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}
+![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/placeholder.png){#fig-synthetic-data fig.pos="h" width="8cm" height="8cm"}
 

### 2022-08-09 dev/paper/paper.tex
updated synthetic data image and other minor

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, sharp corners, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-{\centering \includegraphics[width=8cm,height=8cm]{www/placeholder.png}
+{\centering \includegraphics[width=8cm,height=2cm]{www/synthetic_data.png}
 
@@
 
-\caption{\label{fig-synthetic-data}PLACEHOLDER: A visualization of the
-synthetic classification datasets used in our experiments.}
+\caption{\label{fig-synthetic-data}Synthetic classification datasets
+used in our experiments.}
 

### 2022-08-09 dev/paper/sections/methodology.qmd
updated synthetic data image and other minor

@@
 
+```{julia}
+using Plots, PlotThemes
+theme(:wong)
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
+function plot_data(data,title)
+    plt = plot(title=title)
+    scatter!(data)
+    return plt
+end
+plts = [plot_data(data,name) for (name, data) in catalogue]
+plt = plot(plts..., layout=(1,4), size=(850,200))
+savefig(plt, "dev/paper/www/synthetic_data.png")
+```
+
 We use four synthetic binary classification datasets consisting of 1000 samples each.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
@@
 
-![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/placeholder.png){#fig-synthetic-data fig.pos="h" width="8cm" height="8cm"}
+![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}
 

### 2022-08-10 dev/notebooks/endo_roar_generator.qmd
added another generator that simply penalizes the loss the model incurs on the counterfactual.

---
title: EndoROAR Generator
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("generator")
www_path = www_dir("generator")
```

# `EndoROARGenerator`

```{julia}
using MLJ
N = 1000
X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-5 => 5), cluster_std=0.1)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')
```

```{julia}
using Flux
nn = Chain(Dense(2,1))
using Flux.Optimise: update!, ADAM
opt = ADAM()
epochs = 100
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
show_every = epochs/10
grad_norms = []
using LinearAlgebra
for epoch = 1:epochs
    grads_ = []
    for d in data
        gs = gradient(Flux.params(nn)) do
            l = loss(d...)
        end
        update!(opt, Flux.params(nn), gs)
        grads_ = vcat(grads_..., gs)
    end
    if epoch % show_every == 0
        println("Epoch " * string(epoch))
        @show avg_loss(data)
    end
    grad_norms = vcat(grad_norms..., norm(grads_))
end
```

```{julia}
M = FluxModel(nn)
```

```{julia}
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
T = 100
```

```{julia}
generator = GenericGenerator()
ce = generate_counterfactual(x, target,counterfactual_data, M, generator)
```

```{julia}
x_ = path(ce)
y_ = CounterfactualExplanations.Counterfactuals.counterfactual_label_path(ce)
data_ = zip(x_,y_)
grad_norms = []
loss_ = []
for d in data_
    gs = gradient(Flux.params(nn)) do
        loss(d...)
    end
    loss_ = vcat(loss_...,loss(d...))
    grad_norms = vcat(grad_norms..., norm(gs))
end
```

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals_strict = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = EndoROARGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals_strict = vcat(
      counterfactuals_strict...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
    )
end
```

```{julia}
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals_strict[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
# savefig(plt, joinpath(www_path,"endo_roar_generator.png"))
```

```{julia}
generators = Dict(
    :Generic => GenericGenerator(),
    :ROAR => EndoROARGenerator()
)
counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; convergence=:strict, T=T) for (name, gen) in generators])
```

```{julia}
plts = []
for (name,ce) ∈ counterfactuals
    plt = plot(ce, plot_up_to=minimum([T,T_]), title=name)
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(800,300), layout=(1,2))
```

### 2022-08-10 src/AlgorithmicRecourseDynamics.jl

@@ -10,7 +10,7 @@ using .Models
 
 include("generators/Generators.jl")
 using .Generators
-export GravitationalGenerator
+export GravitationalGenerator, EndoROARGenerator
 
 include("experiments/Experiments.jl")
 using .Experiments

### 2022-08-10 src/data/utils.jl

@@ -1,7 +1,6 @@
 # Train-test split
 using StatsBase
 using CounterfactualExplanations
-using CounterfactualExplanations.DataPreprocessing: unpack
 using Flux
 """
     train_test_split(data::CounterfactualData;test_size=0.2)
@@ -9,7 +8,7 @@ using Flux
 Splits data into train and test split.
 """
 function train_test_split(data::CounterfactualData;test_size=0.2)
-    X,y = unpack(data)
+    X,y = CounterfactualExplanations.DataPreprocessing.unpack(data)
     N = size(y,2)
     classes_ = sort(unique(y))
     n_per_class = round(N/length(classes_))

### 2022-08-10 src/generators/EndoROARGenerator.jl

@@ -1,0 +1,83 @@
+using LinearAlgebra, CounterfactualExplanations
+
+mutable struct EndoROARGenerator <: AbstractGradientBasedGenerator
+    loss::Union{Nothing,Symbol} # loss function
+    complexity::Function # complexity function
+    λ::Union{AbstractFloat,AbstractVector} # strength of penalty
+    opt::Any # optimizer
+    τ::AbstractFloat # tolerance for convergence
+end
+
+# API streamlining:
+using Parameters, Flux
+@with_kw struct EndoROARGeneratorParams
+    opt::Any=Flux.Optimise.Descent()
+    τ::AbstractFloat=1e-5
+end
+
+"""
+    EndoROARGenerator(
+        ;
+        loss::Symbol=:logitbinarycrossentropy,
+        complexity::Function=norm,
+        λ::AbstractFloat=0.1,
+        opt::Any=Flux.Optimise.Descent(),
+        τ::AbstractFloat=1e-5
+    )
+
+An outer constructor method that instantiates a generic generator.
+
+# Examples
+```julia-repl
+generator = EndoROARGenerator()
+```
+"""
+function EndoROARGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=[0.1,5.0],kwargs...)
+    params = EndoROARGeneratorParams(;kwargs...)
+    EndoROARGenerator(loss, complexity, λ, params.opt, params.τ)
+end
+
+using Flux
+function gradient_penalty(generator::EndoROARGenerator, counterfactual_state::CounterfactualState.State)
+    
+    x_ = counterfactual_state.f(counterfactual_state.s′)
+    M = counterfactual_state.M
+    nn = M.model
+    y_ = counterfactual_state.y′
+
+    loss_type = M.likelihood == :classification_binary ? :logitbinarycrossentropy : :logitcrossentropy
+    loss(x, y) = getfield(Flux.Losses,loss_type)(nn(x), y)
+
+    return loss(x_,y_)
+end
+
+# Complexity:
+using Statistics, LinearAlgebra
+using CounterfactualExplanations.CounterfactualState
+import CounterfactualExplanations.Generators: h
+"""
+    h(generator::AbstractGenerator, counterfactual_state::CounterfactualState.State)
+
+The default method to apply the generator complexity penalty to the current counterfactual state for any generator.
+"""
[…]

### 2022-08-10 dev/notebooks/synthetic.qmd
blieh

@@
     :DICE=>DiCEGenerator(),
-    :Gravitational=>GravitationalGenerator(),
+    :Gravitational=>GravitationalGenerator(λ=[0.1,0.5]),
 )

### 2022-08-10 dev/notebooks/real_world.qmd
uh

@@
     :DICE=>DiCEGenerator(),
-    :Gravitational=>GravitationalGenerator(λ=[0.1,2]),
 )

### 2022-08-10 dev/notebooks/synthetic.qmd
uh

--- 2022-08-09 dev/notebooks/synthetic.qmd
@@
     :DICE=>DiCEGenerator(),
-    :Gravitational=>GravitationalGenerator(),
 )

### 2022-08-10 dev/notebooks/endo_roar_generator.qmd
all mitigation strategies somewhat effective

@@
       counterfactuals_strict...,
-      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
+      generate_counterfactual(x, target, counterfactual_data, M, generator; T=T)
     )
@@
 )
-counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; convergence=:strict, T=T) for (name, gen) in generators])
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; T=T) for (name, gen) in generators])
 ```
@@
 ```
+
+```{julia}
+
+```

### 2022-08-10 dev/notebooks/mitigation_strategies.qmd
all mitigation strategies somewhat effective

---
title: Mitigation Strategies
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("mitigation_strategies")
www_path = www_dir("mitigation_strategies")
```

```{julia}
max_obs = 1000
catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
choices = [
    :linearly_separable, 
    :overlapping, 
    :circles, 
    :moons,
]
data_sets = filter(p -> p[1] in choices, catalogue)
```

```{julia}
models = [
    :LogisticRegression, 
    :FluxModel, 
    :FluxEnsemble,
]
generators = Dict(
    :Generic=>GenericGenerator(decision_threshold=0.5),
    :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
    :Gravitational=>GravitationalGenerator(),
    :ROAR=>EndoROARGenerator()
)
```

```{julia}
experiments = set_up_experiments(data_sets,models,generators)
```

```{julia}
using AlgorithmicRecourseDynamics.Models: model_evaluation
plts = []
for (exp_name, exp_) in experiments
    for (M_name, M) in exp_.models
        score = round(model_evaluation(M, exp_.test_data),digits=2)
        plt = plot(M, exp_.test_data, title="$exp_name;\n $M_name ($score)")
        # Errors:
        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
        x_wrongly_labelled = exp_.test_data.X[:,ids]
        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
        plts = vcat(plts..., plt)
    end
end
plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
savefig(plt, joinpath(www_path,"models_test_before.png"))
```

```{julia}
using AlgorithmicRecourseDynamics.Models: model_evaluation
plts = []
for (exp_name, exp_) in experiments
    for (M_name, M) in exp_.models
        score = round(model_evaluation(M, exp_.train_data),digits=2)
        plt = plot(M, exp_.train_data, title="$exp_name;\n $M_name ($score)")
        # Errors:
        ids = findall(vec(round.(probs(M, exp_.train_data.X)) .!= exp_.train_data.y))
        x_wrongly_labelled = exp_.train_data.X[:,ids]
        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
        plts = vcat(plts..., plt)
    end
end
plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
savefig(plt, joinpath(www_path,"models_train_before.png"))
```

## Experiments

```{julia}
n_evals = 5
n_rounds = 50
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 5
n_bootstrap = 1
T = 100
using Serialization
results = run_experiments(
    experiments;
    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T
)
Serialization.serialize(joinpath(output_path,"results.jls"),results)
```

```{julia}
using AlgorithmicRecourseDynamics.Models: model_evaluation
plot_dict = Dict(key => Dict() for (key,val) in results)
fold = 1
for (name, res) in results
    exp_ = res.experiment
    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
    rec_sys = exp_.recourse_systems[fold]
    sys_ids = collect(exp_.system_identifiers)
    M = length(rec_sys)
    for m in 1:M
        model_name, generator_name = sys_ids[m]
        M = rec_sys[m].model
        score = round(model_evaluation(M, exp_.test_data),digits=2)
        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
        # Errors:
        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
        x_wrongly_labelled = exp_.test_data.X[:,ids]
        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
    end
end
plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
for (name, plts) in plot_dict
    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
    savefig(plt, joinpath(www_path,"models_test_after_$(name).png"))
end
```

```{julia}
using AlgorithmicRecourseDynamics.Models: model_evaluation
plot_dict = Dict(key => Dict() for (key,val) in results)
fold = 1
for (name, res) in results
    exp_ = res.experiment
    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
    rec_sys = exp_.recourse_systems[fold]
    sys_ids = collect(exp_.system_identifiers)
    M = length(rec_sys)
    for m in 1:M
        model_name, generator_name = sys_ids[m]
        M = rec_sys[m].model
        data = rec_sys[m].data
        score = round(model_evaluation(M, data),digits=2)
        plt = plot(M, data, title="$name;\n $model_name ($score)")
        # Errors:
        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
        x_wrongly_labelled = data.X[:,ids]
        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
    end
end
plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
for (name, plts) in plot_dict
    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
    savefig(plt, joinpath(www_path,"models_train_after_$(name).png"))
end
```

```{julia}
using Serialization
results = Serialization.deserialize(joinpath(output_path,"results.jls"))
```

### Plots

```{julia}
line_charts = Dict()
errorbar_charts = Dict()
for (data_name, res) in results
    line_charts[data_name] = plot(res)
    errorbar_charts[data_name] = plot(res,50)
end
```

### 2022-08-10 src/experiments/functions.jl

@@ -7,7 +7,6 @@ using Parameters
     seed::Union{Nothing, Int} = nothing
     T::Int = 1000
     μ::AbstractFloat = 0.05
-    γ::AbstractFloat = 0.75
     intersect_::Bool = true
     convergence::Symbol = :threshold_only
     generative_model_params::NamedTuple = (;)
@@ -156,15 +155,14 @@ function update!(experiment::Experiment, recourse_system::RecourseSystem, chosen
 
     # Experiment:
     args = experiment.fixed_parameters
-    T, γ = args.T, args.γ
+    T = args.T
     target = experiment.target
 
     # Generate recourse:
     factuals = select_factual(counterfactual_data,chosen_individuals)
     results = generate_counterfactual(
         factuals, target, counterfactual_data, M, generator; 
-        T=T, γ=γ, num_counterfactuals=experiment.num_counterfactuals,
-        convergence=args.convergence, generative_model_params=args.generative_model_params
+        T=T, num_counterfactuals=experiment.num_counterfactuals, generative_model_params=args.generative_model_params
     );
     
     indices_ = rand(1:experiment.num_counterfactuals,length(results)) # randomly draw from generated counterfactuals

### 2022-08-10 src/generators/GravitationalGenerator.jl

@@ -4,6 +4,7 @@ mutable struct GravitationalGenerator <: AbstractGradientBasedGenerator
     loss::Union{Nothing,Symbol} # loss function
     complexity::Function # complexity function
     λ::Union{AbstractFloat,AbstractVector} # strength of penalty
+    decision_threshold::Union{Nothing,AbstractFloat}
     opt::Any # optimizer
     τ::AbstractFloat # tolerance for convergence
     K::Int # number of K randomly chosen neighbours
@@ -36,9 +37,16 @@ An outer constructor method that instantiates a generic generator.
 generator = GravitationalGenerator()
 ```
 """
-function GravitationalGenerator(;loss::Union{Nothing,Symbol}=nothing,complexity::Function=norm,λ::Union{AbstractFloat,AbstractVector}=[0.1,1.0],kwargs...)
+function GravitationalGenerator(
+    ;
+    loss::Union{Nothing,Symbol}=nothing,
+    complexity::Function=norm,
+    λ::Union{AbstractFloat,AbstractVector}=[0.1,5.0],
+    decision_threshold=nothing,
+    kwargs...
+)
     params = GravitationalGeneratorParams(;kwargs...)
-    GravitationalGenerator(loss, complexity, λ, params.opt, params.τ, params.K, params.centroid)
+    GravitationalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ, params.K, params.centroid)
 end
 
 # Complexity:

### 2022-08-11 dev/notebooks/mitigation_strategies.qmd
mitigation strategies also show some effect for real world data. Finally now also trying in combination with revise

@@
 ```{julia}
-max_obs = 1000
-catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
-choices = [
-    :linearly_separable, 
-    :overlapping, 
-    :circles, 
-    :moons,
-]
-data_sets = filter(p -> p[1] in choices, catalogue)
-```
-
-```{julia}
 models = [
@@
 )
+```
+
+## Synthetic
+
+```{julia}
+max_obs = 1000
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
+choices = [
+    :linearly_separable, 
+    :overlapping, 
+    :circles, 
+    :moons,
+]
+data_sets = filter(p -> p[1] in choices, catalogue)
 ```
@@
 ```
-
-## Experiments
 
@@
 ```
+
+## Real World
+
+```{julia}
+max_obs = 1000
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
+```
+
+```{julia}
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 50
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
+```
+
+```{julia}
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
+```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+T = 100
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,
+    evaluate_every=evaluate_every,
+    n_rounds=n_rounds, 
+    n_folds=n_folds, 
+    n_bootstrap=n_bootstrap, 
+    T=T
+)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+```{julia}
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
+end
+```
+
+### Latent Space Search
+
+```{julia}
+generators = Dict(
+    :REVISE=>GenericGenerator(decision_threshold=0.5),
+    :REVISE_conservative=>GenericGenerator(decision_threshold=0.9),
+    :Gravitational=>GravitationalGenerator(),
+    :ROAR=>EndoROARGenerator()
+)
+```
+
+```{julia}
+generative_model_params = (epochs=250, latent_dim=8)
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
+```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+T = 100
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,
+    save_name_suffix="latent",
+    evaluate_every=evaluate_every,
+    n_rounds=n_rounds, 
+    n_folds=n_folds, 
+    n_bootstrap=n_bootstrap, 
+    T=T,
+    latent_space=true,
+    generative_model_params=generative_model_params
+)
+Serialization.serialize(joinpath(output_path,"results_latent.jls"),results)
+```
+
+### Plots
+
+```{julia}
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    line_charts[data_name] = plot(res)
+    errorbar_charts[data_name] = plot(res,50)
+end
+```

### 2022-08-11 dev/notebooks/real_world.qmd
mitigation strategies also show some effect for real world data. Finally now also trying in combination with revise

@@
 n_bootstrap = 1
-T = 250
+T = 100
 generative_model_params = (epochs=250, latent_dim=8)
@@
     save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
-    convergence=:strict, generative_model_params=generative_model_params, save_name_suffix="strict"
+    generative_model_params=generative_model_params
 )
-Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
 ```
@@
 using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_strict.jls"))
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
 ```
@@
 ```
-
-## Simple convergence
-
-```{julia}
-n_evals = 5
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-n_bootstrap = 1
-using Serialization
-results = run_experiments(
-    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap
-)
-Serialization.serialize(joinpath(output_path,"results.jls"),results)
-```
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results.jls"))
-```
-
-### Plots
-
-```{julia}
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
-end
-```

### 2022-08-11 dev/notebooks/endo_roar_generator.qmd
renamed EndoROAR to ClapROAR as in Classifier-preserving ROAR

@@
 
-# `EndoROARGenerator`
+# `ClapROARGenerator`
 
@@
     λ = [0.1, λ₂]
-    generator = EndoROARGenerator(λ=λ)
+    generator = ClapROARGenerator(λ=λ)
     generators = vcat(generators..., generator)
@@
     :Generic => GenericGenerator(),
-    :ROAR => EndoROARGenerator()
+    :ROAR => ClapROARGenerator()
 )

### 2022-08-11 dev/notebooks/mitigation_strategies.qmd
renamed EndoROAR to ClapROAR as in Classifier-preserving ROAR

@@
     :Generic=>GenericGenerator(decision_threshold=0.5),
+    :REVISE=>REVISEGenerator(),
     :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
     :Gravitational=>GravitationalGenerator(),
-    :ROAR=>EndoROARGenerator()
+    :ClapROAR=>ClapROARGenerator()
 )
@@
     :Gravitational=>GravitationalGenerator(),
-    :ROAR=>EndoROARGenerator()
+    :ROAR=>ClapROARGenerator()
 )

### 2022-08-11 src/experiments/functions.jl

@@ -10,6 +10,7 @@ using Parameters
     intersect_::Bool = true
     convergence::Symbol = :threshold_only
     generative_model_params::NamedTuple = (;)
+    latent_space::Union{Nothing, Bool} = nothing
 end
 
 mutable struct Experiment
@@ -162,7 +163,8 @@ function update!(experiment::Experiment, recourse_system::RecourseSystem, chosen
     factuals = select_factual(counterfactual_data,chosen_individuals)
     results = generate_counterfactual(
         factuals, target, counterfactual_data, M, generator; 
-        T=T, num_counterfactuals=experiment.num_counterfactuals, generative_model_params=args.generative_model_params
+        T=T, num_counterfactuals=experiment.num_counterfactuals, generative_model_params=args.generative_model_params,
+        latent_space=args.latent_space
     );
     
     indices_ = rand(1:experiment.num_counterfactuals,length(results)) # randomly draw from generated counterfactuals

### 2022-08-11 src/generators/Generators.jl

@@ -3,7 +3,7 @@ module Generators
 include("GravitationalGenerator.jl")
 export GravitationalGenerator
 
-include("EndoROARGenerator.jl")
-export EndoROARGenerator
+include("ClapROARGenerator.jl")
+export ClapROARGenerator
     
 end

### 2022-08-11 dev/paper/paper.tex
some rejigging of sections

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, sharp corners, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{sec-related}{%
-\section{Related Work}\label{sec-related}}
+\section{Background}\label{sec-related}}
 
@@
 \hypertarget{sec-method}{%
-\section{Methodology}\label{sec-method}}
+\section{Gradient-Based Recourse Revisited}\label{sec-method}}
 
@@
 in our experiments. We then describe the experimental setup in
-Section~\ref{sec-method-experiment} and introduce several evaluation
+Section~\ref{sec-method-2-experiment} and introduce several evaluation
 metrics used to benchmark the different generators.
@@
 \hypertarget{sec-method-general}{%
-\subsection{A Generalized Framework for Gradient-Based Counterfactual
-Search}\label{sec-method-general}}
+\subsection{From individual recourse
+\ldots{}}\label{sec-method-general}}
 
@@
 search is well-suited for differentiable black-box models like deep
-neural networks.
-
-\hypertarget{from-individual-recourse}{%
-\subsubsection{From individual recourse
-\ldots{}}\label{from-individual-recourse}}
-
-We can restate Equation~\ref{eq-solution} in a more general form that
-encompasses most gradient-based approaches to counterfactual search:
+neural networks. We can restate Equation~\ref{eq-solution} in a more
+general form that encompasses most gradient-based approaches to
+counterfactual search:
 
@@
 \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
-state space to the feature space. In the case of the baseline
-counterfactual generator
+state space to the feature space.
+
+\hypertarget{towards-collective-recourse}{%
+\subsection{\ldots{} towards collective
+recourse}\label{towards-collective-recourse}}
+
+To explicitly address the notion that individual recourse may affect the
+outcome and prospect of other individuals, we propose to extend
+Equation~\ref{eq-general} as follows:
+
+\begin{equation}\protect\hypertarget{eq-collective}{}{
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
+\end{aligned}
+}\label{eq-collective}\end{equation}
+
+Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
+the individual as before and \(\lambda_1\) governs to what extent that
+private cost ought to be penalised. The newly introduced term
+\(g(f(s_k^\prime))\) is meant to capture and address social costs
+incurred by the collective of individuals in response to changes in
+\(\mathbf{s}^\prime\). The trade-off between private and social costs is
+determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
+individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
+nor do we intend to provide a definite answer in this work, if such even
+exists. A straight-forward choice simply extends the baseline approach
+by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
+of only penalizing the distance of the individuals' counterfactual to
+its factual, we propose penalizing its distance to some sensible point
+in the target domain, for example the sample average:
+\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
+\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
+gravitate towards the specified point in the target domain. In the
+remainder of this paper we will therefore refer to this approach as
+\textbf{Gravitational} generator, when we investigate its potential
+usefulness for mitigating endongenous macrodynamics\footnote{Note that
+  despite the naming convention our goal here is not to provide yet
+  another counterfactual generator, but merely investigate the most
+  simple penalty we can think of with respect to its effectiveness.}.
+
+\hypertarget{a-note-on-convergence}{%
+\subsection{A note on convergence}\label{a-note-on-convergence}}
+
+For this simple mitigating strategy underlying the Gravitational
+generator to work as expected, one needs to ensure that counterfactual
+search continues, even after a predetermined threshold probability
+\(\gamma\) has potentially already been reached.
+Figure~\ref{fig-convergence} illustrates this distinction: if one
+chooses to terminate search once the desired threshold is reached (left
+panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
+actually satisfied (compare to right panel). More generally, if
+convergence is defined simply in terms of flipping the predicted label
+with some desired degree of confidence, this corresponds to essentially
+ignoring any parts of the counterfactual search objective that do not
+involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
+be appropriate for some applications, in general this seems like an odd
+convention. Since we nonetheless seen convergence specified simply in
+terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
+  define convergence of Algorithm 1 in this way. The implementation of
+  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
+  also defined in this way.}, we thought it worth making this
+distinction explicit.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}
+
+}
+
+\caption{\label{fig-convergence}Comparison of counterfactual search
+outcome with simple (left) and strict convergence (right).}
+
+\end{figure}
+
+\hypertarget{sec-method-2}{%
+\section{Modeling Endogenous Macrodynamics in Algorithmic
+Recourse}\label{sec-method-2}}
+
+In this \ldots{}
+
+\hypertarget{sec-method-2-experiment}{%
+\subsection{Simulations}\label{sec-method-2-experiment}}
+
+The dynamics illustrated in Figure~\ref{fig-poc} in
+Section~\ref{sec-intro} were generated through a simple experiment that
+aims to simulate the process of algorithmic recourse in practice. We
+begin in the static setting at time time \(t=0\): given some pre-trained
+classifier \(M\) we generate recourse for a random batch of \(B\)
+individuals in the non-target class. Note that we focus our attention on
+classification problems, since classification poses the most common
+practical use-case for algorithmic recourse. In order to simulate the
+dynamical process we suppose that the model \(M\) is retrained following
+the actual implementation of recourse in time \(t=0\). Following the
+update to the model, we assume that at time \(t=1\) recourse is
+generated for yet another random subset of individuals in the non-target
+class. This process is repeated for a number of time periods \(T\). To
+get a clean read on endogenous dynamics we keep the total population of
+samples closed: we allow existing samples to move from factual to
+counterfactual states, but do not allow any entirely new samples to
+enter the population. The experimental setup is summarized in Algorithm
+\ref{algo-experiment}
+
+\begin{algorithm}
+\caption{Experiment}\label{algo-experiment}
+\begin{algorithmic}[1]
+\Procedure{Experiment}{$M,D,G$}
+\State $t\gets 0$
+\While{$t<T$}
+\State $D_B \subset D$
+\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
+\State $M\gets M(D)$ \Comment{Retrain model.}
+\EndWhile
+\State \textbf{return} $M,D$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+A noteworthy practical consideration is the choice of \(T\) and \(B\).
+The higher these values, the more factual instances undergo recourse
+throughout the entire experiment. Of course, this is likely to lead to
+more pronounced domain and model shifts by time \(T\). At the same time,
+it is generally improbable that a very large part of the population
+would request an explanation of the algorithm's decisions. In our
+experiments, we choose the values such that \(T \cdot B\) corresponds to
+the application of recourse on \(25-50\%\) of the negative instances
+from the initial dataset. As we collect data at each time \(t\), we can
+also verify the impact of recourse when it is implemented for a smaller
+number of individuals. Using our framework the experiment can be
+conducted on an arbitrary number of algorithmic recourse generators. As
+all generators make use of the same initial model and initial dataset,
+the differences in domain and model shifts observed throughout the
+rounds depend solely on the employed generator.
+
+\hypertarget{sec-method-2-metrics}{%
+\subsection{Evaluation Metrics}\label{sec-method-2-metrics}}
+
+We formulate two desiderata for the set of metrics used to measure
+domain and model shifts induced by recourse. First, the metrics should
+be applicable regardless of the dataset or classification technique so
+that they allow for the meaningful comparison of the generators in
+various scenarios. As the knowledge of the underlying probability
+distribution is rarely available, the metrics should be empirical and
+non-parametric. This further ensures that we can also measure large
+datasets by sampling from the available data. Moreover, while our study
+was conducted in a two-class classification setting, our choice of
+metrics should remain applicable in the future research on multi- class
+recourse problems. Second, the set of metrics should allow to capture
+various aspects of the previously mentioned magnitude, path, and tempo
+of changes while remaining as small as possible.
+
+\hypertarget{domain-shifts}{%
+\subsubsection{Domain Shifts}\label{domain-shifts}}
+
+To quantify the magnitude of domain shifts we rely on an unbiased
+estimate of the squared population \textbf{Maximum Mean Discrepancy
+(MMD)} given as:
+
+\begin{equation}\protect\hypertarget{eq-mmd}{}{
+\begin{aligned}
+MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+}\label{eq-mmd}\end{equation}
+
+where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
+Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
+identically distributed samples drawn from probability distributions
+\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
+between the kernel mean embeddings of \(p\) and \(q\) in RKHS
+\(\mathcal{H}\). An important consideration is the choice of the kernel
+function \(k(\cdot,\cdot)\). In our implementation we make use of the
+radial basis function (RBF) kernel with a constant length-scale
+parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
+and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
+\(X=\tilde{X}\).
+
+The evaluation metric in Equation~\ref{eq-mmd} is computed after every
+round \(t=1,...,T\) of the experiment. To assess the statistical
+significance of the observed shifts under the null hypothesis that
+samples \(X\) and \(\tilde{X}\) were drawn from the same probability
+distribution we follow
+\protect\hyperlink{ref-arcones1992bootstrap}{{[}24{]}}. To that end, we
+combine the two samples and generate a large number of permutations of
+\(X + \tilde{X}\). Then, we split the permuted data into two new samples
+\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
+original samples. Then under the null hypothesis we should have that
+\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
+\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
+calculated by counting how these two quantities are not equal.
+
+We calculate the MMD for both classes individually based on the ground
+truth labels, i.e.~the labels that samples were assigned in time
+\(t=0\). Throughout our experiments we generally do not expect the
+distribution of the negative class to change over time -- application of
+recourse reduces the size of this class, but since individuals are
+sampled uniformly the distribution should remain unaffected. Conversely,
+unless a recourse generator can perfectly replicate the original
+probability distribution, we expect the MMD of the positive class to
+increase. Thus, when discussing MMD, we generally mean the shift in the
+distribution of the positive class.
+
+Finally, \textbf{feature mean and feature standard deviation} are also
+calculated to verify how the implementation of recourse impacts every
+attribute in the dataset. Although MMD already captures information
+about the expected value and variance, we may also be interested in a
+more granular look at individual features.
+
+\hypertarget{model-shifts}{%
+\subsubsection{Model Shifts}\label{model-shifts}}
+
+As our baseline for quantifying model shifts we measure perturbations to
+the model parameters at each point in time \(t\) following
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
+\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
+between the vectors of parameters before and after retraining the model
+\(M\). We shall refer to this baseline metric simply as
+\textbf{Perturbations}.
+
+We extend the metric in Equation~\ref{eq-mmd} for the purpose of
+quantifying model shifts. Specifically, we introduce \textbf{Predicted
+Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
+features directly, we apply it to the predicted probabilities assigned
+to a set of samples by the model \(M\). If the model shifts, the
+probabilities assigned to each sample will change; again, this metric
+will equal 0 only if the two classifiers are the same. It is worth
+noting that while we apply the technique to samples drawn uniformly from
+the dataset, it can also be employed on arbitrary points in the entire
+feature space (or a subspace). The latter approach is theoretically more
+robust. Unfortunately, in practice this approachs suffers from the curse
+of dimensionality, since it becomes increasingly difficult to select
+enough points to overcome noise as the dimension \(D\) grows.
+
+As an alternative to PP MMD we use a pseudo-distance for the
+\textbf{Disagreement Coefficient} (Disagreement). This metric was
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}25{]}} and
+estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
+classifiers do not agree on the predicted outcome for a randomly chosen
+sample. Thus, it is not relevant whether the classification is correct
+according to the ground truth, but only whether the sample lies on the
+same side of the two respective decision boundaries. In our context,
+this metric quantifies the overlap between the initial model (trained
+before the application of recourse) and the updated model. A
+Disagreement Coefficient unequal to zero is indicative of a model shift.
+The opposite is not true: even if the Disagreement Coefficient is equal
+to zero a model shift may still have occured. This is one reason for why
+PP MMD is our our preferred metric.
+
+Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
+the likelihood that a model assigns a high probability to its
+classification of any given sample. We define the metric simply as
+\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
+predicted logits from a binary classifier and \(\sigma\) denotes the
+sigmoid function. This metric provides an unbiased estimate of the
+binary classifier's tendency to produce high-confidence predictions in
+either one of the two classes. Although the exact values for this metric
+are not important for our study, they can be used to detect model
+shifts. If decisiveness changes over time, then this is indicative of
+the decision boundary moves towards either one of the two classes.
+
+\hypertarget{sec-empirical}{%
+\section{Experiment Setup}\label{sec-empirical}}
+
+\hypertarget{sec-empirical-recourse}{%
+\subsection{Recourse Generators}\label{sec-empirical-recourse}}
+
+In the case of the baseline counterfactual generator
 \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
@@
 
-\hypertarget{towards-collective-recourse}{%
-\subsubsection{\ldots{} towards collective
-recourse}\label{towards-collective-recourse}}
-
-To explicitly address the notion that individual recourse may affect the
-outcome and prospect of other individuals, we propose to extend
-Equation~\ref{eq-general} as follows:
-
-\begin{equation}\protect\hypertarget{eq-collective}{}{
-\begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
-\end{aligned}
-}\label{eq-collective}\end{equation}
-
-Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
-the individual as before and \(\lambda_1\) governs to what extent that
-private cost ought to be penalised. The newly introduced term
-\(g(f(s_k^\prime))\) is meant to capture and address social costs
-incurred by the collective of individuals in response to changes in
-\(\mathbf{s}^\prime\). The trade-off between private and social costs is
-determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
-individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
-nor do we intend to provide a definite answer in this work, if such even
-exists. A straight-forward choice simply extends the baseline approach
-by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
-of only penalizing the distance of the individuals' counterfactual to
-its factual, we propose penalizing its distance to some sensible point
-in the target domain, for example the sample average:
-\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
-\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
-gravitate towards the specified point in the target domain. In the
-remainder of this paper we will therefore refer to this approach as
-\textbf{Gravitational} generator, when we investigate its potential
-usefulness for mitigating endongenous macrodynamics\footnote{Note that
-  despite the naming convention our goal here is not to provide yet
-  another counterfactual generator, but merely investigate the most
-  simple penalty we can think of with respect to its effectiveness.}.
-
-\hypertarget{a-note-on-convergence}{%
-\subsubsection{A note on convergence}\label{a-note-on-convergence}}
-
-For this simple mitigating strategy underlying the Gravitational
-generator to work as expected, one needs to ensure that counterfactual
-search continues, even after a predetermined threshold probability
-\(\gamma\) has potentially already been reached.
-Figure~\ref{fig-convergence} illustrates this distinction: if one
-chooses to terminate search once the desired threshold is reached (left
-panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
-actually satisfied (compare to right panel). More generally, if
-convergence is defined simply in terms of flipping the predicted label
-with some desired degree of confidence, this corresponds to essentially
-ignoring any parts of the counterfactual search objective that do not
-involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
-be appropriate for some applications, in general this seems like an odd
-convention. Since we nonetheless seen convergence specified simply in
-terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
-  define convergence of Algorithm 1 in this way. The implementation of
-  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
-  also defined in this way.}, we thought it worth making this
-distinction explicit.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}
-
-}
-
-\caption{\label{fig-convergence}Comparison of counterfactual search
-outcome with simple (left) and strict convergence (right).}
-
-\end{figure}
-
-\hypertarget{sec-method-experiment}{%
-\subsection{Experimental Setup}\label{sec-method-experiment}}
-
-The dynamics illustrated in Figure~\ref{fig-poc} in
-Section~\ref{sec-intro} were generated through a simple experiment that
-aims to simulate the process of algorithmic recourse in practice. We
-begin in the static setting at time time \(t=0\): given some pre-trained
-classifier \(M\) we generate recourse for a random batch of \(B\)
-individuals in the non-target class. Note that we focus our attention on
-classification problems, since classification poses the most common
-practical use-case for algorithmic recourse. In order to simulate the
-dynamical process we suppose that the model \(M\) is retrained following
-the actual implementation of recourse in time \(t=0\). Following the
-update to the model, we assume that at time \(t=1\) recourse is
-generated for yet another random subset of individuals in the non-target
-class. This process is repeated for a number of time periods \(T\). To
-get a clean read on endogenous dynamics we keep the total population of
-samples closed: we allow existing samples to move from factual to
-counterfactual states, but do not allow any entirely new samples to
-enter the population. The experimental setup is summarized in Algorithm
-\ref{algo-experiment}
-
-\begin{algorithm}
-\caption{Experiment}\label{algo-experiment}
-\begin{algorithmic}[1]
-\Procedure{Experiment}{$M,D,G$}
-\State $t\gets 0$
-\While{$t<T$}
-\State $D_B \subset D$
-\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
-\State $M\gets M(D)$ \Comment{Retrain model.}
-\EndWhile
-\State \textbf{return} $M,D$
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-A noteworthy practical consideration is the choice of \(T\) and \(B\).
-The higher these values, the more factual instances undergo recourse
-throughout the entire experiment. Of course, this is likely to lead to
-more pronounced domain and model shifts by time \(T\). At the same time,
-it is generally improbable that a very large part of the population
-would request an explanation of the algorithm's decisions. In our
-experiments, we choose the values such that \(T \cdot B\) corresponds to
-the application of recourse on \(25-50\%\) of the negative instances
-from the initial dataset. As we collect data at each time \(t\), we can
-also verify the impact of recourse when it is implemented for a smaller
-number of individuals. Using our framework the experiment can be
-conducted on an arbitrary number of algorithmic recourse generators. As
-all generators make use of the same initial model and initial dataset,
-the differences in domain and model shifts observed throughout the
-rounds depend solely on the employed generator.
-
-\hypertarget{sec-method-data}{%
-\subsection{Data}\label{sec-method-data}}
+\hypertarget{sec-empirical-data}{%
+\subsection{Data}\label{sec-empirical-data}}
 
@@
 experience financial difficulties in the next two years
-\protect\hyperlink{ref-gmsc_data}{{[}24{]}}. Originally consisting of
+\protect\hyperlink{ref-gmsc_data}{{[}26{]}}. Originally consisting of
 250,000 instances with 11 numerical attributes, the dataset was randomly
@@
 task of predicting if bank customers are credit-worthy or not
-\protect\hyperlink{ref-germancredit1994}{{[}25{]}}. It consists of 700
+\protect\hyperlink{ref-germancredit1994}{{[}27{]}}. It consists of 700
 positive and 300 negative instances charaterized by 7 numerical and 13
@@
 
-\hypertarget{sec-method-metrics}{%
-\subsection{Evaluation Metrics}\label{sec-method-metrics}}
-
-We formulate two desiderata for the set of metrics used to measure
-domain and model shifts induced by recourse. First, the metrics should
-be applicable regardless of the dataset or classification technique so
-that they allow for the meaningful comparison of the generators in
-various scenarios. As the knowledge of the underlying probability
-distribution is rarely available, the metrics should be empirical and
-non-parametric. This further ensures that we can also measure large
-datasets by sampling from the available data. Moreover, while our study
-was conducted in a two-class classification setting, our choice of
-metrics should remain applicable in the future research on multi- class
-recourse problems. Second, the set of metrics should allow to capture
-various aspects of the previously mentioned magnitude, path, and tempo
-of changes while remaining as small as possible.
-
-\hypertarget{domain-shifts}{%
-\subsubsection{Domain Shifts}\label{domain-shifts}}
-
-To quantify the magnitude of domain shifts we rely on an unbiased
-estimate of the squared population \textbf{Maximum Mean Discrepancy
-(MMD)} given as:
-
-\begin{equation}\protect\hypertarget{eq-mmd}{}{
-\begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
-\end{aligned}
-}\label{eq-mmd}\end{equation}
-
-where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
-Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
-identically distributed samples drawn from probability distributions
-\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
-between the kernel mean embeddings of \(p\) and \(q\) in RKHS
-\(\mathcal{H}\). An important consideration is the choice of the kernel
-function \(k(\cdot,\cdot)\). In our implementation we make use of the
-radial basis function (RBF) kernel with a constant length-scale
-parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
-and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
-\(X=\tilde{X}\).
-
-The evaluation metric in Equation~\ref{eq-mmd} is computed after every
-round \(t=1,...,T\) of the experiment. To assess the statistical
-significance of the observed shifts under the null hypothesis that
-samples \(X\) and \(\tilde{X}\) were drawn from the same probability
-distribution we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
-combine the two samples and generate a large number of permutations of
-\(X + \tilde{X}\). Then, we split the permuted data into two new samples
-\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
-original samples. Then under the null hypothesis we should have that
-\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
-\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
-calculated by counting how these two quantities are not equal.
-
-We calculate the MMD for both classes individually based on the ground
-truth labels, i.e.~the labels that samples were assigned in time
-\(t=0\). Throughout our experiments we generally do not expect the
-distribution of the negative class to change over time -- application of
-recourse reduces the size of this class, but since individuals are
-sampled uniformly the distribution should remain unaffected. Conversely,
-unless a recourse generator can perfectly replicate the original
-probability distribution, we expect the MMD of the positive class to
-increase. Thus, when discussing MMD, we generally mean the shift in the
-distribution of the positive class.
-
-Finally, \textbf{feature mean and feature standard deviation} are also
-calculated to verify how the implementation of recourse impacts every
-attribute in the dataset. Although MMD already captures information
-about the expected value and variance, we may also be interested in a
-more granular look at individual features.
-
-\hypertarget{model-shifts}{%
-\subsubsection{Model Shifts}\label{model-shifts}}
-
-As our baseline for quantifying model shifts we measure perturbations to
-the model parameters at each point in time \(t\) following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
-\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
-between the vectors of parameters before and after retraining the model
-\(M\). We shall refer to this baseline metric simply as
-\textbf{Perturbations}.
-
-We extend the metric in Equation~\ref{eq-mmd} for the purpose of
-quantifying model shifts. Specifically, we introduce \textbf{Predicted
-Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
-features directly, we apply it to the predicted probabilities assigned
-to a set of samples by the model \(M\). If the model shifts, the
-probabilities assigned to each sample will change; again, this metric
-will equal 0 only if the two classifiers are the same. It is worth
-noting that while we apply the technique to samples drawn uniformly from
-the dataset, it can also be employed on arbitrary points in the entire
-feature space (or a subspace). The latter approach is theoretically more
-robust. Unfortunately, in practice this approachs suffers from the curse
-of dimensionality, since it becomes increasingly difficult to select
-enough points to overcome noise as the dimension \(D\) grows.
-
-As an alternative to PP MMD we use a pseudo-distance for the
-\textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
-estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
-classifiers do not agree on the predicted outcome for a randomly chosen
-sample. Thus, it is not relevant whether the classification is correct
-according to the ground truth, but only whether the sample lies on the
-same side of the two respective decision boundaries. In our context,
-this metric quantifies the overlap between the initial model (trained
-before the application of recourse) and the updated model. A
-Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal
-to zero a model shift may still have occured. This is one reason for why
-PP MMD is our our preferred metric.
-
-Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
-the likelihood that a model assigns a high probability to its
-classification of any given sample. We define the metric simply as
-\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
-predicted logits from a binary classifier and \(\sigma\) denotes the
-sigmoid function. This metric provides an unbiased estimate of the
-binary classifier's tendency to produce high-confidence predictions in
-either one of the two classes. Although the exact values for this metric
-are not important for our study, they can be used to detect model
-shifts. If decisiveness changes over time, then this is indicative of
-the decision boundary moves towards either one of the two classes.
-
-\hypertarget{sec-empirical}{%
-\section{Experiments}\label{sec-empirical}}
+\hypertarget{sec-empirical-2}{%
+\section{Experiments}\label{sec-empirical-2}}
+
+\hypertarget{endogenous-macrodynamics}{%
+\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
+
+\hypertarget{potential-mitigation-strategies}{%
+\subsection{Potential Mitigation
+Strategies}\label{potential-mitigation-strategies}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
+\CSLLeftMargin{{[}24{]} }%
+\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
+v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
+\CSLLeftMargin{{[}25{]} }%
+\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
+agnostic active learning,''} in \emph{Proceedings of the 24th
+international conference on machine learning}, 2007, pp. 353--360.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}24{]} }%
+\CSLLeftMargin{{[}26{]} }%
 \CSLRightInline{K. Competition, {``Give me some credit, improve on the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
-\CSLLeftMargin{{[}25{]} }%
+\CSLLeftMargin{{[}27{]} }%
 \CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
 \url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}26{]} }%
-\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
-v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}27{]} }%
-\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
-agnostic active learning,''} in \emph{Proceedings of the 24th
-international conference on machine learning}, 2007, pp. 353--360.}
 

### 2022-08-11 dev/paper/sections/empirical.qmd
some rejigging of sections

@@
-# Experiments {#sec-empirical}
+# Experiment Setup {#sec-empirical}
 
-1.
-Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
-2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
-3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
-    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
-4. Mitigating strategy: penaliye distance from centroid.
+## Recourse Generators {#sec-empirical-recourse}
+
+In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one.
+This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared.
+In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
+
+Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly.
+The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
+In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space.
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
+CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$.
+To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
+The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
+The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals.
+To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ...
+, s_K$ look as different from each other as possible.
+
+Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
+We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro.
+This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
+All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
+
+## Data {#sec-empirical-data}
+
+We have chosen to work with both synthetic and real-world datasets.
+Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
+Following @upadhyay2021towards we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
+Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
+
+### Synthetic data
+
+```{julia}
+using Plots, PlotThemes
+theme(:wong)
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
+function plot_data(data,title)
+    plt = plot(title=title)
+    scatter!(data)
+    return plt
+end
+plts = [plot_data(data,name) for (name, data) in catalogue]
+plt = plot(plts..., layout=(1,4), size=(850,200))
+savefig(plt, "dev/paper/www/synthetic_data.png")
+```
+
+We use four synthetic binary classification datasets consisting of 1000 samples each.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
+Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
+
+![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}
+
+Ex-ante we expect to see that Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
+Thus, the choice of a black-box model may have an impact on the paths of the recourse.
+For generators that use latent space search (@joshi2019towards, @antoran2020getting) or rely on (and have access to) probabilistic models (@antoran2020getting, @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class.
+
+### Real-world data
+
+Additionally, we use two real-world datasets from the Finance domain.
+Firstly, we use the Give Me Some Credit dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data].
+Originally consisting of 250,000 instances with 11 numerical attributes, the dataset was randomly undersampled to result in a balanced subsample made up of 3000 individuals.
+Secondly, we the German Credit dataset which involves the task of predicting if bank customers are credit-worthy or not [@germancredit1994].
+It consists of 700 positive and 300 negative instances charaterized by 7 numerical and 13 categorical attributes.
+We process the dataset in two ways: (1) the values of the "Personal status and sex" feature are aggregated by the two represented genders; (2) the most common values are calculated for all categorical features such that a feature $x_d$ with the mode $\bar{x}_d$ is transformed into a new binary feature $\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}$.
+Binarization ensures that we can use all counterfactual generators in the benchmark.

### 2022-08-11 dev/paper/sections/empirical_2.qmd
some rejigging of sections

# Experiments {#sec-empirical-2}

## Endogenous Macrodynamics

## Potential Mitigation Strategies

### 2022-08-11 dev/paper/sections/methodology.qmd
some rejigging of sections

--- 2022-08-09 dev/paper/sections/methodology.qmd
@@
-# Methodology {#sec-method}
+# Gradient-Based Recourse Revisited {#sec-method}
 
 In the following we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general to introduce the various counterfactual generators we have chosen to use in our experiments.
-We then describe the experimental setup in @sec-method-experiment and introduce several evaluation metrics used to benchmark the different generators.
+We then describe the experimental setup in @sec-method-2-experiment and introduce several evaluation metrics used to benchmark the different generators.
 
-## A Generalized Framework for Gradient-Based Counterfactual Search {#sec-method-general}
+## From individual recourse ... {#sec-method-general}
 
@@
 Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks.
-
-### From individual recourse ...
-
 We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
@@
 Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
-In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one.
-This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared.
-In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
 
-Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly.
-The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
-In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
-CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$.
-To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
-The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
-The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
-Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals.
-To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ...
-, s_K$ look as different from each other as possible.
-
-Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
-We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro.
-This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
-All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
-
-### ... towards collective recourse
+## ... towards collective recourse
 
@@
 
-### A note on convergence
+## A note on convergence
 
@@
 ![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}
-
-## Experimental Setup {#sec-method-experiment}
-
-The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice.
-We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class.
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse.
-In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$.
-Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class.
-This process is repeated for a number of time periods $T$.
-To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population.
-The experimental setup is summarized in Algorithm \ref{algo-experiment}
-
-\begin{algorithm}
-\caption{Experiment}\label{algo-experiment}
-\begin{algorithmic}[1]
-\Procedure{Experiment}{$M,D,G$}
-\State $t\gets 0$
-\While{$t<T$}
-\State $D_B \subset D$
-\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
-\State $M\gets M(D)$ \Comment{Retrain model.}
-\EndWhile
-\State \textbf{return} $M,D$
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-A noteworthy practical consideration is the choice of $T$ and $B$.
-The higher these values, the more factual instances undergo recourse throughout the entire experiment.
-Of course, this is likely to lead to more pronounced domain and model shifts by time $T$.
-At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions.
-In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $25-50\%$ of the negative instances from the initial dataset.
-As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
-Using our framework the experiment can be conducted on an arbitrary number of algorithmic recourse generators.
-As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.
-
-## Data {#sec-method-data}
-
-We have chosen to work with both synthetic and real-world datasets.
-Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
-Following @upadhyay2021towards we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
-Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
-
-### Synthetic data
-
-We use four synthetic binary classification datasets consisting of 1000 samples each.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
-Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
-
-![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/placeholder.png){#fig-synthetic-data fig.pos="h" width="8cm" height="8cm"}
-
-Ex-ante we expect to see that Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
-Thus, the choice of a black-box model may have an impact on the paths of the recourse.
-For generators that use latent space search (@joshi2019towards, @antoran2020getting) or rely on (and have access to) probabilistic models (@antoran2020getting, @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
-Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class.
-
-### Real-world data
-
-Additionally, we use two real-world datasets from the Finance domain.
-Firstly, we use the Give Me Some Credit dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data].
-Originally consisting of 250,000 instances with 11 numerical attributes, the dataset was randomly undersampled to result in a balanced subsample made up of 3000 individuals.
-Secondly, we the German Credit dataset which involves the task of predicting if bank customers are credit-worthy or not [@germancredit1994].
-It consists of 700 positive and 300 negative instances charaterized by 7 numerical and 13 categorical attributes.
-We process the dataset in two ways: (1) the values of the "Personal status and sex" feature are aggregated by the two represented genders; (2) the most common values are calculated for all categorical features such that a feature $x_d$ with the mode $\bar{x}_d$ is transformed into a new binary feature $\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}$.
-Binarization ensures that we can use all counterfactual generators in the benchmark.
-
-## Evaluation Metrics {#sec-method-metrics}
-
-We formulate two desiderata for the set of metrics used to measure domain and model shifts induced by recourse.
-First, the metrics should be applicable regardless of the dataset or classification technique so that they allow for the meaningful comparison of the generators in various scenarios.
-As the knowledge of the underlying probability distribution is rarely available, the metrics should be empirical and non-parametric.
-This further ensures that we can also measure large datasets by sampling from the available data.
-Moreover, while our study was conducted in a two-class classification setting, our choice of metrics should remain applicable in the future research on multi- class recourse problems.
-Second, the set of metrics should allow to capture various aspects of the previously mentioned magnitude, path, and tempo of changes while remaining as small as possible.
-
-### Domain Shifts
-
-To quantify the magnitude of domain shifts we rely on an unbiased estimate of the squared population **Maximum Mean Discrepancy (MMD)** given as:
-
-$$
-\begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
-\end{aligned}
-$$ {#eq-mmd}
-
-where $\mathcal{F}$ is a unit ball in a Reproducing Kernel Hilbert Space H [27], and $X$, $\tilde{X}$ represent independent and identically distributed samples drawn from probability distributions $p$ and $q$ respectively [28].
-MMD is a measure of the distance between the kernel mean embeddings of $p$ and $q$ in RKHS $\mathcal{H}$.
-An important consideration is the choice of the kernel function $k(\cdot,\cdot)$.
-In our implementation we make use of the radial basis function (RBF) kernel with a constant length-scale parameter of $0.5$.
-As RBF captures all moments of distributions $p$ and $q$, we have that $MMD_u^2[F,X,\tilde{X}]=0$ if and only if $X=\tilde{X}$.
-
-The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment.
-To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution we follow @arcones1992bootstrap.
-To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$.
-Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples.
-Then under the null hypothesis we should have that $MMD_u^2[F,X^\prime,\tilde{X}^\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$.
-The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.
-
-We calculate the MMD for both classes individually based on the ground truth labels, i.e.
-the labels that samples were assigned in time $t=0$.
-Throughout our experiments we generally do not expect the distribution of the negative class to change over time – application of recourse reduces the size of this class, but since individuals are sampled uniformly the distribution should remain unaffected.
-Conversely, unless a recourse generator can perfectly replicate the original probability distribution, we expect the MMD of the positive class to increase.
-Thus, when discussing MMD, we generally mean the shift in the distribution of the positive class.
-
-Finally, **feature mean and feature standard deviation** are also calculated to verify how the implementation of recourse impacts every attribute in the dataset.
-Although MMD already captures information about the expected value and variance, we may also be interested in a more granular look at individual features.
-
-### Model Shifts
-
-As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021towards.
-We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$.
-We shall refer to this baseline metric simply as **Perturbations**.
-
-We extend the metric in @eq-mmd for the purpose of quantifying model shifts.
-Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying @eq-mmd to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$.
-If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same.
-It is worth noting that while we apply the technique to samples drawn uniformly from the dataset, it can also be employed on arbitrary points in the entire feature space (or a subspace).
-The latter approach is theoretically more robust.
-Unfortunately, in practice this approachs suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.
-
-As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement).
-This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
-Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
-In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model.
-A Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured.
-This is one reason for why PP MMD is our our preferred metric.
-
-Finally, we introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample.
-We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function.
-This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes.
-Although the exact values for this metric are not important for our study, they can be used to detect model shifts.
-If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.

### 2022-08-11 dev/paper/sections/methodology_2.qmd
some rejigging of sections

--- 2022-07-12 paper/sections/methodology.qmd
@@
-# Methodology {#sec-method}
+# Modeling Endogenous Macrodynamics in Algorithmic Recourse {#sec-method-2}
 
-In the following we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general to introduce the various counterfactual generators we have chosen to use in our experiments.
-We then describe the experimental setup in @sec-method-experiment and introduce several evaluation metrics used to benchmark the different generators.
+In this ...
 
-## A Generalized Framework for Gradient-Based Counterfactual Search {#sec-method-general}
-
-In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro.
-Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks.
-We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
-
-$$
-\begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
-\end{aligned}
-$$ {#eq-general}
-
-Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
-In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one.
-This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared.
-In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
-
-Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly.
-The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
-In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
-CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$.
-To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
-The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
-The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
-Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals.
-To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ...
-, s_K$ look as different from each other as possible.
-
-Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
-We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro.
-This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
-All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
-
-## Experimental Setup {#sec-method-experiment}
+## Simulations {#sec-method-2-experiment}
 
@@
 
-## Data {#sec-method-data}
-
-We have chosen to work with both synthetic and real-world datasets.
-Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
-Following @upadhyay2021towards we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
-Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
-
-### Synthetic data
-
-We use 6 synthetic binary classification datasets consisting of 200-400 samples grouped in normally-distributed clusters.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
-Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
-
-![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}
-
-Ex-ante we expect to see that Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
-Thus, the choice of a black-box model may have an impact on the paths of the recourse.
-For generators that use latent space search (@joshi2019towards, @antoran2020getting) or rely on (and have access to) probabilistic models (@antoran2020getting, @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
-Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class.
-
-### Real-world data
-
-Additionally, we use two real-world datasets from the Finance domain.
-Firstly, we use the Give Me Some Credit dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data].
-Originally consisting of 250,000 instances with 11 numerical attributes, the dataset was randomly undersampled to result in a balanced subsample made up of 3000 individuals.
-Secondly, we the German Credit dataset which involves the task of predicting if bank customers are credit-worthy or not [@germancredit1994].
-It consists of 700 positive and 300 negative instances charaterized by 7 numerical and 13 categorical attributes.
-We process the dataset in two ways: (1) the values of the "Personal status and sex" feature are aggregated by the two represented genders; (2) the most common values are calculated for all categorical features such that a feature $x_d$ with the mode $\bar{x}_d$ is transformed into a new binary feature $\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}$.
-Binarization ensures that we can use all counterfactual generators in the benchmark.
-
-## Evaluation Metrics {#sec-method-metrics}
+## Evaluation Metrics {#sec-method-2-metrics}
 

### 2022-08-11 dev/paper/sections/abstract.qmd
prepared artifacts
progress progress progress

--- 2022-07-05 paper/sections/abstract.qmd
@@
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting: given some classifier we are interested in finding close, actionable, realistic, sparse, diverse and ideally causally founded counterfactuals.
-The ability of CE to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
-Only one recent work considers the implications of exogenous domain and model shifts.
-This project instead focuses on endogenous dynamics, that is shifts that occur when AR is actually implemented by a proportion of individuals.
-Early findings suggest that the involved shifts may be large with important implications on the validity of AR and the overall characteristics of the sample population.
+# Abstract
+
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
+Through this work we aim to close that gap by systematizing and extending existing knowledge.
+We first show that many of the existing methodologies can be collectively described by a generalized framework.
+We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts.
+We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice.
+Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-08-11 dev/paper/sections/methodology_2.qmd
progress progress progress

@@
 
-In this ...
+In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
+We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of `AlgorithmicRecourseDynamics.jl` - the Julia package we introduced earlier.
+The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
 
@@
 
-The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice.
+The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
 We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class.
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse.
-In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$.
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse.
+In order to simulate the dynamic process, we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$.
 Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class.
@@
 \begin{algorithm}
-\caption{Experiment}\label{algo-experiment}
+\caption{Simulation Experiment}\label{algo-experiment}
 \begin{algorithmic}[1]
@@
 As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
-Using our framework the experiment can be conducted on an arbitrary number of algorithmic recourse generators.
-As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.
+
+Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $D$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators.
+The framework we have built facilitates this, making use of multi-threading in order to speed up computations.
+Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
+To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,D,G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
 

### 2022-08-12 dev/paper/paper.tex

@@
 Existing work on Counterfactual Explanations (CE) and Algorithmic
-Recourse (AR) has largely been limited to the static setting: given some
-classifier we are interested in finding close, actionable, realistic,
-sparse, diverse and ideally causally founded counterfactuals. The
-ability of CE to handle dynamics like data and model drift remains a
-largely unexplored research challenge at this point. Only one recent
-work considers the implications of exogenous domain and model shifts.
-This project instead focuses on endogenous dynamics, that is shifts that
-occur when AR is actually implemented by a proportion of individuals.
-Early findings suggest that the involved shifts may be large with
-important implications on the validity of AR and the overall
-characteristics of the sample population.
+Recourse (AR) has largely been limited to the static setting and focused
+on single individuals: given some estimated model the goal is to find
+valid counterfactuals for individual instance that fulfill various
+desiderata. The ability of such counterfactuals to handle dynamics like
+data and model drift remains a largely unexplored research challenge at
+this point. There has also been surprisingly little work on the related
+question of how the actual implementation of recourse by one individual
+may affect other individuals. Through this work we aim to close that gap
+by systematizing and extending existing knowledge. We first show that
+many of the existing methodologies can be collectively described by a
+generalized framework. We then argue that the existing framework fails
+to account for a hidden external cost of recourse, that only reveals
+itself when studying the endogenous dynamics of recourse at the group
+level. Through simulation experiments involving various popular
+counterfactual generators and several benchmark datasets, we generate a
+total XX million Counterfactual Explanations and study the resulting
+domain and model shifts. We find that the induced shifts are substantial
+enough to likely impede the applicability Algorithmic Recourse in
+practice. Fortunately, we find various potential mitigation strategies
+that can be used in combination with existing approaches. Our simulation
+framework for studying recourse dynamics is fast and open-sourced.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 has tackled this issue from two different angles: while some have
-appealed for a strict focus on inherently iterpretable models
+appealed for a strict focus on inherently interpretable models
 \protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
@@
 (AR) to help individuals who face adverse outcomes. An example relevant
-to the Social Sciences is consumer credit: in this context AR can be
+to the Social Sciences is consumer credit: in this context, AR can be
 used to guide individuals in improving their creditworthiness, should
@@
 work has compared different approaches within this static setting
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work we go
+\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work, we go
 one step further and ask ourselves: what happens if recourse is provided
@@
 
-Research on algorithmic recourse has also so far typically addressed the
+Research on Algorithmic Recourse has also so far typically addressed the
 issue from the perspective of one single individual and has indeed been
 referred to as \textbf{individual recourse} in some places. Arguably
-though, most real-world applications that warrant algorithmic recourse
+though, most real-world applications that warrant Algorithmic Recourse
 involve potentially large groups of individuals typically competing for
@@
 knowledge are as follows: firstly, we posit a compelling argument that
-calls for a novel perspective on algorithmic recourse extending our
+calls for a novel perspective on Algorithmic Recourse extending our
 focus from single individuals to groups. Secondly, we introduce an
@@
 \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
-which enables us to study macrodynamics of algorithmic recourse through
+which enables us to study macrodynamics of Algorithmic Recourse through
 simulations that can be fully parallelized. Thirdly, we use this
@@
 Section~\ref{sec-related} places our work in the broader context of
-related literature. Section~\ref{sec-method} presents our methodology
-and data. Section~\ref{sec-empirical} presents our empirical findings
-which are then discussed in the broader context of the literature in
-Section~\ref{sec-discussion}. We also point to some of the limitations
-or our work as well as avenues for future research in
+related literature. Section~\ref{sec-method} posits a generalized
+framework for gradient-based Algorithmic Recourse and introduces the
+notion of hidden external costs. Section~\ref{sec-method-2} sets out our
+experimental framework for modeling endogenous macrodynamics in AR.
+Section~\ref{sec-empirical} presents our experimental results for
+synthetic and real-world datasets. We also present evidence for the
+effectiveness of various proposed mitigation strategies in that section.
+Our findings are then discussed in the broader context of the literature
+in Section~\ref{sec-discussion}. We also point to some of the
+limitations of our work as well as avenues for future research in
 Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
@@
 Section~\ref{sec-related-recourse} discusses the existing research
-within the domain of counterfactual explanations and algorithmic
-recourse. Then, Section~\ref{sec-related-shifts} presents some of the
+within the domain of Counterfactual Explanations and Algorithmic
+Recourse. Then, Section~\ref{sec-related-shifts} presents some of the
 previous work on the measurement of dataset and model shifts.
@@
 \(Y \in \mathcal{Y}\). Then we are interested in minimizing the
-complexity or effort \(H=h(x\prime)\) associated with moving an
-individual \(x\) to a counterfactual state \(x\prime\) such that the
-predicted outcome \(M(x\prime)\) corresponds to some target outcome
-\(t\):
+cost\footnote{Equivalently, others have referred to this quantity as
+  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
+by individual \(x\) when moving to a counterfactual state \(x\prime\)
+such that the predicted outcome \(M(x\prime)\) corresponds to some
+target outcome \(y^*\):
 
 \begin{equation}\protect\hypertarget{eq-obj}{}{
-\min_{x\prime \in \mathcal{x}} c(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
+\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
 }\label{eq-obj}\end{equation}
@@
 \begin{equation}\protect\hypertarget{eq-solution}{}{
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),t) + \lambda h(x\prime)
+x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
 }\label{eq-solution}\end{equation}
 
-In the baseline work and many subsequent approaches the complexity
+In the baseline work
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, the cost
 function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
-distance metric based on the simple intuition that large perturbations
-of \(x\) are costly.
-
-Many approaches for the generation of algorithmic recourse have been
-described in the literature. An October 2020 survey by Karimi et
-al.~layed out 60 algorithms that have been proposed since 2014
+distance metric based on the simple intuition that perturbations of
+\(x\) are costly to the individual. For models that are differentiable
+and produce smooth predictions, Equation~\ref{eq-solution} can be solved
+through gradient descent. This summarizes the approach followed in
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} which we
+shall refer to simply as \textbf{Wachter} - the name of the first author
+- in the remainder of this paper.
+
+Many approaches for the generation of Algorithmic Recourse have been
+described in the literature since 2017. An October 2020 survey by Karimi
+et al.~layed out 60 algorithms that have been proposed since 2014
 \protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
@@
 Efforts so far have largely been directed at improving the quality of
-counterfactual explanations within a static context: given some
-pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are
-interested in generating one or multiple meaningful counterfactual
-explanations for some individual characterized by \(x_i\). The ability
-of counterfactual explanations to handle dynamics like data and model
+Counterfactual Explanations within a static context: given some
+pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
+interested in generating one or multiple meaningful Counterfactual
+Explanations for some individual characterized by \(x\). The ability of
+Counterfactual Explanations to handle dynamics like data and model
 shifts remains a largely unexplored research challenge at this point
@@
 \protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose framework
-for algorithmic recourse (ROAR) that evidently improves robustness to
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose ROAR - a
+framework for Algorithmic Recourse that evidently improves robustness to
 such exogenous shifts.
@@
 counterfactuals. We provide a motivation for doing so from the
-perspective of endogenous macrodynamics of algorithmic recourse.
+perspective of endogenous macrodynamics of Algorithmic Recourse.
 
@@
 
-In the context of algorithmic recourse, domain and model shifts were
+In the context of Algorithmic Recourse, domain and model shifts were
 first brought up by the authors behind ROAR
@@
 Despite the large and growing number of approaches to counterfactual
-search there have been surprisingly few benchmark studies that compare
+search, there have been surprisingly few benchmark studies that compare
 different methodologies. This may be partially due to limited software
@@
 \protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
-framework that can be used to apply and benchmark different
-methodologies; finally,
+framework - CARLA - that can be used to apply and benchmark different
+methodologies; finally, \texttt{CounterfactualExplanations.jl}
 \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
@@
 simulations, we have relied on and extended the Julia implementation due
-to the associated performance benefits.
+to the associated performance benefits. In particular, we have built a
+framework on top of \texttt{CounterfactualExplanations.jl} that extends
+the functionality from static benchmarks to simulation experiments:
+\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
+  available from \ldots{}}. The core concepts implemented in that
+package reflect what is presented in section Section~\ref{sec-method-2}
+of this paper.
 
@@
 
-In the following we first set out a generalized framework for
+In this section we first set out a generalized framework for
 gradient-based counterfactual search in Section~\ref{sec-method-general}
-to introduce the various counterfactual generators we have chosen to use
-in our experiments. We then describe the experimental setup in
-Section~\ref{sec-method-2-experiment} and introduce several evaluation
-metrics used to benchmark the different generators.
+that encapsulates the various individual recourse methods we have chosen
+to use in our experiments. We then introduce the notion of a hidden
+external cost in algorithmic recourse and extend the existing framework
+to explicitly address this cost in the counterfactual search objective.
 
@@
 
-In this work we have chosen to focus on a number of gradient-based
-counterfactual generators to investigate the endogenous dynamics we
-introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
-search is well-suited for differentiable black-box models like deep
-neural networks. We can restate Equation~\ref{eq-solution} in a more
-general form that encompasses most gradient-based approaches to
-counterfactual search:
+We have chosen to focus on gradient-based counterfactual search for two
+reasons: firstly, they can be seen as direct descendants of our baseline
+method - Wachter; secondly, gradient-based search is particularly
+well-suited for differentiable black-box models like deep neural
+networks, which we focus on in this work. In particular, we include
+include the following generators in our simulation experiments below:
+\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
+\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
+\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
+and a greedy approach that relies on probabilistic models
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. Our motivation for
+including these different generators in our analysis, is that they all
+offer slightly different approaches to generate meaningful
+counterfactuals for differentiable black-box models. We hypothesize that
+generating more \textbf{meaningful} counterfactuals should mitigate the
+endogenous dynamics illustrated in Figure~\ref{fig-poc} in
+Section~\ref{sec-intro}. This intuition stems from the underlying idea
+that more meaningful counterfactuals are generated by the same or at
+least a very similar data generating process as the training data. All
+else equal, counterfactuals that fulfill this basic requirement should
+be less prone to trigger domain and model shifts.
+
+As we will see next, all of them can be described by the following
+generalized form of Equation~\ref{eq-solution}:
 
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
 \end{aligned}
@@
 \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
-state space to the feature space.
+state space to the feature space. In Wachter, the state space is the
+feature space: \(f\) is just the identity function and the number of
+counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
+counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
+instead of the feature space directly. The latent embedding is learned
+by a separate generative model that is tasked with learning the data
+generating process (DGP) of \(X\). In this case \(f\) in
+Equation~\ref{eq-general} corresponds to the decoder part of the
+generative model, in other words the function that maps back from the
+latent embedding to the feature space. Provided the generative model is
+well-specified, traversing the latent embedding typically results in
+realistic and plausible counterfactuals, because they are implicitly
+generated by the (learned) DGP
+\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}.
+
+CLUE distinguishes itself from REVISE and other counterfactual
+generators in that it aims to minimize the predictive uncertainty of the
+model in question, \(M\). To quantify predictive uncertainty the authors
+rely on entropy estimates for probabilistic models. The approach
+proposed by \protect\hyperlink{ref-schut2021generating}{{[}7{]}}, which
+we shall refer to as \textbf{Greedy}, also works with the subclass of
+models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
+predictive uncertainty estimates. The authors show that in this setting
+the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
+and meaningful counterfactuals can be generated in a fast and efficient
+manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}7{]}} also
+show that by maximizing the predicted probability of \(x^\prime\) being
+assigned to target class \(y^*\) we also implicitly minimize predictive
+entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
+REVISE in the Bayesian context and we shall therefore refer to both
+approaches collectively as \textbf{Latent Space} generators\footnote{In
+  fact, there are a number of other recently proposed approaches to
+  counterfactual search that also broadly fall in this same category.
+  They largely differ with respect to the chosen generative model: for
+  example, the generator proposed by
+  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}24{]}} relies
+  on normalizing flows.}.
+
+Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
+distinguishes itself from all other generators considered here in that
+it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
+al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} show that
+diverse outcomes can in principal be achieved simply rerunning
+counterfactual search multiple times using stochastic gradient descent
+(or by randomly initializing the counterfactual). In
+\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} diversity is
+explicitly proxied via Determinantal Point Processes (DDP): the authors
+simply introduce DDP as a component of the cost function
+\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
+\(s_1, ... , s_K\) that look as different from each other as possible.
+The implementation of DiCE in the our library of choice -
+\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
+worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
+constant and therefore does not affect the objective function
+Equation~\ref{eq-general}.
 
@@
 
-To explicitly address the notion that individual recourse may affect the
-outcome and prospect of other individuals, we propose to extend
+All of the different approaches introduced above tackle the problem of
+Algorithmic Recourse from the perspective of one single
+individual\footnote{DiCE recognizes that different individuals may have
+  different objective functions, but it does not address the
+  interdependencies between different individuals.}. To explicitly
+address the issue that individual recourse may affect the outcome and
+prospect of other individuals, we propose to extend
 Equation~\ref{eq-general} as follows:
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
 \end{aligned}
@@
 the individual as before and \(\lambda_1\) governs to what extent that
-private cost ought to be penalised. The newly introduced term
-\(g(f(s_k^\prime))\) is meant to capture and address social costs
+private cost ought to be penalized. The newly introduced term
+\(g(f(s_k^\prime))\) is meant to capture and address external costs
 incurred by the collective of individuals in response to changes in
-\(\mathbf{s}^\prime\). The trade-off between private and social costs is
-determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
-individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
-nor do we intend to provide a definite answer in this work, if such even
-exists. A straight-forward choice simply extends the baseline approach
-by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
-of only penalizing the distance of the individuals' counterfactual to
-its factual, we propose penalizing its distance to some sensible point
-in the target domain, for example the sample average:
-\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
-\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
-gravitate towards the specified point in the target domain. In the
-remainder of this paper we will therefore refer to this approach as
-\textbf{Gravitational} generator, when we investigate its potential
-usefulness for mitigating endongenous macrodynamics\footnote{Note that
-  despite the naming convention our goal here is not to provide yet
-  another counterfactual generator, but merely investigate the most
-  simple penalty we can think of with respect to its effectiveness.}.
+\(\mathbf{s}^\prime\). The underlying concept of private and external
+costs is borrowed from Economics and well-established in that field:
+when the decisions or actions by some individual market participant
+generate external costs, then the market is said to suffer from negative
+externalities and considered inefficient
+\protect\hyperlink{ref-pindyck2014microeconomics}{{[}25{]}}. We think
+that this concept describes the endogenous dynamics of algorithmic
+recourse oberserved here very well. As with individual recourse, the
+exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide
+a definite answer in this work, if such even exists. That being said, we
+do propose a few potential mitigation strategies in
+Section~\ref{sec-empirical-2-mitigate}.
+
+\hypertarget{sec-method-2}{%
+\section{Modeling Endogenous Macrodynamics in Algorithmic
+Recourse}\label{sec-method-2}}
+
+In the following we describe the framework we propose for modeling and
+analysing endogenous macrodynamics in Algorithmic Recourse. We first
+describe the basic simulations that were generated to produce the
+findings in this work and also constitute the core of
+\texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we
+introduced earlier. The remainder of this section then introduces
+various evaluation metrics that can be used to benchmark different
+counterfactual generators with respect to how they perform in the
+dynamic setting.
+
+\hypertarget{sec-method-2-experiment}{%
+\subsection{Simulations}\label{sec-method-2-experiment}}
+
+The dynamics illustrated in Figure~\ref{fig-poc} in
+Section~\ref{sec-intro} were generated through a simple experiment that
+aims to simulate the process of Algorithmic Recourse in practice. We
+begin in the static setting at time time \(t=0\): given some pre-trained
+classifier \(M\) we generate recourse for a random batch of \(B\)
+individuals in the non-target class. Note that we focus our attention on
+classification problems, since classification poses the most common
+practical use-case for Algorithmic Recourse. In order to simulate the
+dynamic process, we suppose that the model \(M\) is retrained following
+the actual implementation of recourse in time \(t=0\). Following the
+update to the model, we assume that at time \(t=1\) recourse is
+generated for yet another random subset of individuals in the non-target
+class. This process is repeated for a number of time periods \(T\). To
+get a clean read on endogenous dynamics we keep the total population of
+samples closed: we allow existing samples to move from factual to
+counterfactual states, but do not allow any entirely new samples to
+enter the population. The experimental setup is summarized in Algorithm
+\ref{algo-experiment}
+
+\begin{algorithm}
+\caption{Simulation Experiment}\label{algo-experiment}
+\begin{algorithmic}[1]
+\Procedure{Experiment}{$M,D,G$}
+\State $t\gets 0$
+\While{$t<T$}
+\State $D_B \subset D$
+\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
+\State $M\gets M(D)$ \Comment{Retrain model.}
+\EndWhile
+\State \textbf{return} $M,D$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+A noteworthy practical consideration is the choice of \(T\) and \(B\).
+The higher these values, the more factual instances undergo recourse
+throughout the entire experiment. Of course, this is likely to lead to
+more pronounced domain and model shifts by time \(T\). At the same time,
+it is generally improbable that a very large part of the population
+would request an explanation of the algorithm's decisions. In our
+experiments, we choose the values such that \(T \cdot B\) corresponds to
+the application of recourse on \(25-50\%\) of the negative instances
+from the initial dataset. As we collect data at each time \(t\), we can
+also verify the impact of recourse when it is implemented for a smaller
+number of individuals.
+
+Algorithm \ref{algo-experiment} summarizes the proposed simulation
+experiment for a given dataset \(D\), model \(M\) and generator \(G\),
+but naturally we are interested in comparing simulation outcomes for
+different sources of data, models and generators. The framework we have
+built facilitates this, making use of multi-threading in order to speed
+up computations. Holding the initial model and dataset constant the
+experiments are run for all generators, since our primary concern is to
+benchmark different recourse methods. To ensure that each generator is
+faced with the exact same initial conditions in each round \(t\), the
+candidate batch of individuals from the non-target class is randomly
+drawn from the intersection of all individuals in the non-target class
+across all experiments
+\(\left\{\textsc{Experiment}(M,D,G)\right\}_{j=1}^J\) where \(J\) is the
+total number of generators.
+
+\hypertarget{sec-method-2-metrics}{%
+\subsection{Evaluation Metrics}\label{sec-method-2-metrics}}
+
+We formulate two desiderata for the set of metrics used to measure
+domain and model shifts induced by recourse. First, the metrics should
+be applicable regardless of the dataset or classification technique so
+that they allow for the meaningful comparison of the generators in
+various scenarios. As the knowledge of the underlying probability
+distribution is rarely available, the metrics should be empirical and
+non-parametric. This further ensures that we can also measure large
+datasets by sampling from the available data. Moreover, while our study
+was conducted in a two-class classification setting, our choice of
+metrics should remain applicable in the future research on multi- class
+recourse problems. Second, the set of metrics should allow to capture
+various aspects of the previously mentioned magnitude, path, and tempo
+of changes while remaining as small as possible.
+
+\hypertarget{domain-shifts}{%
+\subsubsection{Domain Shifts}\label{domain-shifts}}
+
+To quantify the magnitude of domain shifts we rely on an unbiased
+estimate of the squared population \textbf{Maximum Mean Discrepancy
+(MMD)} given as:
+
+\begin{equation}\protect\hypertarget{eq-mmd}{}{
+\begin{aligned}
+MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+}\label{eq-mmd}\end{equation}
+
+where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
+Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
+identically distributed samples drawn from probability distributions
+\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
+between the kernel mean embeddings of \(p\) and \(q\) in RKHS
+\(\mathcal{H}\). An important consideration is the choice of the kernel
+function \(k(\cdot,\cdot)\). In our implementation we make use of the
+radial basis function (RBF) kernel with a constant length-scale
+parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
+and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
+\(X=\tilde{X}\).
+
+The evaluation metric in Equation~\ref{eq-mmd} is computed after every
+round \(t=1,...,T\) of the experiment. To assess the statistical
+significance of the observed shifts under the null hypothesis that
+samples \(X\) and \(\tilde{X}\) were drawn from the same probability
+distribution we follow
+\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
+combine the two samples and generate a large number of permutations of
+\(X + \tilde{X}\). Then, we split the permuted data into two new samples
+\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
+original samples. Then under the null hypothesis we should have that
+\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
+\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
+calculated by counting how these two quantities are not equal.
+
+We calculate the MMD for both classes individually based on the ground
+truth labels, i.e.~the labels that samples were assigned in time
+\(t=0\). Throughout our experiments we generally do not expect the
+distribution of the negative class to change over time -- application of
+recourse reduces the size of this class, but since individuals are
+sampled uniformly the distribution should remain unaffected. Conversely,
+unless a recourse generator can perfectly replicate the original
+probability distribution, we expect the MMD of the positive class to
+increase. Thus, when discussing MMD, we generally mean the shift in the
+distribution of the positive class.
+
+Finally, \textbf{feature mean and feature standard deviation} are also
+calculated to verify how the implementation of recourse impacts every
+attribute in the dataset. Although MMD already captures information
+about the expected value and variance, we may also be interested in a
+more granular look at individual features.
+
+\hypertarget{model-shifts}{%
+\subsubsection{Model Shifts}\label{model-shifts}}
+
+As our baseline for quantifying model shifts we measure perturbations to
+the model parameters at each point in time \(t\) following
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
+\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
+between the vectors of parameters before and after retraining the model
+\(M\). We shall refer to this baseline metric simply as
+\textbf{Perturbations}.
+
+We extend the metric in Equation~\ref{eq-mmd} for the purpose of
+quantifying model shifts. Specifically, we introduce \textbf{Predicted
+Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
+features directly, we apply it to the predicted probabilities assigned
+to a set of samples by the model \(M\). If the model shifts, the
+probabilities assigned to each sample will change; again, this metric
+will equal 0 only if the two classifiers are the same. It is worth
+noting that while we apply the technique to samples drawn uniformly from
+the dataset, it can also be employed on arbitrary points in the entire
+feature space (or a subspace). The latter approach is theoretically more
+robust. Unfortunately, in practice this approachs suffers from the curse
+of dimensionality, since it becomes increasingly difficult to select
+enough points to overcome noise as the dimension \(D\) grows.
+
+As an alternative to PP MMD we use a pseudo-distance for the
+\textbf{Disagreement Coefficient} (Disagreement). This metric was
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
+estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
+classifiers do not agree on the predicted outcome for a randomly chosen
+sample. Thus, it is not relevant whether the classification is correct
+according to the ground truth, but only whether the sample lies on the
+same side of the two respective decision boundaries. In our context,
+this metric quantifies the overlap between the initial model (trained
+before the application of recourse) and the updated model. A
+Disagreement Coefficient unequal to zero is indicative of a model shift.
+The opposite is not true: even if the Disagreement Coefficient is equal
+to zero a model shift may still have occured. This is one reason for why
+PP MMD is our our preferred metric.
+
+Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
+the likelihood that a model assigns a high probability to its
+classification of any given sample. We define the metric simply as
+\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
+predicted logits from a binary classifier and \(\sigma\) denotes the
+sigmoid function. This metric provides an unbiased estimate of the
+binary classifier's tendency to produce high-confidence predictions in
+either one of the two classes. Although the exact values for this metric
+are not important for our study, they can be used to detect model
+shifts. If decisiveness changes over time, then this is indicative of
+the decision boundary moves towards either one of the two classes.
+
+\hypertarget{sec-empirical}{%
+\section{Experiment Setup}\label{sec-empirical}}
+
+\hypertarget{sec-empirical-data}{%
+\subsection{Data}\label{sec-empirical-data}}
+
+We have chosen to work with both synthetic and real-world datasets.
+Using synthetic data allows us to impose distributional properties that
+may affect the resulting recourse dynamics. Following
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}, we generate
+synthetic data in \(\mathbb{R}^2\) to also allow for a visual
+interpretation of the results. Real-world data is used in order to
+assess if endogenous dynamics also occur in higher-dimensional settings.
+
+\hypertarget{synthetic-data}{%
+\subsubsection{Synthetic data}\label{synthetic-data}}
+
+We use four synthetic binary classification datasets consisting of 1000
+samples each. The datasets are presented in
+Figure~\ref{fig-synthetic-data} (see also Appendix A for a formal
+description). Samples from the negative class are marked in blue while
+samples of the positive class are marked in orange.
+
+\begin{figure}
+
+{\centering \includegraphics[width=8cm,height=2cm]{www/synthetic_data.png}
+
+}
+
+\caption{\label{fig-synthetic-data}Synthetic classification datasets
+used in our experiments.}
+
+\end{figure}
+
+Ex-ante we expect to see that by construction Wachter will create a new
+cluster of counterfactual instances in the proximity of the initial
+decision boundary. Thus, the choice of a black-box model may have an
+impact on the paths of the recourse. For generators that use latent
+space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
+CLUE \protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on
+(and have access to) probabilistic models (CLUE
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, Greedy
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
+counterfactuals will end up in regions of the target domain that are
+densely populated by training samples. Of course, this is expectation
+hinges on how effective said probabilistic models are at capturing
+predictive uncertainty. Finally, we expect to see the counterfactuals
+generated by DiCE to be uniformly spread around the feature space inside
+the target class\footnote{As we mentioned earlier, the diversity
+  constraint used by DiCE is only effective for when at least two
+  counterfactuals are being generated. We have therefore decided to
+  always generate 5 counterfactuals for each generator and randomly pick
+  one of them.}. In summary, we expect that the endogenous shifts
+induced by Wachter outsize those induced by all other generators, since
+Wachter is the only approach that is not concered with generating what
+we have defined as meaningful counterfactuals.
+
+\hypertarget{real-world-data}{%
+\subsubsection{Real-world data}\label{real-world-data}}
+
+We use three different real-world datasets from the Finance and
+Economics domain, all of which are tabular and can be used for binary
+classification. Firstly, we use the \textbf{Give Me Some Credit} dataset
+which was open-sourced on Kaggle for the task to predict whether a
+borrower is likely to experience financial difficulties in the next two
+years \protect\hyperlink{ref-gmsc_data}{{[}28{]}}. Originally consisting
+of 250,000 instances with 11 numerical attributes. Secondly, we use the
+\textbf{UCI defaultCredit} dataset
+\protect\hyperlink{ref-yeh2009comparisons}{{[}29{]}}, a benchmark
+dataset that can be used to train binary classifiers to predict the
+binary outcome variable, whether credit card clients default on their
+payment. In its raw form it consists of 23 explanatory variables - 4
+categorical features relating to demographic attributes\footnote{These
+  have been ommitted from the analysis. See Section~\ref{sec-limit-data}
+  for details.} and 19 continuous features largely relating to
+individuals' payment histories and amount of credit outstanding. Both of
+these datasets have been used in the literature on Algorithmic Recourse
+before (see for example
+\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
+\protect\hyperlink{ref-joshi2019towards}{{[}8{]}} and
+\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}), presumably
+because they constitute real-world classification tasks involving
+individuals that compete for access to credit.
+
+As a third dataset we include the \textbf{California Housing} dataset
+derived from the 1990 U.S. census and sourced through scikit-learn
+\protect\hyperlink{ref-pace1997sparse}{{[}31{]}}. It consists of 8
+continuous features that can be used to predict the median house price
+for California districts. The continuous outcome variable is binarized
+as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or
+not the median house price of a given district is above or below the
+median of all districts. While we have not seen this dataset used in the
+previous literature on AR, others have used the Boston Housing dataset
+in a similar fashion (see for example
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}). While we
+initially also conducted experiments on that dataset, we eventually
+discarded this dataset, since it has been found to suffer from an
+ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}32{]}}.
+
+Since the simulations involve generating counterfactuals for a
+significant proportion of the entire sample of individuals, we have
+randomly undersampled each dataset to yield balanced subsamples
+consisting of 10,000 individuals each. We have also standardized all
+explanatory features since our chosen classifiers are sensetive to
+scale.
+
+\hypertarget{sec-empirical-classifiers}{%
+\subsection{Classifiers and Generative
+Models}\label{sec-empirical-classifiers}}
+
+For each dataset and generator we look at three different types of
+classifiers all of them built and trained using \texttt{Flux.jl}
+\protect\hyperlink{ref-innes2018fashionable}{{[}33{]}}: firstly, a
+simple linear classifier - \textbf{Logistic Regression} - implemented as
+single linear layer with sigmoid activation; secondly, a multilayer
+perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
+composed of five MLPs following
+\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}34{]}} that serves
+as our only probabilistic classifier. We have chosen to work with deep
+ensembles both for their simplicity and effectiveness at modelling
+predictive uncertainty. They are also the model of choice in
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
+network architectures are kept simple (Table~\ref{tbl-mlp}), since we
+are only marginally concerned with achieving good initial classifier
+performance. For the real-world datasets we using mini-batch training
+and dropout regularization.
+
+The Latent Space generators rely on separate generative models.
+Following the authors of both REVISE and CLUE we use Variational
+Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers,
+we deliberately choose to work with fairly simple architectures
+(Table~\ref{tbl-vae}). More expressive generative models generally also
+lead to more meaningful counterfactuals produced by Latent Space
+generators. But in our view this should simply be considered as a
+vulnerability of counterfactual generators that rely on surrogate models
+to learn what realistic representations of the underlying data.
+
+All classifiers and generative models are retrained for 10 epochs in
+each round \(t\) of the experiment. Rather than retraining models from
+scratch, we initialize all parameters at their previous levels (\(t-1\))
+and compute backpropagate for 10 epochs using the new training data as
+inputs into the existing model.
+
+\begin{table}
+
+\caption{\label{tbl-panel}Model
+Architectures}\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-mlp}MLP}
+
+{\centering
+
+\begin{tabular}[t]{lllll}
+\toprule
+ & Hidden Dim. & Hidden Layers & Batch & Dropout\\
+\midrule
+Synthetic & 32 & 1 & - & -\\
+Real-World & 32 & 2 & 50 & 0.25\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{minipage}%
+\newline
+\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-vae}Variational Autoencoder}
+
+{\centering
+
+\begin{tabular}[t]{lll}
+\toprule
+ & Hidden Dim. & Epochs\\
+\midrule
+Synthetic & 2 & 100\\
+Real-World & 8 & 250\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{minipage}%
+
+\end{table}
+
+\hypertarget{sec-empirical-2}{%
+\section{Experiments}\label{sec-empirical-2}}
+
+\hypertarget{endogenous-macrodynamics}{%
+\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
+
+\hypertarget{potential-mitigation-strategies}{%
+\subsection{Potential Mitigation
+Strategies}\label{potential-mitigation-strategies}}
+
+\hypertarget{sec-empirical-2-mitigate}{%
+\subsubsection{Gravitational Counterfactual
+Explanations}\label{sec-empirical-2-mitigate}}
+
+A straight-forward choice simply extends the baseline approach by
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead of
+only penalizing the distance of the individuals' counterfactual to its
+factual, we propose penalizing its distance to some sensible point in
+the target domain, for example the sample average: \(\bar{\mathbf{x}}\).
+For such a recourse objective, higher choices of \(\lambda_2\) relative
+to \(\lambda_1\) will lead counterfactuals to gravitate towards the
+specified point in the target domain. In the remainder of this paper we
+will therefore refer to this approach as \textbf{Gravitational}
+generator, when we investigate its potential usefulness for mitigating
+endongenous macrodynamics\footnote{Note that despite the naming
+  convention our goal here is not to provide yet another counterfactual
+  generator, but merely investigate the most simple penalty we can think
+  of with respect to its effectiveness.}.
 
 \hypertarget{a-note-on-convergence}{%
-\subsection{A note on convergence}\label{a-note-on-convergence}}
+\paragraph{A note on convergence}\label{a-note-on-convergence}}
 
@@
 
-\hypertarget{sec-method-2}{%
-\section{Modeling Endogenous Macrodynamics in Algorithmic
-Recourse}\label{sec-method-2}}
-
-In this \ldots{}
-
-\hypertarget{sec-method-2-experiment}{%
-\subsection{Simulations}\label{sec-method-2-experiment}}
-
-The dynamics illustrated in Figure~\ref{fig-poc} in
-Section~\ref{sec-intro} were generated through a simple experiment that
-aims to simulate the process of algorithmic recourse in practice. We
-begin in the static setting at time time \(t=0\): given some pre-trained
-classifier \(M\) we generate recourse for a random batch of \(B\)
-individuals in the non-target class. Note that we focus our attention on
-classification problems, since classification poses the most common
-practical use-case for algorithmic recourse. In order to simulate the
-dynamical process we suppose that the model \(M\) is retrained following
-the actual implementation of recourse in time \(t=0\). Following the
-update to the model, we assume that at time \(t=1\) recourse is
-generated for yet another random subset of individuals in the non-target
-class. This process is repeated for a number of time periods \(T\). To
-get a clean read on endogenous dynamics we keep the total population of
-samples closed: we allow existing samples to move from factual to
-counterfactual states, but do not allow any entirely new samples to
-enter the population. The experimental setup is summarized in Algorithm
-\ref{algo-experiment}
-
-\begin{algorithm}
-\caption{Experiment}\label{algo-experiment}
-\begin{algorithmic}[1]
-\Procedure{Experiment}{$M,D,G$}
-\State $t\gets 0$
-\While{$t<T$}
-\State $D_B \subset D$
-\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
-\State $M\gets M(D)$ \Comment{Retrain model.}
-\EndWhile
-\State \textbf{return} $M,D$
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-A noteworthy practical consideration is the choice of \(T\) and \(B\).
-The higher these values, the more factual instances undergo recourse
-throughout the entire experiment. Of course, this is likely to lead to
-more pronounced domain and model shifts by time \(T\). At the same time,
-it is generally improbable that a very large part of the population
-would request an explanation of the algorithm's decisions. In our
-experiments, we choose the values such that \(T \cdot B\) corresponds to
-the application of recourse on \(25-50\%\) of the negative instances
-from the initial dataset. As we collect data at each time \(t\), we can
-also verify the impact of recourse when it is implemented for a smaller
-number of individuals. Using our framework the experiment can be
-conducted on an arbitrary number of algorithmic recourse generators. As
-all generators make use of the same initial model and initial dataset,
-the differences in domain and model shifts observed throughout the
-rounds depend solely on the employed generator.
-
-\hypertarget{sec-method-2-metrics}{%
-\subsection{Evaluation Metrics}\label{sec-method-2-metrics}}
-
-We formulate two desiderata for the set of metrics used to measure
-domain and model shifts induced by recourse. First, the metrics should
-be applicable regardless of the dataset or classification technique so
-that they allow for the meaningful comparison of the generators in
-various scenarios. As the knowledge of the underlying probability
-distribution is rarely available, the metrics should be empirical and
-non-parametric. This further ensures that we can also measure large
-datasets by sampling from the available data. Moreover, while our study
-was conducted in a two-class classification setting, our choice of
-metrics should remain applicable in the future research on multi- class
-recourse problems. Second, the set of metrics should allow to capture
-various aspects of the previously mentioned magnitude, path, and tempo
-of changes while remaining as small as possible.
-
-\hypertarget{domain-shifts}{%
-\subsubsection{Domain Shifts}\label{domain-shifts}}
-
-To quantify the magnitude of domain shifts we rely on an unbiased
-estimate of the squared population \textbf{Maximum Mean Discrepancy
-(MMD)} given as:
-
-\begin{equation}\protect\hypertarget{eq-mmd}{}{
-\begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
-\end{aligned}
-}\label{eq-mmd}\end{equation}
-
-where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
-Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
-identically distributed samples drawn from probability distributions
-\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
-between the kernel mean embeddings of \(p\) and \(q\) in RKHS
-\(\mathcal{H}\). An important consideration is the choice of the kernel
-function \(k(\cdot,\cdot)\). In our implementation we make use of the
-radial basis function (RBF) kernel with a constant length-scale
-parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
-and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
-\(X=\tilde{X}\).
-
-The evaluation metric in Equation~\ref{eq-mmd} is computed after every
-round \(t=1,...,T\) of the experiment. To assess the statistical
-significance of the observed shifts under the null hypothesis that
-samples \(X\) and \(\tilde{X}\) were drawn from the same probability
-distribution we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}24{]}}. To that end, we
-combine the two samples and generate a large number of permutations of
-\(X + \tilde{X}\). Then, we split the permuted data into two new samples
-\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
-original samples. Then under the null hypothesis we should have that
-\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
-\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
-calculated by counting how these two quantities are not equal.
-
-We calculate the MMD for both classes individually based on the ground
-truth labels, i.e.~the labels that samples were assigned in time
-\(t=0\). Throughout our experiments we generally do not expect the
-distribution of the negative class to change over time -- application of
-recourse reduces the size of this class, but since individuals are
-sampled uniformly the distribution should remain unaffected. Conversely,
-unless a recourse generator can perfectly replicate the original
-probability distribution, we expect the MMD of the positive class to
-increase. Thus, when discussing MMD, we generally mean the shift in the
-distribution of the positive class.
-
-Finally, \textbf{feature mean and feature standard deviation} are also
-calculated to verify how the implementation of recourse impacts every
-attribute in the dataset. Although MMD already captures information
-about the expected value and variance, we may also be interested in a
-more granular look at individual features.
-
-\hypertarget{model-shifts}{%
-\subsubsection{Model Shifts}\label{model-shifts}}
-
-As our baseline for quantifying model shifts we measure perturbations to
-the model parameters at each point in time \(t\) following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
-\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
-between the vectors of parameters before and after retraining the model
-\(M\). We shall refer to this baseline metric simply as
-\textbf{Perturbations}.
-
-We extend the metric in Equation~\ref{eq-mmd} for the purpose of
-quantifying model shifts. Specifically, we introduce \textbf{Predicted
-Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
-features directly, we apply it to the predicted probabilities assigned
-to a set of samples by the model \(M\). If the model shifts, the
-probabilities assigned to each sample will change; again, this metric
-will equal 0 only if the two classifiers are the same. It is worth
-noting that while we apply the technique to samples drawn uniformly from
-the dataset, it can also be employed on arbitrary points in the entire
-feature space (or a subspace). The latter approach is theoretically more
-robust. Unfortunately, in practice this approachs suffers from the curse
-of dimensionality, since it becomes increasingly difficult to select
-enough points to overcome noise as the dimension \(D\) grows.
-
-As an alternative to PP MMD we use a pseudo-distance for the
-\textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}25{]}} and
-estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
-classifiers do not agree on the predicted outcome for a randomly chosen
-sample. Thus, it is not relevant whether the classification is correct
-according to the ground truth, but only whether the sample lies on the
-same side of the two respective decision boundaries. In our context,
-this metric quantifies the overlap between the initial model (trained
-before the application of recourse) and the updated model. A
-Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal
-to zero a model shift may still have occured. This is one reason for why
-PP MMD is our our preferred metric.
-
-Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
-the likelihood that a model assigns a high probability to its
-classification of any given sample. We define the metric simply as
-\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
-predicted logits from a binary classifier and \(\sigma\) denotes the
-sigmoid function. This metric provides an unbiased estimate of the
-binary classifier's tendency to produce high-confidence predictions in
-either one of the two classes. Although the exact values for this metric
-are not important for our study, they can be used to detect model
-shifts. If decisiveness changes over time, then this is indicative of
-the decision boundary moves towards either one of the two classes.
-
-\hypertarget{sec-empirical}{%
-\section{Experiment Setup}\label{sec-empirical}}
-
-\hypertarget{sec-empirical-recourse}{%
-\subsection{Recourse Generators}\label{sec-empirical-recourse}}
-
-In the case of the baseline counterfactual generator
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
-the idendity function and the number of counterfactuals \(K\) is equal
-to one. This generator, which we shall refer to as \textbf{Wachter} in
-the following, shall serve as the baseline against which all other
-gradient-based methodologies will be compared. In particular, we include
-include the following generator in our benchmarking exercises: REVISE
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DICE
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and a greedy
-approach that relies on probabilistic models
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}.
-
-Both \textbf{REVISE} and \textbf{CLUE} search counterfactuals in some
-latent embedding \(S \subset \mathcal{S}\) instead of the feature space
-directly. The latent embedding is learned by a separate generative model
-that is tasked with learning the data generating process (DGP) of \(X\).
-In this case \(f\) in Equation~\ref{eq-general} corresponds to the
-decoder part of the generative model, in other words the deterministic
-function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent
-embedding typically results in realistic and plausible counterfactuals,
-because they are implicitly generated by the (learned) DGP
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}. CLUE distinguishes
-itself from REVISE and other counterfactual generators in that it aims
-to minimize the predictive uncertainty of the model in question \(M\).
-To quantify predictive uncertainty the authors rely on entropy estimates
-for probabilistic models. The \textbf{Greedy} approach proposed by
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}} also works with the
-subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can
-produce predictive uncertainty estimates. The authors show that in this
-setting the complexity penalty \(h(\cdot)\) in Equation~\ref{eq-general}
-is redundant and meaningful counterfactuals can be generated in a fast
-and efficient manner through a modified Jacobian-based Saliency Map
-Attack (JSMA). Finally, \textbf{DICE} distinguishes itself from all
-other generators considered here in that it aims to generate a diverse
-set of \(K>1\) counterfactuals. To this end the authors use a complexity
-penalty \(h(\mathbf{s}^\prime)\) that favours diverse outcomes, in the
-sense that \(s_1, ... , s_K\) look as different from each other as
-possible.
-
-Our motivation for including these different generators in our analysis,
-is that they all offer slightly different approaches to generate
-meaningful counterfactuals for differentiable black-box models. We
-hypothesize that generating more \textbf{meaningful} counterfactuals
-should mitigate the endogenous dynamics illustrated in
-Figure~\ref{fig-poc} in Section~\ref{sec-intro}. This intuition stems
-from the underlying idea that more meaningful counterfactuals are
-generated by the same or at least a very similar data generating process
-as the training data. All else equal, counterfactuals that fulfill this
-basic requirement should be less prone to trigger domain and model
-shifts.
-
-\hypertarget{sec-empirical-data}{%
-\subsection{Data}\label{sec-empirical-data}}
-
-We have chosen to work with both synthetic and real-world datasets.
-Using synthetic data allows us to impose distributional properties that
-may affect the resulting recourse dynamics. Following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} we generate
-synthetic data in \(\mathbb{R}^2\) to also allow for a visual
-interpretation of the results. Real-world data is used in order to
-assess if endogenous dynamics also occur in higher-dimensional settings.
-
-\hypertarget{synthetic-data}{%
-\subsubsection{Synthetic data}\label{synthetic-data}}
-
-We use four synthetic binary classification datasets consisting of 1000
-samples each.\footnote{To see how the data is generated see here:
-  \url{https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb}}
-The datasets are presented in Figure~\ref{fig-synthetic-data} (see also
-Appendix A for a formal description). Samples from the negative class
-are marked in blue while samples of the positive class are marked in
-orange.
-
-\begin{figure}
-
-{\centering \includegraphics[width=8cm,height=2cm]{www/synthetic_data.png}
-
-}
-
-\caption{\label{fig-synthetic-data}Synthetic classification datasets
-used in our experiments.}
-
-\end{figure}
-
-Ex-ante we expect to see that Wachter will create a new cluster of
-counterfactual instances in the proximity of the initial decision
-boundary. Thus, the choice of a black-box model may have an impact on
-the paths of the recourse. For generators that use latent space search
-(\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on (and
-have access to) probabilistic models
-(\protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
-counterfactuals will end up in regions of the target domain that are
-densely populated by training samples. Finally, we expect to see the
-counterfactuals generated by DiCE to be uniformly spread around the
-feature space inside the target class.
-
-\hypertarget{real-world-data}{%
-\subsubsection{Real-world data}\label{real-world-data}}
-
-Additionally, we use two real-world datasets from the Finance domain.
-Firstly, we use the Give Me Some Credit dataset which was open-sourced
-on Kaggle for the task to predict whether a borrower is likely to
-experience financial difficulties in the next two years
-\protect\hyperlink{ref-gmsc_data}{{[}26{]}}. Originally consisting of
-250,000 instances with 11 numerical attributes, the dataset was randomly
-undersampled to result in a balanced subsample made up of 3000
-individuals. Secondly, we the German Credit dataset which involves the
-task of predicting if bank customers are credit-worthy or not
-\protect\hyperlink{ref-germancredit1994}{{[}27{]}}. It consists of 700
-positive and 300 negative instances charaterized by 7 numerical and 13
-categorical attributes. We process the dataset in two ways: (1) the
-values of the ``Personal status and sex'' feature are aggregated by the
-two represented genders; (2) the most common values are calculated for
-all categorical features such that a feature \(x_d\) with the mode
-\(\bar{x}_d\) is transformed into a new binary feature
-\(\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}\). Binarization ensures
-that we can use all counterfactual generators in the benchmark.
-
-\hypertarget{sec-empirical-2}{%
-\section{Experiments}\label{sec-empirical-2}}
-
-\hypertarget{endogenous-macrodynamics}{%
-\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
-
-\hypertarget{potential-mitigation-strategies}{%
-\subsection{Potential Mitigation
-Strategies}\label{potential-mitigation-strategies}}
-
 \hypertarget{sec-discussion}{%
@@
 While we believe that this work constitutes a valuable starting point
-for addressing existing issues in algorithmic recourse from a fresh
+for addressing existing issues in Algorithmic Recourse from a fresh
 perspective, we are aware of several of its limitations. In the
@@
 that may well be too simple: individuals seeking recourse are assumed to
-always implement the proposed counterfactual explanations; conversely,
+always implement the proposed Counterfactual Explanations; conversely,
 the agent in charge of the model \(M\) is assumed to always treat
@@
 
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-Largely in line with the existing literature on algorithmic recourse, we
+\hypertarget{sec-limit-data}{%
+\subsection{Data}\label{sec-limit-data}}
+
+Largely in line with the existing literature on Algorithmic Recourse, we
 have limited our analysis of real-world data to three commonly used
@@
 have also typically been analyzed in the existing literature on
-counterfactual explanations and algorithmic recourse, they represent
+Counterfactual Explanations and Algorithmic Recourse, they represent
 only a subset of popular machine learning models employed in practice -
@@
 tabular datasets, such as the ones considered here
-\protect\hyperlink{ref-grinsztajn2022tree}{{[}29{]}}.
+\protect\hyperlink{ref-grinsztajn2022tree}{{[}36{]}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
+\CSLLeftMargin{{[}24{]} }%
+\CSLRightInline{A.-K. Dombrowski, J. E. Gerken, and P. Kessel,
+{``Diffeomorphic explanations with normalizing flows,''} 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
+\CSLLeftMargin{{[}25{]} }%
+\CSLRightInline{R. S. Pindyck and D. L. Rubinfeld,
+\emph{Microeconomics}. Pearson Education, 2014.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}24{]} }%
+\CSLLeftMargin{{[}26{]} }%
 \CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}25{]} }%
+\CSLLeftMargin{{[}27{]} }%
 \CSLRightInline{S. Hanneke, {``A bound on the label complexity of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}26{]} }%
+\CSLLeftMargin{{[}28{]} }%
 \CSLRightInline{K. Competition, {``Give me some credit, improve on the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
-\CSLLeftMargin{{[}27{]} }%
-\CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
-\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}}
+\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
+\CSLLeftMargin{{[}29{]} }%
+\CSLRightInline{I.-C. Yeh and C. Lien, {``The comparisons of data mining
+techniques for the predictive accuracy of probability of default of
+credit card clients,''} \emph{Expert systems with applications}, vol.
+36, no. 2, pp. 2473--2480, 2009.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
+\CSLLeftMargin{{[}30{]} }%
+\CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine
+learning in python,''} \emph{the Journal of machine Learning research},
+vol. 12, pp. 2825--2830, 2011.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
+\CSLLeftMargin{{[}31{]} }%
+\CSLRightInline{R. K. Pace and R. Barry, {``Sparse spatial
+autoregressions,''} \emph{Statistics \& Probability Letters}, vol. 33,
+no. 3, pp. 291--297, 1997.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
+\CSLLeftMargin{{[}32{]} }%
+\CSLRightInline{C. M., {``Racist data destruction? - a boston housing
+dataset controversy,''} 2019, Available:
+\url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
+\CSLLeftMargin{{[}33{]} }%
+\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with
+flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
+\CSLLeftMargin{{[}34{]} }%
+\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell,
+{``Simple and scalable predictive uncertainty estimation using deep
+ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}28{]} }%
+\CSLLeftMargin{{[}35{]} }%
 \CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-\CSLLeftMargin{{[}29{]} }%
+\CSLLeftMargin{{[}36{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do

### 2022-08-12 dev/paper/sections/empirical.qmd

--- 2022-08-08 dev/paper/sections/empirical.qmd
@@
-# Experiments {#sec-empirical}
+# Experiment Setup {#sec-empirical}
 
-1.
-Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
-2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
-3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
-    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
-4. Mitigating strategy: penaliye distance from centroid.
+## Data {#sec-empirical-data}
+
+We have chosen to work with both synthetic and real-world datasets.
+Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
+Following @upadhyay2021towards, we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
+Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
+
+### Synthetic data
+
+```{julia}
+using Plots, PlotThemes
+theme(:wong)
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
+function plot_data(data,title)
+    plt = plot(title=title)
+    scatter!(data)
+    return plt
+end
+plts = [plot_data(data,name) for (name, data) in catalogue]
+plt = plot(plts..., layout=(1,4), size=(850,200))
+savefig(plt, "dev/paper/www/synthetic_data.png")
+```
+
+We use four synthetic binary classification datasets consisting of 1000 samples each.
+The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
+Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
+
+![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}
+
+Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
+Thus, the choice of a black-box model may have an impact on the paths of the recourse.
+For generators that use latent space search (REVISE @joshi2019towards, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
+Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated.
+We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.].
+In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concered with generating what we have defined as meaningful counterfactuals.
+
+### Real-world data
+
+We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification.
+Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data].
+Originally consisting of 250,000 instances with 11 numerical attributes.
+Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
+In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis.
+See @sec-limit-data for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
+Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+
+As a third dataset we include the **California Housing** dataset derived from the 1990 U.S.
+census and sourced through scikit-learn [@pedregosa2011scikit, @pace1997sparse].
+It consists of 8 continuous features that can be used to predict the median house price for California districts.
+The continuous outcome variable is binarized as $\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}$ indicating whether or not the median house price of a given district is above or below the median of all districts.
+While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example @schut2021generating).
+While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem [@carlisle2019racist].
+
+Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 10,000 individuals each.
+We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.
+
+## Classifiers and Generative Models {#sec-empirical-classifiers}
+
+For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
+We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
+They are also the model of choice in @schut2021generating.
+The actual neural network architectures are kept simple (@tbl-mlp), since we are only marginally concerned with achieving good initial classifier performance.
+For the real-world datasets we using mini-batch training and dropout regularization.
+
+The Latent Space generators rely on separate generative models.
+Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose.
+As with the classifiers, we deliberately choose to work with fairly simple architectures (@tbl-vae).
+More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
+But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.
+
+All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment.
+Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model.
+
+::: {#tbl-panel layout-ncol=1}
+
+| | Hidden Dim. | Hidden Layers | Batch | Dropout |
+|------|------|------|-----|-----|
+| Synthetic | 32 | 1 | - | - |
+| Real-World | 32 | 2 | 50 | 0.25 |
+
+: MLP {#tbl-mlp}
+
+|  | Hidden Dim. | Epochs |
+|------|------|------|
+| Synthetic | 2 | 100 | 
+| Real-World | 8 | 250 |
+
+: Variational Autoencoder {#tbl-vae}
+
+Model Architectures
+:::

### 2022-08-12 dev/paper/sections/empirical_2.qmd

@@
 ## Potential Mitigation Strategies
+
+### Gravitational Counterfactual Explanations {#sec-empirical-2-mitigate}
+
+A straight-forward choice simply extends the baseline approach by @wachter2017counterfactual: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{\mathbf{x}}$.
+For such a recourse objective, higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point in the target domain.
+In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+
+#### A note on convergence
+
+For this simple mitigating strategy underlying the Gravitational generator to work as expected, one needs to ensure that counterfactual search continues, even after a predetermined threshold probability $\gamma$ has potentially already been reached.
+@fig-convergence illustrates this distinction: if one chooses to terminate search once the desired threshold is reached (left panel) the gravitational pull towards $\bar{\mathbf{x}}$ is never actually satisfied (compare to right panel).
+More generally, if convergence is defined simply in terms of flipping the predicted label with some desired degree of confidence, this corresponds to essentially ignoring any parts of the counterfactual search objective that do not involve $\ell(M(f(s_k^\prime)),t)$ beyond that point.
+While this may be appropriate for some applications, in general this seems like an odd convention.
+Since we nonetheless seen convergence specified simply in terms of reaching the threshold probability in some places^[@joshi2019towards define convergence of Algorithm 1 in this way.
+The implementation of @wachter2017counterfactual in CARLA is also defined in this way.], we thought it worth making this distinction explicit.
+
+![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}

### 2022-08-12 dev/paper/sections/introduction.qmd

@@
 Despite their shortcomings, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons].
-The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently iterpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models.
+The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models.
 These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.
@@
 Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes.
-An example relevant to the Social Sciences is consumer credit: in this context AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
+An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
 A meaningful recourse recommendation for a denied applicant could be: *"If your net savings rate had been 10% of your monthly income instead of the actual 8%, your application would have been successful.
@@
 More recent work has compared different approaches within this static setting [@pawelczyk2021carla].
-In this work we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly?
+In this work, we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly?
 What types of dynamics are introduced and how do different counterfactual generators compare in this context?
 
-Research on algorithmic recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as **individual recourse** in some places.
-Arguably though, most real-world applications that warrant algorithmic recourse involve potentially large groups of individuals typically competing for scarce resources.
+Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as **individual recourse** in some places.
+Arguably though, most real-world applications that warrant Algorithmic Recourse involve potentially large groups of individuals typically competing for scarce resources.
 Our work demonstrates that in such scenarios, choices made by or for one single individual are likely to affect the broader collective of individuals in ways that current approaches to AR fail to account for.
@@
 To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
-Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on algorithmic recourse extending our focus from single individuals to groups.
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations that can be fully parallelized.
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups.
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
 Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
@@
 The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature.
-@sec-method presents our methodology and data.
-@sec-empirical presents our empirical findings which are then discussed in the broader context of the literature in @sec-discussion.
-We also point to some of the limitations or our work as well as avenues for future research in @sec-limit.
+@sec-method posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
+@sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR.
+@sec-empirical presents our experimental results for synthetic and real-world datasets.
+We also present evidence for the effectiveness of various proposed mitigation strategies in that section.
+Our findings are then discussed in the broader context of the literature in @sec-discussion.
+We also point to some of the limitations of our work as well as avenues for future research in @sec-limit.
 Finally, @sec-conclusion concludes.

### 2022-08-12 dev/paper/sections/limitations.qmd

@@
 
-While we believe that this work constitutes a valuable starting point for addressing existing issues in algorithmic recourse from a fresh perspective, we are aware of several of its limitations.
+While we believe that this work constitutes a valuable starting point for addressing existing issues in Algorithmic Recourse from a fresh perspective, we are aware of several of its limitations.
 In the following we highlight some of these limitations and point to avenues for future research.
@@
 In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021towards.
-We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed counterfactual explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
+We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
 Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary.
@@
 
-## Data
+## Data {#sec-limit-data}
 
-Largely in line with the existing literature on algorithmic recourse, we have limited our analysis of real-world data to three commonly used benchmark datasets that involve binary prediction tasks.
+Largely in line with the existing literature on Algorithmic Recourse, we have limited our analysis of real-world data to three commonly used benchmark datasets that involve binary prediction tasks.
 Future work may benefit from including novel datasets or extending the analysis to multi-class or regression problems, the latter arguably representing the most common objective in Finance and Economics.
@@
 For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks.
-While these sorts of classifiers have also typically been analyzed in the existing literature on counterfactual explanations and algorithmic recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
+While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
 Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here [@borisov2021deep, @grinsztajn2022tree].

### 2022-08-12 dev/paper/sections/methodology.qmd

@@
 
-In the following we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general to introduce the various counterfactual generators we have chosen to use in our experiments.
-We then describe the experimental setup in @sec-method-2-experiment and introduce several evaluation metrics used to benchmark the different generators.
+In this section we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general that encapsulates the various individual recourse methods we have chosen to use in our experiments.
+We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.
 
@@
 
-In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro.
-Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks.
-We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
+We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
+In particular, we include include the following generators in our simulation experiments below: **REVISE** [@joshi2019towards], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
+Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
+We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro.
+This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
+All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
+
+As we will see next, all of them can be described by the following generalized form of @eq-solution:
 
@@
 Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
+In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one.
+Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly.
+The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
+In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
+
+CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$.
+To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
+The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
+The authors show that in this setting the cost function $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+Schut et al.
+[@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$ we also implicitly minimize predictive entropy - as in CLUE.
+In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
+They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].
+
+Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals.
+Wachter et al.
+[@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
+In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $h(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ...
+, s_K$ that look as different from each other as possible.
+The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach.
+It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function @eq-general.
 
@@
 
-To explicitly address the notion that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:
+All of the different approaches introduced above tackle the problem of Algorithmic Recourse from the perspective of one single individual^[DiCE recognizes that different individuals may have different objective functions, but it does not address the interdependencies between different individuals.].
+To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:
 
@@
 
-Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalised.
-The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address social costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
-The trade-off between private and social costs is determined by the ratio between $\lambda_1$ and $\lambda_2$.
+Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized.
+The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
+The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics].
+We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
 As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
-A straight-forward choice simply extends the baseline approach by @wachter2017counterfactual: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{\mathbf{x}}$.
-For such a recourse objective, higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point in the target domain.
-In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
-
-## A note on convergence
-
-For this simple mitigating strategy underlying the Gravitational generator to work as expected, one needs to ensure that counterfactual search continues, even after a predetermined threshold probability $\gamma$ has potentially already been reached.
-@fig-convergence illustrates this distinction: if one chooses to terminate search once the desired threshold is reached (left panel) the gravitational pull towards $\bar{\mathbf{x}}$ is never actually satisfied (compare to right panel).
-More generally, if convergence is defined simply in terms of flipping the predicted label with some desired degree of confidence, this corresponds to essentially ignoring any parts of the counterfactual search objective that do not involve $\ell(M(f(s_k^\prime)),t)$ beyond that point.
-While this may be appropriate for some applications, in general this seems like an odd convention.
-Since we nonetheless seen convergence specified simply in terms of reaching the threshold probability in some places^[@joshi2019towards define convergence of Algorithm 1 in this way.
-The implementation of @wachter2017counterfactual in CARLA is also defined in this way.], we thought it worth making this distinction explicit.
-
-![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}
+That being said, we do propose a few potential mitigation strategies in @sec-empirical-2-mitigate.

### 2022-08-12 dev/paper/sections/related.qmd
some rejigging of sections

--- 2022-07-06 paper/sections/related.qmd
@@
-# Related Work {#sec-related}
+# Background {#sec-related}
 
 In this Section we provide a review of the relevant literature.
-First, @sec-related-recourse discusses the existing research within the domain of counterfactual explanations and algorithmic recourse.
+First, @sec-related-recourse discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
 Then, @sec-related-shifts presents some of the previous work on the measurement of dataset and model shifts.
@@
 Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.
-Then we are interested in minimizing the complexity or effort $H=h(x\prime)$ associated with moving an individual $x$ to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $t$:
+Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $H=h(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:
 
 $$
-\min_{x\prime \in \mathcal{x}} c(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
+\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
 $$ {#eq-obj}
@@
 $$
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),t) + \lambda h(x\prime)
+x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
 $$ {#eq-solution}
 
-In the baseline work and many subsequent approaches the complexity function $h: \mathcal{X} \mapsto \mathbb{R}$ is proxied by some distance metric based on the simple intuition that large perturbations of $x$ are costly.
+In the baseline work [@wachter2017counterfactual], the cost function $h: \mathcal{X} \mapsto \mathbb{R}$ is proxied by some distance metric based on the simple intuition that perturbations of $x$ are costly to the individual.
+For models that are differentiable and produce smooth predictions, @eq-solution can be solved through gradient descent.
+This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter** - the name of the first author - in the remainder of this paper.
 
-Many approaches for the generation of algorithmic recourse have been described in the literature.
+Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017.
 An October 2020 survey by Karimi et al.
@@
 
-Efforts so far have largely been directed at improving the quality of counterfactual explanations within a static context: given some pre-trained classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in generating one or multiple meaningful counterfactual explanations for some individual characterized by $x_i$.
-The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
+Efforts so far have largely been directed at improving the quality of Counterfactual Explanations within a static context: given some pre-trained classifier $M: \mathcal{X} \mapsto \mathcal{Y}$, we are interested in generating one or multiple meaningful Counterfactual Explanations for some individual characterized by $x$.
+The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
 We have been able to identify only one recent work that considers the implications of **exogenous** domain and model shifts in the context of AR [@upadhyay2021towards].
@@
 For example, they might stem from data correction, temporal shifts or geospatial changes [@upadhyay2021towards].
-The authors of [@upadhyay2021towards] propose framework for algorithmic recourse (ROAR) that evidently improves robustness to such exogenous shifts.
+The authors of [@upadhyay2021towards] propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
+
+As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances.
+We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals [@carrizosa2021generating].
+The authors propose an optimization framework that generates collective counterfactuals.
+We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.
 
@@
 
-In the context of algorithmic recourse, domain and model shifts were first brought up by the authors behind ROAR [@upadhyay2021towards].
+In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR [@upadhyay2021towards].
 In their work they refer to model shifts as simply any perturbation $\Delta$ to the parameters of the model in question: $M$.
@@
 
-...
+Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies.
+This may be partially due to limited software availability in this space.
+Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia.
+Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
+In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: `AlgorithmicRecourseDynamics.jl`^[The package is available from ...].
+The core concepts implemented in that package reflect what is presented in section @sec-method-2 of this paper.

### 2022-08-12 dev/paper/paper.tex
only findings left now

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 aims to simulate the process of Algorithmic Recourse in practice. We
-begin in the static setting at time time \(t=0\): given some pre-trained
-classifier \(M\) we generate recourse for a random batch of \(B\)
-individuals in the non-target class. Note that we focus our attention on
-classification problems, since classification poses the most common
-practical use-case for Algorithmic Recourse. In order to simulate the
-dynamic process, we suppose that the model \(M\) is retrained following
-the actual implementation of recourse in time \(t=0\). Following the
-update to the model, we assume that at time \(t=1\) recourse is
-generated for yet another random subset of individuals in the non-target
-class. This process is repeated for a number of time periods \(T\). To
-get a clean read on endogenous dynamics we keep the total population of
-samples closed: we allow existing samples to move from factual to
-counterfactual states, but do not allow any entirely new samples to
-enter the population. The experimental setup is summarized in Algorithm
+begin in the static setting at time time \(t=0\): given some classifier
+\(M\) that was pre-trained on data
+\(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse
+for a random batch of \(B\) individuals in the non-target class
+(\(\mathcal{D}_0\)). Note that we focus our attention on classification
+problems, since classification poses the most common practical use-case
+for Algorithmic Recourse. In order to simulate the dynamic process, we
+suppose that the model \(M\) is retrained following the actual
+implementation of recourse in time \(t=0\). Following the update to the
+model, we assume that at time \(t=1\) recourse is generated for yet
+another random subset of individuals in the non-target class. This
+process is repeated for a number of time periods \(T\). To get a clean
+read on endogenous dynamics we keep the total population of samples
+closed: we allow existing samples to move from factual to counterfactual
+states, but do not allow any entirely new samples to enter the
+population. The experimental setup is summarized in Algorithm
 \ref{algo-experiment}
@@
 \begin{algorithmic}[1]
-\Procedure{Experiment}{$M,D,G$}
+\Procedure{Experiment}{$M,\mathcal{D},G$}
 \State $t\gets 0$
 \While{$t<T$}
-\State $D_B \subset D$
-\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
-\State $M\gets M(D)$ \Comment{Retrain model.}
+\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
+\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
+\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
 \EndWhile
-\State \textbf{return} $M,D$
+\State \textbf{return} $M,\mathcal{D}$
 \EndProcedure
@@
 experiments, we choose the values such that \(T \cdot B\) corresponds to
-the application of recourse on \(25-50\%\) of the negative instances
+the application of recourse on \(\approx50\%\) of the negative instances
 from the initial dataset. As we collect data at each time \(t\), we can
@@
 Algorithm \ref{algo-experiment} summarizes the proposed simulation
-experiment for a given dataset \(D\), model \(M\) and generator \(G\),
-but naturally we are interested in comparing simulation outcomes for
-different sources of data, models and generators. The framework we have
-built facilitates this, making use of multi-threading in order to speed
-up computations. Holding the initial model and dataset constant the
-experiments are run for all generators, since our primary concern is to
-benchmark different recourse methods. To ensure that each generator is
-faced with the exact same initial conditions in each round \(t\), the
-candidate batch of individuals from the non-target class is randomly
-drawn from the intersection of all individuals in the non-target class
-across all experiments
-\(\left\{\textsc{Experiment}(M,D,G)\right\}_{j=1}^J\) where \(J\) is the
-total number of generators.
+experiment for a given dataset \(\mathcal{D}\), model \(M\) and
+generator \(G\), but naturally we are interested in comparing simulation
+outcomes for different sources of data, models and generators. The
+framework we have built facilitates this, making use of multi-threading
+in order to speed up computations. Holding the initial model and dataset
+constant the experiments are run for all generators, since our primary
+concern is to benchmark different recourse methods. To ensure that each
+generator is faced with the exact same initial conditions in each round
+\(t\), the candidate batch of individuals from the non-target class is
+randomly drawn from the intersection of all individuals in the
+non-target class across all experiments
+\(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where
+\(J\) is the total number of generators.
 
@@
 where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
-Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
-identically distributed samples drawn from probability distributions
-\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
-between the kernel mean embeddings of \(p\) and \(q\) in RKHS
-\(\mathcal{H}\). An important consideration is the choice of the kernel
-function \(k(\cdot,\cdot)\). In our implementation we make use of the
-radial basis function (RBF) kernel with a constant length-scale
-parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
-and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
-\(X=\tilde{X}\).
+Space \(\mathcal{H}\)
+\protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}, and \(X\),
+\(\tilde{X}\) represent independent and identically distributed samples
+drawn from probability distributions \(\mathcal{X}\) and
+\(\mathcal{\tilde{X}}\) respectively
+\protect\hyperlink{ref-gretton2012kernel}{{[}27{]}}. MMD is a measure of
+the distance between the kernel mean embeddings of \(\mathcal{X}\) and
+\(\mathcal{\tilde{X}}\) in RKHS \(\mathcal{H}\). An important
+consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
+In our implementation we make use of Gaussian kernel with a constant
+length-scale parameter of \(0.5\). As the Gaussian kernel captures all
+moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we
+have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if \(X=\tilde{X}\).
 
@@
 samples \(X\) and \(\tilde{X}\) were drawn from the same probability
-distribution we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
+distribution, we follow
+\protect\hyperlink{ref-arcones1992bootstrap}{{[}28{]}}. To that end, we
 combine the two samples and generate a large number of permutations of
@@
 truth labels, i.e.~the labels that samples were assigned in time
-\(t=0\). Throughout our experiments we generally do not expect the
+\(t=0\). Throughout our experiments, we generally do not expect the
 distribution of the negative class to change over time -- application of
@@
 
-Finally, \textbf{feature mean and feature standard deviation} are also
-calculated to verify how the implementation of recourse impacts every
-attribute in the dataset. Although MMD already captures information
-about the expected value and variance, we may also be interested in a
-more granular look at individual features.
-
 \hypertarget{model-shifts}{%
@@
 probabilities assigned to each sample will change; again, this metric
-will equal 0 only if the two classifiers are the same. It is worth
-noting that while we apply the technique to samples drawn uniformly from
-the dataset, it can also be employed on arbitrary points in the entire
-feature space (or a subspace). The latter approach is theoretically more
-robust. Unfortunately, in practice this approachs suffers from the curse
-of dimensionality, since it becomes increasingly difficult to select
-enough points to overcome noise as the dimension \(D\) grows.
+will equal 0 only if the two classifiers are the same. We compute PP MMD
+in two ways: firstly, we compute it over samples drawn uniformly from
+the dataset, and, secondly, we compute it over points spanning a
+meshgrid over a subspace of the entire feature space. For the latter
+approach we bound the subspace by the extrema of each feature. While
+this approach is theoretically more robust, unfortunately, it suffers
+from the curse of dimensionality, since it becomes increasingly
+difficult to select enough points to overcome noise as the dimension
+\(D\) grows.
 
@@
 \textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}29{]}} and
 estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
@@
 
+This section presents the exact ingredients and parameter choices
+describing the simulation experiments we ran to produce the findings
+presented in the next section (Section~\ref{sec-empirical-2}). For
+convenience, we use Algorithm \ref{algo-experiment} as a template to
+guide us through this section. A few high-level details upfront: each
+experiment is run for a total of \(T=50\) rounds, where in each round we
+provide recourse to five percent of all individuals in the non-target
+class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in
+  the previous section, we end up providing recourse to a total of
+  \(\approx50\%\) by the end of round \(T=50\).}. All classifiers and
+generative models are retrained for 10 epochs in each round \(t\) of the
+experiment. Rather than retraining models from scratch, we initialize
+all parameters at their previous levels (\(t-1\)) and compute
+backpropagate for 10 epochs using the new training data as inputs into
+the existing model. Evaluation metrics are computed and stored every 10
+rounds.
+
+\hypertarget{sec-empirical-classifiers}{%
+\subsection{\texorpdfstring{\(M\) -- Classifiers and Generative
+Models}{M -- Classifiers and Generative Models}}\label{sec-empirical-classifiers}}
+
+For each dataset and generator we look at three different types of
+classifiers all of them built and trained using \texttt{Flux.jl}
+\protect\hyperlink{ref-innes2018fashionable}{{[}30{]}}: firstly, a
+simple linear classifier - \textbf{Logistic Regression} - implemented as
+single linear layer with sigmoid activation; secondly, a multilayer
+perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
+composed of five MLPs following
+\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}31{]}} that serves
+as our only probabilistic classifier. We have chosen to work with deep
+ensembles both for their simplicity and effectiveness at modelling
+predictive uncertainty. They are also the model of choice in
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
+network architectures are kept simple (Table~\ref{tbl-mlp}), since we
+are only marginally concerned with achieving good initial classifier
+performance. For the real-world datasets we using mini-batch training
+and dropout regularization.
+
+The Latent Space generators rely on separate generative models.
+Following the authors of both REVISE and CLUE we use Variational
+Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers,
+we deliberately choose to work with fairly simple architectures
+(Table~\ref{tbl-vae}). More expressive generative models generally also
+lead to more meaningful counterfactuals produced by Latent Space
+generators. But in our view this should simply be considered as a
+vulnerability of counterfactual generators that rely on surrogate models
+to learn what realistic representations of the underlying data.
+
+\begin{table}
+
+\caption{\label{tbl-panel}Model
+Architectures}\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-mlp}MLP}
+
+{\centering
+
+\begin{tabular}[t]{lllll}
+\toprule
+ & Hidden Dim. & Hidden Layers & Batch & Dropout\\
+\midrule
+Synthetic & 32 & 1 & - & -\\
+Real-World & 32 & 2 & 50 & 0.25\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{minipage}%
+\newline
+\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-vae}Variational Autoencoder}
+
+{\centering
+
+\begin{tabular}[t]{lll}
+\toprule
+ & Hidden Dim. & Epochs\\
+\midrule
+Synthetic & 2 & 100\\
+Real-World & 8 & 250\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{minipage}%
+
+\end{table}
+
 \hypertarget{sec-empirical-data}{%
-\subsection{Data}\label{sec-empirical-data}}
+\subsection{\texorpdfstring{\(\mathcal{D}\) --
+Data}{\textbackslash mathcal\{D\} -- Data}}\label{sec-empirical-data}}
 
@@
 borrower is likely to experience financial difficulties in the next two
-years \protect\hyperlink{ref-gmsc_data}{{[}28{]}}. Originally consisting
+years \protect\hyperlink{ref-gmsc_data}{{[}32{]}}. Originally consisting
 of 250,000 instances with 11 numerical attributes. Secondly, we use the
 \textbf{UCI defaultCredit} dataset
-\protect\hyperlink{ref-yeh2009comparisons}{{[}29{]}}, a benchmark
+\protect\hyperlink{ref-yeh2009comparisons}{{[}33{]}}, a benchmark
 dataset that can be used to train binary classifiers to predict the
@@
 derived from the 1990 U.S. census and sourced through scikit-learn
-\protect\hyperlink{ref-pace1997sparse}{{[}31{]}}. It consists of 8
+\protect\hyperlink{ref-pace1997sparse}{{[}35{]}}. It consists of 8
 continuous features that can be used to predict the median house price
@@
 discarded this dataset, since it has been found to suffer from an
-ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}32{]}}.
+ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}36{]}}.
 
@@
 
-\hypertarget{sec-empirical-classifiers}{%
-\subsection{Classifiers and Generative
-Models}\label{sec-empirical-classifiers}}
-
-For each dataset and generator we look at three different types of
-classifiers all of them built and trained using \texttt{Flux.jl}
-\protect\hyperlink{ref-innes2018fashionable}{{[}33{]}}: firstly, a
-simple linear classifier - \textbf{Logistic Regression} - implemented as
-single linear layer with sigmoid activation; secondly, a multilayer
-perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
-composed of five MLPs following
-\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}34{]}} that serves
-as our only probabilistic classifier. We have chosen to work with deep
-ensembles both for their simplicity and effectiveness at modelling
-predictive uncertainty. They are also the model of choice in
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
-network architectures are kept simple (Table~\ref{tbl-mlp}), since we
-are only marginally concerned with achieving good initial classifier
-performance. For the real-world datasets we using mini-batch training
-and dropout regularization.
-
-The Latent Space generators rely on separate generative models.
-Following the authors of both REVISE and CLUE we use Variational
-Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers,
-we deliberately choose to work with fairly simple architectures
-(Table~\ref{tbl-vae}). More expressive generative models generally also
-lead to more meaningful counterfactuals produced by Latent Space
-generators. But in our view this should simply be considered as a
-vulnerability of counterfactual generators that rely on surrogate models
-to learn what realistic representations of the underlying data.
-
-All classifiers and generative models are retrained for 10 epochs in
-each round \(t\) of the experiment. Rather than retraining models from
-scratch, we initialize all parameters at their previous levels (\(t-1\))
-and compute backpropagate for 10 epochs using the new training data as
-inputs into the existing model.
-
-\begin{table}
-
-\caption{\label{tbl-panel}Model
-Architectures}\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-mlp}MLP}
-
-{\centering
-
-\begin{tabular}[t]{lllll}
-\toprule
- & Hidden Dim. & Hidden Layers & Batch & Dropout\\
-\midrule
-Synthetic & 32 & 1 & - & -\\
-Real-World & 32 & 2 & 50 & 0.25\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-vae}Variational Autoencoder}
-
-{\centering
-
-\begin{tabular}[t]{lll}
-\toprule
- & Hidden Dim. & Epochs\\
-\midrule
-Synthetic & 2 & 100\\
-Real-World & 8 & 250\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{minipage}%
-
-\end{table}
+\hypertarget{g-generators}{%
+\subsection{\texorpdfstring{\(G\) --
+Generators}{G -- Generators}}\label{g-generators}}
+
+All generators introduced earlier are included in the experiments:
+Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
+REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
+\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DiCE
+\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and Greedy
+\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. In addition, we
+introduce two new generators in Section~\ref{sec-empirical-2-mitigate}
+that directly address the issue of endogenous domain and model shifts.
+We also test to what extent it may be beneficial to combine ideas
+underlying the various generators.
 
@@
 
-\hypertarget{potential-mitigation-strategies}{%
+\hypertarget{sec-empirical-2-mitigate}{%
 \subsection{Potential Mitigation
-Strategies}\label{potential-mitigation-strategies}}
-
-\hypertarget{sec-empirical-2-mitigate}{%
+Strategies}\label{sec-empirical-2-mitigate}}
+
+In the following we explore several ideas for strategies that may help
+to mitigate those types of endogenous recourse dynamics we observed in
+the previous subsection. All of them essentially boil down to one simple
+principle: to avoid substantial domain and model shifts, the generated
+counterfactuals should comply as much as possible with the true data
+generating process. This principle is really at the core of Latent Space
+generators, and hence it is not surprising that we have found these
+types of generators to perform comparably well in the previous
+subsection. But as we have mentioned earlier, generators that rely on
+separate generative models carry an additional computational cost and -
+perhaps more importantly - their performance hinges on the performance
+of said generative models. Fortunately, it turns out that we can use a
+number of other, much simpler strategies, which we will discuss now.
+
+\hypertarget{more-conservative-decision-thresholds}{%
+\subsubsection{More Conservative Decision
+Thresholds}\label{more-conservative-decision-thresholds}}
+
+The most obvious and trivial mitigation strategy is to simply choose a
+higher decision threshold \(\gamma\). Under \(\gamma=0.5\),
+counterfactuals will end up near the decision boundary by construction.
+Since this is the region of maximal aleotoric uncertainty, the
+classifier is bound to be thrown off. By simply setting a more
+conservative decision threshold, we can avoid this issue to some extent.
+A potential drawback of this approach is that a classifier with high
+decisiveness may classify samples with high confidence even far away
+from the training data.
+
+\hypertarget{classifier-preserving-roar}{%
+\subsubsection{Classifier Preserving
+ROAR}\label{classifier-preserving-roar}}
+
+Another potential strategy draws inspiration from ROAR
+\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}: to preserve the
+classifier, we propose to simply explicitly penalize the loss it incurs
+when evaluated on the counterfactual \(x^\prime\) at given parameter
+values. Recall that \(g(\cdot)\) denotes what we had defined as the
+external cost in Equation~\ref{eq-collective}. Then formally we let
+
+\begin{equation}\protect\hypertarget{eq-clap}{}{
+\begin{aligned}
+g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
+\end{aligned}
+}\label{eq-clap}\end{equation}
+
+for each counterfactual \(k\) where \(l\) denotes the loss function used
+to train \(M\). This approach, which we shall refer to as
+\textbf{ClapROAR}, is based on the intuition that (endogenous) model
+shifts will be triggered by counterfactuals that increase classifier
+loss. It is closely linked to the idea of choosing higher decision
+threshold, but likely better at avoiding the potential pitfalls
+associated with highly decisive classifiers. It also makes the private
+vs.~external cost trade-off more explicit and hence manageable.
+
+\hypertarget{gravitational-counterfactual-explanations}{%
 \subsubsection{Gravitational Counterfactual
-Explanations}\label{sec-empirical-2-mitigate}}
-
-A straight-forward choice simply extends the baseline approach by
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead of
-only penalizing the distance of the individuals' counterfactual to its
+Explanations}\label{gravitational-counterfactual-explanations}}
+
+Yet another strategy simply extends Wachter as follows: instead of only
+penalizing the distance of the individuals' counterfactual to its
 factual, we propose penalizing its distance to some sensible point in
-the target domain, for example the sample average: \(\bar{\mathbf{x}}\).
-For such a recourse objective, higher choices of \(\lambda_2\) relative
-to \(\lambda_1\) will lead counterfactuals to gravitate towards the
-specified point in the target domain. In the remainder of this paper we
-will therefore refer to this approach as \textbf{Gravitational}
-generator, when we investigate its potential usefulness for mitigating
-endongenous macrodynamics\footnote{Note that despite the naming
-  convention our goal here is not to provide yet another counterfactual
-  generator, but merely investigate the most simple penalty we can think
-  of with respect to its effectiveness.}.
-
-\hypertarget{a-note-on-convergence}{%
-\paragraph{A note on convergence}\label{a-note-on-convergence}}
-
-For this simple mitigating strategy underlying the Gravitational
-generator to work as expected, one needs to ensure that counterfactual
-search continues, even after a predetermined threshold probability
-\(\gamma\) has potentially already been reached.
-Figure~\ref{fig-convergence} illustrates this distinction: if one
-chooses to terminate search once the desired threshold is reached (left
-panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
-actually satisfied (compare to right panel). More generally, if
-convergence is defined simply in terms of flipping the predicted label
-with some desired degree of confidence, this corresponds to essentially
-ignoring any parts of the counterfactual search objective that do not
-involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
-be appropriate for some applications, in general this seems like an odd
-convention. Since we nonetheless seen convergence specified simply in
-terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
-  define convergence of Algorithm 1 in this way. The implementation of
-  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
-  also defined in this way.}, we thought it worth making this
-distinction explicit.
+the target domain, for example the sample average: \(\bar{x}\):
+
+\begin{equation}\protect\hypertarget{eq-clap}{}{
+\begin{aligned}
+g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
+\end{aligned}
+}\label{eq-clap}\end{equation}
+
+Once again we can putting this in the context of
+Equation~\ref{eq-collective}, the former penalty can be thought of here
+as the private cost incurred by the individual, while the latter
+reflects the external cost incurred by other individuals. Higher choices
+of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
+gravitate towards the specified point \(\bar{x}\) in the target domain.
+In the remainder of this paper we will therefore refer to this approach
+as \textbf{Gravitational} generator, when we investigate its potential
+usefulness for mitigating endongenous macrodynamics\footnote{Note that
+  despite the naming convention our goal here is not to provide yet
+  another counterfactual generator, but merely investigate the most
+  simple penalty we can think of with respect to its effectiveness.}.
+
+Figure~\ref{fig-mitigation} shows an illustrative example that
+demonstrates the differences in counterfactual outcomes when using the
+various mitigation strategies compared to the baseline approach, that is
+Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes
+the counterfactual a little further into the target domain; this effect
+is even stronger for ClapROAR; finally, using the Gravitational
+generator the counterfactual ends up all the way inside the target
+domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the
+  Graviational generator and ClapROAR to work as expected, one needs to
+  ensure that counterfactual search continues, independent of the
+  threshold probability \(\gamma\).}.
 
@@
 
-{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/mitigation.png}
 
@@
 
-\caption{\label{fig-convergence}Comparison of counterfactual search
-outcome with simple (left) and strict convergence (right).}
+\caption{\label{fig-mitigation}The differences in counterfactual
+outcomes when using the various mitigation strategies compared to the
+baseline approach, that is Wachter with \(\gamma=0.5\)}
 
 \end{figure}
+
+Our findings \ldots{}
 
@@
 tabular datasets, such as the ones considered here
-\protect\hyperlink{ref-grinsztajn2022tree}{{[}36{]}}.
+\protect\hyperlink{ref-grinsztajn2022tree}{{[}38{]}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
+\CSLLeftMargin{{[}26{]} }%
+\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing
+kernel hilbert spaces in probability and statistics}. Springer Science
+\& Business Media, 2011.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
+\CSLLeftMargin{{[}27{]} }%
+\CSLRightInline{A. Gretton, K. M. Borgwardt, M. J. Rasch, B. Schölkopf,
+and A. Smola, {``A kernel two-sample test,''} \emph{The Journal of
+Machine Learning Research}, vol. 13, no. 1, pp. 723--773, 2012.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}26{]} }%
+\CSLLeftMargin{{[}28{]} }%
 \CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}27{]} }%
+\CSLLeftMargin{{[}29{]} }%
 \CSLRightInline{S. Hanneke, {``A bound on the label complexity of
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
+\CSLLeftMargin{{[}30{]} }%
+\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with
+flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
+\CSLLeftMargin{{[}31{]} }%
+\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell,
+{``Simple and scalable predictive uncertainty estimation using deep
+ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}28{]} }%
+\CSLLeftMargin{{[}32{]} }%
 \CSLRightInline{K. Competition, {``Give me some credit, improve on the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
-\CSLLeftMargin{{[}29{]} }%
+\CSLLeftMargin{{[}33{]} }%
 \CSLRightInline{I.-C. Yeh and C. Lien, {``The comparisons of data mining
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
-\CSLLeftMargin{{[}30{]} }%
+\CSLLeftMargin{{[}34{]} }%
 \CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
-\CSLLeftMargin{{[}31{]} }%
+\CSLLeftMargin{{[}35{]} }%
 \CSLRightInline{R. K. Pace and R. Barry, {``Sparse spatial
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
-\CSLLeftMargin{{[}32{]} }%
+\CSLLeftMargin{{[}36{]} }%
 \CSLRightInline{C. M., {``Racist data destruction? - a boston housing
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
-\CSLLeftMargin{{[}33{]} }%
-\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with
-flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
-\CSLLeftMargin{{[}34{]} }%
-\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell,
-{``Simple and scalable predictive uncertainty estimation using deep
-ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}35{]} }%
+\CSLLeftMargin{{[}37{]} }%
 \CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-\CSLLeftMargin{{[}36{]} }%
+\CSLLeftMargin{{[}38{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do

### 2022-08-12 dev/paper/sections/empirical.qmd
only findings left now

@@
 
-## Data {#sec-empirical-data}
+This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (@sec-empirical-2).
+For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section.
+A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.].
+All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment.
+Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model.
+Evaluation metrics are computed and stored every 10 rounds.
+
+## $M$ -- Classifiers and Generative Models {#sec-empirical-classifiers}
+
+For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
+We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
+They are also the model of choice in @schut2021generating.
+The actual neural network architectures are kept simple (@tbl-mlp), since we are only marginally concerned with achieving good initial classifier performance.
+For the real-world datasets we using mini-batch training and dropout regularization.
+
+The Latent Space generators rely on separate generative models.
+Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose.
+As with the classifiers, we deliberately choose to work with fairly simple architectures (@tbl-vae).
+More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
+But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.
+
+::: {#tbl-panel layout-ncol=1}
+
+| | Hidden Dim. | Hidden Layers | Batch | Dropout |
+|------|------|------|-----|-----|
+| Synthetic | 32 | 1 | - | - |
+| Real-World | 32 | 2 | 50 | 0.25 |
+
+: MLP {#tbl-mlp}
+
+|  | Hidden Dim. | Epochs |
+|------|------|------|
+| Synthetic | 2 | 100 | 
+| Real-World | 8 | 250 |
+
+: Variational Autoencoder {#tbl-vae}
+
+Model Architectures
+:::
+
+## $\mathcal{D}$ -- Data {#sec-empirical-data}
 
@@
 
-## Classifiers and Generative Models {#sec-empirical-classifiers}
+## $G$ -- Generators
 
-For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
-We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
-They are also the model of choice in @schut2021generating.
-The actual neural network architectures are kept simple (@tbl-mlp), since we are only marginally concerned with achieving good initial classifier performance.
-For the real-world datasets we using mini-batch training and dropout regularization.
-
-The Latent Space generators rely on separate generative models.
-Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose.
-As with the classifiers, we deliberately choose to work with fairly simple architectures (@tbl-vae).
-More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
-But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.
-
-All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment.
-Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model.
-
-::: {#tbl-panel layout-ncol=1}
-
-| | Hidden Dim. | Hidden Layers | Batch | Dropout |
-|------|------|------|-----|-----|
-| Synthetic | 32 | 1 | - | - |
-| Real-World | 32 | 2 | 50 | 0.25 |
-
-: MLP {#tbl-mlp}
-
-|  | Hidden Dim. | Epochs |
-|------|------|------|
-| Synthetic | 2 | 100 | 
-| Real-World | 8 | 250 |
-
-: Variational Autoencoder {#tbl-vae}
-
-Model Architectures
-:::
+All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating].
+In addition, we introduce two new generators in @sec-empirical-2-mitigate that directly address the issue of endogenous domain and model shifts.
+We also test to what extent it may be beneficial to combine ideas underlying the various generators.

### 2022-08-12 dev/paper/sections/empirical_2.qmd
only findings left now

--- 2022-08-11 dev/paper/sections/empirical_2.qmd
@@
 
-## Potential Mitigation Strategies
+## Potential Mitigation Strategies {#sec-empirical-2-mitigate}
+
+In the following we explore several ideas for strategies that may help to mitigate those types of endogenous recourse dynamics we observed in the previous subsection.
+All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
+This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous subsection.
+But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational cost and - perhaps more importantly - their performance hinges on the performance of said generative models.
+Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.
+
+### More Conservative Decision Thresholds
+
+The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$.
+Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction.
+Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off.
+By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
+A potential drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
+
+### Classifier Preserving ROAR
+
+Another potential strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
+Recall that $g(\cdot)$ denotes what we had defined as the external cost in @eq-collective.
+Then formally we let
+
+$$
+\begin{aligned}
+g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
+\end{aligned}
+$$ {#eq-clap}
+
+for each counterfactual $k$ where $l$ denotes the loss function used to train $M$.
+This approach, which we shall refer to as **ClapROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
+It also makes the private vs.
+external cost trade-off more explicit and hence manageable.
+
+### Gravitational Counterfactual Explanations
+
+Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{x}$:
+
+$$
+\begin{aligned}
+g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
+\end{aligned}
+$$ {#eq-clap}
+
+Once again we can putting this in the context of @eq-collective, the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
+Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain.
+In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+
+@fig-mitigation shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
+
+```{julia}
+#| eval: false
+using Random
+Random.seed!(2022)
+
+# Data:
+using MLJ
+N = 1000
+X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-5 => 5), cluster_std=0.5)
+ys .= ys.==2
+X = X'
+xs = Flux.unstack(X,2)
+data = zip(xs,ys)
+counterfactual_data = CounterfactualData(X,ys')
+
+# Models:
+using AlgorithmicRecourseDynamics
+M = AlgorithmicRecourseDynamics.Models.FluxModel(counterfactual_data)
+M = AlgorithmicRecourseDynamics.Models.train(M, counterfactual_data)
+
+# Generators:
+generators = Dict(
+    "Generic (γ=0.5)" => GenericGenerator(decision_threshold=0.5),
+    "Generic (γ=0.9)" => GenericGenerator(decision_threshold=0.9),
+    "Gravitational" => GravitationalGenerator(),
+    "ClapROAR" => ClapROARGenerator()
+)
+
+# Counterfactuals
+x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
+y = round(probs(M, x)[1])
+target = ifelse(y==1.0,0.0,1.0) # opposite label as target
+T = 50
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; T=T, latent_space=false) for (name, gen) in generators])
+
+# Plots:
+plts = []
+for (name,ce) ∈ counterfactuals
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false)
+    plts = vcat(plts..., plt)
+end
+plt = plot(plts..., size=(750,200), layout=(1,4))
+savefig(plt, "dev/paper/www/mitigation.png")
+```
+
+![The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$](www/mitigation.png){#fig-mitigation fig.pos="h" width=45%}
+
+Our findings ...

### 2022-08-12 dev/paper/sections/methodology_2.qmd
only findings left now

@@
 The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
-We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class.
+We begin in the static setting at time time $t=0$: given some classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$ we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
 Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse.
@@
 At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions.
-In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $25-50\%$ of the negative instances from the initial dataset.
+In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $\approx50\%$ of the negative instances from the initial dataset.
 As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
 
-Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $D$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators.
+Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $\mathcal{D}$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators.
 The framework we have built facilitates this, making use of multi-threading in order to speed up computations.
 Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
-To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,D,G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
+To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
 
@@
 
-where $\mathcal{F}$ is a unit ball in a Reproducing Kernel Hilbert Space H [27], and $X$, $\tilde{X}$ represent independent and identically distributed samples drawn from probability distributions $p$ and $q$ respectively [28].
-MMD is a measure of the distance between the kernel mean embeddings of $p$ and $q$ in RKHS $\mathcal{H}$.
+where $\mathcal{F}$ is a unit ball in a Reproducing Kernel Hilbert Space $\mathcal{H}$ @berlinet2011reproducing, and $X$, $\tilde{X}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel.
+MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in RKHS $\mathcal{H}$.
 An important consideration is the choice of the kernel function $k(\cdot,\cdot)$.
-In our implementation we make use of the radial basis function (RBF) kernel with a constant length-scale parameter of $0.5$.
-As RBF captures all moments of distributions $p$ and $q$, we have that $MMD_u^2[F,X,\tilde{X}]=0$ if and only if $X=\tilde{X}$.
+In our implementation we make use of Gaussian kernel with a constant length-scale parameter of $0.5$.
+As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD_u^2[F,X,\tilde{X}]=0$ if and only if $X=\tilde{X}$.
 
 The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment.
-To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution we follow @arcones1992bootstrap.
+To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap.
 To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$.
@@
 the labels that samples were assigned in time $t=0$.
-Throughout our experiments we generally do not expect the distribution of the negative class to change over time – application of recourse reduces the size of this class, but since individuals are sampled uniformly the distribution should remain unaffected.
+Throughout our experiments, we generally do not expect the distribution of the negative class to change over time – application of recourse reduces the size of this class, but since individuals are sampled uniformly the distribution should remain unaffected.
 Conversely, unless a recourse generator can perfectly replicate the original probability distribution, we expect the MMD of the positive class to increase.
 Thus, when discussing MMD, we generally mean the shift in the distribution of the positive class.
-
-Finally, **feature mean and feature standard deviation** are also calculated to verify how the implementation of recourse impacts every attribute in the dataset.
-Although MMD already captures information about the expected value and variance, we may also be interested in a more granular look at individual features.
 
@@
 If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same.
-It is worth noting that while we apply the technique to samples drawn uniformly from the dataset, it can also be employed on arbitrary points in the entire feature space (or a subspace).
-The latter approach is theoretically more robust.
-Unfortunately, in practice this approachs suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.
+We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a meshgrid over a subspace of the entire feature space.
+For the latter approach we bound the subspace by the extrema of each feature.
+While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.
 

### 2022-08-12 dev/notebooks/mitigation_strategies.qmd

@@
 )
-Serialization.serialize(joinpath(output_path,"results.jls"),results)
+Serialization.serialize(joinpath(output_path,"results_synthetic.jls"),results)
 ```
@@
 using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+results = Serialization.deserialize(joinpath(output_path,"results_synthetic.jls"))
 ```
@@
 )
-Serialization.serialize(joinpath(output_path,"results.jls"),results)
+Serialization.serialize(joinpath(output_path,"results_real_world.jls"),results)
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_real_world.jls"))
 ```

### 2022-08-12 dev/notebooks/mitigation_strategies.qmd
uh

@@
     :Gravitational=>GravitationalGenerator(),
-    :ROAR=>ClapROARGenerator()
+    :ClapROAR=>ClapROARGenerator()
 )
@@
 ```
-
-```{julia}
-
-```

### 2022-08-12 dev/notebooks/synthetic.qmd

@@
 for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
+    plt = plot(res)
+    savefig(plt, joinpath(www_path, "line_chart_$(data_name).png"))
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    savefig(plt, joinpath(www_path, "errorbar_chart_$(data_name).png"))
+    errorbar_charts[data_name] = plt
 end

### 2022-08-12 dev/paper/paper.tex
almost done, findings still need to be fleshed out though

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 state-of-the-art models like Deep Neural Networks (DNN) have also given
-rise a new type of principal-agent problem in the context of data-driven
-decision-making. It involves a group of \textbf{principals} - i.e.~human
-stakeholders - that fail to understand the behaviour of their
-\textbf{agent} - i.e.~the model used for automated decision-making
-\protect\hyperlink{ref-borch2022machine}{{[}1{]}}.
-
-Models or algorithms that fall into this category are typically referred
-to \textbf{black-box} models. Despite their shortcomings, black-box
-models have grown in popularity in recent years and have at times
-created undesirable societal outcomes
-\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
-has tackled this issue from two different angles: while some have
-appealed for a strict focus on inherently interpretable models
-\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
-different ways to explain the behaviour of black-box models. These two
-sub-domains can be broadly referred to as \textbf{interpretable AI} and
-\textbf{explainable AI} (XAI), respectively.
+rise a new type of problem in the context of data-driven
+decision-making. They are essentially \textbf{black boxes}: so complex,
+opaque and underspecified in the data that it is often impossible to
+understand how they actually arrive at their decision without auxiliary
+tools. Despite this shortcoming, black-box models have grown in
+popularity in recent years and have at times created undesirable
+societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}. The
+scientific community has tackled this issue from two different angles:
+while some have appealed for a strict focus on inherently interpretable
+models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have
+investigated different ways to explain the behaviour of black-box
+models. These two sub-domains can be broadly referred to as
+\textbf{interpretable AI} and \textbf{explainable AI} (XAI),
+respectively.
 
@@
 work has compared different approaches within this static setting
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work, we go
+\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}. In this work, we go
 one step further and ask ourselves: what happens if recourse is provided
@@
 involving a probabilistic classifier and the counterfactual generator
-proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
+proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}:
 the implementation of AR for a subset of individuals leads to a domain
@@
 \caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
-simple Bayesian model trained for binary classification (a); the
+simple linear classifier trained for binary classification (a); the
 implementation of AR for a random subset of individuals leads to a
-domain shift (b); as the classifier is retrained we observe a model
-shift (c); as this process is repeated, the decision boundary moves away
-from the target class (d).}
+noticable domain shift (b); as the classifier is retrained we observe a
+corresponding model shift (c); as this process is repeated, the decision
+boundary moves away from the target class (d).}
 
@@
 experimental framework extending previous work by
-\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
+\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}},
 which enables us to study macrodynamics of Algorithmic Recourse through
@@
 dynamics induced by various popular counterfactual generators including
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}},
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}},
+\protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
+\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and
+\protect\hyperlink{ref-antoran2020getting}{{[}9{]}}. To this end we
 propose a number of novel evaluation metrics that can be used to
@@
 experimental framework for modeling endogenous macrodynamics in AR.
-Section~\ref{sec-empirical} presents our experimental results for
-synthetic and real-world datasets. We also present evidence for the
-effectiveness of various proposed mitigation strategies in that section.
-Our findings are then discussed in the broader context of the literature
-in Section~\ref{sec-discussion}. We also point to some of the
-limitations of our work as well as avenues for future research in
+Section~\ref{sec-empirical} presents our experimental setup, including
+the data, classifiers and counterfactual generators that we have
+employed. Section~\ref{sec-empirical-2} presents the results for
+synthetic and real-world datasets along with evidence for the
+effectiveness of various proposed mitigation strategies. Our findings
+are then discussed in the broader context of the literature in
+Section~\ref{sec-discussion}, before pointing to some of the limitations
+of our work as well as avenues for future research in
 Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
@@
 A framework for Counterfactual Explanations was first proposed in 2017
-by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
+by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has
 served as the baseline for most methodologies that have been proposed
@@
 In the baseline work
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, the cost
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, the cost
 function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
@@
 through gradient descent. This summarizes the approach followed in
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} which we
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we
 shall refer to simply as \textbf{Wachter} - the name of the first author
@@
 et al.~layed out 60 algorithms that have been proposed since 2014
-\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
+\protect\hyperlink{ref-karimi2020survey}{{[}10{]}}. Another survey
 published around the same time by Verma et al.~described 29 algorithms
-\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
+\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. Different
 approaches vary primarily in terms of the objective functions they
@@
 generating counterfactuals that close
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
-\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable
+\protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse
+\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible
 causally founded
-\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.
+\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
 
@@
 shifts remains a largely unexplored research challenge at this point
-\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
+\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. We have been
 able to identify only one recent work that considers the implications of
 \textbf{exogenous} domain and model shifts in the context of AR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
 are strictly of external origin. For example, they might stem from data
 correction, temporal shifts or geospatial changes
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose ROAR - a
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a
 framework for Algorithmic Recourse that evidently improves robustness to
@@
 model behavior towards a group of individuals
-\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
+\protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors
 propose an optimization framework that generates collective
@@
 samples through the application of two-sample tests
-\protect\hyperlink{ref-rabanser2019failing}{{[}17{]}}. This task is a
+\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
 generalization of the anomaly detection problem for large datasets,
@@
 anomaly detection have been summarized
-\protect\hyperlink{ref-chandola2009anomaly}{{[}18{]}}. Another
+\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
 well-established research topic is that of concept drift -- situations
@@
 the output of a model
-\protect\hyperlink{ref-widmer1996learning}{{[}19{]}}. For instance, Gama
+\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
 et al.~offer a review of the adaptive learning techniques which can
-handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}20{]}}.
+handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
 Less previous work is available on the related topic of model drift -
@@
 resistant different machine learning models are to the model drift
-\protect\hyperlink{ref-nelson2015evaluating}{{[}21{]}}. Ackerman et
+\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
 al.~offer a method to detect changes in model performance when ground
 truth is not available
-\protect\hyperlink{ref-ackerman2021machine}{{[}22{]}}.
+\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.
 
@@
 first brought up by the authors behind ROAR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
 they refer to model shifts as simply any perturbation \(\Delta\) to the
@@
 baseline for our analysis here, it is worth noting that in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
 perturbations are mechanically introduced. In contrast we are interested
@@
 availability in this space. Recent work has started to address this gap:
-firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
+firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
 benchmarking study using different algorithmic aproaches and numerous
 tabular datasets; secondly,
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
+\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python
 framework - CARLA - that can be used to apply and benchmark different
 methodologies; finally, \texttt{CounterfactualExplanations.jl}
-\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
+\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}
 provides an extensible, fast and language-agnostic implementation in
@@
 include the following generators in our simulation experiments below:
-\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
-\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
+\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
+\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}},
+\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
 and a greedy approach that relies on probabilistic models
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. Our motivation for
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for
 including these different generators in our analysis, is that they all
@@
 generated by the (learned) DGP
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}.
+\protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.
 
@@
 rely on entropy estimates for probabilistic models. The approach
-proposed by \protect\hyperlink{ref-schut2021generating}{{[}7{]}}, which
+proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which
 we shall refer to as \textbf{Greedy}, also works with the subclass of
@@
 manner through a modified Jacobian-based Saliency Map Attack (JSMA).
-Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}7{]}} also
+Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also
 show that by maximizing the predicted probability of \(x^\prime\) being
@@
   example, the generator proposed by
-  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}24{]}} relies
+  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies
   on normalizing flows.}.
 
-Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
+Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
 distinguishes itself from all other generators considered here in that
 it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
-al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} show that
+al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that
 diverse outcomes can in principal be achieved simply rerunning
@@
 (or by randomly initializing the counterfactual). In
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} diversity is
+\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is
 explicitly proxied via Determinantal Point Processes (DDP): the authors
@@
 externalities and considered inefficient
-\protect\hyperlink{ref-pindyck2014microeconomics}{{[}25{]}}. We think
+\protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think
 that this concept describes the endogenous dynamics of algorithmic
@@
 Space \(\mathcal{H}\)
-\protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}, and \(X\),
+\protect\hyperlink{ref-berlinet2011reproducing}{{[}25{]}}, and \(X\),
 \(\tilde{X}\) represent independent and identically distributed samples
@@
 \(\mathcal{\tilde{X}}\) respectively
-\protect\hyperlink{ref-gretton2012kernel}{{[}27{]}}. MMD is a measure of
+\protect\hyperlink{ref-gretton2012kernel}{{[}26{]}}. MMD is a measure of
 the distance between the kernel mean embeddings of \(\mathcal{X}\) and
@@
 distribution, we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}28{]}}. To that end, we
+\protect\hyperlink{ref-arcones1992bootstrap}{{[}27{]}}. To that end, we
 combine the two samples and generate a large number of permutations of
@@
 the model parameters at each point in time \(t\) following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We define
 \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
@@
 \textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}29{]}} and
+introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and
 estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
@@
 
-Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
+We further introduce \textbf{Decisiveness} as a metric that quantifies
 the likelihood that a model assigns a high probability to its
@@
 the decision boundary moves towards either one of the two classes.
+
+Finally, we also take a look at the out-of-sample \textbf{Performance}
+of our models. To this end, we compute their F-score on a test sample
+that we leave untouched throughout the experiment.
 
@@
 the existing model. Evaluation metrics are computed and stored every 10
-rounds.
+rounds. To account for noise each individual experiment is repeat five
+times.
 
@@
 classifiers all of them built and trained using \texttt{Flux.jl}
-\protect\hyperlink{ref-innes2018fashionable}{{[}30{]}}: firstly, a
+\protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a
 simple linear classifier - \textbf{Logistic Regression} - implemented as
@@
 composed of five MLPs following
-\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}31{]}} that serves
+\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves
 as our only probabilistic classifier. We have chosen to work with deep
@@
 predictive uncertainty. They are also the model of choice in
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}. The actual neural
 network architectures are kept simple (Table~\ref{tbl-mlp}), since we
@@
 may affect the resulting recourse dynamics. Following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}, we generate
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}, we generate
 synthetic data in \(\mathbb{R}^2\) to also allow for a visual
@@
 impact on the paths of the recourse. For generators that use latent
-space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-CLUE \protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on
+space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
+CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}) or rely on
 (and have access to) probabilistic models (CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, Greedy
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
+\protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, Greedy
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}) we expect that
 counterfactuals will end up in regions of the target domain that are
@@
 borrower is likely to experience financial difficulties in the next two
-years \protect\hyperlink{ref-gmsc_data}{{[}32{]}}. Originally consisting
+years \protect\hyperlink{ref-gmsc_data}{{[}31{]}}. Originally consisting
 of 250,000 instances with 11 numerical attributes. Secondly, we use the
 \textbf{UCI defaultCredit} dataset
-\protect\hyperlink{ref-yeh2009comparisons}{{[}33{]}}, a benchmark
+\protect\hyperlink{ref-yeh2009comparisons}{{[}32{]}}, a benchmark
 dataset that can be used to train binary classifiers to predict the
@@
 before (see for example
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}} and
-\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}), presumably
+\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}},
+\protect\hyperlink{ref-joshi2019towards}{{[}7{]}} and
+\protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably
 because they constitute real-world classification tasks involving
@@
 derived from the 1990 U.S. census and sourced through scikit-learn
-\protect\hyperlink{ref-pace1997sparse}{{[}35{]}}. It consists of 8
+\protect\hyperlink{ref-pace1997sparse}{{[}34{]}}. It consists of 8
 continuous features that can be used to predict the median house price
@@
 in a similar fashion (see for example
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}). While we
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}). While we
 initially also conducted experiments on that dataset, we eventually
 discarded this dataset, since it has been found to suffer from an
-ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}36{]}}.
+ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}35{]}}.
 
@@
 All generators introduced earlier are included in the experiments:
-Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
-REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DiCE
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and Greedy
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. In addition, we
+Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}},
+REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, CLUE
+\protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, DiCE
+\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and Greedy
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}. In addition, we
 introduce two new generators in Section~\ref{sec-empirical-2-mitigate}
@@
 
+Below we first present our main findings with respect to the presence of
+endogenous macrodynamics in Algorithmic Recourse. We then go on to
+introduce and test a number of potential mitigation strategies.
+
 \hypertarget{endogenous-macrodynamics}{%
 \subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
+
+In light of the huge number of experiments we have conducted, we shall
+start this section off with a few high-level observations. Across all
+datasets (synthetic and real), classifiers and counterfactual generators
+we observe either most or all of the following dynamics at varying
+degrees:
+
+\begin{itemize}
+\tightlist
+\item
+  Statistically significant domain and model shift as measured by MMD.
+\item
+  A deterioration in out-of-sample model performance as measured by the
+  F-Score evaluated on a test sample. In many cases this drop in
+  performance is substantial.
+\item
+  Significant perturbations to the model parameters as well as an
+  increase in the model's decisiveness.
+\item
+  Disagreement between the original and retrained model, in some cases
+  large.
+\end{itemize}
+
+There is also some clear heterogeneity across the results:
+
+\begin{itemize}
+\tightlist
+\item
+  The observed, adverse dynamics are generally speaking of the highest
+  magnitude for the simple linear classifier. Differences in results for
+  the MLP and Deep Ensemble are mostly negligible.
+\item
+  The reduction in model performance appears to be most severe when
+  classes are not perfectly separable.
+\item
+  With the exception of the Greedy generator, all other generators
+  generally perform somewhat better than Wachter overall as expected.
+\end{itemize}
+
+Granular tables and visualizations summarizing all of our experimental
+results are published in the supplementary appendix. Here we will zoom
+in on illustrative examples that nicely reflect some of the broader
+patterns we have observed and listed above. Figure~\ref{fig-syn}
+presents a small subset of our results for the synthetic dataset with
+overlapping classes. It shows the resulting values for some of our
+evaluation metrics at the end of the experiment, so after all \(T=50\)
+rounds, along with error bars indicating the variation across folds. The
+top row shows the estimated domain shifts. While it is not
+straight-forward to interpret the exact magnitude of MMD, we can see
+clearly that the values are different from zero and there is essentially
+no variation across our five folds. With respect to the domain shifts,
+the Greedy generator actually induces the smallest shifts. In general,
+we have observed the opposite. The second row shows the estimated model
+shifts, where here we have used the meshgrid approach explained earlier.
+As with the domain shifts, the observed values are clearly different
+from zero and variation across folds is once again small. In this case,
+the results for this particular dataset very much reflect the broader
+patterns we have observed: Latent Space generators induce the smallest
+shifts, followed by DiCE, then Wachter and finally Greedy. The same
+broad pattern also emerges in the bottom row: we observe the smallest
+deterioration in model performance for Latent Space generators, albeit
+still a reduction in the F-Score of around 5 percentage points on
+average.
+
+\begin{figure}
+
+{\centering \includegraphics[width=9cm,height=9cm]{www/synthetic_results.png}
+
+}
+
+\caption{\label{fig-syn}Results for synthetic data with overlapping
+classes. The shown model MMD (PP MMD) was computed over a meshgrid of
+points. Error bars indicate the standard deviation across folds.}
+
+\end{figure}
+
+Turning to the real-world data next \ldots{}
+
+\begin{figure}
+
+{\centering \includegraphics[width=9cm,height=9cm]{www/placeholder.png}
+
+}
+
+\caption{\label{fig-real}PLACEHOLDER: Results for real-world as table.}
+
+\end{figure}
 
@@
 Another potential strategy draws inspiration from ROAR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}: to preserve the
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}: to preserve the
 classifier, we propose to simply explicitly penalize the loss it incurs
@@
 
-Our findings \ldots{}
+Our findings indicate that all three mitigation strategies are at least
+at par with Latent Space generators with respect to their effectiveness
+at mitigating domain and model shifts
+(Figure~\ref{fig-mitigate-results}) \ldots{}
+
+An interesting finding is also that the proposed strategies can have a
+complementary effect when used in combination with Latent Space
+generators. In experiments we conducted on the synthetic data, the
+benefits of Latent Space generators were exacerbated further when using
+a more conservative threshold or combining it with the penalties
+underlying Gravitational and ClapROAR.
+
+\begin{figure}
+
+{\centering \includegraphics[width=9cm,height=9cm]{www/synthetic_results.png}
+
+}
+
+\caption{\label{fig-mitigate-results}PLACEHOLDER: Results for mitigating
+strategies.}
+
+\end{figure}
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Shift of focus from individual to group of individuals (related:
-  https://www.researchgate.net/publication/353073138\_Generating\_Collective\_Counterfactual\_Explanations\_in\_Score-Based\_Classification\_via\_Mathematical\_Optimization)
-\item
-  Convergence criterium matters: terminating once threshold probability
-  is reached may not be optimal (see e.g.~REVISE)
-\item
-  Optimizer choice matters: dimensionality is typically low, so no
-  obvious benefit to using ADAM.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    This might be better placed in JuliaCon proceedings, perhaps backed
-    by small blog post on the matter.
-  \end{itemize}
-\item
-  Mitigating strategy: penaliye distance from centroid.
-\end{enumerate}
+Our results in Section~\ref{sec-empirical-2} indicate that
+state-of-the-art approaches to Algorithmic Recourse induce substantial
+domain and model shift, if implemented at scale in practice. These
+induced shifts can and should be considered as an (expected) external
+cost of individual recourse. While they do not affect the individual
+directly as long as we look at the individual in isolation, they can
+been seen to affect the broader group of stakeholders in automated
+data-driven decision-making. We have seen, for example, that
+out-of-sample model performance generally deteriorates in our simulation
+experiments. In practice, this can be seen as a cost to model owners,
+that is the group of stakeholders using the model as decision-making
+tool. As we have set out in the introduction, these model owners will
+generally be unwilling to carry that cost, and hence can be expected to
+stop offering recourse to individuals altogether. This in turn is costly
+to those individuals that would otherwise derive utility from being
+offered recourse.
+
+So, where does this leave us? We would argue that the expected external
+costs of individual recourse should be shared by all stakeholders. The
+most straight-forward way to achieve this is to introduce a penalty for
+external costs in the counterfactual search objective function, as we
+have set out in Equation~\ref{eq-collective}. This will on average lead
+to more costly counterfactual outcomes. But it may help to avoid extreme
+scenarios, in which minimal-cost recourse is reserved to a tiny minority
+of individuals. We have shown various types of shift-mitigating
+strategies that can be used to this end. Since all of these strategies
+can be seen simply as specific adaption of Equation~\ref{eq-collective},
+they can be applied to any of the various counterfactual generators
+studied here.
 
@@
 perspective, we are aware of several of its limitations. In the
-following we highlight some of these limitations and point to avenues
+following, we highlight some of these limitations and point to avenues
 for future research.
+
+\hypertarget{private-vs.-external-costs}{%
+\subsection{Private vs.~External
+Costs}\label{private-vs.-external-costs}}
+
+Perhaps the most crucial shortcoming of our work is that we merely point
+out that there exists a trade-off between private costs to the
+individual and external costs to the collective of stakeholders. We fall
+short of providing any definite answers as to how that trade-off may be
+resolved in practice. The mitigation strategies we have proposed here
+provide a good starting point, but they are ad-hoc, mechanical
+extensions of the existing AR framework. An interesting idea to explore
+in future work could be the potential for Pareto optimal Algorithmic
+Recourse, that is a collective recourse outcome in which no single
+individual can be made better off, without make at least one other
+individual worse off. This type of work would be interdisciplinary and
+could help to formalize some of the concepts presented in this work. For
+the time being, we suggest that the trade-off can be resolved on a
+case-by-case basis.
 
@@
 updated on a regular basis
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also find it
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also find it
 plausible to assume that the implementation of recourse happens
@@
 exogenous shifts of the nature investigated in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also make
+\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also make
 implicit assumptions about the utility functions of the involved agents
@@
 tabular datasets, such as the ones considered here
-\protect\hyperlink{ref-grinsztajn2022tree}{{[}38{]}}.
+\protect\hyperlink{ref-grinsztajn2022tree}{{[}37{]}}.
 
@@
 
-\hypertarget{acknowledgment}{%
-\section*{Acknowledgment}\label{acknowledgment}}
-\addcontentsline{toc}{section}{Acknowledgment}
-
-P. A. thanks \ldots{}
-
-\pagebreak
-\FloatBarrier
+This work has revisited and extended some of the most general and
+defining concepts underlying the literature on Counterfactual
+Explanations and, in particular, Algorithmic Recourse. We demonstrate
+that long-held beliefs as to what defines optimality in AR, are too
+short-sighted to serve as a foundation for applications of recourse in
+practice. Specifically, we run multiple experiments that simulate the
+application of recourse in practice using various popular counterfactual
+generators and find that all of them induce substantial domain and model
+shifts. We argue that these shifts should be considered as an expected
+external cost of individual recourse and call for a paradigm shift from
+individual to collective recourse. By proposing an adapted
+counterfactual search objective that incorporates this cost, we make
+that paradigm shift explicit. We show that this modified objective lends
+itself to mitigation strategies that can be used to effectively decrease
+the magnitude of induced domain and model shifts. Through our work we
+hope to inspire future research on this important topic. To this end we
+have open-sourced all of our code along with a Julia package -
+\texttt{AlgorithmicRecourseDynamics.jl}. The package is built on top of
+\texttt{CounterfactualExplanations.jl} and inherits its extensibility
+\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}.
+That is to say that future researchers should find it relatively easy to
+replicate, modify and extend the simulation experiments presented here
+and apply to their own custom counterfactual generators.
 
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
 \CSLLeftMargin{{[}1{]} }%
-\CSLRightInline{C. Borch, {``Machine learning, knowledge risk, and
-principal-agent problems in automated trading,''} \emph{Technology in
-Society}, p. 101852, 2022.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
-\CSLLeftMargin{{[}2{]} }%
 \CSLRightInline{C. O'neil, \emph{Weapons of math destruction: How big
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
-\CSLLeftMargin{{[}3{]} }%
+\CSLLeftMargin{{[}2{]} }%
 \CSLRightInline{C. Rudin, {``Stop explaining black box machine learning
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
-\CSLLeftMargin{{[}4{]} }%
+\CSLLeftMargin{{[}3{]} }%
 \CSLRightInline{M. Pawelczyk, S. Bielawski, J. van den Heuvel, T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
-\CSLLeftMargin{{[}5{]} }%
+\CSLLeftMargin{{[}4{]} }%
 \CSLRightInline{S. Wachter, B. Mittelstadt, and C. Russell,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
-\CSLLeftMargin{{[}6{]} }%
+\CSLLeftMargin{{[}5{]} }%
 \CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
-\CSLLeftMargin{{[}7{]} }%
+\CSLLeftMargin{{[}6{]} }%
 \CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
@@
 \leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
-\CSLLeftMargin{{[}8{]} }%
+\CSLLeftMargin{{[}7{]} }%
 \CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
-\CSLLeftMargin{{[}9{]} }%
+\CSLLeftMargin{{[}8{]} }%
 \CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
@@
 \leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-\CSLLeftMargin{{[}10{]} }%
+\CSLLeftMargin{{[}9{]} }%
 \CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
-\CSLLeftMargin{{[}11{]} }%
+\CSLLeftMargin{{[}10{]} }%
 \CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
-\CSLLeftMargin{{[}12{]} }%
+\CSLLeftMargin{{[}11{]} }%
 \CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
-\CSLLeftMargin{{[}13{]} }%
+\CSLLeftMargin{{[}12{]} }%
 \CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
@@
 \leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
-\CSLLeftMargin{{[}14{]} }%
+\CSLLeftMargin{{[}13{]} }%
 \CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
-\CSLLeftMargin{{[}15{]} }%
+\CSLLeftMargin{{[}14{]} }%
 \CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
-\CSLLeftMargin{{[}16{]} }%
+\CSLLeftMargin{{[}15{]} }%
 \CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
-\CSLLeftMargin{{[}17{]} }%
+\CSLLeftMargin{{[}16{]} }%
 \CSLRightInline{S. Rabanser, S. Günnemann, and Z. Lipton, {``Failing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
-\CSLLeftMargin{{[}18{]} }%
+\CSLLeftMargin{{[}17{]} }%
 \CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly
@@
 \leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
-\CSLLeftMargin{{[}19{]} }%
+\CSLLeftMargin{{[}18{]} }%
 \CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
-\CSLLeftMargin{{[}20{]} }%
+\CSLLeftMargin{{[}19{]} }%
 \CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
-\CSLLeftMargin{{[}21{]} }%
+\CSLLeftMargin{{[}20{]} }%
 \CSLRightInline{K. Nelson, G. Corbin, M. Anania, M. Kovacs, J. Tobias,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
-\CSLLeftMargin{{[}22{]} }%
+\CSLLeftMargin{{[}21{]} }%
 \CSLRightInline{S. Ackerman, P. Dube, E. Farchi, O. Raz, and M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
-\CSLLeftMargin{{[}23{]} }%
+\CSLLeftMargin{{[}22{]} }%
 \CSLRightInline{R. M. B. de Oliveira and D. Martens, {``A framework and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
-\CSLLeftMargin{{[}24{]} }%
+\CSLLeftMargin{{[}23{]} }%
 \CSLRightInline{A.-K. Dombrowski, J. E. Gerken, and P. Kessel,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
-\CSLLeftMargin{{[}25{]} }%
+\CSLLeftMargin{{[}24{]} }%
 \CSLRightInline{R. S. Pindyck and D. L. Rubinfeld,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
-\CSLLeftMargin{{[}26{]} }%
+\CSLLeftMargin{{[}25{]} }%
 \CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
-\CSLLeftMargin{{[}27{]} }%
+\CSLLeftMargin{{[}26{]} }%
 \CSLRightInline{A. Gretton, K. M. Borgwardt, M. J. Rasch, B. Schölkopf,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}28{]} }%
+\CSLLeftMargin{{[}27{]} }%
 \CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}29{]} }%
+\CSLLeftMargin{{[}28{]} }%
 \CSLRightInline{S. Hanneke, {``A bound on the label complexity of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
-\CSLLeftMargin{{[}30{]} }%
+\CSLLeftMargin{{[}29{]} }%
 \CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
-\CSLLeftMargin{{[}31{]} }%
+\CSLLeftMargin{{[}30{]} }%
 \CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}32{]} }%
+\CSLLeftMargin{{[}31{]} }%
 \CSLRightInline{K. Competition, {``Give me some credit, improve on the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
-\CSLLeftMargin{{[}33{]} }%
+\CSLLeftMargin{{[}32{]} }%
 \CSLRightInline{I.-C. Yeh and C. Lien, {``The comparisons of data mining
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
-\CSLLeftMargin{{[}34{]} }%
+\CSLLeftMargin{{[}33{]} }%
 \CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
-\CSLLeftMargin{{[}35{]} }%
+\CSLLeftMargin{{[}34{]} }%
 \CSLRightInline{R. K. Pace and R. Barry, {``Sparse spatial
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
-\CSLLeftMargin{{[}36{]} }%
+\CSLLeftMargin{{[}35{]} }%
 \CSLRightInline{C. M., {``Racist data destruction? - a boston housing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}37{]} }%
+\CSLLeftMargin{{[}36{]} }%
 \CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-\CSLLeftMargin{{[}38{]} }%
+\CSLLeftMargin{{[}37{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do

### 2022-08-12 dev/paper/sections/discussion.qmd
some work on rephrasing narrative in intro
almost done, findings still need to be fleshed out though

--- 2022-08-08 dev/paper/sections/empirical.qmd
@@
-# Experiments {#sec-empirical}
+# Discussion {#sec-discussion}
 
-1.
-Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
-2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
-3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
-    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
-4. Mitigating strategy: penaliye distance from centroid.
+Our results in @sec-empirical-2 indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice.
+These induced shifts can and should be considered as an (expected) external cost of individual recourse.
+While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making.
+We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments.
+In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool.
+As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
+This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.
+
+So, where does this leave us?
+We would argue that the expected external costs of individual recourse should be shared by all stakeholders.
+The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in @eq-collective.
+This will on average lead to more costly counterfactual outcomes.
+But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
+We have shown various types of shift-mitigating strategies that can be used to this end.
+Since all of these strategies can be seen simply as specific adaption of @eq-collective, they can be applied to any of the various counterfactual generators studied here.

### 2022-08-12 dev/paper/sections/empirical_2.qmd
almost done, findings still need to be fleshed out though

@@
 
+Below we first present our main findings with respect to the presence of endogenous macrodynamics in Algorithmic Recourse.
+We then go on to introduce and test a number of potential mitigation strategies.
+
 ## Endogenous Macrodynamics
+
+In light of the huge number of experiments we have conducted, we shall start this section off with a few high-level observations.
+Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:
+
+- Statistically significant domain and model shift as measured by MMD. 
+- A deterioration in out-of-sample model performance as measured by the F-Score evaluated on a test sample. In many cases this drop in performance is substantial.
+- Significant perturbations to the model parameters as well as an increase in the model's decisiveness.
+- Disagreement between the original and retrained model, in some cases large.
+
+There is also some clear heterogeneity across the results:
+
+- The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
+- The reduction in model performance appears to be most severe when classes are not perfectly separable.
+- With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.
+
+Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix.
+Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
+@fig-syn presents a small subset of our results for the synthetic dataset with overlapping classes.
+It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds.
+The top row shows the estimated domain shifts.
+While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds.
+With respect to the domain shifts, the Greedy generator actually induces the smallest shifts.
+In general, we have observed the opposite.
+The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier.
+As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
+In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
+The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.
+
+![Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of points.
+Error bars indicate the standard deviation across folds.](www/synthetic_results.png){#fig-syn fig.pos="h" width="9cm" height="9cm"}
+
+Turning to the real-world data next ...
+
+![PLACEHOLDER: Results for real-world as table.](www/placeholder.png){#fig-real fig.pos="h" width="9cm" height="9cm"}
 
@@
 
-Our findings ...
+Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts (@fig-mitigate-results) ...
+
+An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
+In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+
+![PLACEHOLDER: Results for mitigating strategies.](www/synthetic_results.png){#fig-mitigate-results fig.pos="h" width="9cm" height="9cm"}

### 2022-08-12 dev/paper/sections/introduction.qmd
almost done, findings still need to be fleshed out though

@@
 Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences.
-While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of principal-agent problem in the context of data-driven decision-making.
-It involves a group of **principals** - i.e.
-human stakeholders - that fail to understand the behaviour of their **agent** - i.e.
-the model used for automated decision-making [@borch2022machine].
-
-Models or algorithms that fall into this category are typically referred to **black-box** models.
-Despite their shortcomings, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons].
+While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making.
+They are essentially **black boxes**: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools.
+Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons].
 The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models.
@@
 
-![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}
+![Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}
 
@@
 @sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR.
-@sec-empirical presents our experimental results for synthetic and real-world datasets.
-We also present evidence for the effectiveness of various proposed mitigation strategies in that section.
-Our findings are then discussed in the broader context of the literature in @sec-discussion.
-We also point to some of the limitations of our work as well as avenues for future research in @sec-limit.
+@sec-empirical presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
+@sec-empirical-2 presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
+Our findings are then discussed in the broader context of the literature in @sec-discussion, before pointing to some of the limitations of our work as well as avenues for future research in @sec-limit.
 Finally, @sec-conclusion concludes.

### 2022-08-12 dev/paper/sections/limitations.qmd
almost done, findings still need to be fleshed out though

@@
 While we believe that this work constitutes a valuable starting point for addressing existing issues in Algorithmic Recourse from a fresh perspective, we are aware of several of its limitations.
-In the following we highlight some of these limitations and point to avenues for future research.
+In the following, we highlight some of these limitations and point to avenues for future research.
+
+## Private vs. External Costs
+
+Perhaps the most crucial shortcoming of our work is that we merely point out that there exists a trade-off between private costs to the individual and external costs to the collective of stakeholders.
+We fall short of providing any definite answers as to how that trade-off may be resolved in practice.
+The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework.
+An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is a collective recourse outcome in which no single individual can be made better off, without make at least one other individual worse off.
+This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work.
+For the time being, we suggest that the trade-off can be resolved on a case-by-case basis.
 

### 2022-08-12 dev/paper/sections/methodology_2.qmd
almost done, findings still need to be fleshed out though

@@
 
-Finally, we introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample.
+We further introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample.
 We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function.
@@
 If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.
+
+Finally, we also take a look at the out-of-sample **Performance** of our models. To this end, we compute their F-score on a test sample that we leave untouched throughout the experiment.

### 2022-08-15 dev/paper/paper.tex
more work on results

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{sec-empirical-2-mitigate}{%
-\subsection{Potential Mitigation
+\section{Potential Mitigation
 Strategies}\label{sec-empirical-2-mitigate}}
@@
 \hypertarget{more-conservative-decision-thresholds}{%
-\subsubsection{More Conservative Decision
+\subsection{More Conservative Decision
 Thresholds}\label{more-conservative-decision-thresholds}}
@@
 \hypertarget{classifier-preserving-roar}{%
-\subsubsection{Classifier Preserving
+\subsection{Classifier Preserving
 ROAR}\label{classifier-preserving-roar}}
@@
 \hypertarget{gravitational-counterfactual-explanations}{%
-\subsubsection{Gravitational Counterfactual
+\subsection{Gravitational Counterfactual
 Explanations}\label{gravitational-counterfactual-explanations}}
@@
 
-Figure~\ref{fig-mitigation} shows an illustrative example that
+Figure~\ref{fig-mitigation} hows an illustrative example that
 demonstrates the differences in counterfactual outcomes when using the
@@
 
-\caption{\label{fig-mitigation}The differences in counterfactual
-outcomes when using the various mitigation strategies compared to the
-baseline approach, that is Wachter with \(\gamma=0.5\)}
+\caption{\label{fig-mitigation}Illustrative example demonstrating the
+properties of the various mitigation strategies.}
 
@@
 at par with Latent Space generators with respect to their effectiveness
-at mitigating domain and model shifts
-(Figure~\ref{fig-mitigate-results}) \ldots{}
+at mitigating domain and model shifts. Figure~\ref{fig-mitigate-results}
+presents a subset of the evaluation metrics for our synthetic data with
+overlapping classes. The top row in Figure~\ref{fig-mitigate-results}
+indicates that while domain shifts are of roughly the same magnitude for
+both Wachter and Latent Space generators, our proposed strategies
+effectively mitigate these shifts. ClapROAR appears to be particularly
+effective, which is positively surprising, since it is designed to
+explicitly address model shifts, not domain shifts. As evident from the
+middle row in Figure~\ref{fig-mitigate-results} model shifts can also be
+reduced: for the deep ensemble Latent Space search yields results that
+are at par with the mitigation strategies, while for both the simple MLP
+and logistic regression our simple strategies are actually more
+effective. The same overall pattern can be observed for out-of-sample
+model performance. With respect to the other synthetic datasets we found
+the following: for the Moons dataset the emerging patterns are largely
+the same, but the estimated model shifts are insignificant as noted
+earlier; the same holds for the Circles dataset, but there is no
+significant reduction in model performance for our neural networks; in
+the case of linearly separable data we find the Gravitational generator
+to be most effective at mitigating shifts.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/mitigation_synthetic_results.png}
+
+}
+
+\caption{\label{fig-mitigate-results}The differences in counterfactual
+outcomes when using the various mitigation strategies compared to the
+baseline approach, that is Wachter with \(\gamma=0.5\). Results for
+synthetic data with overlapping classes. The shown model MMD (PP MMD)
+was computed over a meshgrid of points. Error bars indicate the standard
+deviation across folds.}
+
+\end{figure}
 
@@
 a more conservative threshold or combining it with the penalties
-underlying Gravitational and ClapROAR.
+underlying Gravitational and ClapROAR. A snapshot of the results is
+shown in Figure~\ref{fig-mitigate-latent-results}.
 
@@
 
-{\centering \includegraphics[width=9cm,height=9cm]{www/synthetic_results.png}
+{\centering \includegraphics[width=9cm,height=9cm]{www/mitigation_synthetic_latent_results.png}
 
@@
 
-\caption{\label{fig-mitigate-results}PLACEHOLDER: Results for mitigating
-strategies.}
+\caption{\label{fig-mitigate-latent-results}Combinining various
+mitigation strategies with Latent Space search. Results for synthetic
+data with overlapping classes. The shown model MMD (PP MMD) was computed
+over a meshgrid of points. Error bars indicate the standard deviation
+across folds.}
 

### 2022-08-15 dev/notebooks/mitigation_strategies.qmd
updated kable function to produce latex table

@@
     :Generic=>GenericGenerator(decision_threshold=0.5),
-    :REVISE=>REVISEGenerator(),
+    :Latent=>REVISEGenerator(),
     :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
@@
 
-### Plots
-
 ```{julia}
+using Images
 line_charts = Dict()
@@
 for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
 end
@@
 
+### Chart in paper
+
+```{julia}
+using DataFrames, Statistics
+df = results[:overlapping].output
+df = df[df.n .== maximum(df.n),:]
+gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="Generic" ? "Generic (γ=0.5)" : x for x in X]) => :generator)
+transform!(df_plot, :generator => (X -> [x=="Generic_conservative" ? "Generic (γ=0.9)" : x for x in X]) => :generator)
+
+ncol = length(unique(df_plot.model))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 2.2
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_wrap(name ~ model, scale="free_y", ncol=$ncol) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_synthetic_results.png"), img)
+```
+
+### Latent Space Search
+
+```{julia}
+generators = Dict(
+    :Latent=>GenericGenerator(decision_threshold=0.5),
+    :Latent_conservative=>GenericGenerator(decision_threshold=0.9),
+    :Gravitational=>GravitationalGenerator(),
+    :ClapROAR=>ClapROARGenerator()
+)
+```
+
+```{julia}
+experiments = set_up_experiments(data_sets,models,generators)
+```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_bootstrap = 1
+T = 100
+using Serialization
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T
+)
+Serialization.serialize(joinpath(output_path,"results_synthetic_latent.jls"),results)
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_synthetic_latent.jls"))
+```
+
+```{julia}
+using Images
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_latent_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_latent_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
+end
+```
+
+### Chart in paper
+
+```{julia}
+using DataFrames, Statistics
+df = results[:overlapping].output
+df = df[df.n .== maximum(df.n),:]
+gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="Latent" ? "Latent (γ=0.5)" : x for x in X]) => :generator)
+transform!(df_plot, :generator => (X -> [x=="Latent_conservative" ? "Latent (γ=0.9)" : x for x in X]) => :generator)
+
+ncol = length(unique(df_plot.model))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 2.2
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_wrap(name ~ model, scale="free_y", ncol=$ncol) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_synthetic_latent_results.png"), img)
+```
+
 ## Real World
+
+```{julia}
+generators = Dict(
+    :Generic=>GenericGenerator(decision_threshold=0.5),
+    :Latent=>REVISEGenerator(),
+    :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
+    :Gravitational=>GravitationalGenerator(),
+    :ClapROAR=>ClapROARGenerator()
+)
+```
 
@@
 n_bootstrap = 1
-T = 100
+T = 250
 using Serialization
@@
 ```{julia}
+using Images
 line_charts = Dict()
@@
 n_bootstrap = 1
-T = 100
+T = 200
 using Serialization
@@
 ```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_real_world.jls"))
+```
+
+```{julia}
+using Images
 line_charts = Dict()
@@
 for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_latent_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_latent_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
 end

### 2022-08-15 dev/notebooks/real_world.qmd
updated kable function to produce latex table

@@
 
-## Strict convergence
+## Running Experiment
 
@@
 n_bootstrap = 1
-T = 100
+T = 250
 generative_model_params = (epochs=250, latent_dim=8)
@@
 for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
 end
 ```
+
+### Table in paper
+
+```{julia}
+using AlgorithmicRecourseDynamics: kable
+kable(results, [50])
+```

### 2022-08-15 dev/notebooks/synthetic.qmd
updated kable function to produce latex table

--- 2022-08-10 dev/notebooks/synthetic.qmd
@@
     experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
-    convergence=:strict, save_name_suffix="strict"
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T
 )
-Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
 ```
@@
     plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_train_after_$(name)_strict.png"))
+    savefig(plt, joinpath(www_path,"models_train_after_$(name).png"))
 end
@@
 for (data_name, res) in results
-    line_charts[data_name] = plot(res)
-    errorbar_charts[data_name] = plot(res,50)
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
 end
 ```
+
+### Chart in paper
+
+```{julia}
+using DataFrames, Statistics
+df = results[:overlapping].output
+df = df[df.n .== maximum(df.n),:]
+gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+
+ncol = length(unique(df_plot.model))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 2
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_wrap(name ~ model, scale="free_y", ncol=$ncol) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_synthetic_results.png"), img)
+```

### 2022-08-15 dev/paper/sections/conclusion.qmd

--- 2022-08-12 dev/paper/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  conference]{IEEEtran}
+# Concluding Remarks {#sec-conclusion}
 
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithms
-\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
-\usepackage{graphicx}
-\usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{placeins}
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\title{Endogenous Macrodynamics in Algorithmic Recourse}
-\author{}
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-Existing work on Counterfactual Explanations (CE) and Algorithmic
-Recourse (AR) has largely been limited to the static setting and focused
-on single individuals: given some estimated model the goal is to find
-valid counterfactuals for individual instance that fulfill various
-desiderata. The ability of such counterfactuals to handle dynamics like
-data and model drift remains a largely unexplored research challenge at
-this point. There has also been surprisingly little work on the related
-question of how the actual implementation of recourse by one individual
-may affect other individuals. Through this work we aim to close that gap
-by systematizing and extending existing knowledge. We first show that
-many of the existing methodologies can be collectively described by a
-generalized framework. We then argue that the existing framework fails
-to account for a hidden external cost of recourse, that only reveals
-itself when studying the endogenous dynamics of recourse at the group
-level. Through simulation experiments involving various popular
-counterfactual generators and several benchmark datasets, we generate a
-total XX million Counterfactual Explanations and study the resulting
-domain and model shifts. We find that the induced shifts are substantial
-enough to likely impede the applicability Algorithmic Recourse in
-practice. Fortunately, we find various potential mitigation strategies
-that can be used in combination with existing approaches. Our simulation
-framework for studying recourse dynamics is fast and open-sourced.
-\end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
-
-Recent advances in Artificial Intelligence (AI) have propelled its
-adoption in scientific domains outside of Computer Science including
-Healthcare, Bioinformatics, Genetics and the Social Sciences. While this
-has in many cases brought benefits in terms of efficiency,
-state-of-the-art models like Deep Neural Networks (DNN) have also given
-rise a new type of principal-agent problem in the context of data-driven
-decision-making. It involves a group of \textbf{principals} - i.e.~human
-stakeholders - that fail to understand the behaviour of their
-\textbf{agent} - i.e.~the model used for automated decision-making
-\protect\hyperlink{ref-borch2022machine}{{[}1{]}}.
-
-Models or algorithms that fall into this category are typically referred
-to \textbf{black-box} models. Despite their shortcomings, black-box
-models have grown in popularity in recent years and have at times
-created undesirable societal outcomes
-\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
-has tackled this issue from two different angles: while some have
-appealed for a strict focus on inherently interpretable models
-\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
-different ways to explain the behaviour of black-box models. These two
-sub-domains can be broadly referred to as \textbf{interpretable AI} and
-\textbf{explainable AI} (XAI), respectively.
-
-Among the approaches to XAI that have recently grown in popularity are
-\textbf{Counterfactual Explanations} (CE). They explain how inputs into
-a model need to change for it to produce different outputs.
-Counterfactual Explanations that involve realistic and actionable
-changes can be used for the purpose of \textbf{Algorithmic Recourse}
-(AR) to help individuals who face adverse outcomes. An example relevant
-to the Social Sciences is consumer credit: in this context, AR can be
-used to guide individuals in improving their creditworthiness, should
-they have previously been denied access to credit based on an automated
-decision-making system. A meaningful recourse recommendation for a
-denied applicant could be: \emph{``If your net savings rate had been
-10\% of your monthly income instead of the actual 8\%, your application
-would have been successful. See if you can temporarily cut down on
-consumption.''} In the remainder of this paper we will use both
-terminologies - recourse and counterfactual - interchangeably to refer
-to situations where counterfactuals are generated with the intent to
-provide individual recourse.
-
-Existing work in this field has largely worked in a static setting:
-various approaches have been proposed to generate counterfactuals for a
-given individual that is subject to some pre-trained model. More recent
-work has compared different approaches within this static setting
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work, we go
-one step further and ask ourselves: what happens if recourse is provided
-and implemented repeatedly? What types of dynamics are introduced and
-how do different counterfactual generators compare in this context?
-
-Research on Algorithmic Recourse has also so far typically addressed the
-issue from the perspective of one single individual and has indeed been
-referred to as \textbf{individual recourse} in some places. Arguably
-though, most real-world applications that warrant Algorithmic Recourse
-involve potentially large groups of individuals typically competing for
-scarce resources. Our work demonstrates that in such scenarios, choices
-made by or for one single individual are likely to affect the broader
-collective of individuals in ways that current approaches to AR fail to
-account for. More specifically, we argue that a strict focus on
-minimizing the private costs faced by individuals may be too narrow an
-objective.
-
-Figure~\ref{fig-poc} illustrates this idea for a binary problem
-involving a probabilistic classifier and the counterfactual generator
-proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
-the implementation of AR for a subset of individuals leads to a domain
-shift (b), which in turn triggers a model shift (c). As this game of
-implementing AR and updating the classifier is repeated, the decision
-boundary moves away from training samples that were originally in the
-target class (d). We refer to these types of dynamics as
-\textbf{endogenous} because they are induced by the implementation of
-recourse itself. The term \textbf{macrodynamics} is borrowed from the
-economics literature and used to describe processes involving whole
-groups or societies.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/poc.png}
-
-}
-
-\caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
-simple Bayesian model trained for binary classification (a); the
-implementation of AR for a random subset of individuals leads to a
-domain shift (b); as the classifier is retrained we observe a model
-shift (c); as this process is repeated, the decision boundary moves away
-from the target class (d).}
-
-\end{figure}
-
-We think that these types of endogenous dynamics may be problematic and
-warrant our attention. Firstly, model shifts may inadvertently change
-classification outcomes for individuals who never received and
-implemented recourse. Secondly and relatedly, we observe in
-Figure~\ref{fig-poc} that as the decision boundary moves in the
-direction of the non-target class, counterfactual paths become shorter:
-in the consumer credit example, individuals that previously would have
-been denied credit based on their input features are suddenly considered
-as creditworthy. Average default risk across all borrowers can therefore
-be expected to increase. Conversely, lenders that anticipate such
-dynamics may choose to refrain from offering recourse (and hence credit)
-to more than just a tiny share of individuals. In that latter and
-perhaps more likely scenario, the probability of being offered recourse
-decreases with every individual that implements recourse: in other
-words, the actions of first-movers exert a negative externality on
-future would-be borrowers.
-
-To the best of our knowledge this is the first work investigating
-endogenous macrodynamics in AR. Our contributions to the state of
-knowledge are as follows: firstly, we posit a compelling argument that
-calls for a novel perspective on Algorithmic Recourse extending our
-focus from single individuals to groups. Secondly, we introduce an
-experimental framework extending previous work by
-\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
-which enables us to study macrodynamics of Algorithmic Recourse through
-simulations that can be fully parallelized. Thirdly, we use this
-framework to provide a first in-depth analysis of endogenous recourse
-dynamics induced by various popular counterfactual generators including
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
-propose a number of novel evaluation metrics that can be used to
-quantify and benchmark the macrodynamics introduced by the different
-generators. Finally, we also discuss what drives endogenous dynamics and
-propose strategies to mitigate them.
-
-The remainder of the paper is structured as follows:
-Section~\ref{sec-related} places our work in the broader context of
-related literature. Section~\ref{sec-method} posits a generalized
-framework for gradient-based Algorithmic Recourse and introduces the
-notion of hidden external costs. Section~\ref{sec-method-2} sets out our
-experimental framework for modeling endogenous macrodynamics in AR.
-Section~\ref{sec-empirical} presents our experimental results for
-synthetic and real-world datasets. We also present evidence for the
-effectiveness of various proposed mitigation strategies in that section.
-Our findings are then discussed in the broader context of the literature
-in Section~\ref{sec-discussion}. We also point to some of the
-limitations of our work as well as avenues for future research in
-Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
-concludes.
-
-\hypertarget{sec-related}{%
-\section{Background}\label{sec-related}}
-
-In this Section we provide a review of the relevant literature. First,
-Section~\ref{sec-related-recourse} discusses the existing research
-within the domain of Counterfactual Explanations and Algorithmic
-Recourse. Then, Section~\ref{sec-related-shifts} presents some of the
-previous work on the measurement of dataset and model shifts.
-
-\hypertarget{sec-related-recourse}{%
-\subsection{Algorithmic Recourse}\label{sec-related-recourse}}
-
-A framework for Counterfactual Explanations was first proposed in 2017
-by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
-served as the baseline for most methodologies that have been proposed
-since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
-pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
-\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
-cost\footnote{Equivalently, others have referred to this quantity as
-  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
-by individual \(x\) when moving to a counterfactual state \(x\prime\)
-such that the predicted outcome \(M(x\prime)\) corresponds to some
-target outcome \(y^*\):
-
-\begin{equation}\protect\hypertarget{eq-obj}{}{
-\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
-}\label{eq-obj}\end{equation}
-
-For implementation purposes, Equation~\ref{eq-obj} is typically
-approximated through regularization:
-
-\begin{equation}\protect\hypertarget{eq-solution}{}{
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
-}\label{eq-solution}\end{equation}
-
-In the baseline work
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, the cost
-function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
-distance metric based on the simple intuition that perturbations of
-\(x\) are costly to the individual. For models that are differentiable
-and produce smooth predictions, Equation~\ref{eq-solution} can be solved
-through gradient descent. This summarizes the approach followed in
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} which we
-shall refer to simply as \textbf{Wachter} - the name of the first author
-- in the remainder of this paper.
-
-Many approaches for the generation of Algorithmic Recourse have been
-described in the literature since 2017. An October 2020 survey by Karimi
-et al.~layed out 60 algorithms that have been proposed since 2014
-\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
-published around the same time by Verma et al.~described 29 algorithms
-\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
-approaches vary primarily in terms of the objective functions they
-impose, how they optimize said objective (from brute force through
-gradient-based approaches to graph traversal algorithms), and how the
-ensure that certain requirements for CE are met. Regarding the latter,
-the literature has produced an extensive list of desiderata each
-addressing different needs. To name but a few, we are interested in
-generating counterfactuals that close
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
-\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
-causally founded
-\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.
-
-Efforts so far have largely been directed at improving the quality of
-Counterfactual Explanations within a static context: given some
-pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
-interested in generating one or multiple meaningful Counterfactual
-Explanations for some individual characterized by \(x\). The ability of
-Counterfactual Explanations to handle dynamics like data and model
-shifts remains a largely unexplored research challenge at this point
-\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
-able to identify only one recent work that considers the implications of
-\textbf{exogenous} domain and model shifts in the context of AR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
-are strictly of external origin. For example, they might stem from data
-correction, temporal shifts or geospatial changes
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose ROAR - a
-framework for Algorithmic Recourse that evidently improves robustness to
-such exogenous shifts.
-
-As mentioned earlier, research has so far also generally focused on
-generating counterfactuals for single individuals or instances. We have
-been able to identify only one existing work that investigates black-box
-model behavior towards a group of individuals
-\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
-propose an optimization framework that generates collective
-counterfactuals. We provide a motivation for doing so from the
-perspective of endogenous macrodynamics of Algorithmic Recourse.
-
-\hypertarget{sec-related-shifts}{%
-\subsection{Domain and Model Shifts}\label{sec-related-shifts}}
-
-Much attention has been paid to the detection of dataset shifts --
-situations where the distribution of data changes over time. Rabanser et
-al.~suggest a framework to detect data drift from a minimal number of
-samples through the application of two-sample tests
-\protect\hyperlink{ref-rabanser2019failing}{{[}17{]}}. This task is a
-generalization of the anomaly detection problem for large datasets,
-which aims to answer the question if two sets of samples could have been
-generated from the same probability distribution. Numerous approaches to
-anomaly detection have been summarized
-\protect\hyperlink{ref-chandola2009anomaly}{{[}18{]}}. Another
-well-established research topic is that of concept drift -- situations
-where external variables influence the patterns between the input and
-the output of a model
-\protect\hyperlink{ref-widmer1996learning}{{[}19{]}}. For instance, Gama
-et al.~offer a review of the adaptive learning techniques which can
-handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}20{]}}.
-Less previous work is available on the related topic of model drift -
-changes in model performance over time. Nelson et al.~review how
-resistant different machine learning models are to the model drift
-\protect\hyperlink{ref-nelson2015evaluating}{{[}21{]}}. Ackerman et
-al.~offer a method to detect changes in model performance when ground
-truth is not available
-\protect\hyperlink{ref-ackerman2021machine}{{[}22{]}}.
-
-In the context of Algorithmic Recourse, domain and model shifts were
-first brought up by the authors behind ROAR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
-they refer to model shifts as simply any perturbation \(\Delta\) to the
-parameters of the model in question: \(M\). While this also sets the
-baseline for our analysis here, it is worth noting that in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
-perturbations are mechanically introduced. In contrast we are interested
-in quantifying model shifts that arise endogenously as part of a dynamic
-recourse process. In addition to quantifying the magnitude of shifts
-\(\Delta\), we aim to also analyse the characteristics of changes to the
-model, such as the position of the decision boundary and the overall
-decisiveness of the model. We have not been able to identify previous
-work on this topic.
-
-\hypertarget{sec-related-benchmark}{%
-\subsection{Benchmarking Counterfactual
-Generators}\label{sec-related-benchmark}}
-
-Despite the large and growing number of approaches to counterfactual
-search, there have been surprisingly few benchmark studies that compare
-different methodologies. This may be partially due to limited software
-availability in this space. Recent work has started to address this gap:
-firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
-benchmarking study using different algorithmic aproaches and numerous
-tabular datasets; secondly,
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
-framework - CARLA - that can be used to apply and benchmark different
-methodologies; finally, \texttt{CounterfactualExplanations.jl}
-\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
-provides an extensible, fast and language-agnostic implementation in
-Julia. Since the experiments presented here involve extensive
-simulations, we have relied on and extended the Julia implementation due
-to the associated performance benefits. In particular, we have built a
-framework on top of \texttt{CounterfactualExplanations.jl} that extends
-the functionality from static benchmarks to simulation experiments:
-\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
-  available from \ldots{}}. The core concepts implemented in that
-package reflect what is presented in section Section~\ref{sec-method-2}
-of this paper.
-
-\hypertarget{sec-method}{%
-\section{Gradient-Based Recourse Revisited}\label{sec-method}}
-
-In this section we first set out a generalized framework for
-gradient-based counterfactual search in Section~\ref{sec-method-general}
-that encapsulates the various individual recourse methods we have chosen
-to use in our experiments. We then introduce the notion of a hidden
-external cost in algorithmic recourse and extend the existing framework
-to explicitly address this cost in the counterfactual search objective.
-
-\hypertarget{sec-method-general}{%
-\subsection{From individual recourse
-\ldots{}}\label{sec-method-general}}
-
-We have chosen to focus on gradient-based counterfactual search for two
-reasons: firstly, they can be seen as direct descendants of our baseline
-method - Wachter; secondly, gradient-based search is particularly
-well-suited for differentiable black-box models like deep neural
-networks, which we focus on in this work. In particular, we include
-include the following generators in our simulation experiments below:
-\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
-\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
-and a greedy approach that relies on probabilistic models
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. Our motivation for
-including these different generators in our analysis, is that they all
-offer slightly different approaches to generate meaningful
-counterfactuals for differentiable black-box models. We hypothesize that
-generating more \textbf{meaningful} counterfactuals should mitigate the
-endogenous dynamics illustrated in Figure~\ref{fig-poc} in
-Section~\ref{sec-intro}. This intuition stems from the underlying idea
-that more meaningful counterfactuals are generated by the same or at
-least a very similar data generating process as the training data. All
-else equal, counterfactuals that fulfill this basic requirement should
-be less prone to trigger domain and model shifts.
-
-As we will see next, all of them can be described by the following
-generalized form of Equation~\ref{eq-solution}:
-
-\begin{equation}\protect\hypertarget{eq-general}{}{
-\begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
-\end{aligned}
-}\label{eq-general}\end{equation}
-
-Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
-\(K\)-dimensional array of counterfactual states and
-\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
-state space to the feature space. In Wachter, the state space is the
-feature space: \(f\) is just the identity function and the number of
-counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
-counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
-instead of the feature space directly. The latent embedding is learned
-by a separate generative model that is tasked with learning the data
-generating process (DGP) of \(X\). In this case \(f\) in
-Equation~\ref{eq-general} corresponds to the decoder part of the
-generative model, in other words the function that maps back from the
-latent embedding to the feature space. Provided the generative model is
-well-specified, traversing the latent embedding typically results in
-realistic and plausible counterfactuals, because they are implicitly
-generated by the (learned) DGP
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}.
-
-CLUE distinguishes itself from REVISE and other counterfactual
-generators in that it aims to minimize the predictive uncertainty of the
-model in question, \(M\). To quantify predictive uncertainty the authors
-rely on entropy estimates for probabilistic models. The approach
-proposed by \protect\hyperlink{ref-schut2021generating}{{[}7{]}}, which
-we shall refer to as \textbf{Greedy}, also works with the subclass of
-models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
-predictive uncertainty estimates. The authors show that in this setting
-the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
-and meaningful counterfactuals can be generated in a fast and efficient
-manner through a modified Jacobian-based Saliency Map Attack (JSMA).
-Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}7{]}} also
-show that by maximizing the predicted probability of \(x^\prime\) being
-assigned to target class \(y^*\) we also implicitly minimize predictive
-entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
-REVISE in the Bayesian context and we shall therefore refer to both
-approaches collectively as \textbf{Latent Space} generators\footnote{In
-  fact, there are a number of other recently proposed approaches to
-  counterfactual search that also broadly fall in this same category.
-  They largely differ with respect to the chosen generative model: for
-  example, the generator proposed by
-  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}24{]}} relies
-  on normalizing flows.}.
-
-Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
-distinguishes itself from all other generators considered here in that
-it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
-al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} show that
-diverse outcomes can in principal be achieved simply rerunning
-counterfactual search multiple times using stochastic gradient descent
-(or by randomly initializing the counterfactual). In
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} diversity is
-explicitly proxied via Determinantal Point Processes (DDP): the authors
-simply introduce DDP as a component of the cost function
-\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
-\(s_1, ... , s_K\) that look as different from each other as possible.
-The implementation of DiCE in the our library of choice -
-\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
-worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
-constant and therefore does not affect the objective function
-Equation~\ref{eq-general}.
-
-\hypertarget{towards-collective-recourse}{%
-\subsection{\ldots{} towards collective
-recourse}\label{towards-collective-recourse}}
-
-All of the different approaches introduced above tackle the problem of
-Algorithmic Recourse from the perspective of one single
-individual\footnote{DiCE recognizes that different individuals may have
-  different objective functions, but it does not address the
-  interdependencies between different individuals.}. To explicitly
-address the issue that individual recourse may affect the outcome and
-prospect of other individuals, we propose to extend
-Equation~\ref{eq-general} as follows:
-
-\begin{equation}\protect\hypertarget{eq-collective}{}{
-\begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
-\end{aligned}
-}\label{eq-collective}\end{equation}
-
-Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
-the individual as before and \(\lambda_1\) governs to what extent that
-private cost ought to be penalized. The newly introduced term
-\(g(f(s_k^\prime))\) is meant to capture and address external costs
-incurred by the collective of individuals in response to changes in
-\(\mathbf{s}^\prime\). The underlying concept of private and external
-costs is borrowed from Economics and well-established in that field:
-when the decisions or actions by some individual market participant
-generate external costs, then the market is said to suffer from negative
-externalities and considered inefficient
-\protect\hyperlink{ref-pindyck2014microeconomics}{{[}25{]}}. We think
-that this concept describes the endogenous dynamics of algorithmic
-recourse oberserved here very well. As with individual recourse, the
-exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide
-a definite answer in this work, if such even exists. That being said, we
-do propose a few potential mitigation strategies in
-Section~\ref{sec-empirical-2-mitigate}.
-
-\hypertarget{sec-method-2}{%
-\section{Modeling Endogenous Macrodynamics in Algorithmic
-Recourse}\label{sec-method-2}}
-
-In the following we describe the framework we propose for modeling and
-analysing endogenous macrodynamics in Algorithmic Recourse. We first
-describe the basic simulations that were generated to produce the
-findings in this work and also constitute the core of
-\texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we
-introduced earlier. The remainder of this section then introduces
-various evaluation metrics that can be used to benchmark different
-counterfactual generators with respect to how they perform in the
-dynamic setting.
-
-\hypertarget{sec-method-2-experiment}{%
-\subsection{Simulations}\label{sec-method-2-experiment}}
-
-The dynamics illustrated in Figure~\ref{fig-poc} in
-Section~\ref{sec-intro} were generated through a simple experiment that
-aims to simulate the process of Algorithmic Recourse in practice. We
-begin in the static setting at time time \(t=0\): given some classifier
-\(M\) that was pre-trained on data
-\(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse
-for a random batch of \(B\) individuals in the non-target class
-(\(\mathcal{D}_0\)). Note that we focus our attention on classification
-problems, since classification poses the most common practical use-case
-for Algorithmic Recourse. In order to simulate the dynamic process, we
-suppose that the model \(M\) is retrained following the actual
-implementation of recourse in time \(t=0\). Following the update to the
-model, we assume that at time \(t=1\) recourse is generated for yet
-another random subset of individuals in the non-target class. This
-process is repeated for a number of time periods \(T\). To get a clean
-read on endogenous dynamics we keep the total population of samples
-closed: we allow existing samples to move from factual to counterfactual
-states, but do not allow any entirely new samples to enter the
-population. The experimental setup is summarized in Algorithm
-\ref{algo-experiment}
-
-\begin{algorithm}
-\caption{Simulation Experiment}\label{algo-experiment}
-\begin{algorithmic}[1]
-\Procedure{Experiment}{$M,\mathcal{D},G$}
-\State $t\gets 0$
-\While{$t<T$}
-\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
-\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
-\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
-\EndWhile
-\State \textbf{return} $M,\mathcal{D}$
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-A noteworthy practical consideration is the choice of \(T\) and \(B\).
-The higher these values, the more factual instances undergo recourse
-throughout the entire experiment. Of course, this is likely to lead to
-more pronounced domain and model shifts by time \(T\). At the same time,
-it is generally improbable that a very large part of the population
-would request an explanation of the algorithm's decisions. In our
-experiments, we choose the values such that \(T \cdot B\) corresponds to
-the application of recourse on \(\approx50\%\) of the negative instances
-from the initial dataset. As we collect data at each time \(t\), we can
-also verify the impact of recourse when it is implemented for a smaller
-number of individuals.
-
-Algorithm \ref{algo-experiment} summarizes the proposed simulation
-experiment for a given dataset \(\mathcal{D}\), model \(M\) and
-generator \(G\), but naturally we are interested in comparing simulation
-outcomes for different sources of data, models and generators. The
-framework we have built facilitates this, making use of multi-threading
-in order to speed up computations. Holding the initial model and dataset
-constant the experiments are run for all generators, since our primary
-concern is to benchmark different recourse methods. To ensure that each
-generator is faced with the exact same initial conditions in each round
-\(t\), the candidate batch of individuals from the non-target class is
-randomly drawn from the intersection of all individuals in the
-non-target class across all experiments
-\(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where
-\(J\) is the total number of generators.
-
-\hypertarget{sec-method-2-metrics}{%
-\subsection{Evaluation Metrics}\label{sec-method-2-metrics}}
-
-We formulate two desiderata for the set of metrics used to measure
-domain and model shifts induced by recourse. First, the metrics should
-be applicable regardless of the dataset or classification technique so
-that they allow for the meaningful comparison of the generators in
-various scenarios. As the knowledge of the underlying probability
-distribution is rarely available, the metrics should be empirical and
-non-parametric. This further ensures that we can also measure large
-datasets by sampling from the available data. Moreover, while our study
-was conducted in a two-class classification setting, our choice of
-metrics should remain applicable in the future research on multi- class
-recourse problems. Second, the set of metrics should allow to capture
-various aspects of the previously mentioned magnitude, path, and tempo
-of changes while remaining as small as possible.
-
-\hypertarget{domain-shifts}{%
-\subsubsection{Domain Shifts}\label{domain-shifts}}
-
-To quantify the magnitude of domain shifts we rely on an unbiased
-estimate of the squared population \textbf{Maximum Mean Discrepancy
-(MMD)} given as:
-
-\begin{equation}\protect\hypertarget{eq-mmd}{}{
-\begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
-\end{aligned}
-}\label{eq-mmd}\end{equation}
-
-where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
-Space \(\mathcal{H}\)
-\protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}, and \(X\),
-\(\tilde{X}\) represent independent and identically distributed samples
-drawn from probability distributions \(\mathcal{X}\) and
-\(\mathcal{\tilde{X}}\) respectively
-\protect\hyperlink{ref-gretton2012kernel}{{[}27{]}}. MMD is a measure of
-the distance between the kernel mean embeddings of \(\mathcal{X}\) and
-\(\mathcal{\tilde{X}}\) in RKHS \(\mathcal{H}\). An important
-consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
-In our implementation we make use of Gaussian kernel with a constant
-length-scale parameter of \(0.5\). As the Gaussian kernel captures all
-moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we
-have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if \(X=\tilde{X}\).
-
-The evaluation metric in Equation~\ref{eq-mmd} is computed after every
-round \(t=1,...,T\) of the experiment. To assess the statistical
-significance of the observed shifts under the null hypothesis that
-samples \(X\) and \(\tilde{X}\) were drawn from the same probability
-distribution, we follow
-\protect\hyperlink{ref-arcones1992bootstrap}{{[}28{]}}. To that end, we
-combine the two samples and generate a large number of permutations of
-\(X + \tilde{X}\). Then, we split the permuted data into two new samples
-\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
-original samples. Then under the null hypothesis we should have that
-\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
-\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
-calculated by counting how these two quantities are not equal.
-
-We calculate the MMD for both classes individually based on the ground
-truth labels, i.e.~the labels that samples were assigned in time
-\(t=0\). Throughout our experiments, we generally do not expect the
-distribution of the negative class to change over time -- application of
-recourse reduces the size of this class, but since individuals are
-sampled uniformly the distribution should remain unaffected. Conversely,
-unless a recourse generator can perfectly replicate the original
-probability distribution, we expect the MMD of the positive class to
-increase. Thus, when discussing MMD, we generally mean the shift in the
-distribution of the positive class.
-
-\hypertarget{model-shifts}{%
-\subsubsection{Model Shifts}\label{model-shifts}}
-
-As our baseline for quantifying model shifts we measure perturbations to
-the model parameters at each point in time \(t\) following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
-\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
-between the vectors of parameters before and after retraining the model
-\(M\). We shall refer to this baseline metric simply as
-\textbf{Perturbations}.
-
-We extend the metric in Equation~\ref{eq-mmd} for the purpose of
-quantifying model shifts. Specifically, we introduce \textbf{Predicted
-Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
-features directly, we apply it to the predicted probabilities assigned
-to a set of samples by the model \(M\). If the model shifts, the
-probabilities assigned to each sample will change; again, this metric
-will equal 0 only if the two classifiers are the same. We compute PP MMD
-in two ways: firstly, we compute it over samples drawn uniformly from
-the dataset, and, secondly, we compute it over points spanning a
-meshgrid over a subspace of the entire feature space. For the latter
-approach we bound the subspace by the extrema of each feature. While
-this approach is theoretically more robust, unfortunately, it suffers
-from the curse of dimensionality, since it becomes increasingly
-difficult to select enough points to overcome noise as the dimension
-\(D\) grows.
-
-As an alternative to PP MMD we use a pseudo-distance for the
-\textbf{Disagreement Coefficient} (Disagreement). This metric was
-introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}29{]}} and
-estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
-classifiers do not agree on the predicted outcome for a randomly chosen
-sample. Thus, it is not relevant whether the classification is correct
-according to the ground truth, but only whether the sample lies on the
-same side of the two respective decision boundaries. In our context,
-this metric quantifies the overlap between the initial model (trained
-before the application of recourse) and the updated model. A
-Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal
-to zero a model shift may still have occured. This is one reason for why
-PP MMD is our our preferred metric.
-
-Finally, we introduce \textbf{Decisiveness} as a metric that quantifies
-the likelihood that a model assigns a high probability to its
-classification of any given sample. We define the metric simply as
-\({1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2\) where \(M(x)\) are
-predicted logits from a binary classifier and \(\sigma\) denotes the
-sigmoid function. This metric provides an unbiased estimate of the
-binary classifier's tendency to produce high-confidence predictions in
-either one of the two classes. Although the exact values for this metric
-are not important for our study, they can be used to detect model
-shifts. If decisiveness changes over time, then this is indicative of
-the decision boundary moves towards either one of the two classes.
-
-\hypertarget{sec-empirical}{%
-\section{Experiment Setup}\label{sec-empirical}}
-
-This section presents the exact ingredients and parameter choices
-describing the simulation experiments we ran to produce the findings
-presented in the next section (Section~\ref{sec-empirical-2}). For
-convenience, we use Algorithm \ref{algo-experiment} as a template to
-guide us through this section. A few high-level details upfront: each
-experiment is run for a total of \(T=50\) rounds, where in each round we
-provide recourse to five percent of all individuals in the non-target
-class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in
-  the previous section, we end up providing recourse to a total of
-  \(\approx50\%\) by the end of round \(T=50\).}. All classifiers and
-generative models are retrained for 10 epochs in each round \(t\) of the
-experiment. Rather than retraining models from scratch, we initialize
-all parameters at their previous levels (\(t-1\)) and compute
-backpropagate for 10 epochs using the new training data as inputs into
-the existing model. Evaluation metrics are computed and stored every 10
-rounds.
-
-\hypertarget{sec-empirical-classifiers}{%
-\subsection{\texorpdfstring{\(M\) -- Classifiers and Generative
-Models}{M -- Classifiers and Generative Models}}\label{sec-empirical-classifiers}}
-
-For each dataset and generator we look at three different types of
-classifiers all of them built and trained using \texttt{Flux.jl}
-\protect\hyperlink{ref-innes2018fashionable}{{[}30{]}}: firstly, a
-simple linear classifier - \textbf{Logistic Regression} - implemented as
-single linear layer with sigmoid activation; secondly, a multilayer
-perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
-composed of five MLPs following
-\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}31{]}} that serves
-as our only probabilistic classifier. We have chosen to work with deep
-ensembles both for their simplicity and effectiveness at modelling
-predictive uncertainty. They are also the model of choice in
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
-network architectures are kept simple (Table~\ref{tbl-mlp}), since we
-are only marginally concerned with achieving good initial classifier
-performance. For the real-world datasets we using mini-batch training
-and dropout regularization.
-
-The Latent Space generators rely on separate generative models.
-Following the authors of both REVISE and CLUE we use Variational
-Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers,
-we deliberately choose to work with fairly simple architectures
-(Table~\ref{tbl-vae}). More expressive generative models generally also
-lead to more meaningful counterfactuals produced by Latent Space
-generators. But in our view this should simply be considered as a
-vulnerability of counterfactual generators that rely on surrogate models
-to learn what realistic representations of the underlying data.
-
-\begin{table}
-
-\caption{\label{tbl-panel}Model
-Architectures}\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-mlp}MLP}
-
-{\centering
-
-\begin{tabular}[t]{lllll}
-\toprule
- & Hidden Dim. & Hidden Layers & Batch & Dropout\\
-\midrule
-Synthetic & 32 & 1 & - & -\\
-Real-World & 32 & 2 & 50 & 0.25\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-vae}Variational Autoencoder}
-
-{\centering
-
-\begin{tabular}[t]{lll}
-\toprule
- & Hidden Dim. & Epochs\\
-\midrule
-Synthetic & 2 & 100\\
-Real-World & 8 & 250\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{minipage}%
-
-\end{table}
-
-\hypertarget{sec-empirical-data}{%
-\subsection{\texorpdfstring{\(\mathcal{D}\) --
-Data}{\textbackslash mathcal\{D\} -- Data}}\label{sec-empirical-data}}
-
-We have chosen to work with both synthetic and real-world datasets.
-Using synthetic data allows us to impose distributional properties that
-may affect the resulting recourse dynamics. Following
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}, we generate
-synthetic data in \(\mathbb{R}^2\) to also allow for a visual
-interpretation of the results. Real-world data is used in order to
-assess if endogenous dynamics also occur in higher-dimensional settings.
-
-\hypertarget{synthetic-data}{%
-\subsubsection{Synthetic data}\label{synthetic-data}}
-
-We use four synthetic binary classification datasets consisting of 1000
-samples each. The datasets are presented in
-Figure~\ref{fig-synthetic-data} (see also Appendix A for a formal
-description). Samples from the negative class are marked in blue while
-samples of the positive class are marked in orange.
-
-\begin{figure}
-
-{\centering \includegraphics[width=8cm,height=2cm]{www/synthetic_data.png}
-
-}
-
-\caption{\label{fig-synthetic-data}Synthetic classification datasets
-used in our experiments.}
-
-\end{figure}
-
-Ex-ante we expect to see that by construction Wachter will create a new
-cluster of counterfactual instances in the proximity of the initial
-decision boundary. Thus, the choice of a black-box model may have an
-impact on the paths of the recourse. For generators that use latent
-space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
-CLUE \protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on
-(and have access to) probabilistic models (CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, Greedy
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
-counterfactuals will end up in regions of the target domain that are
-densely populated by training samples. Of course, this is expectation
-hinges on how effective said probabilistic models are at capturing
-predictive uncertainty. Finally, we expect to see the counterfactuals
-generated by DiCE to be uniformly spread around the feature space inside
-the target class\footnote{As we mentioned earlier, the diversity
-  constraint used by DiCE is only effective for when at least two
-  counterfactuals are being generated. We have therefore decided to
-  always generate 5 counterfactuals for each generator and randomly pick
-  one of them.}. In summary, we expect that the endogenous shifts
-induced by Wachter outsize those induced by all other generators, since
-Wachter is the only approach that is not concered with generating what
-we have defined as meaningful counterfactuals.
-
-\hypertarget{real-world-data}{%
-\subsubsection{Real-world data}\label{real-world-data}}
-
-We use three different real-world datasets from the Finance and
-Economics domain, all of which are tabular and can be used for binary
-classification. Firstly, we use the \textbf{Give Me Some Credit} dataset
-which was open-sourced on Kaggle for the task to predict whether a
-borrower is likely to experience financial difficulties in the next two
-years \protect\hyperlink{ref-gmsc_data}{{[}32{]}}. Originally consisting
-of 250,000 instances with 11 numerical attributes. Secondly, we use the
-\textbf{UCI defaultCredit} dataset
-\protect\hyperlink{ref-yeh2009comparisons}{{[}33{]}}, a benchmark
-dataset that can be used to train binary classifiers to predict the
-binary outcome variable, whether credit card clients default on their
-payment. In its raw form it consists of 23 explanatory variables - 4
-categorical features relating to demographic attributes\footnote{These
-  have been ommitted from the analysis. See Section~\ref{sec-limit-data}
-  for details.} and 19 continuous features largely relating to
-individuals' payment histories and amount of credit outstanding. Both of
-these datasets have been used in the literature on Algorithmic Recourse
-before (see for example
-\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
-\protect\hyperlink{ref-joshi2019towards}{{[}8{]}} and
-\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}), presumably
-because they constitute real-world classification tasks involving
-individuals that compete for access to credit.
-
-As a third dataset we include the \textbf{California Housing} dataset
-derived from the 1990 U.S. census and sourced through scikit-learn
-\protect\hyperlink{ref-pace1997sparse}{{[}35{]}}. It consists of 8
-continuous features that can be used to predict the median house price
-for California districts. The continuous outcome variable is binarized
-as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or
-not the median house price of a given district is above or below the
-median of all districts. While we have not seen this dataset used in the
-previous literature on AR, others have used the Boston Housing dataset
-in a similar fashion (see for example
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}). While we
-initially also conducted experiments on that dataset, we eventually
-discarded this dataset, since it has been found to suffer from an
-ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}36{]}}.
-
-Since the simulations involve generating counterfactuals for a
-significant proportion of the entire sample of individuals, we have
-randomly undersampled each dataset to yield balanced subsamples
-consisting of 10,000 individuals each. We have also standardized all
-explanatory features since our chosen classifiers are sensetive to
-scale.
-
-\hypertarget{g-generators}{%
-\subsection{\texorpdfstring{\(G\) --
-Generators}{G -- Generators}}\label{g-generators}}
-
-All generators introduced earlier are included in the experiments:
-Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
-REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
-\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DiCE
-\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and Greedy
-\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. In addition, we
-introduce two new generators in Section~\ref{sec-empirical-2-mitigate}
-that directly address the issue of endogenous domain and model shifts.
-We also test to what extent it may be beneficial to combine ideas
-underlying the various generators.
-
-\hypertarget{sec-empirical-2}{%
-\section{Experiments}\label{sec-empirical-2}}
-
-\hypertarget{endogenous-macrodynamics}{%
-\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
-
-\hypertarget{sec-empirical-2-mitigate}{%
-\subsection{Potential Mitigation
-Strategies}\label{sec-empirical-2-mitigate}}
-
-In the following we explore several ideas for strategies that may help
-to mitigate those types of endogenous recourse dynamics we observed in
-the previous subsection. All of them essentially boil down to one simple
-principle: to avoid substantial domain and model shifts, the generated
-counterfactuals should comply as much as possible with the true data
-generating process. This principle is really at the core of Latent Space
-generators, and hence it is not surprising that we have found these
-types of generators to perform comparably well in the previous
-subsection. But as we have mentioned earlier, generators that rely on
-separate generative models carry an additional computational cost and -
-perhaps more importantly - their performance hinges on the performance
-of said generative models. Fortunately, it turns out that we can use a
-number of other, much simpler strategies, which we will discuss now.
-
-\hypertarget{more-conservative-decision-thresholds}{%
-\subsubsection{More Conservative Decision
-Thresholds}\label{more-conservative-decision-thresholds}}
-
-The most obvious and trivial mitigation strategy is to simply choose a
-higher decision threshold \(\gamma\). Under \(\gamma=0.5\),
-counterfactuals will end up near the decision boundary by construction.
-Since this is the region of maximal aleotoric uncertainty, the
-classifier is bound to be thrown off. By simply setting a more
-conservative decision threshold, we can avoid this issue to some extent.
-A potential drawback of this approach is that a classifier with high
-decisiveness may classify samples with high confidence even far away
-from the training data.
-
-\hypertarget{classifier-preserving-roar}{%
-\subsubsection{Classifier Preserving
-ROAR}\label{classifier-preserving-roar}}
-
-Another potential strategy draws inspiration from ROAR
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}: to preserve the
-classifier, we propose to simply explicitly penalize the loss it incurs
-when evaluated on the counterfactual \(x^\prime\) at given parameter
-values. Recall that \(g(\cdot)\) denotes what we had defined as the
-external cost in Equation~\ref{eq-collective}. Then formally we let
-
-\begin{equation}\protect\hypertarget{eq-clap}{}{
-\begin{aligned}
-g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
-\end{aligned}
-}\label{eq-clap}\end{equation}
-
-for each counterfactual \(k\) where \(l\) denotes the loss function used
-to train \(M\). This approach, which we shall refer to as
-\textbf{ClapROAR}, is based on the intuition that (endogenous) model
-shifts will be triggered by counterfactuals that increase classifier
-loss. It is closely linked to the idea of choosing higher decision
-threshold, but likely better at avoiding the potential pitfalls
-associated with highly decisive classifiers. It also makes the private
-vs.~external cost trade-off more explicit and hence manageable.
-
-\hypertarget{gravitational-counterfactual-explanations}{%
-\subsubsection{Gravitational Counterfactual
-Explanations}\label{gravitational-counterfactual-explanations}}
-
-Yet another strategy simply extends Wachter as follows: instead of only
-penalizing the distance of the individuals' counterfactual to its
-factual, we propose penalizing its distance to some sensible point in
-the target domain, for example the sample average: \(\bar{x}\):
-
-\begin{equation}\protect\hypertarget{eq-clap}{}{
-\begin{aligned}
-g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
-\end{aligned}
-}\label{eq-clap}\end{equation}
-
-Once again we can putting this in the context of
-Equation~\ref{eq-collective}, the former penalty can be thought of here
-as the private cost incurred by the individual, while the latter
-reflects the external cost incurred by other individuals. Higher choices
-of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
-gravitate towards the specified point \(\bar{x}\) in the target domain.
-In the remainder of this paper we will therefore refer to this approach
-as \textbf{Gravitational} generator, when we investigate its potential
-usefulness for mitigating endongenous macrodynamics\footnote{Note that
-  despite the naming convention our goal here is not to provide yet
-  another counterfactual generator, but merely investigate the most
-  simple penalty we can think of with respect to its effectiveness.}.
-
-Figure~\ref{fig-mitigation} shows an illustrative example that
-demonstrates the differences in counterfactual outcomes when using the
-various mitigation strategies compared to the baseline approach, that is
-Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes
-the counterfactual a little further into the target domain; this effect
-is even stronger for ClapROAR; finally, using the Gravitational
-generator the counterfactual ends up all the way inside the target
-domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the
-  Graviational generator and ClapROAR to work as expected, one needs to
-  ensure that counterfactual search continues, independent of the
-  threshold probability \(\gamma\).}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/mitigation.png}
-
-}
-
-\caption{\label{fig-mitigation}The differences in counterfactual
-outcomes when using the various mitigation strategies compared to the
-baseline approach, that is Wachter with \(\gamma=0.5\)}
-
-\end{figure}
-
-Our findings \ldots{}
-
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Shift of focus from individual to group of individuals (related:
-  https://www.researchgate.net/publication/353073138\_Generating\_Collective\_Counterfactual\_Explanations\_in\_Score-Based\_Classification\_via\_Mathematical\_Optimization)
-\item
-  Convergence criterium matters: terminating once threshold probability
-  is reached may not be optimal (see e.g.~REVISE)
-\item
-  Optimizer choice matters: dimensionality is typically low, so no
-  obvious benefit to using ADAM.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    This might be better placed in JuliaCon proceedings, perhaps backed
-    by small blog post on the matter.
-  \end{itemize}
-\item
-  Mitigating strategy: penaliye distance from centroid.
-\end{enumerate}
-
-\hypertarget{sec-limit}{%
-\section{Limitations and Future Work}\label{sec-limit}}
-
-While we believe that this work constitutes a valuable starting point
-for addressing existing issues in Algorithmic Recourse from a fresh
-perspective, we are aware of several of its limitations. In the
-following we highlight some of these limitations and point to avenues
-for future research.
-
-\hypertarget{experimental-setup}{%
-\subsection{Experimental Setup}\label{experimental-setup}}
-
-The experimental setup proposed here is designed to mimic a real-world
-recourse process in a simple fashion. In practice, models are in fact
-updated on a regular basis
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also find it
-plausible to assume that the implementation of recourse happens
-periodically for different individuals, rather that all at once at time
-\(t=0\). That being said, our experimental design is a vast
-over-simplification of potential real-world scenarios. In practice, any
-endogenous shifts that may occur can be expected to be entangled with
-exogenous shifts of the nature investigated in
-\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also make
-implicit assumptions about the utility functions of the involved agents
-that may well be too simple: individuals seeking recourse are assumed to
-always implement the proposed Counterfactual Explanations; conversely,
-the agent in charge of the model \(M\) is assumed to always treat
-individuals that have implemented valid recourse as if they were truly
-now in the target class. Relating this back to the consumer credit
-example, we assume that the would-be borrowers are always willing and
-able to implement recourse and the bank is always willing to provide
-credit as would-be borrowers move across the decision boundary. In
-practice it is doubtful that agents behave according to such simple
-rules. Nonetheless, we think that our simple framework offers a starting
-point for future work on recourse dynamics (both endogenous and
-exogenous dynamics).
-
-\hypertarget{sec-limit-data}{%
-\subsection{Data}\label{sec-limit-data}}
-
-Largely in line with the existing literature on Algorithmic Recourse, we
-have limited our analysis of real-world data to three commonly used
-benchmark datasets that involve binary prediction tasks. Future work may
-benefit from including novel datasets or extending the analysis to
-multi-class or regression problems, the latter arguably representing the
-most common objective in Finance and Economics. It is also worth
-mentioning that the use of real-world datasets considered in this work
-is constrained by the fact that at the time of writing
-\texttt{CounterfactualExplanations.jl} only supports continuous
-features, at least of some of the counterfactual generators considered
-here. The fact that we therefore had to discard discrete features led to
-relatively poor initial performance of our classifiers in some cases.
-While this is indeed a limitation we intend to address in future and
-derivative work, our findings with respect to endogenous macrodynamics
-do not hinge on strong classifier performance.
-
-\hypertarget{classifiers}{%
-\subsection{Classifiers}\label{classifiers}}
-
-For reasons stated earlier we have limited our analysis to
-differentiable linear and non-linear classifiers, in particular logistic
-regression and deep neural networks. While these sorts of classifiers
-have also typically been analyzed in the existing literature on
-Counterfactual Explanations and Algorithmic Recourse, they represent
-only a subset of popular machine learning models employed in practice -
-both black-box and glass-box. Despite the success and popularity of deep
-learning in the context of high-dimensional data such as image, audio
-and video, empirical evidence suggests that other models such as boosted
-decision trees may have an edge when it comes to lower-dimensional
-tabular datasets, such as the ones considered here
-\protect\hyperlink{ref-grinsztajn2022tree}{{[}38{]}}.
-
-\hypertarget{sec-conclusion}{%
-\section{Concluding Remarks}\label{sec-conclusion}}
-
-\hypertarget{acknowledgment}{%
-\section*{Acknowledgment}\label{acknowledgment}}
-\addcontentsline{toc}{section}{Acknowledgment}
-
-P. A. thanks \ldots{}
-
-\pagebreak
-\FloatBarrier
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
-\CSLLeftMargin{{[}1{]} }%
-\CSLRightInline{C. Borch, {``Machine learning, knowledge risk, and
-principal-agent problems in automated trading,''} \emph{Technology in
-Society}, p. 101852, 2022.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
-\CSLLeftMargin{{[}2{]} }%
-\CSLRightInline{C. O'neil, \emph{Weapons of math destruction: How big
-data increases inequality and threatens democracy}. Crown, 2016.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
-\CSLLeftMargin{{[}3{]} }%
-\CSLRightInline{C. Rudin, {``Stop explaining black box machine learning
-models for high stakes decisions and use interpretable models
-instead,''} \emph{Nature Machine Intelligence}, vol. 1, no. 5, pp.
-206--215, 2019.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
-\CSLLeftMargin{{[}4{]} }%
-\CSLRightInline{M. Pawelczyk, S. Bielawski, J. van den Heuvel, T.
-Richter, and G. Kasneci, {``Carla: A python library to benchmark
-algorithmic recourse and counterfactual explanation algorithms,''}
-\emph{arXiv preprint arXiv:2108.00783}, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
-\CSLLeftMargin{{[}5{]} }%
-\CSLRightInline{S. Wachter, B. Mittelstadt, and C. Russell,
-{``Counterfactual explanations without opening the black box: Automated
-decisions and the GDPR,''} \emph{Harv. JL \& Tech.}, vol. 31, p. 841,
-2017.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
-\CSLLeftMargin{{[}6{]} }%
-\CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
-Julia package for Counterfactual Explanations and Algorithmic
-Recourse}}. 2022. Available:
-\url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
-\CSLLeftMargin{{[}7{]} }%
-\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
-counterfactual explanations by implicit minimisation of epistemic and
-aleatoric uncertainties,''} in \emph{International conference on
-artificial intelligence and statistics}, 2021, pp. 1756--1764.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
-\CSLLeftMargin{{[}8{]} }%
-\CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
-Ghosh, {``Towards realistic individual recourse and actionable
-explanations in black-box decision making systems,''} \emph{arXiv
-preprint arXiv:1907.09615}, 2019.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
-\CSLLeftMargin{{[}9{]} }%
-\CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
-machine learning classifiers through diverse counterfactual
-explanations,''} in \emph{Proceedings of the 2020 conference on
-fairness, accountability, and transparency}, 2020, pp. 607--617.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-\CSLLeftMargin{{[}10{]} }%
-\CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
-Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty
-estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
-\CSLLeftMargin{{[}11{]} }%
-\CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
-{``A survey of algorithmic recourse: Definitions, formulations,
-solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050},
-2020.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
-\CSLLeftMargin{{[}12{]} }%
-\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
-explanations for machine learning: A review,''} \emph{arXiv preprint
-arXiv:2010.10596}, 2020.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
-\CSLLeftMargin{{[}13{]} }%
-\CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
-recourse in linear classification,''} in \emph{Proceedings of the
-conference on fairness, accountability, and transparency}, 2019, pp.
-10--19.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
-\CSLLeftMargin{{[}14{]} }%
-\CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
-{``Algorithmic recourse: From counterfactual explanations to
-interventions,''} in \emph{Proceedings of the 2021 ACM conference on
-fairness, accountability, and transparency}, 2021, pp. 353--362.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
-\CSLLeftMargin{{[}15{]} }%
-\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
-robust and reliable algorithmic recourse,''} \emph{arXiv preprint
-arXiv:2102.13620}, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
-\CSLLeftMargin{{[}16{]} }%
-\CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero,
-{``Generating collective counterfactual explanations in score-based
-classification via mathematical optimization,''} 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
-\CSLLeftMargin{{[}17{]} }%
-\CSLRightInline{S. Rabanser, S. Günnemann, and Z. Lipton, {``Failing
-loudly: An empirical study of methods for detecting dataset shift,''}
-\emph{Advances in Neural Information Processing Systems}, vol. 32,
-2019.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
-\CSLLeftMargin{{[}18{]} }%
-\CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly
-detection: A survey,''} \emph{ACM computing surveys (CSUR)}, vol. 41,
-no. 3, pp. 1--58, 2009.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
-\CSLLeftMargin{{[}19{]} }%
-\CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of
-concept drift and hidden contexts,''} \emph{Machine learning}, vol. 23,
-no. 1, pp. 69--101, 1996.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
-\CSLLeftMargin{{[}20{]} }%
-\CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A.
-Bouchachia, {``A survey on concept drift adaptation,''} \emph{ACM
-computing surveys (CSUR)}, vol. 46, no. 4, pp. 1--37, 2014.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
-\CSLLeftMargin{{[}21{]} }%
-\CSLRightInline{K. Nelson, G. Corbin, M. Anania, M. Kovacs, J. Tobias,
-and M. Blowers, {``Evaluating model drift in machine learning
-algorithms,''} in \emph{2015 IEEE symposium on computational
-intelligence for security and defense applications (CISDA)}, 2015, pp.
-1--8.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
-\CSLLeftMargin{{[}22{]} }%
-\CSLRightInline{S. Ackerman, P. Dube, E. Farchi, O. Raz, and M.
-Zalmanovici, {``Machine learning model drift detection via weak data
-slices,''} in \emph{2021 IEEE/ACM third international workshop on deep
-learning for testing and testing for deep learning (DeepTest)}, 2021,
-pp. 1--8.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
-\CSLLeftMargin{{[}23{]} }%
-\CSLRightInline{R. M. B. de Oliveira and D. Martens, {``A framework and
-benchmarking study for counterfactual generating methods on tabular
-data,''} \emph{Applied Sciences}, vol. 11, no. 16, p. 7274, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
-\CSLLeftMargin{{[}24{]} }%
-\CSLRightInline{A.-K. Dombrowski, J. E. Gerken, and P. Kessel,
-{``Diffeomorphic explanations with normalizing flows,''} 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
-\CSLLeftMargin{{[}25{]} }%
-\CSLRightInline{R. S. Pindyck and D. L. Rubinfeld,
-\emph{Microeconomics}. Pearson Education, 2014.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
-\CSLLeftMargin{{[}26{]} }%
-\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing
-kernel hilbert spaces in probability and statistics}. Springer Science
-\& Business Media, 2011.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
-\CSLLeftMargin{{[}27{]} }%
-\CSLRightInline{A. Gretton, K. M. Borgwardt, M. J. Rasch, B. Schölkopf,
-and A. Smola, {``A kernel two-sample test,''} \emph{The Journal of
-Machine Learning Research}, vol. 13, no. 1, pp. 723--773, 2012.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-\CSLLeftMargin{{[}28{]} }%
-\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
-v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-\CSLLeftMargin{{[}29{]} }%
-\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
-agnostic active learning,''} in \emph{Proceedings of the 24th
-international conference on machine learning}, 2007, pp. 353--360.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
-\CSLLeftMargin{{[}30{]} }%
-\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with
-flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
-\CSLLeftMargin{{[}31{]} }%
-\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell,
-{``Simple and scalable predictive uncertainty estimation using deep
-ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-\CSLLeftMargin{{[}32{]} }%
-\CSLRightInline{K. Competition, {``Give me some credit, improve on the
-state of the art in credit scoring by predicting the probability that
-somebody will experience financial distress in the next two years.''}
-\url{https://www.kaggle.com/c/GiveMeSomeCredit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
-\CSLLeftMargin{{[}33{]} }%
-\CSLRightInline{I.-C. Yeh and C. Lien, {``The comparisons of data mining
-techniques for the predictive accuracy of probability of default of
-credit card clients,''} \emph{Expert systems with applications}, vol.
-36, no. 2, pp. 2473--2480, 2009.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
-\CSLLeftMargin{{[}34{]} }%
-\CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine
-learning in python,''} \emph{the Journal of machine Learning research},
-vol. 12, pp. 2825--2830, 2011.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
-\CSLLeftMargin{{[}35{]} }%
-\CSLRightInline{R. K. Pace and R. Barry, {``Sparse spatial
-autoregressions,''} \emph{Statistics \& Probability Letters}, vol. 33,
-no. 3, pp. 291--297, 1997.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
-\CSLLeftMargin{{[}36{]} }%
-\CSLRightInline{C. M., {``Racist data destruction? - a boston housing
-dataset controversy,''} 2019, Available:
-\url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}37{]} }%
-\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
-Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: A
-survey,''} \emph{arXiv preprint arXiv:2110.01889}, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-\CSLLeftMargin{{[}38{]} }%
-\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do
-tree-based models still outperform deep learning on tabular data?''}
-\emph{arXiv preprint arXiv:2207.08815}, 2022.}
-
-\end{CSLReferences}
-
-\pagebreak
-
-\hypertarget{tables}{%
-\section{Tables}\label{tables}}
-
-\ldots{}
-
-\pagebreak
-
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
-
-\ldots{}
-
-\pagebreak
-
-\hypertarget{code}{%
-\section{Code}\label{code}}
-
-\ldots{}
-
-\end{document}
+This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse.
+We demonstrate that long-held beliefs as to what defines optimality in AR, are too short-sighted to serve as a foundation for applications of recourse in practice.
+Specifically, we run multiple experiments that simulate the application of recourse in practice using various popular counterfactual generators and find that all of them induce substantial domain and model shifts.
+We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse.
+By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit.
+We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts.
+Through our work we hope to inspire future research on this important topic.
+To this end we have open-sourced all of our code along with a Julia package - `AlgorithmicRecourseDynamics.jl`.
+The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations].
+That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-08-15 dev/paper/sections/empirical_2.qmd

--- 2022-08-12 dev/paper/sections/empirical_2.qmd
@@
 # Experiments {#sec-empirical-2}
+
+Below we first present our main findings with respect to the presence of endogenous macrodynamics in Algorithmic Recourse.
+We then go on to introduce and test a number of potential mitigation strategies.
 
@@
 
-## Potential Mitigation Strategies
+In light of the huge number of experiments we have conducted, we shall start this section off with a few high-level observations.
+Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:
 
-### Gravitational Counterfactual Explanations {#sec-empirical-2-mitigate}
+- Statistically significant domain and model shift as measured by MMD. 
+- A deterioration in out-of-sample model performance as measured by the F-Score evaluated on a test sample. In many cases this drop in performance is substantial.
+- Significant perturbations to the model parameters as well as an increase in the model's decisiveness.
+- Disagreement between the original and retrained model, in some cases large.
 
-A straight-forward choice simply extends the baseline approach by @wachter2017counterfactual: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{\mathbf{x}}$.
-For such a recourse objective, higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point in the target domain.
-In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+There is also some clear heterogeneity across the results:
 
-#### A note on convergence
+- The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
+- The reduction in model performance appears to be most severe when classes are not perfectly separable.
+- With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.
 
-For this simple mitigating strategy underlying the Gravitational generator to work as expected, one needs to ensure that counterfactual search continues, even after a predetermined threshold probability $\gamma$ has potentially already been reached.
-@fig-convergence illustrates this distinction: if one chooses to terminate search once the desired threshold is reached (left panel) the gravitational pull towards $\bar{\mathbf{x}}$ is never actually satisfied (compare to right panel).
-More generally, if convergence is defined simply in terms of flipping the predicted label with some desired degree of confidence, this corresponds to essentially ignoring any parts of the counterfactual search objective that do not involve $\ell(M(f(s_k^\prime)),t)$ beyond that point.
-While this may be appropriate for some applications, in general this seems like an odd convention.
-Since we nonetheless seen convergence specified simply in terms of reaching the threshold probability in some places^[@joshi2019towards define convergence of Algorithm 1 in this way.
-The implementation of @wachter2017counterfactual in CARLA is also defined in this way.], we thought it worth making this distinction explicit.
+Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix.
+Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
+@fig-syn presents a small subset of our results for the synthetic dataset with overlapping classes.
+It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds.
+The top row shows the estimated domain shifts.
+While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds.
+With respect to the domain shifts, the Greedy generator actually induces the smallest shifts.
+In general, we have observed the opposite.
+The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier.
+As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
+In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
+The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.
 
-![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}
+![Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of points.
+Error bars indicate the standard deviation across folds.](www/synthetic_results.png){#fig-syn fig.pos="h" width="9cm" height="9cm"}
+
+Turning to the real-world data next ...
+
+![PLACEHOLDER: Results for real-world as table.](www/placeholder.png){#fig-real fig.pos="h" width="9cm" height="9cm"}

### 2022-08-15 dev/paper/sections/mitigation.qmd

# Potential Mitigation Strategies {#sec-empirical-2-mitigate}

In the following we explore several ideas for strategies that may help to mitigate those types of endogenous recourse dynamics we observed in the previous subsection. All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process. This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous subsection. But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational cost and - perhaps more importantly - their performance hinges on the performance of said generative models. Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.

## More Conservative Decision Thresholds

The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$. Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction. Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off. By simply setting a more conservative decision threshold, we can avoid this issue to some extent. A potential drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.

## Classifier Preserving ROAR

Another potential strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values. Recall that $g(\cdot)$ denotes what we had defined as the external cost in @eq-collective. Then formally we let

$$
\begin{aligned}
g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
\end{aligned}
$$ {#eq-clap}

for each counterfactual $k$ where $l$ denotes the loss function used to train $M$. This approach, which we shall refer to as **ClapROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss. It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers. It also makes the private vs. external cost trade-off more explicit and hence manageable.

## Gravitational Counterfactual Explanations

Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{x}$:

$$
\begin{aligned}
g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
\end{aligned}
$$ {#eq-clap}

Once again we can putting this in the context of @eq-collective, the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals. Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain. In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].

@fig-mitigation hows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].

```{julia}
#| eval: false
using Random
Random.seed!(2022)

# Data:
using MLJ
N = 1000
X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-5 => 5), cluster_std=0.5)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')

# Models:
using AlgorithmicRecourseDynamics
M = AlgorithmicRecourseDynamics.Models.FluxModel(counterfactual_data)
M = AlgorithmicRecourseDynamics.Models.train(M, counterfactual_data)

# Generators:
generators = Dict(
    "Generic (γ=0.5)" => GenericGenerator(decision_threshold=0.5),
    "Generic (γ=0.9)" => GenericGenerator(decision_threshold=0.9),
    "Gravitational" => GravitationalGenerator(),
    "ClapROAR" => ClapROARGenerator()
)

# Counterfactuals
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
T = 50
counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; T=T, latent_space=false) for (name, gen) in generators])

# Plots:
plts = []
for (name,ce) ∈ counterfactuals
    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false)
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(750,200), layout=(1,4))
savefig(plt, "dev/paper/www/mitigation.png")
```

![Illustrative example demonstrating the properties of the various mitigation strategies.](www/mitigation.png){#fig-mitigation fig.pos="h" width=45%}

Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts. @fig-mitigate-results presents a subset of the evaluation metrics for our synthetic data with overlapping classes. The top row in @fig-mitigate-results indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts. ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts. As evident from the middle row in @fig-mitigate-results model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective. The same overall pattern can be observed for out-of-sample model performance. With respect to the other synthetic datasets we found the following: for the Moons dataset the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.

![The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.](www/mitigation_synthetic_results.png){#fig-mitigate-results fig.pos="h" width=45%}

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. A snapshot of the results is shown in @fig-mitigate-latent-results.

![Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.](www/mitigation_synthetic_latent_results.png){#fig-mitigate-latent-results fig.pos="h" width="9cm" height="9cm"}

### 2022-08-15 paper/paper.Rmd
moved everything to R Mardkown and now template works out well with rticles

@@
 ---
-title: "Overlooked dynamics of algorithmic recourse"
-author:
-- "Patrick Altmeyer (p.altmeyer@tudelft.nl)"
-date: "`r format(Sys.Date(), '%B, %Y')`"
+title: Endogenous Macrodynamics in Algorithmic Recourse
+affiliation:
+  ## use one only of the following
+  # author-columnar: true         ## one column per author
+  institution-columnar: true  ## one column per institution (multiple autors eventually)
+  # wide: true                  ## one column wide author/affiliation fields
+
+  #author:   ## A custom author ordering field that breaks institution grouping.
+  #  - name: Eldon Tyrell
+  #    mark: 4
+  #    email: eldon@starfleet-academy.star
+  #  - name: Michael Shell
+  #    mark: 1
+  #  - name: Roy Batty
+  #    mark: 4
+  #    email: roy@replicant.offworld
+  ## Authors cited in institution field are ignored when author field exists
+
+  #author_multiline: ## Put authors in a given order, with multiline possibility. Authors cited in institution are ignored if exists
+  #  - line:         ## Create a new author line
+  #    - name: Michael Shell
+  #      mark: 1
+  #    - name: Homer Simpson
+  #      mark: 2
+  #      email: homer@thesimsons.com
+  #    - name: Roy Batty
+  #      mark: 4
+  #      email: roy@replicant.offworld
+  #  - line:         ## Create a second author line
+  #    - name: Montgomery Scott
+  #      mark: 3
+  #    - name: Eldon Tyrell
+  #      mark: 4
+
+  # institution:
+  #   - name: Georgia Institute of Technology
+  #     department: School of Electrical and Computer Engineering
+  #     location: Atlanta, Georgia 30332--0250
+  #     email: ece@datech.edu
+  #     mark: 1
+  #     author:
+  #       - name: Michael Shell
+  #   - name: Twentieth Century Fox
+  #     location: Springfield, USA
+  #     mark: 2
+  #     author:
+  #       - name: Homer Simpson
+  #         email: homer@thesimpsons.com
+  #   - name: Starfleet Academy
+  #     location: San Francisco, California 96678-2391
+  #     other: "Telephone: (800) 555--1212, Fax: (888) 555--1212"
+  #     mark: 3
+  #     author:
+  #       - name: Montgomery Scott
+  #   - name: Tyrell Inc.
+  #     location: 123 Replicant Street, Los Angeles, USA 90210--4321
+  #     mark: 4
+  #     author:
+  #       - name: Eldon Tyrell
+  #         email: eldon@starfleet-academy.star
+  #       - name: Roy Batty
+  #         email: roy@replicant.offworld
+keywords: ["keyword 1", "keyword 2"]
+abstract: |
+  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
+Through this work we aim to close that gap by systematizing and extending existing knowledge.
+We first show that many of the existing methodologies can be collectively described by a generalized framework.
+We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts.
+We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice.
+Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+Our simulation framework for studying recourse dynamics is fast and open-sourced.
+
+# use some specific Tex packages if needed. 
+# with_ifpdf: true
+# with_cite: true
+# amsmath need to be true to use with bookdown for referencing equations.
+with_amsmath: true
+# with_algorithmic: true
+# with_array: true
+# with_dblfloatfix: true
+
+bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
 output: 
-  bookdown::pdf_document2: 
-    toc: false
+  bookdown::pdf_book:
+    base_format: rticles::ieee_article
     includes:
-      in_header: ../tex/preamble.tex
-      before_body: ../tex/before_body.tex
-  bookdown::html_document2: default
-fontsize: 12pt
-bibliography: ../bib.bib
-link-citations: true
-abstract: "..."
+      in_header: "tex/preamble.tex"
+      before_body: "tex/before_body.tex"
+#citation_sorting: none   ## used as sorting option of the biblatex package (if selected)
 ---
 
-```{r setup, include=FALSE}
-rm(list=ls())
+```{r, include=FALSE}
+child_docs <- c(
+  "sections/introduction.rmd",
+  "sections/related.rmd",
+  "sections/methodology.rmd",
+  "sections/methodology_2.rmd",
+  "sections/empirical.rmd",
+  "sections/empirical_2.rmd",
+  "sections/mitigation.rmd",
+  "sections/discussion.rmd",
+  "sections/limitations.rmd",
+  "sections/conclusion.rmd"
+)
 knitr::opts_chunk$set(
@@
   message=FALSE,
-  out.width = 450,
-  fig.align = "center"
+  fig.align = "center",
+  out.width = "90%"
 )
@@
 
-```{r introduction, child="introduction.Rmd"}
+```{r, child=child_docs}
 ```
 
-\FloatBarrier
-\pagebreak
+# References {.unnumbered}
 
-# References {-}
+::: {#refs}
+:::
 
-<div id="refs"></div>
-
-\pagebreak
-
-```{r annex, child="appendix.Rmd"}
-```
+{{< include sections/appendix.qmd >}}

### 2022-08-15 paper/paper.tex
moved everything to R Mardkown and now template works out well with rticles

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%% template.tex
+%% from
+%% bare_conf.tex
+%% V1.4b
+%% 2015/08/26
+%% by Michael Shell
+%% See:
+%% http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
+%% conference paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/pkg/ieeetran
+%% and
+%% http://www.ieee.org/
+
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE!
+%% User assumes all risk.
+%% In no event shall the IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%*************************************************************************
+
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. The IEEE's font choices and paper sizes can   ***
+% *** trigger bugs that do not appear when using other class files.       ***                          ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
+\documentclass[conference,final,]{IEEEtran}
+% Some Computer Society conferences also require the compsoc mode option,
+% but others use the standard conference format.
 %
-\documentclass[
-  12pt,
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[conference]{../sty/IEEEtran}
+
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
+
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/pkg/ifpdf
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
+
+% *** CITATION PACKAGES ***
+%
+%\usepackage{cite}
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of the IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off
+% such as if a citation ever needs to be enclosed in parenthesis.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 5.0 (2009-03-20) and later if using hyperref.sty.
+% The latest version can be obtained at:
+% http://www.ctan.org/pkg/cite
+% The documentation is contained in the cite.sty file itself.
+
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation
+% can be obtained at:
+% http://www.ctan.org/pkg/graphicx
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found at:
+% http://www.ctan.org/pkg/epslatex
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
+
+\usepackage{graphicx}
+
+% *** MATH PACKAGES ***
+%
+\usepackage{amsmath}
+\interdisplaylinepenalty=2500
+%\usepackage{amsmath}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics.
+%
+% Note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/amsmath
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/pkg/algorithms
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/pkg/algorithmicx
+
+% *** ALIGNMENT PACKAGES ***
+%
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/array
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+% *** SUBFIGURE PACKAGES ***
+%\ifCLASSOPTIONcompsoc
+%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+%\else
+%  \usepackage[caption=false,font=footnotesize]{subfig}
+%\fi
+% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
+% for subfigure.sty, the latter of which is no longer maintained and is
+% incompatible with some LaTeX packages including fixltx2e. However,
+% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
+% which will override IEEEtran.cls' handling of captions and this will result
+% in non-IEEE style figure/table captions. To prevent this problem, be sure
+% and invoke subfig.sty's "caption=false" package option (available since
+% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
+% handling of captions.
+% Note that the Computer Society format requires a larger sans serif font
+% than the serif footnote size font used in traditional IEEE formatting
+% and thus the need to invoke different subfig.sty package options depending
+% on whether compsoc mode has been enabled.
+%
+% The latest version and documentation of subfig.sty can be obtained at:
+% http://www.ctan.org/pkg/subfig
+
+% *** FLOAT PACKAGES ***
+%
+
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure.
+% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
+% corrections already built into the system in which case a warning will
+% be issued if an attempt is made to load fixltx2e.sty as it is no longer
+% needed.
+% The latest version and documentation can be found at:
+% http://www.ctan.org/pkg/fixltx2e
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/stfloats
+% Do not use the stfloats baselinefloat ability as the IEEE does not allow
+% \baselineskip to stretch. Authors submitting work to the IEEE should note
+% that the IEEE rarely uses double column equations and that authors should try
+% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
+% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
+% such ways.
+% Do not attempt to use stfloats with fixltx2e as they are incompatible.
+% Instead, use Morten Hogholm'a dblfloatfix which combines the features
+% of both fixltx2e and stfloats:
+%
+% \usepackage{dblfloatfix}
+% The latest version can be found at:
+% http://www.ctan.org/pkg/dblfloatfix
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+%\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/url
+% Basically, \url{my_url_here}.
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
+
+%% BEGIN MY ADDITIONS %%
+
+\usepackage[unicode=true]{hyperref}
+
 \hypersetup{
-  pdftitle={Overlooked Dynamics in Algorithmic Recourse},
-  pdfauthor={Patrick Altmeyer (p.altmeyer@tudelft.nl)},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
+            pdfkeywords={keyword 1, keyword 2},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+
+% Pandoc toggle for numbering sections (defaults to be off)
+\setcounter{secnumdepth}{5}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
 \usepackage{longtable,booktabs,array}
@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+
+% Pandoc citation processing
 \newlength{\cslhangindent}
@@
 \setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
 \newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-% Packages:
+
+\IEEEoverridecommandlockouts
+\usepackage{cite}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{graphicx}
+\usepackage{textcomp}
+\usepackage{xcolor}
 \usepackage{placeins}
-\usepackage{hyperref}
-\usepackage{acmart}
-\KOMAoption{captions}{tableheading}
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\renewcommand*\contentsname{Table of contents}
-\renewcommand*\listfigurename{List of Figures}
-\renewcommand*\listtablename{List of Tables}
-\renewcommand*\figurename{Figure}
-\renewcommand*\tablename{Table}
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
+    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
-\title{Overlooked Dynamics in Algorithmic Recourse}
-\author{Patrick Altmeyer (p.altmeyer@tudelft.nl)}
-\date{April, 2022}
+%% END MY ADDITIONS %%
+
+\hyphenation{op-tical net-works semi-conduc-tor}
 
 \begin{document}
+%
+% paper title
+% Titles are generally capitalized except for words such as a, an, and, as,
+% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
+% not capitalized unless they are the first or last word of the title.
+% Linebreaks \\ can be used within to get better formatting as desired.
+% Do not put math or special symbols in the title.
+\title{Endogenous Macrodynamics in Algorithmic Recourse}
+
+% author names and affiliations
+% use a multiple column layout for up to three different
+% affiliations
+
+\author{
+
+%% ---- classic IEEETrans wide authors' list ----------------
+ % -- end affiliation.wide
+%% ----------------------------------------------------------
+
+%% ---- classic IEEETrans one column per institution --------
+ %% -- beg if/affiliation.institution-columnar
+ %% -- end for/affiliation.institution
+ %% -- end if/affiliation.institution-columnar
+%% ----------------------------------------------------------
+
+%% ---- one column per author, classic/default IEEETrans ----
+ %% -- end if/affiliation.institution-columnar
+%% ----------------------------------------------------------
+
+}
+
+% conference papers do not typically use \thanks and this command
+% is locked out in conference mode. If really needed, such as for
+% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
+% after \documentclass
+
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
+%
+%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
+%Homer Simpson\IEEEauthorrefmark{2},
+%James Kirk\IEEEauthorrefmark{3},
+%Montgomery Scott\IEEEauthorrefmark{3} and
+%Eldon Tyrell\IEEEauthorrefmark{4}}
+%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
+%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
+
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% make the title area
 \maketitle
+
+% As a general rule, do not put math, special symbols or citations
+% in the abstract
 \begin{abstract}
-\ldots{}
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
+Through this work we aim to close that gap by systematizing and extending existing knowledge.
+We first show that many of the existing methodologies can be collectively described by a generalized framework.
+We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts.
+We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice.
+Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+Our simulation framework for studying recourse dynamics is fast and open-sourced.
 \end{abstract}
 
-% Include before body: ----
-\pagenumbering{gobble}
+% keywords
+\begin{IEEEkeywords}
+keyword 1; keyword 2
+\end{IEEEkeywords}
 
-\newenvironment{acknowledgements}%
-{\begin{center}%
-\bfseries Acknowledgements
-\end{center}}%
-{\vfill\null}
-\begin{acknowledgements}
-...
-\end{acknowledgements}
-{\vfill\null}
-\textbf{DISCLAIMER:}
-This is a work-in-progress. Please do not distribute or share.
-\pagebreak
+% use for special paper notices
 
-\pagenumbering{arabic}
+% make the title area
+\maketitle
 
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
+% no keywords
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
+
+\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
 }
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
 
-\ldots{}
+\hypertarget{intro}{%
+\section{Introduction}\label{intro}}
 
-\FloatBarrier
-\pagebreak
+Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences.
+While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making.
+They are essentially \textbf{black boxes}: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools.
+Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes (O'neil 2016).
+The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models (Rudin 2019), others have investigated different ways to explain the behaviour of black-box models.
+These two sub-domains can be broadly referred to as \textbf{interpretable AI} and \textbf{explainable AI} (XAI), respectively.
 
-References
+Among the approaches to XAI that have recently grown in popularity are \textbf{Counterfactual Explanations} (CE).
+They explain how inputs into a model need to change for it to produce different outputs.
+Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of \textbf{Algorithmic Recourse} (AR) to help individuals who face adverse outcomes.
+An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system.
+A meaningful recourse recommendation for a denied applicant could be: \emph{``If your net savings rate had been 10\% of your monthly income instead of the actual 8\%, your application would have been successful.
+See if you can temporarily cut down on consumption.''} In the remainder of this paper we will use both terminologies - recourse and counterfactual - interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
+
+Existing work in this field has largely worked in a static setting: various approaches have been proposed to generate counterfactuals for a given individual that is subject to some pre-trained model.
+More recent work has compared different approaches within this static setting (Pawelczyk et al.
+2021).
+In this work, we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly?
+What types of dynamics are introduced and how do different counterfactual generators compare in this context?
+
+Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as \textbf{individual recourse} in some places.
+Arguably though, most real-world applications that warrant Algorithmic Recourse involve potentially large groups of individuals typically competing for scarce resources.
+Our work demonstrates that in such scenarios, choices made by or for one single individual are likely to affect the broader collective of individuals in ways that current approaches to AR fail to account for.
+More specifically, we argue that a strict focus on minimizing the private costs faced by individuals may be too narrow an objective.
+
+Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by Wachter, Mittelstadt, and Russell (2017): the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
+We refer to these types of dynamics as \textbf{endogenous} because they are induced by the implementation of recourse itself.
+The term \textbf{macrodynamics} is borrowed from the economics literature and used to describe processes involving whole groups or societies.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/poc}
+
+}
+
+\caption{Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).}\label{fig:poc}
+\end{figure}
+
+We think that these types of endogenous dynamics may be problematic and warrant our attention.
+Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+Secondly and relatedly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+Average default risk across all borrowers can therefore be expected to increase.
+Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals.
+In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.
+
+To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups.
+Secondly, we introduce an experimental framework extending previous work by Altmeyer (2022), which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including Wachter, Mittelstadt, and Russell (2017), Schut et al.
+(2021), Joshi et al.
+(2019), Mothilal, Sharma, and Tan (2020) and Antorán et al.
+(2020).
+To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators.
+Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.
+
+The remainder of the paper is structured as follows: Section \ref{related} places our work in the broader context of related literature.
+Section \ref{method} posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
+Section \ref{method-2} sets out our experimental framework for modeling endogenous macrodynamics in AR.
+Section \ref{empirical} presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
+Section \ref{empirical-2} presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
+Our findings are then discussed in the broader context of the literature in Section \ref{discussion}, before pointing to some of the limitations of our work as well as avenues for future research in Section \ref{limit}.
+Finally, Section \ref{conclusion} concludes.
+
+\hypertarget{related}{%
+\section{Background}\label{related}}
+
+In this Section we provide a review of the relevant literature.
+First, Subsection \ref{related-recourse} discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
+Then, Subsection \ref{related-shifts} presents some of the previous work on the measurement of dataset and model shifts.
+
+\hypertarget{related-recourse}{%
+\subsection{Algorithmic Recourse}\label{related-recourse}}
+
+A framework for Counterfactual Explanations was first proposed in 2017 by Wachter, Mittelstadt, and Russell (2017) and has served as the baseline for most methodologies that have been proposed since then.
+Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\).
+Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):
+
+\begin{equation}
+\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
+\end{equation}
+
+For implementation purposes, \eqref{eq:obj} is typically approximated through regularization:
+
+\begin{equation}
+x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) \label{eq:solution}
+\end{equation}
+
+In the baseline work (Wachter, Mittelstadt, and Russell 2017), the cost function is proxied by some distance metric based on the simple intuition that perturbations of \(x\) are costly to the individual.
+For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent.
+This summarizes the approach followed in Wachter, Mittelstadt, and Russell (2017) which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.
+
+Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017.
+An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 (Karimi et al.
+2020).
+Another survey published around the same time by Verma et al.~described 29 algorithms (Verma, Dickerson, and Hines 2020).
+Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met.
+Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs.
+To name but a few, we are interested in generating counterfactuals that close (Wachter, Mittelstadt, and Russell 2017), actionable (Ustun, Spangher, and Liu 2019), realistic Schut et al.
+(2021), sparse, diverse (Mothilal, Sharma, and Tan 2020) and if possible causally founded (Karimi, Schölkopf, and Valera 2021).
+
+Efforts so far have largely been directed at improving the quality of Counterfactual Explanations within a static context: given some pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are interested in generating one or multiple meaningful Counterfactual Explanations for some individual characterized by \(x\).
+The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point (Verma, Dickerson, and Hines 2020).
+We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR (Upadhyay, Joshi, and Lakkaraju 2021).
+Exogenous shifts are strictly of external origin.
+For example, they might stem from data correction, temporal shifts or geospatial changes (Upadhyay, Joshi, and Lakkaraju 2021).
+The authors of (Upadhyay, Joshi, and Lakkaraju 2021) propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
+
+As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances.
+We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals (Carrizosa, Ramırez-Ayerbe, and Romero 2021).
+The authors propose an optimization framework that generates collective counterfactuals.
+We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.
+
+\hypertarget{related-shifts}{%
+\subsection{Domain and Model Shifts}\label{related-shifts}}
+
+Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time.
+Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests (Rabanser, Günnemann, and Lipton 2019).
+This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution.
+Numerous approaches to anomaly detection have been summarized (Chandola, Banerjee, and Kumar 2009).
+Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model (Widmer and Kubat 1996).
+For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift (Gama et al.
+2014).
+Less previous work is available on the related topic of model drift - changes in model performance over time.
+Nelson et al.~review how resistant different machine learning models are to the model drift (Nelson et al.
+2015).
+Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available (Ackerman et al.
+2021).
+
+In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR (Upadhyay, Joshi, and Lakkaraju 2021).
+In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\).
+While this also sets the baseline for our analysis here, it is worth noting that in Upadhyay, Joshi, and Lakkaraju (2021) these perturbations are mechanically introduced.
+In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
+In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model.
+We have not been able to identify previous work on this topic.
+
+\hypertarget{related-benchmark}{%
+\subsection{Benchmarking Counterfactual Generators}\label{related-benchmark}}
+
+Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies.
+This may be partially due to limited software availability in this space.
+Recent work has started to address this gap: firstly, Oliveira and Martens (2021) run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, Pawelczyk et al.
+(2021) introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} (Altmeyer 2022) provides an extensible, fast and language-agnostic implementation in Julia.
+Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
+In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is available from \ldots{}}.
+The core concepts implemented in that package reflect what is presented in Section \ref{method-2} of this paper.
+
+\hypertarget{method}{%
+\section{Gradient-Based Recourse Revisited}\label{method}}
+
+In this section we first set out a generalized framework for gradient-based counterfactual search in Section \ref{method-general} that encapsulates the various individual recourse methods we have chosen to use in our experiments.
+We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.
+
+\hypertarget{method-general}{%
+\subsection{From individual recourse \ldots{}}\label{method-general}}
+
+We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
+In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} (Joshi et al.
+2019), \textbf{CLUE} (Antorán et al.
+2020), \textbf{DiCE} (Mothilal, Sharma, and Tan 2020) and a greedy approach that relies on probabilistic models (Schut et al.
+2021).
+Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
+We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}.
+This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
+All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
+
+As we will see next, all of them can be described by the following generalized form of Equation \eqref{eq:general}:
+
+\begin{equation}
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} \label{eq:general}
+\end{aligned} 
+\end{equation}
+
+Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space.
+In Wachter, the state space is the feature space: \(f\) is just the identity function and the number of counterfactuals \(K\) is equal to one.
+Both REVISE and CLUE search counterfactuals in some latent embedding \(S \subset \mathcal{S}\) instead of the feature space directly.
+The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\).
+In this case \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP (Joshi et al.
+2019).
+
+CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\).
+To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
+The approach proposed by Schut et al.
+(2021), which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
+The authors show that in this setting the cost function \(h(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+Schut et al.
+(Schut et al.
+2021) also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE.
+In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
+They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski, Gerken, and Kessel (2021) relies on normalizing flows.}.
+
+Finally, DiCE (Mothilal, Sharma, and Tan 2020) distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals.
+Wachter et al.
+(Wachter, Mittelstadt, and Russell 2017) show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
+In Mothilal, Sharma, and Tan (2020) diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ...
+, s_K\) that look as different from each other as possible.
+The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach.
+It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.
+
+\hypertarget{towards-collective-recourse}{%
+\subsection{\ldots{} towards collective recourse}\label{towards-collective-recourse}}
+
+All of the different approaches introduced above tackle the problem of Algorithmic Recourse from the perspective of one single individual\footnote{DiCE recognizes that different individuals may have different objective functions, but it does not address the interdependencies between different individuals.}.
+To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend Equation \eqref{eq:general} as follows:
+
+\begin{equation}
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  \label{eq:collective}
+\end{aligned} 
+\end{equation}
+
+Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized.
+The newly introduced term \(g(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\).
+The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient (Pindyck and Rubinfeld 2014).
+We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
+As with individual recourse, the exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
+That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.
+
+\hypertarget{method-2}{%
+\section{Modeling Endogenous Macrodynamics in Algorithmic Recourse}\label{method-2}}
+
+In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
+We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we introduced earlier.
+The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
+
+\hypertarget{method-2-experiment}{%
+\subsection{Simulations}\label{method-2-experiment}}
+
+The dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
+We begin in the static setting at time time \(t=0\): given some classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)).
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse.
+In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\).
+Following the update to the model, we assume that at time \(t=1\) recourse is generated for yet another random subset of individuals in the non-target class.
+This process is repeated for a number of time periods \(T\).
+To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population.
+The experimental setup is summarized in Algorithm \ref{algo-experiment}
+
+\begin{algorithm}
+\caption{Simulation Experiment}\label{algo-experiment}
+\begin{algorithmic}[1]
+\Procedure{Experiment}{$M,\mathcal{D},G$}
+\State $t\gets 0$
+\While{$t<T$}
+\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
+\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
+\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
+\EndWhile
+\State \textbf{return} $M,\mathcal{D}$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+A noteworthy practical consideration is the choice of \(T\) and \(B\).
+The higher these values, the more factual instances undergo recourse throughout the entire experiment.
+Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\).
+At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm's decisions.
+In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset.
+As we collect data at each time \(t\), we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
+
+Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset \(\mathcal{D}\), model \(M\) and generator \(G\), but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators.
+The framework we have built facilitates this, making use of multi-threading in order to speed up computations.
+Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
+To ensure that each generator is faced with the exact same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.
+
+\hypertarget{method-2-metrics}{%
+\subsection{Evaluation Metrics}\label{method-2-metrics}}
+
+We formulate two desiderata for the set of metrics used to measure domain and model shifts induced by recourse.
+First, the metrics should be applicable regardless of the dataset or classification technique so that they allow for the meaningful comparison of the generators in various scenarios.
+As the knowledge of the underlying probability distribution is rarely available, the metrics should be empirical and non-parametric.
+This further ensures that we can also measure large datasets by sampling from the available data.
+Moreover, while our study was conducted in a two-class classification setting, our choice of metrics should remain applicable in the future research on multi- class recourse problems.
+Second, the set of metrics should allow to capture various aspects of the previously mentioned magnitude, path, and tempo of changes while remaining as small as possible.
+
+\hypertarget{domain-shifts}{%
+\subsubsection{Domain Shifts}\label{domain-shifts}}
+
+To quantify the magnitude of domain shifts we rely on an unbiased estimate of the squared population \textbf{Maximum Mean Discrepancy (MMD)} given as:
+
+\begin{equation}
+\begin{aligned}
+MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) \label{eq:mmd}
+\end{aligned}
+\end{equation}
+
+where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert Space \(\mathcal{H}\) Berlinet and Thomas-Agnan (2011), and \(X\), \(\tilde{X}\) represent independent and identically distributed samples drawn from probability distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) respectively Gretton et al.
+(2012).
+MMD is a measure of the distance between the kernel mean embeddings of \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) in RKHS \(\mathcal{H}\).
+An important consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
+In our implementation we make use of Gaussian kernel with a constant length-scale parameter of \(0.5\).
+As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if \(X=\tilde{X}\).
+
+The evaluation metric in Equation \eqref{eq:mmd} is computed after every round \(t=1,...,T\) of the experiment.
+To assess the statistical significance of the observed shifts under the null hypothesis that samples \(X\) and \(\tilde{X}\) were drawn from the same probability distribution, we follow Arcones and Gine (1992).
+To that end, we combine the two samples and generate a large number of permutations of \(X + \tilde{X}\).
+Then, we split the permuted data into two new samples \(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the original samples.
+Then under the null hypothesis we should have that \(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to \(MMD_u^2[F,X,\tilde{X}]\).
+The corresponding \(p\)-value can then be calculated by counting how these two quantities are not equal.
+
+We calculate the MMD for both classes individually based on the ground truth labels, i.e.~the labels that samples were assigned in time \(t=0\).
+Throughout our experiments, we generally do not expect the distribution of the negative class to change over time -- application of recourse reduces the size of this class, but since individuals are sampled uniformly the distribution should remain unaffected.
+Conversely, unless a recourse generator can perfectly replicate the original probability distribution, we expect the MMD of the positive class to increase.
+Thus, when discussing MMD, we generally mean the shift in the distribution of the positive class.
+
+\hypertarget{model-shifts}{%
+\subsubsection{Model Shifts}\label{model-shifts}}
+
+As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following Upadhyay, Joshi, and Lakkaraju (2021).
+We define \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance between the vectors of parameters before and after retraining the model \(M\).
+We shall refer to this baseline metric simply as \textbf{Perturbations}.
+
+We extend the metric in Equation \eqref{eq:mmd} for the purpose of quantifying model shifts.
+Specifically, we introduce \textbf{Predicted Probability MMD (PP MMD)}: instead of applying Equation \eqref{eq:mmd} to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model \(M\).
+If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same.
+We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a meshgrid over a subspace of the entire feature space.
+For the latter approach we bound the subspace by the extrema of each feature.
+While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension \(D\) grows.
+
+As an alternative to PP MMD we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement).
+This metric was introduced in Hanneke (2007) and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
+In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model.
+A Disagreement Coefficient unequal to zero is indicative of a model shift.
+The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured.
+This is one reason for why PP MMD is our our preferred metric.
+
+We further introduce \textbf{Decisiveness} as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample.
+We define the metric simply as \({\frac{1}{N}}\sum_{i=0}^N(\sigma(M(x)) - 0.5)^2\) where \(M(x)\) are predicted logits from a binary classifier and \(\sigma\) denotes the sigmoid function.
+This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes.
+Although the exact values for this metric are not important for our study, they can be used to detect model shifts.
+If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.
+
+Finally, we also take a look at the out-of-sample \textbf{Performance} of our models. To this end, we compute their F-score on a test sample that we leave untouched throughout the experiment.
+
+\hypertarget{empirical}{%
+\section{Experiment Setup}\label{empirical}}
+
+This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\ref{empirical-2}).
+For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section.
+A few high-level details upfront: each experiment is run for a total of \(T=50\) rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in the previous section, we end up providing recourse to a total of \(\approx50\%\) by the end of round \(T=50\).}.
+All classifiers and generative models are retrained for 10 epochs in each round \(t\) of the experiment.
+Rather than retraining models from scratch, we initialize all parameters at their previous levels (\(t-1\)) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model.
+Evaluation metrics are computed and stored every 10 rounds.
+To account for noise each individual experiment is repeat five times.
+
+\hypertarget{empirical-classifiers}{%
+\subsection{\texorpdfstring{\(M\) -- Classifiers and Generative Models}{M -- Classifiers and Generative Models}}\label{empirical-classifiers}}
+
+For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} (Innes et al.
+2018): firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following Lakshminarayanan, Pritzel, and Blundell (2016) that serves as our only probabilistic classifier.
+We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
+They are also the model of choice in Schut et al.
+(2021).
+The actual neural network architectures are kept simple ((\textbf{tbl-mlp?})), since we are only marginally concerned with achieving good initial classifier performance.
+For the real-world datasets we using mini-batch training and dropout regularization.
+
+The Latent Space generators rely on separate generative models.
+Following the authors of both REVISE and CLUE we use Variational Autoencoders (\textbf{VAE}) for this purpose.
+As with the classifiers, we deliberately choose to work with fairly simple architectures ((\textbf{tbl-vae?})).
+More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
+But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.
+
+\hypertarget{empirical-data}{%
+\subsection{\texorpdfstring{\(\mathcal{D}\) -- Data}{\textbackslash mathcal\{D\} -- Data}}\label{empirical-data}}
+
+We have chosen to work with both synthetic and real-world datasets.
+Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
+Following Upadhyay, Joshi, and Lakkaraju (2021), we generate synthetic data in \(\mathbb{R}^2\) to also allow for a visual interpretation of the results.
+Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
+
+\hypertarget{synthetic-data}{%
+\subsubsection{Synthetic data}\label{synthetic-data}}
+
+We use four synthetic binary classification datasets consisting of 1000 samples each.
+The datasets are presented in Figure \ref{fig:synthetic-data} (see also Appendix A for a formal description).
+Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/synthetic_data}
+
+}
+
+\caption{Synthetic classification datasets used in our experiments.}\label{fig:synthetic-data}
+\end{figure}
+
+Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
+Thus, the choice of a black-box model may have an impact on the paths of the recourse.
+For generators that use latent space search (REVISE Joshi et al.
+(2019), CLUE Antorán et al.
+(2020)) or rely on (and have access to) probabilistic models (CLUE Antorán et al.
+(2020), Greedy Schut et al.
+(2021)) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
+Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class\footnote{As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated.
+We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.}.
+In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concered with generating what we have defined as meaningful counterfactuals.
+
+\hypertarget{real-world-data}{%
+\subsubsection{Real-world data}\label{real-world-data}}
+
+We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification.
+Firstly, we use the \textbf{Give Me Some Credit} dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years (Competition, n.d.).
+Originally consisting of 250,000 instances with 11 numerical attributes.
+Secondly, we use the \textbf{UCI defaultCredit} dataset (Yeh and Lien 2009), a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
+In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes\footnote{These have been ommitted from the analysis.
+See Section \ref{limit-data} for details.} and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
+Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example Pawelczyk et al.
+(2021), Joshi et al.
+(2019) and Ustun, Spangher, and Liu (2019)), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+
+As a third dataset we include the \textbf{California Housing} dataset derived from the 1990 U.S.
+census and sourced through scikit-learn Pace and Barry (1997).
+It consists of 8 continuous features that can be used to predict the median house price for California districts.
+The continuous outcome variable is binarized as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or not the median house price of a given district is above or below the median of all districts.
+While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example Schut et al.
+(2021)).
+While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem (M.
+2019).
+
+Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 10,000 individuals each.
+We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.
+
+\hypertarget{g-generators}{%
+\subsection{\texorpdfstring{\(G\) -- Generators}{G -- Generators}}\label{g-generators}}
+
+All generators introduced earlier are included in the experiments: Wachter (Wachter, Mittelstadt, and Russell 2017), REVISE (Joshi et al.
+2019), CLUE (Antorán et al.
+2020), DiCE (Mothilal, Sharma, and Tan 2020) and Greedy (Schut et al.
+2021).
+In addition, we introduce two new generators in Section \ref{mitigate} that directly address the issue of endogenous domain and model shifts.
+We also test to what extent it may be beneficial to combine ideas underlying the various generators.
+
+\hypertarget{empirical-2}{%
+\section{Experiments}\label{empirical-2}}
+
+Below we first present our main findings with respect to the presence of endogenous macrodynamics in Algorithmic Recourse.
+We then go on to introduce and test a number of potential mitigation strategies.
+
+\hypertarget{endogenous-macrodynamics}{%
+\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}
+
+In light of the huge number of experiments we have conducted, we shall start this section off with a few high-level observations.
+Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:
+
+\begin{itemize}
+\tightlist
+\item
+  Statistically significant domain and model shift as measured by MMD.
+\item
+  A deterioration in out-of-sample model performance as measured by the F-Score evaluated on a test sample. In many cases this drop in performance is substantial.
+\item
+  Significant perturbations to the model parameters as well as an increase in the model's decisiveness.
+\item
+  Disagreement between the original and retrained model, in some cases large.
+\end{itemize}
+
+There is also some clear heterogeneity across the results:
+
+\begin{itemize}
+\tightlist
+\item
+  The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible.
+\item
+  The reduction in model performance appears to be most severe when classes are not perfectly separable.
+\item
+  With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.
+\end{itemize}
+
+Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix.
+Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
+Figure \ref{fig:syn} presents a small subset of our results for the synthetic dataset with overlapping classes.
+It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all \(T=50\) rounds, along with error bars indicating the variation across folds.
+The top row shows the estimated domain shifts.
+While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds.
+With respect to the domain shifts, the Greedy generator actually induces the smallest shifts.
+In general, we have observed the opposite.
+The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier.
+As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
+In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
+The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/synthetic_results}
+
+}
+
+\caption{Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of points.
+Error bars indicate the standard deviation across folds.}\label{fig:syn}
+\end{figure}
+
+Turning to the real-world data next \ldots{}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/real_world_results}
+
+}
+
+\caption{Results for real-world datasets.}\label{fig:real}
+\end{figure}
+
+\hypertarget{mitigate}{%
+\section{Potential Mitigation Strategies}\label{mitigate}}
+
+In the following we explore several ideas for strategies that may help to mitigate those types of endogenous recourse dynamics we observed in the previous subsection.
+All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
+This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous subsection.
+But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational cost and - perhaps more importantly - their performance hinges on the performance of said generative models.
+Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.
+
+\hypertarget{more-conservative-decision-thresholds}{%
+\subsection{More Conservative Decision Thresholds}\label{more-conservative-decision-thresholds}}
+
+The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold \(\gamma\).
+Under \(\gamma=0.5\), counterfactuals will end up near the decision boundary by construction.
+Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off.
+By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
+A potential drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
+
+\hypertarget{classifier-preserving-roar}{%
+\subsection{Classifier Preserving ROAR}\label{classifier-preserving-roar}}
+
+Another potential strategy draws inspiration from ROAR Upadhyay, Joshi, and Lakkaraju (2021): to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
+Recall that \(g(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
+Then formally we let
+
+\begin{equation}
+\begin{aligned}
+g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) \label{eq:clap}
+\end{aligned}
+\end{equation}
+
+for each counterfactual \(k\) where \(l\) denotes the loss function used to train \(M\).
+This approach, which we shall refer to as \textbf{ClapROAR}, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
+It also makes the private vs.~external cost trade-off more explicit and hence manageable.
+
+\hypertarget{gravitational-counterfactual-explanations}{%
+\subsection{Gravitational Counterfactual Explanations}\label{gravitational-counterfactual-explanations}}
+
+Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: \(\bar{x}\):
+
+\begin{equation}
+\begin{aligned}
+g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  \label{eq:grav}
+\end{aligned}
+\end{equation}
+
+Once again we can putting this in the context of Equation \eqref{eq:collective}, the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
+Higher choices of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to gravitate towards the specified point \(\bar{x}\) in the target domain.
+In the remainder of this paper we will therefore refer to this approach as \textbf{Gravitational} generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics\footnote{Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.}.
+
+Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/mitigation}
+
+}
+
+\caption{Illustrative example demonstrating the properties of the various mitigation strategies.}\label{fig:mitigation}
+\end{figure}
+
+Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts.
+Figure \ref{fig:mitigate-results} presents a subset of the evaluation metrics for our synthetic data with overlapping classes.
+The top row in Figure \ref{fig:mitigate-results} indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
+ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
+As evident from the middle row in Figure \ref{fig:mitigate-results} model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
+The same overall pattern can be observed for out-of-sample model performance.
+With respect to the other synthetic datasets we found the following: for the Moons dataset the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/mitigation_synthetic_results}
+
+}
+
+\caption{The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$.
+Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of points.
+Error bars indicate the standard deviation across folds.}\label{fig:mitigate-results}
+\end{figure}
+
+An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
+In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+A snapshot of the results is shown in Figure \ref{fig:mitigate-latent-results}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/mitigation_synthetic_latent_results}
+
+}
+
+\caption{Combinining various mitigation strategies with Latent Space search.
+Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of points.
+Error bars indicate the standard deviation across folds.}\label{fig:mitigate-latent-results}
+\end{figure}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Our results in Section \ref{empirical-2} indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice.
+These induced shifts can and should be considered as an (expected) external cost of individual recourse.
+While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making.
+We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments.
+In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool.
+As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
+This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.
+
+So, where does this leave us?
+We would argue that the expected external costs of individual recourse should be shared by all stakeholders.
+The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \eqref{eq:collective}.
+This will on average lead to more costly counterfactual outcomes.
+But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
+We have shown various types of shift-mitigating strategies that can be used to this end.
+Since all of these strategies can be seen simply as specific adaption of Equation \eqref{eq:collective}, they can be applied to any of the various counterfactual generators studied here.
+
+\hypertarget{limit}{%
+\section{Limitations and Future Work}\label{limit}}
+
+While we believe that this work constitutes a valuable starting point for addressing existing issues in Algorithmic Recourse from a fresh perspective, we are aware of several of its limitations.
+In the following, we highlight some of these limitations and point to avenues for future research.
+
+\hypertarget{private-vs.-external-costs}{%
+\subsection{Private vs.~External Costs}\label{private-vs.-external-costs}}
+
+Perhaps the most crucial shortcoming of our work is that we merely point out that there exists a trade-off between private costs to the individual and external costs to the collective of stakeholders.
+We fall short of providing any definite answers as to how that trade-off may be resolved in practice.
+The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework.
+An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is a collective recourse outcome in which no single individual can be made better off, without make at least one other individual worse off.
+This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work.
+For the time being, we suggest that the trade-off can be resolved on a case-by-case basis.
+
+\hypertarget{experimental-setup}{%
+\subsection{Experimental Setup}\label{experimental-setup}}
+
+The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion.
+In practice, models are in fact updated on a regular basis (Upadhyay, Joshi, and Lakkaraju 2021).
+We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time \(t=0\).
+That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in Upadhyay, Joshi, and Lakkaraju (2021).
+We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model \(M\) is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
+Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary.
+In practice it is doubtful that agents behave according to such simple rules.
+Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).
+
+\hypertarget{limit-data}{%
+\subsection{Data}\label{limit-data}}
+
+Largely in line with the existing literature on Algorithmic Recourse, we have limited our analysis of real-world data to three commonly used benchmark datasets that involve binary prediction tasks.
+Future work may benefit from including novel datasets or extending the analysis to multi-class or regression problems, the latter arguably representing the most common objective in Finance and Economics.
+It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing \texttt{CounterfactualExplanations.jl} only supports continuous features, at least of some of the counterfactual generators considered here.
+The fact that we therefore had to discard discrete features led to relatively poor initial performance of our classifiers in some cases.
+While this is indeed a limitation we intend to address in future and derivative work, our findings with respect to endogenous macrodynamics do not hinge on strong classifier performance.
+
+\hypertarget{classifiers}{%
+\subsection{Classifiers}\label{classifiers}}
+
+For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks.
+While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here Grinsztajn, Oyallon, and Varoquaux (2022).
+
+\hypertarget{conclusion}{%
+\section{Concluding Remarks}\label{conclusion}}
+
+This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse.
+We demonstrate that long-held beliefs as to what defines optimality in AR, are too short-sighted to serve as a foundation for applications of recourse in practice.
+Specifically, we run multiple experiments that simulate the application of recourse in practice using various popular counterfactual generators and find that all of them induce substantial domain and model shifts.
+We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse.
+By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit.
+We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts.
+Through our work we hope to inspire future research on this important topic.
+To this end we have open-sourced all of our code along with a Julia package - \texttt{AlgorithmicRecourseDynamics.jl}.
+The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility (Altmeyer 2022).
+That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
+Ackerman, Samuel, Parijat Dube, Eitan Farchi, Orna Raz, and Marcel Zalmanovici.
+2021.
+{``Machine Learning Model Drift Detection via Weak Data Slices.''} In \emph{2021 IEEE/ACM Third International Workshop on Deep Learning for Testing and Testing for Deep Learning (DeepTest)}, 1--8.
+IEEE.
+
+\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
+Altmeyer, Patrick.
+2022.
+\emph{{CounterfactualExplanations.jl - a Julia package for Counterfactual Explanations and Algorithmic Recourse}} (version 0.1.2).
+\url{https://github.com/pat-alt/CounterfactualExplanations.jl}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
+Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José Miguel Hernández-Lobato.
+2020.
+{``Getting a Clue: A Method for Explaining Uncertainty Estimates.''} \emph{arXiv Preprint arXiv:2006.06848}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
+Arcones, Miguel A, and Evarist Gine. 1992. {``On the Bootstrap of u and v Statistics.''} \emph{The Annals of Statistics}, 655--74.
+
+\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
+Berlinet, Alain, and Christine Thomas-Agnan. 2011. \emph{Reproducing Kernel Hilbert Spaces in Probability and Statistics}. Springer Science \& Business Media.
+
+\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
+Borisov, Vadim, Tobias Leemann, Kathrin Seßler, Johannes Haug, Martin Pawelczyk, and Gjergji Kasneci.
+2021.
+{``Deep Neural Networks and Tabular Data: A Survey.''} \emph{arXiv Preprint arXiv:2110.01889}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
+Carrizosa, Emilio, Jasone Ramırez-Ayerbe, and Dolores Romero. 2021. {``Generating Collective Counterfactual Explanations in Score-Based Classification via Mathematical Optimization.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
+Chandola, Varun, Arindam Banerjee, and Vipin Kumar. 2009. {``Anomaly Detection: A Survey.''} \emph{ACM Computing Surveys (CSUR)} 41 (3): 1--58.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
+Competition, Kaggle.
+n.d.
+{``Give Me Some Credit, Improve on the State of the Art in Credit Scoring by Predicting the Probability That Somebody Will Experience Financial Distress in the Next Two Years.''} \url{https://www.kaggle.com/c/GiveMeSomeCredit}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
+Dombrowski, Ann-Kathrin, Jan E Gerken, and Pan Kessel.
+2021.
+{``Diffeomorphic Explanations with Normalizing Flows.''} In \emph{ICML Workshop on Invertible Neural Networks, Normalizing Flows, and Explicit Likelihood Models}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
+Gama, João, Indrė Žliobaitė, Albert Bifet, Mykola Pechenizkiy, and Abdelhamid Bouchachia. 2014. {``A Survey on Concept Drift Adaptation.''} \emph{ACM Computing Surveys (CSUR)} 46 (4): 1--37.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
+Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf, and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The Journal of Machine Learning Research} 13 (1): 723--73.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
+Grinsztajn, Léo, Edouard Oyallon, and Gaël Varoquaux. 2022. {``Why Do Tree-Based Models Still Outperform Deep Learning on Tabular Data?''} \emph{arXiv Preprint arXiv:2207.08815}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
+Hanneke, Steve. 2007. {``A Bound on the Label Complexity of Agnostic Active Learning.''} In \emph{Proceedings of the 24th International Conference on Machine Learning}, 353--60.
+
+\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
+Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and Viral Shah.
+2018.
+{``Fashionable Modelling with Flux.''} \emph{arXiv Preprint arXiv:1811.01457}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and Joydeep Ghosh.
+2019.
+{``Towards Realistic Individual Recourse and Actionable Explanations in Black-Box Decision Making Systems.''} \emph{arXiv Preprint arXiv:1907.09615}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
+Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel Valera.
+2020.
+{``A Survey of Algorithmic Recourse: Definitions, Formulations, Solutions, and Prospects.''} \emph{arXiv Preprint arXiv:2010.04050}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
+Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera.
+2021.
+{``Algorithmic Recourse: From Counterfactual Explanations to Interventions.''} In \emph{Proceedings of the 2021 ACM Conference on Fairness, Accountability, and Transparency}, 353--62.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
+Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2016. {``Simple and Scalable Predictive Uncertainty Estimation Using Deep Ensembles.''} \emph{arXiv Preprint arXiv:1612.01474}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
+M., Carlisle. 2019. {``Racist Data Destruction? - a Boston Housing Dataset Controversy.''} \url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
+Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan.
+2020.
+{``Explaining Machine Learning Classifiers Through Diverse Counterfactual Explanations.''} In \emph{Proceedings of the 2020 Conference on Fairness, Accountability, and Transparency}, 607--17.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
+Nelson, Kevin, George Corbin, Mark Anania, Matthew Kovacs, Jeremy Tobias, and Misty Blowers.
+2015.
+{``Evaluating Model Drift in Machine Learning Algorithms.''} In \emph{2015 IEEE Symposium on Computational Intelligence for Security and Defense Applications (CISDA)}, 1--8.
+IEEE.
+
+\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
+O'neil, Cathy. 2016. \emph{Weapons of Math Destruction: How Big Data Increases Inequality and Threatens Democracy}. Crown.
+
+\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
+Oliveira, Raphael Mazzine Barbosa de, and David Martens. 2021. {``A Framework and Benchmarking Study for Counterfactual Generating Methods on Tabular Data.''} \emph{Applied Sciences} 11 (16): 7274.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
+Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3): 291--97.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
+Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias Richter, and Gjergji Kasneci.
+2021.
+{``Carla: A Python Library to Benchmark Algorithmic Recourse and Counterfactual Explanation Algorithms.''} \emph{arXiv Preprint arXiv:2108.00783}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
+Pedregosa, Fabian, Gaël Varoquaux, Alexandre Gramfort, Vincent Michel, Bertrand Thirion, Olivier Grisel, Mathieu Blondel, et al.
+2011.
+{``Scikit-Learn: Machine Learning in Python.''} \emph{The Journal of Machine Learning Research} 12: 2825--30.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
+Pindyck, Robert S, and Daniel L Rubinfeld. 2014. \emph{Microeconomics}. Pearson Education.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
+Rabanser, Stephan, Stephan Günnemann, and Zachary Lipton.
+2019.
+{``Failing Loudly: An Empirical Study of Methods for Detecting Dataset Shift.''} \emph{Advances in Neural Information Processing Systems} 32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
+Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning Models for High Stakes Decisions and Use Interpretable Models Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
+Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan Sacaleanu, Yarin Gal, et al.
+2021.
+{``Generating Interpretable Counterfactual Explanations by Implicit Minimisation of Epistemic and Aleatoric Uncertainties.''} In \emph{International Conference on Artificial Intelligence and Statistics}, 1756--64.
+PMLR.
+
+\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
+Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021. {``Towards Robust and Reliable Algorithmic Recourse.''} \emph{arXiv Preprint arXiv:2102.13620}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
+Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable Recourse in Linear Classification.''} In \emph{Proceedings of the Conference on Fairness, Accountability, and Transparency}, 10--19.
+
+\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
+Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual Explanations for Machine Learning: A Review.''} \emph{arXiv Preprint arXiv:2010.10596}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
+Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017. {``Counterfactual Explanations Without Opening the Black Box: Automated Decisions and the GDPR.''} \emph{Harv. JL \& Tech.} 31: 841.
+
+\leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
+Widmer, Gerhard, and Miroslav Kubat. 1996. {``Learning in the Presence of Concept Drift and Hidden Contexts.''} \emph{Machine Learning} 23 (1): 69--101.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
+Yeh, I-Cheng, and Che-hui Lien.
+2009.
+{``The Comparisons of Data Mining Techniques for the Predictive Accuracy of Probability of Default of Credit Card Clients.''} \emph{Expert Systems with Applications} 36 (2): 2473--80.
+
 \end{CSLReferences}
 
-\pagebreak
-
-\hypertarget{appendix-appendix}{%
-\section{(APPENDIX) Appendix}\label{appendix-appendix}}
-
-\hypertarget{tables}{%
-\section{Tables}\label{tables}}
-
-\ldots{}
-
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
-
-\ldots{}
-
-\hypertarget{code}{%
-\section{Code}\label{code}}
-
-\ldots{}
+\{\{\textless{} include sections/appendix.qmd \textgreater\}\}
 

### 2022-08-15 paper/sections/discussion.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/discussion.qmd
@@
-# Discussion {#sec-discussion}
+# Discussion {#discussion}
 
-Our results in @sec-empirical-2 indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice.
+Our results in Section \@ref(empirical-2) indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice.
 These induced shifts can and should be considered as an (expected) external cost of individual recourse.
@@
 We would argue that the expected external costs of individual recourse should be shared by all stakeholders.
-The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in @eq-collective.
+The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective).
 This will on average lead to more costly counterfactual outcomes.
@@
 We have shown various types of shift-mitigating strategies that can be used to this end.
-Since all of these strategies can be seen simply as specific adaption of @eq-collective, they can be applied to any of the various counterfactual generators studied here.
+Since all of these strategies can be seen simply as specific adaption of Equation \@ref(eq:collective), they can be applied to any of the various counterfactual generators studied here.

### 2022-08-15 paper/sections/empirical.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/empirical.qmd
@@
-# Experiment Setup {#sec-empirical}
+# Experiment Setup {#empirical}
 
-This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (@sec-empirical-2).
+This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)).
 For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section.
@@
 Evaluation metrics are computed and stored every 10 rounds.
+To account for noise each individual experiment is repeat five times.
 
-## $M$ -- Classifiers and Generative Models {#sec-empirical-classifiers}
+## $M$ -- Classifiers and Generative Models {#empirical-classifiers}
 
@@
 
-::: {#tbl-panel layout-ncol=1}
+<!-- | | Hidden Dim. | Hidden Layers | Batch | Dropout | -->
+<!-- |------|------|------|-----|-----| -->
+<!-- | Synthetic | 32 | 1 | - | - | -->
+<!-- | Real-World | 32 | 2 | 50 | 0.25 | -->
 
-| | Hidden Dim. | Hidden Layers | Batch | Dropout |
-|------|------|------|-----|-----|
-| Synthetic | 32 | 1 | - | - |
-| Real-World | 32 | 2 | 50 | 0.25 |
+<!-- : MLP {#tbl-mlp}  -->
 
-: MLP {#tbl-mlp}
+<!-- |  | Hidden Dim. | Epochs | -->
+<!-- |------|------|------| -->
+<!-- | Synthetic | 2 | 100 |  -->
+<!-- | Real-World | 8 | 250 |  -->
 
-|  | Hidden Dim. | Epochs |
-|------|------|------|
-| Synthetic | 2 | 100 | 
-| Real-World | 8 | 250 |
+<!-- : Variational Autoencoder {#tbl-vae} -->
 
-: Variational Autoencoder {#tbl-vae}
-
-Model Architectures
-:::
-
-## $\mathcal{D}$ -- Data {#sec-empirical-data}
+## $\mathcal{D}$ -- Data {#empirical-data}
 
@@
 
-```{julia}
+```{julia, eval=FALSE, echo=FALSE}
 using Plots, PlotThemes
@@
 We use four synthetic binary classification datasets consisting of 1000 samples each.
-The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description).
+The datasets are presented in Figure \@ref(fig:synthetic-data) (see also Appendix A for a formal description).
 Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
 
-![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}
+```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments."}
+knitr::include_graphics("www/synthetic_data.png")
+```
 
@@
 In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis.
-See @sec-limit-data for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
+See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
 Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
@@
 All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating].
-In addition, we introduce two new generators in @sec-empirical-2-mitigate that directly address the issue of endogenous domain and model shifts.
+In addition, we introduce two new generators in Section \@ref(mitigate) that directly address the issue of endogenous domain and model shifts.
 We also test to what extent it may be beneficial to combine ideas underlying the various generators.

### 2022-08-15 paper/sections/empirical_2.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-15 dev/paper/sections/empirical_2.qmd
@@
-# Experiments {#sec-empirical-2}
+# Experiments {#empirical-2}
 
@@
 Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
-@fig-syn presents a small subset of our results for the synthetic dataset with overlapping classes.
+Figure \@ref(fig:syn) presents a small subset of our results for the synthetic dataset with overlapping classes.
 It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds.
@@
 
-![Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of points.
-Error bars indicate the standard deviation across folds.](www/synthetic_results.png){#fig-syn fig.pos="h" width="9cm" height="9cm"}
+```{r syn, fig.cap="Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds."}
+knitr::include_graphics("www/synthetic_results.png")
+```
 
@@
 
-![PLACEHOLDER: Results for real-world as table.](www/placeholder.png){#fig-real fig.pos="h" width="9cm" height="9cm"}
+```{r real, fig.cap="Results for real-world datasets."}
+knitr::include_graphics("www/real_world_results.png")
+```

### 2022-08-15 paper/sections/introduction.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/introduction.qmd
@@
-# Introduction {#sec-intro}
+# Introduction {#intro}
 
@@
 
-@fig-poc illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 
-![Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}
+```{r poc, fig.cap="Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d)."}
+knitr::include_graphics("www/poc.png")
+```
 
@@
 Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
-Secondly and relatedly, we observe in @fig-poc that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
+Secondly and relatedly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
 Average default risk across all borrowers can therefore be expected to increase.
@@
 
-The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature.
-@sec-method posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
-@sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR.
-@sec-empirical presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
-@sec-empirical-2 presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
-Our findings are then discussed in the broader context of the literature in @sec-discussion, before pointing to some of the limitations of our work as well as avenues for future research in @sec-limit.
-Finally, @sec-conclusion concludes.
+The remainder of the paper is structured as follows: Section \@ref(related) places our work in the broader context of related literature.
+Section \@ref(method) posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
+Section \@ref(method-2) sets out our experimental framework for modeling endogenous macrodynamics in AR.
+Section \@ref(empirical) presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
+Section \@ref(empirical-2)  presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
+Our findings are then discussed in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit).
+Finally, Section \@ref(conclusion) concludes.

### 2022-08-15 paper/sections/methodology.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/methodology.qmd
@@
-# Gradient-Based Recourse Revisited {#sec-method}
+# Gradient-Based Recourse Revisited {#method}
 
-In this section we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general that encapsulates the various individual recourse methods we have chosen to use in our experiments.
+In this section we first set out a generalized framework for gradient-based counterfactual search in Section \@ref(method-general) that encapsulates the various individual recourse methods we have chosen to use in our experiments.
 We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.
 
-## From individual recourse ... {#sec-method-general}
+## From individual recourse ... {#method-general}
 
@@
 Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
-We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro.
+We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro).
 This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
@@
 
-As we will see next, all of them can be described by the following generalized form of @eq-solution:
+As we will see next, all of them can be described by the following generalized form of Equation \@ref(eq:general):
 
-$$
+\begin{equation}
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} (\#eq:general)
 \end{aligned} 
-$$ {#eq-general}
+\end{equation}
 
@@
 The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
-In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
+In this case $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
 Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
@@
 The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
-The authors show that in this setting the cost function $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+The authors show that in this setting the cost function $h(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
 Schut et al.
@@
 The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach.
-It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function @eq-general.
+It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).
 
@@
 All of the different approaches introduced above tackle the problem of Algorithmic Recourse from the perspective of one single individual^[DiCE recognizes that different individuals may have different objective functions, but it does not address the interdependencies between different individuals.].
-To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:
+To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend Equation \@ref(eq:general) as follows:
 
-$$
+\begin{equation}
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  (\#eq:collective)
 \end{aligned} 
-$$ {#eq-collective}
+\end{equation}
 
@@
 As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
-That being said, we do propose a few potential mitigation strategies in @sec-empirical-2-mitigate.
+That being said, we do propose a few potential mitigation strategies in Section \@ref(mitigate).

### 2022-08-15 paper/sections/methodology_2.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/methodology_2.qmd
@@
-# Modeling Endogenous Macrodynamics in Algorithmic Recourse {#sec-method-2}
+# Modeling Endogenous Macrodynamics in Algorithmic Recourse {#method-2}
 
@@
 
-## Simulations {#sec-method-2-experiment}
+## Simulations {#method-2-experiment}
 
-The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
+The dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro) were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
 We begin in the static setting at time time $t=0$: given some classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$ we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
@@
 
-## Evaluation Metrics {#sec-method-2-metrics}
+## Evaluation Metrics {#method-2-metrics}
 
@@
 
-$$
+\begin{equation}
 \begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) (\#eq:mmd)
 \end{aligned}
-$$ {#eq-mmd}
+\end{equation}
 
@@
 
-The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment.
+The evaluation metric in Equation \@ref(eq:mmd) is computed after every round $t=1,...,T$ of the experiment.
 To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap.
@@
 
-We extend the metric in @eq-mmd for the purpose of quantifying model shifts.
-Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying @eq-mmd to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$.
+We extend the metric in Equation \@ref(eq:mmd) for the purpose of quantifying model shifts.
+Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying Equation \@ref(eq:mmd) to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$.
 If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same.
@@
 We further introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample.
-We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function.
+We define the metric simply as ${\frac{1}{N}}\sum_{i=0}^N(\sigma(M(x)) - 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function.
 This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes.

### 2022-08-15 paper/sections/mitigation.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-15 dev/paper/sections/mitigation.qmd
@@
-# Potential Mitigation Strategies {#sec-empirical-2-mitigate}
+# Potential Mitigation Strategies {#mitigate}
 
@@
 Another potential strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
-Recall that $g(\cdot)$ denotes what we had defined as the external cost in @eq-collective.
+Recall that $g(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
 Then formally we let
 
-$$
+\begin{equation}
 \begin{aligned}
-g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
+g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) (\#eq:clap)
 \end{aligned}
-$$ {#eq-clap}
+\end{equation}
 
@@
 
-$$
+\begin{equation}
 \begin{aligned}
-g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
+g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  (\#eq:grav)
 \end{aligned}
-$$ {#eq-clap}
+\end{equation}
 
-Once again we can putting this in the context of @eq-collective, the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
+Once again we can putting this in the context of Equation \@ref(eq:collective), the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
 Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain.
@@
 
-@fig-mitigation hows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
+Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
 
-```{julia}
-#| eval: false
+```{julia, eval=FALSE, echo=FALSE}
 using Random
@@
 
-![Illustrative example demonstrating the properties of the various mitigation strategies.](www/mitigation.png){#fig-mitigation fig.pos="h" width=45%}
+```{r mitigation, fig.cap="Illustrative example demonstrating the properties of the various mitigation strategies."}
+knitr::include_graphics("www/mitigation.png")
+```
 
 Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts.
-@fig-mitigate-results presents a subset of the evaluation metrics for our synthetic data with overlapping classes.
-The top row in @fig-mitigate-results indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
+Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes.
+The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
 ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
-As evident from the middle row in @fig-mitigate-results model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
+As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
 The same overall pattern can be observed for out-of-sample model performance.
@@
 
-![The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$.
+```{r mitigate-results, fig.cap="The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\\gamma=0.5$.
 Results for synthetic data with overlapping classes.
 The shown model MMD (PP MMD) was computed over a meshgrid of points.
-Error bars indicate the standard deviation across folds.](www/mitigation_synthetic_results.png){#fig-mitigate-results fig.pos="h" width=45%}
+Error bars indicate the standard deviation across folds."}
+knitr::include_graphics("www/mitigation_synthetic_results.png")
+```
 
@@
 In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
-A snapshot of the results is shown in @fig-mitigate-latent-results.
+A snapshot of the results is shown in Figure \@ref(fig:mitigate-latent-results).
 
-![Combinining various mitigation strategies with Latent Space search.
+```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with Latent Space search.
 Results for synthetic data with overlapping classes.
 The shown model MMD (PP MMD) was computed over a meshgrid of points.
-Error bars indicate the standard deviation across folds.](www/mitigation_synthetic_latent_results.png){#fig-mitigate-latent-results fig.pos="h" width="9cm" height="9cm"}
+Error bars indicate the standard deviation across folds."}
+knitr::include_graphics("www/mitigation_synthetic_latent_results.png")
+```

### 2022-08-15 paper/sections/related.rmd
moved everything to R Mardkown and now template works out well with rticles

--- 2022-08-12 dev/paper/sections/related.qmd
@@
-# Background {#sec-related}
+# Background {#related}
 
 In this Section we provide a review of the relevant literature.
-First, @sec-related-recourse discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
-Then, @sec-related-shifts presents some of the previous work on the measurement of dataset and model shifts.
+First, Subsection \@ref(related-recourse) discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
+Then, Subsection \@ref(related-shifts) presents some of the previous work on the measurement of dataset and model shifts.
 
-## Algorithmic Recourse {#sec-related-recourse}
+## Algorithmic Recourse {#related-recourse}
 
@@
 
-$$
-\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
-$$ {#eq-obj}
+\begin{equation}
+\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
+\end{equation}
 
-For implementation purposes, @eq-obj is typically approximated through regularization:
+For implementation purposes, \@ref(eq:obj) is typically approximated through regularization:
 
-$$
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
-$$ {#eq-solution}
+\begin{equation}
+x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) (\#eq:solution)
+\end{equation}
 
-In the baseline work [@wachter2017counterfactual], the cost function $h: \mathcal{X} \mapsto \mathbb{R}$ is proxied by some distance metric based on the simple intuition that perturbations of $x$ are costly to the individual.
-For models that are differentiable and produce smooth predictions, @eq-solution can be solved through gradient descent.
+In the baseline work [@wachter2017counterfactual], the cost function  is proxied by some distance metric based on the simple intuition that perturbations of $x$ are costly to the individual.
+For models that are differentiable and produce smooth predictions, \@ref(eq:solution) can be solved through gradient descent.
 This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter** - the name of the first author - in the remainder of this paper.
@@
 
-## Domain and Model Shifts {#sec-related-shifts}
+## Domain and Model Shifts {#related-shifts}
 
@@
 
-## Benchmarking Counterfactual Generators {#sec-related-benchmark}
+## Benchmarking Counterfactual Generators {#related-benchmark}
 
@@
 In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: `AlgorithmicRecourseDynamics.jl`^[The package is available from ...].
-The core concepts implemented in that package reflect what is presented in section @sec-method-2 of this paper.
+The core concepts implemented in that package reflect what is presented in Section \@ref(method-2) of this paper.

### 2022-08-15 paper/paper.tex
some minor things

@@
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
-In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is available from \ldots{}}.
+In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \href{(https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)}{\texttt{AlgorithmicRecourseDynamics.jl}}\footnote{The code is available from the following anonymized Github repository: \url{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}.}.
 The core concepts implemented in that package reflect what is presented in Section \ref{method-2} of this paper.
@@
 In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
-We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we introduced earlier.
+We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}} - the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
@@
 Through our work we hope to inspire future research on this important topic.
-To this end we have open-sourced all of our code along with a Julia package - \texttt{AlgorithmicRecourseDynamics.jl}.
+To this end we have open-sourced all of our code along with a Julia package - \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}.
 The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility (Altmeyer 2022).

### 2022-08-15 paper/sections/methodology_2.rmd
some minor things

@@
 In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
-We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of `AlgorithmicRecourseDynamics.jl` - the Julia package we introduced earlier.
+We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) - the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.

### 2022-08-15 paper/sections/related.rmd
some minor things

@@
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
-In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: `AlgorithmicRecourseDynamics.jl`^[The package is available from ...].
+In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: [`AlgorithmicRecourseDynamics.jl`]((https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md))^[The code is available from the following anonymized Github repository: [https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).].
 The core concepts implemented in that package reflect what is presented in Section \@ref(method-2) of this paper.

### 2022-08-16 paper/paper.tex
progress on results, formatting and comments

@@
 %
+\usepackage{cite}
 %\usepackage{cite}
@@
 We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
-Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice.
+Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate a total of XX million counterfactuals and study the resulting domain and model shifts.
+We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources.
 Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
@@
 They are essentially \textbf{black boxes}: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools.
-Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes (O'neil 2016).
-The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models (Rudin 2019), others have investigated different ways to explain the behaviour of black-box models.
+Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}.
+The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have investigated different ways to explain the behaviour of black-box models.
 These two sub-domains can be broadly referred to as \textbf{interpretable AI} and \textbf{explainable AI} (XAI), respectively.
@@
 Existing work in this field has largely worked in a static setting: various approaches have been proposed to generate counterfactuals for a given individual that is subject to some pre-trained model.
-More recent work has compared different approaches within this static setting (Pawelczyk et al.
-2021).
+More recent work has compared different approaches within this static setting \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}.
 In this work, we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly?
@@
 
-Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by Wachter, Mittelstadt, and Russell (2017): the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups.
-Secondly, we introduce an experimental framework extending previous work by Altmeyer (2022), which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including Wachter, Mittelstadt, and Russell (2017), Schut et al.
-(2021), Joshi et al.
-(2019), Mothilal, Sharma, and Tan (2020) and Antorán et al.
-(2020).
+Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}.
 To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators.
@@
 
-A framework for Counterfactual Explanations was first proposed in 2017 by Wachter, Mittelstadt, and Russell (2017) and has served as the baseline for most methodologies that have been proposed since then.
+A framework for Counterfactual Explanations was first proposed in 2017 by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then.
 Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\).
-Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):
+Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):
 
 \begin{equation}
-\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
+\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
 \end{equation}
@@
 \begin{equation}
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) \label{eq:solution}
+x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) \label{eq:solution}
 \end{equation}
 
-In the baseline work (Wachter, Mittelstadt, and Russell 2017), the cost function is proxied by some distance metric based on the simple intuition that perturbations of \(x\) are costly to the individual.
+In the baseline work \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, the cost function is proxied by some distance metric based on the simple intuition that perturbations of \(x\) are costly to the individual.
 For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent.
-This summarizes the approach followed in Wachter, Mittelstadt, and Russell (2017) which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.
+This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.
 
 Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017.
-An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 (Karimi et al.
-2020).
-Another survey published around the same time by Verma et al.~described 29 algorithms (Verma, Dickerson, and Hines 2020).
+An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 \protect\hyperlink{ref-karimi2020survey}{{[}10{]}}.
+Another survey published around the same time by Verma et al.~described 29 algorithms \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}.
 Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met.
 Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs.
-To name but a few, we are interested in generating counterfactuals that close (Wachter, Mittelstadt, and Russell 2017), actionable (Ustun, Spangher, and Liu 2019), realistic Schut et al.
-(2021), sparse, diverse (Mothilal, Sharma, and Tan 2020) and if possible causally founded (Karimi, Schölkopf, and Valera 2021).
+To name but a few, we are interested in generating counterfactuals that close \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible causally founded \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
 
 Efforts so far have largely been directed at improving the quality of Counterfactual Explanations within a static context: given some pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are interested in generating one or multiple meaningful Counterfactual Explanations for some individual characterized by \(x\).
-The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point (Verma, Dickerson, and Hines 2020).
-We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR (Upadhyay, Joshi, and Lakkaraju 2021).
+The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}.
+We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
 Exogenous shifts are strictly of external origin.
-For example, they might stem from data correction, temporal shifts or geospatial changes (Upadhyay, Joshi, and Lakkaraju 2021).
-The authors of (Upadhyay, Joshi, and Lakkaraju 2021) propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
+For example, they might stem from data correction, temporal shifts or geospatial changes \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+The authors of \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
 
 As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances.
-We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals (Carrizosa, Ramırez-Ayerbe, and Romero 2021).
+We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals \protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}.
 The authors propose an optimization framework that generates collective counterfactuals.
@@
 Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time.
-Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests (Rabanser, Günnemann, and Lipton 2019).
+Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests \protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}.
 This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution.
-Numerous approaches to anomaly detection have been summarized (Chandola, Banerjee, and Kumar 2009).
-Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model (Widmer and Kubat 1996).
-For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift (Gama et al.
-2014).
+Numerous approaches to anomaly detection have been summarized \protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}.
+Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}.
+For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
 Less previous work is available on the related topic of model drift - changes in model performance over time.
-Nelson et al.~review how resistant different machine learning models are to the model drift (Nelson et al.
-2015).
-Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available (Ackerman et al.
-2021).
-
-In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR (Upadhyay, Joshi, and Lakkaraju 2021).
+Nelson et al.~review how resistant different machine learning models are to the model drift \protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}.
+Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available \protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.
+
+In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
 In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\).
-While this also sets the baseline for our analysis here, it is worth noting that in Upadhyay, Joshi, and Lakkaraju (2021) these perturbations are mechanically introduced.
+While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these perturbations are mechanically introduced.
 In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, Oliveira and Martens (2021) run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, Pawelczyk et al.
-(2021) introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} (Altmeyer 2022) provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
@@
 We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
-In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} (Joshi et al.
-2019), \textbf{CLUE} (Antorán et al.
-2020), \textbf{DiCE} (Mothilal, Sharma, and Tan 2020) and a greedy approach that relies on probabilistic models (Schut et al.
-2021).
+In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
 Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} \label{eq:general}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} \label{eq:general}
 \end{aligned} 
@@
 In this case \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP (Joshi et al.
-2019).
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.
 
@@
 To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
-The approach proposed by Schut et al.
-(2021), which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
-The authors show that in this setting the cost function \(h(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+The approach proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
+The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
 Schut et al.
-(Schut et al.
-2021) also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE.
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE.
 In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
-They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski, Gerken, and Kessel (2021) relies on normalizing flows.}.
-
-Finally, DiCE (Mothilal, Sharma, and Tan 2020) distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals.
+They largely differ with respect to the chosen generative model: for example, the generator proposed by \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.
+
+Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals.
 Wachter et al.
-(Wachter, Mittelstadt, and Russell 2017) show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
-In Mothilal, Sharma, and Tan (2020) diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ...
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
+In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ...
 , s_K\) that look as different from each other as possible.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  \label{eq:collective}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  \label{eq:collective}
 \end{aligned} 
@@
 
-Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized.
-The newly introduced term \(g(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\).
-The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient (Pindyck and Rubinfeld 2014).
+Here \(\text{cost}(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized.
+The newly introduced term \(\text{extcost}(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\).
+The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}.
 We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
-As with individual recourse, the exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
+As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
 That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.
@@
 \begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) \label{eq:mmd}
+MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) \label{eq:mmd}
 \end{aligned}
@@
 
-where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert Space \(\mathcal{H}\) Berlinet and Thomas-Agnan (2011), and \(X\), \(\tilde{X}\) represent independent and identically distributed samples drawn from probability distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) respectively Gretton et al.
-(2012).
-MMD is a measure of the distance between the kernel mean embeddings of \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) in RKHS \(\mathcal{H}\).
+where \(X=\{x_1,...,x_m\}\), \(\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}\) represent independent and identically distributed samples drawn from probability distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) respectively \protect\hyperlink{ref-gretton2012kernel}{{[}25{]}}.
+MMD is a measure of the distance between the kernel mean embeddings of \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) in a Reproducing Kernel Hilbert Space, \(\mathcal{H}\) \protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}.
 An important consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
 In our implementation we make use of Gaussian kernel with a constant length-scale parameter of \(0.5\).
-As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if \(X=\tilde{X}\).
+As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD(X,\tilde{X})=0\) if and only if \(X=\tilde{X}\).
 
 The evaluation metric in Equation \eqref{eq:mmd} is computed after every round \(t=1,...,T\) of the experiment.
-To assess the statistical significance of the observed shifts under the null hypothesis that samples \(X\) and \(\tilde{X}\) were drawn from the same probability distribution, we follow Arcones and Gine (1992).
+To assess the statistical significance of the observed shifts under the null hypothesis that samples \(X\) and \(\tilde{X}\) were drawn from the same probability distribution, we follow \protect\hyperlink{ref-arcones1992bootstrap}{{[}27{]}}.
 To that end, we combine the two samples and generate a large number of permutations of \(X + \tilde{X}\).
 Then, we split the permuted data into two new samples \(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the original samples.
-Then under the null hypothesis we should have that \(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to \(MMD_u^2[F,X,\tilde{X}]\).
+Then under the null hypothesis we should have that \(MMD(X^\prime,\tilde{X}^\prime)\) be approximately equal to \(MMD(X,\tilde{X})\).
 The corresponding \(p\)-value can then be calculated by counting how these two quantities are not equal.
@@
 
-As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following Upadhyay, Joshi, and Lakkaraju (2021).
+As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
 We define \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance between the vectors of parameters before and after retraining the model \(M\).
@@
 As an alternative to PP MMD we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement).
-This metric was introduced in Hanneke (2007) and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
@@
 If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.
+A potential caveat of this metric in the context of our experiments is that it will to some degree get inflated simply through retraining the model.
 
@@
 
-For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} (Innes et al.
-2018): firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following Lakshminarayanan, Pritzel, and Blundell (2016) that serves as our only probabilistic classifier.
+For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier.
 We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
-They are also the model of choice in Schut et al.
-(2021).
-The actual neural network architectures are kept simple ((\textbf{tbl-mlp?})), since we are only marginally concerned with achieving good initial classifier performance.
+They are also the model of choice in \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
+The actual neural network architectures are kept simple (Table \ref{tab:mlp}), since we are only marginally concerned with achieving good initial classifier performance.
 For the real-world datasets we using mini-batch training and dropout regularization.
 
-The Latent Space generators rely on separate generative models.
+The Latent Space generator relies on a separate generative model.
 Following the authors of both REVISE and CLUE we use Variational Autoencoders (\textbf{VAE}) for this purpose.
-As with the classifiers, we deliberately choose to work with fairly simple architectures ((\textbf{tbl-vae?})).
+As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \ref{tab:vae}).
 More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
@@
 
+\begin{table}
+
+\caption{\label{tab:mlp}MLP architecture and training parameters.}
+\centering
+\begin{tabular}[t]{lrrll}
+\toprule
+  & Hidden Dim. & Hidden Layers & Batch & Dropout\\
+\midrule
+Synthetic & 32 & 1 & - & -\\
+Real-World & 32 & 2 & 50 & 0.25\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:vae}VAE architecture and training parameters.}
+\centering
+\begin{tabular}[t]{lrr}
+\toprule
+  & Hidden Dim. & Epochs\\
+\midrule
+Synthetic & 2 & 100\\
+Real-World & 8 & 250\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \hypertarget{empirical-data}{%
@@
 Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
-Following Upadhyay, Joshi, and Lakkaraju (2021), we generate synthetic data in \(\mathbb{R}^2\) to also allow for a visual interpretation of the results.
+Following \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}, we generate synthetic data in \(\mathbb{R}^2\) to also allow for a visual interpretation of the results.
 Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
@@
 Thus, the choice of a black-box model may have an impact on the paths of the recourse.
-For generators that use latent space search (REVISE Joshi et al.
-(2019), CLUE Antorán et al.
-(2020)) or rely on (and have access to) probabilistic models (CLUE Antorán et al.
-(2020), Greedy Schut et al.
-(2021)) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+For generators that use latent space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}) or rely on (and have access to) probabilistic models (CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
 Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
@@
 We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification.
-Firstly, we use the \textbf{Give Me Some Credit} dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years (Competition, n.d.).
+Firstly, we use the \textbf{Give Me Some Credit} dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years \protect\hyperlink{ref-gmsc_data}{{[}31{]}}.
 Originally consisting of 250,000 instances with 11 numerical attributes.
-Secondly, we use the \textbf{UCI defaultCredit} dataset (Yeh and Lien 2009), a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
+Secondly, we use the \textbf{UCI defaultCredit} dataset \protect\hyperlink{ref-yeh2009comparisons}{{[}32{]}}, a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
 In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes\footnote{These have been ommitted from the analysis.
 See Section \ref{limit-data} for details.} and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
-Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example Pawelczyk et al.
-(2021), Joshi et al.
-(2019) and Ustun, Spangher, and Liu (2019)), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}} and \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
 
 As a third dataset we include the \textbf{California Housing} dataset derived from the 1990 U.S.
-census and sourced through scikit-learn Pace and Barry (1997).
+census and sourced through scikit-learn \protect\hyperlink{ref-pace1997sparse}{{[}34{]}}.
 It consists of 8 continuous features that can be used to predict the median house price for California districts.
 The continuous outcome variable is binarized as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or not the median house price of a given district is above or below the median of all districts.
-While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example Schut et al.
-(2021)).
-While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem (M.
-2019).
+While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example \protect\hyperlink{ref-schut2021generating}{{[}6{]}}).
+While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}35{]}}.
 
@@
 
-All generators introduced earlier are included in the experiments: Wachter (Wachter, Mittelstadt, and Russell 2017), REVISE (Joshi et al.
-2019), CLUE (Antorán et al.
-2020), DiCE (Mothilal, Sharma, and Tan 2020) and Greedy (Schut et al.
-2021).
+All generators introduced earlier are included in the experiments: Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
 In addition, we introduce two new generators in Section \ref{mitigate} that directly address the issue of endogenous domain and model shifts.
@@
 
-Below we first present our main findings with respect to the presence of endogenous macrodynamics in Algorithmic Recourse.
-We then go on to introduce and test a number of potential mitigation strategies.
+Using the framework introduced in the previous section we are interested in answering the following research questions in this section:
+
+\begin{itemize}
+\tightlist
+\item
+  Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
+\item
+  If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+\item
+  Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
+\item
+  What seem to be drivers of endogenous dynamics in Algorithmic Recourse?
+\end{itemize}
+
+Below we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.
 
@@
 
-In light of the huge number of experiments we have conducted, we shall start this section off with a few high-level observations.
+We start this section off with a few high-level observations.
 Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:
@@
 \item
-  The reduction in model performance appears to be most severe when classes are not perfectly separable.
+  The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial performance of the classifier was weak to begin with.
 \item
-  With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.
+  With the exception of the Greedy generator, all other generators generally perform somewhat better overall than the baseline (Wachter) as expected.
 \end{itemize}
 
-Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix.
-Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
-Figure \ref{fig:syn} presents a small subset of our results for the synthetic dataset with overlapping classes.
+Focusing first on synthetic data, Figure \ref{fig:syn} presents our findings for the dataset with overlapping classes.
 It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all \(T=50\) rounds, along with error bars indicating the variation across folds.
+
 The top row shows the estimated domain shifts.
@@
 In general, we have observed the opposite.
-The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier.
+
+The second row shows the estimated model shifts, where here we have used the grid approach explained earlier.
 As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
 In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
-The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.
+
+The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
+Related to this the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
+We also note that the final classifiers are more decisive, although as we noted earlier this may to some extent just be a byproduct of retraining the model throughout the course of the experiment.
+
+Figure \ref{fig:syn} also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently.
+Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP.
+It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.
 
@@
 \caption{Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of points.
+The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points.
 Error bars indicate the standard deviation across folds.}\label{fig:syn}
@@
 
-Turning to the real-world data next \ldots{}
+Turning to the real-world data we will go through the findings presented in Figure \ref{fig:real} where each column corresponds to one of the three data sets.
+The results shown here are for the deep ensemble, which once again largely resemble those for the MLP.
+Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts.
+While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
+
+The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large.
+Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
+Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.
+
+Out-of-sample model performance also deteriorates across the board and substantially so: the smallest average reduction in F-Scores of around 15-20 percentage points is observed for the California Housing dataset.
+For this dataset we achieved the highest initial model performance of just under 90 percent, indicating once again that weaker classifiers may be more exposed to endogenous dynamics.
+As with the synthetic data, the estimates for logistic regression are qualitatively in line with the above, but quantitatively even more pronounced.
+
+Zooming in on individual features we \ldots{}
 
@@
 
-\caption{Results for real-world datasets.}\label{fig:real}
+\caption{Results for deep ensemble using real-world datasets.
+The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points.
+Error bars indicate the standard deviation across folds.}\label{fig:real}
 \end{figure}
@@
 
-In the following we explore several ideas for strategies that may help to mitigate those types of endogenous recourse dynamics we observed in the previous subsection.
+Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrent our attention, in this section we introduce and test a number of potential mitigation strategies.
 All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
-This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous subsection.
-But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational cost and - perhaps more importantly - their performance hinges on the performance of said generative models.
+This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section.
+But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and - perhaps more importantly - their performance hinges on the performance of said generative models.
 Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.
@@
 The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold \(\gamma\).
+This threshold determines when a proposed counterfactual should be considered as valid.
 Under \(\gamma=0.5\), counterfactuals will end up near the decision boundary by construction.
@@
 
-Another potential strategy draws inspiration from ROAR Upadhyay, Joshi, and Lakkaraju (2021): to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
-Recall that \(g(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
+Another potential strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
+Recall that \(\text{extcost}(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
 Then formally we let
@@
 \begin{aligned}
-g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) \label{eq:clap}
+\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) \label{eq:clap}
 \end{aligned}
@@
 \begin{aligned}
-g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  \label{eq:grav}
+\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  \label{eq:grav}
 \end{aligned}
@@
 The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion.
-In practice, models are in fact updated on a regular basis (Upadhyay, Joshi, and Lakkaraju 2021).
+In practice, models are in fact updated on a regular basis \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
 We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time \(t=0\).
 That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
-In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in Upadhyay, Joshi, and Lakkaraju (2021).
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
 We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model \(M\) is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
@@
 While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here Grinsztajn, Oyallon, and Varoquaux (2022).
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}36{]}}, \protect\hyperlink{ref-grinsztajn2022tree}{{[}37{]}}).
 
@@
 This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse.
-We demonstrate that long-held beliefs as to what defines optimality in AR, are too short-sighted to serve as a foundation for applications of recourse in practice.
-Specifically, we run multiple experiments that simulate the application of recourse in practice using various popular counterfactual generators and find that all of them induce substantial domain and model shifts.
+We demonstrate that long-held beliefs as to what defines optimality in AR, may not be applicable in situations that involve decisions over scarce resources.
+Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts.
 We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse.
@@
 To this end we have open-sourced all of our code along with a Julia package - \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}.
-The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility (Altmeyer 2022).
+The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}.
 That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
+\CSLLeftMargin{{[}1{]} }%
+\CSLRightInline{C. O'neil, \emph{Weapons of math destruction: How big data increases inequality and threatens democracy}. Crown, 2016.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
+\CSLLeftMargin{{[}2{]} }%
+\CSLRightInline{C.
+Rudin, {``Stop explaining black box machine learning models for high stakes decisions and use interpretable models instead,''} \emph{Nature Machine Intelligence}, vol.
+1, no.
+5, pp.
+206--215, 2019.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
+\CSLLeftMargin{{[}3{]} }%
+\CSLRightInline{M.
+Pawelczyk, S.
+Bielawski, J.
+van den Heuvel, T.
+Richter, and G.
+Kasneci, {``Carla: A python library to benchmark algorithmic recourse and counterfactual explanation algorithms,''} \emph{arXiv preprint arXiv:2108.00783}, 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
+\CSLLeftMargin{{[}4{]} }%
+\CSLRightInline{S.
+Wachter, B.
+Mittelstadt, and C.
+Russell, {``Counterfactual explanations without opening the black box: Automated decisions and the GDPR,''} \emph{Harv.
+JL \& Tech.}, vol.
+31, p.
+841, 2017.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
+\CSLLeftMargin{{[}5{]} }%
+\CSLRightInline{P.
+Altmeyer, \emph{{CounterfactualExplanations.jl - a Julia package for Counterfactual Explanations and Algorithmic Recourse}}.
+2022.
+Available: \url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
+\CSLLeftMargin{{[}6{]} }%
+\CSLRightInline{L.
+Schut \emph{et al.}, {``Generating interpretable counterfactual explanations by implicit minimisation of epistemic and aleatoric uncertainties,''} in \emph{International conference on artificial intelligence and statistics}, 2021, pp.
+1756--1764.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
+\CSLLeftMargin{{[}7{]} }%
+\CSLRightInline{S.
+Joshi, O.
+Koyejo, W.
+Vijitbenjaronk, B.
+Kim, and J.
+Ghosh, {``Towards realistic individual recourse and actionable explanations in black-box decision making systems,''} \emph{arXiv preprint arXiv:1907.09615}, 2019.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
+\CSLLeftMargin{{[}8{]} }%
+\CSLRightInline{R.
+K.
+Mothilal, A.
+Sharma, and C.
+Tan, {``Explaining machine learning classifiers through diverse counterfactual explanations,''} in \emph{Proceedings of the 2020 conference on fairness, accountability, and transparency}, 2020, pp.
+607--617.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
+\CSLLeftMargin{{[}9{]} }%
+\CSLRightInline{J.
+Antorán, U.
+Bhatt, T.
+Adel, A.
+Weller, and J.
+M.
+Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
+\CSLLeftMargin{{[}10{]} }%
+\CSLRightInline{A.-H.
+Karimi, G.
+Barthe, B.
+Schölkopf, and I.
+Valera, {``A survey of algorithmic recourse: Definitions, formulations, solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050}, 2020.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
+\CSLLeftMargin{{[}11{]} }%
+\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual explanations for machine learning: A review,''} \emph{arXiv preprint arXiv:2010.10596}, 2020.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
+\CSLLeftMargin{{[}12{]} }%
+\CSLRightInline{B.
+Ustun, A.
+Spangher, and Y.
+Liu, {``Actionable recourse in linear classification,''} in \emph{Proceedings of the conference on fairness, accountability, and transparency}, 2019, pp.
+10--19.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
+\CSLLeftMargin{{[}13{]} }%
+\CSLRightInline{A.-H.
+Karimi, B.
+Schölkopf, and I.
+Valera, {``Algorithmic recourse: From counterfactual explanations to interventions,''} in \emph{Proceedings of the 2021 ACM conference on fairness, accountability, and transparency}, 2021, pp.
+353--362.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
+\CSLLeftMargin{{[}14{]} }%
+\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards robust and reliable algorithmic recourse,''} \emph{arXiv preprint arXiv:2102.13620}, 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
+\CSLLeftMargin{{[}15{]} }%
+\CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero, {``Generating collective counterfactual explanations in score-based classification via mathematical optimization,''} 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
+\CSLLeftMargin{{[}16{]} }%
+\CSLRightInline{S.
+Rabanser, S.
+Günnemann, and Z.
+Lipton, {``Failing loudly: An empirical study of methods for detecting dataset shift,''} \emph{Advances in Neural Information Processing Systems}, vol.
+32, 2019.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
+\CSLLeftMargin{{[}17{]} }%
+\CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly detection: A survey,''} \emph{ACM computing surveys (CSUR)}, vol. 41, no. 3, pp. 1--58, 2009.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
+\CSLLeftMargin{{[}18{]} }%
+\CSLRightInline{G. Widmer and M. Kubat, {``Learning in the presence of concept drift and hidden contexts,''} \emph{Machine learning}, vol. 23, no. 1, pp. 69--101, 1996.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
+\CSLLeftMargin{{[}19{]} }%
+\CSLRightInline{J. Gama, I. Žliobaitė, A. Bifet, M. Pechenizkiy, and A. Bouchachia, {``A survey on concept drift adaptation,''} \emph{ACM computing surveys (CSUR)}, vol. 46, no. 4, pp. 1--37, 2014.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
+\CSLLeftMargin{{[}20{]} }%
+\CSLRightInline{K.
+Nelson, G.
+Corbin, M.
+Anania, M.
+Kovacs, J.
+Tobias, and M.
+Blowers, {``Evaluating model drift in machine learning algorithms,''} in \emph{2015 IEEE symposium on computational intelligence for security and defense applications (CISDA)}, 2015, pp.
+1--8.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-ackerman2021machine}{}}%
-Ackerman, Samuel, Parijat Dube, Eitan Farchi, Orna Raz, and Marcel Zalmanovici.
-2021.
-{``Machine Learning Model Drift Detection via Weak Data Slices.''} In \emph{2021 IEEE/ACM Third International Workshop on Deep Learning for Testing and Testing for Deep Learning (DeepTest)}, 1--8.
-IEEE.
-
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
-Altmeyer, Patrick.
-2022.
-\emph{{CounterfactualExplanations.jl - a Julia package for Counterfactual Explanations and Algorithmic Recourse}} (version 0.1.2).
-\url{https://github.com/pat-alt/CounterfactualExplanations.jl}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José Miguel Hernández-Lobato.
-2020.
-{``Getting a Clue: A Method for Explaining Uncertainty Estimates.''} \emph{arXiv Preprint arXiv:2006.06848}.
+\CSLLeftMargin{{[}21{]} }%
+\CSLRightInline{S.
+Ackerman, P.
+Dube, E.
+Farchi, O.
+Raz, and M.
+Zalmanovici, {``Machine learning model drift detection via weak data slices,''} in \emph{2021 IEEE/ACM third international workshop on deep learning for testing and testing for deep learning (DeepTest)}, 2021, pp.
+1--8.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
+\CSLLeftMargin{{[}22{]} }%
+\CSLRightInline{R.
+M.
+B.
+de Oliveira and D.
+Martens, {``A framework and benchmarking study for counterfactual generating methods on tabular data,''} \emph{Applied Sciences}, vol.
+11, no.
+16, p.
+7274, 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
+\CSLLeftMargin{{[}23{]} }%
+\CSLRightInline{A.-K. Dombrowski, J. E. Gerken, and P. Kessel, {``Diffeomorphic explanations with normalizing flows,''} 2021.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
+\CSLLeftMargin{{[}24{]} }%
+\CSLRightInline{R. S. Pindyck and D. L. Rubinfeld, \emph{Microeconomics}. Pearson Education, 2014.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
+\CSLLeftMargin{{[}25{]} }%
+\CSLRightInline{A.
+Gretton, K.
+M.
+Borgwardt, M.
+J.
+Rasch, B.
+Schölkopf, and A.
+Smola, {``A kernel two-sample test,''} \emph{The Journal of Machine Learning Research}, vol.
+13, no.
+1, pp.
+723--773, 2012.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
+\CSLLeftMargin{{[}26{]} }%
+\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing kernel hilbert spaces in probability and statistics}. Springer Science \& Business Media, 2011.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
-Arcones, Miguel A, and Evarist Gine. 1992. {``On the Bootstrap of u and v Statistics.''} \emph{The Annals of Statistics}, 655--74.
-
-\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
-Berlinet, Alain, and Christine Thomas-Agnan. 2011. \emph{Reproducing Kernel Hilbert Spaces in Probability and Statistics}. Springer Science \& Business Media.
+\CSLLeftMargin{{[}27{]} }%
+\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
+\CSLLeftMargin{{[}28{]} }%
+\CSLRightInline{S. Hanneke, {``A bound on the label complexity of agnostic active learning,''} in \emph{Proceedings of the 24th international conference on machine learning}, 2007, pp. 353--360.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
+\CSLLeftMargin{{[}29{]} }%
+\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
+\CSLLeftMargin{{[}30{]} }%
+\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell, {``Simple and scalable predictive uncertainty estimation using deep ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
+\CSLLeftMargin{{[}31{]} }%
+\CSLRightInline{K.
+Competition, {``Give me some credit, improve on the state of the art in credit scoring by predicting the probability that somebody will experience financial distress in the next two years.''} \url{https://www.kaggle.com/c/GiveMeSomeCredit}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
+\CSLLeftMargin{{[}32{]} }%
+\CSLRightInline{I.-C.
+Yeh and C.
+Lien, {``The comparisons of data mining techniques for the predictive accuracy of probability of default of credit card clients,''} \emph{Expert systems with applications}, vol.
+36, no.
+2, pp.
+2473--2480, 2009.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
+\CSLLeftMargin{{[}33{]} }%
+\CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine learning in python,''} \emph{the Journal of machine Learning research}, vol. 12, pp. 2825--2830, 2011.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
+\CSLLeftMargin{{[}34{]} }%
+\CSLRightInline{R. K. Pace and R. Barry, {``Sparse spatial autoregressions,''} \emph{Statistics \& Probability Letters}, vol. 33, no. 3, pp. 291--297, 1997.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
+\CSLLeftMargin{{[}35{]} }%
+\CSLRightInline{C. M., {``Racist data destruction? - a boston housing dataset controversy,''} 2019, Available: \url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-Borisov, Vadim, Tobias Leemann, Kathrin Seßler, Johannes Haug, Martin Pawelczyk, and Gjergji Kasneci.
-2021.
-{``Deep Neural Networks and Tabular Data: A Survey.''} \emph{arXiv Preprint arXiv:2110.01889}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-carrizosa2021generating}{}}%
-Carrizosa, Emilio, Jasone Ramırez-Ayerbe, and Dolores Romero. 2021. {``Generating Collective Counterfactual Explanations in Score-Based Classification via Mathematical Optimization.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chandola2009anomaly}{}}%
-Chandola, Varun, Arindam Banerjee, and Vipin Kumar. 2009. {``Anomaly Detection: A Survey.''} \emph{ACM Computing Surveys (CSUR)} 41 (3): 1--58.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
-Competition, Kaggle.
-n.d.
-{``Give Me Some Credit, Improve on the State of the Art in Credit Scoring by Predicting the Probability That Somebody Will Experience Financial Distress in the Next Two Years.''} \url{https://www.kaggle.com/c/GiveMeSomeCredit}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dombrowski2021diffeomorphic}{}}%
-Dombrowski, Ann-Kathrin, Jan E Gerken, and Pan Kessel.
-2021.
-{``Diffeomorphic Explanations with Normalizing Flows.''} In \emph{ICML Workshop on Invertible Neural Networks, Normalizing Flows, and Explicit Likelihood Models}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gama2014survey}{}}%
-Gama, João, Indrė Žliobaitė, Albert Bifet, Mykola Pechenizkiy, and Abdelhamid Bouchachia. 2014. {``A Survey on Concept Drift Adaptation.''} \emph{ACM Computing Surveys (CSUR)} 46 (4): 1--37.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
-Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf, and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The Journal of Machine Learning Research} 13 (1): 723--73.
+\CSLLeftMargin{{[}36{]} }%
+\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M. Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: A survey,''} \emph{arXiv preprint arXiv:2110.01889}, 2021.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
-Grinsztajn, Léo, Edouard Oyallon, and Gaël Varoquaux. 2022. {``Why Do Tree-Based Models Still Outperform Deep Learning on Tabular Data?''} \emph{arXiv Preprint arXiv:2207.08815}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
-Hanneke, Steve. 2007. {``A Bound on the Label Complexity of Agnostic Active Learning.''} In \emph{Proceedings of the 24th International Conference on Machine Learning}, 353--60.
-
-\leavevmode\vadjust pre{\hypertarget{ref-innes2018fashionable}{}}%
-Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and Viral Shah.
-2018.
-{``Fashionable Modelling with Flux.''} \emph{arXiv Preprint arXiv:1811.01457}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and Joydeep Ghosh.
-2019.
-{``Towards Realistic Individual Recourse and Actionable Explanations in Black-Box Decision Making Systems.''} \emph{arXiv Preprint arXiv:1907.09615}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
-Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel Valera.
-2020.
-{``A Survey of Algorithmic Recourse: Definitions, Formulations, Solutions, and Prospects.''} \emph{arXiv Preprint arXiv:2010.04050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
-Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera.
-2021.
-{``Algorithmic Recourse: From Counterfactual Explanations to Interventions.''} In \emph{Proceedings of the 2021 ACM Conference on Fairness, Accountability, and Transparency}, 353--62.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
-Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2016. {``Simple and Scalable Predictive Uncertainty Estimation Using Deep Ensembles.''} \emph{arXiv Preprint arXiv:1612.01474}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-carlisle2019racist}{}}%
-M., Carlisle. 2019. {``Racist Data Destruction? - a Boston Housing Dataset Controversy.''} \url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
-Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan.
-2020.
-{``Explaining Machine Learning Classifiers Through Diverse Counterfactual Explanations.''} In \emph{Proceedings of the 2020 Conference on Fairness, Accountability, and Transparency}, 607--17.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nelson2015evaluating}{}}%
-Nelson, Kevin, George Corbin, Mark Anania, Matthew Kovacs, Jeremy Tobias, and Misty Blowers.
-2015.
-{``Evaluating Model Drift in Machine Learning Algorithms.''} In \emph{2015 IEEE Symposium on Computational Intelligence for Security and Defense Applications (CISDA)}, 1--8.
-IEEE.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
-O'neil, Cathy. 2016. \emph{Weapons of Math Destruction: How Big Data Increases Inequality and Threatens Democracy}. Crown.
-
-\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
-Oliveira, Raphael Mazzine Barbosa de, and David Martens. 2021. {``A Framework and Benchmarking Study for Counterfactual Generating Methods on Tabular Data.''} \emph{Applied Sciences} 11 (16): 7274.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
-Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3): 291--97.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
-Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias Richter, and Gjergji Kasneci.
-2021.
-{``Carla: A Python Library to Benchmark Algorithmic Recourse and Counterfactual Explanation Algorithms.''} \emph{arXiv Preprint arXiv:2108.00783}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
-Pedregosa, Fabian, Gaël Varoquaux, Alexandre Gramfort, Vincent Michel, Bertrand Thirion, Olivier Grisel, Mathieu Blondel, et al.
-2011.
-{``Scikit-Learn: Machine Learning in Python.''} \emph{The Journal of Machine Learning Research} 12: 2825--30.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pindyck2014microeconomics}{}}%
-Pindyck, Robert S, and Daniel L Rubinfeld. 2014. \emph{Microeconomics}. Pearson Education.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabanser2019failing}{}}%
-Rabanser, Stephan, Stephan Günnemann, and Zachary Lipton.
-2019.
-{``Failing Loudly: An Empirical Study of Methods for Detecting Dataset Shift.''} \emph{Advances in Neural Information Processing Systems} 32.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
-Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning Models for High Stakes Decisions and Use Interpretable Models Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan Sacaleanu, Yarin Gal, et al.
-2021.
-{``Generating Interpretable Counterfactual Explanations by Implicit Minimisation of Epistemic and Aleatoric Uncertainties.''} In \emph{International Conference on Artificial Intelligence and Statistics}, 1756--64.
-PMLR.
-
-\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
-Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021. {``Towards Robust and Reliable Algorithmic Recourse.''} \emph{arXiv Preprint arXiv:2102.13620}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
-Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable Recourse in Linear Classification.''} In \emph{Proceedings of the Conference on Fairness, Accountability, and Transparency}, 10--19.
-
-\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
-Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual Explanations for Machine Learning: A Review.''} \emph{arXiv Preprint arXiv:2010.10596}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017. {``Counterfactual Explanations Without Opening the Black Box: Automated Decisions and the GDPR.''} \emph{Harv. JL \& Tech.} 31: 841.
-
-\leavevmode\vadjust pre{\hypertarget{ref-widmer1996learning}{}}%
-Widmer, Gerhard, and Miroslav Kubat. 1996. {``Learning in the Presence of Concept Drift and Hidden Contexts.''} \emph{Machine Learning} 23 (1): 69--101.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
-Yeh, I-Cheng, and Che-hui Lien.
-2009.
-{``The Comparisons of Data Mining Techniques for the Predictive Accuracy of Probability of Default of Credit Card Clients.''} \emph{Expert Systems with Applications} 36 (2): 2473--80.
+\CSLLeftMargin{{[}37{]} }%
+\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} \emph{arXiv preprint arXiv:2207.08815}, 2022.}
 
@@
 
-\{\{\textless{} include sections/appendix.qmd \textgreater\}\}
-
 \end{document}

### 2022-08-16 paper/sections/conclusion.rmd
moved everything to R Mardkown and now template works out well with rticles
some minor things
progress on results, formatting and comments

--- 2022-08-15 dev/paper/sections/conclusion.qmd
@@
-# Concluding Remarks {#sec-conclusion}
+# Concluding Remarks {#conclusion}
 
 This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse.
-We demonstrate that long-held beliefs as to what defines optimality in AR, are too short-sighted to serve as a foundation for applications of recourse in practice.
-Specifically, we run multiple experiments that simulate the application of recourse in practice using various popular counterfactual generators and find that all of them induce substantial domain and model shifts.
+We demonstrate that long-held beliefs as to what defines optimality in AR, may not be applicable in situations that involve decisions over scarce resources.
+Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts.
 We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse.
@@
 Through our work we hope to inspire future research on this important topic.
-To this end we have open-sourced all of our code along with a Julia package - `AlgorithmicRecourseDynamics.jl`.
+To this end we have open-sourced all of our code along with a Julia package - [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).
 The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations].

### 2022-08-16 paper/sections/empirical.rmd
progress on results, formatting and comments

@@
 They are also the model of choice in @schut2021generating.
-The actual neural network architectures are kept simple (@tbl-mlp), since we are only marginally concerned with achieving good initial classifier performance.
+The actual neural network architectures are kept simple (Table \@ref(tab:mlp)), since we are only marginally concerned with achieving good initial classifier performance.
 For the real-world datasets we using mini-batch training and dropout regularization.
 
-The Latent Space generators rely on separate generative models.
+The Latent Space generator relies on a separate generative model.
 Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose.
-As with the classifiers, we deliberately choose to work with fairly simple architectures (@tbl-vae).
+As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \@ref(tab:vae)).
 More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
@@
 
-<!-- | | Hidden Dim. | Hidden Layers | Batch | Dropout | -->
-<!-- |------|------|------|-----|-----| -->
-<!-- | Synthetic | 32 | 1 | - | - | -->
-<!-- | Real-World | 32 | 2 | 50 | 0.25 | -->
+```{r mlp}
+tab <- data.frame(
+  "Hidden Dim." = c(32,32),
+  "Hidden Layers." = c(1,2),
+  "Batch" = c("-",50),
+  "Dropout" = c("-",0.25)
+)
+rownames(tab) <- c("Synthetic", "Real-World")
+knitr::kable(
+  tab, booktabs = TRUE,
+  caption = 'MLP architecture and training parameters.',
+  col.names = c("Hidden Dim.","Hidden Layers", "Batch", "Dropout")
+)
+```
 
-<!-- : MLP {#tbl-mlp}  -->
-
-<!-- |  | Hidden Dim. | Epochs | -->
-<!-- |------|------|------| -->
-<!-- | Synthetic | 2 | 100 |  -->
-<!-- | Real-World | 8 | 250 |  -->
-
-<!-- : Variational Autoencoder {#tbl-vae} -->
+```{r vae}
+tab <- data.frame(
+  "Hidden Dim." = c(2,8),
+  "Epochs" = c(100,250)
+)
+rownames(tab) <- c("Synthetic", "Real-World")
+knitr::kable(
+  tab, booktabs = TRUE,
+  caption = 'VAE architecture and training parameters.',
+  col.names = c("Hidden Dim.","Epochs")
+)
+```
 

### 2022-08-16 paper/sections/empirical_2.rmd
progress on results, formatting and comments

@@
 
-Below we first present our main findings with respect to the presence of endogenous macrodynamics in Algorithmic Recourse.
-We then go on to introduce and test a number of potential mitigation strategies.
+Using the framework introduced in the previous section we are interested in answering the following research questions in this section:
+
+- Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
+- If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+- Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
+- What seem to be drivers of endogenous dynamics in Algorithmic Recourse?
+
+Below we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.
 
@@
 - The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
-- The reduction in model performance appears to be most severe when classes are not perfectly separable.
-- With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.
+- The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial performance of the classifier was weak to begin with.
+- With the exception of the Greedy generator, all other generators generally perform somewhat better overall than the baseline (Wachter) as expected.
 
-Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix.
-Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above.
-Figure \@ref(fig:syn) presents a small subset of our results for the synthetic dataset with overlapping classes.
+Focusing first on synthetic data, Figure \@ref(fig:syn) presents our findings for the dataset with overlapping classes.
 It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds.
+
 The top row shows the estimated domain shifts.
@@
 In general, we have observed the opposite.
-The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier.
+
+The second row shows the estimated model shifts, where here we have used the grid approach explained earlier.
 As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
 In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
-The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.
 
-```{r syn, fig.cap="Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds."}
+The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
+Related to this the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
+We also note that the final classifiers are more decisive, although as we noted earlier this may to some extent just be a byproduct of retraining the model throughout the course of the experiment.
+
+Figure \@ref(fig:syn) also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently.
+Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP.
+It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.
+
+```{r syn, fig.cap="Results for synthetic data with overlapping classes.
+The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points.
+Error bars indicate the standard deviation across folds."}
 knitr::include_graphics("www/synthetic_results.png")
@@
 
-Turning to the real-world data next ...
+Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real) where each column corresponds to one of the three data sets.
+The results shown here are for the deep ensemble, which once again largely resemble those for the MLP.
+Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts.
+While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
 
-```{r real, fig.cap="Results for real-world datasets."}
+The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large.
+Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
+Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.
+
+Out-of-sample model performance also deteriorates across the board and substantially so: the smallest average reduction in F-Scores of around 15-20 percentage points is observed for the California Housing dataset.
+For this dataset we achieved the highest initial model performance of just under 90 percent, indicating once again that weaker classifiers may be more exposed to endogenous dynamics.
+As with the synthetic data, the estimates for logistic regression are qualitatively in line with the above, but quantitatively even more pronounced.
+
+Zooming in on individual features we ...
+
+```{r real, fig.cap="Results for deep ensemble using real-world datasets.
+The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points.
+Error bars indicate the standard deviation across folds."}
 knitr::include_graphics("www/real_world_results.png")

### 2022-08-16 paper/sections/limitations.rmd
moved everything to R Mardkown and now template works out well with rticles
progress on results, formatting and comments

--- 2022-08-12 dev/paper/sections/limitations.qmd
@@
-# Limitations and Future Work {#sec-limit}
+# Limitations and Future Work {#limit}
 
@@
 
-## Data {#sec-limit-data}
+## Data {#limit-data}
 
@@
 While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here [@borisov2021deep, @grinsztajn2022tree].
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (@borisov2021deep, @grinsztajn2022tree).

### 2022-08-16 paper/sections/methodology.rmd
progress on results, formatting and comments

@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} (\#eq:general)
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} (\#eq:general)
 \end{aligned} 
@@
 The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
-The authors show that in this setting the cost function $h(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
+The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
 Schut et al.
@@
 [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
-In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $h(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ...
+In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ...
 , s_K$ that look as different from each other as possible.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  (\#eq:collective)
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  (\#eq:collective)
 \end{aligned} 
@@
 
-Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized.
-The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
+Here $\text{cost}(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized.
+The newly introduced term $\text{extcost}(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
 The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics].
 We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
-As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
+As with individual recourse, the exact choice of $\text{extcost}(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
 That being said, we do propose a few potential mitigation strategies in Section \@ref(mitigate).

### 2022-08-16 paper/sections/methodology_2.rmd
progress on results, formatting and comments

@@
 \begin{aligned}
-MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) (\#eq:mmd)
+MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j) (\#eq:mmd)
 \end{aligned}
@@
 
-where $\mathcal{F}$ is a unit ball in a Reproducing Kernel Hilbert Space $\mathcal{H}$ @berlinet2011reproducing, and $X$, $\tilde{X}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel.
-MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in RKHS $\mathcal{H}$.
+where $X=\{x_1,...,x_m\}$, $\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel.
+MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in a Reproducing Kernel Hilbert Space, $\mathcal{H}$ [@berlinet2011reproducing].
 An important consideration is the choice of the kernel function $k(\cdot,\cdot)$.
 In our implementation we make use of Gaussian kernel with a constant length-scale parameter of $0.5$.
-As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD_u^2[F,X,\tilde{X}]=0$ if and only if $X=\tilde{X}$.
+As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.
 
@@
 Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples.
-Then under the null hypothesis we should have that $MMD_u^2[F,X^\prime,\tilde{X}^\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$.
+Then under the null hypothesis we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$.
 The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

### 2022-08-16 paper/sections/mitigation.rmd
progress on results, formatting and comments

@@
 The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$.
+This threshold determines when a proposed counterfactual should be considered as valid.
 Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction.
@@
 Another potential strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
-Recall that $g(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
+Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
 Then formally we let
@@
 \begin{aligned}
-g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) (\#eq:clap)
+\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) (\#eq:clap)
 \end{aligned}
@@
 \begin{aligned}
-g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  (\#eq:grav)
+\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  (\#eq:grav)
 \end{aligned}
@@
 plt = plot(plts..., size=(750,200), layout=(1,4))
-savefig(plt, "dev/paper/www/mitigation.png")
+savefi\text{extcost}(plt, "dev/paper/www/mitigation.png")
 ```

### 2022-08-16 paper/paper.tex
more progress

@@
 If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same.
-We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a meshgrid over a subspace of the entire feature space.
+We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a mesh grid over a subspace of the entire feature space.
 For the latter approach we bound the subspace by the extrema of each feature.
@@
 
-Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 10,000 individuals each.
+Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 2,500 individuals each.
 We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.
@@
 Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP.
-It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.
+It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy, that explicitly address this quantity, fail to work as expected.
+
+The findings for the other synthetic datasets are broadly consistent with the observations above.
+For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
+For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.}.
+Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \ref{fig:syn}.
+The same broadly holds for the Moons data, For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.
+
+Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment.
+That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.
 
@@
 \caption{Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points.
+The shown model MMD (PP MMD) was computed over a mesh grid of 1,000 points.
 Error bars indicate the standard deviation across folds.}\label{fig:syn}
@@
 
-The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large.
+The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated PP MMD is statistically significant and large.
 Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
@@
 \caption{Results for deep ensemble using real-world datasets.
-The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points.
+The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid.
 Error bars indicate the standard deviation across folds.}\label{fig:real}
@@
 Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of points.
+The shown model MMD (PP MMD) was computed over a mesh grid of points.
 Error bars indicate the standard deviation across folds.}\label{fig:mitigate-results}
@@
 An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
-A snapshot of the results is shown in Figure \ref{fig:mitigate-latent-results}.
+In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+In Figure \ref{fig:mitigate-latent-results} the conventional Latent Space generator with \(\gamma=0.5\) serves as our baseline.
+Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.
 
@@
 Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of points.
+The shown model MMD (PP MMD) was computed over a mesh grid of points.
 Error bars indicate the standard deviation across folds.}\label{fig:mitigate-latent-results}
+\end{figure}
+
+Finally, Figure \ref{fig:mitigate-real-world-results} shows the results for our real-world data.
+We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \ref{fig:mitigate-results}\footnote{Estimated domain shifts (not shown) were largely insubstantial, as in Figure \ref{fig:real} in the previous section.}.
+Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
+For the GMSC dataset we observe no notable differences.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{www/mitigation_real_world_results}
+
+}
+
+\caption{The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$.
+Results for deep ensemble using real-world datasets.
+The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid.
+Error bars indicate the standard deviation across folds.}\label{fig:mitigate-real-world-results}
 \end{figure}

### 2022-08-16 paper/sections/empirical_2.rmd
more progress

@@
 Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP.
-It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.
+It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy, that explicitly address this quantity, fail to work as expected.
+
+The findings for the other synthetic datasets are broadly consistent with the observations above.
+For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
+For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.].
+Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn).
+The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.
+
+Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment.
+That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.
 
 ```{r syn, fig.cap="Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points.
+The shown model MMD (PP MMD) was computed over a mesh grid of 1,000 points.
 Error bars indicate the standard deviation across folds."}
@@
 
-The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large.
+The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated PP MMD is statistically significant and large.
 Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
@@
 ```{r real, fig.cap="Results for deep ensemble using real-world datasets.
-The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points.
+The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid.
 Error bars indicate the standard deviation across folds."}

### 2022-08-16 paper/sections/mitigation.rmd
more progress

@@
 Results for synthetic data with overlapping classes.
-The shown model MMD (PP MMD) was computed over a meshgrid of points.
+The shown model MMD (PP MMD) was computed over a mesh grid of points.
 Error bars indicate the standard deviation across folds."}
@@
 An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
-A snapshot of the results is shown in Figure \@ref(fig:mitigate-latent-results).
+In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline.
+Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.
 
@@
 ```
+
+Finally, Figure \@ref(fig:mitigate-real-world-results) shows the results for our real-world data.
+We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.].
+Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
+For the GMSC dataset we observe no notable differences.
+
+```{r mitigate-real-world-results, fig.cap="The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\\gamma=0.5$.
+Results for deep ensemble using real-world datasets.
+The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid.
+Error bars indicate the standard deviation across folds."}
+knitr::include_graphics("www/mitigation_real_world_results.png")
+```

### 2022-08-16 paper/paper.tex
toned down and added more examples

@@
 
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}{Lemma}[section]
+\newtheorem{corollary}{Corollary}[section]
+\newtheorem{proposition}{Proposition}[section]
+\newtheorem{conjecture}{Conjecture}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Exercise}[section]
+\theoremstyle{definition}
+\newtheorem{hypothesis}{Hypothesis}[section]
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
 \begin{document}
@@
 
-Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 We think that these types of endogenous dynamics may be problematic and warrant our attention.
-Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
-Secondly and relatedly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
-Average default risk across all borrowers can therefore be expected to increase.
-Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals.
-In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.
+From a purely technical we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+Secondly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
+We would argue that in some practical applications this can be expected to generate costs for the involved stakeholders.
+To follow our argument, consider the following two examples:
+
+\begin{example}[Consumer Credit]
+\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their credit worthiness.
+Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction.
+Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan.
+This represents a cost to the retail bank.
+\end{example}
+
+\begin{example}[Student Admission]
+\protect\hypertarget{exm:student}{}\label{exm:student}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a university in their student admission process.
+Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction.
+Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more liekly to fail their degree than students that were admitted in previous years.
+The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse.
+This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
+\end{example}
+
+Both of these examples are exaggerated simplifications of potential real-world scenarios, but they serve to illustrate the point that recourse for one single individual may exert negative externalities on other individuals.
 
@@
 
-Zooming in on individual features we \ldots{}
-
 \begin{figure}
@@
 In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool.
-As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
+As we have set out in Example \ref{exm:student} of our introduction, these model owners may be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
 This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.
@@
 This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse.
-We demonstrate that long-held beliefs as to what defines optimality in AR, may not be applicable in situations that involve decisions over scarce resources.
+We demonstrate that long-held beliefs as to what defines optimality in AR, may not be suitable in contexts that involves large groups of individuals facing adverse outcomes.
 Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts.
-We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse.
+We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse in these types of situations.
 By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit.

### 2022-08-16 paper/sections/introduction.rmd
toned down and added more examples

@@
 
-Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c).
+Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 We think that these types of endogenous dynamics may be problematic and warrant our attention.
-Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
-Secondly and relatedly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy.
-Average default risk across all borrowers can therefore be expected to increase.
-Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals.
-In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.
+From a purely technical we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+Secondly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
+We would argue that in some practical applications this can be expected to generate costs for the involved stakeholders.
+To follow our argument, consider the following two examples:
+
+::: {.example #consumer name="Consumer Credit"}
+Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their credit worthiness.
+Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction.
+Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan.
+This represents a cost to the retail bank.
+:::
+
+::: {.example #student name="Student Admission"}
+Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a university in their student admission process.
+Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction.
+Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more liekly to fail their degree than students that were admitted in previous years.
+The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse.
+This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
+:::
+
+Both of these examples are exaggerated simplifications of potential real-world scenarios, but they serve to illustrate the point that recourse for one single individual may exert negative externalities on other individuals.
 

### 2022-08-17 paper/paper.tex
more polishing

@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 
-\caption{Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).}\label{fig:poc}
+\caption{Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.}\label{fig:poc}
 \end{figure}
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} \label{eq:general}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} \label{eq:general}
 \end{aligned} 
@@
 Wachter et al.
-\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
-In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ...
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)\footnote{Note, in fact, that \eqref{eq:general} naturally lends itself to that idea: setting \(K\) to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel.
+Here, \(yloss(\cdot)\) is first broadcasted over elements of \(\mathbf{s}^\prime\) and then aggregated.
+This is exactly how counterfactual search is implemented in \texttt{CounterfactualExplanations.jl}.}.
+In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and thereby produce counterfactuals \(s_1, ...
 , s_K\) that look as different from each other as possible.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  \label{eq:collective}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \{ {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(\mathbf{s}^\prime))} + \lambda_2 {\text{extcost}(f(\mathbf{s}^\prime))} \}  \label{eq:collective}
 \end{aligned} 
@@
 
-Here \(\text{cost}(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized.
-The newly introduced term \(\text{extcost}(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\).
+Here \(\text{cost}(f(\mathbf{s}^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized.
+The newly introduced term \(\text{extcost}(f(\mathbf{s}^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\).
 The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}.
-We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
+We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well.
 As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.
@@
 The dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
-We begin in the static setting at time time \(t=0\): given some classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse.
+We begin in the static setting at time time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)).
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse\footnote{To keep notation simple we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.
+
 In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\).
@@
 \Procedure{Experiment}{$M,\mathcal{D},G$}
+\State $E\gets \emptyset$ \Comment{Initialize evaluation $E$.}
 \State $t\gets 0$
 \While{$t<T$}
-\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
-\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
+\State $\text{batch} \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$  (assignment).}
+\State $\text{batch}\gets G(\text{batch})$ \Comment{Generate counterfactuals.}
 \State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
+\State $E\gets \text{eval}(M,\mathcal{D}) \cup E$ \Comment{Update evaluation.}
+\State $t\gets t+1$ \Comment{Increment $t$.}
 \EndWhile
-\State \textbf{return} $M,\mathcal{D}$
+\State \textbf{return} $E, M,\mathcal{D}$
 \EndProcedure
@@
 
-A noteworthy practical consideration is the choice of \(T\) and \(B\).
-The higher these values, the more factual instances undergo recourse throughout the entire experiment.
+Note that the operation in line 4 is an assignment, rather than a copy operation, so any updates to `batch' will also affect \(\mathcal{D}\).
+The function \(\text{eval}(M,\mathcal{D})\) loosely denotes the computation of various evaluation metrics introduced below.
+In practice, these metrics can also be computed at regular intervals as opposed to every round.
+
+Along with any other fixed parameters affecting the counterfactual search, the parameters \(T\) and \(B\) are assumed as given in Algorithm \ref{algo-experiment}.
+Still, it worth noting that the higher these values, the more factual instances undergo recourse throughout the entire experiment.
 Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\).
-At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm's decisions.
 In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset.
-As we collect data at each time \(t\), we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
+As we compute evaluation metrics at regular intervals throughout the procedure, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
 
@@
 They are also the model of choice in \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
-The actual neural network architectures are kept simple (Table \ref{tab:mlp}), since we are only marginally concerned with achieving good initial classifier performance.
-For the real-world datasets we using mini-batch training and dropout regularization.
+The actual neural network architectures are kept simple (top half of Table \ref{tab:architecture}), since we are only marginally concerned with achieving good initial classifier performance.
 
@@
 Following the authors of both REVISE and CLUE we use Variational Autoencoders (\textbf{VAE}) for this purpose.
-As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \ref{tab:vae}).
+As with the classifiers, we deliberately choose to work with fairly simple architectures (bottom half of Table \ref{tab:architecture}).
 More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
@@
 
-\caption{\label{tab:mlp}MLP architecture and training parameters.}
+\caption{\label{tab:architecture}Neural network architectures and training parameters.}
 \centering
-\begin{tabular}[t]{lrrll}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llrlrllr}
 \toprule
-  & Hidden Dim. & Hidden Layers & Batch & Dropout\\
+ & Data & Hidden Dim. & Latent Dim. & Hidden Layers & Batch & Dropout & Epochs\\
 \midrule
-Synthetic & 32 & 1 & - & -\\
-Real-World & 32 & 2 & 50 & 0.25\\
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textbf{MLP}}\\
+\hspace{1em} & Synthetic & 32 & - & 1 & - & - & 100\\
+\cmidrule{2-8}
+\hspace{1em} & Real-World & 32 & - & 2 & 50 & 0.25 & 100\\
+\cmidrule{1-8}
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textbf{VAE}}\\
+\hspace{1em} & Synthetic & 32 & 2 & 1 & - & - & 100\\
+\cmidrule{2-8}
+\hspace{1em} & Real-World & 32 & 8 & 1 & - & - & 250\\
 \bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:vae}VAE architecture and training parameters.}
-\centering
-\begin{tabular}[t]{lrr}
-\toprule
-  & Hidden Dim. & Epochs\\
-\midrule
-Synthetic & 2 & 100\\
-Real-World & 8 & 250\\
-\bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
@@
 
-We use four synthetic binary classification datasets consisting of 1000 samples each.
-The datasets are presented in Figure \ref{fig:synthetic-data} (see also Appendix A for a formal description).
+We use four synthetic binary classification datasets consisting of 1000 samples each: \textbf{Overlapping}, \textbf{Linearly Separable}, \textbf{Circles} and \textbf{Moons}.
+The datasets are presented in Figure \ref{fig:synthetic-data}.
 Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
@@
 
-\caption{Synthetic classification datasets used in our experiments.}\label{fig:synthetic-data}
+\caption{Synthetic classification datasets used in our experiments.
+Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange.}\label{fig:synthetic-data}
 \end{figure}
@@
 
+To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments and we find them substantial enough to be considered costly; secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect.
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process.
+
 \hypertarget{mitigate}{%
@@
 \begin{aligned}
-\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) \label{eq:clap}
+\text{extcost}(f(\mathbf{s}^\prime)) = l(M(f(\mathbf{s}^\prime)),y^\prime) \label{eq:clap}
 \end{aligned}
@@
 \begin{aligned}
-\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  \label{eq:grav}
+\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  \label{eq:grav}
 \end{aligned}
@@
 
-\caption{Illustrative example demonstrating the properties of the various mitigation strategies.}\label{fig:mitigation}
+\caption{Illustrative example demonstrating the properties of the various mitigation strategies.
+Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange.}\label{fig:mitigation}
 \end{figure}

### 2022-08-17 paper/sections/empirical.rmd
more polishing

@@
 They are also the model of choice in @schut2021generating.
-The actual neural network architectures are kept simple (Table \@ref(tab:mlp)), since we are only marginally concerned with achieving good initial classifier performance.
-For the real-world datasets we using mini-batch training and dropout regularization.
+The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.
 
@@
 Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose.
-As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \@ref(tab:vae)).
+As with the classifiers, we deliberately choose to work with fairly simple architectures (bottom half of Table \@ref(tab:architecture)).
 More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators.
@@
 
-```{r mlp}
+```{r architecture}
 tab <- data.frame(
-  "Hidden Dim." = c(32,32),
-  "Hidden Layers." = c(1,2),
-  "Batch" = c("-",50),
-  "Dropout" = c("-",0.25)
+  "Model" = c("MLP","MLP","VAE","VAE"),
+  "Data" = c("Synthetic", "Real-World", "Synthetic", "Real-World"),
+  "Hidden" = c(32,32,32,32),
+  "Latent" = c("-","-",2,8),
+  "Layers" = c(1,2,1,1),
+  "Batch" = c("-",50,"-","-"),
+  "Dropout" = c("-",0.25,"-","-"),
+  "Epochs" = c(100,100,100,250)
 )
-rownames(tab) <- c("Synthetic", "Real-World")
-knitr::kable(
+library(kableExtra)
+kbl(
   tab, booktabs = TRUE,
-  caption = 'MLP architecture and training parameters.',
-  col.names = c("Hidden Dim.","Hidden Layers", "Batch", "Dropout")
-)
-```
-
-```{r vae}
-tab <- data.frame(
-  "Hidden Dim." = c(2,8),
-  "Epochs" = c(100,250)
-)
-rownames(tab) <- c("Synthetic", "Real-World")
-knitr::kable(
-  tab, booktabs = TRUE,
-  caption = 'VAE architecture and training parameters.',
-  col.names = c("Hidden Dim.","Epochs")
-)
+  caption = 'Neural network architectures and training parameters.',
+  col.names = c("Model","Data","Hidden Dim.","Latent Dim.","Hidden Layers", "Batch", "Dropout", "Epochs")
+) |> 
+  collapse_rows(1:2, row_group_label_position = 'stack') |>
+  kable_styling(latex_options = c("scale_down"))
 ```
@@
 
-We use four synthetic binary classification datasets consisting of 1000 samples each.
-The datasets are presented in Figure \@ref(fig:synthetic-data) (see also Appendix A for a formal description).
+We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons**.
+The datasets are presented in Figure \@ref(fig:synthetic-data).
 Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
 
-```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments."}
+```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments.
+Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange."}
 knitr::include_graphics("www/synthetic_data.png")

### 2022-08-17 paper/sections/methodology.rmd
more polishing

@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} (\#eq:general)
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} (\#eq:general)
 \end{aligned} 
@@
 Wachter et al.
-[@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual).
-In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ...
+[@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)^[Note, in fact, that \@ref(eq:general) naturally lends itself to that idea: setting $K$ to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel.
+Here, $yloss(\cdot)$ is first broadcasted over elements of $\mathbf{s}^\prime$ and then aggregated.
+This is exactly how counterfactual search is implemented in `CounterfactualExplanations.jl`.].
+In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and thereby produce counterfactuals $s_1, ...
 , s_K$ that look as different from each other as possible.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  (\#eq:collective)
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \{ {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(\mathbf{s}^\prime))} + \lambda_2 {\text{extcost}(f(\mathbf{s}^\prime))} \}  (\#eq:collective)
 \end{aligned} 
@@
 
-Here $\text{cost}(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized.
-The newly introduced term $\text{extcost}(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
+Here $\text{cost}(f(\mathbf{s}^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized.
+The newly introduced term $\text{extcost}(f(\mathbf{s}^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$.
 The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics].
-We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well.
+We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well.
 As with individual recourse, the exact choice of $\text{extcost}(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists.

### 2022-08-17 paper/sections/methodology_2.rmd
more polishing

@@
 The dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro) were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
-We begin in the static setting at time time $t=0$: given some classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$ we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse.
+We begin in the static setting at time time $t=0$: firstly, we have some binary classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$, where $\mathcal{D}_0$ and $\mathcal{D}_1$ denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].
+
 In order to simulate the dynamic process, we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$.
@@
 
-A noteworthy practical consideration is the choice of $T$ and $B$.
-The higher these values, the more factual instances undergo recourse throughout the entire experiment.
+Note that the operation in line 4 is an assignment, rather than a copy operation, so any updates to 'batch' will also affect $\mathcal{D}$.
+The function $\text{eval}(M,\mathcal{D})$ loosely denotes the computation of various evaluation metrics introduced below.
+In practice, these metrics can also be computed at regular intervals as opposed to every round.
+
+Along with any other fixed parameters affecting the counterfactual search, the parameters $T$ and $B$ are assumed as given in Algorithm \ref{algo-experiment}.
+Still, it worth noting that the higher these values, the more factual instances undergo recourse throughout the entire experiment.
 Of course, this is likely to lead to more pronounced domain and model shifts by time $T$.
-At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions.
 In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $\approx50\%$ of the negative instances from the initial dataset.
-As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
+As we compute evaluation metrics at regular intervals throughout the procedure, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.
 

### 2022-08-17 dev/notebooks/mitigation_strategies.qmd
minor things

@@
 transform!(df_plot, :generator => (X -> [x=="Generic_conservative" ? "Generic (γ=0.9)" : x for x in X]) => :generator)
+transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
 
@@
 using RCall
-scale_ = 2.2
+scale_ = 2.0
 R"""
+library(data.table)
+df_plot <- data.table($df_plot)
+model_order <- c("Linear", "MLP", "Deep Ensemble")
+df_plot[,model:=factor(model, levels=model_order)]
 library(ggplot2)
@@
     geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
-    facet_wrap(name ~ model, scale="free_y", ncol=$ncol) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(model), 
+        scales = "free_y"
+    ) +
     labs(y = "Value") + 
@@
         axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()
-    )
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    ) +
+    guides(fill=guide_legend(ncol=3))
 temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
 """
@@
 transform!(df_plot, :generator => (X -> [x=="Latent_conservative" ? "Latent (γ=0.9)" : x for x in X]) => :generator)
+transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
 
@@
 using RCall
-scale_ = 2.2
+scale_ = 1.9
 R"""
+library(data.table)
+df_plot <- data.table($df_plot)
+model_order <- c("Linear", "MLP", "Deep Ensemble")
+df_plot[,model:=factor(model, levels=model_order)]
 library(ggplot2)
@@
         axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()
-    )
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    ) +
+    guides(fill=guide_legend(ncol=4))
 temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
 """
@@
 ```{julia}
-max_obs = 1000
+max_obs = 2500
 data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
@@
 n_bootstrap = 1
+n_samples = 10000
 T = 250
@@
 
-### Latent Space Search
+### Chart in paper
 
 ```{julia}
-generators = Dict(
-    :REVISE=>GenericGenerator(decision_threshold=0.5),
-    :REVISE_conservative=>GenericGenerator(decision_threshold=0.9),
-    :Gravitational=>GravitationalGenerator(),
-    :ClapROAR=>ClapROARGenerator()
-)
+using DataFrames, Statistics
+model_ = :FluxEnsemble
+df = DataFrame() 
+for (key, val) in results
+    df_ = deepcopy(val.output)
+    df_.dataset .= key
+    df = vcat(df,df_)
+end
+df = df[df.n .== maximum(df.n),:]
+df = df[df.model .== model_,:]
+filter!(:value => x -> !any(f -> f(x), (ismissing, isnothing, isnan)), df)
+gdf = groupby(df, [:generator, :dataset, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.!=:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :dataset => (X -> [x=="cal_housing" ? "California Housing" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="credit_default" ? "Credit Default" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="gmsc" ? "GMSC" : x for x in X]) => :dataset)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="Generic" ? "Generic (γ=0.5)" : x for x in X]) => :generator)
+transform!(df_plot, :generator => (X -> [x=="Generic_conservative" ? "Generic (γ=0.9)" : x for x in X]) => :generator)
+
+ncol = length(unique(df_plot.dataset))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 2.0
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(dataset), 
+        scales = "free_y"
+    ) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    ) +
+    guides(fill=guide_legend(ncol=3))
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.85) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_real_world_results.png"), img)
 ```
-
-```{julia}
-generative_model_params = (epochs=250, latent_dim=8)
-experiments = set_up_experiments(
-    data_sets,models,generators; 
-    pre_train_models=100, model_params=model_params, 
-    data_loader=data_loader
-)
-```
-
-```{julia}
-n_evals = 5
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-n_bootstrap = 1
-T = 200
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,
-    save_name_suffix="latent",
-    evaluate_every=evaluate_every,
-    n_rounds=n_rounds, 
-    n_folds=n_folds, 
-    n_bootstrap=n_bootstrap, 
-    T=T,
-    latent_space=true,
-    generative_model_params=generative_model_params
-)
-Serialization.serialize(joinpath(output_path,"results_latent.jls"),results)
-```
-
-### Plots
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_real_world.jls"))
-```
-
-```{julia}
-using Images
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    plt = plot(res)
-    Images.save(joinpath(www_path, "line_chart_latent_$(data_name).png"), plt)
-    line_charts[data_name] = plt
-    plt = plot(res,maximum(res.output.n))
-    Images.save(joinpath(www_path, "errorbar_chart_latent_$(data_name).png"), plt)
-    errorbar_charts[data_name] = plt
-end
-```

### 2022-08-17 dev/paper/sections/empirical.qmd
minor things

@@
 Evaluation metrics are computed and stored every 10 rounds.
+To account for noise each individual experiment is repeat five times.
 
@@
 function plot_data(data,title)
-    plt = plot(title=title)
+    plt = plot(title=uppercasefirst(replace(string(title),"_" => " ")))
     scatter!(data)
@@
 plt = plot(plts..., layout=(1,4), size=(850,200))
-savefig(plt, "dev/paper/www/synthetic_data.png")
+savefig(plt, "paper/www/synthetic_data.png")
 ```

### 2022-08-17 src/base.jl

@@ -12,11 +12,11 @@ is_logging(io) = isa(io, Base.TTY) == false || (get(ENV, "CI", nothing) == "true
 
 function collect_output(
     experiment::Experiment, recourse_system::RecourseSystem, chosen_individuals::Union{Nothing,AbstractArray}, k::Int, n::Int, m::Int;
-    n_bootstrap=1000
+    n_bootstrap=1000, n_samples=1000
 )
 
     # Evaluate:
-    output = evaluate_system(recourse_system, experiment, n=n_bootstrap)
+    output = evaluate_system(recourse_system, experiment, n=n_bootstrap, n_samples=n_samples)
     
     # Add additional information:
     output.k .= k
@@ -43,7 +43,7 @@ end
 A wrapper function that runs the experiment for endogenous models shifts.
 """
 function run!(
-    experiment::Experiment; evaluate_every=10, n_bootstrap=1000, forward=false, show_progress=!is_logging(stderr), fixed_parameters...
+    experiment::Experiment; evaluate_every=10, n_bootstrap=1000, n_samples=1000, forward=false, show_progress=!is_logging(stderr), fixed_parameters...
 )
 
     # Load fixed hyperparameters:
@@ -91,7 +91,7 @@ function run!(
                 end
                 # Evaluate:
                 if n % evaluate_every == 0 
-                    output_checkpoint = collect_output(experiment, recourse_system, chosen_individuals[m], k, n, m, n_bootstrap=n_bootstrap)
+                    output_checkpoint = collect_output(experiment, recourse_system, chosen_individuals[m], k, n, m, n_bootstrap=n_bootstrap, n_samples=n_samples)
                     output[m] = vcat(output[m], output_checkpoint, cols=:union)
                 end
             end

### 2022-08-17 paper/paper.Rmd

@@
 # with_ifpdf: true
-# with_cite: true
+with_cite: true
 # amsmath need to be true to use with bookdown for referencing equations.
@@
 bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
+csl: ieee.csl
 output: 
@@
       before_body: "tex/before_body.tex"
+link-citations: true
 #citation_sorting: none   ## used as sorting option of the biblatex package (if selected)
@@
 :::
-
-{{< include sections/appendix.qmd >}}

### 2022-08-17 paper/sections/discussion.rmd

@@
 In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool.
-As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
+As we have set out in Example \@ref(exm:student) of our introduction, these model owners may be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether.
 This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

### 2022-08-17 paper/sections/empirical_2.rmd
toned down and added more examples

@@
 
-Zooming in on individual features we ...
-
 ```{r real, fig.cap="Results for deep ensemble using real-world datasets.
@@
 ```
+
+To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments and we find them substantial enough to be considered costly; secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect.
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process.

### 2022-08-17 paper/sections/introduction.rmd

@@
 
-```{r poc, fig.cap="Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d)."}
+```{r poc, fig.cap="Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class."}
 knitr::include_graphics("www/poc.png")

### 2022-08-17 paper/sections/mitigation.rmd

@@
 \begin{aligned}
-\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) (\#eq:clap)
+\text{extcost}(f(\mathbf{s}^\prime)) = l(M(f(\mathbf{s}^\prime)),y^\prime) (\#eq:clap)
 \end{aligned}
@@
 \begin{aligned}
-\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  (\#eq:grav)
+\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  (\#eq:grav)
 \end{aligned}
@@
 
-```{r mitigation, fig.cap="Illustrative example demonstrating the properties of the various mitigation strategies."}
+```{r mitigation, fig.cap="Illustrative example demonstrating the properties of the various mitigation strategies.
+Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange."}
 knitr::include_graphics("www/mitigation.png")

### 2022-08-17 paper/sections/related.rmd

@@
 Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.
-Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $H=h(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:
+Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:
 
 \begin{equation}
-\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
+\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
 \end{equation}
@@
 \begin{equation}
-x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) (\#eq:solution)
+x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) (\#eq:solution)
 \end{equation}

### 2022-08-17 README.qmd
another attempt at py 3.8, but not working. Experiment code throwing and error that looks CARLA related.
setting up Julia framework
setting up to have separate paper branch that will serve as static artifact of code at time of publication to ensure replicability

--- 2022-07-05 README.qmd
@@
 ---
-format: gfm
+format: 
+    gfm:
+        wrap: none
 ---
 
-# Algorithmic Recourse
+# AlgorithmicRecourseDynamics.jl - Research Paper 📝
 
-## Abstract
+**Note** ⚠: You are on the `#original-paper` branch of `AlgorithmicRecourseDynamics.jl`.
+This branch is a static artifact corresponding to the state of the package at the time the paper was first published.
+It can be used to replicate the original findings of the paper.
+For an up-to-date version of the package, please switch to the [`#main`](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl) branch.
 
-{{< include paper/sections/abstract.qmd >}}
+## At a Glance
 
-![](www/poc.png)
+The paper titles **Endogenous Macrodynamics in Algorithmic Recourse** is currently under review and not yet published. You can find
+a preprint along with other resources right here on this branch of the
+repository:
 
-## Code
+- [Paper](paper/paper.pdf)
+- [Notebooks](dev/notebooks/)
+- [Artifacts](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/releases/tag/artifacts) (including data and experimental results)
 
-...
+In this work we investigate what happens if Algorithmic Recourse is actually implemented by a large number of individuals.
+The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.
+
+![](paper/www/poc.png)
+
+## Paper Abstract
+
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
+Through this work we aim to close that gap by systematizing and extending existing knowledge.
+We first show that many of the existing methodologies can be collectively described by a generalized framework.
+We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
+We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources.
+Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-08-17 paper/paper.Rmd
setting up to have separate paper branch that will serve as static artifact of code at time of publication to ensure replicability

@@
   #         email: roy@replicant.offworld
-keywords: ["keyword 1", "keyword 2"]
+keywords: ["Algorithmic Recourse", "Counterfactual Explanations", "Explainable AI", "Dynamic Systems"]
 abstract: |
@@
 We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
-Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate a total of XX million counterfactuals and study the resulting domain and model shifts.
+Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
 We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources.

### 2022-08-17 dev/notebooks/real_world.qmd

@@
 ```{julia}
-max_obs = 1000
+max_obs = 2500
 data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
+choices = [
+    :cal_housing, 
+    :credit_default, 
+    :gmsc, 
+]
+data_sets = filter(p -> p[1] in choices, data_sets)
 ```
@@
     experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T, n_samples=n_samples,
     generative_model_params=generative_model_params
@@
 
-### Table in paper
+### Chart in paper
 
 ```{julia}
-using AlgorithmicRecourseDynamics: kable
-kable(results, [50])
+using DataFrames, Statistics
+model_ = :FluxEnsemble
+df = DataFrame() 
+for (key, val) in results
+    df_ = deepcopy(val.output)
+    df_.dataset .= key
+    df = vcat(df,df_)
+end
+df = df[df.n .== maximum(df.n),:]
+df = df[df.model .== model_,:]
+filter!(:value => x -> !any(f -> f(x), (ismissing, isnothing, isnan)), df)
+gdf = groupby(df, [:generator, :dataset, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :model_performance] for name in df_plot.name],:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+df_plot.name .= [r[:name] == "mmd" ? "$(r[:name])_$(r[:scope])" : r[:name] for r in eachrow(df_plot)]
+transform!(df_plot, :dataset => (X -> [x=="cal_housing" ? "California Housing" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="credit_default" ? "Credit Default" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="gmsc" ? "GMSC" : x for x in X]) => :dataset)
+transform!(df_plot, :name => (X -> [x=="mmd_domain" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_model" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
+
+ncol = length(unique(df_plot.dataset))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 1.75
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(dataset), 
+        scales = "free_y"
+    ) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_real_world_results.png"), img)
 ```

### 2022-08-17 dev/notebooks/synthetic.qmd

@@
 
+#### Line Chart
+
 ```{julia}
@@
 df = results[:overlapping].output
-df = df[df.n .== maximum(df.n),:]
+# df = df[df.n .== maximum(df.n),:]
 gdf = groupby(df, [:generator, :model, :n, :name, :scope])
 df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
-df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[[name in [:decisiveness, :disagreement, :mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
 df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
 df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="decisiveness" ? "Decisiveness" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="disagreement" ? "Disagreement" : x for x in X]) => :name)
 transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
@@
 transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
+transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
 
@@
 using RCall
-scale_ = 2
+scale_ = 1.5
 R"""
 library(ggplot2)
-plt <- ggplot($df_plot) +
-    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
-    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
-    facet_wrap(name ~ model, scale="free_y", ncol=$ncol) +
+plt <- ggplot($df_plot, aes(x=n, y=mean, ymin=ymin, ymax=ymax, color=generator)) +
+    geom_ribbon(aes(fill=generator), alpha=0.5) +
+    geom_line() +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(model), 
+        scales = "free_y"
+    ) +
     labs(y = "Value") + 
@@
         axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
     )
 temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_synthetic_results_line.png"), img)
+```
+
+#### Errorbar
+
+```{julia}
+using DataFrames, Statistics
+df = results[:overlapping].output
+df = df[[x ∈ [50] for x in df.n],:]
+gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:decisiveness, :disagreement, :mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="decisiveness" ? "Decisiveness" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="disagreement" ? "Disagreement" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
+transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
+
+ncol = length(unique(df_plot.model))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 1.5
+R"""
+library(data.table)
+df_plot <- data.table($df_plot)
+name_order <- c(
+    "MMD (domain)",
+    "MMD (model)",
+    "Performance",
+    "Disagreement",
+    "Decisiveness"
+)
+df_plot[,name:=factor(name, levels=name_order)]
+model_order <- c("Linear", "MLP", "Deep Ensemble")
+df_plot[,model:=factor(model, levels=model_order)]
+library(ggplot2)
+plt <- ggplot(df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange(aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=c(0.9,0.9)), size=0.5) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(model), 
+        scales = "free_y"
+    ) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
 """

### 2022-08-17 dev/notebooks/data_preprocessing/synthetic_data.qmd
sorted out deps and set up online appendix

---
title: Generating Synthetic Data
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
output_path = output_dir("synthetic")
www_path = www_dir("synthetic")
data_path = data_dir("synthetic")
```

```{julia}
using MLJ, DataFrames, CSV
n = 1000
p = 2

using Random
Random.seed!(42)

# Linearly separable:
X, y = make_blobs(n, p; centers=2, center_box=(-2 => 2), cluster_std=0.1)
df = DataFrame(X)
df.target .= ifelse.(y.==1,0,1)
CSV.write(joinpath(data_path, "linearly_separable.csv"),df)

# Overlapping:
X, y = make_blobs(n, p; centers=2, center_box=(-2 => 2), cluster_std=0.5)
df = DataFrame(X)
df.target .= ifelse.(y.==1,0,1)
CSV.write(joinpath(data_path, "overlapping.csv"),df)

# Circles:
X, y = make_circles(n; noise=0.15, factor=0.01)
df = DataFrame(X)
df.target = y
CSV.write(joinpath(data_path, "circles.csv"),df)

# Moon:
X, y = make_moons(n)
df = DataFrame(X)
df.target = y
CSV.write(joinpath(data_path, "moons.csv"),df)
```

### 2022-08-17 paper/appendix.qmd
sorted out deps and set up online appendix

---
title: Supplementary Appendix
format:
    html: 
        standalone: true
        code-fold: true
        execute:
            echo: true
            eval: false
            warning: false
    gfm:
        wrap: none
        execute:
            echo: false
            eval: false
            warning: false
jupyter: julia-1.7
---

This is a supplementary appendix to the research paper **Endogenous Macrodynamics in Algorithmic Recourse**. It contains all of the experimental results, including those not highlighted in the actual paper. It also contains additional information about the proposed counterfactual generators.

# Experimental Results {#sec-results}

{{< include ../dev/notebooks/experiments/synthetic.qmd >}}

{{< include ../dev/notebooks/experiments/real_world.qmd >}}

{{< include ../dev/notebooks/experiments/mitigation_strategies.qmd >}}

# Generators {#sec-generators}

{{< include ../dev/notebooks/generators/gravitational_generator.qmd >}}

{{< include ../dev/notebooks/generators/clap_roar_generator.qmd >}}

### 2022-08-17 paper/appendix.qmd
uh

@@
     html: 
-        standalone: true
+        self-contained: true
         code-fold: true
@@
             echo: true
-            eval: false
-            warning: false
-    gfm:
-        wrap: none
-        execute:
-            echo: false
             eval: false

### 2022-08-17 README.qmd
work on online appendix. Just need to finish mitigation strats and sort out tables.

@@
 - [Notebooks](dev/notebooks/)
+- [Supplementary Appendix](build/dev/notebooks/appendix.html) (download the HTML and view in browser)
 - [Artifacts](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/releases/tag/artifacts) (including data and experimental results)

### 2022-08-17 dev/notebooks/appendix.qmd
work on online appendix. Just need to finish mitigation strats and sort out tables.

--- 2022-08-17 paper/appendix.qmd
@@
 
-{{< include ../dev/notebooks/experiments/synthetic.qmd >}}
+{{< include experiments/synthetic.qmd >}}
 
-{{< include ../dev/notebooks/experiments/real_world.qmd >}}
+{{< include experiments/real_world.qmd >}}
 
-{{< include ../dev/notebooks/experiments/mitigation_strategies.qmd >}}
+{{< include experiments/mitigation_strategies.qmd >}}
 
@@
 
-{{< include ../dev/notebooks/generators/gravitational_generator.qmd >}}
+{{< include generators/gravitational_generator.qmd >}}
 
-{{< include ../dev/notebooks/generators/clap_roar_generator.qmd >}}
+{{< include generators/clap_roar_generator.qmd >}}

### 2022-08-17 dev/notebooks/experiments/mitigation_strategies.qmd
sorted out deps and set up online appendix
work on online appendix. Just need to finish mitigation strats and sort out tables.

--- 2022-08-17 dev/notebooks/mitigation_strategies.qmd
@@
----
-title: Mitigation Strategies
-jupyter: julia-1.7
----
+## Mitigation Strategies
 
 ```{julia}
+#| eval: true
 using Pkg; Pkg.activate("dev")
@@
 ```{julia}
+#| eval: true
 include("dev/utils.jl")
@@
 
-## Synthetic
+### Synthetic
 

### 2022-08-17 dev/notebooks/experiments/real_world.qmd
sorted out deps and set up online appendix
work on online appendix. Just need to finish mitigation strats and sort out tables.

--- 2022-08-17 dev/notebooks/real_world.qmd
@@
----
-title: Experiments with Credit Data
-jupyter: julia-1.7
----
+## Real-World Data
 
 ```{julia}
+#| eval: true
 using Pkg; Pkg.activate("dev")
@@
 output_path = output_dir("real_world")
-www_path = www_dir("real_world")
+www_path = www_dir("real_world");
 ```
@@
 
-## Running Experiment
+### Experiment
 
@@
 ```
+
+### Plots
 
@@
 
+#### Line Charts
+
+@fig-real-line shows the evolution of the evaluation metrics over the course of the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Credit Default"
+#|      - "GMSC"
+#| layout-ncol: 1
+#| label: fig-real-line
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+#### Error Bar Charts
+
+@fig-real-error shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Credit Default"
+#|      - "GMSC"
+#| layout-ncol: 1
+#| label: fig-real-error
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
 ### Chart in paper
+
+@fig-real-paper shows the chart that went into the paper.
 
@@
 ```
+
+```{julia}
+#| label: fig-real-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_real_world_results.png"))
+```

### 2022-08-17 dev/notebooks/experiments/synthetic.qmd
sorted out deps and set up online appendix
uh
work on online appendix. Just need to finish mitigation strats and sort out tables.

--- 2022-08-17 dev/notebooks/synthetic.qmd
@@
----
-title: Experiments with Synthetic Data
-jupyter: julia-1.7
----
+## Synthetic Data {#sec-app-synthetic}
+
+This notebook was used to run the experiments for the synthetic datasets. In the following we first run the experiments and then generate visualizations and tables.
+
+### Experiments
 
 ```{julia}
+#| eval: true
 using Pkg; Pkg.activate("dev")
@@
 ```{julia}
+#| eval: true
 include("dev/utils.jl")
 using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
+using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra, Images
 theme(:wong)
 output_path = output_dir("synthetic")
-www_path = www_dir("synthetic")
+www_path = www_dir("synthetic");
 ```
@@
 
+@fig-test-before shows the test data before running the experiment.
+
+```{julia}
+#| label: fig-test-before
+#| fig-cap: "Test data before experiment"
+#| eval: true
+load(joinpath(www_path,"models_test_before.png"))
+```
+
 ```{julia}
@@
 
-## Strict convergence
+@fig-train-before shows the training data before running the experiment.
+
+```{julia}
+#| label: fig-train-before
+#| fig-cap: "Training data before experiment"
+#| eval: true
+load(joinpath(www_path,"models_train_before.png"))
+```
 
@@
     plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_test_after_$(name)_strict.png"))
+    savefig(plt, joinpath(www_path,"models_test_after_$(name).png"))
+end
+```
+
+@fig-test-after shows the test data after running the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Test data after experiment"
+#| fig-subcap: 
+#|      - "DICE"
+#|      - "Generic"
+#|      - "Greedy"
+#|      - "Latent"
+#| layout-ncol: 1
+#| label: fig-test-after
+img_files = readdir(www_path)[contains.(readdir(www_path),"models_test_after")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
 end
@@
 
+@fig-train-after shows the training data after running the experiment.
+
 ```{julia}
+#| eval: true
+#| fig-cap: "Training data after experiment"
+#| fig-subcap: 
+#|      - "DICE"
+#|      - "Generic"
+#|      - "Greedy"
+#|      - "Latent"
+#| layout-ncol: 1
+#| label: fig-train-after
+img_files = readdir(www_path)[contains.(readdir(www_path),"models_train_after")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+### Plots
+
+```{julia}
+#| eval: true
 using Serialization
@@
 ```
-
-### Plots
 
@@
 
-### Chart in paper
+#### Line Charts
 
-#### Line Chart
+@fig-line shows the evolution of the evaluation metrics over the course of the experiment.
 
 ```{julia}
-using DataFrames, Statistics
-df = results[:overlapping].output
-# df = df[df.n .== maximum(df.n),:]
-gdf = groupby(df, [:generator, :model, :n, :name, :scope])
-df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
-df_plot = df_plot[[name in [:decisiveness, :disagreement, :mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
-df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
-df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
-transform!(df_plot, :name => (X -> [x=="decisiveness" ? "Decisiveness" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="disagreement" ? "Disagreement" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
-transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
-transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
-
-ncol = length(unique(df_plot.model))
-nrow = length(unique(df_plot.name))
-
-using RCall
-scale_ = 1.5
-R"""
-library(ggplot2)
-plt <- ggplot($df_plot, aes(x=n, y=mean, ymin=ymin, ymax=ymax, color=generator)) +
-    geom_ribbon(aes(fill=generator), alpha=0.5) +
-    geom_line() +
-    facet_grid(
-        rows = vars(name),
-        cols =  vars(model), 
-        scales = "free_y"
-    ) +
-    labs(y = "Value") + 
-    scale_fill_discrete(name="Generator:") +
-    scale_colour_discrete(name="Generator:") +
-    theme(
-        axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(),
-        legend.position="bottom"
-    )
-temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
-"""
-
-img = Images.load(rcopy(R"temp_path"))
-Images.save(joinpath(www_path,"paper_synthetic_results_line.png"), img)
+#| eval: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-line
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
 ```
 
-#### Errorbar
+#### Error Bar Charts
+
+@fig-error shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-error
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+### Tables
+
+@tbl-results shows a summary of all results.
+
+```{julia}
+#| label: tbl-results
+#| tbl-cap: "Summary of all results"
+#| eval: true
+#| results: asis
+using AlgorithmicRecourseDynamics: kable
+kable(results, [50]; format="html")
+```
+
+### Chart in paper {#sec-app-synthetic-paper}
+
+@fig-paper shows the chart that went into the paper.
 
@@
 ```
+
+```{julia}
+#| label: fig-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_synthetic_results.png"))
+```

### 2022-08-17 paper/paper.tex
work on online appendix. Just need to finish mitigation strats and sort out tables.

@@
             pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
-            pdfkeywords={keyword 1, keyword 2},
+            pdfkeywords={Algorithmic Recourse, Counterfactual Explanations, Explainable AI, Dynamic Systems},
             pdfborder={0 0 0},
@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
 \begin{IEEEkeywords}
-keyword 1; keyword 2
+Algorithmic Recourse; Counterfactual Explanations; Explainable AI; Dynamic Systems
 \end{IEEEkeywords}

### 2022-08-17 README.qmd
bleh

@@
 
-# AlgorithmicRecourseDynamics.jl - Research Paper 📝
+# AlgorithmicRecourseDynamics
+
+<!-- [![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/stable) -->
+<!-- [![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/dev) -->
+[![Build Status](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml?query=branch%3Amain)
+<!-- [![Coverage](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl) -->
+
+`AlgorithmicRecourseDynamics.jl` is a Julia package for modelling Algorithmic Recourse Dynamics.
+
+## Research Paper 📝
 

### 2022-08-18 dev/notebooks/appendix.qmd
just missing the tables now

@@
             warning: false
+        toc: true
 jupyter: julia-1.7

### 2022-08-18 dev/notebooks/experiments/mitigation_strategies.qmd
just missing the tables now

@@
 using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
+using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra, Images
 theme(:wong)
 output_path = output_dir("mitigation_strategies")
-www_path = www_dir("mitigation_strategies")
+www_path = www_dir("mitigation_strategies");
 ```
@@
 ```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plts = []
-for (exp_name, exp_) in experiments
-    for (M_name, M) in exp_.models
-        score = round(model_evaluation(M, exp_.test_data),digits=2)
-        plt = plot(M, exp_.test_data, title="$exp_name;\n $M_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
-        x_wrongly_labelled = exp_.test_data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plts = vcat(plts..., plt)
-    end
-end
-plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-savefig(plt, joinpath(www_path,"models_test_before.png"))
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plts = []
-for (exp_name, exp_) in experiments
-    for (M_name, M) in exp_.models
-        score = round(model_evaluation(M, exp_.train_data),digits=2)
-        plt = plot(M, exp_.train_data, title="$exp_name;\n $M_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, exp_.train_data.X)) .!= exp_.train_data.y))
-        x_wrongly_labelled = exp_.train_data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plts = vcat(plts..., plt)
-    end
-end
-plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-savefig(plt, joinpath(www_path,"models_train_before.png"))
-```
-
-```{julia}
 n_evals = 5
@@
 
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        score = round(model_evaluation(M, exp_.test_data),digits=2)
-        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
-        x_wrongly_labelled = exp_.test_data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_test_after_$(name).png"))
-end
-```
-
-```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
-plot_dict = Dict(key => Dict() for (key,val) in results)
-fold = 1
-for (name, res) in results
-    exp_ = res.experiment
-    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
-    rec_sys = exp_.recourse_systems[fold]
-    sys_ids = collect(exp_.system_identifiers)
-    M = length(rec_sys)
-    for m in 1:M
-        model_name, generator_name = sys_ids[m]
-        M = rec_sys[m].model
-        data = rec_sys[m].data
-        score = round(model_evaluation(M, data),digits=2)
-        plt = plot(M, data, title="$name;\n $model_name ($score)")
-        # Errors:
-        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
-        x_wrongly_labelled = data.X[:,ids]
-        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
-        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
-    end
-end
-plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
-for (name, plts) in plot_dict
-    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
-    savefig(plt, joinpath(www_path,"models_train_after_$(name).png"))
-end
-```
+### Plots
 
@@
 
+#### Line Charts
+
+@fig-mit-line shows the evolution of the evaluation metrics over the course of the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Circles"
+#|      - "Credit Default"
+#|      - "GMSC"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-mit-line
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart") .&& .!contains.(readdir(www_path),"latent")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+#### Error Bar Charts
+
+@fig-mit-error shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Circles"
+#|      - "Credit Default"
+#|      - "GMSC"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-mit-error
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart") .&& .!contains.(readdir(www_path),"latent")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
 ### Chart in paper
+
+@fig-mit-paper shows the chart that went into the paper.
 
@@
 
+```{julia}
+#| label: fig-mit-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_synthetic_results.png"))
+```
+
 ### Latent Space Search
@@
 
+### Plots
+
+#### Line Charts
+
+@fig-mit-line-latent shows the evolution of the evaluation metrics over the course of the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-mit-line-latent
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart") .&& contains.(readdir(www_path),"latent")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+#### Error Bar Charts
+
+@fig-mit-error-latent shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-mit-error-latent
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart") .&& contains.(readdir(www_path),"latent")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
 ### Chart in paper
+
+@fig-mit-latent-paper shows the chart that went into the paper.
 
@@
 Images.save(joinpath(www_path,"paper_synthetic_latent_results.png"), img)
+```
+
+```{julia}
+#| label: fig-mit-latent-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_synthetic_latent_results.png"))
 ```
@@
 
+@fig-mit-latent-paper shows the chart that went into the paper.
+
 ```{julia}
@@
 ```
+
+```{julia}
+#| label: fig-mit-real-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_real_world_results.png"))
+```

### 2022-08-18 dev/notebooks/experiments/synthetic.qmd
just missing the tables now

@@
 using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+results = Serialization.deserialize(joinpath(output_path,"results.jls"));
 ```
@@
 
-### Tables
-
-@tbl-results shows a summary of all results.
+### Bootstrap
 
 ```{julia}
-#| label: tbl-results
-#| tbl-cap: "Summary of all results"
+n_bootstrap = 1
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap.html")
+R"""
+dt <- DT::datatable($df) |>
+    DT::formatRound(columns=c("value"), digits=3)
+DT::saveWidget(dt, $save_path)
+"""
+```
+
+```{julia}
 #| eval: true
-#| results: asis
-using AlgorithmicRecourseDynamics: kable
-kable(results, [50]; format="html")
+using Gumbo
+save_path = joinpath(output_path, "bootstrap.html")
+parsehtml(read(save_path, String))
 ```

### 2022-08-18 dev/notebooks/experiments/mitigation_strategies.qmd
cant get tables to work, charts will have to do

@@
 end
+```
+
+### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap_synthetic.csv")
+using CSV
+CSV.write(save_path)
 ```
@@
 
+### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap_latent.csv")
+using CSV
+CSV.write(save_path)
+```
+
 ### Chart in paper
@@
 
+### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap_real_world.csv")
+using CSV
+CSV.write(save_path)
+```
+
 ### Chart in paper

### 2022-08-18 dev/notebooks/experiments/real_world.qmd
cant get tables to work, charts will have to do

@@
 
+### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap.csv")
+using CSV
+CSV.write(save_path)
+```
+
 ### Chart in paper

### 2022-08-18 dev/notebooks/experiments/synthetic.qmd
cant get tables to work, charts will have to do

--- 2022-08-17 dev/notebooks/experiments/synthetic.qmd
@@
 
-### Tables
-
-@tbl-results shows a summary of all results.
+### Bootstrap
 
 ```{julia}
-#| label: tbl-results
-#| tbl-cap: "Summary of all results"
-#| eval: true
-#| results: asis
-using AlgorithmicRecourseDynamics: kable
-kable(results, [50]; format="html")
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap.csv")
+using CSV
+CSV.write(save_path)
 ```

### 2022-08-18 dev/notebooks/generators/clap_roar_generator.qmd
sorted out deps and set up online appendix
uh
cant get tables to work, charts will have to do

--- 2022-08-11 dev/notebooks/endo_roar_generator.qmd
@@
----
-title: EndoROAR Generator
-jupyter: julia-1.7
----
+## ClapROAR Generator
 
@@
 ```
-
-# `ClapROARGenerator`
 
@@
 plt = plot(plts..., size=(800,300), layout=(1,2))
+display(plt)
 ```
-
-```{julia}
-
-```

### 2022-08-18 paper/sections/empirical.rmd
bleh
cant get tables to work, charts will have to do

@@
 We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.].
-In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concered with generating what we have defined as meaningful counterfactuals.
+In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.
 
@@
 Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
-In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis.
+In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been omitted from the analysis.
 See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
@@
 Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 2,500 individuals each.
-We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.
+We have also standardized all explanatory features since our chosen classifiers are sensitive to scale.
 

### 2022-08-18 paper/sections/empirical_2.rmd
cant get tables to work, charts will have to do

@@
 For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
-For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.].
+For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.].
 Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn).

### 2022-08-18 paper/sections/introduction.rmd
cant get tables to work, charts will have to do

@@
 Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction.
-Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more liekly to fail their degree than students that were admitted in previous years.
+Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years.
 The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse.

### 2022-08-18 paper/sections/mitigation.rmd
cant get tables to work, charts will have to do

@@
 
-Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrent our attention, in this section we introduce and test a number of potential mitigation strategies.
+Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we introduce and test a number of potential mitigation strategies.
 All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
@@
 
-Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
+Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
 

### 2022-08-18 paper/sections/related.rmd
cant get tables to work, charts will have to do

@@
 An October 2020 survey by Karimi et al.
-layed out 60 algorithms that have been proposed since 2014 [@karimi2020survey].
+laid out 60 algorithms that have been proposed since 2014 [@karimi2020survey].
 Another survey published around the same time by Verma et al.
@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.

### 2022-08-18 paper/paper.tex
fully anonymised

@@
 Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017.
-An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 \protect\hyperlink{ref-karimi2020survey}{{[}10{]}}.
+An October 2020 survey by Karimi et al.~laid out 60 algorithms that have been proposed since 2014 \protect\hyperlink{ref-karimi2020survey}{{[}10{]}}.
 Another survey published around the same time by Verma et al.~described 29 algorithms \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}.
@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
@@
 We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.}.
-In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concered with generating what we have defined as meaningful counterfactuals.
+In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.
 
@@
 Secondly, we use the \textbf{UCI defaultCredit} dataset \protect\hyperlink{ref-yeh2009comparisons}{{[}32{]}}, a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
-In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes\footnote{These have been ommitted from the analysis.
+In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes\footnote{These have been omitted from the analysis.
 See Section \ref{limit-data} for details.} and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
@@
 For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
-For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.}.
+For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.}.
 Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \ref{fig:syn}.
@@
 
-Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
+Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
 
@@
 
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+Granular results for all of our experiments can be found in this supplementary \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/build/dev/notebooks/appendix.html}{appendix}.
+
 \end{document}

### 2022-08-18 docs/make.jl

@@ -5,7 +5,7 @@ DocMeta.setdocmeta!(AlgorithmicRecourseDynamics, :DocTestSetup, :(using Algorith
 
 makedocs(;
     modules=[AlgorithmicRecourseDynamics],
-    authors="Patrick Altmeyer",
+    authors="Anonymous",
     repo="https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/{commit}{path}#{line}",
     sitename="AlgorithmicRecourseDynamics.jl",
     format=Documenter.HTML(;

### 2022-08-18 paper/paper.tex
Arie's final round of comments

@@
 \newtheorem{corollary}{Corollary}[section]
-\newtheorem{proposition}{Proposition}[section]
+\newtheorem{proposition}{Research Question}[section]
 \newtheorem{conjecture}{Conjecture}[section]
@@
 To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
-Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups.
-Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}.
-To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators.
-Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.
-
-The remainder of the paper is structured as follows: Section \ref{related} places our work in the broader context of related literature.
-Section \ref{method} posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
-Section \ref{method-2} sets out our experimental framework for modeling endogenous macrodynamics in AR.
-Section \ref{empirical} presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
-Section \ref{empirical-2} presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
-Our findings are then discussed in the broader context of the literature in Section \ref{discussion}, before pointing to some of the limitations of our work as well as avenues for future research in Section \ref{limit}.
-Finally, Section \ref{conclusion} concludes.
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \ref{related} and \ref{method}).
+Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\ref{method-2}).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (\ref{empirical} and \ref{empirical-2}).
+Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \ref{mitigate}).
+Finally, we discuss our findings in the broader context of the literature in Section \ref{discussion}, before pointing to some of the limitations of our work as well as avenues for future research in Section \ref{limit}.
+Section \ref{conclusion} concludes.
 
@@
 In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
-We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}} - the Julia package we introduced earlier.
+We introduce this framework with the ambition to shed light on the following research questions:
+
+\begin{proposition}[Endogenous Shifts]
+\protect\hypertarget{prp:shifts}{}\label{prp:shifts}Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
+\end{proposition}
+
+\begin{proposition}[Costs]
+\protect\hypertarget{prp:costs}{}\label{prp:costs}If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+\end{proposition}
+
+\begin{proposition}[Heterogeneity]
+\protect\hypertarget{prp:het}{}\label{prp:het}Do different counterfactual generators yield significantly different outcomes in this context?
+Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
+\end{proposition}
+
+\begin{proposition}[Drivers]
+\protect\hypertarget{prp:drive}{}\label{prp:drive}What are drivers of endogenous dynamics in Algorithmic Recourse?
+\end{proposition}
+
+Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}} - the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
@@
 
-Using the framework introduced in the previous section we are interested in answering the following research questions in this section:
-
-\begin{itemize}
-\tightlist
-\item
-  Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
-\item
-  If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
-\item
-  Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
-\item
-  What seem to be drivers of endogenous dynamics in Algorithmic Recourse?
-\end{itemize}
-
 Below we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.
@@
 
-We start this section off with a few high-level observations.
+We start this section off with the key high-level observations.
 Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:
@@
 
-To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments and we find them substantial enough to be considered costly; secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect.
-The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process.
+To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments (RQ \ref{prp:shifts}) and we find them substantial enough to be considered costly (RQ \ref{prp:costs}); secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect (RQ \ref{prp:het}).
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process (RQ \ref{prp:drive}).
 
 \hypertarget{mitigate}{%
-\section{Potential Mitigation Strategies}\label{mitigate}}
-
-Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we introduce and test a number of potential mitigation strategies.
+\section{Mitigation Strategies and Experiments}\label{mitigate}}
+
+Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we ask ourselves:
+
+\begin{proposition}[Mitigation Strategies]
+\protect\hypertarget{prp:mitigate}{}\label{prp:mitigate}What are potential mitigation strategies with respect to endogenous macrodynamics in AR?
+\end{proposition}
+
+We proposed and test a number of simple mitigation strategies.
 All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
@@
 By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
-A potential drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
+A drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
 
@@
 
-Another potential strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
+Another strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
 Recall that \(\text{extcost}(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
@@
 Higher choices of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to gravitate towards the specified point \(\bar{x}\) in the target domain.
-In the remainder of this paper we will therefore refer to this approach as \textbf{Gravitational} generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics\footnote{Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.}.
+In the remainder of this paper we will therefore refer to this approach as \textbf{Gravitational} generator, when we investigate its usefulness for mitigating endongenous macrodynamics\footnote{Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.}.
 

### 2022-08-18 paper/sections/empirical_2.rmd
Arie's final round of comments

@@
 # Experiments {#empirical-2}
-
-Using the framework introduced in the previous section we are interested in answering the following research questions in this section:
-
-- Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
-- If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
-- Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
-- What seem to be drivers of endogenous dynamics in Algorithmic Recourse?
 
@@
 
-To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments and we find them substantial enough to be considered costly; secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect.
-The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process.
+To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments (RQ \@ref(prp:shifts)) and we find them substantial enough to be considered costly (RQ \@ref(prp:costs)); secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect (RQ \@ref(prp:het)).
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process (RQ \@ref(prp:drive)).

### 2022-08-18 paper/sections/introduction.rmd
Arie's final round of comments

@@
 To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
-Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups.
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized.
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting.
-To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators.
-Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.
-
-The remainder of the paper is structured as follows: Section \@ref(related) places our work in the broader context of related literature.
-Section \@ref(method) posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs.
-Section \@ref(method-2) sets out our experimental framework for modeling endogenous macrodynamics in AR.
-Section \@ref(empirical) presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed.
-Section \@ref(empirical-2)  presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies.
-Our findings are then discussed in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit).
-Finally, Section \@ref(conclusion) concludes.
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \@ref(related) and \@ref(method)).
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)).
+Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \@ref(mitigate)).
+Finally, we discuss our findings in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit).
+Section \@ref(conclusion) concludes.

### 2022-08-18 paper/sections/methodology_2.rmd
Arie's final round of comments

@@
 In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
-We first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) - the Julia package we introduced earlier.
+We introduce this framework with the ambition to shed light on the following research questions:
+
+::: {.proposition #shifts name="Endogenous Shifts"}
+Does the repeated implementation of recourse provided by state-of-the-art generators lead to shifts in the domain and model?
+:::
+
+::: {.proposition #costs name="Costs"}
+If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+:::
+
+::: {.proposition #het name="Heterogeneity"}
+Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
+:::
+
+::: {.proposition #drive name="Drivers"}
+What are drivers of endogenous dynamics in Algorithmic Recourse?
+:::
+
+Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) - the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.

### 2022-08-18 paper/sections/mitigation.rmd
Arie's final round of comments

@@
-# Potential Mitigation Strategies {#mitigate}
+# Mitigation Strategies and Experiments {#mitigate}
 
-Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we introduce and test a number of potential mitigation strategies.
+Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we ask ourselves:
+
+::: {.proposition #mitigate name="Mitigation Strategies"}
+What are potential mitigation strategies with respect to endogenous macrodynamics in AR?
+:::
+
+We proposed and test a number of simple mitigation strategies.
 All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
@@
 
-Another potential strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
+Another strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
 Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
@@
 Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain.
-In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
 

### 2022-08-19 paper/paper.Rmd
cant get tables to work, charts will have to do
fully anonymised
incorporated Cynthia's comments

@@
 abstract: |
-  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
 The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
@@
 Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources.
+We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve competition for scarce resources.
 Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
@@
 :::
+
+# Appendix
+
+Granular results for all of our experiments can be found in this supplementary [appendix](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/build/dev/notebooks/appendix.html).

### 2022-08-19 paper/paper.tex
incorporated Cynthia's comments

@@
 \begin{abstract}
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
 The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
@@
 Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences.
-While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making.
+While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise to a new type of problem in the context of data-driven decision-making.
 They are essentially \textbf{black boxes}: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools.
-Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}.
+Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes \protect\hyperlink{ref-oneil2016weapons}{{[}1{]}}.
 The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have investigated different ways to explain the behaviour of black-box models.
@@
 A meaningful recourse recommendation for a denied applicant could be: \emph{``If your net savings rate had been 10\% of your monthly income instead of the actual 8\%, your application would have been successful.
-See if you can temporarily cut down on consumption.''} In the remainder of this paper we will use both terminologies - recourse and counterfactual - interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
+See if you can temporarily cut down on consumption.''} In the remainder of this paper we will use both terminologies---recourse and counterfactual---interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
 
@@
 We think that these types of endogenous dynamics may be problematic and warrant our attention.
-From a purely technical we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
 Secondly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
-We would argue that in some practical applications this can be expected to generate costs for the involved stakeholders.
+We would argue that in some practical applications, this can be expected to generate costs for the involved stakeholders.
 To follow our argument, consider the following two examples:
@@
 \begin{example}[Consumer Credit]
-\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their credit worthiness.
+\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness.
 Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction.
@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \ref{related} and \ref{method}).
-Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\ref{method-2}).
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (\ref{empirical} and \ref{empirical-2}).
+Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\ref{method-2}).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (\ref{empirical} and \ref{empirical-2}).
 Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \ref{mitigate}).
@@
 
-In this Section we provide a review of the relevant literature.
+In this section we provide a review of the relevant literature.
 First, Subsection \ref{related-recourse} discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
@@
 Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\).
-Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):
+Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x^\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x^\prime\) such that the predicted outcome \(M(x^\prime)\) corresponds to some target outcome \(y^*\):
 
 \begin{equation}
-\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
+\min_{x^\prime \in \mathcal{X}} \text{cost}(x^\prime) \ \ \ \mbox{s. t.} \ \ \ M(x^\prime) = y^* \label{eq:obj}
 \end{equation}
@@
 \begin{equation}
-x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) \label{eq:solution}
+x^\prime = \arg \min_{x^\prime}  \text{yloss}(M(x^\prime),y^*) + \lambda \text{cost}(x^\prime) \label{eq:solution}
 \end{equation}
@@
 For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent.
-This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.
+This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we shall refer to simply as \textbf{Wachter}, the name of the first author, in the remainder of this paper.
 
@@
 Another survey published around the same time by Verma et al.~described 29 algorithms \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}.
-Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met.
+Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how they ensure that certain requirements for CE are met.
 Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs.
-To name but a few, we are interested in generating counterfactuals that close \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible causally founded \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
+To name but a few, we are interested in generating counterfactuals that are close \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible causally founded \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
 
@@
 The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}.
-We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+We have been able to identify only one recent work by Upadhyay et al.~that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 Exogenous shifts are strictly of external origin.
-For example, they might stem from data correction, temporal shifts or geospatial changes \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
-The authors of \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
+For example, they might stem from data correction, temporal shifts or geospatial changes \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
+Upadhyay et al.
+\protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}} propose ROAR: a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
 
@@
 Numerous approaches to anomaly detection have been summarized \protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}.
-Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}.
+Another well-established research topic is that of concept drift: situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}.
 For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
-Less previous work is available on the related topic of model drift - changes in model performance over time.
+Less previous work is available on the related topic of model drift: changes in model performance over time.
 Nelson et al.~review how resistant different machine learning models are to the model drift \protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}.
@@
 
-In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\).
-While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these perturbations are mechanically introduced.
-In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
+While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}} these perturbations are mechanically introduced.
+In contrast, we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
 In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model.
@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, \protect\hyperlink{ref-deoliveira2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
@@
 
-In this section we first set out a generalized framework for gradient-based counterfactual search in Section \ref{method-general} that encapsulates the various individual recourse methods we have chosen to use in our experiments.
-We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.
+In this section we first set out a generalized framework for gradient-based counterfactual search that encapsulates the various individual recourse methods we have chosen to use in our experiments (Section \ref{method-general}).
+We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective (Section \ref{method-collective}).
 
@@
 
-We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
-In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
+We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
+In particular, we include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
 Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
@@
 The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\).
-In this case \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.
+In this case, \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}.
 
 CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\).
-To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
-The approach proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
+To quantify predictive uncertainty, Antoran et al.
+\protect\hyperlink{ref-antoran2020getting}{{[}9{]}} rely on entropy estimates for probabilistic models.
+The greedy approach proposed by Schut et al.
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
 The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
 Schut et al.
-\protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE.
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\), we also implicitly minimize predictive entropy (as in CLUE).
 In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
@@
 , s_K\) that look as different from each other as possible.
-The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach.
+The implementation of DiCE in our library of choice---\texttt{CounterfactualExplanations.jl}---uses that exact approach.
 It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.
 
-\hypertarget{towards-collective-recourse}{%
-\subsection{\ldots{} towards collective recourse}\label{towards-collective-recourse}}
+\hypertarget{method-collective}{%
+\subsection{\ldots{} towards collective recourse}\label{method-collective}}
 
@@
 
-In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse.
+In the following we describe the framework we propose for modeling and analyzing endogenous macrodynamics in Algorithmic Recourse.
 We introduce this framework with the ambition to shed light on the following research questions:
@@
 
-Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}} - the Julia package we introduced earlier.
+Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}---the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
@@
 We begin in the static setting at time time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse\footnote{To keep notation simple we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse\footnote{To keep notation simple, we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.
 
@@
 To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population.
-The experimental setup is summarized in Algorithm \ref{algo-experiment}
+The experimental setup is summarized in Algorithm \ref{algo-experiment}.
 
@@
 The framework we have built facilitates this, making use of multi-threading in order to speed up computations.
-Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
+Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
 To ensure that each generator is faced with the exact same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.
@@
 An important consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
-In our implementation we make use of Gaussian kernel with a constant length-scale parameter of \(0.5\).
+In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of \(0.5\).
 As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD(X,\tilde{X})=0\) if and only if \(X=\tilde{X}\).
@@
 Then, we split the permuted data into two new samples \(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the original samples.
-Then under the null hypothesis we should have that \(MMD(X^\prime,\tilde{X}^\prime)\) be approximately equal to \(MMD(X,\tilde{X})\).
+Then under the null hypothesis, we should have that \(MMD(X^\prime,\tilde{X}^\prime)\) be approximately equal to \(MMD(X,\tilde{X})\).
 The corresponding \(p\)-value can then be calculated by counting how these two quantities are not equal.
@@
 
-As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 We define \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance between the vectors of parameters before and after retraining the model \(M\).
@@
 
-As an alternative to PP MMD we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement).
-This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+As an alternative to PP MMD, we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement).
+This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise each individual experiment is repeat five times.
+To account for noise, each individual experiment is repeat five times.
 
 \hypertarget{empirical-classifiers}{%
-\subsection{\texorpdfstring{\(M\) -- Classifiers and Generative Models}{M -- Classifiers and Generative Models}}\label{empirical-classifiers}}
-
-For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier.
+\subsection{\texorpdfstring{\(M\)---Classifiers and Generative Models}{M---Classifiers and Generative Models}}\label{empirical-classifiers}}
+
+For each dataset and generator, we look at three different types of classifiers, all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier---\textbf{Logistic Regression}---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier.
 We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
@@
 \hypertarget{empirical-data}{%
-\subsection{\texorpdfstring{\(\mathcal{D}\) -- Data}{\textbackslash mathcal\{D\} -- Data}}\label{empirical-data}}
+\subsection{\texorpdfstring{\(\mathcal{D}\)---Data}{\textbackslash mathcal\{D\}---Data}}\label{empirical-data}}
 
@@
 Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
-Following \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}, we generate synthetic data in \(\mathbb{R}^2\) to also allow for a visual interpretation of the results.
+Following \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}, we generate synthetic data in \(\mathbb{R}^2\) to also allow for a visual interpretation of the results.
 Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
@@
 The datasets are presented in Figure \ref{fig:synthetic-data}.
-Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
 
@@
 
-Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
+Ex-ante we expect to see that by construction, Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary as we saw in Figure \ref{fig:poc}.
 Thus, the choice of a black-box model may have an impact on the paths of the recourse.
-For generators that use latent space search (REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}) or rely on (and have access to) probabilistic models (CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
-Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
+For generators that use latent space search (REVISE \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}) or rely on (and have access to) probabilistic models (CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+Of course, this expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
 Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class\footnote{As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated.
@@
 We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification.
-Firstly, we use the \textbf{Give Me Some Credit} dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years \protect\hyperlink{ref-gmsc_data}{{[}31{]}}.
-Originally consisting of 250,000 instances with 11 numerical attributes.
-Secondly, we use the \textbf{UCI defaultCredit} dataset \protect\hyperlink{ref-yeh2009comparisons}{{[}32{]}}, a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
-In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes\footnote{These have been omitted from the analysis.
+Firstly, we use the \textbf{Give Me Some Credit} dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years \protect\hyperlink{ref-kaggle2011give}{{[}31{]}}, originally consisting of 250,000 instances with 11 numerical attributes.
+Secondly, we use the \textbf{UCI defaultCredit} dataset \protect\hyperlink{ref-yeh2009comparisons}{{[}32{]}}, a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable whether credit card clients default on their payment.
+In its raw form it consists of 23 explanatory variables: 4 categorical features relating to demographic attributes\footnote{These have been omitted from the analysis.
 See Section \ref{limit-data} for details.} and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
-Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}} and \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}} and \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
 
@@
 While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example \protect\hyperlink{ref-schut2021generating}{{[}6{]}}).
-While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}35{]}}.
+While we initially also conducted experiments on that dataset, we eventually discarded this dataset due to surrounding ethical concerns \protect\hyperlink{ref-carlisle2019racist}{{[}35{]}}.
 
@@
 
-\hypertarget{g-generators}{%
-\subsection{\texorpdfstring{\(G\) -- Generators}{G -- Generators}}\label{g-generators}}
-
-All generators introduced earlier are included in the experiments: Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, REVISE \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
+\hypertarget{ggenerators}{%
+\subsection{\texorpdfstring{\(G\)---Generators}{G---Generators}}\label{ggenerators}}
+
+All generators introduced earlier are included in the experiments: Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, REVISE \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, CLUE \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and Greedy \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
 In addition, we introduce two new generators in Section \ref{mitigate} that directly address the issue of endogenous domain and model shifts.
@@
 
-Below we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.
+Below, we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.
 
@@
 The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
-Related to this the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
+Related to this, the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
 We also note that the final classifiers are more decisive, although as we noted earlier this may to some extent just be a byproduct of retraining the model throughout the course of the experiment.
@@
 
-Turning to the real-world data we will go through the findings presented in Figure \ref{fig:real} where each column corresponds to one of the three data sets.
+Turning to the real-world data we will go through the findings presented in Figure \ref{fig:real}, where each column corresponds to one of the three data sets.
 The results shown here are for the deep ensemble, which once again largely resemble those for the MLP.
@@
 This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section.
-But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and - perhaps more importantly - their performance hinges on the performance of said generative models.
+But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and, perhaps more importantly, their performance hinges on the performance of said generative models.
 Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.
@@
 Under \(\gamma=0.5\), counterfactuals will end up near the decision boundary by construction.
-Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off.
+Since this is the region of maximal aleatoric uncertainty, the classifier is bound to be thrown off.
 By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
@@
 
-\hypertarget{classifier-preserving-roar}{%
-\subsection{Classifier Preserving ROAR}\label{classifier-preserving-roar}}
-
-Another strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
+\hypertarget{classifier-preserving-roar-claproar}{%
+\subsection{Classifier Preserving ROAR (ClaPROAR)}\label{classifier-preserving-roar-claproar}}
+
+Another strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
 Recall that \(\text{extcost}(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
@@
 for each counterfactual \(k\) where \(l\) denotes the loss function used to train \(M\).
-This approach, which we shall refer to as \textbf{ClapROAR}, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
-It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
+This approach, which we shall refer to as \textbf{ClaPROAR}, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
 It also makes the private vs.~external cost trade-off more explicit and hence manageable.
@@
 
-Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: \(\bar{x}\):
+Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average \(\bar{x}\):
 
@@
 
-Once again we can putting this in the context of Equation \eqref{eq:collective}, the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
+Once again we can putting this in the context of Equation \eqref{eq:collective}: the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
 Higher choices of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to gravitate towards the specified point \(\bar{x}\) in the target domain.
-In the remainder of this paper we will therefore refer to this approach as \textbf{Gravitational} generator, when we investigate its usefulness for mitigating endongenous macrodynamics\footnote{Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.}.
-
-Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
+In the remainder of this paper, we will therefore refer to this approach as \textbf{Gravitational} generator, when we investigate its usefulness for mitigating endogenous macrodynamics\footnote{Note that despite the naming conventions, our goal here is not to provide yet more counterfactual generators.
+Rather than looking at them as isolated entities, we believe and demonstrate that different approaches can be effectively combined.}.
+
+Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is, Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClaPROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Gravitational generator and ClaPROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
 
@@
 The top row in Figure \ref{fig:mitigate-results} indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
-ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
+ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
 As evident from the middle row in Figure \ref{fig:mitigate-results} model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
 The same overall pattern can be observed for out-of-sample model performance.
-With respect to the other synthetic datasets we found the following: for the Moons dataset the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.
+With respect to the other synthetic datasets, for the Moons dataset, the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.
 
@@
 
-An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators.
+In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
 In Figure \ref{fig:mitigate-latent-results} the conventional Latent Space generator with \(\gamma=0.5\) serves as our baseline.
@@
 Finally, Figure \ref{fig:mitigate-real-world-results} shows the results for our real-world data.
-We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \ref{fig:mitigate-results}\footnote{Estimated domain shifts (not shown) were largely insubstantial, as in Figure \ref{fig:real} in the previous section.}.
-Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
+We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both model shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in Figure \ref{fig:mitigate-results}\footnote{Estimated domain shifts (not shown) were largely insubstantial, as in Figure \ref{fig:real} in the previous section.}.
+Still, both ClaPROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
 For the GMSC dataset we observe no notable differences.
@@
 We would argue that the expected external costs of individual recourse should be shared by all stakeholders.
-The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \eqref{eq:collective}.
-This will on average lead to more costly counterfactual outcomes.
-But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
+The most straightforward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \eqref{eq:collective}.
+This will on average lead to more costly counterfactual outcomes, but may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
 We have shown various types of shift-mitigating strategies that can be used to this end.
@@
 The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework.
-An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is a collective recourse outcome in which no single individual can be made better off, without make at least one other individual worse off.
+An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is, a collective recourse outcome in which no single individual can be made better off, without making at least one other individual worse off.
 This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work.
-For the time being, we suggest that the trade-off can be resolved on a case-by-case basis.
 
@@
 The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion.
-In practice, models are in fact updated on a regular basis \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+In practice, models are in fact updated on a regular basis \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time \(t=0\).
 That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
-In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}.
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model \(M\) is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
@@
 For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks.
-While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}36{]}}, \protect\hyperlink{ref-grinsztajn2022tree}{{[}37{]}}).
+While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice.
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}36{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}37{]}}).
 
@@
 Through our work we hope to inspire future research on this important topic.
-To this end we have open-sourced all of our code along with a Julia package - \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}.
-The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}.
+To this end we have open-sourced all of our code along with a Julia package: \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}.
+The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}.
 That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-o2016weapons}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-oneil2016weapons}{}}%
 \CSLLeftMargin{{[}1{]} }%
-\CSLRightInline{C. O'neil, \emph{Weapons of math destruction: How big data increases inequality and threatens democracy}. Crown, 2016.}
+\CSLRightInline{C. O'Neil, \emph{Weapons of math destruction: {How} big data increases inequality and threatens democracy}. {Crown}, 2016.}
 
@@
 Richter, and G.
-Kasneci, {``Carla: A python library to benchmark algorithmic recourse and counterfactual explanation algorithms,''} \emph{arXiv preprint arXiv:2108.00783}, 2021.}
+Kasneci, {``Carla: A python library to benchmark algorithmic recourse and counterfactual explanation algorithms,''} 2021.
+Available: \url{https://arxiv.org/abs/2108.00783}}
 
@@
 Mittelstadt, and C.
-Russell, {``Counterfactual explanations without opening the black box: Automated decisions and the GDPR,''} \emph{Harv.
+Russell, {``Counterfactual explanations without opening the black box: {Automated} decisions and the {GDPR},''} \emph{Harv.
 JL \& Tech.}, vol.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022counterfactualexplanations}{}}%
 \CSLLeftMargin{{[}5{]} }%
 \CSLRightInline{P.
-Altmeyer, \emph{{CounterfactualExplanations.jl - a Julia package for Counterfactual Explanations and Algorithmic Recourse}}.
-2022.
+Altmeyer, {``{CounterfactualExplanations}.jl - a {Julia} package for {Counterfactual Explanations} and {Algorithmic Recourse}.''} 2022.
 Available: \url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
@@
 \CSLRightInline{L.
-Schut \emph{et al.}, {``Generating interpretable counterfactual explanations by implicit minimisation of epistemic and aleatoric uncertainties,''} in \emph{International conference on artificial intelligence and statistics}, 2021, pp.
+Schut \emph{et al.}, {``Generating {Interpretable Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and {Aleatoric Uncertainties},''} in \emph{International {Conference} on {Artificial Intelligence} and {Statistics}}, 2021, pp.
 1756--1764.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-joshi2019realistic}{}}%
 \CSLLeftMargin{{[}7{]} }%
@@
 Kim, and J.
-Ghosh, {``Towards realistic individual recourse and actionable explanations in black-box decision making systems,''} \emph{arXiv preprint arXiv:1907.09615}, 2019.}
+Ghosh, {``Towards realistic individual recourse and actionable explanations in black-box decision making systems,''} 2019.
+Available: \url{https://arxiv.org/abs/1907.09615}}
 
@@
 Sharma, and C.
-Tan, {``Explaining machine learning classifiers through diverse counterfactual explanations,''} in \emph{Proceedings of the 2020 conference on fairness, accountability, and transparency}, 2020, pp.
+Tan, {``Explaining machine learning classifiers through diverse counterfactual explanations,''} in \emph{Proceedings of the 2020 {Conference} on {Fairness}, {Accountability}, and {Transparency}}, 2020, pp.
 607--617.}
@@
 M.
-Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}
+Hernández-Lobato, {``Getting a clue: {A} method for explaining uncertainty estimates,''} 2020.
+Available: \url{https://arxiv.org/abs/2006.06848}}
 
@@
 Schölkopf, and I.
-Valera, {``A survey of algorithmic recourse: Definitions, formulations, solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050}, 2020.}
+Valera, {``A survey of algorithmic recourse: Definitions, formulations, solutions, and prospects,''} 2020.
+Available: \url{https://arxiv.org/abs/2010.04050}}
 
@@
 \CSLLeftMargin{{[}11{]} }%
-\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual explanations for machine learning: A review,''} \emph{arXiv preprint arXiv:2010.10596}, 2020.}
+\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual explanations for machine learning: {A} review,''} 2020. Available: \url{https://arxiv.org/abs/2010.10596}}
 
@@
 Spangher, and Y.
-Liu, {``Actionable recourse in linear classification,''} in \emph{Proceedings of the conference on fairness, accountability, and transparency}, 2019, pp.
+Liu, {``Actionable recourse in linear classification,''} in \emph{Proceedings of the {Conference} on {Fairness}, {Accountability}, and {Transparency}}, 2019, pp.
 10--19.}
@@
 Schölkopf, and I.
-Valera, {``Algorithmic recourse: From counterfactual explanations to interventions,''} in \emph{Proceedings of the 2021 ACM conference on fairness, accountability, and transparency}, 2021, pp.
+Valera, {``Algorithmic recourse: From counterfactual explanations to interventions,''} in \emph{Proceedings of the 2021 {ACM Conference} on {Fairness}, {Accountability}, and {Transparency}}, 2021, pp.
 353--362.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021robust}{}}%
 \CSLLeftMargin{{[}14{]} }%
-\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards robust and reliable algorithmic recourse,''} \emph{arXiv preprint arXiv:2102.13620}, 2021.}
+\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards {Robust} and {Reliable Algorithmic Recourse},''} 2021. Available: \url{https://arxiv.org/abs/2102.13620}}
 
@@
 \CSLLeftMargin{{[}15{]} }%
-\CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero, {``Generating collective counterfactual explanations in score-based classification via mathematical optimization,''} 2021.}
+\CSLRightInline{E. Carrizosa, J. Ramırez-Ayerbe, and D. Romero, {``Generating {Collective Counterfactual Explanations} in {Score-Based Classification} via {Mathematical Optimization},''} 2021.}
 
@@
 Günnemann, and Z.
-Lipton, {``Failing loudly: An empirical study of methods for detecting dataset shift,''} \emph{Advances in Neural Information Processing Systems}, vol.
+Lipton, {``Failing loudly: {An} empirical study of methods for detecting dataset shift,''} \emph{Advances in Neural Information Processing Systems}, vol.
 32, 2019.}
@@
 \CSLLeftMargin{{[}17{]} }%
-\CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly detection: A survey,''} \emph{ACM computing surveys (CSUR)}, vol. 41, no. 3, pp. 1--58, 2009.}
+\CSLRightInline{V. Chandola, A. Banerjee, and V. Kumar, {``Anomaly detection: {A} survey,''} \emph{ACM computing surveys (CSUR)}, vol. 41, no. 3, pp. 1--58, 2009.}
 
@@
 Tobias, and M.
-Blowers, {``Evaluating model drift in machine learning algorithms,''} in \emph{2015 IEEE symposium on computational intelligence for security and defense applications (CISDA)}, 2015, pp.
+Blowers, {``Evaluating model drift in machine learning algorithms,''} in \emph{2015 {IEEE Symposium} on {Computational Intelligence} for {Security} and {Defense Applications} ({CISDA})}, 2015, pp.
 1--8.}
@@
 Raz, and M.
-Zalmanovici, {``Machine learning model drift detection via weak data slices,''} in \emph{2021 IEEE/ACM third international workshop on deep learning for testing and testing for deep learning (DeepTest)}, 2021, pp.
+Zalmanovici, {``Machine {Learning Model Drift Detection Via Weak Data Slices},''} in \emph{2021 {IEEE}/{ACM Third International Workshop} on {Deep Learning} for {Testing} and {Testing} for {Deep Learning} ({DeepTest})}, 2021, pp.
 1--8.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-de2021framework}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-deoliveira2021framework}{}}%
 \CSLLeftMargin{{[}22{]} }%
@@
 \CSLLeftMargin{{[}24{]} }%
-\CSLRightInline{R. S. Pindyck and D. L. Rubinfeld, \emph{Microeconomics}. Pearson Education, 2014.}
+\CSLRightInline{R. S. Pindyck and D. L. Rubinfeld, \emph{Microeconomics}. {Pearson Education}, 2014.}
 
@@
 \CSLLeftMargin{{[}26{]} }%
-\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing kernel hilbert spaces in probability and statistics}. Springer Science \& Business Media, 2011.}
+\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing kernel {Hilbert} spaces in probability and statistics}. {Springer Science \& Business Media}, 2011.}
 
@@
 \CSLLeftMargin{{[}27{]} }%
-\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
+\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of {U} and {V} statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
 
@@
 \CSLLeftMargin{{[}28{]} }%
-\CSLRightInline{S. Hanneke, {``A bound on the label complexity of agnostic active learning,''} in \emph{Proceedings of the 24th international conference on machine learning}, 2007, pp. 353--360.}
+\CSLRightInline{S. Hanneke, {``A bound on the label complexity of agnostic active learning,''} in \emph{Proceedings of the 24th international conference on {Machine} learning}, 2007, pp. 353--360.}
 
@@
 \CSLLeftMargin{{[}29{]} }%
-\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with flux,''} \emph{arXiv preprint arXiv:1811.01457}, 2018.}
+\CSLRightInline{M. Innes \emph{et al.}, {``Fashionable modelling with flux,''} 2018. Available: \url{https://arxiv.org/abs/1811.01457}}
 
@@
 \CSLLeftMargin{{[}30{]} }%
-\CSLRightInline{B. Lakshminarayanan, A. Pritzel, and C. Blundell, {``Simple and scalable predictive uncertainty estimation using deep ensembles,''} \emph{arXiv preprint arXiv:1612.01474}, 2016.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
+\CSLRightInline{B.
+Lakshminarayanan, A.
+Pritzel, and C.
+Blundell, {``Simple and scalable predictive uncertainty estimation using deep ensembles,''} 2016.
+Available: \url{https://arxiv.org/abs/1612.01474}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kaggle2011give}{}}%
 \CSLLeftMargin{{[}31{]} }%
-\CSLRightInline{K.
-Competition, {``Give me some credit, improve on the state of the art in credit scoring by predicting the probability that somebody will experience financial distress in the next two years.''} \url{https://www.kaggle.com/c/GiveMeSomeCredit}}
+\CSLRightInline{Kaggle, {``Give me some credit, {Improve} on the state of the art in credit scoring by predicting the probability that somebody will experience financial distress in the next two years.''} {Kaggle}, 2011.
+Available: \url{https://www.kaggle.com/c/GiveMeSomeCredit}}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikit}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-pedregosa2011scikitlearn}{}}%
 \CSLLeftMargin{{[}33{]} }%
-\CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: Machine learning in python,''} \emph{the Journal of machine Learning research}, vol. 12, pp. 2825--2830, 2011.}
+\CSLRightInline{F. Pedregosa \emph{et al.}, {``Scikit-learn: {Machine} learning in {Python},''} \emph{the Journal of machine Learning research}, vol. 12, pp. 2825--2830, 2011.}
 
@@
 \CSLLeftMargin{{[}35{]} }%
-\CSLRightInline{C. M., {``Racist data destruction? - a boston housing dataset controversy,''} 2019, Available: \url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}}
+\CSLRightInline{M. Carlisle, {``Racist data destruction? - a {Boston} housing dataset controversy,''} 2019, Available: \url{https://medium.com/@docintangible/racist-data-destruction-113e3eff54a8}}
 
@@
 \CSLLeftMargin{{[}36{]} }%
-\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M. Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: A survey,''} \emph{arXiv preprint arXiv:2110.01889}, 2021.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
+\CSLRightInline{V.
+Borisov, T.
+Leemann, K.
+Seßler, J.
+Haug, M.
+Pawelczyk, and G.
+Kasneci, {``Deep neural networks and tabular data: {A} survey,''} 2021.
+Available: \url{https://arxiv.org/abs/2110.01889}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022why}{}}%
 \CSLLeftMargin{{[}37{]} }%
-\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} \emph{arXiv preprint arXiv:2207.08815}, 2022.}
+\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} 2022. Available: \url{https://arxiv.org/abs/2207.08815}}
 

### 2022-08-19 paper/sections/conclusion.rmd
incorporated Cynthia's comments

@@
 Through our work we hope to inspire future research on this important topic.
-To this end we have open-sourced all of our code along with a Julia package - [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).
-The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations].
+To this end we have open-sourced all of our code along with a Julia package: [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).
+The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022counterfactualexplanations].
 That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-08-19 paper/sections/discussion.rmd
incorporated Cynthia's comments

@@
 We would argue that the expected external costs of individual recourse should be shared by all stakeholders.
-The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective).
-This will on average lead to more costly counterfactual outcomes.
-But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
+The most straightforward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective).
+This will on average lead to more costly counterfactual outcomes, but may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals.
 We have shown various types of shift-mitigating strategies that can be used to this end.

### 2022-08-19 paper/sections/empirical.rmd
incorporated Cynthia's comments

@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise each individual experiment is repeat five times.
+To account for noise, each individual experiment is repeat five times.
 
-## $M$ -- Classifiers and Generative Models {#empirical-classifiers}
+## $M$---Classifiers and Generative Models {#empirical-classifiers}
 
-For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
+For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
 We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
@@
 
-## $\mathcal{D}$ -- Data {#empirical-data}
+## $\mathcal{D}$---Data {#empirical-data}
 
@@
 Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics.
-Following @upadhyay2021towards, we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
+Following @upadhyay2021robust, we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results.
 Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.
@@
 The datasets are presented in Figure \@ref(fig:synthetic-data).
-Samples from the negative class are marked in blue while samples of the positive class are marked in orange.
 
@@
 
-Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary.
+Ex-ante we expect to see that by construction, Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary as we saw in Figure \@ref(fig:poc).
 Thus, the choice of a black-box model may have an impact on the paths of the recourse.
-For generators that use latent space search (REVISE @joshi2019towards, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
-Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
+For generators that use latent space search (REVISE @joshi2019realistic, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples.
+Of course, this expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty.
 Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated.
@@
 We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification.
-Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data].
-Originally consisting of 250,000 instances with 11 numerical attributes.
-Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment.
-In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been omitted from the analysis.
+Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@kaggle2011give], originally consisting of 250,000 instances with 11 numerical attributes.
+Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable whether credit card clients default on their payment.
+In its raw form it consists of 23 explanatory variables: 4 categorical features relating to demographic attributes^[These have been omitted from the analysis.
 See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
-Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
 
 As a third dataset we include the **California Housing** dataset derived from the 1990 U.S.
-census and sourced through scikit-learn [@pedregosa2011scikit, @pace1997sparse].
+census and sourced through scikit-learn [@pedregosa2011scikitlearn, @pace1997sparse].
 It consists of 8 continuous features that can be used to predict the median house price for California districts.
@@
 While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example @schut2021generating).
-While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem [@carlisle2019racist].
+While we initially also conducted experiments on that dataset, we eventually discarded this dataset due to surrounding ethical concerns [@carlisle2019racist].
 
@@
 
-## $G$ -- Generators
+## $G$---Generators
 
-All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating].
+All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019realistic], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating].
 In addition, we introduce two new generators in Section \@ref(mitigate) that directly address the issue of endogenous domain and model shifts.

### 2022-08-19 paper/sections/empirical_2.rmd
incorporated Cynthia's comments

@@
 
-Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real) where each column corresponds to one of the three data sets.
+Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real), where each column corresponds to one of the three data sets.
 The results shown here are for the deep ensemble, which once again largely resemble those for the MLP.

### 2022-08-19 paper/sections/introduction.rmd
incorporated Cynthia's comments

@@
 Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences.
-While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making.
+While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise to a new type of problem in the context of data-driven decision-making.
 They are essentially **black boxes**: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools.
-Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons].
+Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@oneil2016weapons].
 The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models.
@@
 A meaningful recourse recommendation for a denied applicant could be: *"If your net savings rate had been 10% of your monthly income instead of the actual 8%, your application would have been successful.
-See if you can temporarily cut down on consumption."* In the remainder of this paper we will use both terminologies - recourse and counterfactual - interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
+See if you can temporarily cut down on consumption."* In the remainder of this paper we will use both terminologies---recourse and counterfactual---interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.
 
@@
 We think that these types of endogenous dynamics may be problematic and warrant our attention.
-From a purely technical we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
+From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
 Secondly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
-We would argue that in some practical applications this can be expected to generate costs for the involved stakeholders.
+We would argue that in some practical applications, this can be expected to generate costs for the involved stakeholders.
 To follow our argument, consider the following two examples:
@@
 ::: {.example #consumer name="Consumer Credit"}
-Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their credit worthiness.
+Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness.
 Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction.
@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \@ref(related) and \@ref(method)).
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)).
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)).
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019realistic, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)).
 Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \@ref(mitigate)).

### 2022-08-19 paper/sections/limitations.rmd
incorporated Cynthia's comments

@@
 The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion.
-In practice, models are in fact updated on a regular basis [@upadhyay2021towards].
+In practice, models are in fact updated on a regular basis [@upadhyay2021robust].
 We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$.
 That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
-In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021towards.
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021robust.
 We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
@@
 For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks.
-While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (@borisov2021deep, @grinsztajn2022tree).
+While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice.
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (@borisov2021deep, @grinsztajn2022why).

### 2022-08-19 paper/sections/methodology.rmd
incorporated Cynthia's comments

@@
 
-In this section we first set out a generalized framework for gradient-based counterfactual search in Section \@ref(method-general) that encapsulates the various individual recourse methods we have chosen to use in our experiments.
-We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.
+In this section we first set out a generalized framework for gradient-based counterfactual search that encapsulates the various individual recourse methods we have chosen to use in our experiments (Section \@ref(method-general)).
+We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective (Section \@ref(method-collective)).
 
@@
 
-We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
-In particular, we include include the following generators in our simulation experiments below: **REVISE** [@joshi2019towards], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
+We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work.
+In particular, we include the following generators in our simulation experiments below: **REVISE** [@joshi2019realistic], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].
 Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models.
@@
 The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
-In this case $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].
+In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
+Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019realistic].
 
 CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$.
-To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models.
-The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
+To quantify predictive uncertainty, Antoran et al.
+[@antoran2020getting] rely on entropy estimates for probabilistic models.
+The greedy approach proposed by Schut et al.
+@schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
 The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
 Schut et al.
-[@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$ we also implicitly minimize predictive entropy - as in CLUE.
+[@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE).
 In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
@@
 , s_K$ that look as different from each other as possible.
-The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach.
+The implementation of DiCE in our library of choice---`CounterfactualExplanations.jl`---uses that exact approach.
 It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).
 
-## ... towards collective recourse
+## ... towards collective recourse {#method-collective}
 

### 2022-08-19 paper/sections/methodology_2.rmd
incorporated Cynthia's comments

@@
 
-Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) - the Julia package we introduced earlier.
+Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)---the Julia package we introduced earlier.
 The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.
@@
 We begin in the static setting at time time $t=0$: firstly, we have some binary classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$, where $\mathcal{D}_0$ and $\mathcal{D}_1$ denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].
+Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple, we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].
 
@@
 The framework we have built facilitates this, making use of multi-threading in order to speed up computations.
-Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
+Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
 To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
@@
 An important consideration is the choice of the kernel function $k(\cdot,\cdot)$.
-In our implementation we make use of Gaussian kernel with a constant length-scale parameter of $0.5$.
+In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of $0.5$.
 As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.
@@
 Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples.
-Then under the null hypothesis we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$.
+Then under the null hypothesis, we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$.
 The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.
@@
 
-As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021towards.
+As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021robust.
 We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$.
@@
 
-As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement).
-This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+As an alternative to PP MMD, we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement).
+This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.

### 2022-08-19 paper/sections/mitigation.rmd
incorporated Cynthia's comments

@@
 Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction.
-Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off.
+Since this is the region of maximal aleatoric uncertainty, the classifier is bound to be thrown off.
 By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
@@
 
-## Classifier Preserving ROAR
+## Classifier Preserving ROAR (ClaPROAR)
 
-Another strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
+Another strategy draws inspiration from ROAR @upadhyay2021robust: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
 Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
@@
 for each counterfactual $k$ where $l$ denotes the loss function used to train $M$.
-This approach, which we shall refer to as **ClapROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
-It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
+This approach, which we shall refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
 It also makes the private vs.
@@
 
-Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{x}$:
+Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average $\bar{x}$:
 
@@
 
-Once again we can putting this in the context of Equation \@ref(eq:collective), the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
+Once again we can putting this in the context of Equation \@ref(eq:collective): the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals.
 Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain.
-In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].
+In the remainder of this paper, we will therefore refer to this approach as **Gravitational** generator, when we investigate its usefulness for mitigating endogenous macrodynamics^[Note that despite the naming conventions, our goal here is not to provide yet more counterfactual generators.
+Rather than looking at them as isolated entities, we believe and demonstrate that different approaches can be effectively combined.].
 
-Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
+Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is, Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClaPROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClaPROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
 
@@
     "Gravitational" => GravitationalGenerator(),
-    "ClapROAR" => ClapROARGenerator()
+    "ClaPROAR" => ClaPROARGenerator()
 )
@@
 The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
-ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
+ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
 As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
 The same overall pattern can be observed for out-of-sample model performance.
-With respect to the other synthetic datasets we found the following: for the Moons dataset the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.
+With respect to the other synthetic datasets, for the Moons dataset, the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.
 
@@
 
-An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.
+An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators.
+In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
 In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline.
@@
 Finally, Figure \@ref(fig:mitigate-real-world-results) shows the results for our real-world data.
-We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.].
-Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
+We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both model shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in Figure \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.].
+Still, both ClaPROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points.
 For the GMSC dataset we observe no notable differences.

### 2022-08-19 paper/sections/related.rmd
Arie's final round of comments
incorporated Cynthia's comments

@@
 
-In this Section we provide a review of the relevant literature.
+In this section we provide a review of the relevant literature.
 First, Subsection \@ref(related-recourse) discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse.
@@
 Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.
-Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:
+Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:
 
 \begin{equation}
-\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
+\min_{x^\prime \in \mathcal{X}} \text{cost}(x^\prime) \ \ \ \mbox{s. t.} \ \ \ M(x^\prime) = y^* (\#eq:obj)
 \end{equation}
@@
 \begin{equation}
-x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) (\#eq:solution)
+x^\prime = \arg \min_{x^\prime}  \text{yloss}(M(x^\prime),y^*) + \lambda \text{cost}(x^\prime) (\#eq:solution)
 \end{equation}
@@
 For models that are differentiable and produce smooth predictions, \@ref(eq:solution) can be solved through gradient descent.
-This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter** - the name of the first author - in the remainder of this paper.
+This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter**, the name of the first author, in the remainder of this paper.
 
@@
 described 29 algorithms [@verma2020counterfactual].
-Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met.
+Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how they ensure that certain requirements for CE are met.
 Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs.
-To name but a few, we are interested in generating counterfactuals that close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and if possible causally founded [@karimi2021algorithmic].
+To name but a few, we are interested in generating counterfactuals that are close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019realistic, @schut2021generating], sparse, diverse [@mothilal2020explaining] and if possible causally founded [@karimi2021algorithmic].
 
@@
 The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual].
-We have been able to identify only one recent work that considers the implications of **exogenous** domain and model shifts in the context of AR [@upadhyay2021towards].
+We have been able to identify only one recent work by Upadhyay et al.
+that considers the implications of **exogenous** domain and model shifts in the context of AR [@upadhyay2021robust].
 Exogenous shifts are strictly of external origin.
-For example, they might stem from data correction, temporal shifts or geospatial changes [@upadhyay2021towards].
-The authors of [@upadhyay2021towards] propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
+For example, they might stem from data correction, temporal shifts or geospatial changes [@upadhyay2021robust].
+Upadhyay et al.
+[@upadhyay2021robust] propose ROAR: a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.
 
@@
 Numerous approaches to anomaly detection have been summarized [@chandola2009anomaly].
-Another well-established research topic is that of concept drift – situations where external variables influence the patterns between the input and the output of a model [@widmer1996learning].
+Another well-established research topic is that of concept drift: situations where external variables influence the patterns between the input and the output of a model [@widmer1996learning].
 For instance, Gama et al.
 offer a review of the adaptive learning techniques which can handle concept drift [@gama2014survey].
-Less previous work is available on the related topic of model drift - changes in model performance over time.
+Less previous work is available on the related topic of model drift: changes in model performance over time.
 Nelson et al.
@@
 
-In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR [@upadhyay2021towards].
+In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR [@upadhyay2021robust].
 In their work they refer to model shifts as simply any perturbation $\Delta$ to the parameters of the model in question: $M$.
-While this also sets the baseline for our analysis here, it is worth noting that in @upadhyay2021towards these perturbations are mechanically introduced.
-In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
+While this also sets the baseline for our analysis here, it is worth noting that in @upadhyay2021robust these perturbations are mechanically introduced.
+In contrast, we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process.
 In addition to quantifying the magnitude of shifts $\Delta$, we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model.
@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, @deoliveira2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022counterfactualexplanations] provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.

### 2022-08-19 README.qmd
anonymised further (removed anything pointing to original git repo and user

--- 2022-08-17 README.qmd
@@
 
-# AlgorithmicRecourseDynamics.jl - Research Paper 📝
+# AlgorithmicRecourseDynamics
 
-**Note** ⚠: You are on the `#original-paper` branch of `AlgorithmicRecourseDynamics.jl`.
+`AlgorithmicRecourseDynamics.jl` is a Julia package for modelling Algorithmic Recourse Dynamics.
+
+## Research Paper 📝
+
+**Note** ⚠: You are browsing the `#original-paper` branch of `AlgorithmicRecourseDynamics.jl`.
 This branch is a static artifact corresponding to the state of the package at the time the paper was first published.
 It can be used to replicate the original findings of the paper.
-For an up-to-date version of the package, please switch to the [`#main`](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl) branch.
+Only this branch is currently accessible as an anonymised git repository.
+The main repository is private and will will be open-sourced after the review process.
 
@@
 - [Notebooks](dev/notebooks/)
-- [Artifacts](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/releases/tag/artifacts) (including data and experimental results)
+- [Supplementary Appendix](build/dev/notebooks/appendix.html) (download the HTML and view in browser)
+- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review)
 

### 2022-08-19 docs/make.jl

@@ -1,25 +1,0 @@
-using AlgorithmicRecourseDynamics
-using Documenter
-
-DocMeta.setdocmeta!(AlgorithmicRecourseDynamics, :DocTestSetup, :(using AlgorithmicRecourseDynamics); recursive=true)
-
-makedocs(;
-    modules=[AlgorithmicRecourseDynamics],
-    authors="Anonymous",
-    repo="https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/{commit}{path}#{line}",
-    sitename="AlgorithmicRecourseDynamics.jl",
-    format=Documenter.HTML(;
-        prettyurls=get(ENV, "CI", "false") == "true",
-        canonical="https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl",
-        edit_link="main",
-        assets=String[],
-    ),
-    pages=[
-        "Home" => "index.md",
-    ],
-)
-
-deploydocs(;
-    repo="github.com/pat-alt/AlgorithmicRecourseDynamics.jl",
-    devbranch="main",
-)

### 2022-08-19 paper/paper.Rmd
rerendered

@@
 
-bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
+bibliography: ../bib.bib
 csl: ieee.csl

### 2022-08-19 README.qmd
final things on supp appendix

@@
         wrap: none
+        html-math-method: webtex
 ---
@@
 
-**Note** ⚠: You are browsing the `#original-paper` branch of `AlgorithmicRecourseDynamics.jl`.
+**Note** ⚠: You are browsing the (anonymised) [`#original-paper`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) branch of `AlgorithmicRecourseDynamics.jl`.
 This branch is a static artifact corresponding to the state of the package at the time the paper was first published.
@@
 - [Notebooks](dev/notebooks/)
-- [Supplementary Appendix](build/dev/notebooks/appendix.html) (download the HTML and view in browser)
-- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review)
+- [Supplementary Appendix](build/dev/notebooks/appendix.html) generated from notebooks (download the HTML and view in browser)
+- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review process)
 
@@
 
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata.
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
 The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.

### 2022-08-19 paper/paper.Rmd
final things on supp appendix

@@
 ---
-title: Endogenous Macrodynamics in Algorithmic Recourse
+title: "SoK: Endogenous Macrodynamics in Algorithmic Recourse"
 affiliation:

### 2022-08-19 paper/paper.tex
final things on supp appendix

@@
 \hypersetup{
-            pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
+            pdftitle={SoK: Endogenous Macrodynamics in Algorithmic Recourse},
             pdfkeywords={Algorithmic Recourse, Counterfactual Explanations, Explainable AI, Dynamic Systems},
@@
 % Do not put math or special symbols in the title.
-\title{Endogenous Macrodynamics in Algorithmic Recourse}
+\title{SoK: Endogenous Macrodynamics in Algorithmic Recourse}
 
@@
 To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
-Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \ref{related} and \ref{method}).
-Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\ref{method-2}).
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (\ref{empirical} and \ref{empirical-2}).
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \ref{related} and \ref{method}).
+Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \ref{method-2}).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (Sections \ref{empirical} and \ref{empirical-2}).
 Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \ref{mitigate}).

### 2022-08-19 paper/sections/introduction.rmd
final things on supp appendix

@@
 To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR.
-Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \@ref(related) and \@ref(method)).
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)).
-Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019realistic, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)).
+Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \@ref(related) and \@ref(method)).
+Secondly, we introduce an experimental framework extending previous work by @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \@ref(method-2)).
+Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019realistic, @mothilal2020explaining and @antoran2020getting (Sections \@ref(empirical) and \@ref(empirical-2)).
 Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \@ref(mitigate)).

### 2022-08-19 README.qmd
gor rid of mathjax in readme

@@
 In this work we investigate what happens if Algorithmic Recourse is actually implemented by a large number of individuals.
-The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.
+The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class (y=0) are marked in blue and samples of the positive class (y=1) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.
 

### 2022-08-19 paper/paper.Rmd
one final thing

@@
 Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve competition for scarce resources.
+We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve large groups of individuals.
 Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.

### 2022-08-19 paper/paper.tex
one final thing

@@
 
-The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated PP MMD is statistically significant and large.
+The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: the estimated PP MMD is statistically significant and large for all datasets.
 Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.

### 2022-08-19 paper/sections/empirical_2.rmd
one final thing

@@
 
-The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated PP MMD is statistically significant and large.
+The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): the estimated PP MMD is statistically significant and large for all datasets.
 Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.

### 2022-10-31 README.qmd
deps and reviewer comments

@@
 ---
+
+[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/stable)
+[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/dev)
+[![Build Status](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml?query=branch%3Amain)
+[![Coverage](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl)
 

### 2022-10-31 dev/submissions/satml/revision.qmd
deps and reviewer comments

---
format:
    pdf: default
---

## General

Dear reviewers,

We were very grateful to receive such thoughtful comments from all of you. Of course, we were also very happy to get overwhelmingly very positive feedback from you --- it has been truly rewarding to see that you all see value in this type of work.

This leads us straight to the point on SoK, that two of you made. We were both happy and somewhat surprised to see that two of you consider this as more of a genuinely novel contribution, than a systematization of knowledge. We did not think our work was novel enough, because a) we build on a very simple observation (see Reviewer vV4a's comments) and b) the algorithmic strategies we suggest are very simple. We do hope that we provide a novel perspective on Algorithmic Recourse, but we have done so merely by restating the existing framework in a way that explicitly addresses the notion of external costs that we propose. All of that being said, if the consensus is that this paper is novel enough for the "research paper" track, then we are happy to go for that instead. That is provided this is possible, of course, and the consensus also reflects the view of the PC.

We also have a general question with respect to prioritization of requested changes. When you've had a chance to read our individual comments on all of your individual reviews, could you let us know if you strongly disagree with us on any particular point? Upon also seeing other reviewer's comments, do you have a sense of what we should prioritize?

Once again, many thanks for your time and thoughtful feedback. It is really nice to see so much engagement.

## Reviewer ebVS

### Review

This paper presents an study of negative dynamics that happen when using algorithmic recourse. The authors provide a framework to study this problem and conduct an empirical evaluation on synthetic and real datasets.

PROS:

Novel angle on algorithmic recourse
Characterization of negative dynamics and the situations in which they happen
Thorough empirical evaluation
Proposal of fixes and evaluations
Honest discussion of limitations
CONS

Paper content does not match an SoK. I do not understand why the authors choose to present this paper as an SoK. There is no systematization of knowledge. Instead, the authors provide new results on the area.
I enjoyed reading this paper. The authors make a good job at explaining the problem and motivating why it is relevant contextualizing their contributions with respect to the state of the art. The proposed treatment is simple, yet effective to show clearly the negative effects that considering recourse as static in time can have.

I only have two remarks:

The authors could do a better job at interpreting what the metrics they propose mean. While mathematically they make sense, I missed more intuition about why they capture the problems the authors are trying to highlight.
In the countermeasure section, I would have liked to see more connection to the motivating examples given at the beginning of the paper. How would these countermeasures help in the case of the University?
As a minor comment, in several places in the paper citations are treated as subjects of sentences (e.g., "presented by [4]"). This is non-desirable. I recommend to revise to make the authors the subject (e.g., "presented by X [4])

### Response

Dear reviewer,

Thank you very much, we are happy to see that you enjoyed reading our paper.

Regarding your two main remarks:

1. Metrics: We think this is a fair point with respect to MMD. Here, in particular, we tried to make sure to rigorously lay out the proposed method(s). In our understanding MMD is a go-to measure in these kinds of setting, so we relied on that and a detailed description of what it entails. Do you have a specific suggestion as to how we can address intuition here? With respect to the remaining measures we think we do provide intuition, but if there is anything in particular that you were missing here please shout.

2. Linking the countermeasures back to the examples is a good idea. If after addressing other points we still have space, we will add a sentence or two. Does that sound good? Please see the general comment regarding prioritization.

With respect to SoK, please also see the general comment.

## Reviewer vV4a

### Review

This paper builds on a simple observation: when people successfully implement the actions that counterfactual explanations have suggested to achieve a different classification, such actions may induce concept drift if changes to the features do not result in a corresponding change in the outcome that these feature values are expected to predict. Or to put it differently, if acting on counterfactual explanations actually changes the observed relationship between X and Y, then a model's predictive accuracy will decline as more people take these actions. The authors build on this observation to point out that the cost of concept drift is likely to affect other people, not just the institution using the model to make predictions. This could be true for two different reasons. First, an institution that finds that the accuracy of its decisions declines as more people make use of counterfactual explanations may cease to provide counterfactual explanations altogether. Second, if the institution recognizes that it needs to retrain the model to deal with drift, **the retrained model will likely move the decision boundary even further away from the people who have not yet attempted to make any changes, thereby increasing the costs they would face in seeking to achieve recourse.** Much of the paper is spent trying to formalize these dynamics and to evaluate how serious the problem is when using different methods for generating counterfactual explanations. The authors run experiments on a range of real-world and synthetic datasets, find that the problem is often quite significant across all of the methods tested, with some faring better than others. Finally, the authors consider a number of mitigation techniques, most premised on the idea that the best way to avoid these dynamics is to ensure that counterfactual explanations suggest actions that are in line with the true data generating process.

As I mentioned in the title field, I think this is a **very nice paper and is well above the bar for acceptance**. It's well motivated, it engages quite effectively with the existing and relevant literature, it formulates the problem in a very sensible way, and it does a wonderful job presenting all of the formal and experimental work in a very clear and accessible manner.

All that said, it also strikes me that the paper is built upon an observation that is not at all surprising, even if it is not one that is given enough attention in the literature on counterfactual explanations: **taking the actions suggested by a counterfactual explanation may not change the underlying property that a model is trying to predict**. For example, it should be obvious that it's possible to change the values of the features that are considered by a credit scoring model without changing the underlying likelihood of default. The reason for this is again obvious: most machine learning models are only learning statistical patterns in the training data; they are not learning the causal relationship between features and an outcome of interest. What is necessary to achieve recourse from a model (i.e., achieve a different classification) may be very different than what is necessary to actually achieve the quality that the model is try to predict.

The likely dynamics that fall out from this observation are pretty obvious: acting on counterfactual explanations can cause concept drift (i.e., alternating the observed relationship between X and Y); it will decrease accuracy due to concept drift will impose costs on the decision maker; the decision maker will likely try to find some way to reduce these costs, either by ceasing to provide counterfactual explanations or by retraining the model; retraining the model **will push the decision boundary further away from the initial people who have not yet attempted to achieve recourse**. It seems a bit odd to bring to bear so much machinery on this problem when the underlying issue is rather obvious.

Perhaps this didn't strike the authors as particularly obvious because they did not engage with the related work on strategic classification, which has focused on the fact that preventing gaming of machine learning models is really "causal modeling in disguise": https://proceedings.mlr.press/v119/miller20b.html. While much of this work assumes that decision subjects might be acting adversarially (i.e., making changes to feature values in bad faith, knowing that such changes will not affect the underlying property that they are meant to predict), the main observations of this work apply equally well to this setting. Indeed, as Barocas, Hardt, and Narayanan point out, "**gaming can be a problem even when decision subjects are not acting adversarially. Job seekers may expend considerable effort and money to obtain meaningless credentials that they are told matter in their industry, only to find that while this helps them land a job, it does not make them any better prepared to actually perform it." https://fairmlbook.org/legitimacy.html.** Again, with this other scholarship in mind, it is not at all surprising that the authors conclude that "to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process".

I still think this work is quite valuable, especially because it does such a nice job demonstrating these dynamics with real-world data sets using a range of counterfactual explanation techniques. **But I would caution the authors against staking out such a strong position on the novelty of the initial observation that motivates much of this work.**

### Requested Changes:

As I mentioned at the end of the review, I suggest that the authors adopt a more modest position with respect to the novelty of their work. In doing that, I also encourage them to take a closer look at the work on strategic classification and to try to link some of the observations in that work to the problem they are seeking to describe. I believe the paper would be strengthened significantly in drawing such connections and using these observations to better motivate and ground their work.

### Response

Dear reviewer,

Thank you very much for the detailed review --- it is really great to see that you have engaged with the paper so much.

Turning straight to your main point of criticism, not relating this to existing literature on strategic classification was indeed an oversight on our end. It should be possible to tone down the narrative a bit and point to the literature you have brought to our attention. In light of this, an interesting avenue for future research would be to test approaches to counterfactual generation that do explicitly incorporate causal knowledge (e.g. [Karimi et al. (2020)](https://arxiv.org/abs/2002.06278)), which we briefly reference in the paper. Since we are pushing the page limit (and also in light of the quick turnover now), we do feel that it makes sense to merely point these additional considerations out, rather then attempt to address them in this paper. Do you agree with this?

One other tiny thing: we observe that the decision boundary does in fact "moves away from the target class", therefore making recourse even less costly to "future" recourse seekers. Of course, there is a catch: we would expect that recourse providers who are aware of such dynamics refuse to provide any recourse at all, so ultimately the cost to future recourse seekers is that they would be denied recourse. We tried to illustrate this through example I.2, but perhaps we need to make this even clearer?

## Reviewer cxf3

### Review

This paper examines the problems of model and distribution shifts which arise when individuals act on recourses received for the model. It argues for a departure from studying the problem of recourse for individual users toward a collective and longitudinal setting. The new problem is formalized with a number of evaluation metrics derived from related literature or developed by the authors. The paper further explores several algorithmic mitigation approaches to prevent the shifts while providing users with useful recourses. The paper supports its arguments with an empirical simulation using multiple base models, recourse models, and datasets.

The paper is very well-written. It might serve as a good foundation for future studies of the problem, especially since the authors plan to release their simulation framework. It might not seem like a typical **SoK paper in that it does not survey a body of work over an existing problem, but rather derives and grounds a new problem using prior literature and an empirical study.** To my mind (I am only tangentially familiar with the algorithmic recourse literature) the paper provides high pedagogical value.

I enjoyed reading the paper and believe it could be accepted as is. One particular omission which the authors might consider adding are plots showing how the shift metrics change over the course of the simulation.

When it comes to the out-of-sample error measure, I am wondering: to what extent do you expect the out of sample error increase to be a problem in practice? That is, what would the sample that is set aside in the experiments correspond to? Would the data distribution among users who did not yet receive recourse remain intact? Or might one assume that they'd also learn what they need to do to receive a positive outcome from the classifier through a backchannel?

### Requested Changes

I would have liked to to see plots showing how the shift metrics change over the course of the simulation.

### Response

Dear reviewer,

thank you very much for your review, we are happy to see that you enjoyed reading our paper.

With respect to your main points:

1. We considered showing the sort of chart you have in mind, but ultimately resorted to reserving this for the supplementary appendix that we point to in the paper. The reason we decided to omit these types of charts from the main paper, was that they only provide marginal additional information and are somewhat harder to read than the bar charts. In particular, we observed that over the course of the simulation the induced shifts gradually increase roughly proportionately for the different generators, until they level off. From the paper:

"Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment. That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment."

Currently, due to anonymization the online appendix only lives as a HTML document in our repo. We were planning to make it easily accessible through Github pages, once the paper is released. We can then provide a hyperlink to the relevant charts. Does this sufficiently address the issue in your view? If not we could consider replacing one of the bar charts with a line chart showing the changes over time.

2. Regarding the out-of-sample error, there are two different things going on here. Firstly, to address the issue of uncertainty around our proposed shift measures, we use a form of cross validation: we simply rerun the experiment multiple times for the sample sample and leverage the fact that individuals are randomly selected for recourse. This should not be an issue in practice, since it doesn't actually involve any data splitting. Secondly, though, to assess model performance we reserve a holdout set (this is probably what you are referring to). It's a good point. Our immediate suggestion would be to add an additional layer of cross-validation, this time with respect to data splitting. So the standard procedure of training on (K-1) out of K folds for each k. As a matter of fact, we might consider getting rid of the first CV procedure and simply replacing it with this one.

Regarding your point on SoK, please so our general comment.

### 2022-10-31 src/data/functions.jl

@@ -1,6 +1,9 @@
-using Pkg.Artifacts
+using LazyArtifacts
+using CounterfactualExplanations
+using CSV
+using DataFrames
+using StatsBase
 
-using CSV, DataFrames, CounterfactualExplanations
 function load_synthetic(max_obs::Union{Nothing, Int}=nothing)
     data_dir = joinpath(artifact"data","data/synthetic")
     files = readdir(data_dir)
@@ -20,7 +23,6 @@ function load_synthetic(max_obs::Union{Nothing, Int}=nothing)
     return data
 end
 
-using CounterfactualExplanations
 function load_real_world(max_obs::Union{Nothing, Int}=nothing)
     data_dir = joinpath(artifact"data","data/real_world")
     files = readdir(data_dir)
@@ -41,7 +43,7 @@ function load_real_world(max_obs::Union{Nothing, Int}=nothing)
     return data
 end
 
-using StatsBase
+
 function scale(X, dim)
     dt = fit(ZScoreTransform, X, dim=dim)
     X_scaled = StatsBase.transform(dt, X)

### 2022-10-31 src/models/utils.jl

@@ -1,4 +1,9 @@
-using MLUtils, CounterfactualExplanations
+using CounterfactualExplanations
+using CounterfactualExplanations.DataPreprocessing
+using Flux
+using MLJ
+using MLUtils
+
 """
     data_loader(data::CounterfactualData)
 
@@ -11,9 +16,6 @@ function data_loader(data::CounterfactualData)
     return data
 end
 
-using CounterfactualExplanations
-using CounterfactualExplanations.DataPreprocessing
-using MLJ, Flux
 function model_evaluation(M::CounterfactualExplanations.AbstractFittedModel, test_data::CounterfactualData)
     X, y = DataPreprocessing.unpack(test_data)
     m = MulticlassFScore()

### 2022-11-03 dev/submissions/satml/revision.qmd
incorporated some revsions

@@
 With respect to SoK, please also see the general comment.
+
+### 2nd Response
+
+Thanks again!
+For MMD we have added a sentence explaining that intuitively, larger values indicate that it is more likely that the two distributions are different.
+And that "In our context, large values therefore indicate that a domain shift indeed seems to have occurred." We state this just before we go on to explain the bootstrap procedure.
+We have also now linked the mitigation strategies back to the university example.
 
@@
 Regarding your point on SoK, please so our general comment.
+
+### 2nd Response
+
+Thanks again! We have added the remark now as a footnote. We'll aim to polish the online appendix further, to make it as easy as possible to direct readers to the charts showing dynamics.

### 2022-11-03 paper/paper.Rmd
incorporated some revsions

@@
 ---
-title: "SoK: Endogenous Macrodynamics in Algorithmic Recourse"
+title: "Endogenous Macrodynamics in Algorithmic Recourse"
 affiliation:

### 2022-11-03 paper/paper.tex
incorporated some revsions

@@
 \hypersetup{
-            pdftitle={SoK: Endogenous Macrodynamics in Algorithmic Recourse},
+            pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
             pdfkeywords={Algorithmic Recourse, Counterfactual Explanations, Explainable AI, Dynamic Systems},
@@
 % Do not put math or special symbols in the title.
-\title{SoK: Endogenous Macrodynamics in Algorithmic Recourse}
+\title{Endogenous Macrodynamics in Algorithmic Recourse}
 
@@
 
-Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
+Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by Wachter et al.
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \ref{related} and \ref{method}).
-Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \ref{method-2}).
+Secondly, we introduce an experimental framework extending previous work by Altmeyer \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \ref{method-2}).
 Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (Sections \ref{empirical} and \ref{empirical-2}).
@@
 
-A framework for Counterfactual Explanations was first proposed in 2017 by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then.
+A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al.
+\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then.
 Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\).
@@
 In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
-They largely differ with respect to the chosen generative model: for example, the generator proposed by \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.
+They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al.
+\protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.
 
@@
 As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD(X,\tilde{X})=0\) if and only if \(X=\tilde{X}\).
-
-The evaluation metric in Equation \eqref{eq:mmd} is computed after every round \(t=1,...,T\) of the experiment.
+Conversely, larger values \(MMD(X,\tilde{X})>0\) indicate that it is more likely that \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) are different distributions.
+In our context, large values therefore indicate that a domain shift indeed seems to have occurred.
+
 To assess the statistical significance of the observed shifts under the null hypothesis that samples \(X\) and \(\tilde{X}\) were drawn from the same probability distribution, we follow \protect\hyperlink{ref-arcones1992bootstrap}{{[}27{]}}.
@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise, each individual experiment is repeat five times.
+To account for noise, each individual experiment is repeat five times.\footnote{In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
+An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.}
 
@@
 Figure \ref{fig:mitigation} shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is, Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClaPROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the Gravitational generator and ClaPROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability \(\gamma\).}.
+Linking these ideas back to Example \ref{exm:student}, the mitigation strategies help ensure that the recommended recourse actions are substantial enough to truly lead to an increase in the probability that the admitted student eventually graduates.
 
@@
 That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
-In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in Upadhyay et al.
+\protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}.
 We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model \(M\) is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
-Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary.
-In practice it is doubtful that agents behave according to such simple rules.
-Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).
+
+\hypertarget{causal-modelling}{%
+\subsection{Causal Modelling}\label{causal-modelling}}
+
+In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge.
+The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
+In the context of Example \ref{exm:student}, the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee.
+Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating.
+The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}.
+Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
+\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.
 
@@
 While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}36{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}37{]}}).
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}37{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}38{]}}).
 
@@
 Richter, and G.
-Kasneci, {``Carla: A python library to benchmark algorithmic recourse and counterfactual explanation algorithms,''} 2021.
+Kasneci, {``Carla: {A} python library to benchmark algorithmic recourse and counterfactual explanation algorithms,''} 2021.
 Available: \url{https://arxiv.org/abs/2108.00783}}
@@
 \CSLRightInline{P.
-Altmeyer, {``{CounterfactualExplanations}.jl - a {Julia} package for {Counterfactual Explanations} and {Algorithmic Recourse}.''} 2022.
+Altmeyer, {``{CounterfactualExplanations}.{Jl} - a {Julia} package for {Counterfactual Explanations} and {Algorithmic Recourse}.''} 2022.
 Available: \url{https://github.com/pat-alt/CounterfactualExplanations.jl}}
@@
 Schölkopf, and I.
-Valera, {``A survey of algorithmic recourse: Definitions, formulations, solutions, and prospects,''} 2020.
+Valera, {``A survey of algorithmic recourse: {Definitions}, formulations, solutions, and prospects,''} 2020.
 Available: \url{https://arxiv.org/abs/2010.04050}}
@@
 Schölkopf, and I.
-Valera, {``Algorithmic recourse: From counterfactual explanations to interventions,''} in \emph{Proceedings of the 2021 {ACM Conference} on {Fairness}, {Accountability}, and {Transparency}}, 2021, pp.
+Valera, {``Algorithmic recourse: {From} counterfactual explanations to interventions,''} in \emph{Proceedings of the 2021 {ACM Conference} on {Fairness}, {Accountability}, and {Transparency}}, 2021, pp.
 353--362.}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-miller2020strategic}{}}%
+\CSLLeftMargin{{[}36{]} }%
+\CSLRightInline{J.
+Miller, S.
+Milli, and M.
+Hardt, {``Strategic {Classification} is {Causal Modeling} in {Disguise},''} in \emph{Proceedings of the 37th {International Conference} on {Machine Learning}}, Nov.
+2020, pp.
+6917--6926.
+Accessed: Nov.
+03, 2022.
+{[}Online{]}.
+Available: \url{https://proceedings.mlr.press/v119/miller20b.html}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}36{]} }%
+\CSLLeftMargin{{[}37{]} }%
 \CSLRightInline{V.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022why}{}}%
-\CSLLeftMargin{{[}37{]} }%
+\CSLLeftMargin{{[}38{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} 2022. Available: \url{https://arxiv.org/abs/2207.08815}}

### 2022-11-03 paper/sections/empirical.rmd
incorporated some revsions

@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise, each individual experiment is repeat five times.
+To account for noise, each individual experiment is repeat five times.^[In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
+An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.]
 

### 2022-11-03 paper/sections/introduction.rmd
incorporated some revsions

@@
 
-Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
+Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by Wachter et al.
+@wachter2017counterfactual: the implementation of AR for a subset of individuals immediately leads to a visible domain shift in the (orange) target class (b), which in turn triggers a model shift (c).
 As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d).
@@
 Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \@ref(related) and \@ref(method)).
-Secondly, we introduce an experimental framework extending previous work by @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \@ref(method-2)).
+Secondly, we introduce an experimental framework extending previous work by Altmeyer @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \@ref(method-2)).
 Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019realistic, @mothilal2020explaining and @antoran2020getting (Sections \@ref(empirical) and \@ref(empirical-2)).

### 2022-11-03 paper/sections/limitations.rmd
incorporated some revsions

@@
 That being said, our experimental design is a vast over-simplification of potential real-world scenarios.
-In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021robust.
+In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in Upadhyay et al.
+@upadhyay2021robust.
 We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.
-Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary.
-In practice it is doubtful that agents behave according to such simple rules.
-Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).
+
+## Causal Modelling
+
+In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge.
+The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
+In the context of Example \@ref(exm:student), the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee.
+Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating.
+The implementation of such changes is typically described as **gaming** @miller2020strategic.
+Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
+@karimi2021algorithmic.
 

### 2022-11-03 paper/sections/methodology.rmd
incorporated some revsions

@@
 In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category.
-They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].
+They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al.
+@dombrowski2021diffeomorphic relies on normalizing flows.].
 

### 2022-11-03 paper/sections/methodology_2.rmd
incorporated some revsions

@@
 As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.
+Conversely, larger values $MMD(X,\tilde{X})>0$ indicate that it is more likely that $\mathcal{X}$ and $\mathcal{\tilde{X}}$ are different distributions.
+In our context, large values therefore indicate that a domain shift indeed seems to have occurred.
 
-The evaluation metric in Equation \@ref(eq:mmd) is computed after every round $t=1,...,T$ of the experiment.
 To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap.

### 2022-11-03 paper/sections/related.rmd
incorporated some revsions

@@
 
-A framework for Counterfactual Explanations was first proposed in 2017 by @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then.
+A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al.
+@wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then.
 Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.

### 2022-11-03 src/models/deep_ensemble.jl

@@ -1,4 +1,10 @@
+using CounterfactualExplanations
+using Flux
+using Flux.Optimise: update!
+using LinearAlgebra
 using Parameters
+using Statistics
+
 @with_kw struct FluxEnsembleParams 
     loss::Symbol = :logitbinarycrossentropy
     opt::Symbol = :Adam
@@ -6,46 +12,14 @@ using Parameters
     data_loader::Function = data_loader
 end
 
-
-import CounterfactualExplanations.Models: FluxEnsemble
-function FluxEnsemble(data::CounterfactualData, K::Int=5;kwargs...)
-    X, y = CounterfactualExplanations.DataPreprocessing.unpack(data)
-    input_dim = size(X,1)
-    output_dim = length(unique(y))
-    output_dim = output_dim==2 ? output_dim=1 : output_dim # adjust in case binary
-    ensemble = build_ensemble(K;input_dim=input_dim, output_dim=output_dim,kwargs...)
-
-    if output_dim==1
-        M = FluxEnsemble(ensemble; likelihood=:classification_binary)
-    else
-        M = FluxEnsemble(ensemble; likelihood=:classification_multi)
-    end
-
-    return M
-end
-
-using Flux
-"""
-    build_ensemble(K::Int;kw=(input_dim=2,n_hidden=32,output_dim=1))
-
-Helper function that builds an ensemble of `K` models.
-"""
-function build_ensemble(K::Int;kwargs...)
-    ensemble = [build_mlp(;kwargs...) for i in 1:K]
-    return ensemble
-end
-
-using Flux
-using Flux.Optimise: update!
-using CounterfactualExplanations
-using CounterfactualExplanations.Models: FluxEnsemble
+@info println(names(CounterfactualExplanations.Models))
 
 """
-    train(M::FluxEnsemble, data::CounterfactualData; kwargs...)
+    train(M::CounterfactualExplanations.Models.FluxEnsemble, data::CounterfactualData; kwargs...)
 
 Wrapper function to retrain.
 """
-function train(M::FluxEnsemble, data::CounterfactualData; kwargs...)
+function train(M::CounterfactualExplanations.Models.FluxEnsemble, data::CounterfactualData; kwargs...)
 
     args = FluxEnsembleParams(; kwargs...)
 
@@ -68,9 +42,33 @@ function train(M::FluxEnsemble, data::CounterfactualData; kwargs...)
     
 end
 
-using LinearAlgebra, Flux, Statistics
+"""
[…]

### 2022-11-03 src/models/mlp.jl

@@ -1,4 +1,10 @@
+using CounterfactualExplanations
+using Flux
+using Flux.Optimise: update!
+using LinearAlgebra
 using Parameters
+using Statistics
+
 @with_kw struct FluxModelParams 
     loss::Symbol = :logitbinarycrossentropy
     opt::Symbol = :Adam
@@ -6,15 +12,12 @@ using Parameters
     data_loader::Function = data_loader
 end
 
-using Flux
-using Flux.Optimise: update!
-using CounterfactualExplanations
 """
-    train(M::FluxModel, data::CounterfactualData; kwargs...)
+    train(M::CounterfactualExplanations.Models.FluxModel, data::CounterfactualData; kwargs...)
 
 Wrapper function to retrain `FluxModel`.
 """
-function train(M::FluxModel, data::CounterfactualData; kwargs...)
+function train(M::CounterfactualExplanations.Models.FluxModel, data::CounterfactualData; kwargs...)
 
     args = FluxModelParams(; kwargs...)
 
@@ -34,7 +37,6 @@ function train(M::FluxModel, data::CounterfactualData; kwargs...)
     
 end
 
-using Statistics
 function forward!(model::Flux.Chain, data; loss::Symbol, opt::Symbol, n_epochs::Int=10)
 
     # Loss:
@@ -121,8 +123,7 @@ function build_mlp(;
 
 end
 
-import CounterfactualExplanations.Models: FluxModel
-function FluxModel(data::CounterfactualData;kwargs...)
+function CounterfactualExplanations.Models.FluxModel(data::CounterfactualData;kwargs...)
     X, y = CounterfactualExplanations.DataPreprocessing.unpack(data)
     input_dim = size(X,1)
     output_dim = length(unique(y))
@@ -152,8 +153,7 @@ function LogisticRegression(data::CounterfactualData;kwargs...)
     return M
 end
 
-using LinearAlgebra, Flux, Statistics
-function perturbation(model::FluxModel, new_model::FluxModel)
+function perturbation(model::CounterfactualExplanations.Models.FluxModel, new_model::CounterfactualExplanations.Models.FluxModel)
     mlp = model.model
     new_mlp = new_model.model
     Δ = mean(map(x -> norm(x)/length(x),Flux.params(new_mlp).-Flux.params(mlp)))

### 2022-11-03 dev/submissions/satml/revision.qmd
cut example

@@
 
+### 2nd Response
+
+Thanks - we have now added a subsection to the section on limitations:
+
+"In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge.
+The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
+The implementation of such changes is typically described as gaming [36].
+Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
+[13]."
+
+Does this sound reasonable?
+By adding it as a limitation we make it clear that this is in fact a shortcoming of our current work, which should help with toning down the narrative.
+Beyond that, were there any specific points in the paper that struck you as too boastful?
+
+We also thought about potentially linking this story on causality back to one of the examples, but have so far struggled to find any obvious non-causal attribute and are anyway struggling with space.
+We thought of this: "In the context of Example I.2, the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university’s hiring committee.
+Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students’ chances of graduating." Another example we could think of is motivational statements that may or may not be causally linked to student performance.
+
+As for your point on retraining, ideally, what we'd want is to actually intervene and retrain on observed outcomes after the intervention.
+We would then expect that the model would "catch the fact that at least some people who relied on AR actually failed to experience the predicted outcome (e.g., repayment or completing their degree)".
+This would set the baseline for a very interesting study of the actual causal effect of algorithmic recourse on actual (as opposed to predicted) outcomes.
+But since here we're dealing with observational real-world data, we were more limited in our analysis.
+
+To be clear, during retraining we consider new predicted labels as new observed labels.
+The alternative would be to retrain with existing labels in which case we would indeed also expect the decision boundary to shift in the direction of the target class.
+Both approaches are simplifications and not quite optimal (which we recognize in the limitations section).
+Our approach is simplified, because we assume that the agent in charge of the model accepts the new labels as ground truth.
+The alternative approach is also not optimal, because it assumes that changes to features have no affect on actual outcomes.
+
+Again, since we never actually observe the outcome after implementation of recourse, we are limited here.
+But having thought about this again now, an interesting idea for future work might be to generate synthetic data from a simple causal model.
+
+Thanks again!
+
 ## Reviewer cxf3

### 2022-11-03 paper/paper.tex
cut example

@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-In the context of Example \ref{exm:student}, the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee.
-Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating.
 The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}.

### 2022-11-03 paper/sections/limitations.rmd
cut example

@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-In the context of Example \@ref(exm:student), the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee.
-Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating.
 The implementation of such changes is typically described as **gaming** @miller2020strategic.

### 2022-11-04 paper/paper.tex
trimmed down to meet page limit

@@
 
-Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as \textbf{individual recourse} in some places.
-Arguably though, most real-world applications that warrant Algorithmic Recourse involve potentially large groups of individuals typically competing for scarce resources.
-Our work demonstrates that in such scenarios, choices made by or for one single individual are likely to affect the broader collective of individuals in ways that current approaches to AR fail to account for.
-More specifically, we argue that a strict focus on minimizing the private costs faced by individuals may be too narrow an objective.
+Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of a single individual.
+Arguably though, most real-world applications that warrant AR involve potentially large groups of individuals typically competing for scarce resources.
+Our work demonstrates that in such scenarios, choices made by or for a single individual are likely to affect the broader collective of individuals in ways that many current approaches to AR fail to account for.
+More specifically, we argue that a strict focus on minimizing the private costs to individuals may be too narrow an objective.
 
@@
 
-We think that these types of endogenous dynamics may be problematic and warrant our attention.
+We think that these types of endogenous dynamics may be problematic and deserve our attention.
 From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
 Secondly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
-We would argue that in some practical applications, this can be expected to generate costs for the involved stakeholders.
+We think that in some practical applications, this can be expected to generate costs for involved stakeholders.
 To follow our argument, consider the following two examples:
@@
 
-The dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
+The dynamics illustrated in Figure \ref{fig:poc} in were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
 We begin in the static setting at time time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse\footnote{To keep notation simple, we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.
+Note that we focus our attention on classification problems, since classification poses the most common use-case for recourse\footnote{To keep notation simple, we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.
 
@@
 
-We use four synthetic binary classification datasets consisting of 1000 samples each: \textbf{Overlapping}, \textbf{Linearly Separable}, \textbf{Circles} and \textbf{Moons}.
-The datasets are presented in Figure \ref{fig:synthetic-data}.
+We use four synthetic binary classification datasets consisting of 1000 samples each: \textbf{Overlapping}, \textbf{Linearly Separable}, \textbf{Circles} and \textbf{Moons} (\ref{fig:synthetic-data}).
 
@@
 The continuous outcome variable is binarized as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or not the median house price of a given district is above or below the median of all districts.
-While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example \protect\hyperlink{ref-schut2021generating}{{[}6{]}}).
+While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
 While we initially also conducted experiments on that dataset, we eventually discarded this dataset due to surrounding ethical concerns \protect\hyperlink{ref-carlisle2019racist}{{[}35{]}}.
@@
 The top row shows the estimated domain shifts.
-While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds.
+While it is difficult to interpret the exact magnitude of MMD, we can see that the values are clearly different from zero and there is essentially no variation across our five folds.
 With respect to the domain shifts, the Greedy generator actually induces the smallest shifts.
@@
 For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.}.
-Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \ref{fig:syn}.
+Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \ref{fig:syn}.
 The same broadly holds for the Moons data, For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.
 
-Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment.
+Finally, it is also worth noting that the observed dynamics and patterns are consistent throughout the course of the experiment.
 That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.
@@
 
-To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments (RQ \ref{prp:shifts}) and we find them substantial enough to be considered costly (RQ \ref{prp:costs}); secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect (RQ \ref{prp:het}).
-The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process (RQ \ref{prp:drive}).
+To recap, we answer our research questions: firstly, endogenous dynamics do emerge in our experiments (RQ \ref{prp:shifts}) and we find them substantial enough to be considered costly (RQ \ref{prp:costs}); secondly, the choice of the counterfactual generator matters, with Latent Space search generally having a dampening effect (RQ \ref{prp:het}).
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to individuals and outcomes that comply with the data generating process (RQ \ref{prp:drive}).
 
@@
 But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and, perhaps more importantly, their performance hinges on the performance of said generative models.
-Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.
+Fortunately, it turns out that we can use a number of other, much simpler strategies.
 
@@
 The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold \(\gamma\).
-This threshold determines when a proposed counterfactual should be considered as valid.
+This threshold determines when a counterfactual should be considered as valid.
 Under \(\gamma=0.5\), counterfactuals will end up near the decision boundary by construction.
 Since this is the region of maximal aleatoric uncertainty, the classifier is bound to be thrown off.
-By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
+By setting a more conservative threshold, we can avoid this issue to some extent.
 A drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
@@
 
-Another strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
+Another strategy draws inspiration from ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}: to preserve the classifier, we propose to explicitly penalize the loss it incurs when evaluated on the counterfactual \(x^\prime\) at given parameter values.
 Recall that \(\text{extcost}(\cdot)\) denotes what we had defined as the external cost in Equation \eqref{eq:collective}.
-Then formally we let
+Formally, we let
 
@@
 
-Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average \(\bar{x}\):
+Yet another strategy extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the subsample average \(\bar{x}^*=\text{mean}(x)\), \(x \in \mathcal{D}_1\):
 
@@
 \begin{aligned}
-\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  \label{eq:grav}
+\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x}^*)  \label{eq:grav}
 \end{aligned}

### 2022-11-04 paper/sections/empirical.rmd
trimmed down to meet page limit

@@
 
-We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons**.
-The datasets are presented in Figure \@ref(fig:synthetic-data).
+We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons** (\@ref(fig:synthetic-data)).
 
@@
 The continuous outcome variable is binarized as $\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}$ indicating whether or not the median house price of a given district is above or below the median of all districts.
-While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example @schut2021generating).
+While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion @schut2021generating.
 While we initially also conducted experiments on that dataset, we eventually discarded this dataset due to surrounding ethical concerns [@carlisle2019racist].

### 2022-11-04 paper/sections/empirical_2.rmd
trimmed down to meet page limit

@@
 For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.].
-Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn).
+Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \@ref(fig:syn).
 The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.
 
-Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment.
+Finally, it is also worth noting that the observed dynamics and patterns are consistent throughout the course of the experiment.
 That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.
@@
 
-To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments (RQ \@ref(prp:shifts)) and we find them substantial enough to be considered costly (RQ \@ref(prp:costs)); secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect (RQ \@ref(prp:het)).
-The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process (RQ \@ref(prp:drive)).
+To recap, we answer our research questions: firstly, endogenous dynamics do emerge in our experiments (RQ \@ref(prp:shifts)) and we find them substantial enough to be considered costly (RQ \@ref(prp:costs)); secondly, the choice of the counterfactual generator matters, with Latent Space search generally having a dampening effect (RQ \@ref(prp:het)).
+The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to individuals and outcomes that comply with the data generating process (RQ \@ref(prp:drive)).

### 2022-11-04 paper/sections/introduction.rmd
trimmed down to meet page limit

@@
 
-Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of one single individual and has indeed been referred to as **individual recourse** in some places.
-Arguably though, most real-world applications that warrant Algorithmic Recourse involve potentially large groups of individuals typically competing for scarce resources.
-Our work demonstrates that in such scenarios, choices made by or for one single individual are likely to affect the broader collective of individuals in ways that current approaches to AR fail to account for.
-More specifically, we argue that a strict focus on minimizing the private costs faced by individuals may be too narrow an objective.
+Research on Algorithmic Recourse has also so far typically addressed the issue from the perspective of a single individual.
+Arguably though, most real-world applications that warrant AR involve potentially large groups of individuals typically competing for scarce resources.
+Our work demonstrates that in such scenarios, choices made by or for a single individual are likely to affect the broader collective of individuals in ways that many current approaches to AR fail to account for.
+More specifically, we argue that a strict focus on minimizing the private costs to individuals may be too narrow an objective.
 
@@
 
-We think that these types of endogenous dynamics may be problematic and warrant our attention.
+We think that these types of endogenous dynamics may be problematic and deserve our attention.
 From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse.
 Secondly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter.
-We would argue that in some practical applications, this can be expected to generate costs for the involved stakeholders.
+We think that in some practical applications, this can be expected to generate costs for involved stakeholders.
 To follow our argument, consider the following two examples:

### 2022-11-04 paper/sections/methodology_2.rmd
trimmed down to meet page limit

@@
 
-The dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro) were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
+The dynamics illustrated in Figure \@ref(fig:poc) in were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice.
 We begin in the static setting at time time $t=0$: firstly, we have some binary classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$, where $\mathcal{D}_0$ and $\mathcal{D}_1$ denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$).
-Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple, we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].
+Note that we focus our attention on classification problems, since classification poses the most common use-case for recourse^[To keep notation simple, we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].
 

### 2022-11-04 paper/sections/mitigation.rmd
incorporated some revsions
trimmed down to meet page limit

@@
 The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$.
-This threshold determines when a proposed counterfactual should be considered as valid.
+This threshold determines when a counterfactual should be considered as valid.
 Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction.
 Since this is the region of maximal aleatoric uncertainty, the classifier is bound to be thrown off.
-By simply setting a more conservative decision threshold, we can avoid this issue to some extent.
+By setting a more conservative threshold, we can avoid this issue to some extent.
 A drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.
@@
 
-Another strategy draws inspiration from ROAR @upadhyay2021robust: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
+Another strategy draws inspiration from ROAR @upadhyay2021robust: to preserve the classifier, we propose to explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values.
 Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective).
-Then formally we let
+Formally, we let
 
@@
 
-Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average $\bar{x}$:
+Yet another strategy extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the subsample average $\bar{x}^*=\text{mean}(x)$, $x \in \mathcal{D}_1$:
 
@@
 \begin{aligned}
-\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  (\#eq:grav)
+\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x}^*)  (\#eq:grav)
 \end{aligned}
@@
 Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is, Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClaPROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClaPROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].
+Linking these ideas back to Example \@ref(exm:student), the mitigation strategies help ensure that the recommended recourse actions are substantial enough to truly lead to an increase in the probability that the admitted student eventually graduates.
 

### 2022-11-08 paper/paper.tex
revision

@@
 Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\).
-Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x^\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x^\prime\) such that the predicted outcome \(M(x^\prime)\) corresponds to some target outcome \(y^*\):
+Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}.} \(C=\text{cost}(x^\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x^\prime\) such that the predicted outcome \(M(x^\prime)\) corresponds to some target outcome \(y^*\):
 
@@
 This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data.
-All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.
+All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger shifts.
 
@@
 Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
-To ensure that each generator is faced with the exact same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.
+To ensure that each generator is faced with the same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.
 
@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise, each individual experiment is repeat five times.\footnote{In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
+To account for noise, each individual experiment is repeated five times.\footnote{In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
 An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.}
@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}.
-Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}, although they need not be driven by adversarial intentions: in Example \ref{exm:student}, student applicants may dutifully focus on acquiring credentials that help them be admitted to university, but ultimately not improve their chances of success at completing their degree \protect\hyperlink{ref-barocas2017fairness}{{[}37{]}}.
+Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
 Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
@@
 While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice.
-Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}37{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}38{]}}).
+Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}38{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}39{]}}).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-barocas2017fairness}{}}%
+\CSLLeftMargin{{[}37{]} }%
+\CSLRightInline{S. Barocas, M. Hardt, and A. Narayanan, {``Fairness in machine learning,''} \emph{Nips tutorial}, vol. 1, p. 2, 2017.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
-\CSLLeftMargin{{[}37{]} }%
+\CSLLeftMargin{{[}38{]} }%
 \CSLRightInline{V.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022why}{}}%
-\CSLLeftMargin{{[}38{]} }%
+\CSLLeftMargin{{[}39{]} }%
 \CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} 2022. Available: \url{https://arxiv.org/abs/2207.08815}}

### 2022-11-08 paper/sections/empirical.rmd
revision

@@
 Evaluation metrics are computed and stored every 10 rounds.
-To account for noise, each individual experiment is repeat five times.^[In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
+To account for noise, each individual experiment is repeated five times.^[In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse.
 An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.]

### 2022-11-08 paper/sections/limitations.rmd
revision

@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-The implementation of such changes is typically described as **gaming** @miller2020strategic.
-Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them be admitted to university, but ultimately not improve their chances of success at completing their degree @barocas2017fairness.
+Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
 Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.

### 2022-11-08 paper/sections/methodology_2.rmd
revision

@@
 Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods.
-To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
+To ensure that each generator is faced with the same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.
 

### 2022-11-08 paper/sections/related.rmd
revision

@@
 Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$.
-Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:
+Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*.] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:
 

### 2022-11-08 paper/paper.tex
final things

@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, \protect\hyperlink{ref-deoliveira2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, \protect\hyperlink{ref-deoliveira2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}} \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
-In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \href{(https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)}{\texttt{AlgorithmicRecourseDynamics.jl}}\footnote{The code is available from the following anonymized Github repository: \url{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}.}.
+In particular, we have built a framework on top of \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}} that extends the functionality from static benchmarks to simulation experiments: \href{(https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)}{\texttt{AlgorithmicRecourseDynamics.jl}}\footnote{The code is available from the following anonymized Github repository: \url{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}.}.
 The core concepts implemented in that package reflect what is presented in Section \ref{method-2} of this paper.
@@
 Here, \(yloss(\cdot)\) is first broadcasted over elements of \(\mathbf{s}^\prime\) and then aggregated.
-This is exactly how counterfactual search is implemented in \texttt{CounterfactualExplanations.jl}.}.
+This is exactly how counterfactual search is implemented in \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}.}.
 In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and thereby produce counterfactuals \(s_1, ...
 , s_K\) that look as different from each other as possible.
-The implementation of DiCE in our library of choice---\texttt{CounterfactualExplanations.jl}---uses that exact approach.
+The implementation of DiCE in our library of choice---\href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}---uses that exact approach.
 It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.
@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}, although they need not be driven by adversarial intentions: in Example \ref{exm:student}, student applicants may dutifully focus on acquiring credentials that help them be admitted to university, but ultimately not improve their chances of success at completing their degree \protect\hyperlink{ref-barocas2017fairness}{{[}37{]}}.
+The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}, although they need not be driven by adversarial intentions: in Example \ref{exm:student}, student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree \protect\hyperlink{ref-barocas2017fairness}{{[}37{]}}.
 Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
@@
 Future work may benefit from including novel datasets or extending the analysis to multi-class or regression problems, the latter arguably representing the most common objective in Finance and Economics.
-It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing \texttt{CounterfactualExplanations.jl} only supports continuous features, at least of some of the counterfactual generators considered here.
+It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}} only supports continuous features, at least for some of the counterfactual generators considered here.
 The fact that we therefore had to discard discrete features led to relatively poor initial performance of our classifiers in some cases.
@@
 To this end we have open-sourced all of our code along with a Julia package: \href{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}{\texttt{AlgorithmicRecourseDynamics.jl}}.
-The package is built on top of \texttt{CounterfactualExplanations.jl} and inherits its extensibility \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}.
+The package is built on top of \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}} and inherits its extensibility \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}.
 That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-11-08 paper/sections/conclusion.rmd
final things

@@
 To this end we have open-sourced all of our code along with a Julia package: [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).
-The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022counterfactualexplanations].
+The package is built on top of [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl) and inherits its extensibility [@altmeyer2022counterfactualexplanations].
 That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-11-08 paper/sections/limitations.rmd
final things

--- 2022-11-03 paper/sections/limitations.rmd
@@
 The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome.
-The implementation of such changes is typically described as **gaming** @miller2020strategic.
-Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work.
+The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree @barocas2017fairness.
+Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
 Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
@@
 Future work may benefit from including novel datasets or extending the analysis to multi-class or regression problems, the latter arguably representing the most common objective in Finance and Economics.
-It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing `CounterfactualExplanations.jl` only supports continuous features, at least of some of the counterfactual generators considered here.
+It is also worth mentioning that the use of real-world datasets considered in this work is constrained by the fact that at the time of writing [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl) only supports continuous features, at least for some of the counterfactual generators considered here.
 The fact that we therefore had to discard discrete features led to relatively poor initial performance of our classifiers in some cases.

### 2022-11-08 paper/sections/methodology.rmd
revision
final things

@@
 Here, $yloss(\cdot)$ is first broadcasted over elements of $\mathbf{s}^\prime$ and then aggregated.
-This is exactly how counterfactual search is implemented in `CounterfactualExplanations.jl`.].
+This is exactly how counterfactual search is implemented in [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl).].
 In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and thereby produce counterfactuals $s_1, ...
 , s_K$ that look as different from each other as possible.
-The implementation of DiCE in our library of choice---`CounterfactualExplanations.jl`---uses that exact approach.
+The implementation of DiCE in our library of choice---[`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl)---uses that exact approach.
 It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

### 2022-11-08 paper/sections/related.rmd
final things

@@
 This may be partially due to limited software availability in this space.
-Recent work has started to address this gap: firstly, @deoliveira2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022counterfactualexplanations] provides an extensible, fast and language-agnostic implementation in Julia.
+Recent work has started to address this gap: firstly, @deoliveira2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework---CARLA---that can be used to apply and benchmark different methodologies; finally, [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl) [@altmeyer2022counterfactualexplanations] provides an extensible, fast and language-agnostic implementation in Julia.
 Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
-In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: [`AlgorithmicRecourseDynamics.jl`]((https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md))^[The code is available from the following anonymized Github repository: [https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).].
+In particular, we have built a framework on top of [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl) that extends the functionality from static benchmarks to simulation experiments: [`AlgorithmicRecourseDynamics.jl`]((https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md))^[The code is available from the following anonymized Github repository: [https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).].
 The core concepts implemented in that package reflect what is presented in Section \@ref(method-2) of this paper.

### 2022-12-14 paper/paper.Rmd
first go at camera ready

@@
   bookdown::pdf_book:
+    keep_tex:  true
     base_format: rticles::ieee_article

### 2022-12-14 paper/paper.Rmd
trying to sort out author details

@@
 
-  # institution:
-  #   - name: Georgia Institute of Technology
-  #     department: School of Electrical and Computer Engineering
-  #     location: Atlanta, Georgia 30332--0250
-  #     email: ece@datech.edu
-  #     mark: 1
-  #     author:
-  #       - name: Michael Shell
+  institution:
+    - name: |
+        *Delft University of Technology*
+      department: |
+        *Faculty of Electrical Engineering, Mathematics and Computer Science*
+      location: 2628 XE Delft, The Netherlands
+      mark: 1
+      author:
+        - name: Patrick Altmeyer
+          email: P.Altmeyer@tudelft.nl
+        - name: Giovan Angela
+          # email: G.J.A.Angela@student.tudelft.nl
+        - name: Aleksander Buszydlik
+          # email: A.J.Buszydlik@student.tudelft.nl
+        - name: Karol Dobiczek
+          # email: K.T.Dobiczek@student.tudelft.nl
+        - name: Arie van Deursen
+          # email: Arie.vanDeursen@tudelft.nl
+        - name: Cynthia C. S. Liem
+          # email: C.C.S.Liem@tudelft.nl
   #   - name: Twentieth Century Fox

### 2022-12-15 paper/paper.Rmd
trimmed down

@@
 abstract: |
-  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
-The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely focused on single individuals in a static environment: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge.
 There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
-Through this work we aim to close that gap by systematizing and extending existing knowledge.
+Through this work we aim to close that gap.
 We first show that many of the existing methodologies can be collectively described by a generalized framework.
-We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+We then argue that the existing framework does not account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
 Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve large groups of individuals.
-Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in some situations.
+Fortunately, we find various strategies to mitigate these concerns.
 Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-12-15 paper/paper.tex
first go at camera ready
trying to sort out author details
trimmed down

@@
 \newtheorem{corollary}{Corollary}[section]
-\newtheorem{proposition}{Research Question}[section]
+\newtheorem{proposition}{Proposition}[section]
 \newtheorem{conjecture}{Conjecture}[section]
@@
  %% -- beg if/affiliation.institution-columnar
+\IEEEauthorblockN{
+  %% -- beg for/affiliation.institution.author
+Patrick Altmeyer,
+  %% -- beg for/affiliation.institution.author
+Giovan Angela,
+  %% -- beg for/affiliation.institution.author
+Aleksander Buszydlik,
+  %% -- beg for/affiliation.institution.author
+Karol Dobiczek,
+  %% -- beg for/affiliation.institution.author
+Arie van Deursen,
+  %% -- beg for/affiliation.institution.author
+Cynthia C. S. Liem %% -- end for/affiliation.institution.author
+}
+\IEEEauthorblockA{\emph{Faculty of Electrical Engineering, Mathematics and Computer Science}\\
+\emph{Delft University of Technology}\\
+2628 XE Delft, The Netherlands
+  %% -- beg for/affiliation.institution.author
+\\P.Altmeyer@tudelft.nl
+  %% -- beg for/affiliation.institution.author
+  %% -- beg for/affiliation.institution.author
+  %% -- beg for/affiliation.institution.author
+  %% -- beg for/affiliation.institution.author
+  %% -- beg for/affiliation.institution.author
+ %% -- end for/affiliation.institution.author
+}
  %% -- end for/affiliation.institution
@@
 \begin{abstract}
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
-The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
+Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely focused on single individuals in a static environment: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
+The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge.
 There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
-Through this work we aim to close that gap by systematizing and extending existing knowledge.
+Through this work we aim to close that gap.
 We first show that many of the existing methodologies can be collectively described by a generalized framework.
-We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
+We then argue that the existing framework does not account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
 Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve large groups of individuals.
-Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
+We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in some situations.
+Fortunately, we find various strategies to mitigate these concerns.
 Our simulation framework for studying recourse dynamics is fast and open-sourced.
@@
 
-Both of these examples are exaggerated simplifications of potential real-world scenarios, but they serve to illustrate the point that recourse for one single individual may exert negative externalities on other individuals.
+Both examples are exaggerated simplifications of potential real-world scenarios, but they serve to illustrate the point that recourse for one single individual may exert negative externalities on other individuals.
 
@@
 For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent.
-This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we shall refer to simply as \textbf{Wachter}, the name of the first author, in the remainder of this paper.
+This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we refer to simply as \textbf{Wachter}, the name of the first author, in the remainder of this paper.
 
@@
 
-Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space.
-In Wachter, the state space is the feature space: \(f\) is just the identity function and the number of counterfactuals \(K\) is equal to one.
-Both REVISE and CLUE search counterfactuals in some latent embedding \(S \subset \mathcal{S}\) instead of the feature space directly.
+Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is a \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space.
+In Wachter, the state space is the feature space: \(f\) is the identity function and the number of counterfactuals \(K\) is one.
+Both REVISE and CLUE search counterfactuals in some latent space \(\mathcal{S}\) instead of the feature space directly.
 The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\).
-In this case, \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}.
+In this case, \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, that is the function that maps back from the latent space to inputs.
+Provided the generative model is well-specified, traversing the latent embedding typically yields meaningful counterfactuals, since they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}.
 
@@
 The greedy approach proposed by Schut et al.
-\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
+\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates.
 The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).
@@
 \begin{proposition}[Costs]
-\protect\hypertarget{prp:costs}{}\label{prp:costs}If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+\protect\hypertarget{prp:costs}{}\label{prp:costs}If so, are these dynamics substantial enough to be considered costly to stakeholders involved in real-world automated decision-making processes?
 \end{proposition}
@@
 As an alternative to PP MMD, we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement).
-This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers disagree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
-In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model.
+In our context, this metric quantifies the overlap between the initial model (trained before the application of AR) and the updated model.
 A Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured.
-This is one reason for why PP MMD is our our preferred metric.
+The opposite is not true: even if the Disagreement Coefficient is equal to zero, a model shift may still have occured.
+This is one reason for why PP MMD is our preferred metric.
 
@@
 
-For each dataset and generator, we look at three different types of classifiers, all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier---\textbf{Logistic Regression}---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier.
+For each dataset and generator, we look at three different types of classifiers, all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier---\textbf{Logistic Regression}---implemented as a single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier.
 We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
 They are also the model of choice in \protect\hyperlink{ref-schut2021generating}{{[}6{]}}.
-The actual neural network architectures are kept simple (top half of Table \ref{tab:architecture}), since we are only marginally concerned with achieving good initial classifier performance.
+The network architectures are kept simple (top half of Table \ref{tab:architecture}), since we are only marginally concerned with achieving good initial classifier performance.
 
@@
 We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.}.
-In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.
+In summary, we expect that the endogenous shifts induced by Wachter outsize those of all other generators, since Wachter is not explicitly concerned with generating what we have defined as meaningful counterfactuals.
 
@@
 See Section \ref{limit-data} for details.} and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
-Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}} and \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+Both datasets have been used in the literature on AR before (see for example \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}} and \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
 
@@
 \item
-  The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible.
+  The observed dynamics are generally of the highest magnitude for the linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible.
 \item
-  The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial performance of the classifier was weak to begin with.
+  The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial model performance was weak to begin with.
 \item
@@
 As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
-In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
-
-The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
+In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space (LS) generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
+
+The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for LS generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
 Related to this, the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
@@
 The findings for the other synthetic datasets are broadly consistent with the observations above.
-For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
-For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.}.
+For the Moons data the same broad patterns emerge, although in this case it is less evident that LS generators induce relatively smaller shifts.
+For the Circles data, it also appears at first sight that LS search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.}.
 Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \ref{fig:syn}.
@@
 Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts.
-While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
+While LS search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
 
 The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: the estimated PP MMD is statistically significant and large for all datasets.
-Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
+Here we find no evidence that LS search helps to mitigate model shifts, as we did before for the synthetic data.
 Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.
@@
 All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process.
-This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section.
+This principle is really at the core of Latent Space (LS) generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section.
 But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and, perhaps more importantly, their performance hinges on the performance of said generative models.
@@
 for each counterfactual \(k\) where \(l\) denotes the loss function used to train \(M\).
-This approach, which we shall refer to as \textbf{ClaPROAR}, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+This approach, which we refer to as \textbf{ClaPROAR}, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
 It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
@@
 
-Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts.
+Our findings indicate that all three mitigation strategies are at least at par with LS generators with respect to their effectiveness at mitigating domain and model shifts.
 Figure \ref{fig:mitigate-results} presents a subset of the evaluation metrics for our synthetic data with overlapping classes.
-The top row in Figure \ref{fig:mitigate-results} indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
+The top row in Figure \ref{fig:mitigate-results} indicates that while domain shifts are of roughly the same magnitude for both Wachter and LS generators, our proposed strategies effectively mitigate these shifts.
 ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
-As evident from the middle row in Figure \ref{fig:mitigate-results} model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
+As evident from the middle row in Figure \ref{fig:mitigate-results} model shifts can also be reduced: for the deep ensemble LS search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
 The same overall pattern can be observed for out-of-sample model performance.
@@
 
-An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
-In Figure \ref{fig:mitigate-latent-results} the conventional Latent Space generator with \(\gamma=0.5\) serves as our baseline.
+An interesting finding is also that the proposed strategies have a complementary effect when used in combination with LS generators.
+In experiments we conducted on the synthetic data, the benefits of LS generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
+In Figure \ref{fig:mitigate-latent-results} the conventional LS generator with \(\gamma=0.5\) serves as our baseline.
 Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.
@@
 
-\caption{Combinining various mitigation strategies with Latent Space search.
+\caption{Combinining various mitigation strategies with LS search.
 Results for synthetic data with overlapping classes.
@@
 The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}, although they need not be driven by adversarial intentions: in Example \ref{exm:student}, student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree \protect\hyperlink{ref-barocas2017fairness}{{[}37{]}}.
-Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
-Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
+Preventing such actions may help to avoid the dynamics we have pointed to in this work.
+Future work would likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
 \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

### 2022-12-15 paper/sections/empirical.rmd
trimmed down

@@
 
-For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
+For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as a single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier.
 We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty.
 They are also the model of choice in @schut2021generating.
-The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.
+The network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.
 
@@
 We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.].
-In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.
+In summary, we expect that the endogenous shifts induced by Wachter outsize those of all other generators, since Wachter is not explicitly concerned with generating what we have defined as meaningful counterfactuals.
 
@@
 See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding.
-Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
+Both datasets have been used in the literature on AR before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.
 

### 2022-12-15 paper/sections/empirical_2.rmd
trimmed down

@@
 
-- The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
-- The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial performance of the classifier was weak to begin with.
+- The observed dynamics are generally of the highest magnitude for the linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
+- The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial model performance was weak to begin with.
 - With the exception of the Greedy generator, all other generators generally perform somewhat better overall than the baseline (Wachter) as expected.
@@
 As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small.
-In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
+In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space (LS) generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy.
 
-The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
+The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for LS generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average.
 Related to this, the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals.
@@
 The findings for the other synthetic datasets are broadly consistent with the observations above.
-For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts.
-For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.].
+For the Moons data the same broad patterns emerge, although in this case it is less evident that LS generators induce relatively smaller shifts.
+For the Circles data, it also appears at first sight that LS search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.].
 Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \@ref(fig:syn).
@@
 Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts.
-While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
+While LS search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.
 
 The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): the estimated PP MMD is statistically significant and large for all datasets.
-Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data.
+Here we find no evidence that LS search helps to mitigate model shifts, as we did before for the synthetic data.
 Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

### 2022-12-15 paper/sections/limitations.rmd
trimmed down

@@
 The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree @barocas2017fairness.
-Preventing such actions may help to avoid the macrodynamics we have pointed to in this work.
-Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
+Preventing such actions may help to avoid the dynamics we have pointed to in this work.
+Future work would likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al.
 @karimi2021algorithmic.

### 2022-12-15 paper/sections/methodology.rmd
trimmed down

@@
 
-Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
-In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one.
-Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly.
+Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is a $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space.
+In Wachter, the state space is the feature space: $f$ is the identity function and the number of counterfactuals $K$ is one.
+Both REVISE and CLUE search counterfactuals in some latent space $\mathcal{S}$ instead of the feature space directly.
 The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$.
-In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space.
-Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019realistic].
+In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, that is the function that maps back from the latent space to inputs.
+Provided the generative model is well-specified, traversing the latent embedding typically yields meaningful counterfactuals, since they are implicitly generated by the (learned) DGP [@joshi2019realistic].
 
@@
 The greedy approach proposed by Schut et al.
-@schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
+@schut2021generating, which we refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates.
 The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA).

### 2022-12-15 paper/sections/methodology_2.rmd
trimmed down

@@
 ::: {.proposition #costs name="Costs"}
-If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
+If so, are these dynamics substantial enough to be considered costly to stakeholders involved in real-world automated decision-making processes?
 :::
@@
 As an alternative to PP MMD, we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement).
-This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample.
+This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers disagree on the predicted outcome for a randomly chosen sample.
 Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries.
-In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model.
+In our context, this metric quantifies the overlap between the initial model (trained before the application of AR) and the updated model.
 A Disagreement Coefficient unequal to zero is indicative of a model shift.
-The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured.
-This is one reason for why PP MMD is our our preferred metric.
+The opposite is not true: even if the Disagreement Coefficient is equal to zero, a model shift may still have occured.
+This is one reason for why PP MMD is our preferred metric.
 

### 2022-12-15 paper/sections/mitigation.rmd
trimmed down

@@
 for each counterfactual $k$ where $l$ denotes the loss function used to train $M$.
-This approach, which we shall refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
+This approach, which we refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss.
 It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers.
@@
 
-Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts.
+Our findings indicate that all three mitigation strategies are at least at par with LS generators with respect to their effectiveness at mitigating domain and model shifts.
 Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes.
-The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts.
+The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and LS generators, our proposed strategies effectively mitigate these shifts.
 ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts.
-As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
+As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble LS search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective.
 The same overall pattern can be observed for out-of-sample model performance.
@@
 
-An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators.
-In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
-In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline.
+An interesting finding is also that the proposed strategies have a complementary effect when used in combination with LS generators.
+In experiments we conducted on the synthetic data, the benefits of LS generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR.
+In Figure \@ref(fig:mitigate-latent-results) the conventional LS generator with $\gamma=0.5$ serves as our baseline.
 Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.
 
-```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with Latent Space search.
+```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with LS search.
 Results for synthetic data with overlapping classes.

### 2022-12-15 paper/sections/related.rmd
trimmed down

@@
 For models that are differentiable and produce smooth predictions, \@ref(eq:solution) can be solved through gradient descent.
-This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter**, the name of the first author, in the remainder of this paper.
+This summarizes the approach followed in @wachter2017counterfactual which we refer to simply as **Wachter**, the name of the first author, in the remainder of this paper.
 

### 2022-12-16 README.qmd
sorted out package deps and compat

--- 2022-08-19 README.qmd
@@
 format: 
-    gfm:
+  commonmark:
+    variant: -raw_html
     wrap: none
-        html-math-method: webtex
+execute: 
+  freeze: auto
+  echo: true
+  eval: true
+  output: false
+crossref:
+  fig-prefix: Figure
+  tbl-prefix: Table
+bibliography: bib.bib
+jupyter: julia-1.8
 ---
 
+[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/stable)
+[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/dev)
+[![Build Status](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml?query=branch%3Amain)
+[![Coverage](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl)
+[![Code Style: Blue](https://img.shields.io/badge/code%20style-blue-4495d1.svg)](https://github.com/invenia/BlueStyle)
+[![ColPrac: Contributor's Guide on Collaborative Practices for Community Packages](https://img.shields.io/badge/ColPrac-Contributor's%20Guide-blueviolet)](https://github.com/SciML/ColPrac)
+[![Twitter Badge](https://img.shields.io/twitter/url/https/twitter.com/paltmey.svg?style=social&label=Follow%20%40paltmey)](https://twitter.com/paltmey)
+
 # AlgorithmicRecourseDynamics
-
-`AlgorithmicRecourseDynamics.jl` is a Julia package for modelling Algorithmic Recourse Dynamics.
-
-## Research Paper 📝
-
-**Note** ⚠: You are browsing the (anonymised) [`#original-paper`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) branch of `AlgorithmicRecourseDynamics.jl`.
-This branch is a static artifact corresponding to the state of the package at the time the paper was first published.
-It can be used to replicate the original findings of the paper.
-Only this branch is currently accessible as an anonymised git repository.
-The main repository is private and will will be open-sourced after the review process.
-
-## At a Glance
-
-The paper titles **Endogenous Macrodynamics in Algorithmic Recourse** is currently under review and not yet published. You can find
-a preprint along with other resources right here on this branch of the
-repository:
-
-- [Paper](paper/paper.pdf)
-- [Notebooks](dev/notebooks/)
-- [Supplementary Appendix](build/dev/notebooks/appendix.html) generated from notebooks (download the HTML and view in browser)
-- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review process)
-
-In this work we investigate what happens if Algorithmic Recourse is actually implemented by a large number of individuals.
-The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class (y=0) are marked in blue and samples of the positive class (y=1) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.
-
-![](paper/www/poc.png)
-
-## Paper Abstract
-
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
-The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point.
-There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
-Through this work we aim to close that gap by systematizing and extending existing knowledge.
-We first show that many of the existing methodologies can be collectively described by a generalized framework.
-We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
-Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in situations that involve competition for scarce resources.
-Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches.
-Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-12-16 docs/src/_intro.qmd
sorted out package deps and compat

`AlgorithmicRecourseDynamics.jl` is a Julia package for modeling Algorithmic Recourse Dynamics.

## Research Paper 📝

**Note** ⚠: You are browsing the [`#original-paper`](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/tree/original-paper) branch of `AlgorithmicRecourseDynamics.jl`. This branch is a static artifact corresponding to the state of the package at the time the paper was first published. It can be used to replicate the original findings of the paper.

## At a Glance

You can find resources relevant to the paper right here on this branch of the
repository:

- [Paper](paper/paper.pdf)
- [Notebooks](dev/notebooks/)
- [Supplementary Appendix](build/dev/notebooks/appendix.html) generated from notebooks (download the HTML and view in browser)
- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review process)

In this work we investigate what happens if Algorithmic Recourse is actually implemented by a large number of individuals. The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class (y=0) are marked in blue and samples of the positive class (y=1) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.

![](paper/www/poc.png)

## Paper Abstract

Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely focused on single individuals in a static environment: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework does not account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in some situations. Fortunately, we find various strategies to mitigate these concerns. Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-12-16 docs/src/paper/experiments/synthetic.qmd
sorted out package deps and compat

--- 2022-08-17 dev/notebooks/experiments/synthetic.qmd
@@
 
-### Tables
-
-@tbl-results shows a summary of all results.
+### Bootstrap
 
 ```{julia}
-#| label: tbl-results
-#| tbl-cap: "Summary of all results"
-#| eval: true
-#| results: asis
-using AlgorithmicRecourseDynamics: kable
-kable(results, [50]; format="html")
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in 1:length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.data .= key
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap.csv")
+using CSV
+CSV.write(save_path, df)
 ```

### 2022-12-16 docs/src/paper/generators/gravitational_generator.qmd
sorted out package deps and compat

## Gravitational Generator

```{julia}
#| eval: true
using Pkg; Pkg.activate("dev")
```

```{julia}
#| eval: true
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("generator")
www_path = www_dir("generator")
```

```{julia}
#| eval: true
using MLJ
N = 1000
X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-5 => 5), cluster_std=0.5)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')
```

```{julia}
#| eval: true
using Flux
nn = Chain(Dense(2,1))
using Flux.Optimise: update!, ADAM
opt = ADAM()
epochs = 100
loss(x, y) = Flux.Losses.logitbinarycrossentropy(nn(x), y)
avg_loss(data) = mean(map(d -> loss(d[1],d[2]), data))
show_every = epochs/10
for epoch = 1:epochs
  for d in data
    gs = gradient(Flux.params(nn)) do
      l = loss(d...)
    end
    update!(opt, Flux.params(nn), gs)
  end
  if epoch % show_every == 0
    println("Epoch " * string(epoch))
    @show avg_loss(data)
  end
end
```

```{julia}
#| eval: true
M = FluxModel(nn)
```

```{julia}
#| eval: true
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
T = 100
```

```{julia}
#| eval: true
Λ₂ = [0.1, 1, 5]
counterfactuals_strict = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals_strict = vcat(
      counterfactuals_strict...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; T=T)
    )
end
```

```{julia}
#| eval: true
#| fig-cap: |
#|    Generating counterfactuals through `GravitationalGenerator`. The magnitude of attraction ($\lambda_2) increases from left to right.
#| label: grav
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals_strict[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
display(plt)
```

### 2022-12-16 docs/src/paper/proof_of_concept.qmd
sorted out package deps and compat

---
title: Proof of concept
jupyter: julia-1.7
---

```{julia}
using Pkg; Pkg.activate("dev")
```

In this notebook we will use toy data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.

```{julia}
include("dev/utils.jl")
using AlgorithmicRecourseDynamics
using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
theme(:wong)
output_path = output_dir("poc")
www_path = www_dir("poc")
```

## Classifiers

```{julia}
using MLJ
N = 1000
X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-2 => 2), cluster_std=0.1)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')
```

### Logistic Regression

```{julia}
using Flux
model = Chain(Dense(2,1))
M = FluxModel(model)
```

```{julia}
n_epochs = 10
using AlgorithmicRecourseDynamics.Models
M = Models.train(M,counterfactual_data; n_epochs=n_epochs)
```

```{julia}
plt_nn = plot(M,counterfactual_data;zoom=0)
savefig(plt_nn, joinpath(www_path, "nn_contour.png"))
```

![](../artifacts/upload/www/synthetic/nn_contour.png)

### Laplace Redux

```{julia}
λ = 0.1
model_laplace = Laplace(M.model, λ=λ)
LaplaceRedux.fit!(model_laplace, data)
```

```{julia}
M_laplace = Models.LaplaceModel(model_laplace)
```

```{julia}
plt_laplace = plot(M_laplace, counterfactual_data;zoom=0)
savefig(plt_laplace, joinpath(www_path, "la_contour.png"))
```

![](../artifacts/upload/www/synthetic/la_contour.png)

## Single Round

```{julia}
# Models:
models = (Bayesian=M_laplace, Plugin=M)
# Generators:
generators = (
    Greedy=GreedyGenerator(loss=:logitbinarycrossentropy), 
    Generic=GenericGenerator(loss=:logitbinarycrossentropy),
    REVISE=REVISEGenerator(loss=:logitbinarycrossentropy)
)
```

### Generate counterfactual

```{julia}
cb = false
γ = 0.75
mod = M_laplace
gen = generators[:Greedy]
plt_original = plot(mod,counterfactual_data;zoom=0,colorbar=false,title="(a)")
```

```{julia}
μ = 0.10
candidates = findall(ys.==0)
chosen = rand(candidates, Int(round(μ*length(candidates))))
X′ = copy(X)
y′ = copy(ys)
using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
for i in chosen
    x = X[:,i]
    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen,γ=γ)
    X′[:,i] = counterfactual(outcome)
    y′[i] = first(counterfactual_label(outcome))
end
counterfactual_data′ = CounterfactualData(X′,y′')
plt_single = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(b)")
```

### Retrain

```{julia}
mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
plt_single_retrained = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(c)")
```

### Repeat

```{julia}
i = 2
while i <= 10
    counterfactual_data′ = CounterfactualData(X′,y′')
    candidates = findall(y′.==0)
    chosen = rand(candidates, Int(round(μ*length(candidates))))
    H₀ = mod.model.H # use posterior as new prior
    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs, H₀=H₀)
    for i in chosen
        x = X′[:,i]
        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen; γ=γ)
        X′[:,i] = counterfactual(outcome)
        y′[i] = first(counterfactual_label(outcome))
    end
    i += 1
end
plt_single_repeat = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(d)")
```

```{julia}
plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
savefig(plt, joinpath(www_path, "poc.png"))
```

![](../artifacts/upload/www/synthetic/poc.png)

### 2022-12-16 docs/src/paper/index.qmd
moving into docs

---
title: Endogenous Macrodynamics in Algorithmic Recourse
subtitle: Research Paper
---

```@meta
CurrentModule = ConformalPrediction
```

This part of the documentation is specific to the research paper "Endogenous Macrodynamics in Algorithmic Recourse". To allow for reproducibility of the results at the time of publication, there is a specific package environment linked to this part of the documentation that **will not be updated**. It can be activated as follows:

```{julia}
using Pkg; Pkg.activate("docs/src/paper")
```

The status of this package environment is shown below:

```{julia}
#| output: true

Pkg.status()
```

### 2022-12-16 docs/src/paper/proof_of_concept.qmd
moving into docs

@@
 ```{julia}
-using Pkg; Pkg.activate("dev")
+#| echo: false
+
+# Environment:
+using Pkg; Pkg.activate("docs/src/paper")
+
+# Deps:
+using CounterfactualExplanations
+using Flux
+using LaplaceRedux
+using MLJBase
+using Plots
+using Random
+
+# Setup
+Random.seed!(1234)
+theme(:wong)
+include("docs/src/utils.jl")    # some helper functions
+output_path = output_dir("poc") # output directory for artifacts
+www_path = www_dir("poc")       # output directory for images
 ```
 
-In this notebook we will use toy data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.
-
-```{julia}
-include("dev/utils.jl")
-using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
-theme(:wong)
-output_path = output_dir("poc")
-www_path = www_dir("poc")
-```
+To start with, we will look at a proof-of-concept that demonstrates the main observation underlying that paper is framed around.
+In particular, we will use synthetic data to see how endogenous domain shifts and the resulting model shifts can have implications on the validity and cost of algorithmic recourse.
 
@@
 ```{julia}
-using MLJ
 N = 1000
-X, ys = make_blobs(N, 2; centers=2, as_table=false, center_box=(-2 => 2), cluster_std=0.1)
+xmax = 2
+X, ys = make_blobs(
+    N, 2; 
+    centers=2, as_table=false, center_box=(-xmax => xmax), cluster_std=0.1
+)
 ys .= ys.==2
@@
 ```{julia}
-using Flux
+n_epochs = 10
 model = Chain(Dense(2,1))
-M = FluxModel(model)
+Models.forward!(model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
 ```
@@
 ```{julia}
-n_epochs = 10
-using AlgorithmicRecourseDynamics.Models
-M = Models.train(M,counterfactual_data; n_epochs=n_epochs)
+model = Laplace(model, likelihood=:classification)
+LaplaceRedux.fit!(model, data)      # fit Laplace Approximation
+optimize_prior!(model)              # optimize prior
+mod = LaplaceReduxModel(model)
 ```
@@
 ```{julia}
-plt_nn = plot(M,counterfactual_data;zoom=0)
-savefig(plt_nn, joinpath(www_path, "nn_contour.png"))
+#| output: true
+#| label: fig-model
+#| fig-cap: "The baseline model: contours indicate the predicted label; dots indicate observed data points."
+
+plt_original = plot(mod, counterfactual_data; zoom=0, colorbar=false, title="(a)")
 ```
 
-![](../artifacts/upload/www/synthetic/nn_contour.png)
-
-### Laplace Redux
-
-```{julia}
-λ = 0.1
-model_laplace = Laplace(M.model, λ=λ)
-LaplaceRedux.fit!(model_laplace, data)
-```
-
-```{julia}
-M_laplace = Models.LaplaceModel(model_laplace)
-```
-
-```{julia}
-plt_laplace = plot(M_laplace, counterfactual_data;zoom=0)
-savefig(plt_laplace, joinpath(www_path, "la_contour.png"))
-```
-
-![](../artifacts/upload/www/synthetic/la_contour.png)
-
 ## Single Round
-
-```{julia}
-# Models:
-models = (Bayesian=M_laplace, Plugin=M)
-# Generators:
-generators = (
-    Greedy=GreedyGenerator(loss=:logitbinarycrossentropy), 
-    Generic=GenericGenerator(loss=:logitbinarycrossentropy),
-    REVISE=REVISEGenerator(loss=:logitbinarycrossentropy)
-)
-```
 
@@
 ```{julia}
-cb = false
 γ = 0.75
-mod = M_laplace
-gen = generators[:Greedy]
-plt_original = plot(mod,counterfactual_data;zoom=0,colorbar=false,title="(a)")
+gen = GenericGenerator(;decision_threshold=γ)
 ```
@@
     x = X[:,i]
-    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen,γ=γ)
+    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen)
     X′[:,i] = counterfactual(outcome)
@@
 ```{julia}
-mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs)
+Models.forward!(mod.model.model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
+LaplaceRedux.fit!(mod.model, data)      # fit Laplace Approximation
+optimize_prior!(mod.model)              # optimize prior
 plt_single_retrained = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(c)")
@@
     H₀ = mod.model.H # use posterior as new prior
-    mod = Models.train(mod, counterfactual_data′; n_epochs=n_epochs, H₀=H₀)
+    Models.forward!(mod.model.model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
+    LaplaceRedux.fit!(mod.model, data)      # fit Laplace Approximation
+    optimize_prior!(mod.model)              # optimize prior
     for i in chosen
         x = X′[:,i]
-        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen; γ=γ)
+        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen)
         X′[:,i] = counterfactual(outcome)
@@
 plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
-savefig(plt, joinpath(www_path, "poc.png"))
+# savefig(plt, joinpath(www_path, "poc.png"))
+```
+
+```{julia}
+
 ```

### 2022-12-17 docs/src/paper/proof_of_concept.qmd
bleh

@@
 title: Proof of concept
-jupyter: julia-1.7
 ---
@@
 # Deps:
+using AlgorithmicRecourseDynamics
+using AlgorithmicRecourseDynamics.Models
 using CounterfactualExplanations
@@
 using LaplaceRedux
+using Markdown
 using MLJBase
@@
 # Setup
-Random.seed!(1234)
+Random.seed!(2023)
 theme(:wong)
@@
 
-## Classifiers
+## Data
+
+We begin by generating the synthetic data for a simple binary classification problem.
+For illustrative purposes we will use data that is linearly separable.
+The chart below shows the data $\mathcal{D}$ at time zero, before any implementation of recourse.
 
 ```{julia}
+#| output: true
+#| label: fig-data
+#| fig-cap: "Linearly separable synthetic data"
+
 N = 1000
@@
 counterfactual_data = CounterfactualData(X,ys')
+plot()
+scatter!(counterfactual_data)
 ```
 
-### Logistic Regression
+## Classifier
+
+To model this data $\mathcal{D}$ we will use a linear classifier. In particular, as in the paper, we will build a logistic regression model in `Flux.jl`: a single layer with sigmoid activation.
 
 ```{julia}
-n_epochs = 10
+n_epochs = 100
 model = Chain(Dense(2,1))
-Models.forward!(model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
+mod = FluxModel(model)
+Models.train(mod, counterfactual_data; n_epochs=n_epochs)
 ```
 
-```{julia}
-model = Laplace(model, likelihood=:classification)
-LaplaceRedux.fit!(model, data)      # fit Laplace Approximation
-optimize_prior!(model)              # optimize prior
-mod = LaplaceReduxModel(model)
-```
+The chart (a) below shows the linear separation of the two classes.
 
@@
 
-## Single Round
+## Implementation of Recourse
 
@@
 ```{julia}
-γ = 0.75
-gen = GenericGenerator(;decision_threshold=γ)
+γ = 0.50
+μ = 0.10
+Markdown.parse(
+    """
+    To generate counterfactual explanations we will rely on the most generic approach.
+As our decision threshold we will use $(γ*100)% here.
+In other words, the counterfactual is considered as valid, as soon as the classifier is more convinced that it belongs to the target class (blue) than the non-target class (orange).
+In each round we will implement recourse for $(μ * 100)% of the individuals in the non-target class.
+    """
+)
 ```
@@
 ```{julia}
-μ = 0.10
+opt = Flux.Adam(0.01)
+gen = GenericGenerator(;decision_threshold=γ, opt=opt)
+```
+
+The chart (b) below shows the recourse outcome, which we denote here as $\mathcal{D}^{\prime}$.
+The obvious observation at this point is that the resulting counterfactuals, while valid, are clearly distinguishable from the factuals that were always in the target class.
+This is not a new observation and nor is it entirely surprising.
+In fact, a lot of recent work in this field has tried to address this issue.
+In this work we wonder what happens when we let these sorts of dynamics play out further in practice.
+While the outcome in (b) is not surprising, it may be much harder to observe so clearly it in practice (when the data is more complex).
+
+```{julia}
+#| output: true
+#| label: fig-round-1
+#| fig-cap: "The recourse outcome after one round."
+
 candidates = findall(ys.==0)
-chosen = rand(candidates, Int(round(μ*length(candidates))))
+chosen_individuals = rand(candidates, Int(round(μ*length(candidates))))
 X′ = copy(X)
@@
 using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
-for i in chosen
-    x = X[:,i]
-    outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen)
-    X′[:,i] = counterfactual(outcome)
-    y′[i] = first(counterfactual_label(outcome))
-end
+factuals = select_factual(counterfactual_data,chosen_individuals)
+outcome = generate_counterfactual(factuals, 1, counterfactual_data, mod, gen; initialization=:identity)
+X′[:,chosen_individuals] = reduce(hcat, @.(selectdim(counterfactual(outcome), 3, 1)))
+y′[chosen_individuals] = reduce(vcat,@.(selectdim(counterfactual_label(outcome),3,1)))
 counterfactual_data′ = CounterfactualData(X′,y′')
@@
 
+Suppose the agent in charge of the black-box system has provided recourse to a share of individuals leading to the outcome in chart (b).
+In practice, models are regularly updated through retraining to account for concept drift, for example.
+For our experiments, we assume that the agent accepts $\mathcal{D}^{\prime}$ as its new ground truth.
+To isolate the endogenous effects we are interested in here from any other effect, we further assume away any exogenous changes to the data that we might expect to occur in practice.
+Retraining the model on $\mathcal{D}^{\prime}$ leads to a shift of the decision boundary **in the direction of the non-target class**.
+
 ```{julia}
-Models.forward!(mod.model.model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
-LaplaceRedux.fit!(mod.model, data)      # fit Laplace Approximation
-optimize_prior!(mod.model)              # optimize prior
+#| output: true
+#| label: fig-retrain
+#| fig-cap: "The retrained model."
+
+mod = Models.train(mod, counterfactual_data′)
 plt_single_retrained = plot(mod,counterfactual_data′;zoom=0,colorbar=false,title="(c)")
@@
 ### Repeat
+
+We finally go on to repeat this process of recourse followed by model updates for multiple round.
+The chart below presents the different stages of the experiment side-by-side, where panel (d) represents the outcome after ten rounds.
+
+At first glance it seems that costs to individuals seeking recourse are gradually reduced as the decision boundary moves into the direction of the non-target class: they need to exert less effort to move to valid counterfactual states.
+The problem with this idea is, of course, that there is no free lunch.
+This reduction inflicts a burden on the agent in charge of the black-box: the group of individuals that is now classified as target class individuals looks entirely different from the original group.
+
+Why is this a problem?
+Let's, for example, that the two synthetic features accurately describe the credit worthiness of individual seeking loans, where credit-worthiness increases in the South-West direction.
+Non-target class individuals (orange) are denied credit, while target class individuals (blue) receive a loan.
+Then the population of borrowers in (d) is much more risky than in (a).
+Clearly, any lender (bank) aware of such dynamics would avoid them in practice.
+They might choose not to offer recourse in the first place, generating a cost to all individuals seeking recourse.
+Alternatively, they may reward first movers, but stop offering recourse after a few rounds.
+
+This last point makes it clear that the implementation of recourse by one individual may generate external costs for other individuals. This notion motivates the ideas set out in the paper.
 
@@
     candidates = findall(y′.==0)
-    chosen = rand(candidates, Int(round(μ*length(candidates))))
-    H₀ = mod.model.H # use posterior as new prior
-    Models.forward!(mod.model.model, data; loss=:logitbinarycrossentropy, opt=:Adam, n_epochs=n_epochs)
-    LaplaceRedux.fit!(mod.model, data)      # fit Laplace Approximation
-    optimize_prior!(mod.model)              # optimize prior
-    for i in chosen
-        x = X′[:,i]
-        outcome = generate_counterfactual(x, 1, counterfactual_data, mod, gen)
-        X′[:,i] = counterfactual(outcome)
-        y′[i] = first(counterfactual_label(outcome))
-    end
+    chosen_individuals = rand(candidates, Int(round(μ*length(candidates))))
+    Models.train(mod, counterfactual_data′)
+    factuals = select_factual(counterfactual_data′,chosen_individuals)
+    outcome = generate_counterfactual(factuals, 1, counterfactual_data′, mod, gen; initialization=:identity)
+    X′[:,chosen_individuals] = reduce(hcat, @.(selectdim(counterfactual(outcome), 3, 1)))
+    y′[chosen_individuals] = reduce(vcat,@.(selectdim(counterfactual_label(outcome),3,1)))
     i += 1
@@
 ```{julia}
+#| output: true
+#| label: fig-final
+#| fig-cap: "The different stages of the experiment."
+
 plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
-# savefig(plt, joinpath(www_path, "poc.png"))
+savefig(plt, joinpath(www_path, "poc.png"))
+plt
 ```
-
-```{julia}
-
-```
-
-![](../artifacts/upload/www/synthetic/poc.png)

### 2022-12-17 docs/make.jl

@@ -23,6 +23,7 @@ makedocs(;
         "🏠 Home" => "index.md",
         "🎓 Research Paper" => [
             "Overview" => "paper/index.md",
+            "Proof-of-Concept" => "paper/proof_of_concept.md",
         ],
         "🧐 Reference" => "_reference.md",
     ],

### 2022-12-19 docs/src/paper/appendix.qmd
verifying that results check out with updated package

---
title: Supplementary Appendix
---

This is a supplementary appendix to the research paper **Endogenous Macrodynamics in Algorithmic Recourse**. It contains all of the experimental results, including those not highlighted in the actual paper. It also links to additional information about the proposed mitigation strategies.

## Experimental Results {#sec-results}

{{< include experiments/_synthetic.qmd >}}

{{< include experiments/_real_world.qmd >}}

{{< include experiments/_mitigation_strategies.qmd >}}

## Generators {#sec-generators}

The generators have been moved into [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl). For more information so the package [documentation](https://pat-alt.github.io/CounterfactualExplanations.jl/dev).

### 2022-12-19 docs/src/paper/experiments/_real_world.qmd
verifying that results check out with updated package

--- 2022-08-17 dev/notebooks/experiments/real_world.qmd
@@
 
+### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
+using DataFrames
+df = DataFrame()
+for (key, val) in results
+    n_folds = length(val.experiment.recourse_systems)
+    for fold in 1:n_folds
+        for i in 1:length(val.experiment.system_identifiers)
+            rec_sys = val.experiment.recourse_systems[fold][i]
+            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+            df_.data .= key
+            df_.model .= model_name
+            df_.generator .= gen_name
+            df_.fold .= fold
+            df = vcat(df, df_)
+        end
+    end
+end
+df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+using RCall
+save_path = joinpath(output_path, "bootstrap.csv")
+using CSV
+CSV.write(save_path, df)
+```
+
 ### Chart in paper

### 2022-12-19 docs/src/paper/experiments/_synthetic.qmd
verifying that results check out with updated package

--- 2022-12-16 docs/src/paper/experiments/synthetic.qmd
@@
 
-This notebook was used to run the experiments for the synthetic datasets. In the following we first run the experiments and then generate visualizations and tables.
+This notebook was used to run the experiments for the synthetic datasets and can be used to reproduce the results in the paper.
+In the following we first run the experiments and then generate visualizations and tables.
 
@@
 ```{julia}
-#| eval: true
-using Pkg; Pkg.activate("dev")
-```
+#| echo: false
 
-```{julia}
-#| eval: true
-include("dev/utils.jl")
-using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra, Images
-theme(:wong)
-output_path = output_dir("synthetic")
-www_path = www_dir("synthetic");
+include("docs/src/paper/setup.jl")
+eval(setup)
 ```
@@
 
+@fig-test-before shows the test data before running the experiment.
+
 ```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
+#| label: fig-test-before
+#| fig-cap: "Test data before experiment"
+#| eval: true
+
 plts = []
@@
 savefig(plt, joinpath(www_path,"models_test_before.png"))
+plt
 ```
 
-@fig-test-before shows the test data before running the experiment.
+@fig-train-before shows the training data before running the experiment.
 
 ```{julia}
-#| label: fig-test-before
-#| fig-cap: "Test data before experiment"
+#| label: fig-train-before
+#| fig-cap: "Training data before experiment"
 #| eval: true
-load(joinpath(www_path,"models_test_before.png"))
-```
 
-```{julia}
 using AlgorithmicRecourseDynamics.Models: model_evaluation
@@
 savefig(plt, joinpath(www_path,"models_train_before.png"))
-```
-
-@fig-train-before shows the training data before running the experiment.
-
-```{julia}
-#| label: fig-train-before
-#| fig-cap: "Training data before experiment"
-#| eval: true
-load(joinpath(www_path,"models_train_before.png"))
+plt
 ```
@@
 T = 100
-using Serialization
 results = run_experiments(
@@
 ```{julia}
-using AlgorithmicRecourseDynamics.Models: model_evaluation
 plot_dict = Dict(key => Dict() for (key,val) in results)
@@
 #| label: fig-test-after
+#| output: true
+
 img_files = readdir(www_path)[contains.(readdir(www_path),"models_test_after")]
@@
 for img in img_files
-    display(load(img))
+    display(Images.load(img))
 end
@@
 ```{julia}
-#| eval: true
+#| output: true
 #| fig-cap: "Line Charts"
@@
 ```{julia}
-#| eval: true
+#| output: true
 #| fig-cap: "Error Bar Charts"
@@
 df = results[:overlapping].output
-df = df[[x ∈ [50] for x in df.n],:]
+df = df[[x ∈ maximum(df.n) for x in df.n],:]
 gdf = groupby(df, [:generator, :model, :n, :name, :scope])
@@
 temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+ggsave(temp_path, width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
 """

### 2022-12-19 docs/src/paper/proof_of_concept.qmd
verifying that results check out with updated package

@@
 
-# Environment:
-using Pkg; Pkg.activate("docs/src/paper")
-
-# Deps:
-using AlgorithmicRecourseDynamics
-using AlgorithmicRecourseDynamics.Models
-using CounterfactualExplanations
-using Flux
-using LaplaceRedux
-using Markdown
-using MLJBase
-using Plots
-using Random
-
-# Setup
-Random.seed!(2023)
-theme(:wong)
-include("docs/src/utils.jl")    # some helper functions
-output_path = output_dir("poc") # output directory for artifacts
-www_path = www_dir("poc")       # output directory for images
+include("docs/src/paper/setup.jl")
+eval(setup)
 ```
@@
 y′ = copy(ys)
-using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
 factuals = select_factual(counterfactual_data,chosen_individuals)

### 2022-12-19 docs/src/paper/setup.jl

@@ -1,0 +1,27 @@
+setup = quote
+    # Environment:
+    using Pkg; Pkg.activate("docs/src/paper")
+
+    # Deps:
+    using AlgorithmicRecourseDynamics
+    using AlgorithmicRecourseDynamics.Models
+    using AlgorithmicRecourseDynamics.Models: model_evaluation
+    using CounterfactualExplanations
+    using CounterfactualExplanations: counterfactual, counterfactual_label
+    using DataFrames
+    using Flux
+    using Images
+    using LaplaceRedux
+    using Markdown
+    using MLJBase
+    using Plots
+    using Random
+    using Serialization
+
+    # Setup
+    Random.seed!(2023)              # global seed to allow for reproducibility
+    theme(:wong)
+    include("docs/src/utils.jl")    # some helper functions
+    output_path = output_dir("poc") # output directory for artifacts
+    www_path = www_dir("poc")       # output directory for images
+end

### 2022-12-19 src/experiments/functions.jl

@@ -141,7 +141,7 @@ end
 
 using CounterfactualExplanations.DataPreprocessing: unpack
 using CounterfactualExplanations
-using CounterfactualExplanations.Counterfactuals: counterfactual, counterfactual_label
+using CounterfactualExplanations: counterfactual, counterfactual_label
 using ..Models
 """
 
@@ -170,7 +170,7 @@ function update!(experiment::Experiment, recourse_system::RecourseSystem, chosen
     indices_ = rand(1:experiment.num_counterfactuals,length(results)) # randomly draw from generated counterfactuals
 
     X′ = reduce(hcat,@.(selectdim(counterfactual(results),3,indices_)))
-    y′ = reduce(hcat,@.(selectdim(counterfactual_label(results),3,indices_)))
+    y′ = reduce(hcat,@.(selectdim(counterfactual_label(results),1,indices_)))
     
     X[:,chosen_individuals] = X′
     y[:,chosen_individuals] = y′

### 2022-12-20 docs/src/paper/experiments/_mitigation_strategies.qmd
helper function for bootstrap with caching and multi-threading

--- 2022-08-18 dev/notebooks/experiments/mitigation_strategies.qmd
@@
-## Mitigation Strategies
+### Mitigation Strategies
 
 ```{julia}
-#| eval: true
-using Pkg; Pkg.activate("dev")
-```
+#| echo: false
 
-```{julia}
-#| eval: true
-include("dev/utils.jl")
-using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra, Images
-theme(:wong)
+include("docs/src/paper/setup.jl")
+eval(setup)
 output_path = output_dir("mitigation_strategies")
-www_path = www_dir("mitigation_strategies");
+www_path = www_dir("mitigation_strategies")
 ```
@@
 
-### Synthetic
+#### Synthetic
 
@@
 n_folds = 5
-n_bootstrap = 1
 T = 100
@@
     experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T
 )
@@
 
-### Plots
+#### Plots
 
@@
 
-#### Line Charts
+##### Line Charts
 
@@
 
-#### Error Bar Charts
+##### Error Bar Charts
 
@@
 
-### Chart in paper
+#### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_synthetic.csv"))
+```
+
+#### Chart in paper
 
@@
 
-### Latent Space Search
+#### Latent Space Search
 
@@
 
-### Plots
+#### Plots
 
-#### Line Charts
+##### Line Charts
 
@@
 
-### Chart in paper
+#### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_latent.csv"))
+```
+
+#### Chart in paper
 
@@
 
-## Real World
+#### Bootstrap
 
 ```{julia}
-generators = Dict(
-    :Generic=>GenericGenerator(decision_threshold=0.5),
-    :Latent=>REVISEGenerator(),
-    :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
-    :Gravitational=>GravitationalGenerator(),
-    :ClapROAR=>ClapROARGenerator()
-)
+n_bootstrap = 1000
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_real_world.csv"))
 ```
 
-```{julia}
-max_obs = 2500
-data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
-```
-
-```{julia}
-using CounterfactualExplanations.DataPreprocessing: unpack
-bs = 50
-function data_loader(data::CounterfactualData)
-    X, y = unpack(data)
-    data = Flux.DataLoader((X,y),batchsize=bs)
-    return data
-end
-model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
-```
-
-```{julia}
-experiments = set_up_experiments(
-    data_sets,models,generators; 
-    pre_train_models=100, model_params=model_params, 
-    data_loader=data_loader
-)
-```
-
-```{julia}
-n_evals = 5
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-n_bootstrap = 1
-n_samples = 10000
-T = 250
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,
-    evaluate_every=evaluate_every,
-    n_rounds=n_rounds, 
-    n_folds=n_folds, 
-    n_bootstrap=n_bootstrap, 
-    T=T
-)
-Serialization.serialize(joinpath(output_path,"results_real_world.jls"),results)
-```
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_real_world.jls"))
-```
-
-```{julia}
-using Images
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    plt = plot(res)
-    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
-    line_charts[data_name] = plt
-    plt = plot(res,maximum(res.output.n))
-    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
-    errorbar_charts[data_name] = plt
-end
-```
-
-### Chart in paper
+#### Chart in paper
 

### 2022-12-20 docs/src/paper/experiments/_real_world.qmd
helper function for bootstrap with caching and multi-threading

--- 2022-08-17 dev/notebooks/experiments/real_world.qmd
@@
-## Real-World Data
+### Real-World Data
 
 ```{julia}
-#| eval: true
-using Pkg; Pkg.activate("dev")
-```
+#| echo: false
 
-```{julia}
-#| eval: true
-include("dev/utils.jl")
-using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
-theme(:wong)
+include("docs/src/paper/setup.jl")
+eval(setup)
 output_path = output_dir("real_world")
@@
 
-### Experiment
+#### Experiment
 
@@
 n_folds = 5
-n_bootstrap = 1
 n_samples = 10000
@@
     experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T, n_samples=n_samples,
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T, n_samples=n_samples,
     generative_model_params=generative_model_params
@@
 
-### Chart in paper
+#### Bootstrap
+
+```{julia}
+n_bootstrap = 1000
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap.csv"))
+```
+
+#### Chart in paper
 

### 2022-12-20 docs/src/paper/experiments/_synthetic.qmd
helper function for bootstrap with caching and multi-threading

@@
-## Synthetic Data {#sec-app-synthetic}
+### Synthetic Data {#sec-app-synthetic}
 
@@
 
-### Experiments
+#### Experiments
 
@@
 eval(setup)
+output_path = output_dir("synthetic") # output directory for artifacts
+www_path = www_dir("synthetic")       # output directory for images
 ```
@@
 
-### Plots
+#### Plots
 
 ```{julia}
-#| eval: true
-using Serialization
 results = Serialization.deserialize(joinpath(output_path,"results.jls"));
@@
 
-### Bootstrap
+#### Bootstrap
 
@@
 n_bootstrap = 1000
-using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
-using DataFrames
-df = DataFrame()
-for (key, val) in results
-    n_folds = length(val.experiment.recourse_systems)
-    for fold in 1:n_folds
-        for i in 1:length(val.experiment.system_identifiers)
-            rec_sys = val.experiment.recourse_systems[fold][i]
-            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
-            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
-            df_.data .= key
-            df_.model .= model_name
-            df_.generator .= gen_name
-            df_.fold .= fold
-            df = vcat(df, df_)
-        end
-    end
-end
-df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
-using RCall
-save_path = joinpath(output_path, "bootstrap.csv")
-using CSV
-CSV.write(save_path, df)
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap.csv"))
 ```
 
-### Chart in paper {#sec-app-synthetic-paper}
+#### Chart in paper {#sec-app-synthetic-paper}
 
@@
 
-using RCall
 scale_ = 1.5

### 2022-12-20 src/data/Data.jl

@@ -1,5 +1,8 @@
 module Data
 
+using AlgorithmicRecourseDynamics
+using CounterfactualExplanations
+
 include("utils.jl")
 include("functions.jl")
 export download_data_sets

### 2022-12-20 src/data/utils.jl

@@ -1,7 +1,7 @@
-# Train-test split
-using StatsBase
-using CounterfactualExplanations
+using CounterfactualExplanations.DataPreprocessing: CounterfactualData
 using Flux
+using StatsBase
+
 """
     train_test_split(data::CounterfactualData;test_size=0.2)
 

### 2022-12-20 src/models/Models.jl

@@ -1,5 +1,8 @@
 module Models
 
+using AlgorithmicRecourseDynamics
+using CounterfactualExplanations
+
 include("utils.jl")
 
 # Models:

### 2022-12-20 src/models/utils.jl

@@ -1,5 +1,5 @@
-using CounterfactualExplanations
-using CounterfactualExplanations.DataPreprocessing
+using CounterfactualExplanations: AbstractFittedModel
+using CounterfactualExplanations.DataPreprocessing: CounterfactualData
 using Flux
 using MLJBase
 using MLUtils
@@ -16,8 +16,8 @@ function data_loader(data::CounterfactualData)
     return data
 end
 
-function model_evaluation(M::CounterfactualExplanations.AbstractFittedModel, test_data::CounterfactualData)
-    X, y = DataPreprocessing.unpack(test_data)
+function model_evaluation(M::AbstractFittedModel, test_data::CounterfactualData)
+    X, y = CounterfactualExplanations.DataPreprocessing.unpack(test_data)
     m = MulticlassFScore()
     binary = M.likelihood == :classification_binary
     if binary

### 2022-12-20 src/post_processing.jl

@@ -1,16 +1,49 @@
-using Images
+using AlgorithmicRecourseDynamics: is_logging
+using CSV
 using DataFrames
-using RCall
+using Images
 using Plots
+using ProgressMeter: Progress, next!
+using RCall
 using Statistics
 
+function run_bootstrap(
+    results::Dict{Symbol, ExperimentResults}, n_bootstrap::Int=1000; 
+    filename::String="bootstrapped_results.csv", show_progress=!is_logging(stderr)
+)
+    df = DataFrame()
+    for (key, val) in results
+        n_folds = length(val.experiment.recourse_systems)
+        p_fold = Progress(n_folds; desc="Progress on folds:", showspeed=true, enabled=show_progress, output = stderr)
+        for fold in 1:n_folds
+            N = length(val.experiment.system_identifiers)
+            p_sys = Progress(N; desc="Progress on systems:", showspeed=true, enabled=show_progress, output = stderr)
+            Threads.@threads for i in 1:N
+                rec_sys = val.experiment.recourse_systems[fold][i]
+                model_name, gen_name = collect(val.experiment.system_identifiers)[i]
+                df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
+                df_.data .= key
+                df_.model .= model_name
+                df_.generator .= gen_name
+                df_.fold .= fold
+                df = vcat(df, df_)
+                next!(p_sys, showvalues = [(:Model, model_name), (:Generator, gen_name)])
+            end
+            next!(p_fold)
+        end
+    end
+    df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
+    CSV.write(filename, df)
+    return df
+end
+
 function Plots.plot(results::ExperimentResults, variable::Symbol=:mmd, scope::Symbol=:model; size=3, title=nothing, kwargs...)
-    
+
     df = results.output
     @assert variable in unique(df.name) "Not a valid variable."
 
     gdf = groupby(df, [:generator, :model, :n, :name, :scope])
-    df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
[…]

### 2022-12-20 docs/src/paper/data_preprocessing/_real_world_data.qmd
work on data

--- 2022-08-08 dev/notebooks/data_preprocessing.qmd/real_world_data.qmd
@@
 ---
-title: Preprocessing Data
-jupyter: julia-1.7
+title: Preprocessing Real-World Data
 ---
@@
 ```{julia}
-using Pkg; Pkg.activate("dev")
-```
+#| echo: false
 
-```{julia}
-include("dev/utils.jl")
-using AlgorithmicRecourseDynamics
-using CounterfactualExplanations, Flux, Plots, PlotThemes, Random, LaplaceRedux, LinearAlgebra
-theme(:wong)
+include("docs/src/paper/setup.jl")
+eval(setup)
 output_path = output_dir("real_world")
@@
 
-Fetching the data using Python's `sklearn`:
+Fetching the data using Python's `sklearn` (run this in the Python REPL):
 
-```{python}
+```python
 from sklearn.datasets import fetch_california_housing
@@
 df["target"] = y.values
-data_path = "../../artifacts/upload/data/real_world"
+data_path = "dev/artifacts/upload/data/real_world"
 import os
+if not os.path.isdir(os.path.join(data_path,"raw")):
+    os.makedirs(os.path.join(data_path,"raw"))
 df.to_csv(os.path.join(data_path,"raw/cal_housing.csv"), index=False)
@@
 ```{julia}
-using CSV, DataFrames, Statistics, StatsBase
 df = CSV.read(joinpath(data_path, "raw/cal_housing.csv"), DataFrame)
@@
 X = Matrix(df[:,Not(:target)])
-dt = fit(ZScoreTransform, X, dims=1)
+dt = StatsBase.fit(ZScoreTransform, X, dims=1)
 StatsBase.transform!(dt, X)

### 2022-12-20 docs/src/paper/setup.jl

@@ -10,16 +10,19 @@ setup = quote
     using AlgorithmicRecourseDynamics.Models: model_evaluation
     using CounterfactualExplanations
     using CounterfactualExplanations: counterfactual, counterfactual_label
+    using CSV
     using DataFrames
     using Flux
     using Images
     using LaplaceRedux
     using Markdown
     using MLJBase
+    using MLUtils
     using Plots
     using Random
     using RCall
     using Serialization
+    using StatsBase
 
     # Setup
     Random.seed!(2023)              # global seed to allow for reproducibility

### 2022-12-20 docs/src/paper/data_preprocessing/_real_world_data.qmd
rejigged data pre-processing

@@
 MLUtils.numobs(data::DataFrame) = nrow(data)
-df_balanced = getobs(undersample(df, df.target;shuffle=true))
+df_balanced = getobs(undersample(df, df.target;shuffle=true))[1]
 ```
@@
 ```{julia}
-CSV.write(joinpath(data_path, "cal_housing.csv"), df)
+CSV.write(joinpath(data_path, "cal_housing.csv"), df_balanced)
 ```
@@
 ```{julia}
-using CSV, DataFrames, Statistics, StatsBase
 df = CSV.read(joinpath(data_path, "raw/cs-training.csv"), DataFrame)
@@
 ```{julia}
-using CSV, DataFrames, Statistics, StatsBase
 df = CSV.read(joinpath(data_path, "raw/UCI_Credit_Card.csv"), DataFrame)
-select!(df, Not([:ID, :SEX, :EDUCATION, :MARRIAGE]))
+select!(df, Not([:ID]))
 rename!(df, "default.payment.next.month" => :target)
 dropmissing!(df)
+df.SEX = categorical(df.SEX)
+df.EDUCATION = categorical(df.EDUCATION)
+df.MARRIAGE = categorical(df.MARRIAGE)
 mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
-# Features:
-X = Matrix(df[:,Not(:target)])
-dt = fit(ZScoreTransform, X, dims=1)
-StatsBase.transform!(dt, X)
-# Target:
-y = df.target
-# Data:
-df = DataFrame(X,:auto)
-df.target = y
 ```
@@
 ```{julia}
-using MLUtils
-using MLUtils: undersample
 # Make DataFrames.jl work

### 2022-12-21 docs/src/paper/data_preprocessing/_real_world_data.qmd
working on models for real-world data

@@
 
-Loading the data into Julia session:
+Loading the data into Julia session.
 
@@
 df = CSV.read(joinpath(data_path, "raw/cal_housing.csv"), DataFrame)
-# Features:
-X = Matrix(df[:,Not(:target)])
-dt = StatsBase.fit(ZScoreTransform, X, dims=1)
-StatsBase.transform!(dt, X)
+# # Features:
+# X = Matrix(df[:,Not(:target)])
+# dt = StatsBase.fit(ZScoreTransform, X, dims=1)
+# StatsBase.transform!(dt, X)
+# df = DataFrame(X,:auto)
 # Target:
 y = df.target
-y = Float64.(y .>= median(y)); # binary target
+y = Float64.(y .>= median(y)); # binary target (positive outcome)
 # Data:
-df = DataFrame(X,:auto)
 df.target = y
@@
 
+Random undersampling to balance the data:
+
 ```{julia}
-using MLUtils: undersample
-# Make DataFrames.jl work
-MLUtils.getobs(data::DataFrame, i) = data[i,:]
-MLUtils.numobs(data::DataFrame) = nrow(data)
 df_balanced = getobs(undersample(df, df.target;shuffle=true))[1]
@@
 
+All features are continuous:
+
 ```{julia}
-CSV.write(joinpath(data_path, "cal_housing.csv"), df_balanced)
+schema(df_balanced)
+```
+
+Turning the data into `CounterfactualData`:
+
+```{julia}
+X = Matrix(df_balanced[:,Not(:target)])
+X = permutedims(X)
+y = permutedims(df_balanced.target)
+data = CounterfactualData(X,y)
+```
+
+Saving the data:
+
+```{julia}
+CSV.write(joinpath(data_path, "cal_housing.csv"), df_balanced)              # binary file
+Serialization.serialize(joinpath(data_path,"cal_housing.jls"), data)        # CounterfactualData
 ```
@@
 ## Give Me Some Credit
+
+Loading and basic preprocessing:
 
@@
 mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
-# Features:
-X = Matrix(df[:,Not(:target)])
-dt = StatsBase.fit(ZScoreTransform, X, dims=1)
-StatsBase.transform!(dt, X)
+# # Features:
+# X = Matrix(df[:,Not(:target)])
+# dt = StatsBase.fit(ZScoreTransform, X, dims=1)
+# StatsBase.transform!(dt, X)
+# df = DataFrame(X,:auto)
 # Target:
-y = df.target
-# Data:
-df = DataFrame(X,:auto)
-df.target = y
+df.target .= map(y -> y == 0 ? 1 : 0, df.target)        # postive outcome = no delinquency
 ```
 
+Balancing:
+
 ```{julia}
-using MLUtils
-using MLUtils: undersample
-# Make DataFrames.jl work
-MLUtils.getobs(data::DataFrame, i) = data[i,:]
-MLUtils.numobs(data::DataFrame) = nrow(data)
 df_balanced = getobs(undersample(df, df.target;shuffle=true))[1]
@@
 
+All features are continuous:
+
 ```{julia}
-CSV.write(joinpath(data_path, "gmsc.csv"), df_balanced)
+schema(df_balanced)
+```
+
+Turning the data into `CounterfactualData`:
+
+```{julia}
+X = Matrix(df_balanced[:,Not(:target)])
+X = permutedims(X)
+y = permutedims(df_balanced.target)
+data = CounterfactualData(X,y)
+```
+
+Saving:
+
+```{julia}
+CSV.write(joinpath(data_path, "gmsc.csv"), df_balanced)         # binary file
+Serialization.serialize(joinpath(data_path,"gmsc.jls"), data)   # CounterfactualData
 ```
@@
 ## UCI Credit Card Default
+
+Loading and basic preprocessing:
 
@@
 mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
+df.target .= map(y -> y == 0 ? 1 : 0, df.target)        # postive outcome = no default
 ```
 
+Balancing:
+
 ```{julia}
-# Make DataFrames.jl work
-MLUtils.getobs(data::DataFrame, i) = data[i,:]
-MLUtils.numobs(data::DataFrame) = nrow(data)
 df_balanced = getobs(undersample(df, df.target;shuffle=true))[1]
@@
 
+**Not** all features are continuous:
+
 ```{julia}
-CSV.write(joinpath(data_path, "credit_default.csv"), df_balanced)
+schema(df_balanced)
 ```
+
+One-hot encoding:
+
+```{julia}
+hot = OneHotEncoder()
+mach = MLJBase.fit!(machine(hot, df_balanced))
+df_balanced = MLJBase.transform(mach, df_balanced)
+schema(df_balanced)
+```
+
+Categorical indices:
+
+```{julia}
+features_categorical = [
+    [2,3],
+    collect(4:10),
+    collect(11:14)
+]
+```
+
+Preparing for use with `CounterfactualExplanations.jl`:
+
+```{julia}
+X = Matrix(df_balanced[:,Not(:target)])
+X = permutedims(X)
+y = permutedims(df_balanced.target)
+data = CounterfactualData(
+    X, y;
+    features_categorical = features_categorical
+)
+```
+
+Saving:
+
+```{julia}
+CSV.write(joinpath(data_path, "credit_default.csv"), df_balanced)         # binary file
+Serialization.serialize(joinpath(data_path,"credit_default.jls"), data)   # CounterfactualData
+```

### 2022-12-21 docs/src/paper/experiments/_real_world.qmd
working on models for real-world data

@@
 output_path = output_dir("real_world")
-www_path = www_dir("real_world");
+www_path = www_dir("real_world")
+data_path = data_dir("real_world")
 ```
@@
 max_obs = 2500
-data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs; data_dir=data_path)
 choices = [
@@
 using CounterfactualExplanations.DataPreprocessing: unpack
-bs = 50
+bs = 500
 function data_loader(data::CounterfactualData)
@@
 end
-model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
+model_params = (batch_norm=false,n_hidden=64,n_layers=3,dropout=true,p_dropout=0.5)
 ```
@@
 n_samples = 10000
-T = 250
+T = 100
 generative_model_params = (epochs=250, latent_dim=8)
-using Serialization
 results = run_experiments(
@@
 ```{julia}
-using Serialization
 results = Serialization.deserialize(joinpath(output_path,"results.jls"))

### 2022-12-21 src/base.jl

@@ -175,7 +175,10 @@ function set_up_experiment(
 
     # Pretrain:
     if !isnothing(pre_train_models)
-        map!(model -> Models.train(model, data_train; n_epochs=pre_train_models, kwargs...), values(models))
+        for (key, model) in models
+            @info "Training $key"
+            Models.train(model, data_train; n_epochs=pre_train_models, kwargs...)
+        end
     end
 
     experiment = Experiment(data_train, data_test, target, models, deepcopy(generators), num_counterfactuals)
@@ -218,7 +221,11 @@ function set_up_experiments(
         kwargs...
     )
 
-    experiments = Dict(key => set_up_single(data) for (key, data) in catalogue)
+    experiments = Dict{Symbol, Experiment}()
+    for (key, data) in catalogue
+        @info "Setting up $(key)"
+        experiments[key] = set_up_single(data)
+    end
 
     return experiments
 end

### 2022-12-21 src/data/functions.jl

@@ -2,20 +2,21 @@ using LazyArtifacts
 using CounterfactualExplanations.DataPreprocessing: CounterfactualData
 using CSV
 using DataFrames
+using Serialization
 using StatsBase
 
-function load_synthetic(max_obs::Union{Nothing, Int}=nothing)
-    data_dir = joinpath(artifact"data","data/synthetic")
+function load_synthetic(max_obs::Union{Nothing,Int}=nothing)
+    data_dir = joinpath(artifact"data", "data/synthetic")
     files = readdir(data_dir)
-    files = files[contains.(files,".csv")]
+    files = files[contains.(files, ".csv")]
     data = map(files) do file
         df = CSV.read(joinpath(data_dir, file), DataFrame)
-        X = convert(Matrix, hcat(df.x1,df.x2)')
+        X = convert(Matrix, hcat(df.x1, df.x2)')
         y = convert(Matrix, df.target')
-        data = CounterfactualData(X,y)
+        data = CounterfactualData(X, y)
         if !isnothing(max_obs)
             n_classes = length(unique(y))
-            data = undersample(data, Int(round(max_obs/n_classes)))
+            data = undersample(data, Int(round(max_obs / n_classes)))
         end
         (Symbol(replace(file, ".csv" => "")) => data)
     end
@@ -23,21 +24,19 @@ function load_synthetic(max_obs::Union{Nothing, Int}=nothing)
     return data
 end
 
-function load_real_world(max_obs::Union{Nothing, Int}=nothing)
-    data_dir = joinpath(artifact"data","data/real_world")
+function load_real_world(max_obs::Union{Nothing,Int}=nothing; data_dir::Union{Nothing, String}=nothing)
+    if isnothing(data_dir)
+        data_dir = joinpath(artifact"data", "data/real_world")
+    end
     files = readdir(data_dir)
-    files = files[contains.(files,".csv")]
+    files = files[contains.(files, ".jls")]
     data = map(files) do file
-        df = CSV.read(joinpath(data_dir, file), DataFrame)
-        X = Matrix(df[:,Not(:target)])
-        X = permutedims(X)
-        y = convert(Matrix, df.target')
-        data = CounterfactualData(X,y)
+        counterfactual_data = Serialization.deserialize(joinpath(data_dir, file))
         if !isnothing(max_obs)
-            n_classes = length(unique(y))
-            data = undersample(data, Int(round(max_obs/n_classes)))
[…]

### 2022-12-21 src/experiments/functions.jl

@@ -170,6 +170,7 @@ function update_experiment!(experiment::Experiment, recourse_system::RecourseSys
 
     X′ = reduce(hcat, @.(selectdim(counterfactual(results), 3, indices_)))
     y′ = reduce(hcat, @.(selectdim(counterfactual_label(results), 1, indices_)))
+    println(y′)
 
     X[:, chosen_individuals] = X′
     y[:, chosen_individuals] = y′

### 2022-12-21 docs/src/paper/data_preprocessing/_real_world_data.qmd
I just don't know why this isn't working for cat data

@@
 df = CSV.read(joinpath(data_path, "raw/cal_housing.csv"), DataFrame)
-# # Features:
-# X = Matrix(df[:,Not(:target)])
-# dt = StatsBase.fit(ZScoreTransform, X, dims=1)
-# StatsBase.transform!(dt, X)
-# df = DataFrame(X,:auto)
 # Target:
@@
 
+Feature transformation:
+
+```{julia}
+transformer = Standardizer(count=true)
+mach = MLJBase.fit!(machine(transformer, df_balanced[:,Not(:target)]))
+X = MLJBase.transform(mach, df_balanced[:,Not(:target)])
+schema(X)
+```
+
 Turning the data into `CounterfactualData`:
@@
 ```{julia}
-X = Matrix(df_balanced[:,Not(:target)])
+X = Matrix(X)
 X = permutedims(X)
@@
 mapcols!(x -> eltype(x) <: AbstractString ? parse.(Int, x) : x, df)
-# # Features:
-# X = Matrix(df[:,Not(:target)])
-# dt = StatsBase.fit(ZScoreTransform, X, dims=1)
-# StatsBase.transform!(dt, X)
-# df = DataFrame(X,:auto)
-# Target:
 df.target .= map(y -> y == 0 ? 1 : 0, df.target)        # postive outcome = no delinquency
@@
 
-One-hot encoding:
+Feature transformation:
 
 ```{julia}
-hot = OneHotEncoder()
-mach = MLJBase.fit!(machine(hot, df_balanced))
-df_balanced = MLJBase.transform(mach, df_balanced)
-schema(df_balanced)
+transformer = Standardizer(count=true) |> ContinuousEncoder()
+mach = MLJBase.fit!(machine(transformer, df_balanced[:,Not(:target)]))
+X = MLJBase.transform(mach, df_balanced[:,Not(:target)])
+schema(X)
 ```

### 2022-12-21 docs/src/paper/experiments/_real_world.qmd
I just don't know why this isn't working for cat data

@@
 end
-model_params = (batch_norm=false,n_hidden=64,n_layers=3,dropout=true,p_dropout=0.5)
+model_params = (batch_norm=false,n_hidden=64,n_layers=3,dropout=true,p_dropout=0.1)
 ```

### 2022-12-21 src/data/utils.jl

@@ -1,6 +1,28 @@
 using CounterfactualExplanations.DataPreprocessing: CounterfactualData
+using DataFrames
 using Flux
 using StatsBase
+
+function Base.hcat(data::CounterfactualData, more_data::CounterfactualData)
+
+    data = deepcopy(data)
+    more_data = deepcopy(more_data)
+
+    @assert all(data.features_categorical .== more_data.features_categorical) "Datasets have different categorical indices."
+    @assert all(data.features_continuous .== more_data.features_continuous) "Datasets have different continous indices."
+
+    data.X = hcat(data.X, more_data.X)
+    data.y = hcat(data.y, more_data.y)
+
+    return data
+end
+
+function DataFrames.subset(data::CounterfactualData, idx::Vector{Int})
+    dsub = deepcopy(data)
+    dsub.X = dsub.X[:,idx]
+    dsub.y = dsub.y[:,idx]
+    return dsub
+end
 
 """
     train_test_split(data::CounterfactualData;test_size=0.2)
@@ -8,14 +30,14 @@ using StatsBase
 Splits data into train and test split.
 """
 function train_test_split(data::CounterfactualData;test_size=0.2)
-    X,y = CounterfactualExplanations.DataPreprocessing.unpack(data)
+    X, y = CounterfactualExplanations.DataPreprocessing.unpack(data)
     N = size(y,2)
     classes_ = sort(unique(y))
     n_per_class = round(N/length(classes_))
     test_idx = sort(reduce(vcat,[sample(findall(vec(y.==cls)), Int(floor(test_size * n_per_class)),replace=false) for cls in classes_]))
     train_idx = setdiff(1:N, test_idx)
-    train_data = CounterfactualData(X[:,train_idx], y[:,train_idx])
-    test_data = CounterfactualData(X[:,test_idx], y[:,test_idx])
+    train_data = subset(data, train_idx)
+    test_data = subset(data, test_idx)
     return train_data, test_data
 end
 

### 2022-12-21 docs/src/paper/experiments/_real_world.qmd
uh

@@
 generators = Dict(
-    :Greedy=>GreedyGenerator(), 
-    :Generic=>GenericGenerator(),
+    # :Greedy=>GreedyGenerator(), 
+    # :Generic=>GenericGenerator(),
     :REVISE=>REVISEGenerator(),
-    :DICE=>DiCEGenerator(),
+    # :DICE=>DiCEGenerator(),
 )

### 2022-12-21 docs/src/paper/experiments/_real_world.qmd
next bug

--- 2022-12-21 docs/src/paper/experiments/_real_world.qmd
@@
     :cal_housing, 
-    :credit_default, 
+    # :credit_default, 
     :gmsc, 

### 2022-12-21 docs/src/paper/experiments/_mitigation_strategies.qmd
uh all running

@@
 ```{julia}
-n_bootstrap = 1000
+n_bootstrap = 10
 df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_synthetic.csv"))
@@
 ```{julia}
-n_bootstrap = 1000
+n_bootstrap = 10
 df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_latent.csv"))
@@
 
+## Real World
+
+```{julia}
+generators = Dict(
+    :Generic=>GenericGenerator(decision_threshold=0.5),
+    :Latent=>REVISEGenerator(),
+    :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
+    :Gravitational=>GravitationalGenerator(),
+    :ClapROAR=>ClapROARGenerator()
+)
+```
+
+```{julia}
+max_obs = 2500
+data_path = data_dir("real_world")
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs; data_dir=data_path)
+choices = [
+    :cal_housing, 
+    :credit_default, 
+    :gmsc, 
+]
+data_sets = filter(p -> p[1] in choices, data_sets)
+```
+
+```{julia}
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 500
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=false,n_hidden=64,n_layers=3,dropout=true,p_dropout=0.1)
+```
+
+```{julia}
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
+```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_samples = 10000
+T = 100
+generative_model_params = (epochs=250, latent_dim=8)
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T, n_samples=n_samples,
+    generative_model_params=generative_model_params
+)
+Serialization.serialize(joinpath(output_path,"results_real_world.jls"),results)
+```
+
+```{julia}
+using Serialization
+results = Serialization.deserialize(joinpath(output_path,"results_real_world.jls"))
+```
+
+```{julia}
+using Images
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
+end
+```
+
 #### Bootstrap
@@
 ```{julia}
-n_bootstrap = 1000
+n_bootstrap = 10
 df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_real_world.csv"))

### 2022-12-21 src/base.jl

@@ -66,7 +66,7 @@ function run!(
     # Pre-allocate memory:
     output = [DataFrame() for i in 1:M]
 
-    p_fold = Progress(K; desc="Total Progress:", showspeed=true, enabled=show_progress, output=stderr, color=:yellow)
+    p_fold = Progress(K; desc="Progress on folds:", showspeed=true, enabled=show_progress, output=stderr, color=:yellow)
     @info "Running experiment ..."
     for k in 1:K
         recourse_systems = experiment.recourse_systems[k]
@@ -96,7 +96,7 @@ function run!(
                     output[m] = vcat(output[m], output_checkpoint, cols=:union)
                 end
             end
-            next!(p_round, showvalues=[(:Fold, k//K), (:Round, n//N)])
+            next!(p_round, showvalues=[(:Fold, "$k/$K"), (:Round, "$n/$N")])
         end
         next!(p_fold)
     end

### 2022-12-21 src/post_processing.jl

@@ -29,9 +29,9 @@ function run_bootstrap(
                 df_.generator .= gen_name
                 df_.fold .= fold
                 df = vcat(df, df_)
-                next!(p_sys, showvalues = [(:Model, model_name), (:Generator, gen_name), (:System, i//N)])
+                next!(p_sys, showvalues = [(:Model, model_name), (:Generator, gen_name), (:System, "$i/$N")])
             end
-            next!(p_fold, showvalues = [(:Fold, fold//n_folds)])
+            next!(p_fold, showvalues = [(:Fold, "$fold/$n_folds")])
         end
         next!(p_total)
     end

### 2022-12-22 docs/src/paper/appendix copy.qmd
uff

--- 2022-12-21 docs/src/paper/experiments/_mitigation_strategies.qmd
@@
+---
+title: Supplementary Appendix
+---
+
+This is a supplementary appendix to the research paper **Endogenous Macrodynamics in Algorithmic Recourse**.
+It contains all of the experimental results, including those not highlighted in the actual paper.
+It also links to additional information about the proposed mitigation strategies.
+
+## Experimental Results {#sec-results}
+
+### Synthetic Data {#sec-app-synthetic}
+
+This notebook was used to run the experiments for the synthetic datasets and can be used to reproduce the results in the paper.
+In the following we first run the experiments and then generate visualizations and tables.
+
+#### Experiments
+
+```{julia}
+#| echo: false
+
+include("docs/src/paper/setup.jl")
+eval(setup)
+output_path = output_dir("synthetic") # output directory for artifacts
+www_path = www_dir("synthetic")       # output directory for images
+```
+
+```{julia}
+max_obs = 1000
+catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
+choices = [
+    :linearly_separable, 
+    :overlapping, 
+    :circles, 
+    :moons,
+]
+data_sets = filter(p -> p[1] in choices, catalogue)
+```
+
+```{julia}
+models = [
+    :LogisticRegression, 
+    :FluxModel, 
+    :FluxEnsemble,
+]
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator(),
+)
+```
+
+```{julia}
+experiments = set_up_experiments(data_sets,models,generators)
+```
+
+@fig-test-before shows the test data before running the experiment.
+
+```{julia}
+#| label: fig-test-before
+#| fig-cap: "Test data before experiment"
+#| eval: true
+
+plts = []
+for (exp_name, exp_) in experiments
+    for (M_name, M) in exp_.models
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$exp_name;\n $M_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plts = vcat(plts..., plt)
+    end
+end
+plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+savefig(plt, joinpath(www_path,"models_test_before.png"))
+plt
+```
+
+@fig-train-before shows the training data before running the experiment.
+
+```{julia}
+#| label: fig-train-before
+#| fig-cap: "Training data before experiment"
+#| eval: true
+
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plts = []
+for (exp_name, exp_) in experiments
+    for (M_name, M) in exp_.models
+        score = round(model_evaluation(M, exp_.train_data),digits=2)
+        plt = plot(M, exp_.train_data, title="$exp_name;\n $M_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.train_data.X)) .!= exp_.train_data.y))
+        x_wrongly_labelled = exp_.train_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plts = vcat(plts..., plt)
+    end
+end
+plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+savefig(plt, joinpath(www_path,"models_train_before.png"))
+plt
+```
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+T = 100
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T
+)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+```{julia}
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        score = round(model_evaluation(M, exp_.test_data),digits=2)
+        plt = plot(M, exp_.test_data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, exp_.test_data.X)) .!= exp_.test_data.y))
+        x_wrongly_labelled = exp_.test_data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_test_after_$(name).png"))
+end
+```
+
+@fig-test-after shows the test data after running the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Test data after experiment"
+#| fig-subcap: 
+#|      - "DICE"
+#|      - "Generic"
+#|      - "Greedy"
+#|      - "Latent"
+#| layout-ncol: 1
+#| label: fig-test-after
+#| output: true
+
+img_files = readdir(www_path)[contains.(readdir(www_path),"models_test_after")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(Images.load(img))
+end
+```
+
+```{julia}
+using AlgorithmicRecourseDynamics.Models: model_evaluation
+plot_dict = Dict(key => Dict() for (key,val) in results)
+fold = 1
+for (name, res) in results
+    exp_ = res.experiment
+    plot_dict[name] = Dict(key => [] for (key,val) in exp_.generators)
+    rec_sys = exp_.recourse_systems[fold]
+    sys_ids = collect(exp_.system_identifiers)
+    M = length(rec_sys)
+    for m in 1:M
+        model_name, generator_name = sys_ids[m]
+        M = rec_sys[m].model
+        data = rec_sys[m].data
+        score = round(model_evaluation(M, data),digits=2)
+        plt = plot(M, data, title="$name;\n $model_name ($score)")
+        # Errors:
+        ids = findall(vec(round.(probs(M, data.X)) .!= data.y))
+        x_wrongly_labelled = data.X[:,ids]
+        scatter!(plt, x_wrongly_labelled[1,:], x_wrongly_labelled[2,:], ms=7.5, color=:red, label="")
+        plot_dict[name][generator_name] = vcat(plot_dict[name][generator_name], plt)
+    end
+end
+plot_dict = Dict(key => reduce(vcat, [plots[key] for plots in values(plot_dict)]) for (key, value) in generators)
+for (name, plts) in plot_dict
+    plt = plot(plts..., layout=(length(choices),length(models)),size=(length(choices)*300,length(models)*300))
+    savefig(plt, joinpath(www_path,"models_train_after_$(name).png"))
+end
+```
+
+@fig-train-after shows the training data after running the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Training data after experiment"
+#| fig-subcap: 
+#|      - "DICE"
+#|      - "Generic"
+#|      - "Greedy"
+#|      - "Latent"
+#| layout-ncol: 1
+#| label: fig-train-after
+img_files = readdir(www_path)[contains.(readdir(www_path),"models_train_after")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(Images.load(img))
+end
+```
+
+#### Plots
+
+```{julia}
+results = Serialization.deserialize(joinpath(output_path,"results.jls"));
+```
+
+```{julia}
+using Images
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
+end
+```
+
+##### Line Charts
+
+@fig-line shows the evolution of the evaluation metrics over the course of the experiment.
+
+```{julia}
+#| output: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-line
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(Images.load(img))
+end
+```
+
+##### Error Bar Charts
+
+@fig-error shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| output: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "Circles"
+#|      - "Linearly Separable"
+#|      - "Moons"
+#|      - "Overlapping"
+#| layout-ncol: 1
+#| label: fig-error
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(Images.load(img))
+end
+```
+
+#### Bootstrap
+
+```{julia}
+n_bootstrap = 10
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap.csv"))
+```
+
+#### Chart in paper {#sec-app-synthetic-paper}
+
+@fig-paper shows the chart that went into the paper.
+
+```{julia}
+using DataFrames, Statistics
+df = results[:overlapping].output
+df = df[[x ∈ maximum(df.n) for x in df.n],:]
+gdf = groupby(df, [:generator, :model, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:decisiveness, :disagreement, :mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
+df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+transform!(df_plot, :name => (X -> [x=="decisiveness" ? "Decisiveness" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="disagreement" ? "Disagreement" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
+transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
+transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
+
+ncol = length(unique(df_plot.model))
+nrow = length(unique(df_plot.name))
+
+scale_ = 1.5
+R"""
+library(data.table)
+df_plot <- data.table($df_plot)
+name_order <- c(
+    "MMD (domain)",
+    "MMD (model)",
+    "Performance",
+    "Disagreement",
+    "Decisiveness"
+)
+df_plot[,name:=factor(name, levels=name_order)]
+model_order <- c("Linear", "MLP", "Deep Ensemble")
+df_plot[,model:=factor(model, levels=model_order)]
+library(ggplot2)
+plt <- ggplot(df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange(aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=c(0.9,0.9)), size=0.5) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(model), 
+        scales = "free_y"
+    ) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path, width=$ncol * $scale_,height=$nrow * $scale_ * 0.75) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_synthetic_results.png"), img)
+```
+
+```{julia}
+#| label: fig-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_synthetic_results.png"))
+```
+
+### Real-World Data
+
+```{julia}
+#| echo: false
+
+include("docs/src/paper/setup.jl")
+eval(setup)
+output_path = output_dir("real_world")
+www_path = www_dir("real_world")
+data_path = data_dir("real_world")
+```
+
+```{julia}
+max_obs = 2500
+data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs; data_dir=data_path)
+choices = [
+    :cal_housing, 
+    :credit_default, 
+    :gmsc, 
+]
+data_sets = filter(p -> p[1] in choices, data_sets)
+```
+
+```{julia}
+using CounterfactualExplanations.DataPreprocessing: unpack
+bs = 500
+function data_loader(data::CounterfactualData)
+    X, y = unpack(data)
+    data = Flux.DataLoader((X,y),batchsize=bs)
+    return data
+end
+model_params = (batch_norm=false,n_hidden=64,n_layers=3,dropout=true,p_dropout=0.1)
+```
+
+```{julia}
+models = [
+    :LogisticRegression, 
+    :FluxModel, 
+    :FluxEnsemble
+]
+generators = Dict(
+    :Greedy=>GreedyGenerator(), 
+    :Generic=>GenericGenerator(),
+    :REVISE=>REVISEGenerator(),
+    :DICE=>DiCEGenerator(),
+)
+```
+
+```{julia}
+experiments = set_up_experiments(
+    data_sets,models,generators; 
+    pre_train_models=100, model_params=model_params, 
+    data_loader=data_loader
+)
+```
+
+#### Experiment
+
+```{julia}
+n_evals = 5
+n_rounds = 50
+evaluate_every = Int(round(n_rounds/n_evals))
+n_folds = 5
+n_samples = 10000
+T = 100
+generative_model_params = (epochs=250, latent_dim=8)
+results = run_experiments(
+    experiments;
+    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T, n_samples=n_samples,
+    generative_model_params=generative_model_params
+)
+Serialization.serialize(joinpath(output_path,"results.jls"),results)
+```
+
+#### Plots
+
+```{julia}
+results = Serialization.deserialize(joinpath(output_path,"results.jls"))
+```
+
+```{julia}
+using Images
+line_charts = Dict()
+errorbar_charts = Dict()
+for (data_name, res) in results
+    plt = plot(res)
+    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
+    line_charts[data_name] = plt
+    plt = plot(res,maximum(res.output.n))
+    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
+    errorbar_charts[data_name] = plt
+end
+```
+
+##### Line Charts
+
+@fig-real-line shows the evolution of the evaluation metrics over the course of the experiment.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Line Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Credit Default"
+#|      - "GMSC"
+#| layout-ncol: 1
+#| label: fig-real-line
+img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+##### Error Bar Charts
+
+@fig-real-error shows the evaluation metrics at the end of the experiments.
+
+```{julia}
+#| eval: true
+#| fig-cap: "Error Bar Charts"
+#| fig-subcap: 
+#|      - "California Housing"
+#|      - "Credit Default"
+#|      - "GMSC"
+#| layout-ncol: 1
+#| label: fig-real-error
+img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart")]
+img_files = joinpath.(www_path,img_files)
+for img in img_files
+    display(load(img))
+end
+```
+
+#### Bootstrap
+
+```{julia}
+n_bootstrap = 10
+df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap.csv"))
+```
+
+#### Chart in paper
+
+@fig-real-paper shows the chart that went into the paper.
+
+```{julia}
+using DataFrames, Statistics
+model_ = :FluxEnsemble
+df = DataFrame() 
+for (key, val) in results
+    df_ = deepcopy(val.output)
+    df_.dataset .= key
+    df = vcat(df,df_)
+end
+df = df[df.n .== maximum(df.n),:]
+df = df[df.model .== model_,:]
+filter!(:value => x -> !any(f -> f(x), (ismissing, isnothing, isnan)), df)
+gdf = groupby(df, [:generator, :dataset, :n, :name, :scope])
+df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
+df_plot = df_plot[[name in [:mmd, :model_performance] for name in df_plot.name],:]
+df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
+df_plot.name .= [r[:name] == "mmd" ? "$(r[:name])_$(r[:scope])" : r[:name] for r in eachrow(df_plot)]
+transform!(df_plot, :dataset => (X -> [x=="cal_housing" ? "California Housing" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="credit_default" ? "Credit Default" : x for x in X]) => :dataset)
+transform!(df_plot, :dataset => (X -> [x=="gmsc" ? "GMSC" : x for x in X]) => :dataset)
+transform!(df_plot, :name => (X -> [x=="mmd_domain" ? "MMD (domain)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="mmd_model" ? "MMD (model)" : x for x in X]) => :name)
+transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
+transform!(df_plot, :generator => (X -> [x=="REVISE" ? "Latent" : x for x in X]) => :generator)
+
+ncol = length(unique(df_plot.dataset))
+nrow = length(unique(df_plot.name))
+
+using RCall
+scale_ = 1.75
+R"""
+library(ggplot2)
+plt <- ggplot($df_plot) +
+    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
+    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
+    facet_grid(
+        rows = vars(name),
+        cols =  vars(dataset), 
+        scales = "free_y"
+    ) +
+    labs(y = "Value") + 
+    scale_fill_discrete(name="Generator:") +
+    scale_colour_discrete(name="Generator:") +
+    theme(
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        legend.position="bottom"
+    )
+temp_path <- file.path(tempdir(), "plot.png")
+ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
+"""
+
+img = Images.load(rcopy(R"temp_path"))
+Images.save(joinpath(www_path,"paper_real_world_results.png"), img)
+```
+
+```{julia}
+#| label: fig-real-paper
+#| fig-cap: "Chart in paper"
+#| eval: true
+Images.load(joinpath(www_path,"paper_real_world_results.png"))
+```
+
 ### Mitigation Strategies
@@
 ```
+
+## Generators {#sec-generators}
+
+The generators have been moved into [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl).
+For more information so the package [documentation](https://pat-alt.github.io/CounterfactualExplanations.jl/dev).

### 2022-12-22 docs/src/paper/experiments/_index.qmd
uff

--- 2022-12-20 docs/src/paper/experiments/_mitigation_strategies.qmd
@@
-### Mitigation Strategies
+---
+title: Supplementary Appendix
+---
 
-```{julia}
-#| echo: false
+This is a supplementary appendix to the research paper **Endogenous Macrodynamics in Algorithmic Recourse**.
+It contains all of the experimental results, including those not highlighted in the actual paper.
+It also links to additional information about the proposed mitigation strategies.
 
-include("docs/src/paper/setup.jl")
-eval(setup)
-output_path = output_dir("mitigation_strategies")
-www_path = www_dir("mitigation_strategies")
-```
+## Experimental Results {#sec-results}
 
-```{julia}
-models = [
-    :LogisticRegression, 
-    :FluxModel, 
-    :FluxEnsemble,
-]
-generators = Dict(
-    :Generic=>GenericGenerator(decision_threshold=0.5),
-    :Latent=>REVISEGenerator(),
-    :Generic_conservative=>GenericGenerator(decision_threshold=0.9),
-    :Gravitational=>GravitationalGenerator(),
-    :ClapROAR=>ClapROARGenerator()
-)
-```
+{{< include _synthetic.qmd >}}
 
-#### Synthetic
+{{< include _real_world.qmd >}}
 
-```{julia}
-max_obs = 1000
-catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic(max_obs)
-choices = [
-    :linearly_separable, 
-    :overlapping, 
-    :circles, 
-    :moons,
-]
-data_sets = filter(p -> p[1] in choices, catalogue)
-```
+{{< include _mitigation_strategies.qmd >}}
 
-```{julia}
-experiments = set_up_experiments(data_sets,models,generators)
-```
+## Generators {#sec-generators}
 
-```{julia}
-n_evals = 5
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-T = 100
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T
-)
-Serialization.serialize(joinpath(output_path,"results_synthetic.jls"),results)
-```
-
-#### Plots
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_synthetic.jls"))
-```
-
-```{julia}
-using Images
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    plt = plot(res)
-    Images.save(joinpath(www_path, "line_chart_$(data_name).png"), plt)
-    line_charts[data_name] = plt
-    plt = plot(res,maximum(res.output.n))
-    Images.save(joinpath(www_path, "errorbar_chart_$(data_name).png"), plt)
-    errorbar_charts[data_name] = plt
-end
-```
-
-##### Line Charts
-
-@fig-mit-line shows the evolution of the evaluation metrics over the course of the experiment.
-
-```{julia}
-#| eval: true
-#| fig-cap: "Line Charts"
-#| fig-subcap: 
-#|      - "California Housing"
-#|      - "Circles"
-#|      - "Credit Default"
-#|      - "GMSC"
-#|      - "Linearly Separable"
-#|      - "Moons"
-#|      - "Overlapping"
-#| layout-ncol: 1
-#| label: fig-mit-line
-img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart") .&& .!contains.(readdir(www_path),"latent")]
-img_files = joinpath.(www_path,img_files)
-for img in img_files
-    display(load(img))
-end
-```
-
-##### Error Bar Charts
-
-@fig-mit-error shows the evaluation metrics at the end of the experiments.
-
-```{julia}
-#| eval: true
-#| fig-cap: "Error Bar Charts"
-#| fig-subcap: 
-#|      - "California Housing"
-#|      - "Circles"
-#|      - "Credit Default"
-#|      - "GMSC"
-#|      - "Linearly Separable"
-#|      - "Moons"
-#|      - "Overlapping"
-#| layout-ncol: 1
-#| label: fig-mit-error
-img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart") .&& .!contains.(readdir(www_path),"latent")]
-img_files = joinpath.(www_path,img_files)
-for img in img_files
-    display(load(img))
-end
-```
-
-#### Bootstrap
-
-```{julia}
-n_bootstrap = 1000
-df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_synthetic.csv"))
-```
-
-#### Chart in paper
-
-@fig-mit-paper shows the chart that went into the paper.
-
-```{julia}
-using DataFrames, Statistics
-df = results[:overlapping].output
-df = df[df.n .== maximum(df.n),:]
-gdf = groupby(df, [:generator, :model, :n, :name, :scope])
-df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
-df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
-df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
-df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
-transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
-transform!(df_plot, :generator => (X -> [x=="Generic" ? "Generic (γ=0.5)" : x for x in X]) => :generator)
-transform!(df_plot, :generator => (X -> [x=="Generic_conservative" ? "Generic (γ=0.9)" : x for x in X]) => :generator)
-transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
-
-ncol = length(unique(df_plot.model))
-nrow = length(unique(df_plot.name))
-
-using RCall
-scale_ = 2.0
-R"""
-library(data.table)
-df_plot <- data.table($df_plot)
-model_order <- c("Linear", "MLP", "Deep Ensemble")
-df_plot[,model:=factor(model, levels=model_order)]
-library(ggplot2)
-plt <- ggplot($df_plot) +
-    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
-    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
-    facet_grid(
-        rows = vars(name),
-        cols =  vars(model), 
-        scales = "free_y"
-    ) +
-    labs(y = "Value") + 
-    scale_fill_discrete(name="Generator:") +
-    scale_colour_discrete(name="Generator:") +
-    theme(
-        axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(),
-        legend.position="bottom"
-    ) +
-    guides(fill=guide_legend(ncol=3))
-temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
-"""
-
-img = Images.load(rcopy(R"temp_path"))
-Images.save(joinpath(www_path,"paper_synthetic_results.png"), img)
-```
-
-```{julia}
-#| label: fig-mit-paper
-#| fig-cap: "Chart in paper"
-#| eval: true
-Images.load(joinpath(www_path,"paper_synthetic_results.png"))
-```
-
-#### Latent Space Search
-
-```{julia}
-generators = Dict(
-    :Latent=>GenericGenerator(decision_threshold=0.5),
-    :Latent_conservative=>GenericGenerator(decision_threshold=0.9),
-    :Gravitational=>GravitationalGenerator(),
-    :ClapROAR=>ClapROARGenerator()
-)
-```
-
-```{julia}
-experiments = set_up_experiments(data_sets,models,generators)
-```
-
-```{julia}
-n_evals = 5
-n_rounds = 50
-evaluate_every = Int(round(n_rounds/n_evals))
-n_folds = 5
-T = 100
-using Serialization
-results = run_experiments(
-    experiments;
-    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, T=T
-)
-Serialization.serialize(joinpath(output_path,"results_synthetic_latent.jls"),results)
-```
-
-```{julia}
-using Serialization
-results = Serialization.deserialize(joinpath(output_path,"results_synthetic_latent.jls"))
-```
-
-```{julia}
-using Images
-line_charts = Dict()
-errorbar_charts = Dict()
-for (data_name, res) in results
-    plt = plot(res)
-    Images.save(joinpath(www_path, "line_chart_latent_$(data_name).png"), plt)
-    line_charts[data_name] = plt
-    plt = plot(res,maximum(res.output.n))
-    Images.save(joinpath(www_path, "errorbar_chart_latent_$(data_name).png"), plt)
-    errorbar_charts[data_name] = plt
-end
-```
-
-#### Plots
-
-##### Line Charts
-
-@fig-mit-line-latent shows the evolution of the evaluation metrics over the course of the experiment.
-
-```{julia}
-#| eval: true
-#| fig-cap: "Line Charts"
-#| fig-subcap: 
-#|      - "Circles"
-#|      - "Linearly Separable"
-#|      - "Moons"
-#|      - "Overlapping"
-#| layout-ncol: 1
-#| label: fig-mit-line-latent
-img_files = readdir(www_path)[contains.(readdir(www_path),"line_chart") .&& contains.(readdir(www_path),"latent")]
-img_files = joinpath.(www_path,img_files)
-for img in img_files
-    display(load(img))
-end
-```
-
-##### Error Bar Charts
-
-@fig-mit-error-latent shows the evaluation metrics at the end of the experiments.
-
-```{julia}
-#| eval: true
-#| fig-cap: "Error Bar Charts"
-#| fig-subcap: 
-#|      - "Circles"
-#|      - "Linearly Separable"
-#|      - "Moons"
-#|      - "Overlapping"
-#| layout-ncol: 1
-#| label: fig-mit-error-latent
-img_files = readdir(www_path)[contains.(readdir(www_path),"errorbar_chart") .&& contains.(readdir(www_path),"latent")]
-img_files = joinpath.(www_path,img_files)
-for img in img_files
-    display(load(img))
-end
-```
-
-#### Bootstrap
-
-```{julia}
-n_bootstrap = 1000
-df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_latent.csv"))
-```
-
-#### Chart in paper
-
-@fig-mit-latent-paper shows the chart that went into the paper.
-
-```{julia}
-using DataFrames, Statistics
-df = results[:overlapping].output
-df = df[df.n .== maximum(df.n),:]
-gdf = groupby(df, [:generator, :model, :n, :name, :scope])
-df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
-df_plot = df_plot[[name in [:mmd, :mmd_grid, :model_performance] for name in df_plot.name],:]
-df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.==:model),:]
-df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
-transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (domain)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="mmd_grid" ? "MMD (model)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
-transform!(df_plot, :generator => (X -> [x=="Latent" ? "Latent (γ=0.5)" : x for x in X]) => :generator)
-transform!(df_plot, :generator => (X -> [x=="Latent_conservative" ? "Latent (γ=0.9)" : x for x in X]) => :generator)
-transform!(df_plot, :model => (X -> [x=="FluxEnsemble" ? "Deep Ensemble" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="FluxModel" ? "MLP" : x for x in X]) => :model)
-transform!(df_plot, :model => (X -> [x=="LogisticRegression" ? "Linear" : x for x in X]) => :model)
-
-ncol = length(unique(df_plot.model))
-nrow = length(unique(df_plot.name))
-
-using RCall
-scale_ = 1.9
-R"""
-library(data.table)
-df_plot <- data.table($df_plot)
-model_order <- c("Linear", "MLP", "Deep Ensemble")
-df_plot[,model:=factor(model, levels=model_order)]
-library(ggplot2)
-plt <- ggplot($df_plot) +
-    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
-    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
-    facet_grid(
-        rows = vars(name),
-        cols =  vars(model), 
-        scales = "free_y"
-    ) +
-    labs(y = "Value") + 
-    scale_fill_discrete(name="Generator:") +
-    scale_colour_discrete(name="Generator:") +
-    theme(
-        axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(),
-        legend.position="bottom"
-    ) +
-    guides(fill=guide_legend(ncol=4))
-temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.8) 
-"""
-
-img = Images.load(rcopy(R"temp_path"))
-Images.save(joinpath(www_path,"paper_synthetic_latent_results.png"), img)
-```
-
-```{julia}
-#| label: fig-mit-latent-paper
-#| fig-cap: "Chart in paper"
-#| eval: true
-Images.load(joinpath(www_path,"paper_synthetic_latent_results.png"))
-```
-
-#### Bootstrap
-
-```{julia}
-n_bootstrap = 1000
-df = run_bootstrap(results, n_bootstrap; filename=joinpath(output_path,"bootstrap_real_world.csv"))
-```
-
-#### Chart in paper
-
-@fig-mit-latent-paper shows the chart that went into the paper.
-
-```{julia}
-using DataFrames, Statistics
-model_ = :FluxEnsemble
-df = DataFrame() 
-for (key, val) in results
-    df_ = deepcopy(val.output)
-    df_.dataset .= key
-    df = vcat(df,df_)
-end
-df = df[df.n .== maximum(df.n),:]
-df = df[df.model .== model_,:]
-filter!(:value => x -> !any(f -> f(x), (ismissing, isnothing, isnan)), df)
-gdf = groupby(df, [:generator, :dataset, :n, :name, :scope])
-df_plot = combine(gdf, :value => (x -> [(mean(x),mean(x)+std(x),mean(x)-std(x))]) => [:mean, :ymax, :ymin])
-df_plot = df_plot[[name in [:mmd, :model_performance] for name in df_plot.name],:]
-df_plot = df_plot[.!(df_plot.name.==:mmd .&& df_plot.scope.!=:model),:]
-df_plot = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df_plot)
-transform!(df_plot, :dataset => (X -> [x=="cal_housing" ? "California Housing" : x for x in X]) => :dataset)
-transform!(df_plot, :dataset => (X -> [x=="credit_default" ? "Credit Default" : x for x in X]) => :dataset)
-transform!(df_plot, :dataset => (X -> [x=="gmsc" ? "GMSC" : x for x in X]) => :dataset)
-transform!(df_plot, :name => (X -> [x=="mmd" ? "MMD (model)" : x for x in X]) => :name)
-transform!(df_plot, :name => (X -> [x=="model_performance" ? "Performance" : x for x in X]) => :name)
-transform!(df_plot, :generator => (X -> [x=="Generic" ? "Generic (γ=0.5)" : x for x in X]) => :generator)
-transform!(df_plot, :generator => (X -> [x=="Generic_conservative" ? "Generic (γ=0.9)" : x for x in X]) => :generator)
-
-ncol = length(unique(df_plot.dataset))
-nrow = length(unique(df_plot.name))
-
-using RCall
-scale_ = 2.0
-R"""
-library(ggplot2)
-plt <- ggplot($df_plot) +
-    geom_bar(aes(x=n, y=mean, fill=generator), stat="identity", alpha=0.5, position="dodge") +
-    geom_pointrange( aes(x=n, y=mean, ymin=ymin, ymax=ymax, colour=generator), alpha=0.9, position=position_dodge(width=0.9), size=0.5) +
-    facet_grid(
-        rows = vars(name),
-        cols =  vars(dataset), 
-        scales = "free_y"
-    ) +
-    labs(y = "Value") + 
-    scale_fill_discrete(name="Generator:") +
-    scale_colour_discrete(name="Generator:") +
-    theme(
-        axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(),
-        legend.position="bottom"
-    ) +
-    guides(fill=guide_legend(ncol=3))
-temp_path <- file.path(tempdir(), "plot.png")
-ggsave(temp_path,width=$ncol * $scale_,height=$nrow * $scale_ * 0.85) 
-"""
-
-img = Images.load(rcopy(R"temp_path"))
-Images.save(joinpath(www_path,"paper_real_world_results.png"), img)
-```
-
-```{julia}
-#| label: fig-mit-real-paper
-#| fig-cap: "Chart in paper"
-#| eval: true
-Images.load(joinpath(www_path,"paper_real_world_results.png"))
-```
+The generators have been moved into [`CounterfactualExplanations.jl`](https://github.com/pat-alt/CounterfactualExplanations.jl).
+For more information so the package [documentation](https://pat-alt.github.io/CounterfactualExplanations.jl/dev).

### 2022-12-22 docs/src/_intro.qmd
helper function for bootstrap with caching and multi-threading
moved stuff out of here

@@
 
-## Research Paper 📝
-
-**Note** ⚠: You are browsing the [`#original-paper`](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/tree/original-paper) branch of `AlgorithmicRecourseDynamics.jl`.
-This branch is a static artifact corresponding to the state of the package at the time the paper was first published.
-It can be used to replicate the original findings of the paper.
-
-## At a Glance
-
-You can find resources relevant to the paper right here on this branch of the
-repository:
-
-- [Paper](paper/paper.pdf)
-- [Notebooks](dev/notebooks/)
-- [Supplementary Appendix](build/dev/notebooks/appendix.html) generated from notebooks (download the HTML and view in browser)
-- [Artifacts]() (including data and experimental results; link currently exluded due to double-blind review process)
+## Related Research Paper 📝
 
@@
 ![](paper/www/poc.png)
-
-## Paper Abstract
-
-Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely focused on single individuals in a static environment: given some estimated model, the goal is to find valid counterfactuals for an individual instance that fulfill various desiderata.
-The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge.
-There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals.
-Through this work we aim to close that gap.
-We first show that many of the existing methodologies can be collectively described by a generalized framework.
-We then argue that the existing framework does not account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level.
-Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate large numbers of counterfactuals and study the resulting domain and model shifts.
-We find that the induced shifts are substantial enough to likely impede the applicability of Algorithmic Recourse in some situations.
-Fortunately, we find various strategies to mitigate these concerns.
-Our simulation framework for studying recourse dynamics is fast and open-sourced.

### 2022-12-23 README.qmd
moved stuff out of here
cleaned up a bit

@@
 
-[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/stable)
-[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/CounterfactualExplanations.jl/dev)
-[![Build Status](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/CounterfactualExplanations.jl/actions/workflows/CI.yml?query=branch%3Amain)
-[![Coverage](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/CounterfactualExplanations.jl)
+[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl/stable) 
+[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl/dev) 
+[![Build Status](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml?query=branch%3Amain)
+[![Coverage](https://codecov.io/gh/pat-alt/AlgorithmicRecourseDynamics.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/AlgorithmicRecourseDynamics.jl) 
 [![Code Style: Blue](https://img.shields.io/badge/code%20style-blue-4495d1.svg)](https://github.com/invenia/BlueStyle) 
-[![ColPrac: Contributor's Guide on Collaborative Practices for Community Packages](https://img.shields.io/badge/ColPrac-Contributor's%20Guide-blueviolet)](https://github.com/SciML/ColPrac)
+[![ColPrac: Contributor’s Guide on Collaborative Practices for Community Packages](https://img.shields.io/badge/ColPrac-Contributor's%20Guide-blueviolet.png)](https://github.com/SciML/ColPrac) 
 [![Twitter Badge](https://img.shields.io/twitter/url/https/twitter.com/paltmey.svg?style=social&label=Follow%20%40paltmey)](https://twitter.com/paltmey)
@@
 # AlgorithmicRecourseDynamics
+
+{{< include docs/src/_intro.qmd >}}

### 2022-12-23 docs/src/_intro.qmd
cleaned up a bit

@@
-`AlgorithmicRecourseDynamics.jl` is a Julia package for modeling Algorithmic Recourse Dynamics.
+```{julia}
+#| echo: false
+include("docs/setup_docs.jl")
+eval(setup_docs)
+```
+
+`AlgorithmicRecourseDynamics.jl` is a small package for modeling Algorithmic Recourse Dynamics. It builds on `CounterfactualExplanations`, a package for generating counterfactual explanations.
+
+## Basic Usage
+
+### Data and Model
+
+```{julia}
+N = 1000
+xmax = 2
+X, ys = make_blobs(
+    N, 2; 
+    centers=2, as_table=false, center_box=(-xmax => xmax), cluster_std=0.1
+)
+ys .= ys.==2
+X = X'
+counterfactual_data = CounterfactualData(X,ys')
+```
+
+```{julia}
+n_epochs = 100
+model = Chain(Dense(2,1))
+mod = FluxModel(model)
+
+generator = GenericGenerator()
+```
+
+```{julia}
+#| output: true
+
+data_train, data_test = Data.train_test_split(counterfactual_data)
+Models.train(mod, data_train; n_epochs=n_epochs)
+plt_original = plot(mod, counterfactual_data; zoom=0, colorbar=false)
+display(plt_original)
+```
+
+### Simulation
+
+```{julia}
+models = Dict(:mymodel => mod)
+generators = Dict(:wachter => generator)
+experiment = set_up_experiment(data_train, data_test, models, generators)
+```
+
+```{julia}
+#| output: true
+
+run!(experiment)
+```
+
+```{julia}
+#| output: true
+
+new_data = experiment.recourse_systems[1][1].data
+new_model = experiment.recourse_systems[1][1].model
+plt_original = plot(new_model, new_data; zoom=0, colorbar=false)
+```
 
@@
 
-In this work we investigate what happens if Algorithmic Recourse is actually implemented by a large number of individuals.
-The chart below illustrates what we mean by Endogenous Macrodynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class (y=0) are marked in blue and samples of the positive class (y=1) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.
-
-![](paper/www/poc.png)
+The package was developed for a research project that investigates the dynamics of various counterfactual generators.

### 2022-12-23 docs/setup_docs.jl

@@ -1,0 +1,20 @@
+setup_docs = quote
+
+    using Pkg
+    Pkg.activate("docs")
+
+    using AlgorithmicRecourseDynamics
+    using AlgorithmicRecourseDynamics.Data
+    using AlgorithmicRecourseDynamics.Experiments
+    using AlgorithmicRecourseDynamics.Models
+    using AlgorithmicRecourseDynamics: run!
+    using CounterfactualExplanations
+    using Flux
+    using MLJBase
+    using Plots
+    using Random
+
+    Random.seed!(2023)
+    theme(:wong)
+
+end

### 2022-12-23 src/experiments/functions.jl

@@ -10,7 +10,7 @@ using StatsBase
     n_rounds::Int = 10
     n_folds::Int = 5
     seed::Union{Nothing,Int} = nothing
-    T::Int = 1000
+    T::Int = 100
     μ::AbstractFloat = 0.05
     intersect_::Bool = true
     convergence::Symbol = :threshold_only

### 2022-12-23 docs/src/_intro.qmd

@@
 
-### Data and Model
+Below we first generate some synthetic data for a binary classification task and instantiate an instance of `CounterfactualData`.
 
@@
 
+We then define a simple model for the task and prepare it for use with `CounterfactualExplanations.jl`:
+
 ```{julia}
@@
 mod = FluxModel(model)
+```
 
+To generate algorithmic recourse, we will use a simple generic generator:
+
+```{julia}
 generator = GenericGenerator()
 ```
+
+Finally we train our model on a subset of the data. The chart below shows the results.
 
@@
 
+To model the dynamics of algorithmic recourse, we use simulations, in which we repeatedly select as subset of individuals from the non-target class, generate and implement recourse for all of them and finally retrain the model.
+To set this experiment up, we can use the code below:
+
 ```{julia}
@@
 
+Finally, we just run the experiment using default parameter settings that specify the number of rounds, the proportion of individuals to select for recourse and related aspects:
+
 ```{julia}
@@
 ```
+
+The chart below shows the data and predictions at the end of the simulation:
 
@@
 The package was developed for a research project that investigates the dynamics of various counterfactual generators.
+You can find the details [here](https://github.com/pat-alt/endogenous-macrodynamics-in-algorithmic-recourse).

### 2023-02-23 README.qmd
updated domain

@@
 
-[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl/stable) 
-[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl/dev) 
-[![Build Status](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml?query=branch%3Amain)
-[![Coverage](https://codecov.io/gh/pat-alt/AlgorithmicRecourseDynamics.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/pat-alt/AlgorithmicRecourseDynamics.jl) 
+[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://juliatrustworthyai.github.io/AlgorithmicRecourseDynamics.jl/stable) 
+[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://juliatrustworthyai.github.io/AlgorithmicRecourseDynamics.jl/dev) 
+[![Build Status](https://github.com/juliatrustworthyai/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/juliatrustworthyai/AlgorithmicRecourseDynamics.jl/actions/workflows/CI.yml?query=branch%3Amain)
+[![Coverage](https://codecov.io/gh/juliatrustworthyai/AlgorithmicRecourseDynamics.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/juliatrustworthyai/AlgorithmicRecourseDynamics.jl) 
 [![Code Style: Blue](https://img.shields.io/badge/code%20style-blue-4495d1.svg)](https://github.com/invenia/BlueStyle) 

### 2023-02-23 docs/src/_intro.qmd
updated domain

@@
 
-`AlgorithmicRecourseDynamics.jl` is a small package for modeling Algorithmic Recourse Dynamics. It builds on `CounterfactualExplanations`, a package for generating counterfactual explanations.
+`AlgorithmicRecourseDynamics.jl` is a small package for modelling Algorithmic Recourse Dynamics that was developed as part of a research paper by @altmeyer2023endogenous.
+It builds on `CounterfactualExplanations`, a package for generating counterfactual explanations.
 
-## Basic Usage
+## 🔍 Basic Usage
 
@@
 
-## Related Research Paper 📝
+## 📝 Related Research Paper
 
-The package was developed for a research project that investigates the dynamics of various counterfactual generators.
+The research paper by @altmeyer2023endogenous investigates the dynamics of various counterfactual generators.
 You can find the details [here](https://github.com/pat-alt/endogenous-macrodynamics-in-algorithmic-recourse).
+
+## 🎓 References

### 2023-02-23 docs/make.jl

@@ -11,11 +11,11 @@ DocMeta.setdocmeta!(AlgorithmicRecourseDynamics, :DocTestSetup, ex_meta; recursi
 makedocs(;
     modules = [AlgorithmicRecourseDynamics],
     authors = "Patrick Altmeyer",
-    repo = "https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/{commit}{path}#{line}",
+    repo = "https://github.com/juliatrustworthyai/AlgorithmicRecourseDynamics.jl/blob/{commit}{path}#{line}",
     sitename = "AlgorithmicRecourseDynamics.jl",
     format = Documenter.HTML(;
         prettyurls = get(ENV, "CI", "false") == "true",
-        canonical = "https://pat-alt.github.io/AlgorithmicRecourseDynamics.jl",
+        canonical = "https://juliatrustworthyai.github.io/AlgorithmicRecourseDynamics.jl",
         edit_link = "main",
         assets = String[],
     ),
@@ -25,4 +25,4 @@ makedocs(;
     ],
 )
 
-deploydocs(; repo = "github.com/pat-alt/AlgorithmicRecourseDynamics.jl", devbranch = "main")
+deploydocs(; repo = "github.com/juliatrustworthyai/AlgorithmicRecourseDynamics.jl", devbranch = "main")

### 2023-03-10 dev/_intro.qmd
build based on f58d905

--- 2023-02-23 docs/src/_intro.qmd
@@
 data_train, data_test = Data.train_test_split(counterfactual_data)
-Models.train(mod, data_train; n_epochs=n_epochs)
+CounterfactualExplanations.Models.train(mod, data_train; n_epochs=n_epochs)
 plt_original = plot(mod, counterfactual_data; zoom=0, colorbar=false)
